### 2013-10-18 tex/MSThesis.tex
initial commit

%\documentclass[11pt]{amsart}
\documentclass{au}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% For completion by editorial office:
 \presentedby{\dots}
 \received{\dots}{\dots}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage[colorlinks=true,urlcolor=black,linkcolor=black,citecolor=black]{hyperref}

\usepackage{amsmath,amssymb,url,mathrsfs}
\usepackage{enumerate}
\usepackage{tikz}
\usetikzlibrary{calc}
\usepackage{scalefnt}
\numberwithin{equation}{section}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{notation}[theorem]{Notation}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}[theorem]{Example}
\newtheorem*{remarks}{Remarks}

\newcommand{\todo}[1]{~\\[5pt]{\bf TODO(wjd):} #1\\[5pt]}
%\newcommand{\suchthat}{\ensuremath{:}}    % (old notation)
\newcommand{\suchthat}{\ensuremath{\mid}}  % (new notation)
%\newcommand{\defeq}{\ensuremath{:=}}      % (old notation)
\newcommand{\defeq}{\ensuremath{=}}        % (new notation)
\newcommand{\betai}{\ensuremath{\beta^i}}  % \beta^{\bB_i}   % (old notation)
\newcommand{\betaj}{\ensuremath{\beta^j}}  % \beta^{\bB_j}
\newcommand{\betak}{\ensuremath{\beta^k}}  % \beta^{\bB_k} or \beta^{\bB_{k}}
\newcommand{\betal}{\ensuremath{\beta^\ell}}  %\beta^{\bB_\ell} or \beta^{\bB_{\ell}}
\newcommand{\uacalc}{{\small UAC}alc}
\newcommand{\Tuma}{T\r{u}ma}
\newcommand{\Palfy}{P\'alfy}
\newcommand{\Pudlak}{Pudl\'ak}
\newcommand{\PP}{P\'alfy-Pudl\'ak}
\newcommand{\PAP}{P\'alfy\ and Pudl\'ak}
\newcommand{\<}{\ensuremath{\langle}}
\renewcommand{\>}{\ensuremath{\rangle}}
\newcommand{\beps}{\ensuremath{\boldsymbol{\varepsilon}}}
\newcommand{\bN}{\ensuremath{\mathbf{N}}}
\newcommand{\bA}{\ensuremath{\mathbf{A}}}
\newcommand{\bB}{\ensuremath{\mathbf{B}}}
\newcommand{\bBi}{\ensuremath{\mathbf{B}_i}}
\newcommand{\sB}{\ensuremath{\mathcal{B}}}
\newcommand{\sC}{\ensuremath{\mathcal{C}}}
\newcommand{\sS}{\ensuremath{\mathscr{S}}}
\newcommand{\sT}{\ensuremath{\mathscr{T}}}
\newcommand{\sI}{\ensuremath{\mathcal{I}}}
\newcommand{\sE}{\ensuremath{\mathcal{E}}}
\newcommand{\sO}{\ensuremath{\mathcal{O}}}
%\newcommand{\id}{\ensuremath{\mathrm{id}}}
\DeclareMathOperator{\id}{id}
%\newcommand{\Eq}{\ensuremath{\mathrm{Eq}}}
\DeclareMathOperator{\Eq}{Eq}
%\newcommand{\Cg}{\ensuremath{\mathrm{Cg}}}
\DeclareMathOperator{\Cg}{Cg}
%\newcommand{\Con}{\ensuremath{\mathrm{Con\,}}}
\DeclareMathOperator{\Con}{Con}
%\newcommand{\Sub}{\ensuremath{\mathrm{Sub}}}
\DeclareMathOperator{\Sub}{Sub}
%\newcommand{\Pol}{\ensuremath{\mathrm{Pol}}}
\DeclareMathOperator{\Pol}{Pol}
%\newcommand{\Clo}{\ensuremath{\mathrm{Clo}}}
\DeclareMathOperator{\Clo}{Clo}
\newcommand{\ps}[1]{\ensuremath{^{(#1)}}}
\newcommand{\piB}{\ensuremath{\pi_B}}
\newcommand{\hpsi}{\ensuremath{\widehat{\psi}}}
\newcommand{\htheta}{\ensuremath{\widehat{\theta}}}
\newcommand{\supi}{\ensuremath{^{i}}}
\newcommand{\supj}{\ensuremath{^{j}}}
\renewcommand{\leq}{\ensuremath{\leqslant}}
\renewcommand{\nleq}{\ensuremath{\nleqslant}}
\renewcommand{\geq}{\ensuremath{\geqslant}}
\newcommand{\meet}{\ensuremath{\wedge}}
\newcommand{\join}{\ensuremath{\vee}}
\newcommand{\Meet}{\ensuremath{\bigwedge}}
\renewcommand{\Join}{\ensuremath{\bigvee}}
\newcommand{\resB}{\ensuremath{|_{_B}}}
\newcommand{\resBi}{\ensuremath{|_{_{B_i}}}}
\newcommand{\eps}{\ensuremath{\varepsilon}}
\newcommand{\hatmap}{\ensuremath{\widehat{\phantom{x}}}}
\newcommand{\one}{\ensuremath{\mathbf{1}}}
\newcommand{\two}{\ensuremath{\mathbf{2}}}
\newcommand{\three}{\ensuremath{\mathbf{3}}}
\newcommand{\tbeta}{\ensuremath{\widetilde{\beta}}}
\newcommand{\ttheta}{\ensuremath{\widetilde{\theta}}}
\newcommand{\hbeta}{\ensuremath{\widehat{\beta}}}
\newcommand{\cick}{\ensuremath{\sC_i}}
%\newcommand{\CE}{\ensuremath{\sC_r^{\sE}}}
\newcommand{\CE}{\ensuremath{\sB_r^n}}
\newcommand{\CO}{\ensuremath{\sC_r^{\sO}}}
\newcommand{\CICK}{\ensuremath{c_i/\beta \cup c^{K+1}_i/\beta^{\bB_{K+1}}}}
\newcommand{\GAP}{{\small GAP}}
\newcommand{\dotsize}{.8pt}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%  FRONT MATTER                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title[Approximating Eigenvalues of Large Stochastic Matrices]{A Lanczos Procedure for Approximating Eigenvalues of Large Stochastic Matrices}

%% First author (Note: The order of the items here is important!)
\author{William DeMeo}
\email{williamdemeo@gmail.com}
\urladdr{http://www.math.sc.edu/~demeow}
\address{Department of Mathematics\\
University of South Carolina\\Columbia 29208\\USA}

\thanks{The author wishes to thank a number of people who contributed to this
  work.  In particular, the primary referee made many insightful recommendations
  which led vast improvements, not only in the overall exposition, but also in
  the specific algebraic constructions, theorem statements, and proofs.
  This article is dedicated to Ralph Freese and Bill Lampe, who declined to be
  named as co-authors despite their substantial contributions to the project.}

%% Dedication (Optional)
\dedicatory{This article is dedicated to Ralph Freese and Bill
  Lampe.}

%% AMS subject classification; see http://www.ams.org/msc
%% Only one Primary. Possibly several Secondary.
\subjclass[2010]{Primary: 08A30; Secondary: 08A60, 06B10.} %, 06B15.}

%% Keywords and phrases
\keywords{congruence lattice, finite algebra, finite lattice
  representations}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%  MAIN MATTER                                                  %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstract}
The rate at which a Markov chain converges to a given probability distribution
has long been an active area of research. Well known bounds on this rate of
convergence involve the subdominant eigenvalue of the chain’s underlying
transition probability matrix. However, many transition probability matrices are
so large that we are unable to store even a vector of the matrix in fast
computer memory. Thus, traditional methods for approximating eigenvalues are
rendered useless.

In this paper we demonstrate that, if the Markov chain is reversible, and we
understand the structure of the chain, we can derive the coefficients of the
traditional Lanczos algorithm without storing a single vector. We do this by
considering the variational properties of observables on the chain’s state
space. In the process we present the classical theory which relates the
information contained in the Lanczos coefficients to the eigenvalues of the
Markov chain. 
\end{abstract}

\maketitle

%%%%%%%%%%%%%%%%%%%%%
\chapter{Introduction}
The rate at which a Markov chain converges to a given probability distribution has long
been an active area of research. This is not surprising considering this problem’s relevance to the ar-
eas of statistics, statistical mechanics, and computer science. Markov Chain Monte Carlo (MCMC)
algorithms provide important examples. These algorithms come in handy when we encounter a
complicated probability distribution from which we want to draw random samples. In statistical
mechanics, we might wish to estimate the phase average of a function on the state space. Goodman
and Sokal [6] examine Monte Carlo methods in this context. Examples from statistics occur in the
Bayesian paradigm when we are forced to simulate an unwieldy posterior distribution (see, e.g.,
Geman and Geman

To implement the MCMC algorithm, we invent a Markov chain that converges to the
desired distribution (this is often accomplished using the Metropolis algorithm
described in Chapter 5). Realizations of the chain will eventually represent
samples from this distribution. Sometimes ``eventually'' -- meaning all but
finitely many terms of the chain -- is just not enough. We need more practical
results. In particular, we want to know how many terms of the chain should be 
discarded before we are sampling from a distribution that is close (in total variation distance) to
% 03.txt
the distribution of interest. This is the purpose of bounding convergence rates for Markov chains.

Often the Markov chains encountered in this context satisfy a condition known in the
physics literature as detailed balance. Probabilists call chains with this property reversible. This
simply means that the chain has the same probability law whether time moves
forward or 
backward.\footnote{This is not a precise definition. In particular the chain must
  have started from its stationary distribution. Full rigor is postponed until Section 2.1.}
In this paper, we consider the rate at which such chains converge to a \emph{stationary distribution}.\footnote{This and other italicized terms are defined in Section 2.1.}

There are a number of different methods in common use for bounding convergence rates of
Markov chains, and a good review of these methods with many references can be found in  More
recently developed methods, employing logarithmic Sobolev inequalities, are reviewed in  Most
of the bounds in common use involve the sub—dominant eigenvalue of the Markov chain's transition
probability matrix, and thus require good approximations to such eigenvalues. In many applications,
however, the transition probability matrix is so large that it becomes impossible to store even a
single vector of the matrix in conventional computer memory. These so called out-of-core problems
are not amenable to traditional eigenvalue algorithms\footnote{By ``traditional
  eigenvalue algorithms'' we refer to those found, for example, in Golub and Van
  Loan[5]. See also the book by Demmel [1] for a more recent discussion.}
without modification. This paper develops 
such a modification for the Markov chain eigenvalue problem. In particular it develops a method
for approximating the first few eigenvalues of a transition probability matrix when we know the
general structure of the underlying Markov chain. The method does not require storage of large
matrices or vectors. Instead we need only simulate the Markov chain, and conduct a statistical
analysis of the simulation.

Here is a look at what follows. Section 2.1 contains a review of the relevant Markov chain
theory. Readers conversant in the asymptotic theory of Markov chains might wish to at least skim
Section 2.1, if only to become familiar with our notation. Section 2.2 describes functions on the state
% 04.txt
space of the Markov process. This section and Chapter 3 develop the context in which we formulate
the new ideas of the paper. In the last section of Chapter 3, Section 3.3, we present the familiar
Krylov subspace and explain why this represents our best approximation to a subspace containing
extremal eigenvectors of the transition probability matrix.\footnote{or, more
  precisely, a similarity transformation of this matrix.} The first section of
Chapter 4 describes the \emph{Lanczos algorithm} for generating an orthonormal basis
for the Krylov subspace. As it stands, this algorithm is useless for an
out-of-core problem such as ours since, by definition of such problems, 
it requires too much data movement; all the computing time is spent swapping data between slow
and fast memory (e.g. between the hard disk and cache). Therefore, we discuss alternatives to
Lanczos and demonstrate that the \emph{Lanczos coeficients} are readily available through simulations of
the Markov chain, which fact allows us to avoid the standard algorithm altogether. Following this
is a chapter describing the Metropolis algorithm used to produce a reversible stochastic matrix.
It is here that we experiment with the procedure described in Section 4.2 and approximate the
extremal eigenvalues of the matrix, without storing any of its vectors. Finally, Chapter 6 concludes
the paper.
% 05.txt

\chapter{Markov Chains}

\section{General Theory}

This review of Markov chain theory can be found in any good probability text. The present
discussion is most similar to that of Durrett [3], to which we refer the reader desiring greater detail.

\subsection{The Basic Setup}

Heuristically, a Markov chain is a stochastic process with a lack of memory property. Here
this means that the future of the process, given its past behavior and its present state, depends only
on its present state. This is the probabilistic analogue of a familiar property of classical particle
systems. Given the position and velocities of all particles at time t, the equations of motion can be
completely solved for the future evolution of the system. Thus, information describing the behavior
of the process prior to time t is superﬂuous. To be a bit more precise, if technical, we need the
following definitions.

\bibliographystyle{spmpsci}
\bibliography{wjd}

\end{document}

%%%%%%%%%%%%%%%%%%%%%  END DOCUMENT %%%%%%%%%%%%%%%%%%%%%%%%%

Definition 2.1.1 Let (S, 8) be a measurable space. A sequence X,., n 2 0, of random variables

taking values in S is said to be a Markov chain with respect to the filtration a(Xo, . . . ,X,,) if for
% 06.txt
a.llBES,

p(x,,+1 e B | a(Xo, . . . ,X,.)) = P(X,.+1 e B 1 o(X,,)) (2.1)

Equation (2.1) merely states that, given the present location of X,,, the past is irrelevant for

predicting the location of X,,+1.

Definition 2.1.2 A function p : S x S —> R is said to be a transition probability if:
1. for each :1: E S, A —> p(a:, A) is a probability measure on (S, S).
2. for each A E 8, :1: —> p(z, A) is a measurable function.

We call X,, a Markov chain with transition probabilities 12,, if
P(Xn+l E B l a(Xn)) =Pn(XnaB) (22)

The spaces (S, S) that we encounter below are standard Borel spaces, so the existence of the
transition probabilities follows from the existence of regular conditional probabilities on Borel spaces
— a standard measure theory result (see e.g. [3] page 230).

Suppose we are given an initial probability distribution p on (S, S) and a sequence p,, of

transition probabilities. We can define a consistent set of finite dimensional distributions by

P<X,- e B,-,0 s 1 s n) = / #(d1‘o) / Po($o.d-'51) . .. / Pn-l(97n—1ad$n) (2.3)
B0 B1 B1:
Furthermore, denote our probability space by
(9,?) = (S{0,1,...},8{0,l,...})

We call this sequence space and it is defined more explicitly by

sf°»1=-~-} = {(wo,w1,...) :w.- e S}

S{°*1""} = ¢7(w :40; E A,- E 3)

% 07.txt
The Markov chain that we will study on this space is simply X" (w) = w,., the coordinate maps.
Then, by the Kolmogorov extension theorem, there exists a unique probability measure P,, on
(9,?) so that the X" (w) have finite dimensional distributions (2.3).

If instead of p, we begin with the initial distribution 6,, i.e. point mass at 2:, then we
denote the probability measure by P3. With such measures defined for each 2:, we can in turn

define distributions P,,, given any initial distribution p, by

PM = / u<da.->P.(A>

That the foregoing construction — which, recall, was derived merely from an initial distri-
bution p and a sequence p,, of transition probabilities — satisfies Definition (2.2) of a Markov chain
is not obvious, and a proof can be found in

To state the converse of the foregoing, if X,, is a Markov chain with transition probabilities
p,, and initial distribution p, then its finite dimensional distributions are given by (2.3). Proof of
this is also found in

Now that we have put the theory on a firm, if abstract, foundation, we can bring the
discussion down to earth by making the forgoing a little more concrete. First, we specialize our
study of Markov chains by assuming that our chain is temporally homogeneous, which means that
the transition probabilities do not depend on time; i.e. p,,(w,,, B) E p(w,., B). (This is the stochastic
analogue of a conservative system.) Next we assume that our state space S is finite, and suppose
for all states 1', j E S that p(i, j) Z 0, and E, p(z', j) = 1 for all i. In this case, equation (2.2) takes

a more intuitive form:
P(Xn+l =j I X7: = 7:) =P(iaj)

and our transition probabilities become
% 08.txt
If P is a matrix whose (i, j)th element is the transition probability p(i, 1') then P is a stochastic

matriar, that is, a matrix with elements pgj satisfying

Pij 20: zjpij =1: (i7j=1121"‘)d)

We also refer to P as the transition probability matrix.

Without loss of generality, we can further suppose our Markov chain is irreducible. This
means that, for any states 2', 1‘, starting in state i the chain will make a transition to state 1' at
some future time with positive probability. This state of affairs is often described by saying that
all states communicate. We lose no generality with this assumption because any reducible Markov
chain can be factored into irreducible classes of states which can each be studied separately.

The final two conditions we place on the Markov chains considered below will cost us
some generality. Nonetheless, there remain many examples of chains meeting these conditions,
thus making the present study worthwhile. Furthermore, it may be the case that, with a little
more work, we will be able to drop these conditions in future studies. The first condition is that
the chain is aperiodic. If we let I, = {n 2 1 : p"(:c,a:) > 0}, we call a Markov chain aperiodic if,
for any state 2:, the greatest common divisor of I, is 1. The second assumption is that our chain is

reversible. This characterization is understood in terms of the following definition.
Definition 2.1.3 A measure a is called reversible if it satisfies

;4(w)p(I. y) = u(y)p(y. w)

for all :c and y.

We call a Markov chain reversible if its stationary distribution (defined in Section 2.1.2) is reversible.
% 09.txt
\subsection{A Convergence Theorem}

In succeeding arguments, we use some results concerning the asymptotic behavior of

Markov diains. These results require a few more definitions.

Definition 2.1.4 A measure 1r is said to be a stationary measure if

Z«(z)p<z.y> = we) (2.4)

2
Equation 2.4 says P,,(X1 = y) = 1r(y), and by induction that P,,(X,, = y) = 7r(y) for all n 2 1. If 1r
is a probability measure, then we call 1r a stationary distribution. It represents an equilibrium for
the chain in the sense that, if X0 has distribution 1r, then so does X,. for all 11..

When the Markov chain is irreducible and aperiodic, the distribution of the state at time
n converges pointwise to 1r as n —> oo, regardless of the initial state. It is convenient to state
this convergence result in terms of the Markov chain's transition probability matrix P. Before
doing so, we note that irreducibility of a Markov chain is equivalent to irreducibility (in the usual
matrix theory sense) of its transition probability matrix. Furthermore, it turns out that a transition
probability matrix of an aperiodic Markov chain falls into that class of matrices often called acyclic,
but for simplicity we will call such stochastic matrices aperiodic. With this terminology, we can

state the convergence theorem in terms of the transition probability matrix P.

% Theorem 2.1.1
\begin{theorem}
\label{thm-2.1.1}
Suppose P is irreducible, aperiodic, and has stationary distribution 7r. Then as n —> oo,
p"(i,J') -> #0")
\end{theorem}

The symbol p"(i, j) is to be understood as the (ij) element of the nth power of P.
A Markov chain whose transition probability matrix satisfies the hypotheses of
Theorem~\ref{thm-2.1.1} is called ergodic. If we simulate an ergodic chain for
sufficiently many steps, having 
% 10.txt
begun in any initial state, the final state is a sample point from a distribution which is close to 1r.

To make this statement more precise requires that we define “close”.

Definition 2.1.5 Let 1r be a probability measure on S. Then the total variation distance at time

n with initial state I is given by
AA") = l|P"(z, A) - 7r(-4)||Tv = 1333; I P"(I.A) - 7r(A) I

In what follows, we will measure rate of convergence using the function 'r,(e), the first time after

which the total variation distance is always less than 5. That is,
-r,,(e) = min{m : A,(n) 3 E for all n. 2 m}

To consider the connection between convergence rates and eigenvalues, note that an ape-
riodic stochastic matrix P has real eigenvalues 1 = /\o > A1 2 A2 3  Z /\¢_1 2 -1, where
d = |S'] is the dimension of the state space. For an ergodic chain, the rate of convergence to the
stationary distribution 7r is bounded by a function of the subdominant eigenvalue. By subdominant
eigenvalue we mean that eigenvalue which is second-largest in absolute value, and we denote this
Am“ E max{).1,|A¢_1|}. The function bounding the rate of convergence is given by the following

theorem:

% Theorem 2.1.2
\begin{theorem}
\label{thm-2.1.2}
The quantity T1(€) satisfies
1. 13(5) 5 (1 — /\m,_x)“(ln1r(a:)‘1 +1ne‘1)
2. maxxes r,(e) 2 %/\,,,,x(1 — ).,,,,x)‘1ln(2e)’1
\end{theorem}

As this theorem shows, if we have an upper bound on the subdominant eigenvalue, then we have
an upper bound on the function 11(5). In what follows, we will derive an approximation to the
% 11.txt
subdominant eigenvalue and supply error bounds. Together, an approximation and error bounds
for Am“ provide enough information to make Theorem~\ref{thm-2.1.2} useful.

\section{Functions on the State Space}

Recall that X,.(w) = w,. 6 S denotes the state in which the Markov chain exists at time
11. Suppose that 11> = {¢1, . . . , ¢,,} is a collection of p observables, or functions defined on the state
space S. Furthermore, let these observables be real valued, so ¢.~ : S —> R. It is often useful, as we
will see below, to assume that none of the observables is a constant function. Suppose now that
the state space S is finite with at possible states. Then, since an observable is simply a map of the
state space, we can think of each ¢,- as a vector of d real numbers — the d values that it takes on at
the different states.

Now assume the Markov chain is irreducible, and let 1r denote its stationary distribution.
If we start the chain from its stationary distribution - i.e. suppose X0 has distribution 1r - then
X,. is a stationary process. Furthermore, for each i, ¢,(X,.) is a stationary stochastic process with

mean

E.¢.- = Z) vr(x)¢.(z)

165

and autocovariance function

Cir (¢i(Xn)v ¢i (Xn+a))

E7!’  ‘ E1r¢i) (¢i(Xn+a) " E1r¢i)l

Z Pn(Xn = z1X7I+S = y)  — E1r¢i)  _ E1r¢i)

z,yES
(2.5)

which, by the definition of conditional probability, we can write as,

2 P«(X.. = z)P.(X..+.. = y I X. = 1)(¢s($) — Em) (my) ~ Em)
1.1165
% 12.txt
Z 7r($)P1y (MI) - Em.-) (¢:(1I) - E1r¢i)

:c,y€S

Here pgy denotes the element in row :1: and column y of P’, the sth power of the transition probability

matrix. Similarly, we define the cross-covariance between the function ¢,- at time n and q5,- at time

n+sas

E1!’  _ E7r¢:') (¢i(Xn+a) " E1r¢i)l (2-6)

C1r(¢i(Xn)a (#1 (Xn+a))

Z) vr(x)p;,, (¢.-(w) — Em.-) (my) — Em)

z,yES

Now let (<I>) denote the matrix of mean vectors whose jth column is E,,¢,-1, where 1 = (1, . . . , 1)‘,
and let H = diag(1r(w1), . . . , 1r(wd)), be the d X d diagonal matrix with stationary probabilities 1r(w)
on the main diagonal, and zeros elsewhere. Finally, denoting by C(s) the p X p covariance matrix

whose (i,j)th element is C, (qfi,-(X,.),¢_,< (X,.+,)), we have

0(0) = E(‘1’(Xn) - (‘I>))(‘I’(Xn) - (‘PW
= (‘P-(‘1’))‘1'1(<1’-(‘I>))
0(8) = E(<I>(Xn) — (‘I>))(‘1’(Xn+a) - (‘1’))’

(‘P - (‘1’))'1'IP‘(<1> — (<1>))

Below, we will also find it useful to have at our disposal a new matrix that is similar to the
transition probability matrix. We have in mind the matrix M E Ill/2PI'I‘1/2. As is easily verified,
this allows us to write the covariance matrix as

can = <<I>—<<I>>)°n%*M*n%<<I>—<<I>>>

\II‘M’\II (2.7)

where we have defined ‘II E 111/ 2(<I> —  Recall that our main motivation is that, for out-of-core
problems, traditional eigenvalue algorithms are inadequate. With this fact, and the form (2.7) in
% 13.txt
mind, we will consider using the covariance of observables on the state space to implement the
(otherwise useless) Rayleigh-Ritz procedure, which we describe below. This procedure requires
that M be symmetric. As the next fact demonstrates, this required symmetry is the reason for our

interest in the reversibility property of the Markov chain.

Fact 2.2.1 The matrix M is symmetric if and only if the Markov chain is reversible (i.e. iﬂ the

process satisfies the detailed balance condition).

Proof:

M is symmetric

4:, (1-I1/2P1-I-1/2): = H1/2Pn—1/2
<.—> r1-%*P°n%* = H1/2911-1/2
¢> 1>'r1= = up
~:=> (IIP)‘ = l'IP

Element-wise, the final equality implies that 7|‘,'p,'j = 1rjpj.-. According to Definition 2.1.3, this states

that p is a reversible measure.

% 14.txt

\chapter{Invariant and Approximate Invariant Subspaces}

\section{Invariant Subspaces}
Definition 3.1.1 A subspace S Q R" with the property that

1:ES'=Ma:€S

is said to be invariant for M.

Recall that, having chosen observables <I> = ((151, . . . , qfip), we constructed the covariance
matrix C(s) = \II‘M"\II. If the column space of \II, which we denote by ran(\II), is an invariant
subspace for M, the definition implies Mifij E  That is, for each 111,- there exists a vector
t of coeﬂicients such that M1/Jj = f=1t,-1/)5,-. This is true for all j and, putting each vector of
coefficients into a matrix T, we see that M\II = \IIT. Conversely, M\II = IIIT implies that M1/Jj is a

linear combination of columns of \II, so ran(\II) is invariant. We have thus proved the following
% 15.txt
Fact 3.1.1 The subspace ran(\II) is invariant for M if and only if there exists T E R”? such that

M\II = \IIT.

Consequently,

Fact 3.1.2 )\(T) Q ).(M)

Proof of 3.1.2:

z\E)\(T) _=_ 3v€R"3Tv=).v
<=> \IITv = A\IIv
<=> M\II-u = /\\IIv
< /\ E /\(M)

The second ¢> follows from fact 3.1.1.
Facts 3.1.1 and 3.1.2 are theoretically useful. To see why, write the equation of fact 3.1.1

as follows:

\IIT = M\II
\II‘\IIT = \II‘M\II
T = (\1:'~1r)-‘\1:'M\1:

In this form, we recognize that T = C‘1(0)C(1). That is, using only the covariance of observables
on the state space, we can generate a matrix T with the property A(T) (_Z ).(M). Recalling that
M = H1/2Pl'I‘1/2, we see that M is similar to our original stochastic matrix P, and thus ).(M) =

).(P).

\section{Approximate Invariant Subspace}

We qualiﬁed the foregoing by stating that the facts are only theoretically useful. This

is because when choosing observables of the state space we aren't sure that they will satisfy the
% 16.txt
primary assumption underlying facts 3.1.1 and 3.1.2. Recall the assumption: ran(\II) is an invariant
subspace for M, where ‘II = H1/2(<I> —  It may well be the case that there is a 1p; 6 ran(\II) such
that M1/zj ¢ ran(\II), thereby violating the assumption. However, lacking an invariant subspace, we

might hope that our observables at least provide an approximate invariant subspace.

Deﬁnition 3.2.1 If the columns of ‘II E R“? are independent and the residual matrix E E

M\II —— \IIT is small for some T 6 R7”, then ran(\II) deﬁnes an approximate invariant subspace.

To see how an approximate invariant subspace can be useful in approximating eigenvalues of M,
we extract a theorem from Golub and Van Loan 
% Theorem 3.2.1
\begin{theorem}
\label{thm-3.2.1}  
Suppose M 6 Rd” and T 6 RP"? are symmetric and let E E MQ — QT, where Q 6 Rd"? satisﬁes

Q‘Q = I. Then there exist p1,... ,;z,, E A(T) and A1,. . . , A,, E A(M) such that
ll‘k - Akl S \/§|lE"

fork=1,...,p.
\end{theorem}

If the subspace ran(Q) is an approximate invariant subspace, then the deﬁnition implies that
there is a choice T rendering the error IIEII2 small, and thus, the eigenvalues of T provide a good
approximation to those of M.

When considering the preceding ideas, it becomes apparent that their application will
present new — but hopefully less prohibitive — problems. As these problems summarize the rt of
the paper, now is a good time to examine them. First, the preceding theorem assumes a matrix Q
whose columns form an orthonormal basis for the approximate invariant subspace. For our problem,
derivation of such a Q is tricky, and we must prepare for this. Next, having an approximate invariant

subspace at our disposal merely tells us that there exists some matrix T which makes the error
% 17.txt
"EH2 small. We must discover the form of such a T. Moreover, it is natural to seek that T which
minimizes IIEII2 for a given approximate invariant subspace. Finally, in order to apply these ideas
to real examples of our eigenvalue problem, we must ﬁnd a practical way to generate the most
appropriate approximate invariant subspace. We now address each of these issues in the order
raised.

Recall the matrix I1 E III/2(<I> — (<I>)). The columns of this matrix, though independent
(by choice of independent observables), are not necessarily orthonormal. However, consider the
polar decomposition ‘II = QZ where Q‘Q = I and Z2 = \Il‘\II is symmetric positive semideﬁnite.1

Note that Z = (\II‘\II)l/2 is nonsingular, so Q has the form
Q = ~1:z-1 = \II(\II‘\II)“/2

and it is clear that ran(Q) = ran(\II). Perhaps ra.n(Q) is a. useful approximation to the invariant
subspace for M. If ran(Q) is not itself invariant, we have the error E = MQ — QT. It is easy to
show, as we do in the following proof, that the T which minimizes HE“; is T = Q‘MQ. This yields

the following,
Theorem 3.2.2

If \I1 = QZ is the polar decomposition of \II, then the matrix
T = Q‘MQ = (\1:*\1:)-1/’\1:‘M\I:(\Ir‘\I:)-1/2
minimizes IIEII2 = IIMQ - QTII2

Proof:
We prove the result by establishing the following

(Hal: Given M 6 Rd” suppose Q 6 Rd"? satisﬁes Q'Q =1. Then,

min ||MQ — QTII2 = H(1- QQ')MQll2

TERI‘?

‘Recall that the polar decomposition is derived from the SVD, \II = UZJV‘, by letting Q = UV‘ and Z = VEV‘.

% 18.txt
and T = Q’MQ is the minimizer.
To verify the claim, observe the following application of the Pythagorean theorem:

For any T E R.”"” we have

IIMQ — QT||§ = IIMQ - QQ'MQ + QQ'MQ - QT|l§
||(I - QQ’)MQ + Q(Q'MQ - T) H3

||(1 - QQ‘)MQ||§ + |lQ(Q'MQ - T)||§ (3-1)

2 l|(I — QQ')MQ||§

Equality (3.1) holds since I — QQ' projects MQ onto the subspace orthogonal to ran(Q). Thus,
the two terms in the expression on the right are orthogonal, and the Pythagorean theorem yields
equality. The concluding inequality establishes that the minimizing T is that which annihilates the
second term in (3.1); i.e. T = Q‘MQ.

The second equality in Theorem 3.2.2 is a consequence of the polar decomposition, in

which Q = \IIZ“ = \I:(~I:‘\I:)-1/2. Therefore,
T = (x1z'x1:)-1/2x1:‘M~1:(\1:*~1z)—1/2

is the minimizer.

3.3 The Krylov Subspace

Suppose the columns of a matrix Q E R“? provide an orthonormal basis for an approxi-

mate invariant subspace. Then we have seen,
1. T = Q'MQ minimizes ||E||2 = ||MQ — QT||2 and
2. E|p.1,...,p,, E z\(T) and /\1,...,A, 6 »\(M) such that

ll-‘I: - Ml S \/§|lE||2

% 19.txt
19
fork=1,...,p.

Given an approximate invariant subspace ran(Q) of dimension 1), these facts tell us what matrix we
should use to approximate p elements of M’s spectrum. Now all we lack is a. description of ran(Q).
That is, we have not speciﬁed which approximate invariant subspace would best suit our objective

of approximating the subdominant eigenvalue A,,.a2:(M). 'I‘o do so requires the following deﬁnition:

Deﬁnition 3.3.1 The Raleigh quotient of a symmetric matrix M and a nonzero vector 3 is

p(2:, M) E (:z:‘Ma:)/(z‘a:).

We will also denote this by p(z) when the context makes clear what matrix is involved.

To ﬁnd the approximate invariant subspace most appropriate for our problem, we choose
each dimension successively, providing justiﬁcation at each step. Thus, start with one observable
42 on the state space, and let zpl E 111/ 2(¢ — E,,¢). That is, 1/:1 represents a centered (mean zero)
observable whose ith coordinate is weighted by . First notice that the deﬁnition (which comes
directly from the deﬁnition of ‘II following equation (2.7)) is such that 1/)1 is a constant vector if
and only if :15 is oonstant on the state space, in which case 1/); E 0. This fact provides one reason
observables that are constant on the state space are not interesting. Second, recall that the row
sums of the matrix P are all one and, therefore, the eigenvector corresponding to the eigenvalue
Ao(P) = 1 is the constant vector. By deﬁnition of M E 1'11/2Pl'I‘1/2, we see that the constant vector
is also the eigenvector corresponding to A0  This will play an important role in what follows, as
it allows us to focus primarily on the subdominant eigenvalue z\max(M) E max{}q (M), [A44 
rather than on A0  (which we already know is 1).

Now, notice that

lP(¢1,M)l 5 max |P(x,M)l = »\max(M) (3-2)
% 20.txt
where the max is taken over all non-constant vectors. Since our interest centers on An-.3_x(M), we
would like a. 1/11 which makes the left hand side of (3.2) large. This would be achieved if 1/)1 were to
lie in the space spanned by, say, the first two eigenvectors of the Markov chain (sometimes referred
to as the slowest modes of the process). However, this subspace spans only two dimensions of the
entire d-dimensional space, and it is more likely that ¢1, at best, only comes close to lying in the
subspace of interest. Now, given 1/11, a judicious choice for the second dimension l/)2, and hence
W; E [1/)1 1/:2], would be that which makes max,,¢o |p(\II2a)| is as large as possible. To establish that

this is indeed the pertinent objective, note the following:

_ a‘\1:5Mx1:2a
I;1;g<Ip(\I'2a,M)l = 13:3: ————a,W§q,2a
= M
mrggf,mIp(z. )1
5 maxlp(z,M)l (3-3)
= ).,,m(M)

Again, the max on the right side of (3.3) is over non-constant vectors. In words, we wish to chose
\II-2 = [1/:1 ¢2] so that there is a vector a 6 R2 making |p(\II2a,M)| close to )«max(M).

Now, p(1/11) changes most rapidly in the direction of the gradient Vp(¢1).

W0  (iiiiiiv---viiiléi)
= ,,;,l (Mam — p<¢.)«m (3.4)

So, to maximize the left hand side of (3.3), ‘I12 should be chosen so that the subspace ra.n(\II2)

contains the gradient vector. That is, we must choose 1/22 so that

VP(¢'1) E F3l1{¢1y¢2} E F-'%!1(‘I’2) (3-5)

Clearly, equation (3.4) implies Vp(1/:1) E ran{z/11,M1,b1}. Therefore, if ran{1/:1,1/)2} = ra.n{¢1,M¢1},

then (3.5) is satisﬁed.
% 21.txt
In general, having chosen \II,, E [1/)1 - -- 1/11,] so that

ran{'pl9¢21' " 1¢k} = ran{¢1vM1[)la ' ‘ ' vMk—11/’k}

we must chose 1/1;.“ so that, for any non-zero vector u 6 R",

VP(‘I’I:a) 6 ra‘n{¢11 ‘21 - ° - 31/Jk-F1} 
Now,
vp<~I~.a) = gmm — P(‘I’kﬂ)‘I’k“)

and therefore,

Vp(‘II;,a.) E ran(M\II;,) U ra.n(\IIk) = ra.n{¢1, M1/)1, . . . , Mktﬁl}

Thus, requirement (3.6) is met when ran{1/:1, 1/:2, . . . , 1/4+1} = ran{1/)1, M1/:1, . . . , Mk:/11}.

In conclusion, the p-dimensional approximate invariant subspace which is most suitable

for our objective is

/C(Mn/mp) 2 ran{«/:1,M¢1,... ,M"“«/21}

This is known as the K rylov subspace. Therefore, in answer to the problem posed at the outset
of this section, if we take the columns of Q to be an orthonormal basis for IC(M,z/)1, p), then the

eigenvalues of T = Q‘MQ should provide good estimates of p extremal eigenvalues of M.

% 22.txt
\chapter{Lanczos Procedures}

The following will review the usual procedure for generating an orthonormal basis for
lC(M, «/11, p), and conclude by showing that, for our problem, we can find the matrix Q‘MQ without
actually carrying out this procedure. Note that this conclusion is essential since the procedure

requires matrix vector multiplication - an operation we have assumed impossible for large enough :1.

4.1 The General Lanczos Algorithm
Let 21;“) E M"¢ and consider the d x :1 matrix
«I 2 [«/»,M«/»,  ,M"“¢] 2 [«b‘°>,«/»“’,...,¢<‘-‘>1

Notice that M\II = [1/:(1),... ,1/;(d'1),Md¢r] and, assuming for the moment that \I1 is non-singular,

we can compute the vector h = —\II‘1M"1/J. Thus,

M\II = \II[e2,. . . ,e,1, —h] E IIIH

% 23.txt
23

Equivalently,
0 0  0 -—h1
1 0  0 —h2
HE   3 ; =\II“M\II (4.1)
0 "hl1—1
1 —h,;

H is a companion matrix which means that its characteristic polynomial is p(:r) = 2:‘ + 23:1 h,-:r"‘1.
Since H is similar to M, ﬁnding the eigenvalues of M is equivalent to ﬁnding the roots of p(a:).
However, this is of little practical use since ﬁnding h, constructing 12(1), and ﬁnding its roots is
probably a harder problem than the one we started with. Instead, the value of decomposition (4.1)
derives from its upper Hessenberg form. We exploit this property below.

Let ‘II = QR be the QR decomposition of \II. Since ‘I! is assumed nonsingular,
~II;‘M~r.: = (R"Q‘)M(QR) = H
Therefore,
Q‘MQ = RHR-1 2 T

Since R and R“1 are both upper triangular and H is upper Hessenberg, T = R.HR'1 is also upper
Hessenberg. Furthermore, since M is symmetric, it is clear that T‘ = Q‘MQ = T. That is, T is

an upper Hessenberg matrix that is symmetric. Therefore, T is tridiagonal and we can write it as

follows:

T =    (4.2)

ﬁd—1

0  174-1 ad

% 24.txt
24

Equating columns j on both sides of MQ = QT yields
M113‘ = B,-—1qj—1 + 0:91 + ﬁjqj+1 (43)

Since the columns of Q are orthonormal, multiplying both sides of (4.3) by q_,- and q,-+1 yields
oz, = q§Mq,- and ﬂj = q;-+1Mq,-.
The foregoing justifies what is called the Lanczos algorithm, performed as follows:

The Lanczos algorithm for (partial) reduction to symmetric tridiaganal form

91 = ¢/||¢||2, [30 = 0, 40 = 0

for j=1top

z=Mq_,-

5:‘ = ||Z||2
if 5, = 0, quit
¢Ij+1 = 2/51

end for

The qj computed by the Lanczos algorithm are often called the Lanczos vectors. If the
loop in the algorithm is terminated because ﬁg = 0, this indicates that an exact invariant subspace
has been computed, and is given by ran{q1 - - - qk}. Otherwise, we usually halt the algorithm after
p steps, in which case the algorithm converg to approximations of at most 12 eigenvalues.‘

Note that equation (4.3) results from the equation MQ = QT. The latter equation holds
since, starting with d vectors in our subspace, Q is a d Xd matrix whose columns form an orthonormal

basis for all of R‘, which is clearly an invariant “sub” space. The Lanczos algorithm above, however,

‘A detailed discussion of such convergence (and misconvergence) is given in

% 25.txt
25

proceeds for only p steps, producing a d x p matrix Q, whose columns form an orthonormal basis
for the approximate invariant subspace IC(M, ¢, p), and a p X p matrix T, = Q'MQ. In that case,
IIMQ = QTp|[g =  is non-zero and gives the error bound described in Theorem 3.2.1. Now,

writing the full d x :1 matrix T of (4.2) as

01 51
/71

(4.4)

allows us to describe the error bound ||E||-2 in terms of the submatrix Tm, and, hence, in terms of

ﬁzz-
Theorem 4.1.1

HT, and T,“ are the matrices appearing in (4.4), and if the p columns of Q are computed by the

Lanczos algorithm, then there exist pl, . . . ,;t,, E /\(T,) and A1, . . . , A, E A(M) such that

ill» - /\I=i S iiTP'4”2 = 512

fork=1,...,p.

For a proof, see page 365 of [1].

% 26.txt
26

4.2 A Lanczos Procedure for Markov Chains

Notice that the Lanczos algorithm requires the computation of Mqj. Throughout this
paper we have assumed that M 6 Rd“, and that d is so large as to make the matrix vector
multiplication Mqj impossible. Now, the Lanczos algorithm constructs an orthonormal basis for
the Krylov subspace. For our purposes, this construction is not essential. What is essential is that
we ﬁnd a way to generate the Lanczos coejﬁcients oz,-, ﬁj, (j = 1, . . . , p), and whence the matrix
T, without performing the operation Mqj required by the Lanczos algorithm. We now address this
problem.

Begin with the centered and weighted observable 1/1 = 111/ 2 (qS—E,,¢), and let q1 = 11;/||1/JH2.
To simplify notation, let ¢(n) E ¢(X,.). The form of the first Lanczos coeﬂicient a1 is straight

forward:

01 = qiMq1

I/1‘M1/I
¢’¢

and now the special properties of the Markov chain setting play an important role. For, regard the

following

¢‘M«/2 = (¢ — E.¢)‘II%‘MII%(¢ - Em)

Cw (d>(n), ¢(" + 1)) (4-5)

and

W1!’ = (<15 - E«¢)'H(¢ - E«¢)

Varn (¢(")) (4-6)

Recall that the deﬁnition of M implies H%‘MI'I% = HP, and this justiﬁes equation (4.5). Putting

% 27.txt
27

these equations together, we have a. nice form for :11:

_ c, (run), an + 1))
"‘ ' Ta:,(¢(n))

The next Lanczos coeﬂicient, B1, is only slightly more complicated.

[31 = ||M91-0t1tI1||2

[(Mq1 — 0l1¢11)i(M¢]1 - 0101)] 1/2

[t1iM2¢I1 — ail 1/2

(4.7)

Again recalling the deﬁnition M E 111/ 2P1'I‘1/2, a. simple manipulation veriﬁes

1'li‘M’l'[% = HP’, whereby it follows that

(«s — E,,¢)‘HP”(¢ — Em)

"‘M2"‘ (¢ — E,¢)*n<¢ — Em)

C: (¢("), ¢(n + 2))
V3-Ur

Inserting the last expression into equation (4.7) gives

_ c,(¢(n).¢<n+2)) c,(¢<n>,¢(n+1)> ’ "’
"“ jvar,<¢<n)> "(jvar,<¢<n)) ‘)

Notice that we can estimate these expressions of oz; and ﬂl by running simulations of the Markov
chain. To do so, we run the Markov chain starting from a random initial state, compute the value of
the observable at each step of the chain and, from the data, compute estimates of the covariances
and variance of the observable.

Before generalizing this for the jth Lanczos coefficient, we consider one more coeﬂicient
in detail. Recall that the second basis element q; in the Lanczos algorithm is given by 1;; =

(Mql — oz1q1)/ﬁl. From this we can derive an expression for (12 which, though not as neat as those

% 28.txt
28

for oz; and [31, is nonetheless tractable.

I12 ‘IiMq2

1
[7(M91 — 0t1¢I1)tM(M¢I1 - 11191)
1

1
= f,§(qiM3q1 - 2a1qlM2<11 + a’9iM¢11)
1
1
= —2(<1iM3q1 - 2015i ‘ all (4-8)

1.
The last equality follows from equation (4.7), which can be written as q{M2q1 = off’ + (if. The
preceding paragraphs deﬁne all but one of the quantities in the expression (4.8). The one remaining
is q§M3q1, which can also be expressed in terms of covariances of <15, and the analysis is almost

identical to that for q{M2q1:

(¢ — E,¢)'IIP3(¢ — Em)

"lM3"‘ (¢ — E,¢s)‘H(¢ - Em»)

Cu (¢(")a ¢(n + 3))
Varn(¢(n))

To generalize the foregoing for higher order Lanczos coeﬂicients requires a recurrence
relation between (aj+1, ﬂj) and (aj, ﬂ,~_1). As we will see, such a relation involves higher order
Lanczos vectors q,~ — vectors we do not wish to compute. Recall that the only “vector” we can
handle is our initial observable. Therefore, we must also establish a recurrence among the q,-’s, thus
making them available inductively from our observable vector.

To begin with the ﬂfs, recall that ﬁj E ||Mqj — ajq_.,- — ,H_,~_1q,-_;|[2. Whence,

Bf (M411 — °‘j‘Ij - ﬁj-1qj—1)’(MlIj - Otjqj - 51-19141)

q;-M291‘ - 0? - B,z_1 (4-9)

Some simple algebra and the deﬁnitions aj E q§Mqj and ﬁJ~_1 E q;-Mq,--1 prove equality (4.9). The
% 29.txt
recurrence for 0114.1 is equally straight forward, though the ﬁnal form is not as neat:

2
0tj+1 ¢Ij+1M'1j+1

1
§(M¢15 - 11111;‘ - ﬂj—1<Ij—1)'M(Mqj - Otjqj - /3j—1qj—1)
J

1
§(q;-M-“*4; — 2a,~q;M*q,- — 2x3;+’_1q;--1M*q,- + «I? + 2a,-5,51 + am (4.10)
J
From (4.9), q;~M2q,~ = oz? + ﬂf + ﬂ}_1 and, upon plugging this into (4.10), we arrive at
1
a,~+1 = E(q§»M3q_,,- — (1? — 2ozJ~ﬁJ2 — 2ﬂj_1q;»_1M2qj + ﬁ2_1) (4.11)
J

The expressions above involve the higher order Lanczos vectors qj. In the present context,
these vectors will not be readily available. However, another recursion — this time relating forms
involving q,-+1 to forms involving qj and qj__1 — will make these expressions accessible by induction
on our initial observable. The general forms of interest are q;+1M"qj+1 and qj-+1M"qJ~. Again, we

begin with the simpler of the two:

¢I;'+1Mk9j = ij(Mqj - Otjqj - ﬂj—1t1j—1)'M"¢1j
= i(<1§M"“¢Ij - C‘j‘I_;'Mk‘Ij - ﬂj—1q§M"qj-1)
The second is,
‘I_;'+1Mkqj+1 = ﬂi?(M<Ij - Otjqj - .3j—1<Ij—1)‘Mk(Mqj — Orjqj - ﬂj—1¢1j—1)

1
ﬁ(‘1§M'°"<1:‘ + °‘:2"I§M'°<J:' + ﬁ?—z<Ii—1M"<I:‘—1
J
—2a,-q;-M*+‘q,~ + 2am,--1q;-M*q,--1 — zza,--1q;-M*+‘q]--1) (4.12)

The point of all this is that each pair of coeﬂicients (oz,-+1, /3,-) can be expressed as a
function of the covariance between ¢(n) and ¢(n+ m), for m = 0, . . . , 21' + 1. Since the expressions
are based on four recurrence relations, they become quite complicated as 1' gets large. However,
deﬁning these recursions in a. computer program is trivial, and we use this fact in Section 5.4 when

applying the foregoing to a particular problem.

% 30.txt
30

Chapter 5

An Application: Convergence to

Gibbs Measure on the Ising Lattice

5.1 Physical Motivation

Before considering the details of the Metropolis algorithm, it helps to understand the
setting in which it was developed. To do so we consider an typical example, the Ising model.
Consider a model of a system consisting of 71 “sites”, each site taking values in {-1, +1}. It might
help to imagine these sites as equally spaced points on a line or a circle, though we are not restricted
to such cases. Then our state space 5' consists of all possible n—dimensional permutations of -1’s
and +1’s. We reprent this as S’ = {-1, +1}", and note that |S| = 2". We can think ofeach site as
a node existing in one of two possible positions, say “on" or “off”. Each state 1' E S is represented
by a unique permutation a.- of —1’s and +1’s. Let a,-(k) = :|:l denote the position (on or off) of
the kth node when the system is in state 1'. An observable ¢ = qS(a,-) on this space is a function of

the permutations 47,-, 1' 6 {1,. . . , 2"}.

% 31.txt
31

The energy of the system, when in state i, is deﬁned by the Hamiltonian

H07-‘) = - Z Ua(i)0s(k)

<j Ic>
where the sum runs over all neart neighbor pairs. In statistical mechanics we are often concerned

with the Gibbs measure of state i, deﬁned by
1r(a,-) = Z'1 exp{—[3H(a,~)} (5.1)
where Z ‘1 is the normalization constant, or partition function. For the Ising model, we put

Z = §:9XP{~ﬁH(<7:‘)}
i=1

so that 7r is a probability measure. Now, letting 1: denote the so called Boltzman constant, if a
classical mechanical system is in thermal equilibrium with its surroundings with absolute temper-
ature T, and is in state i with energy H (05), then the probability density in phase-space of the
point representing state i is given by (5.1) with [3 = (lcT)“. The central result of ergodic theory
implies that we can also interpret 1r(o;) as the proportion of time the system spends in state i. If

the system is observed at a random time, the expectation E,,¢ of any observable ¢ is thus

.31 ¢(0a) 9"l>{-ﬂH(0s‘)}

E” = 2?;.exp{—nH(a.-)}

(5.2)

Perhaps the number of sites n in our Ising model is so large that it is impractical or impossible to
evaluate (5.2). We might instead consider importance sampling and generate states with probability

density

exp{—aH<a.o}
2?;.exp{—zaH<a.->} (53)

Then «I: is itself an unbiased estimator of (5.2).

1r(0a') =

If evaluation of (5.2) is difficult, there is no reason to believe that evaluation of (5.3) is any

easier. However, Metropolis and his collaborators [9] contrived a method for producing an ergodic

% 32.txt
32

Markov chain with transition probability matrix K whose elements Ic(:::, y) satisfy

2«<z>k<x.y) = «<2»

I

By Deﬁnition 2.1.4 this means that the Markov chain has stationary distribution 7r. Since the
drain is ergodic, we know that it will eventually converge to the desired distribution. So, our main
problem — the rate at which such a Markov chain will converge — is clearly at issue in the application
of the Metropolis algorithm. Moreover, as we will see below, the Metropolis algorithm ensures that
the Markov chain has the desired stationary distribution by requiring that the chain satisfy the
stronger condition of reversibility; i.e. detailed balance. Therefore, Markov chains produced by the
Metropolis algorithm satisfy the assumptions of this paper, which fact conﬁrms that our theory

can be used to bound convergence rates of all such chains.

5.2 Description of the Metropolis Algorithm

The following explains how the Metropolis Algorithm is carried out. The exposition is
similar to that given in Hammersley and Handscomb’s book Monte Carlo Methods 
To begin, choose an arbitrary symmetric Markov chain; that is, a Markov chain whose

transition probability matrix P is symmetric. P is called the proposition matrix, and its elements,
P071"), Satisfy
pa, 1) 2 0. pm‘) = pm). 2,-pm‘) = 1 vi e S (5-4)

Using P, we will derive a new transition probability matrix K which satisﬁes

Z vr(z')k(='.:') = arm (5.5)

I

For 1' aé j deﬁne

kw) = p<-yum")/«<¢> it nu)/1r(='> <1 (5.6)

P(i,J') if 7r(J')/7r(i) Z1

% 33.txt
33

For 2' = j deﬁne

kw) = pm) (1—  p(-3;"), where J. = {j = 2‘ 7‘ an nu)/«(='> < 1}

Notice that 1r(i) > 0 implies k(i,j) _>_ 0 and

;k<«',j> = pow)  { (1 —  pm) +  +]§§p<«",:')
= P(is'5) + 2 P(i1.1') + Z P(iy.7') (5-7)
jel.» jug

Equation (5.7) shows that, for each i, 2, k(z', j) = 1, which fact conﬁrms that K is a stochastic

matrix.

Next we check that K satisﬁes the detailed balance condition. First suppose that 1r(z') =

1r(j). Then (5.4), (5.6), and symmetry of P imply
k(i,J') = p(i2.7') = p(jvi) = k(j1i)

and, since 1r(i) = 1r(j), the detailed balance condition, 1r(i)k(i,j) = 7r(j)k(i,j), clearly holds. Now

suppose that 1r(i) > 1r(j). Then (5.4) and (5.6) imply k(j,i) =p(j, 1') and
k(i,J') = P(iyJ')7r(J')/7r(i) = P(J3i)7r(J')/7r(i) = k(.1'» i)7r(J')/1r(i)

That is, 1r(i)lc(i,j) = 1r(j)k(i,j), as we set out to prove.
To complete the justification of the Metropolis algorithm, we note that the detailed balance
condition, 7r(i)k(i, j) = 1r(j)k(i, j), implies that 1r is indeed the stationary distribution for a Markov

chain with transition probability matrix K. To see this notice that

27r(i)k(i..7') = Zﬂijlkiiajl = 1r(J') Z/€(i,J‘) = 7r(J')

I i
We now summarize the tasks performed when implementing the Metropolis algorithm.

First, pick as a proposition matrix any symmetric stochastic matrix and denote it by P. Begin

% 34.txt
34

the simulation of the Markov chain in state i, and then take one step of the proposition Markov
chain to arrive in state j. That is, choose j according to the mass function p(z', -). Next, compute
1r(j)/1r(i). If 7r(j)/1r(i) 2 1, accept j as the new state. If1r(j)/1r(i') < 1, then with probability
1r(j)/7r(z') accept j as the new state; otherwise, with probability 1 — 1r(j)/1r(i), take i as the new
state. Performing each of these tasks produces one step of the Markov chain represented by the

transition probability matrix K.

5.3 The Ising Model

Returning to the Ising model considered in section 5.1, we consider how the Metropolis
algorithm is applied to such a mode]. Again, start the simulated chain in state X0 = 1' and let

X1 = j be a state chosen according to p(i, -), where p(i, j) = p(j,i). Next, compute

7"(‘7j)/7"(‘7i) = eXP{-ﬁlH(<7j) - H(0i)l}

Suppose the energy H (Uj) of the new state satisﬁes H (U,-) 3 H (05). Then 1r(a_,-)/1r(a,-) Z 1 and we

move to state j. On the other hand, if H(aj) > H(a;), then 1r(aj)/7r(a,-) < 1 and we take

j with probability e-M"

X1 (5.8)

i with probability 1 — e-W’
where AH E H (0,) — H (:75) is the change in energy from state j to state 1'.
There are many ways to choose a proposition matrix P. The simplest derives from the
so called Glauber dynamics, which proceeds as follows: choose one of the 11. sites at random with
probability 1/n — let .9 denote the chosen site — and change the position of this site; i.e., multiply

its value by -1. The resulting state j is the permutation of -1’s and +1’s given by

-0,-(k) if]: = .9
a,~(Ic)= (k=1,...,d)

a,-(Ic) if]: 95 s

% 35.txt
35

The procedure we employ below involves a slight variation: choose site .9 at random uniformly
from the 11 possible sites. However, instead of changing the value of the chosen site with absolute
certainty, as in Glauber dynamics, we “ﬂip the switch” with probability 1/2. Therefore, when k yé .9

we have 0',-(k) = a.~(k), and when k = 3,

—a,~(s) with probability 1/2
01(3) =

a,-(3) with probability 1/2
The advantage of this procedure is that the proposition matrix, before altering it with the Metropo-
lis algorithm, is a bit more interesting than that produced by traditional Glauber dynamics. For,
it has 1/2 along the main diagonal and, thus, is aperiodic and converges to a stationary distribu-
tion. Glauber dynamics has period 2, so its proposition matrix does not satisfy the hypotheses of

our convergence theorem (Theorem 2.1.2). Keep in mind, however, that any proposition matrix

modiﬁed by the Metropolis algorithm satisﬁes all of our hypotheses.

5.4 Computations

5.4.1 Computer Programs

Constructing a computer program to simulate realizations of the Markov chain described
above (i.e. with transition matrix K) is straight forward, and we do so using Matlab and the
program ising.m which appears in the Appendix.

When performing simulations in the present context, there are a few important aspects
to consider. First, our program begins with a “hot start", which means that the initial state is
picked by assigning each site to :l:1 with probability 1/2. Perhaps we should instead have picked
our initial state from the stationary distribution. On the contrary, this would defeat the whole

purpose of our analysis. Recall that the method has been developed to deal with cases for which

% 36.txt
36

sampl from the stationary distribution are not quickly attainable. Furthermore, we are trying to
ﬁgure out how long it will take for observations from the simulated chain to represent samples from
the stationary distribution. Second, the method we have developed depends solely on covariances
of our observable which we estimate from the data. produced by the simulations. However, such
estimates require that the data come from the stationary distribution. Therefore, we must discard a
number of observations before using the data to estimate covariances. The number of observations
to discard depends on how long must we wait until we have data from the stationary distribution,
and we are back where we started — almost. When employing our procedure, we can discard a
conservative (very large) number of observations and perform our analysis, deriving a bound on
the convergence time. Then, in all future studies, we will know how much data should be discarded
before samples come (approximately) from the stationary distribution.

In considering what observable to use on the state space described above, perhaps the
most obvious is simply the number of nodes in the on position (i.e. the number of +l’s). Recalling
that 0, (k) = i1 denotes the position (on or off) of the kth node when the system is in state j, this

observable is deﬁned by,
«M = zxca.-ac) = +1). 1 = 1.---.ISl
I:

where X is the indicator function. The program ising.m simulates the Markov chain, computes the
values of this observable (as well as the number of 1’s squared) and writes it to a ﬁle called phi . dat.
The program then proceeds to compute the ﬁrst few Lanczos coefficients using covariances produced
by Matlab, so that we can check them against the rults we get from the 1anczos.c program,
which we now describe.

The lanczos .c program (listed in the Appendix along with its dependencies) implements
the recursive relations described in section 4.2 to compute the Lanczos coefﬁcients by the new

method. It accepts input from the user specifying any observable, any number of iterations, and any

% 37.txt
37

TABLE 1: ESTIMATES or /\,m(K) = 0.998746 USING ¢1

sim. # /32
1 0.995107 0.006960 0.995220 0.600243
2 0.995099 0.005891 0.994863 0.477393
3 0.994862 0.009671 0.994925 N/ A
4 0.995321 0.009319 0.995654 N/A
5 0.994787 0.004872 0.405446 0.994742

number of desired Lauczos coeﬁicients. That is, the observable could have come from a simulation
of any reversible Markov chain and not just our special Metropolis chain. The output of this
program is a T matrix containing the maximum number of Lauczos coeﬂicients (up to the number
requested by the user) before an approximate invariant subspace was reached. The eigenvalues of
the nonzero part of this matrix, in accordance with the foregoing theory, should well approximate

eigenvalues of the Markov chain’s stochastic matrix.

5.4.2 Results

Table 1 shows the results of 5 simulations each of 100,000 iterations (discarding the ﬁrst
5,000 observations) for the one dimensional Ising lattice with 10 nodes and [7 = 1. Displayed is
the estimate of the ﬁrst Lauczos coeﬂicient a1 (which is the eigenvalue estimate after one step; i.e.
).m,x(T1)), along with its error bound ﬁl. Appearing in the third column is the largest eigenvalue of
the 2 x 2 matrix T2, A0 (T2), and ﬁg appears in the fourth column when it is was computable. When
it was not computable, it was probably because an approximate invariant subspace was reached
at the ﬁrst step, so ﬁg involved expressions which were close to machine epsilon and, thus, could
not be computed accurately. Table 2 shows the same information when using a second observable,
¢2 E ¢§ (i.e. the number of +1’s squared).

The true value of the subdominant eigenvalue for this problem is /\max(K) = 0.998746,

and the next largest is A2(K) = 0.993303. These were computed directly from the transition matrix

% 38.txt
38

TABLE 2: ESTIMATES OF >.,,,,,.,,,(K) = 0.998746 USING z/>2

sim. # B2
1 0.994262 0.006142 0.994222 N / A
2 0.994333 0.002974 0.994332 N / A
3 0.994307 0.008892 0.994374 0.726320
4 0.994436 0.012884 0.994685 0.361404
5 0.993789 0.006909 0.993820 0.788710

using the Matlab program tpm.m listed in the Appendix. They should be very accurate, and it
takes the Matlab routine eig() just over 20 minutes of cpu time on a Sun Ultra to ﬁnd all the
eigenvalues of K.

As shown by Theorem 4.1.1, the [Ts provide conservative error bounds on the eigenvalue
estimates from the previous Lanczos step. Unfortunately, for this example, ,6 appears to be too
conservative as it does not even allow us to bound our estimates away from 1. Furthermore, it
seems that the estimates generated by the method above fall systematically below the true value of
the subdominant eigenvalue. In every simulation, an approximate invariant subspace (to machine
tolerance) was reached after computing at most 2 pairs of Lanczos coefficients. For ¢1 we see that
it was reached after computing only one pair on simulations 3 and 4, similarly for ¢2 on simulations
1 and 2. These observations indicate that the chosen observable is close to the slowest mode (or
eigenfunction) of the process. That is, we should observe fast convergence on an approximate
invariant space. However, the space found might contain only the second slowest mode of the
process and, in that case, our eigenvalue estimates would be closer to the third largest eigenvalue,
instead of the second largest as desired.‘

The matrix T was produced by the program lanczos . c and its eigenvalues were computed
by Matlab, both operations taking a few seconds. Granted, comparing a few seconds to the 20

minutes it takes Matlab is not justiﬁed since a simple call to the routine eig() is far from the

‘Recall that we explained above why we will never converge to the eigenspace corresponding to the largest eigen-
value (which is always 1). That would require a (nearly) constant observable function.

% 39.txt
39

most eﬂicient traditional method for computing only a few eigenvalues of K. However, recall that
the primary motivation for the new method are those examples where the matrix K is so large
that we can’t store even a single vector in fast memory. Once we start swapping data to and from
cheap memory, the time required by traditional algorithms can become years, even lifetimes. Such
examples are not exceptional. Consider, for instance, the 1000 node problem which produces a.

2“’°° x 21°°° transition matrix.

% 40.txt
40

Chapter 6

Conclusion

We conclude by noting a few aspects of the special cases to which the foregoing applies.
We are interested in the subdominant eigenvalue of a large transition probability matrix. The

foregoing can be used to approximate these eigenvalues when the following conditions are satisﬁed:
o The Markov chain is reversible (detailed balance condition).

0 We can easily simulate realizations of the chain as well as functions (observables) of these

realizations.

When these conditions hold, we can side step the traditional Lanczos algorithm, and get at the
values of the Lanczos coeﬂicients via the variational properties of observables on the state space.
There are many ideas in this paper that should be expanded upon in future studies. Per-
haps most obvious is the desire to ﬁnd analogous techniques for non-reversible (i.e. non-symmetric)
Markov chains. This is a considerable problem since the Lanczos methods described above are no
longer applicable. Perhaps the Amoldi algorithm ([1], algorithm 6.9), or the non-symmetric Lanczos

algorithm can also be usefully specialized for stochastic matrices.

% 41.txt
41

When applying the Lanczos algorithm, we usually choose a single starting vector (in the
present context, an observable), perform the algorithm for a number of steps, and gather results.
Then we do the same with a new starting vector and compare the results to those obtained with
the ﬁrst vector. Proceeding in this way for a number of starting vectors, we can provide statistical
evidence that we have, indeed, located the eigenvalues of interest. However, when there are two
observables 451, :15; of particular interest, perhaps we would beneﬁt by considering both observables
simultaneously. Future studies might develop an algorithm which incorporates both observables,

again via ¢.- = H1/2(¢.' - E,.¢.-). In this case, a Lanczos algorithm would be based on the space

Ic*(M,¢1,¢»2.p) 2 ran{«/»mz»2,M«/»1,M«/:2, . . .,MP—‘¢1.M"—‘«/22}

The Lanczos coeﬁicients would then involve not only the autocovariances C, (¢,<(n),¢,-(n + m)),

but also the cross-covariances C, (¢.~(n), ¢j(n + m)) (deﬁned in (2.6)).

% 42.txt
Appendix: computer programs

Z Hatlab code ising.m

Z
Z Written by William J. De Meo on 12/15/97

2 last modified 1/10/98

Z Inputs:

2 d = number of nodes of the ising lattice (e.g. d=100)
Z n = number of iterations (e.g. n=1000)

Z beta = Annealing schedule (e.g. beta = 2,

1 when beta -> 0 will always go to nee state
Z vhen beta -> infty will never go to state of higher energy)
Z

X = zeros(d,1);
Energy = zeros(n,1);

Ave=0;

X Initialize using hot start
for i=1:d,

if rand < .5

x(i) = -1;

else

X(i) = 1;

end
end

Z Initialize at state of high energy (not used)

if 0
X(1:2:d-1)=1;
X(2:2:d)=—1;
and

Z Initialize energy

H=0;

for i=1:(d-1)

H = H - X(i)*X(i+1);
end

2 Initialize observable
phi = zeros(n+1,1);

for i=1:d
if X(i) == 1
phi(1)= phi(1)+1;
end

end

42

% 43.txt
43

for j=1:n

site=0;
while site==

site = round(rand#d);
end
Z Compute new energy (with new X(site) = —x(site)):
if site ==

Hnew = H + 2:x(1)~x(2);
elseif site == d

Hnew = H + 2*X(d)tX(d-1);
else

Hnew = H + 2*X(site)*(X(site-1) + X(site+1));
end

Z Change to new state in two ways:
X with probability 1/2, flip the switch
if rand < .5
if (Hnew <= H) I (rand < exp(-beta*(Hnew-H)))
x(site) = -x(site);
H = Hnew;
Z Compute new value of observable
if X(site) -= 1
phi(j+1) = phi(j) + 1;

else
phi(j+1) = phi(j) — 1;
end
else phi(j+1)=phi(j);
end
else phi(j+1)=phi(j);
end
Energy(j) = H;
end
p1ot(Energy)

% A second observable
phi2 = zeros(n+1,1);
for i=1:n+1

phi2(i) = phi(i)*phi(i);
end

Z Write first observable to phi.dat:

fid = fopen(’/quean.a/masters/deme5025/research/nyuthesis/programs/c/phi.dat’,’w’);
fprintf(fid,’%f\n’,phi);
fc1ose(fid);

Z Write second observable to phi2.dat:
fid = fopen(’/quean.a/masters/deme5025/research/nyuthesis/programs/c/phi2.dat’,’w’);

% 44.txt
fprintf(fid,’%f\n’,phi2);
fc1ose(fid);

1 Compute a few Lanczos coefficients by new method
START=1001;

V1=zeros(2);

C1=zeros(2);

C2=zeros(2);

C3=zeros(2);

V1 = cov(phi(START:n+1),phi(START:n+1));

V1 = V1(1,1)5

C1 = cov(phi(START:n),phi(START+1:n+1));
C1 = C1(1,2);

C2 = cov(phi(START:n-1),phi(START+2:n+1));
C2 = C2(1,2);

C3 = cov(phi(START:n-2),phi(START+3:n+1));

C3 = C3(1,2);

a1pha1 = C1/V1;

betai = sqrt(C2/V1 ' (C1/V1)‘2);

a1pha2 = (1/(beta1’2))*(C3/V1 - 2*a1pha1*(C2/V1) + a1pha1‘3);

T1 = Ealphal betal; betal a1pha2]

v1 = cov(phi2(START:n+1),phi2(START:n+1));
v1 = v1(1,1);

c1 = cov(phi2(START:n),phi2(START+1:n+1));
c1 = c1(1,2);

c2 = cov(phi2(START:n—1),phi2(START+2:n+1));
c2 = c2(1,2);

c3 = cov(phi2(START:n-2),phi2(START+3:n+1));
c3 = c3(1,2);

alphal = C1/V1;
betal = sqrt(C2/V1 - (C1/V1)‘2);
a1pha2 = (1/(beta1‘2))#(C3/V1 - 2ta1pha1*(C2/V1) + a1pha1”3);

T2 = [a1pha1 betai; betal a1pha2]

fid = fopen(’Tmats.dat’,’v’);
fprintf(fid,’%f\n',T1);
fprintf(fid,’%f\n’,T2);
fc1ose(fid);

ZZZ end ising.m

44
% 45.txt
\% Matlab code tpm.m for generating transition probability matrix
Z and computing its eigenvalues directly

1

Z Written by Hilliam J. De Heo on 1/10/98

1

Z Inputs:

Z d = number of nodes of the ising lattice (e.g. d=10)

Z beta = Annealing schedule (e.g. beta - 2,

1 when beta -> 0 will always go to new state
1 when beta -> infty will never go to state of higher energy)
2

disp(’bui1ding proposition matrix...’)

states = 2‘d;

A = 0;

for i=0:(d-1)

E=eye(2”i);

A = [A E;E A];

end

B = .5#eye(states);

A = B + .5-(1/d)*A;

2 our modified Glauber dynamics requires the B and the .5¥(1/d)

disp(’...done’)

1 display pattern of nonzero entries
1 spy(A)

Z check that all row sums are 1
check=O;
for i=1:states

check = check + (not(sum(A(i,1:states))<.99));
end
1 if not all rows sum to 1, print check = (# of rows with sum=1)
if not(check==states)

check

error(’Row sums are not all 1’)
end

Z construct matrix of states
E = zeros(d,states);
f1ip=-1;
for i=1:d
for j=1:2‘(i-1):states
flip = -1*f1ip;
for k=0:2“(i-1)

45

% 46.txt
E(i,j+k) = flip;
and
end
end
E=E’;

Z display first 64 states
Z E(1:64,:)

Z compute energy of each state
H = zeros(states,1);
for i=1:states

for j=1:d-1
H(i) = H(i) - E(i,j)tE(i,j+1);
end
end

disp(’building Hetropolis transition matrix...')

for i=1:states
for j=i+1:states
if not(A(i,j)==0)
if(H(j)>H(i))
alt = A(i,j)t(1-exp(-beta*(H(j)-H(i))));
A(i,i) I A(i,i)+a1t;
A(i,j) = A(i,j)-alt;
elseif(H(j)<H(i))
alt = A(j,i)t(1-exp(—beta#(H(i)-H(j))));
A(j,j) = A(j,j)+a1t:
A(j,i) = A(j,i)-alt;
end
end
end
end

disp(’...done’)

check=0;
for i=1:states

check = check + (not(sum(A(i,1:states))<.99));
end

Z if not all rows sum to 1, print check = (# of rows with aum=1)

if not(check==states)

check

error(’Row sums are not all 1’)
end

46

% 47.txt
47

disp(’computing eigenvalues of tpm...’)
cput = cputime;

evals = eig(A);

ecput = tputime - cput;

disp(’the CPU time (in secs) for computing eigenvalues of tpm: ’)

ecput
ZZZ end tpm.m

/¥#*##¥###*##*#l####¥#‘###$ﬂ###ﬁi##*U3*#**#*##**##¥***ll¥#*##¥ii

* 1anczos.c main program for computing Lanczos coefficients *
3 #
* Created by William J. De Heo #
t on 1/7/98 *
# Last modified 1/10/98 #

*¥#######**3It¥*#¥¥¥#**##**¥##¥#####**####*#*###*##**###¥##*##*/

#inc1ude <stdlib.h>
#include <stdio.h>
#include <math.h>
#inc1ude "prototypes.h"

/# Machine constants on dino (Sun Ultra 1 at Courant) t/
#define HACHEPS 1.15828708535533604981e-16

#define SQRTEPS 1.07623746699106147048e-08

ﬂdefine HAX_NAH 100

void read_name(char *);

/* Prototypes from the file functions.c ¥/

double a1pha(int j);

double betasq(int j);

double form1(int j, int k);

double form2(int j, int k);

double moment(doub1e ‘data, long n, double tave, double tvar, double tcov, long k);

/# external variables to be used by functions */
double *phi,tcov, tvar, tave;

% 48.txt
main()

{

char tfilename;

FILE *ofp;

double temp=O, *T;

long i, j, nrou, nlanc, START;
int f1ag=O;

filename = cma11oc(HAx_NAHE);

printf("\nName of file containing observed values: “);

read_name(fi1ename);

printf(“\nTota1 number of observations in file (iterations):

scanf(“Zu”,&nrov);

printf(“\nNumber of leading observations to discard: “);

scanf("Zu",&START);

printf("\nNumber of Lanczos coefficients desired: “);

scanf("%u",&n1anc);

vhi1e(2*n1anc > nrow—START)

{
printf("\nNot enough data for that many coefficients.\n“);
printf("\nEnter a smaller number of Lanczos coefficients: “);
scanf("Xu“,&n1anc);

}

II);

phi = dma11oc(nrow);
cov=dma11oc(2#n1anc);
var=dma11oc(1);
ave=dma11oc(1);

T = dma11oc(n1anc#n1anc); for(i=0;i<n1anc;i++) T[i]=(doub1e)0;

/* observable phi is stored contiguously column-wise (e.g. by HATLAB) t/
mat1abread(phi, nrow, 1,fi1ename);

/* send the observable, offset by START */
moment(phi+START,nrou-START,ave, var, cov,2*n1anc-1);

/* First column of T */

T[0] = a1pha(1);
if((temp=betasq(1))>HACHEPS*10)
{

T[1]=sqrt(temp);

/* General column of T */

J'=2:

for(i=1;i<n1anc-1, f1ag==O;)
{

T[itn1anc+i-1]=sqrt(betasq(i));

48

main()

{

char tfilename;

FILE tofp;

double temp=0, #T;

long i, j, nrow, nlanc, START;
int f1ag=0;

filename = cma11oc(HAx_NAH);

princf("\nName of file containing observed values: “);
read_name(fi1ename);

printf("\nTota1 number of observations in file (iterations): “);
scanf("2u“,&nrow);

printf("\nNumber of leading observations to discard: “);
scanf("Zu",&START);

printf("\nNumber of Lanczos coefficients desired: “);
scanf("%u“,&n1anc);

whi1e(2*n1anc > nrou-START)

{
printf("\nNot enough data for that many coefficients.\n");

printf("\nEnter a smaller number of Lanczos coefficients: “);
scanf("Zu",&n1anc);
}

phi = dma11oc(urow);
cov=dma1loc(2*n1anc);
var=dma11oc(1);
ave=dma11oc(1);

T = dma11oc(n1anc*n1anc); for(i=0;i<n1anc;i++) T[i]=(doub1e)O;

/* observable phi is stored contiguously column-wise (e.g. by HATLAB) */
mat1abread(phi, nrov, 1,fi1ename);

/* send the observable, offset by START #/
moment(phi+START,nrov-START,ave, var, cov,2*n1anc-1);

/# First column of T */

T[0] = a1pha(1);
if((temp=betasq(1))>HACHEPS#10)
{

T[1]=sqrt(temp);

/* General column of T */

J‘=2:

for(i-1;i<n1anc-1, f1ag==0;)
{

T[i*nlanc+i-1]=sqrt(betasq(i));

48

% 49.txt
T[i*n1anc+i] = a1pha(i+1);
if((temp = betasq(i+1))>HACmEPS#10)
{

T[i#n1anc+i+1] = sqrt(temp);

i++;
}
else

f1ag=1;

}

/t Last column */
if(f1ag!=1 it ((temp= betasq(n1anc-1))>HACHEPSt10))

{
T[n1anc#n1anc-2] = sqrt(temp);
T[n1anctn1anc-1] = a1pha(n1anc);
printf("\nbeta(1) = %1f\nbeta(Zd) = Zlf (last beta)",
T[1],(nlanc-1),T[nlanc*n1anc-2]);
}
else
{
printf("\nApproximate invariant space reached at step %d.",i);
printf("\nbeta(1) = Z1f\nbeta(Zd) = Zlf (last accurate beta)“.
T[1],i,T[itn1anc+i—1]);
printf("\nbeta(Zd)“2: Zlf (first spurious resu1t)“,i+1,temp);
}

printf("\nThe matrix T is: \n");
matprint(T,n1anc,n1anc);

}

else

{

printf("\nmain(): Approximate invariant space reached at first step.“);

49

printf("\na1pha(1) - Z1f\nbeta(1)”2: Zlf (first spurious result)“, T[O], temp);

}
ofp = fopen("Tmat.m","v“);
check(ofp);
mat1aburite(T,n1anc,n1anc,ofp);
fc1ose(ofp);
}
void read_name(char tname)
{
int c, i = 0;
while ((c = getchar()) != EOF it c != ’ ’ it c !- ’\n’)
name[i++] = c;
name[i] = ’\0’;
}

/#t* end 1anczos.c *#*/

% 50.txt
50

/llnkttttttttinlntint*1!!!##1##!ttitilltit###t¥#tt¥ttItt####t#ttII=##mlltttt

* functions.c functions required by 1anczos.c t
# t
* Created by Hilliam J. De Meo #
* on 1/7/98 *
t Last modified 1/10/98 :

¥¥##¥$¥##¥¥#$¥¥¥###¥¥$*I#****#**####*######¥#¥#¥#$#¥¥#l¥**#****/

#inc1ude <math.h>

#define START 1000

#define ITER 10000

/t Machine constants on dino (Sun Ultra 1 at Courant) */

#define HACHEPS 1.15828708535533604981e-16

#define SQRTEPS 1.07623746699106147048e-O8

double a1pha(int j);

double betasq(int j);

double form1(int j, int k);

double form2(int j, int k);

double moment(doub1e tdata, long n, double tave, double tvar, double tcov, int k);

/t external variables to be used by functions */
double *phi,tcov, *var, tave;

double alpha(int j)
{

/* alpha is never called with j < 1 */
if(j==1)
return
form1(1,1);
else if(j>1)
return
((doub1e)1/betasq(j-1))*
(form1(j-1,3) — pou(a1pha(j-1),3)
- 2 * (alpha(j-1)#betasq(j-1) + sqrt(betasq(j-2))#form2(j-1,2))
+ betasq(j-2));
}

double betasq(int j)

{
if(j==0)
return (doub1e)O;
else if(j>0)
return (form1(j,2) - pow(a1pha(j),2) - betasq(j-1));
}

double form1(int j, int k)
{

% 51.txt
51

double form14,alphal,alpha1sq,form12,form13;
if(j==0)

return (doub1e)0;
else if(j==1)
{

/# printf("\nvar = %1f, cov(%d) = %1f \n“,*var,k,cov[k]);t/

return (cov[k])/(*var); /* the only real value */
}
else if(j>1)
{

return ((double)1/betasq(j-1))

* ( form1(j-1,k+2)
+ pow(a1pha(j-1),2)*form1(j-1,k)
+ 2t(a1pha(j-1) t sqrt(betasq(j-2)) * form2(j-1,k)
- a1pha(j-1)*form1(j-1,k+1)
- sqrt(betasq(j—2))#form2(j-1,k+1) )
+ betasq(j-2)*form1(j-2,k)):
}
}

double form2(int j, int k)
{
/* form2 is never called with j < 1 */
if(j==1)
return (doub1e)0;
else if(j>1)
return
(pov(betasq(j-1),—.5)) 3
(form1(j-1,k+1) - a1pha(j-1)*form1(j-1,k)
- sqrt(betasq(j-2)) # form2(j—1,k));
}

/* moment() function for computing var and cov(k)
arguments:
data = a nxi array of doubles
n = length of data[]
ave =(on exit)= the average of data[]
var =(on exit)= the variance of data[]
cov =(on exit)= the covariance of data[i] and data[i+j] for i-1,...,k
k = the max lag for cov above
*/
double moment(double tdata, long n, double tave, double tvar, double tcov, int k)
{
/* Centered about data[0] algorithm: #/

long i,j;
double avel, ave2;

% 52.txt
52

*ave=O;tvar=0; ave1=ave2=O;
for(j=0;j<=k;j++)
cov[j]=0;

for(i=1;i<n;i++)
{
rave += (data[i] - data[0]);
tvar += (data[i] - data[0])#(data[i] - data[0]);
}
tvar /= (doub1e)(n-1);
Ivar -= (((*ave)/(doub1e)n) * ‘(*ave)/(doub1e)(n-1)));
/# *ave = ((*ave)/(doub1e)n) + data[0]; <- that’s the true average (not needed)*/

for(j=0;j<=k;j++)
{

for(i=0;i<(n-j);i++)

{

avel += (data[i] - data[0]);
ave2 += (data[i+j] - data[0]);
cov[j] += (data[i] - data[0])*(data[i+j] — data[0]);

}

avel/=(doub1e)(n-j); ave2/=(doub1e)(n-j);

cov[j] = ((cov[j] - (double)(n-j)tave1¢ave2)/(double)(n-j-1));
}

}

/trt end funccions.c ###/

### 2013-10-18 tex/Appendix.tex
initial commit of tex source for nyu thesis style setup

--- 2013-10-18 tex/MSThesis.tex
@@
-%\documentclass[11pt]{amsart}
-\documentclass{au}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% For completion by editorial office:
- \presentedby{\dots}
- \received{\dots}{\dots}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\usepackage[colorlinks=true,urlcolor=black,linkcolor=black,citecolor=black]{hyperref}
-
-\usepackage{amsmath,amssymb,url,mathrsfs}
-\usepackage{enumerate}
-\usepackage{tikz}
-\usetikzlibrary{calc}
-\usepackage{scalefnt}
-\numberwithin{equation}{section}
-
-\theoremstyle{plain}
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{proposition}[theorem]{Proposition}
-\newtheorem{corollary}[theorem]{Corollary}
-
-\theoremstyle{definition}
-\newtheorem{definition}[theorem]{Definition}
-\newtheorem{notation}[theorem]{Notation}
-\newtheorem{remark}[theorem]{Remark}
-\newtheorem{example}[theorem]{Example}
-\newtheorem*{remarks}{Remarks}
-
-\newcommand{\todo}[1]{~\\[5pt]{\bf TODO(wjd):} #1\\[5pt]}
-%\newcommand{\suchthat}{\ensuremath{:}}    % (old notation)
-\newcommand{\suchthat}{\ensuremath{\mid}}  % (new notation)
-%\newcommand{\defeq}{\ensuremath{:=}}      % (old notation)
-\newcommand{\defeq}{\ensuremath{=}}        % (new notation)
-\newcommand{\betai}{\ensuremath{\beta^i}}  % \beta^{\bB_i}   % (old notation)
-\newcommand{\betaj}{\ensuremath{\beta^j}}  % \beta^{\bB_j}
-\newcommand{\betak}{\ensuremath{\beta^k}}  % \beta^{\bB_k} or \beta^{\bB_{k}}
-\newcommand{\betal}{\ensuremath{\beta^\ell}}  %\beta^{\bB_\ell} or \beta^{\bB_{\ell}}
-\newcommand{\uacalc}{{\small UAC}alc}
-\newcommand{\Tuma}{T\r{u}ma}
-\newcommand{\Palfy}{P\'alfy}
-\newcommand{\Pudlak}{Pudl\'ak}
-\newcommand{\PP}{P\'alfy-Pudl\'ak}
-\newcommand{\PAP}{P\'alfy\ and Pudl\'ak}
-\newcommand{\<}{\ensuremath{\langle}}
-\renewcommand{\>}{\ensuremath{\rangle}}
-\newcommand{\beps}{\ensuremath{\boldsymbol{\varepsilon}}}
-\newcommand{\bN}{\ensuremath{\mathbf{N}}}
-\newcommand{\bA}{\ensuremath{\mathbf{A}}}
-\newcommand{\bB}{\ensuremath{\mathbf{B}}}
-\newcommand{\bBi}{\ensuremath{\mathbf{B}_i}}
-\newcommand{\sB}{\ensuremath{\mathcal{B}}}
-\newcommand{\sC}{\ensuremath{\mathcal{C}}}
-\newcommand{\sS}{\ensuremath{\mathscr{S}}}
-\newcommand{\sT}{\ensuremath{\mathscr{T}}}
-\newcommand{\sI}{\ensuremath{\mathcal{I}}}
-\newcommand{\sE}{\ensuremath{\mathcal{E}}}
-\newcommand{\sO}{\ensuremath{\mathcal{O}}}
-%\newcommand{\id}{\ensuremath{\mathrm{id}}}
-\DeclareMathOperator{\id}{id}
-%\newcommand{\Eq}{\ensuremath{\mathrm{Eq}}}
-\DeclareMathOperator{\Eq}{Eq}
-%\newcommand{\Cg}{\ensuremath{\mathrm{Cg}}}
-\DeclareMathOperator{\Cg}{Cg}
-%\newcommand{\Con}{\ensuremath{\mathrm{Con\,}}}
-\DeclareMathOperator{\Con}{Con}
-%\newcommand{\Sub}{\ensuremath{\mathrm{Sub}}}
-\DeclareMathOperator{\Sub}{Sub}
-%\newcommand{\Pol}{\ensuremath{\mathrm{Pol}}}
-\DeclareMathOperator{\Pol}{Pol}
-%\newcommand{\Clo}{\ensuremath{\mathrm{Clo}}}
-\DeclareMathOperator{\Clo}{Clo}
-\newcommand{\ps}[1]{\ensuremath{^{(#1)}}}
-\newcommand{\piB}{\ensuremath{\pi_B}}
-\newcommand{\hpsi}{\ensuremath{\widehat{\psi}}}
-\newcommand{\htheta}{\ensuremath{\widehat{\theta}}}
-\newcommand{\supi}{\ensuremath{^{i}}}
-\newcommand{\supj}{\ensuremath{^{j}}}
-\renewcommand{\leq}{\ensuremath{\leqslant}}
-\renewcommand{\nleq}{\ensuremath{\nleqslant}}
-\renewcommand{\geq}{\ensuremath{\geqslant}}
-\newcommand{\meet}{\ensuremath{\wedge}}
-\newcommand{\join}{\ensuremath{\vee}}
-\newcommand{\Meet}{\ensuremath{\bigwedge}}
-\renewcommand{\Join}{\ensuremath{\bigvee}}
-\newcommand{\resB}{\ensuremath{|_{_B}}}
-\newcommand{\resBi}{\ensuremath{|_{_{B_i}}}}
-\newcommand{\eps}{\ensuremath{\varepsilon}}
-\newcommand{\hatmap}{\ensuremath{\widehat{\phantom{x}}}}
-\newcommand{\one}{\ensuremath{\mathbf{1}}}
-\newcommand{\two}{\ensuremath{\mathbf{2}}}
-\newcommand{\three}{\ensuremath{\mathbf{3}}}
-\newcommand{\tbeta}{\ensuremath{\widetilde{\beta}}}
-\newcommand{\ttheta}{\ensuremath{\widetilde{\theta}}}
-\newcommand{\hbeta}{\ensuremath{\widehat{\beta}}}
-\newcommand{\cick}{\ensuremath{\sC_i}}
-%\newcommand{\CE}{\ensuremath{\sC_r^{\sE}}}
-\newcommand{\CE}{\ensuremath{\sB_r^n}}
-\newcommand{\CO}{\ensuremath{\sC_r^{\sO}}}
-\newcommand{\CICK}{\ensuremath{c_i/\beta \cup c^{K+1}_i/\beta^{\bB_{K+1}}}}
-\newcommand{\GAP}{{\small GAP}}
-\newcommand{\dotsize}{.8pt}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{document}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%  FRONT MATTER                                                 %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\title[Approximating Eigenvalues of Large Stochastic Matrices]{A Lanczos Procedure for Approximating Eigenvalues of Large Stochastic Matrices}
-
-%% First author (Note: The order of the items here is important!)
-\author{William DeMeo}
-\email{williamdemeo@gmail.com}
-\urladdr{http://www.math.sc.edu/~demeow}
-\address{Department of Mathematics\\
-University of South Carolina\\Columbia 29208\\USA}
-
-\thanks{The author wishes to thank a number of people who contributed to this
-  work.  In particular, the primary referee made many insightful recommendations
-  which led vast improvements, not only in the overall exposition, but also in
-  the specific algebraic constructions, theorem statements, and proofs.
-  This article is dedicated to Ralph Freese and Bill Lampe, who declined to be
-  named as co-authors despite their substantial contributions to the project.}
-
-%% Dedication (Optional)
-\dedicatory{This article is dedicated to Ralph Freese and Bill
-  Lampe.}
-
-%% AMS subject classification; see http://www.ams.org/msc
-%% Only one Primary. Possibly several Secondary.
-\subjclass[2010]{Primary: 08A30; Secondary: 08A60, 06B10.} %, 06B15.}
-
-%% Keywords and phrases
-\keywords{congruence lattice, finite algebra, finite lattice
-  representations}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%  MAIN MATTER                                                  %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{abstract}
-The rate at which a Markov chain converges to a given probability distribution
-has long been an active area of research. Well known bounds on this rate of
-convergence involve the subdominant eigenvalue of the chain’s underlying
-transition probability matrix. However, many transition probability matrices are
-so large that we are unable to store even a vector of the matrix in fast
-computer memory. Thus, traditional methods for approximating eigenvalues are
-rendered useless.
-
-In this paper we demonstrate that, if the Markov chain is reversible, and we
-understand the structure of the chain, we can derive the coefficients of the
-traditional Lanczos algorithm without storing a single vector. We do this by
-considering the variational properties of observables on the chain’s state
-space. In the process we present the classical theory which relates the
-information contained in the Lanczos coefficients to the eigenvalues of the
-Markov chain. 
-\end{abstract}
-
-\maketitle
-
-%%%%%%%%%%%%%%%%%%%%%
-\chapter{Introduction}
-The rate at which a Markov chain converges to a given probability distribution has long
-been an active area of research. This is not surprising considering this problem’s relevance to the ar-
-eas of statistics, statistical mechanics, and computer science. Markov Chain Monte Carlo (MCMC)
-algorithms provide important examples. These algorithms come in handy when we encounter a
-complicated probability distribution from which we want to draw random samples. In statistical
-mechanics, we might wish to estimate the phase average of a function on the state space. Goodman
-and Sokal [6] examine Monte Carlo methods in this context. Examples from statistics occur in the
-Bayesian paradigm when we are forced to simulate an unwieldy posterior distribution (see, e.g.,
-Geman and Geman
-
-To implement the MCMC algorithm, we invent a Markov chain that converges to the
-desired distribution (this is often accomplished using the Metropolis algorithm
-described in Chapter 5). Realizations of the chain will eventually represent
-samples from this distribution. Sometimes ``eventually'' -- meaning all but
-finitely many terms of the chain -- is just not enough. We need more practical
-results. In particular, we want to know how many terms of the chain should be 
-discarded before we are sampling from a distribution that is close (in total variation distance) to
-% 03.txt
-the distribution of interest. This is the purpose of bounding convergence rates for Markov chains.
-
-Often the Markov chains encountered in this context satisfy a condition known in the
-physics literature as detailed balance. Probabilists call chains with this property reversible. This
-simply means that the chain has the same probability law whether time moves
-forward or 
-backward.\footnote{This is not a precise definition. In particular the chain must
-  have started from its stationary distribution. Full rigor is postponed until Section 2.1.}
-In this paper, we consider the rate at which such chains converge to a \emph{stationary distribution}.\footnote{This and other italicized terms are defined in Section 2.1.}
-
-There are a number of different methods in common use for bounding convergence rates of
-Markov chains, and a good review of these methods with many references can be found in  More
-recently developed methods, employing logarithmic Sobolev inequalities, are reviewed in  Most
-of the bounds in common use involve the sub—dominant eigenvalue of the Markov chain's transition
-probability matrix, and thus require good approximations to such eigenvalues. In many applications,
-however, the transition probability matrix is so large that it becomes impossible to store even a
-single vector of the matrix in conventional computer memory. These so called out-of-core problems
-are not amenable to traditional eigenvalue algorithms\footnote{By ``traditional
-  eigenvalue algorithms'' we refer to those found, for example, in Golub and Van
-  Loan[5]. See also the book by Demmel [1] for a more recent discussion.}
-without modification. This paper develops 
-such a modification for the Markov chain eigenvalue problem. In particular it develops a method
-for approximating the first few eigenvalues of a transition probability matrix when we know the
-general structure of the underlying Markov chain. The method does not require storage of large
-matrices or vectors. Instead we need only simulate the Markov chain, and conduct a statistical
-analysis of the simulation.
-
-Here is a look at what follows. Section 2.1 contains a review of the relevant Markov chain
-theory. Readers conversant in the asymptotic theory of Markov chains might wish to at least skim
-Section 2.1, if only to become familiar with our notation. Section 2.2 describes functions on the state
-% 04.txt
-space of the Markov process. This section and Chapter 3 develop the context in which we formulate
-the new ideas of the paper. In the last section of Chapter 3, Section 3.3, we present the familiar
-Krylov subspace and explain why this represents our best approximation to a subspace containing
-extremal eigenvectors of the transition probability matrix.\footnote{or, more
-  precisely, a similarity transformation of this matrix.} The first section of
-Chapter 4 describes the \emph{Lanczos algorithm} for generating an orthonormal basis
-for the Krylov subspace. As it stands, this algorithm is useless for an
-out-of-core problem such as ours since, by definition of such problems, 
-it requires too much data movement; all the computing time is spent swapping data between slow
-and fast memory (e.g. between the hard disk and cache). Therefore, we discuss alternatives to
-Lanczos and demonstrate that the \emph{Lanczos coeficients} are readily available through simulations of
-the Markov chain, which fact allows us to avoid the standard algorithm altogether. Following this
-is a chapter describing the Metropolis algorithm used to produce a reversible stochastic matrix.
-It is here that we experiment with the procedure described in Section 4.2 and approximate the
-extremal eigenvalues of the matrix, without storing any of its vectors. Finally, Chapter 6 concludes
-the paper.
-% 05.txt
-
-\chapter{Markov Chains}
-
-\section{General Theory}
-
-This review of Markov chain theory can be found in any good probability text. The present
-discussion is most similar to that of Durrett [3], to which we refer the reader desiring greater detail.
-
-\subsection{The Basic Setup}
-
-Heuristically, a Markov chain is a stochastic process with a lack of memory property. Here
-this means that the future of the process, given its past behavior and its present state, depends only
-on its present state. This is the probabilistic analogue of a familiar property of classical particle
-systems. Given the position and velocities of all particles at time t, the equations of motion can be
-completely solved for the future evolution of the system. Thus, information describing the behavior
-of the process prior to time t is superﬂuous. To be a bit more precise, if technical, we need the
-following definitions.
-
-\bibliographystyle{spmpsci}
-\bibliography{wjd}
-
-\end{document}
-
-%%%%%%%%%%%%%%%%%%%%%  END DOCUMENT %%%%%%%%%%%%%%%%%%%%%%%%%
-
-Definition 2.1.1 Let (S, 8) be a measurable space. A sequence X,., n 2 0, of random variables
-
-taking values in S is said to be a Markov chain with respect to the filtration a(Xo, . . . ,X,,) if for
-% 06.txt
-a.llBES,
-
-p(x,,+1 e B | a(Xo, . . . ,X,.)) = P(X,.+1 e B 1 o(X,,)) (2.1)
-
-Equation (2.1) merely states that, given the present location of X,,, the past is irrelevant for
-
-predicting the location of X,,+1.
-
-Definition 2.1.2 A function p : S x S —> R is said to be a transition probability if:
-1. for each :1: E S, A —> p(a:, A) is a probability measure on (S, S).
-2. for each A E 8, :1: —> p(z, A) is a measurable function.
-
-We call X,, a Markov chain with transition probabilities 12,, if
-P(Xn+l E B l a(Xn)) =Pn(XnaB) (22)
-
-The spaces (S, S) that we encounter below are standard Borel spaces, so the existence of the
-transition probabilities follows from the existence of regular conditional probabilities on Borel spaces
-— a standard measure theory result (see e.g. [3] page 230).
-
-Suppose we are given an initial probability distribution p on (S, S) and a sequence p,, of
-
-transition probabilities. We can define a consistent set of finite dimensional distributions by
-
-P<X,- e B,-,0 s 1 s n) = / #(d1‘o) / Po($o.d-'51) . .. / Pn-l(97n—1ad$n) (2.3)
-B0 B1 B1:
-Furthermore, denote our probability space by
-(9,?) = (S{0,1,...},8{0,l,...})
-
-We call this sequence space and it is defined more explicitly by
-
-sf°»1=-~-} = {(wo,w1,...) :w.- e S}
-
-S{°*1""} = ¢7(w :40; E A,- E 3)
-
-% 07.txt
-The Markov chain that we will study on this space is simply X" (w) = w,., the coordinate maps.
-Then, by the Kolmogorov extension theorem, there exists a unique probability measure P,, on
-(9,?) so that the X" (w) have finite dimensional distributions (2.3).
-
-If instead of p, we begin with the initial distribution 6,, i.e. point mass at 2:, then we
-denote the probability measure by P3. With such measures defined for each 2:, we can in turn
-
-define distributions P,,, given any initial distribution p, by
-
-PM = / u<da.->P.(A>
-
-That the foregoing construction — which, recall, was derived merely from an initial distri-
-bution p and a sequence p,, of transition probabilities — satisfies Definition (2.2) of a Markov chain
-is not obvious, and a proof can be found in
-
-To state the converse of the foregoing, if X,, is a Markov chain with transition probabilities
-p,, and initial distribution p, then its finite dimensional distributions are given by (2.3). Proof of
-this is also found in
-
-Now that we have put the theory on a firm, if abstract, foundation, we can bring the
-discussion down to earth by making the forgoing a little more concrete. First, we specialize our
-study of Markov chains by assuming that our chain is temporally homogeneous, which means that
-the transition probabilities do not depend on time; i.e. p,,(w,,, B) E p(w,., B). (This is the stochastic
-analogue of a conservative system.) Next we assume that our state space S is finite, and suppose
-for all states 1', j E S that p(i, j) Z 0, and E, p(z', j) = 1 for all i. In this case, equation (2.2) takes
-
-a more intuitive form:
-P(Xn+l =j I X7: = 7:) =P(iaj)
-
-and our transition probabilities become
-% 08.txt
-If P is a matrix whose (i, j)th element is the transition probability p(i, 1') then P is a stochastic
-
-matriar, that is, a matrix with elements pgj satisfying
-
-Pij 20: zjpij =1: (i7j=1121"‘)d)
-
-We also refer to P as the transition probability matrix.
-
-Without loss of generality, we can further suppose our Markov chain is irreducible. This
-means that, for any states 2', 1‘, starting in state i the chain will make a transition to state 1' at
-some future time with positive probability. This state of affairs is often described by saying that
-all states communicate. We lose no generality with this assumption because any reducible Markov
-chain can be factored into irreducible classes of states which can each be studied separately.
-
-The final two conditions we place on the Markov chains considered below will cost us
-some generality. Nonetheless, there remain many examples of chains meeting these conditions,
-thus making the present study worthwhile. Furthermore, it may be the case that, with a little
-more work, we will be able to drop these conditions in future studies. The first condition is that
-the chain is aperiodic. If we let I, = {n 2 1 : p"(:c,a:) > 0}, we call a Markov chain aperiodic if,
-for any state 2:, the greatest common divisor of I, is 1. The second assumption is that our chain is
-
-reversible. This characterization is understood in terms of the following definition.
-Definition 2.1.3 A measure a is called reversible if it satisfies
-
-;4(w)p(I. y) = u(y)p(y. w)
-
-for all :c and y.
-
-We call a Markov chain reversible if its stationary distribution (defined in Section 2.1.2) is reversible.
-% 09.txt
-\subsection{A Convergence Theorem}
-
-In succeeding arguments, we use some results concerning the asymptotic behavior of
-
-Markov diains. These results require a few more definitions.
-
-Definition 2.1.4 A measure 1r is said to be a stationary measure if
-
-Z«(z)p<z.y> = we) (2.4)
-
-2
-Equation 2.4 says P,,(X1 = y) = 1r(y), and by induction that P,,(X,, = y) = 7r(y) for all n 2 1. If 1r
-is a probability measure, then we call 1r a stationary distribution. It represents an equilibrium for
-the chain in the sense that, if X0 has distribution 1r, then so does X,. for all 11..
-
-When the Markov chain is irreducible and aperiodic, the distribution of the state at time
-n converges pointwise to 1r as n —> oo, regardless of the initial state. It is convenient to state
-this convergence result in terms of the Markov chain's transition probability matrix P. Before
-doing so, we note that irreducibility of a Markov chain is equivalent to irreducibility (in the usual
-matrix theory sense) of its transition probability matrix. Furthermore, it turns out that a transition
-probability matrix of an aperiodic Markov chain falls into that class of matrices often called acyclic,
-but for simplicity we will call such stochastic matrices aperiodic. With this terminology, we can
-
-state the convergence theorem in terms of the transition probability matrix P.
-
-% Theorem 2.1.1
-\begin{theorem}
-\label{thm-2.1.1}
-Suppose P is irreducible, aperiodic, and has stationary distribution 7r. Then as n —> oo,
-p"(i,J') -> #0")
-\end{theorem}
-
-The symbol p"(i, j) is to be understood as the (ij) element of the nth power of P.
-A Markov chain whose transition probability matrix satisfies the hypotheses of
-Theorem~\ref{thm-2.1.1} is called ergodic. If we simulate an ergodic chain for
-sufficiently many steps, having 
-% 10.txt
-begun in any initial state, the final state is a sample point from a distribution which is close to 1r.
-
-To make this statement more precise requires that we define “close”.
-
-Definition 2.1.5 Let 1r be a probability measure on S. Then the total variation distance at time
-
-n with initial state I is given by
-AA") = l|P"(z, A) - 7r(-4)||Tv = 1333; I P"(I.A) - 7r(A) I
-
-In what follows, we will measure rate of convergence using the function 'r,(e), the first time after
-
-which the total variation distance is always less than 5. That is,
--r,,(e) = min{m : A,(n) 3 E for all n. 2 m}
-
-To consider the connection between convergence rates and eigenvalues, note that an ape-
-riodic stochastic matrix P has real eigenvalues 1 = /\o > A1 2 A2 3  Z /\¢_1 2 -1, where
-d = |S'] is the dimension of the state space. For an ergodic chain, the rate of convergence to the
-stationary distribution 7r is bounded by a function of the subdominant eigenvalue. By subdominant
-eigenvalue we mean that eigenvalue which is second-largest in absolute value, and we denote this
-Am“ E max{).1,|A¢_1|}. The function bounding the rate of convergence is given by the following
-
-theorem:
-
-% Theorem 2.1.2
-\begin{theorem}
-\label{thm-2.1.2}
-The quantity T1(€) satisfies
-1. 13(5) 5 (1 — /\m,_x)“(ln1r(a:)‘1 +1ne‘1)
-2. maxxes r,(e) 2 %/\,,,,x(1 — ).,,,,x)‘1ln(2e)’1
-\end{theorem}
-
-As this theorem shows, if we have an upper bound on the subdominant eigenvalue, then we have
-an upper bound on the function 11(5). In what follows, we will derive an approximation to the
-% 11.txt
-subdominant eigenvalue and supply error bounds. Together, an approximation and error bounds
-for Am“ provide enough information to make Theorem~\ref{thm-2.1.2} useful.
-
-\section{Functions on the State Space}
-
-Recall that X,.(w) = w,. 6 S denotes the state in which the Markov chain exists at time
-11. Suppose that 11> = {¢1, . . . , ¢,,} is a collection of p observables, or functions defined on the state
-space S. Furthermore, let these observables be real valued, so ¢.~ : S —> R. It is often useful, as we
-will see below, to assume that none of the observables is a constant function. Suppose now that
-the state space S is finite with at possible states. Then, since an observable is simply a map of the
-state space, we can think of each ¢,- as a vector of d real numbers — the d values that it takes on at
-the different states.
-
-Now assume the Markov chain is irreducible, and let 1r denote its stationary distribution.
-If we start the chain from its stationary distribution - i.e. suppose X0 has distribution 1r - then
-X,. is a stationary process. Furthermore, for each i, ¢,(X,.) is a stationary stochastic process with
-
-mean
-
-E.¢.- = Z) vr(x)¢.(z)
-
-165
-
-and autocovariance function
-
-Cir (¢i(Xn)v ¢i (Xn+a))
-
-E7!’  ‘ E1r¢i) (¢i(Xn+a) " E1r¢i)l
-
-Z Pn(Xn = z1X7I+S = y)  — E1r¢i)  _ E1r¢i)
-
-z,yES
-(2.5)
-
-which, by the definition of conditional probability, we can write as,
-
-2 P«(X.. = z)P.(X..+.. = y I X. = 1)(¢s($) — Em) (my) ~ Em)
-1.1165
-% 12.txt
-Z 7r($)P1y (MI) - Em.-) (¢:(1I) - E1r¢i)
-
-:c,y€S
-
-Here pgy denotes the element in row :1: and column y of P’, the sth power of the transition probability
-
-matrix. Similarly, we define the cross-covariance between the function ¢,- at time n and q5,- at time
-
-n+sas
-
-E1!’  _ E7r¢:') (¢i(Xn+a) " E1r¢i)l (2-6)
-
-C1r(¢i(Xn)a (#1 (Xn+a))
-
-Z) vr(x)p;,, (¢.-(w) — Em.-) (my) — Em)
-
-z,yES
-
-Now let (<I>) denote the matrix of mean vectors whose jth column is E,,¢,-1, where 1 = (1, . . . , 1)‘,
-and let H = diag(1r(w1), . . . , 1r(wd)), be the d X d diagonal matrix with stationary probabilities 1r(w)
-on the main diagonal, and zeros elsewhere. Finally, denoting by C(s) the p X p covariance matrix
-
-whose (i,j)th element is C, (qfi,-(X,.),¢_,< (X,.+,)), we have
-
-0(0) = E(‘1’(Xn) - (‘I>))(‘I’(Xn) - (‘PW
-= (‘P-(‘1’))‘1'1(<1’-(‘I>))
-0(8) = E(<I>(Xn) — (‘I>))(‘1’(Xn+a) - (‘1’))’
-
-(‘P - (‘1’))'1'IP‘(<1> — (<1>))
-
-Below, we will also find it useful to have at our disposal a new matrix that is similar to the
-transition probability matrix. We have in mind the matrix M E Ill/2PI'I‘1/2. As is easily verified,
-this allows us to write the covariance matrix as
-
-can = <<I>—<<I>>)°n%*M*n%<<I>—<<I>>>
-
-\II‘M’\II (2.7)
-
-where we have defined ‘II E 111/ 2(<I> —  Recall that our main motivation is that, for out-of-core
-problems, traditional eigenvalue algorithms are inadequate. With this fact, and the form (2.7) in
-% 13.txt
-mind, we will consider using the covariance of observables on the state space to implement the
-(otherwise useless) Rayleigh-Ritz procedure, which we describe below. This procedure requires
-that M be symmetric. As the next fact demonstrates, this required symmetry is the reason for our
-
-interest in the reversibility property of the Markov chain.
-
-Fact 2.2.1 The matrix M is symmetric if and only if the Markov chain is reversible (i.e. iﬂ the
-
-process satisfies the detailed balance condition).
-
-Proof:
-
-M is symmetric
-
-4:, (1-I1/2P1-I-1/2): = H1/2Pn—1/2
-<.—> r1-%*P°n%* = H1/2911-1/2
-¢> 1>'r1= = up
-~:=> (IIP)‘ = l'IP
-
-Element-wise, the final equality implies that 7|‘,'p,'j = 1rjpj.-. According to Definition 2.1.3, this states
-
-that p is a reversible measure.
-
-% 14.txt
-
-\chapter{Invariant and Approximate Invariant Subspaces}
-
-\section{Invariant Subspaces}
-Definition 3.1.1 A subspace S Q R" with the property that
-
-1:ES'=Ma:€S
-
-is said to be invariant for M.
-
-Recall that, having chosen observables <I> = ((151, . . . , qfip), we constructed the covariance
-matrix C(s) = \II‘M"\II. If the column space of \II, which we denote by ran(\II), is an invariant
-subspace for M, the definition implies Mifij E  That is, for each 111,- there exists a vector
-t of coeﬂicients such that M1/Jj = f=1t,-1/)5,-. This is true for all j and, putting each vector of
-coefficients into a matrix T, we see that M\II = \IIT. Conversely, M\II = IIIT implies that M1/Jj is a
-
-linear combination of columns of \II, so ran(\II) is invariant. We have thus proved the following
-% 15.txt
-Fact 3.1.1 The subspace ran(\II) is invariant for M if and only if there exists T E R”? such that
-
-M\II = \IIT.
-
-Consequently,
-
-Fact 3.1.2 )\(T) Q ).(M)
-
-Proof of 3.1.2:
-
-z\E)\(T) _=_ 3v€R"3Tv=).v
-<=> \IITv = A\IIv
-<=> M\II-u = /\\IIv
-< /\ E /\(M)
-
-The second ¢> follows from fact 3.1.1.
-Facts 3.1.1 and 3.1.2 are theoretically useful. To see why, write the equation of fact 3.1.1
-
-as follows:
-
-\IIT = M\II
-\II‘\IIT = \II‘M\II
-T = (\1:'~1r)-‘\1:'M\1:
-
-In this form, we recognize that T = C‘1(0)C(1). That is, using only the covariance of observables
-on the state space, we can generate a matrix T with the property A(T) (_Z ).(M). Recalling that
-M = H1/2Pl'I‘1/2, we see that M is similar to our original stochastic matrix P, and thus ).(M) =
-
-).(P).
-
-\section{Approximate Invariant Subspace}
-
-We qualiﬁed the foregoing by stating that the facts are only theoretically useful. This
-
-is because when choosing observables of the state space we aren't sure that they will satisfy the
-% 16.txt
-primary assumption underlying facts 3.1.1 and 3.1.2. Recall the assumption: ran(\II) is an invariant
-subspace for M, where ‘II = H1/2(<I> —  It may well be the case that there is a 1p; 6 ran(\II) such
-that M1/zj ¢ ran(\II), thereby violating the assumption. However, lacking an invariant subspace, we
-
-might hope that our observables at least provide an approximate invariant subspace.
-
-Deﬁnition 3.2.1 If the columns of ‘II E R“? are independent and the residual matrix E E
-
-M\II —— \IIT is small for some T 6 R7”, then ran(\II) deﬁnes an approximate invariant subspace.
-
-To see how an approximate invariant subspace can be useful in approximating eigenvalues of M,
-we extract a theorem from Golub and Van Loan 
-% Theorem 3.2.1
-\begin{theorem}
-\label{thm-3.2.1}  
-Suppose M 6 Rd” and T 6 RP"? are symmetric and let E E MQ — QT, where Q 6 Rd"? satisﬁes
-
-Q‘Q = I. Then there exist p1,... ,;z,, E A(T) and A1,. . . , A,, E A(M) such that
-ll‘k - Akl S \/§|lE"
-
-fork=1,...,p.
-\end{theorem}
-
-If the subspace ran(Q) is an approximate invariant subspace, then the deﬁnition implies that
-there is a choice T rendering the error IIEII2 small, and thus, the eigenvalues of T provide a good
-approximation to those of M.
-
-When considering the preceding ideas, it becomes apparent that their application will
-present new — but hopefully less prohibitive — problems. As these problems summarize the rt of
-the paper, now is a good time to examine them. First, the preceding theorem assumes a matrix Q
-whose columns form an orthonormal basis for the approximate invariant subspace. For our problem,
-derivation of such a Q is tricky, and we must prepare for this. Next, having an approximate invariant
-
-subspace at our disposal merely tells us that there exists some matrix T which makes the error
-% 17.txt
-"EH2 small. We must discover the form of such a T. Moreover, it is natural to seek that T which
-minimizes IIEII2 for a given approximate invariant subspace. Finally, in order to apply these ideas
-to real examples of our eigenvalue problem, we must ﬁnd a practical way to generate the most
-appropriate approximate invariant subspace. We now address each of these issues in the order
-raised.
-
-Recall the matrix I1 E III/2(<I> — (<I>)). The columns of this matrix, though independent
-(by choice of independent observables), are not necessarily orthonormal. However, consider the
-polar decomposition ‘II = QZ where Q‘Q = I and Z2 = \Il‘\II is symmetric positive semideﬁnite.1
-
-Note that Z = (\II‘\II)l/2 is nonsingular, so Q has the form
-Q = ~1:z-1 = \II(\II‘\II)“/2
-
-and it is clear that ran(Q) = ran(\II). Perhaps ra.n(Q) is a. useful approximation to the invariant
-subspace for M. If ran(Q) is not itself invariant, we have the error E = MQ — QT. It is easy to
-show, as we do in the following proof, that the T which minimizes HE“; is T = Q‘MQ. This yields
-
-the following,
-Theorem 3.2.2
-
-If \I1 = QZ is the polar decomposition of \II, then the matrix
-T = Q‘MQ = (\1:*\1:)-1/’\1:‘M\I:(\Ir‘\I:)-1/2
-minimizes IIEII2 = IIMQ - QTII2
-
-Proof:
-We prove the result by establishing the following
-
-(Hal: Given M 6 Rd” suppose Q 6 Rd"? satisﬁes Q'Q =1. Then,
-
-min ||MQ — QTII2 = H(1- QQ')MQll2
-
-TERI‘?
-
-‘Recall that the polar decomposition is derived from the SVD, \II = UZJV‘, by letting Q = UV‘ and Z = VEV‘.
-
-% 18.txt
-and T = Q’MQ is the minimizer.
-To verify the claim, observe the following application of the Pythagorean theorem:
-
-For any T E R.”"” we have
-
-IIMQ — QT||§ = IIMQ - QQ'MQ + QQ'MQ - QT|l§
-||(I - QQ’)MQ + Q(Q'MQ - T) H3
-
-||(1 - QQ‘)MQ||§ + |lQ(Q'MQ - T)||§ (3-1)
-
-2 l|(I — QQ')MQ||§
-
-Equality (3.1) holds since I — QQ' projects MQ onto the subspace orthogonal to ran(Q). Thus,
-the two terms in the expression on the right are orthogonal, and the Pythagorean theorem yields
-equality. The concluding inequality establishes that the minimizing T is that which annihilates the
-second term in (3.1); i.e. T = Q‘MQ.
-
-The second equality in Theorem 3.2.2 is a consequence of the polar decomposition, in
-
-which Q = \IIZ“ = \I:(~I:‘\I:)-1/2. Therefore,
-T = (x1z'x1:)-1/2x1:‘M~1:(\1:*~1z)—1/2
-
-is the minimizer.
-
-3.3 The Krylov Subspace
-
-Suppose the columns of a matrix Q E R“? provide an orthonormal basis for an approxi-
-
-mate invariant subspace. Then we have seen,
-1. T = Q'MQ minimizes ||E||2 = ||MQ — QT||2 and
-2. E|p.1,...,p,, E z\(T) and /\1,...,A, 6 »\(M) such that
-
-ll-‘I: - Ml S \/§|lE||2
-
-% 19.txt
-19
-fork=1,...,p.
-
-Given an approximate invariant subspace ran(Q) of dimension 1), these facts tell us what matrix we
-should use to approximate p elements of M’s spectrum. Now all we lack is a. description of ran(Q).
-That is, we have not speciﬁed which approximate invariant subspace would best suit our objective
-
-of approximating the subdominant eigenvalue A,,.a2:(M). 'I‘o do so requires the following deﬁnition:
-
-Deﬁnition 3.3.1 The Raleigh quotient of a symmetric matrix M and a nonzero vector 3 is
-
-p(2:, M) E (:z:‘Ma:)/(z‘a:).
-
-We will also denote this by p(z) when the context makes clear what matrix is involved.
-
-To ﬁnd the approximate invariant subspace most appropriate for our problem, we choose
-each dimension successively, providing justiﬁcation at each step. Thus, start with one observable
-42 on the state space, and let zpl E 111/ 2(¢ — E,,¢). That is, 1/:1 represents a centered (mean zero)
-observable whose ith coordinate is weighted by . First notice that the deﬁnition (which comes
-directly from the deﬁnition of ‘II following equation (2.7)) is such that 1/)1 is a constant vector if
-and only if :15 is oonstant on the state space, in which case 1/); E 0. This fact provides one reason
-observables that are constant on the state space are not interesting. Second, recall that the row
-sums of the matrix P are all one and, therefore, the eigenvector corresponding to the eigenvalue
-Ao(P) = 1 is the constant vector. By deﬁnition of M E 1'11/2Pl'I‘1/2, we see that the constant vector
-is also the eigenvector corresponding to A0  This will play an important role in what follows, as
-it allows us to focus primarily on the subdominant eigenvalue z\max(M) E max{}q (M), [A44 
-rather than on A0  (which we already know is 1).
-
-Now, notice that
-
-lP(¢1,M)l 5 max |P(x,M)l = »\max(M) (3-2)
-% 20.txt
-where the max is taken over all non-constant vectors. Since our interest centers on An-.3_x(M), we
-would like a. 1/11 which makes the left hand side of (3.2) large. This would be achieved if 1/)1 were to
-lie in the space spanned by, say, the first two eigenvectors of the Markov chain (sometimes referred
-to as the slowest modes of the process). However, this subspace spans only two dimensions of the
-entire d-dimensional space, and it is more likely that ¢1, at best, only comes close to lying in the
-subspace of interest. Now, given 1/11, a judicious choice for the second dimension l/)2, and hence
-W; E [1/)1 1/:2], would be that which makes max,,¢o |p(\II2a)| is as large as possible. To establish that
-
-this is indeed the pertinent objective, note the following:
-
-_ a‘\1:5Mx1:2a
-I;1;g<Ip(\I'2a,M)l = 13:3: ————a,W§q,2a
-= M
-mrggf,mIp(z. )1
-5 maxlp(z,M)l (3-3)
-= ).,,m(M)
-
-Again, the max on the right side of (3.3) is over non-constant vectors. In words, we wish to chose
-\II-2 = [1/:1 ¢2] so that there is a vector a 6 R2 making |p(\II2a,M)| close to )«max(M).
-
-Now, p(1/11) changes most rapidly in the direction of the gradient Vp(¢1).
-
-W0  (iiiiiiv---viiiléi)
-= ,,;,l (Mam — p<¢.)«m (3.4)
-
-So, to maximize the left hand side of (3.3), ‘I12 should be chosen so that the subspace ra.n(\II2)
-
-contains the gradient vector. That is, we must choose 1/22 so that
-
-VP(¢'1) E F3l1{¢1y¢2} E F-'%!1(‘I’2) (3-5)
-
-Clearly, equation (3.4) implies Vp(1/:1) E ran{z/11,M1,b1}. Therefore, if ran{1/:1,1/)2} = ra.n{¢1,M¢1},
-
-then (3.5) is satisﬁed.
-% 21.txt
-In general, having chosen \II,, E [1/)1 - -- 1/11,] so that
-
-ran{'pl9¢21' " 1¢k} = ran{¢1vM1[)la ' ‘ ' vMk—11/’k}
-
-we must chose 1/1;.“ so that, for any non-zero vector u 6 R",
-
-VP(‘I’I:a) 6 ra‘n{¢11 ‘21 - ° - 31/Jk-F1} 
-Now,
-vp<~I~.a) = gmm — P(‘I’kﬂ)‘I’k“)
-
-and therefore,
-
-Vp(‘II;,a.) E ran(M\II;,) U ra.n(\IIk) = ra.n{¢1, M1/)1, . . . , Mktﬁl}
-
-Thus, requirement (3.6) is met when ran{1/:1, 1/:2, . . . , 1/4+1} = ran{1/)1, M1/:1, . . . , Mk:/11}.
-
-In conclusion, the p-dimensional approximate invariant subspace which is most suitable
-
-for our objective is
-
-/C(Mn/mp) 2 ran{«/:1,M¢1,... ,M"“«/21}
-
-This is known as the K rylov subspace. Therefore, in answer to the problem posed at the outset
-of this section, if we take the columns of Q to be an orthonormal basis for IC(M,z/)1, p), then the
-
-eigenvalues of T = Q‘MQ should provide good estimates of p extremal eigenvalues of M.
-
-% 22.txt
-\chapter{Lanczos Procedures}
-
-The following will review the usual procedure for generating an orthonormal basis for
-lC(M, «/11, p), and conclude by showing that, for our problem, we can find the matrix Q‘MQ without
-actually carrying out this procedure. Note that this conclusion is essential since the procedure
-
-requires matrix vector multiplication - an operation we have assumed impossible for large enough :1.
-
-4.1 The General Lanczos Algorithm
-Let 21;“) E M"¢ and consider the d x :1 matrix
-«I 2 [«/»,M«/»,  ,M"“¢] 2 [«b‘°>,«/»“’,...,¢<‘-‘>1
-
-Notice that M\II = [1/:(1),... ,1/;(d'1),Md¢r] and, assuming for the moment that \I1 is non-singular,
-
-we can compute the vector h = —\II‘1M"1/J. Thus,
-
-M\II = \II[e2,. . . ,e,1, —h] E IIIH
-
-% 23.txt
-23
-
-Equivalently,
-0 0  0 -—h1
-1 0  0 —h2
-HE   3 ; =\II“M\II (4.1)
-0 "hl1—1
-1 —h,;
-
-H is a companion matrix which means that its characteristic polynomial is p(:r) = 2:‘ + 23:1 h,-:r"‘1.
-Since H is similar to M, ﬁnding the eigenvalues of M is equivalent to ﬁnding the roots of p(a:).
-However, this is of little practical use since ﬁnding h, constructing 12(1), and ﬁnding its roots is
-probably a harder problem than the one we started with. Instead, the value of decomposition (4.1)
-derives from its upper Hessenberg form. We exploit this property below.
-
-Let ‘II = QR be the QR decomposition of \II. Since ‘I! is assumed nonsingular,
-~II;‘M~r.: = (R"Q‘)M(QR) = H
-Therefore,
-Q‘MQ = RHR-1 2 T
-
-Since R and R“1 are both upper triangular and H is upper Hessenberg, T = R.HR'1 is also upper
-Hessenberg. Furthermore, since M is symmetric, it is clear that T‘ = Q‘MQ = T. That is, T is
-
-an upper Hessenberg matrix that is symmetric. Therefore, T is tridiagonal and we can write it as
-
-follows:
-
-T =    (4.2)
-
-ﬁd—1
-
-0  174-1 ad
-
-% 24.txt
-24
-
-Equating columns j on both sides of MQ = QT yields
-M113‘ = B,-—1qj—1 + 0:91 + ﬁjqj+1 (43)
-
-Since the columns of Q are orthonormal, multiplying both sides of (4.3) by q_,- and q,-+1 yields
-oz, = q§Mq,- and ﬂj = q;-+1Mq,-.
-The foregoing justifies what is called the Lanczos algorithm, performed as follows:
-
-The Lanczos algorithm for (partial) reduction to symmetric tridiaganal form
-
-91 = ¢/||¢||2, [30 = 0, 40 = 0
-
-for j=1top
-
-z=Mq_,-
-
-5:‘ = ||Z||2
-if 5, = 0, quit
-¢Ij+1 = 2/51
-
-end for
-
-The qj computed by the Lanczos algorithm are often called the Lanczos vectors. If the
-loop in the algorithm is terminated because ﬁg = 0, this indicates that an exact invariant subspace
-has been computed, and is given by ran{q1 - - - qk}. Otherwise, we usually halt the algorithm after
-p steps, in which case the algorithm converg to approximations of at most 12 eigenvalues.‘
-
-Note that equation (4.3) results from the equation MQ = QT. The latter equation holds
-since, starting with d vectors in our subspace, Q is a d Xd matrix whose columns form an orthonormal
-
-basis for all of R‘, which is clearly an invariant “sub” space. The Lanczos algorithm above, however,
-
-‘A detailed discussion of such convergence (and misconvergence) is given in
-
-% 25.txt
-25
-
-proceeds for only p steps, producing a d x p matrix Q, whose columns form an orthonormal basis
-for the approximate invariant subspace IC(M, ¢, p), and a p X p matrix T, = Q'MQ. In that case,
-IIMQ = QTp|[g =  is non-zero and gives the error bound described in Theorem 3.2.1. Now,
-
-writing the full d x :1 matrix T of (4.2) as
-
-01 51
-/71
-
-(4.4)
-
-allows us to describe the error bound ||E||-2 in terms of the submatrix Tm, and, hence, in terms of
-
-ﬁzz-
-Theorem 4.1.1
-
-HT, and T,“ are the matrices appearing in (4.4), and if the p columns of Q are computed by the
-
-Lanczos algorithm, then there exist pl, . . . ,;t,, E /\(T,) and A1, . . . , A, E A(M) such that
-
-ill» - /\I=i S iiTP'4”2 = 512
-
-fork=1,...,p.
-
-For a proof, see page 365 of [1].
-
-% 26.txt
-26
-
-4.2 A Lanczos Procedure for Markov Chains
-
-Notice that the Lanczos algorithm requires the computation of Mqj. Throughout this
-paper we have assumed that M 6 Rd“, and that d is so large as to make the matrix vector
-multiplication Mqj impossible. Now, the Lanczos algorithm constructs an orthonormal basis for
-the Krylov subspace. For our purposes, this construction is not essential. What is essential is that
-we ﬁnd a way to generate the Lanczos coejﬁcients oz,-, ﬁj, (j = 1, . . . , p), and whence the matrix
-T, without performing the operation Mqj required by the Lanczos algorithm. We now address this
-problem.
-
-Begin with the centered and weighted observable 1/1 = 111/ 2 (qS—E,,¢), and let q1 = 11;/||1/JH2.
-To simplify notation, let ¢(n) E ¢(X,.). The form of the first Lanczos coeﬂicient a1 is straight
-
-forward:
-
-01 = qiMq1
-
-I/1‘M1/I
-¢’¢
-
-and now the special properties of the Markov chain setting play an important role. For, regard the
-
-following
-
-¢‘M«/2 = (¢ — E.¢)‘II%‘MII%(¢ - Em)
-
-Cw (d>(n), ¢(" + 1)) (4-5)
-
-and
-
-W1!’ = (<15 - E«¢)'H(¢ - E«¢)
-
-Varn (¢(")) (4-6)
-
-Recall that the deﬁnition of M implies H%‘MI'I% = HP, and this justiﬁes equation (4.5). Putting
-
-% 27.txt
-27
-
-these equations together, we have a. nice form for :11:
-
-_ c, (run), an + 1))
-"‘ ' Ta:,(¢(n))
-
-The next Lanczos coeﬂicient, B1, is only slightly more complicated.
-
-[31 = ||M91-0t1tI1||2
-
-[(Mq1 — 0l1¢11)i(M¢]1 - 0101)] 1/2
-
-[t1iM2¢I1 — ail 1/2
-
-(4.7)
-
-Again recalling the deﬁnition M E 111/ 2P1'I‘1/2, a. simple manipulation veriﬁes
-
-1'li‘M’l'[% = HP’, whereby it follows that
-
-(«s — E,,¢)‘HP”(¢ — Em)
-
-"‘M2"‘ (¢ — E,¢)*n<¢ — Em)
-
-C: (¢("), ¢(n + 2))
-V3-Ur
-
-Inserting the last expression into equation (4.7) gives
-
-_ c,(¢(n).¢<n+2)) c,(¢<n>,¢(n+1)> ’ "’
-"“ jvar,<¢<n)> "(jvar,<¢<n)) ‘)
-
-Notice that we can estimate these expressions of oz; and ﬂl by running simulations of the Markov
-chain. To do so, we run the Markov chain starting from a random initial state, compute the value of
-the observable at each step of the chain and, from the data, compute estimates of the covariances
-and variance of the observable.
-
-Before generalizing this for the jth Lanczos coefficient, we consider one more coeﬂicient
-in detail. Recall that the second basis element q; in the Lanczos algorithm is given by 1;; =
-
-(Mql — oz1q1)/ﬁl. From this we can derive an expression for (12 which, though not as neat as those
-
-% 28.txt
-28
-
-for oz; and [31, is nonetheless tractable.
-
-I12 ‘IiMq2
-
-1
-[7(M91 — 0t1¢I1)tM(M¢I1 - 11191)
-1
-
-1
-= f,§(qiM3q1 - 2a1qlM2<11 + a’9iM¢11)
-1
-1
-= —2(<1iM3q1 - 2015i ‘ all (4-8)
-
-1.
-The last equality follows from equation (4.7), which can be written as q{M2q1 = off’ + (if. The
-preceding paragraphs deﬁne all but one of the quantities in the expression (4.8). The one remaining
-is q§M3q1, which can also be expressed in terms of covariances of <15, and the analysis is almost
-
-identical to that for q{M2q1:
-
-(¢ — E,¢)'IIP3(¢ — Em)
-
-"lM3"‘ (¢ — E,¢s)‘H(¢ - Em»)
-
-Cu (¢(")a ¢(n + 3))
-Varn(¢(n))
-
-To generalize the foregoing for higher order Lanczos coeﬂicients requires a recurrence
-relation between (aj+1, ﬂj) and (aj, ﬂ,~_1). As we will see, such a relation involves higher order
-Lanczos vectors q,~ — vectors we do not wish to compute. Recall that the only “vector” we can
-handle is our initial observable. Therefore, we must also establish a recurrence among the q,-’s, thus
-making them available inductively from our observable vector.
-
-To begin with the ﬂfs, recall that ﬁj E ||Mqj — ajq_.,- — ,H_,~_1q,-_;|[2. Whence,
-
-Bf (M411 — °‘j‘Ij - ﬁj-1qj—1)’(MlIj - Otjqj - 51-19141)
-
-q;-M291‘ - 0? - B,z_1 (4-9)
-
-Some simple algebra and the deﬁnitions aj E q§Mqj and ﬁJ~_1 E q;-Mq,--1 prove equality (4.9). The
-% 29.txt
-recurrence for 0114.1 is equally straight forward, though the ﬁnal form is not as neat:
-
-2
-0tj+1 ¢Ij+1M'1j+1
-
-1
-§(M¢15 - 11111;‘ - ﬂj—1<Ij—1)'M(Mqj - Otjqj - /3j—1qj—1)
-J
-
-1
-§(q;-M-“*4; — 2a,~q;M*q,- — 2x3;+’_1q;--1M*q,- + «I? + 2a,-5,51 + am (4.10)
-J
-From (4.9), q;~M2q,~ = oz? + ﬂf + ﬂ}_1 and, upon plugging this into (4.10), we arrive at
-1
-a,~+1 = E(q§»M3q_,,- — (1? — 2ozJ~ﬁJ2 — 2ﬂj_1q;»_1M2qj + ﬁ2_1) (4.11)
-J
-
-The expressions above involve the higher order Lanczos vectors qj. In the present context,
-these vectors will not be readily available. However, another recursion — this time relating forms
-involving q,-+1 to forms involving qj and qj__1 — will make these expressions accessible by induction
-on our initial observable. The general forms of interest are q;+1M"qj+1 and qj-+1M"qJ~. Again, we
-
-begin with the simpler of the two:
-
-¢I;'+1Mk9j = ij(Mqj - Otjqj - ﬂj—1t1j—1)'M"¢1j
-= i(<1§M"“¢Ij - C‘j‘I_;'Mk‘Ij - ﬂj—1q§M"qj-1)
-The second is,
-‘I_;'+1Mkqj+1 = ﬂi?(M<Ij - Otjqj - .3j—1<Ij—1)‘Mk(Mqj — Orjqj - ﬂj—1¢1j—1)
-
-1
-ﬁ(‘1§M'°"<1:‘ + °‘:2"I§M'°<J:' + ﬁ?—z<Ii—1M"<I:‘—1
-J
-—2a,-q;-M*+‘q,~ + 2am,--1q;-M*q,--1 — zza,--1q;-M*+‘q]--1) (4.12)
-
-The point of all this is that each pair of coeﬂicients (oz,-+1, /3,-) can be expressed as a
-function of the covariance between ¢(n) and ¢(n+ m), for m = 0, . . . , 21' + 1. Since the expressions
-are based on four recurrence relations, they become quite complicated as 1' gets large. However,
-deﬁning these recursions in a. computer program is trivial, and we use this fact in Section 5.4 when
-
-applying the foregoing to a particular problem.
-
-% 30.txt
-30
-
-Chapter 5
-
-An Application: Convergence to
-
-Gibbs Measure on the Ising Lattice
-
-5.1 Physical Motivation
-
-Before considering the details of the Metropolis algorithm, it helps to understand the
-setting in which it was developed. To do so we consider an typical example, the Ising model.
-Consider a model of a system consisting of 71 “sites”, each site taking values in {-1, +1}. It might
-help to imagine these sites as equally spaced points on a line or a circle, though we are not restricted
-to such cases. Then our state space 5' consists of all possible n—dimensional permutations of -1’s
-and +1’s. We reprent this as S’ = {-1, +1}", and note that |S| = 2". We can think ofeach site as
-a node existing in one of two possible positions, say “on" or “off”. Each state 1' E S is represented
-by a unique permutation a.- of —1’s and +1’s. Let a,-(k) = :|:l denote the position (on or off) of
-the kth node when the system is in state 1'. An observable ¢ = qS(a,-) on this space is a function of
-
-the permutations 47,-, 1' 6 {1,. . . , 2"}.
-
-% 31.txt
-31
-
-The energy of the system, when in state i, is deﬁned by the Hamiltonian
-
-H07-‘) = - Z Ua(i)0s(k)
-
-<j Ic>
-where the sum runs over all neart neighbor pairs. In statistical mechanics we are often concerned
-
-with the Gibbs measure of state i, deﬁned by
-1r(a,-) = Z'1 exp{—[3H(a,~)} (5.1)
-where Z ‘1 is the normalization constant, or partition function. For the Ising model, we put
-
-Z = §:9XP{~ﬁH(<7:‘)}
-i=1
-
-so that 7r is a probability measure. Now, letting 1: denote the so called Boltzman constant, if a
-classical mechanical system is in thermal equilibrium with its surroundings with absolute temper-
-ature T, and is in state i with energy H (05), then the probability density in phase-space of the
-point representing state i is given by (5.1) with [3 = (lcT)“. The central result of ergodic theory
-implies that we can also interpret 1r(o;) as the proportion of time the system spends in state i. If
-
-the system is observed at a random time, the expectation E,,¢ of any observable ¢ is thus
-
-.31 ¢(0a) 9"l>{-ﬂH(0s‘)}
-
-E” = 2?;.exp{—nH(a.-)}
-
-(5.2)
-
-Perhaps the number of sites n in our Ising model is so large that it is impractical or impossible to
-evaluate (5.2). We might instead consider importance sampling and generate states with probability
-
-density
-
-exp{—aH<a.o}
-2?;.exp{—zaH<a.->} (53)
-
-Then «I: is itself an unbiased estimator of (5.2).
-
-1r(0a') =
-
-If evaluation of (5.2) is difficult, there is no reason to believe that evaluation of (5.3) is any
-
-easier. However, Metropolis and his collaborators [9] contrived a method for producing an ergodic
-
-% 32.txt
-32
-
-Markov chain with transition probability matrix K whose elements Ic(:::, y) satisfy
-
-2«<z>k<x.y) = «<2»
-
-I
-
-By Deﬁnition 2.1.4 this means that the Markov chain has stationary distribution 7r. Since the
-drain is ergodic, we know that it will eventually converge to the desired distribution. So, our main
-problem — the rate at which such a Markov chain will converge — is clearly at issue in the application
-of the Metropolis algorithm. Moreover, as we will see below, the Metropolis algorithm ensures that
-the Markov chain has the desired stationary distribution by requiring that the chain satisfy the
-stronger condition of reversibility; i.e. detailed balance. Therefore, Markov chains produced by the
-Metropolis algorithm satisfy the assumptions of this paper, which fact conﬁrms that our theory
-
-can be used to bound convergence rates of all such chains.
-
-5.2 Description of the Metropolis Algorithm
-
-The following explains how the Metropolis Algorithm is carried out. The exposition is
-similar to that given in Hammersley and Handscomb’s book Monte Carlo Methods 
-To begin, choose an arbitrary symmetric Markov chain; that is, a Markov chain whose
-
-transition probability matrix P is symmetric. P is called the proposition matrix, and its elements,
-P071"), Satisfy
-pa, 1) 2 0. pm‘) = pm). 2,-pm‘) = 1 vi e S (5-4)
-
-Using P, we will derive a new transition probability matrix K which satisﬁes
-
-Z vr(z')k(='.:') = arm (5.5)
-
-I
-
-For 1' aé j deﬁne
-
-kw) = p<-yum")/«<¢> it nu)/1r(='> <1 (5.6)
-
-P(i,J') if 7r(J')/7r(i) Z1
-
-% 33.txt
-33
-
-For 2' = j deﬁne
-
-kw) = pm) (1—  p(-3;"), where J. = {j = 2‘ 7‘ an nu)/«(='> < 1}
-
-Notice that 1r(i) > 0 implies k(i,j) _>_ 0 and
-
-;k<«',j> = pow)  { (1 —  pm) +  +]§§p<«",:')
-= P(is'5) + 2 P(i1.1') + Z P(iy.7') (5-7)
-jel.» jug
-
-Equation (5.7) shows that, for each i, 2, k(z', j) = 1, which fact conﬁrms that K is a stochastic
-
-matrix.
-
-Next we check that K satisﬁes the detailed balance condition. First suppose that 1r(z') =
-
-1r(j). Then (5.4), (5.6), and symmetry of P imply
-k(i,J') = p(i2.7') = p(jvi) = k(j1i)
-
-and, since 1r(i) = 1r(j), the detailed balance condition, 1r(i)k(i,j) = 7r(j)k(i,j), clearly holds. Now
-
-suppose that 1r(i) > 1r(j). Then (5.4) and (5.6) imply k(j,i) =p(j, 1') and
-k(i,J') = P(iyJ')7r(J')/7r(i) = P(J3i)7r(J')/7r(i) = k(.1'» i)7r(J')/1r(i)
-
-That is, 1r(i)lc(i,j) = 1r(j)k(i,j), as we set out to prove.
-To complete the justification of the Metropolis algorithm, we note that the detailed balance
-condition, 7r(i)k(i, j) = 1r(j)k(i, j), implies that 1r is indeed the stationary distribution for a Markov
-
-chain with transition probability matrix K. To see this notice that
-
-27r(i)k(i..7') = Zﬂijlkiiajl = 1r(J') Z/€(i,J‘) = 7r(J')
-
-I i
-We now summarize the tasks performed when implementing the Metropolis algorithm.
-
-First, pick as a proposition matrix any symmetric stochastic matrix and denote it by P. Begin
-
-% 34.txt
-34
-
-the simulation of the Markov chain in state i, and then take one step of the proposition Markov
-chain to arrive in state j. That is, choose j according to the mass function p(z', -). Next, compute
-1r(j)/1r(i). If 7r(j)/1r(i) 2 1, accept j as the new state. If1r(j)/1r(i') < 1, then with probability
-1r(j)/7r(z') accept j as the new state; otherwise, with probability 1 — 1r(j)/1r(i), take i as the new
-state. Performing each of these tasks produces one step of the Markov chain represented by the
-
-transition probability matrix K.
-
-5.3 The Ising Model
-
-Returning to the Ising model considered in section 5.1, we consider how the Metropolis
-algorithm is applied to such a mode]. Again, start the simulated chain in state X0 = 1' and let
-
-X1 = j be a state chosen according to p(i, -), where p(i, j) = p(j,i). Next, compute
-
-7"(‘7j)/7"(‘7i) = eXP{-ﬁlH(<7j) - H(0i)l}
-
-Suppose the energy H (Uj) of the new state satisﬁes H (U,-) 3 H (05). Then 1r(a_,-)/1r(a,-) Z 1 and we
-
-move to state j. On the other hand, if H(aj) > H(a;), then 1r(aj)/7r(a,-) < 1 and we take
-
-j with probability e-M"
-
-X1 (5.8)
-
-i with probability 1 — e-W’
-where AH E H (0,) — H (:75) is the change in energy from state j to state 1'.
-There are many ways to choose a proposition matrix P. The simplest derives from the
-so called Glauber dynamics, which proceeds as follows: choose one of the 11. sites at random with
-probability 1/n — let .9 denote the chosen site — and change the position of this site; i.e., multiply
-
-its value by -1. The resulting state j is the permutation of -1’s and +1’s given by
-
--0,-(k) if]: = .9
-a,~(Ic)= (k=1,...,d)
-
-a,-(Ic) if]: 95 s
-
-% 35.txt
-35
-
-The procedure we employ below involves a slight variation: choose site .9 at random uniformly
-from the 11 possible sites. However, instead of changing the value of the chosen site with absolute
-certainty, as in Glauber dynamics, we “ﬂip the switch” with probability 1/2. Therefore, when k yé .9
-
-we have 0',-(k) = a.~(k), and when k = 3,
-
-—a,~(s) with probability 1/2
-01(3) =
-
-a,-(3) with probability 1/2
-The advantage of this procedure is that the proposition matrix, before altering it with the Metropo-
-lis algorithm, is a bit more interesting than that produced by traditional Glauber dynamics. For,
-it has 1/2 along the main diagonal and, thus, is aperiodic and converges to a stationary distribu-
-tion. Glauber dynamics has period 2, so its proposition matrix does not satisfy the hypotheses of
-
-our convergence theorem (Theorem 2.1.2). Keep in mind, however, that any proposition matrix
-
-modiﬁed by the Metropolis algorithm satisﬁes all of our hypotheses.
-
-5.4 Computations
-
-5.4.1 Computer Programs
-
-Constructing a computer program to simulate realizations of the Markov chain described
-above (i.e. with transition matrix K) is straight forward, and we do so using Matlab and the
-program ising.m which appears in the Appendix.
-
-When performing simulations in the present context, there are a few important aspects
-to consider. First, our program begins with a “hot start", which means that the initial state is
-picked by assigning each site to :l:1 with probability 1/2. Perhaps we should instead have picked
-our initial state from the stationary distribution. On the contrary, this would defeat the whole
-
-purpose of our analysis. Recall that the method has been developed to deal with cases for which
-
-% 36.txt
-36
-
-sampl from the stationary distribution are not quickly attainable. Furthermore, we are trying to
-ﬁgure out how long it will take for observations from the simulated chain to represent samples from
-the stationary distribution. Second, the method we have developed depends solely on covariances
-of our observable which we estimate from the data. produced by the simulations. However, such
-estimates require that the data come from the stationary distribution. Therefore, we must discard a
-number of observations before using the data to estimate covariances. The number of observations
-to discard depends on how long must we wait until we have data from the stationary distribution,
-and we are back where we started — almost. When employing our procedure, we can discard a
-conservative (very large) number of observations and perform our analysis, deriving a bound on
-the convergence time. Then, in all future studies, we will know how much data should be discarded
-before samples come (approximately) from the stationary distribution.
-
-In considering what observable to use on the state space described above, perhaps the
-most obvious is simply the number of nodes in the on position (i.e. the number of +l’s). Recalling
-that 0, (k) = i1 denotes the position (on or off) of the kth node when the system is in state j, this
-
-observable is deﬁned by,
-«M = zxca.-ac) = +1). 1 = 1.---.ISl
-I:
-
-where X is the indicator function. The program ising.m simulates the Markov chain, computes the
-values of this observable (as well as the number of 1’s squared) and writes it to a ﬁle called phi . dat.
-The program then proceeds to compute the ﬁrst few Lanczos coefficients using covariances produced
-by Matlab, so that we can check them against the rults we get from the 1anczos.c program,
-which we now describe.
-
-The lanczos .c program (listed in the Appendix along with its dependencies) implements
-the recursive relations described in section 4.2 to compute the Lanczos coefﬁcients by the new
-
-method. It accepts input from the user specifying any observable, any number of iterations, and any
-
-% 37.txt
-37
-
-TABLE 1: ESTIMATES or /\,m(K) = 0.998746 USING ¢1
-
-sim. # /32
-1 0.995107 0.006960 0.995220 0.600243
-2 0.995099 0.005891 0.994863 0.477393
-3 0.994862 0.009671 0.994925 N/ A
-4 0.995321 0.009319 0.995654 N/A
-5 0.994787 0.004872 0.405446 0.994742
-
-number of desired Lauczos coeﬁicients. That is, the observable could have come from a simulation
-of any reversible Markov chain and not just our special Metropolis chain. The output of this
-program is a T matrix containing the maximum number of Lauczos coeﬂicients (up to the number
-requested by the user) before an approximate invariant subspace was reached. The eigenvalues of
-the nonzero part of this matrix, in accordance with the foregoing theory, should well approximate
-
-eigenvalues of the Markov chain’s stochastic matrix.
-
-5.4.2 Results
-
-Table 1 shows the results of 5 simulations each of 100,000 iterations (discarding the ﬁrst
-5,000 observations) for the one dimensional Ising lattice with 10 nodes and [7 = 1. Displayed is
-the estimate of the ﬁrst Lauczos coeﬂicient a1 (which is the eigenvalue estimate after one step; i.e.
-).m,x(T1)), along with its error bound ﬁl. Appearing in the third column is the largest eigenvalue of
-the 2 x 2 matrix T2, A0 (T2), and ﬁg appears in the fourth column when it is was computable. When
-it was not computable, it was probably because an approximate invariant subspace was reached
-at the ﬁrst step, so ﬁg involved expressions which were close to machine epsilon and, thus, could
-not be computed accurately. Table 2 shows the same information when using a second observable,
-¢2 E ¢§ (i.e. the number of +1’s squared).
-
-The true value of the subdominant eigenvalue for this problem is /\max(K) = 0.998746,
-
-and the next largest is A2(K) = 0.993303. These were computed directly from the transition matrix
-
-% 38.txt
-38
-
-TABLE 2: ESTIMATES OF >.,,,,,.,,,(K) = 0.998746 USING z/>2
-
-sim. # B2
-1 0.994262 0.006142 0.994222 N / A
-2 0.994333 0.002974 0.994332 N / A
-3 0.994307 0.008892 0.994374 0.726320
-4 0.994436 0.012884 0.994685 0.361404
-5 0.993789 0.006909 0.993820 0.788710
-
-using the Matlab program tpm.m listed in the Appendix. They should be very accurate, and it
-takes the Matlab routine eig() just over 20 minutes of cpu time on a Sun Ultra to ﬁnd all the
-eigenvalues of K.
-
-As shown by Theorem 4.1.1, the [Ts provide conservative error bounds on the eigenvalue
-estimates from the previous Lanczos step. Unfortunately, for this example, ,6 appears to be too
-conservative as it does not even allow us to bound our estimates away from 1. Furthermore, it
-seems that the estimates generated by the method above fall systematically below the true value of
-the subdominant eigenvalue. In every simulation, an approximate invariant subspace (to machine
-tolerance) was reached after computing at most 2 pairs of Lanczos coefficients. For ¢1 we see that
-it was reached after computing only one pair on simulations 3 and 4, similarly for ¢2 on simulations
-1 and 2. These observations indicate that the chosen observable is close to the slowest mode (or
-eigenfunction) of the process. That is, we should observe fast convergence on an approximate
-invariant space. However, the space found might contain only the second slowest mode of the
-process and, in that case, our eigenvalue estimates would be closer to the third largest eigenvalue,
-instead of the second largest as desired.‘
-
-The matrix T was produced by the program lanczos . c and its eigenvalues were computed
-by Matlab, both operations taking a few seconds. Granted, comparing a few seconds to the 20
-
-minutes it takes Matlab is not justiﬁed since a simple call to the routine eig() is far from the
-
-‘Recall that we explained above why we will never converge to the eigenspace corresponding to the largest eigen-
-value (which is always 1). That would require a (nearly) constant observable function.
-
-% 39.txt
-39
-
-most eﬂicient traditional method for computing only a few eigenvalues of K. However, recall that
-the primary motivation for the new method are those examples where the matrix K is so large
-that we can’t store even a single vector in fast memory. Once we start swapping data to and from
-cheap memory, the time required by traditional algorithms can become years, even lifetimes. Such
-examples are not exceptional. Consider, for instance, the 1000 node problem which produces a.
-
-2“’°° x 21°°° transition matrix.
-
-% 40.txt
-40
-
-Chapter 6
-
-Conclusion
-
-We conclude by noting a few aspects of the special cases to which the foregoing applies.
-We are interested in the subdominant eigenvalue of a large transition probability matrix. The
-
-foregoing can be used to approximate these eigenvalues when the following conditions are satisﬁed:
-o The Markov chain is reversible (detailed balance condition).
-
-0 We can easily simulate realizations of the chain as well as functions (observables) of these
-
-realizations.
-
-When these conditions hold, we can side step the traditional Lanczos algorithm, and get at the
-values of the Lanczos coeﬂicients via the variational properties of observables on the state space.
-There are many ideas in this paper that should be expanded upon in future studies. Per-
-haps most obvious is the desire to ﬁnd analogous techniques for non-reversible (i.e. non-symmetric)
-Markov chains. This is a considerable problem since the Lanczos methods described above are no
-longer applicable. Perhaps the Amoldi algorithm ([1], algorithm 6.9), or the non-symmetric Lanczos
-
-algorithm can also be usefully specialized for stochastic matrices.
-
-% 41.txt
-41
-
-When applying the Lanczos algorithm, we usually choose a single starting vector (in the
-present context, an observable), perform the algorithm for a number of steps, and gather results.
-Then we do the same with a new starting vector and compare the results to those obtained with
-the ﬁrst vector. Proceeding in this way for a number of starting vectors, we can provide statistical
-evidence that we have, indeed, located the eigenvalues of interest. However, when there are two
-observables 451, :15; of particular interest, perhaps we would beneﬁt by considering both observables
-simultaneously. Future studies might develop an algorithm which incorporates both observables,
-
-again via ¢.- = H1/2(¢.' - E,.¢.-). In this case, a Lanczos algorithm would be based on the space
-
-Ic*(M,¢1,¢»2.p) 2 ran{«/»mz»2,M«/»1,M«/:2, . . .,MP—‘¢1.M"—‘«/22}
-
-The Lanczos coeﬁicients would then involve not only the autocovariances C, (¢,<(n),¢,-(n + m)),
-
-but also the cross-covariances C, (¢.~(n), ¢j(n + m)) (deﬁned in (2.6)).
-
 % 42.txt

### 2013-10-18 tex/Application.tex
initial commit of tex source for nyu thesis style setup

--- 2013-10-18 tex/MSThesis.tex
@@
-%\documentclass[11pt]{amsart}
-\documentclass{au}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% For completion by editorial office:
- \presentedby{\dots}
- \received{\dots}{\dots}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\usepackage[colorlinks=true,urlcolor=black,linkcolor=black,citecolor=black]{hyperref}
-
-\usepackage{amsmath,amssymb,url,mathrsfs}
-\usepackage{enumerate}
-\usepackage{tikz}
-\usetikzlibrary{calc}
-\usepackage{scalefnt}
-\numberwithin{equation}{section}
-
-\theoremstyle{plain}
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{proposition}[theorem]{Proposition}
-\newtheorem{corollary}[theorem]{Corollary}
-
-\theoremstyle{definition}
-\newtheorem{definition}[theorem]{Definition}
-\newtheorem{notation}[theorem]{Notation}
-\newtheorem{remark}[theorem]{Remark}
-\newtheorem{example}[theorem]{Example}
-\newtheorem*{remarks}{Remarks}
-
-\newcommand{\todo}[1]{~\\[5pt]{\bf TODO(wjd):} #1\\[5pt]}
-%\newcommand{\suchthat}{\ensuremath{:}}    % (old notation)
-\newcommand{\suchthat}{\ensuremath{\mid}}  % (new notation)
-%\newcommand{\defeq}{\ensuremath{:=}}      % (old notation)
-\newcommand{\defeq}{\ensuremath{=}}        % (new notation)
-\newcommand{\betai}{\ensuremath{\beta^i}}  % \beta^{\bB_i}   % (old notation)
-\newcommand{\betaj}{\ensuremath{\beta^j}}  % \beta^{\bB_j}
-\newcommand{\betak}{\ensuremath{\beta^k}}  % \beta^{\bB_k} or \beta^{\bB_{k}}
-\newcommand{\betal}{\ensuremath{\beta^\ell}}  %\beta^{\bB_\ell} or \beta^{\bB_{\ell}}
-\newcommand{\uacalc}{{\small UAC}alc}
-\newcommand{\Tuma}{T\r{u}ma}
-\newcommand{\Palfy}{P\'alfy}
-\newcommand{\Pudlak}{Pudl\'ak}
-\newcommand{\PP}{P\'alfy-Pudl\'ak}
-\newcommand{\PAP}{P\'alfy\ and Pudl\'ak}
-\newcommand{\<}{\ensuremath{\langle}}
-\renewcommand{\>}{\ensuremath{\rangle}}
-\newcommand{\beps}{\ensuremath{\boldsymbol{\varepsilon}}}
-\newcommand{\bN}{\ensuremath{\mathbf{N}}}
-\newcommand{\bA}{\ensuremath{\mathbf{A}}}
-\newcommand{\bB}{\ensuremath{\mathbf{B}}}
-\newcommand{\bBi}{\ensuremath{\mathbf{B}_i}}
-\newcommand{\sB}{\ensuremath{\mathcal{B}}}
-\newcommand{\sC}{\ensuremath{\mathcal{C}}}
-\newcommand{\sS}{\ensuremath{\mathscr{S}}}
-\newcommand{\sT}{\ensuremath{\mathscr{T}}}
-\newcommand{\sI}{\ensuremath{\mathcal{I}}}
-\newcommand{\sE}{\ensuremath{\mathcal{E}}}
-\newcommand{\sO}{\ensuremath{\mathcal{O}}}
-%\newcommand{\id}{\ensuremath{\mathrm{id}}}
-\DeclareMathOperator{\id}{id}
-%\newcommand{\Eq}{\ensuremath{\mathrm{Eq}}}
-\DeclareMathOperator{\Eq}{Eq}
-%\newcommand{\Cg}{\ensuremath{\mathrm{Cg}}}
-\DeclareMathOperator{\Cg}{Cg}
-%\newcommand{\Con}{\ensuremath{\mathrm{Con\,}}}
-\DeclareMathOperator{\Con}{Con}
-%\newcommand{\Sub}{\ensuremath{\mathrm{Sub}}}
-\DeclareMathOperator{\Sub}{Sub}
-%\newcommand{\Pol}{\ensuremath{\mathrm{Pol}}}
-\DeclareMathOperator{\Pol}{Pol}
-%\newcommand{\Clo}{\ensuremath{\mathrm{Clo}}}
-\DeclareMathOperator{\Clo}{Clo}
-\newcommand{\ps}[1]{\ensuremath{^{(#1)}}}
-\newcommand{\piB}{\ensuremath{\pi_B}}
-\newcommand{\hpsi}{\ensuremath{\widehat{\psi}}}
-\newcommand{\htheta}{\ensuremath{\widehat{\theta}}}
-\newcommand{\supi}{\ensuremath{^{i}}}
-\newcommand{\supj}{\ensuremath{^{j}}}
-\renewcommand{\leq}{\ensuremath{\leqslant}}
-\renewcommand{\nleq}{\ensuremath{\nleqslant}}
-\renewcommand{\geq}{\ensuremath{\geqslant}}
-\newcommand{\meet}{\ensuremath{\wedge}}
-\newcommand{\join}{\ensuremath{\vee}}
-\newcommand{\Meet}{\ensuremath{\bigwedge}}
-\renewcommand{\Join}{\ensuremath{\bigvee}}
-\newcommand{\resB}{\ensuremath{|_{_B}}}
-\newcommand{\resBi}{\ensuremath{|_{_{B_i}}}}
-\newcommand{\eps}{\ensuremath{\varepsilon}}
-\newcommand{\hatmap}{\ensuremath{\widehat{\phantom{x}}}}
-\newcommand{\one}{\ensuremath{\mathbf{1}}}
-\newcommand{\two}{\ensuremath{\mathbf{2}}}
-\newcommand{\three}{\ensuremath{\mathbf{3}}}
-\newcommand{\tbeta}{\ensuremath{\widetilde{\beta}}}
-\newcommand{\ttheta}{\ensuremath{\widetilde{\theta}}}
-\newcommand{\hbeta}{\ensuremath{\widehat{\beta}}}
-\newcommand{\cick}{\ensuremath{\sC_i}}
-%\newcommand{\CE}{\ensuremath{\sC_r^{\sE}}}
-\newcommand{\CE}{\ensuremath{\sB_r^n}}
-\newcommand{\CO}{\ensuremath{\sC_r^{\sO}}}
-\newcommand{\CICK}{\ensuremath{c_i/\beta \cup c^{K+1}_i/\beta^{\bB_{K+1}}}}
-\newcommand{\GAP}{{\small GAP}}
-\newcommand{\dotsize}{.8pt}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{document}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%  FRONT MATTER                                                 %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\title[Approximating Eigenvalues of Large Stochastic Matrices]{A Lanczos Procedure for Approximating Eigenvalues of Large Stochastic Matrices}
-
-%% First author (Note: The order of the items here is important!)
-\author{William DeMeo}
-\email{williamdemeo@gmail.com}
-\urladdr{http://www.math.sc.edu/~demeow}
-\address{Department of Mathematics\\
-University of South Carolina\\Columbia 29208\\USA}
-
-\thanks{The author wishes to thank a number of people who contributed to this
-  work.  In particular, the primary referee made many insightful recommendations
-  which led vast improvements, not only in the overall exposition, but also in
-  the specific algebraic constructions, theorem statements, and proofs.
-  This article is dedicated to Ralph Freese and Bill Lampe, who declined to be
-  named as co-authors despite their substantial contributions to the project.}
-
-%% Dedication (Optional)
-\dedicatory{This article is dedicated to Ralph Freese and Bill
-  Lampe.}
-
-%% AMS subject classification; see http://www.ams.org/msc
-%% Only one Primary. Possibly several Secondary.
-\subjclass[2010]{Primary: 08A30; Secondary: 08A60, 06B10.} %, 06B15.}
-
-%% Keywords and phrases
-\keywords{congruence lattice, finite algebra, finite lattice
-  representations}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%  MAIN MATTER                                                  %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{abstract}
-The rate at which a Markov chain converges to a given probability distribution
-has long been an active area of research. Well known bounds on this rate of
-convergence involve the subdominant eigenvalue of the chain’s underlying
-transition probability matrix. However, many transition probability matrices are
-so large that we are unable to store even a vector of the matrix in fast
-computer memory. Thus, traditional methods for approximating eigenvalues are
-rendered useless.
-
-In this paper we demonstrate that, if the Markov chain is reversible, and we
-understand the structure of the chain, we can derive the coefficients of the
-traditional Lanczos algorithm without storing a single vector. We do this by
-considering the variational properties of observables on the chain’s state
-space. In the process we present the classical theory which relates the
-information contained in the Lanczos coefficients to the eigenvalues of the
-Markov chain. 
-\end{abstract}
-
-\maketitle
-
-%%%%%%%%%%%%%%%%%%%%%
-\chapter{Introduction}
-The rate at which a Markov chain converges to a given probability distribution has long
-been an active area of research. This is not surprising considering this problem’s relevance to the ar-
-eas of statistics, statistical mechanics, and computer science. Markov Chain Monte Carlo (MCMC)
-algorithms provide important examples. These algorithms come in handy when we encounter a
-complicated probability distribution from which we want to draw random samples. In statistical
-mechanics, we might wish to estimate the phase average of a function on the state space. Goodman
-and Sokal [6] examine Monte Carlo methods in this context. Examples from statistics occur in the
-Bayesian paradigm when we are forced to simulate an unwieldy posterior distribution (see, e.g.,
-Geman and Geman
-
-To implement the MCMC algorithm, we invent a Markov chain that converges to the
-desired distribution (this is often accomplished using the Metropolis algorithm
-described in Chapter 5). Realizations of the chain will eventually represent
-samples from this distribution. Sometimes ``eventually'' -- meaning all but
-finitely many terms of the chain -- is just not enough. We need more practical
-results. In particular, we want to know how many terms of the chain should be 
-discarded before we are sampling from a distribution that is close (in total variation distance) to
-% 03.txt
-the distribution of interest. This is the purpose of bounding convergence rates for Markov chains.
-
-Often the Markov chains encountered in this context satisfy a condition known in the
-physics literature as detailed balance. Probabilists call chains with this property reversible. This
-simply means that the chain has the same probability law whether time moves
-forward or 
-backward.\footnote{This is not a precise definition. In particular the chain must
-  have started from its stationary distribution. Full rigor is postponed until Section 2.1.}
-In this paper, we consider the rate at which such chains converge to a \emph{stationary distribution}.\footnote{This and other italicized terms are defined in Section 2.1.}
-
-There are a number of different methods in common use for bounding convergence rates of
-Markov chains, and a good review of these methods with many references can be found in  More
-recently developed methods, employing logarithmic Sobolev inequalities, are reviewed in  Most
-of the bounds in common use involve the sub—dominant eigenvalue of the Markov chain's transition
-probability matrix, and thus require good approximations to such eigenvalues. In many applications,
-however, the transition probability matrix is so large that it becomes impossible to store even a
-single vector of the matrix in conventional computer memory. These so called out-of-core problems
-are not amenable to traditional eigenvalue algorithms\footnote{By ``traditional
-  eigenvalue algorithms'' we refer to those found, for example, in Golub and Van
-  Loan[5]. See also the book by Demmel [1] for a more recent discussion.}
-without modification. This paper develops 
-such a modification for the Markov chain eigenvalue problem. In particular it develops a method
-for approximating the first few eigenvalues of a transition probability matrix when we know the
-general structure of the underlying Markov chain. The method does not require storage of large
-matrices or vectors. Instead we need only simulate the Markov chain, and conduct a statistical
-analysis of the simulation.
-
-Here is a look at what follows. Section 2.1 contains a review of the relevant Markov chain
-theory. Readers conversant in the asymptotic theory of Markov chains might wish to at least skim
-Section 2.1, if only to become familiar with our notation. Section 2.2 describes functions on the state
-% 04.txt
-space of the Markov process. This section and Chapter 3 develop the context in which we formulate
-the new ideas of the paper. In the last section of Chapter 3, Section 3.3, we present the familiar
-Krylov subspace and explain why this represents our best approximation to a subspace containing
-extremal eigenvectors of the transition probability matrix.\footnote{or, more
-  precisely, a similarity transformation of this matrix.} The first section of
-Chapter 4 describes the \emph{Lanczos algorithm} for generating an orthonormal basis
-for the Krylov subspace. As it stands, this algorithm is useless for an
-out-of-core problem such as ours since, by definition of such problems, 
-it requires too much data movement; all the computing time is spent swapping data between slow
-and fast memory (e.g. between the hard disk and cache). Therefore, we discuss alternatives to
-Lanczos and demonstrate that the \emph{Lanczos coeficients} are readily available through simulations of
-the Markov chain, which fact allows us to avoid the standard algorithm altogether. Following this
-is a chapter describing the Metropolis algorithm used to produce a reversible stochastic matrix.
-It is here that we experiment with the procedure described in Section 4.2 and approximate the
-extremal eigenvalues of the matrix, without storing any of its vectors. Finally, Chapter 6 concludes
-the paper.
-% 05.txt
-
-\chapter{Markov Chains}
-
-\section{General Theory}
-
-This review of Markov chain theory can be found in any good probability text. The present
-discussion is most similar to that of Durrett [3], to which we refer the reader desiring greater detail.
-
-\subsection{The Basic Setup}
-
-Heuristically, a Markov chain is a stochastic process with a lack of memory property. Here
-this means that the future of the process, given its past behavior and its present state, depends only
-on its present state. This is the probabilistic analogue of a familiar property of classical particle
-systems. Given the position and velocities of all particles at time t, the equations of motion can be
-completely solved for the future evolution of the system. Thus, information describing the behavior
-of the process prior to time t is superﬂuous. To be a bit more precise, if technical, we need the
-following definitions.
-
-\bibliographystyle{spmpsci}
-\bibliography{wjd}
-
-\end{document}
-
-%%%%%%%%%%%%%%%%%%%%%  END DOCUMENT %%%%%%%%%%%%%%%%%%%%%%%%%
-
-Definition 2.1.1 Let (S, 8) be a measurable space. A sequence X,., n 2 0, of random variables
-
-taking values in S is said to be a Markov chain with respect to the filtration a(Xo, . . . ,X,,) if for
-% 06.txt
-a.llBES,
-
-p(x,,+1 e B | a(Xo, . . . ,X,.)) = P(X,.+1 e B 1 o(X,,)) (2.1)
-
-Equation (2.1) merely states that, given the present location of X,,, the past is irrelevant for
-
-predicting the location of X,,+1.
-
-Definition 2.1.2 A function p : S x S —> R is said to be a transition probability if:
-1. for each :1: E S, A —> p(a:, A) is a probability measure on (S, S).
-2. for each A E 8, :1: —> p(z, A) is a measurable function.
-
-We call X,, a Markov chain with transition probabilities 12,, if
-P(Xn+l E B l a(Xn)) =Pn(XnaB) (22)
-
-The spaces (S, S) that we encounter below are standard Borel spaces, so the existence of the
-transition probabilities follows from the existence of regular conditional probabilities on Borel spaces
-— a standard measure theory result (see e.g. [3] page 230).
-
-Suppose we are given an initial probability distribution p on (S, S) and a sequence p,, of
-
-transition probabilities. We can define a consistent set of finite dimensional distributions by
-
-P<X,- e B,-,0 s 1 s n) = / #(d1‘o) / Po($o.d-'51) . .. / Pn-l(97n—1ad$n) (2.3)
-B0 B1 B1:
-Furthermore, denote our probability space by
-(9,?) = (S{0,1,...},8{0,l,...})
-
-We call this sequence space and it is defined more explicitly by
-
-sf°»1=-~-} = {(wo,w1,...) :w.- e S}
-
-S{°*1""} = ¢7(w :40; E A,- E 3)
-
-% 07.txt
-The Markov chain that we will study on this space is simply X" (w) = w,., the coordinate maps.
-Then, by the Kolmogorov extension theorem, there exists a unique probability measure P,, on
-(9,?) so that the X" (w) have finite dimensional distributions (2.3).
-
-If instead of p, we begin with the initial distribution 6,, i.e. point mass at 2:, then we
-denote the probability measure by P3. With such measures defined for each 2:, we can in turn
-
-define distributions P,,, given any initial distribution p, by
-
-PM = / u<da.->P.(A>
-
-That the foregoing construction — which, recall, was derived merely from an initial distri-
-bution p and a sequence p,, of transition probabilities — satisfies Definition (2.2) of a Markov chain
-is not obvious, and a proof can be found in
-
-To state the converse of the foregoing, if X,, is a Markov chain with transition probabilities
-p,, and initial distribution p, then its finite dimensional distributions are given by (2.3). Proof of
-this is also found in
-
-Now that we have put the theory on a firm, if abstract, foundation, we can bring the
-discussion down to earth by making the forgoing a little more concrete. First, we specialize our
-study of Markov chains by assuming that our chain is temporally homogeneous, which means that
-the transition probabilities do not depend on time; i.e. p,,(w,,, B) E p(w,., B). (This is the stochastic
-analogue of a conservative system.) Next we assume that our state space S is finite, and suppose
-for all states 1', j E S that p(i, j) Z 0, and E, p(z', j) = 1 for all i. In this case, equation (2.2) takes
-
-a more intuitive form:
-P(Xn+l =j I X7: = 7:) =P(iaj)
-
-and our transition probabilities become
-% 08.txt
-If P is a matrix whose (i, j)th element is the transition probability p(i, 1') then P is a stochastic
-
-matriar, that is, a matrix with elements pgj satisfying
-
-Pij 20: zjpij =1: (i7j=1121"‘)d)
-
-We also refer to P as the transition probability matrix.
-
-Without loss of generality, we can further suppose our Markov chain is irreducible. This
-means that, for any states 2', 1‘, starting in state i the chain will make a transition to state 1' at
-some future time with positive probability. This state of affairs is often described by saying that
-all states communicate. We lose no generality with this assumption because any reducible Markov
-chain can be factored into irreducible classes of states which can each be studied separately.
-
-The final two conditions we place on the Markov chains considered below will cost us
-some generality. Nonetheless, there remain many examples of chains meeting these conditions,
-thus making the present study worthwhile. Furthermore, it may be the case that, with a little
-more work, we will be able to drop these conditions in future studies. The first condition is that
-the chain is aperiodic. If we let I, = {n 2 1 : p"(:c,a:) > 0}, we call a Markov chain aperiodic if,
-for any state 2:, the greatest common divisor of I, is 1. The second assumption is that our chain is
-
-reversible. This characterization is understood in terms of the following definition.
-Definition 2.1.3 A measure a is called reversible if it satisfies
-
-;4(w)p(I. y) = u(y)p(y. w)
-
-for all :c and y.
-
-We call a Markov chain reversible if its stationary distribution (defined in Section 2.1.2) is reversible.
-% 09.txt
-\subsection{A Convergence Theorem}
-
-In succeeding arguments, we use some results concerning the asymptotic behavior of
-
-Markov diains. These results require a few more definitions.
-
-Definition 2.1.4 A measure 1r is said to be a stationary measure if
-
-Z«(z)p<z.y> = we) (2.4)
-
-2
-Equation 2.4 says P,,(X1 = y) = 1r(y), and by induction that P,,(X,, = y) = 7r(y) for all n 2 1. If 1r
-is a probability measure, then we call 1r a stationary distribution. It represents an equilibrium for
-the chain in the sense that, if X0 has distribution 1r, then so does X,. for all 11..
-
-When the Markov chain is irreducible and aperiodic, the distribution of the state at time
-n converges pointwise to 1r as n —> oo, regardless of the initial state. It is convenient to state
-this convergence result in terms of the Markov chain's transition probability matrix P. Before
-doing so, we note that irreducibility of a Markov chain is equivalent to irreducibility (in the usual
-matrix theory sense) of its transition probability matrix. Furthermore, it turns out that a transition
-probability matrix of an aperiodic Markov chain falls into that class of matrices often called acyclic,
-but for simplicity we will call such stochastic matrices aperiodic. With this terminology, we can
-
-state the convergence theorem in terms of the transition probability matrix P.
-
-% Theorem 2.1.1
-\begin{theorem}
-\label{thm-2.1.1}
-Suppose P is irreducible, aperiodic, and has stationary distribution 7r. Then as n —> oo,
-p"(i,J') -> #0")
-\end{theorem}
-
-The symbol p"(i, j) is to be understood as the (ij) element of the nth power of P.
-A Markov chain whose transition probability matrix satisfies the hypotheses of
-Theorem~\ref{thm-2.1.1} is called ergodic. If we simulate an ergodic chain for
-sufficiently many steps, having 
-% 10.txt
-begun in any initial state, the final state is a sample point from a distribution which is close to 1r.
-
-To make this statement more precise requires that we define “close”.
-
-Definition 2.1.5 Let 1r be a probability measure on S. Then the total variation distance at time
-
-n with initial state I is given by
-AA") = l|P"(z, A) - 7r(-4)||Tv = 1333; I P"(I.A) - 7r(A) I
-
-In what follows, we will measure rate of convergence using the function 'r,(e), the first time after
-
-which the total variation distance is always less than 5. That is,
--r,,(e) = min{m : A,(n) 3 E for all n. 2 m}
-
-To consider the connection between convergence rates and eigenvalues, note that an ape-
-riodic stochastic matrix P has real eigenvalues 1 = /\o > A1 2 A2 3  Z /\¢_1 2 -1, where
-d = |S'] is the dimension of the state space. For an ergodic chain, the rate of convergence to the
-stationary distribution 7r is bounded by a function of the subdominant eigenvalue. By subdominant
-eigenvalue we mean that eigenvalue which is second-largest in absolute value, and we denote this
-Am“ E max{).1,|A¢_1|}. The function bounding the rate of convergence is given by the following
-
-theorem:
-
-% Theorem 2.1.2
-\begin{theorem}
-\label{thm-2.1.2}
-The quantity T1(€) satisfies
-1. 13(5) 5 (1 — /\m,_x)“(ln1r(a:)‘1 +1ne‘1)
-2. maxxes r,(e) 2 %/\,,,,x(1 — ).,,,,x)‘1ln(2e)’1
-\end{theorem}
-
-As this theorem shows, if we have an upper bound on the subdominant eigenvalue, then we have
-an upper bound on the function 11(5). In what follows, we will derive an approximation to the
-% 11.txt
-subdominant eigenvalue and supply error bounds. Together, an approximation and error bounds
-for Am“ provide enough information to make Theorem~\ref{thm-2.1.2} useful.
-
-\section{Functions on the State Space}
-
-Recall that X,.(w) = w,. 6 S denotes the state in which the Markov chain exists at time
-11. Suppose that 11> = {¢1, . . . , ¢,,} is a collection of p observables, or functions defined on the state
-space S. Furthermore, let these observables be real valued, so ¢.~ : S —> R. It is often useful, as we
-will see below, to assume that none of the observables is a constant function. Suppose now that
-the state space S is finite with at possible states. Then, since an observable is simply a map of the
-state space, we can think of each ¢,- as a vector of d real numbers — the d values that it takes on at
-the different states.
-
-Now assume the Markov chain is irreducible, and let 1r denote its stationary distribution.
-If we start the chain from its stationary distribution - i.e. suppose X0 has distribution 1r - then
-X,. is a stationary process. Furthermore, for each i, ¢,(X,.) is a stationary stochastic process with
-
-mean
-
-E.¢.- = Z) vr(x)¢.(z)
-
-165
-
-and autocovariance function
-
-Cir (¢i(Xn)v ¢i (Xn+a))
-
-E7!’  ‘ E1r¢i) (¢i(Xn+a) " E1r¢i)l
-
-Z Pn(Xn = z1X7I+S = y)  — E1r¢i)  _ E1r¢i)
-
-z,yES
-(2.5)
-
-which, by the definition of conditional probability, we can write as,
-
-2 P«(X.. = z)P.(X..+.. = y I X. = 1)(¢s($) — Em) (my) ~ Em)
-1.1165
-% 12.txt
-Z 7r($)P1y (MI) - Em.-) (¢:(1I) - E1r¢i)
-
-:c,y€S
-
-Here pgy denotes the element in row :1: and column y of P’, the sth power of the transition probability
-
-matrix. Similarly, we define the cross-covariance between the function ¢,- at time n and q5,- at time
-
-n+sas
-
-E1!’  _ E7r¢:') (¢i(Xn+a) " E1r¢i)l (2-6)
-
-C1r(¢i(Xn)a (#1 (Xn+a))
-
-Z) vr(x)p;,, (¢.-(w) — Em.-) (my) — Em)
-
-z,yES
-
-Now let (<I>) denote the matrix of mean vectors whose jth column is E,,¢,-1, where 1 = (1, . . . , 1)‘,
-and let H = diag(1r(w1), . . . , 1r(wd)), be the d X d diagonal matrix with stationary probabilities 1r(w)
-on the main diagonal, and zeros elsewhere. Finally, denoting by C(s) the p X p covariance matrix
-
-whose (i,j)th element is C, (qfi,-(X,.),¢_,< (X,.+,)), we have
-
-0(0) = E(‘1’(Xn) - (‘I>))(‘I’(Xn) - (‘PW
-= (‘P-(‘1’))‘1'1(<1’-(‘I>))
-0(8) = E(<I>(Xn) — (‘I>))(‘1’(Xn+a) - (‘1’))’
-
-(‘P - (‘1’))'1'IP‘(<1> — (<1>))
-
-Below, we will also find it useful to have at our disposal a new matrix that is similar to the
-transition probability matrix. We have in mind the matrix M E Ill/2PI'I‘1/2. As is easily verified,
-this allows us to write the covariance matrix as
-
-can = <<I>—<<I>>)°n%*M*n%<<I>—<<I>>>
-
-\II‘M’\II (2.7)
-
-where we have defined ‘II E 111/ 2(<I> —  Recall that our main motivation is that, for out-of-core
-problems, traditional eigenvalue algorithms are inadequate. With this fact, and the form (2.7) in
-% 13.txt
-mind, we will consider using the covariance of observables on the state space to implement the
-(otherwise useless) Rayleigh-Ritz procedure, which we describe below. This procedure requires
-that M be symmetric. As the next fact demonstrates, this required symmetry is the reason for our
-
-interest in the reversibility property of the Markov chain.
-
-Fact 2.2.1 The matrix M is symmetric if and only if the Markov chain is reversible (i.e. iﬂ the
-
-process satisfies the detailed balance condition).
-
-Proof:
-
-M is symmetric
-
-4:, (1-I1/2P1-I-1/2): = H1/2Pn—1/2
-<.—> r1-%*P°n%* = H1/2911-1/2
-¢> 1>'r1= = up
-~:=> (IIP)‘ = l'IP
-
-Element-wise, the final equality implies that 7|‘,'p,'j = 1rjpj.-. According to Definition 2.1.3, this states
-
-that p is a reversible measure.
-
-% 14.txt
-
-\chapter{Invariant and Approximate Invariant Subspaces}
-
-\section{Invariant Subspaces}
-Definition 3.1.1 A subspace S Q R" with the property that
-
-1:ES'=Ma:€S
-
-is said to be invariant for M.
-
-Recall that, having chosen observables <I> = ((151, . . . , qfip), we constructed the covariance
-matrix C(s) = \II‘M"\II. If the column space of \II, which we denote by ran(\II), is an invariant
-subspace for M, the definition implies Mifij E  That is, for each 111,- there exists a vector
-t of coeﬂicients such that M1/Jj = f=1t,-1/)5,-. This is true for all j and, putting each vector of
-coefficients into a matrix T, we see that M\II = \IIT. Conversely, M\II = IIIT implies that M1/Jj is a
-
-linear combination of columns of \II, so ran(\II) is invariant. We have thus proved the following
-% 15.txt
-Fact 3.1.1 The subspace ran(\II) is invariant for M if and only if there exists T E R”? such that
-
-M\II = \IIT.
-
-Consequently,
-
-Fact 3.1.2 )\(T) Q ).(M)
-
-Proof of 3.1.2:
-
-z\E)\(T) _=_ 3v€R"3Tv=).v
-<=> \IITv = A\IIv
-<=> M\II-u = /\\IIv
-< /\ E /\(M)
-
-The second ¢> follows from fact 3.1.1.
-Facts 3.1.1 and 3.1.2 are theoretically useful. To see why, write the equation of fact 3.1.1
-
-as follows:
-
-\IIT = M\II
-\II‘\IIT = \II‘M\II
-T = (\1:'~1r)-‘\1:'M\1:
-
-In this form, we recognize that T = C‘1(0)C(1). That is, using only the covariance of observables
-on the state space, we can generate a matrix T with the property A(T) (_Z ).(M). Recalling that
-M = H1/2Pl'I‘1/2, we see that M is similar to our original stochastic matrix P, and thus ).(M) =
-
-).(P).
-
-\section{Approximate Invariant Subspace}
-
-We qualiﬁed the foregoing by stating that the facts are only theoretically useful. This
-
-is because when choosing observables of the state space we aren't sure that they will satisfy the
-% 16.txt
-primary assumption underlying facts 3.1.1 and 3.1.2. Recall the assumption: ran(\II) is an invariant
-subspace for M, where ‘II = H1/2(<I> —  It may well be the case that there is a 1p; 6 ran(\II) such
-that M1/zj ¢ ran(\II), thereby violating the assumption. However, lacking an invariant subspace, we
-
-might hope that our observables at least provide an approximate invariant subspace.
-
-Deﬁnition 3.2.1 If the columns of ‘II E R“? are independent and the residual matrix E E
-
-M\II —— \IIT is small for some T 6 R7”, then ran(\II) deﬁnes an approximate invariant subspace.
-
-To see how an approximate invariant subspace can be useful in approximating eigenvalues of M,
-we extract a theorem from Golub and Van Loan 
-% Theorem 3.2.1
-\begin{theorem}
-\label{thm-3.2.1}  
-Suppose M 6 Rd” and T 6 RP"? are symmetric and let E E MQ — QT, where Q 6 Rd"? satisﬁes
-
-Q‘Q = I. Then there exist p1,... ,;z,, E A(T) and A1,. . . , A,, E A(M) such that
-ll‘k - Akl S \/§|lE"
-
-fork=1,...,p.
-\end{theorem}
-
-If the subspace ran(Q) is an approximate invariant subspace, then the deﬁnition implies that
-there is a choice T rendering the error IIEII2 small, and thus, the eigenvalues of T provide a good
-approximation to those of M.
-
-When considering the preceding ideas, it becomes apparent that their application will
-present new — but hopefully less prohibitive — problems. As these problems summarize the rt of
-the paper, now is a good time to examine them. First, the preceding theorem assumes a matrix Q
-whose columns form an orthonormal basis for the approximate invariant subspace. For our problem,
-derivation of such a Q is tricky, and we must prepare for this. Next, having an approximate invariant
-
-subspace at our disposal merely tells us that there exists some matrix T which makes the error
-% 17.txt
-"EH2 small. We must discover the form of such a T. Moreover, it is natural to seek that T which
-minimizes IIEII2 for a given approximate invariant subspace. Finally, in order to apply these ideas
-to real examples of our eigenvalue problem, we must ﬁnd a practical way to generate the most
-appropriate approximate invariant subspace. We now address each of these issues in the order
-raised.
-
-Recall the matrix I1 E III/2(<I> — (<I>)). The columns of this matrix, though independent
-(by choice of independent observables), are not necessarily orthonormal. However, consider the
-polar decomposition ‘II = QZ where Q‘Q = I and Z2 = \Il‘\II is symmetric positive semideﬁnite.1
-
-Note that Z = (\II‘\II)l/2 is nonsingular, so Q has the form
-Q = ~1:z-1 = \II(\II‘\II)“/2
-
-and it is clear that ran(Q) = ran(\II). Perhaps ra.n(Q) is a. useful approximation to the invariant
-subspace for M. If ran(Q) is not itself invariant, we have the error E = MQ — QT. It is easy to
-show, as we do in the following proof, that the T which minimizes HE“; is T = Q‘MQ. This yields
-
-the following,
-Theorem 3.2.2
-
-If \I1 = QZ is the polar decomposition of \II, then the matrix
-T = Q‘MQ = (\1:*\1:)-1/’\1:‘M\I:(\Ir‘\I:)-1/2
-minimizes IIEII2 = IIMQ - QTII2
-
-Proof:
-We prove the result by establishing the following
-
-(Hal: Given M 6 Rd” suppose Q 6 Rd"? satisﬁes Q'Q =1. Then,
-
-min ||MQ — QTII2 = H(1- QQ')MQll2
-
-TERI‘?
-
-‘Recall that the polar decomposition is derived from the SVD, \II = UZJV‘, by letting Q = UV‘ and Z = VEV‘.
-
-% 18.txt
-and T = Q’MQ is the minimizer.
-To verify the claim, observe the following application of the Pythagorean theorem:
-
-For any T E R.”"” we have
-
-IIMQ — QT||§ = IIMQ - QQ'MQ + QQ'MQ - QT|l§
-||(I - QQ’)MQ + Q(Q'MQ - T) H3
-
-||(1 - QQ‘)MQ||§ + |lQ(Q'MQ - T)||§ (3-1)
-
-2 l|(I — QQ')MQ||§
-
-Equality (3.1) holds since I — QQ' projects MQ onto the subspace orthogonal to ran(Q). Thus,
-the two terms in the expression on the right are orthogonal, and the Pythagorean theorem yields
-equality. The concluding inequality establishes that the minimizing T is that which annihilates the
-second term in (3.1); i.e. T = Q‘MQ.
-
-The second equality in Theorem 3.2.2 is a consequence of the polar decomposition, in
-
-which Q = \IIZ“ = \I:(~I:‘\I:)-1/2. Therefore,
-T = (x1z'x1:)-1/2x1:‘M~1:(\1:*~1z)—1/2
-
-is the minimizer.
-
-3.3 The Krylov Subspace
-
-Suppose the columns of a matrix Q E R“? provide an orthonormal basis for an approxi-
-
-mate invariant subspace. Then we have seen,
-1. T = Q'MQ minimizes ||E||2 = ||MQ — QT||2 and
-2. E|p.1,...,p,, E z\(T) and /\1,...,A, 6 »\(M) such that
-
-ll-‘I: - Ml S \/§|lE||2
-
-% 19.txt
-19
-fork=1,...,p.
-
-Given an approximate invariant subspace ran(Q) of dimension 1), these facts tell us what matrix we
-should use to approximate p elements of M’s spectrum. Now all we lack is a. description of ran(Q).
-That is, we have not speciﬁed which approximate invariant subspace would best suit our objective
-
-of approximating the subdominant eigenvalue A,,.a2:(M). 'I‘o do so requires the following deﬁnition:
-
-Deﬁnition 3.3.1 The Raleigh quotient of a symmetric matrix M and a nonzero vector 3 is
-
-p(2:, M) E (:z:‘Ma:)/(z‘a:).
-
-We will also denote this by p(z) when the context makes clear what matrix is involved.
-
-To ﬁnd the approximate invariant subspace most appropriate for our problem, we choose
-each dimension successively, providing justiﬁcation at each step. Thus, start with one observable
-42 on the state space, and let zpl E 111/ 2(¢ — E,,¢). That is, 1/:1 represents a centered (mean zero)
-observable whose ith coordinate is weighted by . First notice that the deﬁnition (which comes
-directly from the deﬁnition of ‘II following equation (2.7)) is such that 1/)1 is a constant vector if
-and only if :15 is oonstant on the state space, in which case 1/); E 0. This fact provides one reason
-observables that are constant on the state space are not interesting. Second, recall that the row
-sums of the matrix P are all one and, therefore, the eigenvector corresponding to the eigenvalue
-Ao(P) = 1 is the constant vector. By deﬁnition of M E 1'11/2Pl'I‘1/2, we see that the constant vector
-is also the eigenvector corresponding to A0  This will play an important role in what follows, as
-it allows us to focus primarily on the subdominant eigenvalue z\max(M) E max{}q (M), [A44 
-rather than on A0  (which we already know is 1).
-
-Now, notice that
-
-lP(¢1,M)l 5 max |P(x,M)l = »\max(M) (3-2)
-% 20.txt
-where the max is taken over all non-constant vectors. Since our interest centers on An-.3_x(M), we
-would like a. 1/11 which makes the left hand side of (3.2) large. This would be achieved if 1/)1 were to
-lie in the space spanned by, say, the first two eigenvectors of the Markov chain (sometimes referred
-to as the slowest modes of the process). However, this subspace spans only two dimensions of the
-entire d-dimensional space, and it is more likely that ¢1, at best, only comes close to lying in the
-subspace of interest. Now, given 1/11, a judicious choice for the second dimension l/)2, and hence
-W; E [1/)1 1/:2], would be that which makes max,,¢o |p(\II2a)| is as large as possible. To establish that
-
-this is indeed the pertinent objective, note the following:
-
-_ a‘\1:5Mx1:2a
-I;1;g<Ip(\I'2a,M)l = 13:3: ————a,W§q,2a
-= M
-mrggf,mIp(z. )1
-5 maxlp(z,M)l (3-3)
-= ).,,m(M)
-
-Again, the max on the right side of (3.3) is over non-constant vectors. In words, we wish to chose
-\II-2 = [1/:1 ¢2] so that there is a vector a 6 R2 making |p(\II2a,M)| close to )«max(M).
-
-Now, p(1/11) changes most rapidly in the direction of the gradient Vp(¢1).
-
-W0  (iiiiiiv---viiiléi)
-= ,,;,l (Mam — p<¢.)«m (3.4)
-
-So, to maximize the left hand side of (3.3), ‘I12 should be chosen so that the subspace ra.n(\II2)
-
-contains the gradient vector. That is, we must choose 1/22 so that
-
-VP(¢'1) E F3l1{¢1y¢2} E F-'%!1(‘I’2) (3-5)
-
-Clearly, equation (3.4) implies Vp(1/:1) E ran{z/11,M1,b1}. Therefore, if ran{1/:1,1/)2} = ra.n{¢1,M¢1},
-
-then (3.5) is satisﬁed.
-% 21.txt
-In general, having chosen \II,, E [1/)1 - -- 1/11,] so that
-
-ran{'pl9¢21' " 1¢k} = ran{¢1vM1[)la ' ‘ ' vMk—11/’k}
-
-we must chose 1/1;.“ so that, for any non-zero vector u 6 R",
-
-VP(‘I’I:a) 6 ra‘n{¢11 ‘21 - ° - 31/Jk-F1} 
-Now,
-vp<~I~.a) = gmm — P(‘I’kﬂ)‘I’k“)
-
-and therefore,
-
-Vp(‘II;,a.) E ran(M\II;,) U ra.n(\IIk) = ra.n{¢1, M1/)1, . . . , Mktﬁl}
-
-Thus, requirement (3.6) is met when ran{1/:1, 1/:2, . . . , 1/4+1} = ran{1/)1, M1/:1, . . . , Mk:/11}.
-
-In conclusion, the p-dimensional approximate invariant subspace which is most suitable
-
-for our objective is
-
-/C(Mn/mp) 2 ran{«/:1,M¢1,... ,M"“«/21}
-
-This is known as the K rylov subspace. Therefore, in answer to the problem posed at the outset
-of this section, if we take the columns of Q to be an orthonormal basis for IC(M,z/)1, p), then the
-
-eigenvalues of T = Q‘MQ should provide good estimates of p extremal eigenvalues of M.
-
-% 22.txt
-\chapter{Lanczos Procedures}
-
-The following will review the usual procedure for generating an orthonormal basis for
-lC(M, «/11, p), and conclude by showing that, for our problem, we can find the matrix Q‘MQ without
-actually carrying out this procedure. Note that this conclusion is essential since the procedure
-
-requires matrix vector multiplication - an operation we have assumed impossible for large enough :1.
-
-4.1 The General Lanczos Algorithm
-Let 21;“) E M"¢ and consider the d x :1 matrix
-«I 2 [«/»,M«/»,  ,M"“¢] 2 [«b‘°>,«/»“’,...,¢<‘-‘>1
-
-Notice that M\II = [1/:(1),... ,1/;(d'1),Md¢r] and, assuming for the moment that \I1 is non-singular,
-
-we can compute the vector h = —\II‘1M"1/J. Thus,
-
-M\II = \II[e2,. . . ,e,1, —h] E IIIH
-
-% 23.txt
-23
-
-Equivalently,
-0 0  0 -—h1
-1 0  0 —h2
-HE   3 ; =\II“M\II (4.1)
-0 "hl1—1
-1 —h,;
-
-H is a companion matrix which means that its characteristic polynomial is p(:r) = 2:‘ + 23:1 h,-:r"‘1.
-Since H is similar to M, ﬁnding the eigenvalues of M is equivalent to ﬁnding the roots of p(a:).
-However, this is of little practical use since ﬁnding h, constructing 12(1), and ﬁnding its roots is
-probably a harder problem than the one we started with. Instead, the value of decomposition (4.1)
-derives from its upper Hessenberg form. We exploit this property below.
-
-Let ‘II = QR be the QR decomposition of \II. Since ‘I! is assumed nonsingular,
-~II;‘M~r.: = (R"Q‘)M(QR) = H
-Therefore,
-Q‘MQ = RHR-1 2 T
-
-Since R and R“1 are both upper triangular and H is upper Hessenberg, T = R.HR'1 is also upper
-Hessenberg. Furthermore, since M is symmetric, it is clear that T‘ = Q‘MQ = T. That is, T is
-
-an upper Hessenberg matrix that is symmetric. Therefore, T is tridiagonal and we can write it as
-
-follows:
-
-T =    (4.2)
-
-ﬁd—1
-
-0  174-1 ad
-
-% 24.txt
-24
-
-Equating columns j on both sides of MQ = QT yields
-M113‘ = B,-—1qj—1 + 0:91 + ﬁjqj+1 (43)
-
-Since the columns of Q are orthonormal, multiplying both sides of (4.3) by q_,- and q,-+1 yields
-oz, = q§Mq,- and ﬂj = q;-+1Mq,-.
-The foregoing justifies what is called the Lanczos algorithm, performed as follows:
-
-The Lanczos algorithm for (partial) reduction to symmetric tridiaganal form
-
-91 = ¢/||¢||2, [30 = 0, 40 = 0
-
-for j=1top
-
-z=Mq_,-
-
-5:‘ = ||Z||2
-if 5, = 0, quit
-¢Ij+1 = 2/51
-
-end for
-
-The qj computed by the Lanczos algorithm are often called the Lanczos vectors. If the
-loop in the algorithm is terminated because ﬁg = 0, this indicates that an exact invariant subspace
-has been computed, and is given by ran{q1 - - - qk}. Otherwise, we usually halt the algorithm after
-p steps, in which case the algorithm converg to approximations of at most 12 eigenvalues.‘
-
-Note that equation (4.3) results from the equation MQ = QT. The latter equation holds
-since, starting with d vectors in our subspace, Q is a d Xd matrix whose columns form an orthonormal
-
-basis for all of R‘, which is clearly an invariant “sub” space. The Lanczos algorithm above, however,
-
-‘A detailed discussion of such convergence (and misconvergence) is given in
-
-% 25.txt
-25
-
-proceeds for only p steps, producing a d x p matrix Q, whose columns form an orthonormal basis
-for the approximate invariant subspace IC(M, ¢, p), and a p X p matrix T, = Q'MQ. In that case,
-IIMQ = QTp|[g =  is non-zero and gives the error bound described in Theorem 3.2.1. Now,
-
-writing the full d x :1 matrix T of (4.2) as
-
-01 51
-/71
-
-(4.4)
-
-allows us to describe the error bound ||E||-2 in terms of the submatrix Tm, and, hence, in terms of
-
-ﬁzz-
-Theorem 4.1.1
-
-HT, and T,“ are the matrices appearing in (4.4), and if the p columns of Q are computed by the
-
-Lanczos algorithm, then there exist pl, . . . ,;t,, E /\(T,) and A1, . . . , A, E A(M) such that
-
-ill» - /\I=i S iiTP'4”2 = 512
-
-fork=1,...,p.
-
-For a proof, see page 365 of [1].
-
-% 26.txt
-26
-
-4.2 A Lanczos Procedure for Markov Chains
-
-Notice that the Lanczos algorithm requires the computation of Mqj. Throughout this
-paper we have assumed that M 6 Rd“, and that d is so large as to make the matrix vector
-multiplication Mqj impossible. Now, the Lanczos algorithm constructs an orthonormal basis for
-the Krylov subspace. For our purposes, this construction is not essential. What is essential is that
-we ﬁnd a way to generate the Lanczos coejﬁcients oz,-, ﬁj, (j = 1, . . . , p), and whence the matrix
-T, without performing the operation Mqj required by the Lanczos algorithm. We now address this
-problem.
-
-Begin with the centered and weighted observable 1/1 = 111/ 2 (qS—E,,¢), and let q1 = 11;/||1/JH2.
-To simplify notation, let ¢(n) E ¢(X,.). The form of the first Lanczos coeﬂicient a1 is straight
-
-forward:
-
-01 = qiMq1
-
-I/1‘M1/I
-¢’¢
-
-and now the special properties of the Markov chain setting play an important role. For, regard the
-
-following
-
-¢‘M«/2 = (¢ — E.¢)‘II%‘MII%(¢ - Em)
-
-Cw (d>(n), ¢(" + 1)) (4-5)
-
-and
-
-W1!’ = (<15 - E«¢)'H(¢ - E«¢)
-
-Varn (¢(")) (4-6)
-
-Recall that the deﬁnition of M implies H%‘MI'I% = HP, and this justiﬁes equation (4.5). Putting
-
-% 27.txt
-27
-
-these equations together, we have a. nice form for :11:
-
-_ c, (run), an + 1))
-"‘ ' Ta:,(¢(n))
-
-The next Lanczos coeﬂicient, B1, is only slightly more complicated.
-
-[31 = ||M91-0t1tI1||2
-
-[(Mq1 — 0l1¢11)i(M¢]1 - 0101)] 1/2
-
-[t1iM2¢I1 — ail 1/2
-
-(4.7)
-
-Again recalling the deﬁnition M E 111/ 2P1'I‘1/2, a. simple manipulation veriﬁes
-
-1'li‘M’l'[% = HP’, whereby it follows that
-
-(«s — E,,¢)‘HP”(¢ — Em)
-
-"‘M2"‘ (¢ — E,¢)*n<¢ — Em)
-
-C: (¢("), ¢(n + 2))
-V3-Ur
-
-Inserting the last expression into equation (4.7) gives
-
-_ c,(¢(n).¢<n+2)) c,(¢<n>,¢(n+1)> ’ "’
-"“ jvar,<¢<n)> "(jvar,<¢<n)) ‘)
-
-Notice that we can estimate these expressions of oz; and ﬂl by running simulations of the Markov
-chain. To do so, we run the Markov chain starting from a random initial state, compute the value of
-the observable at each step of the chain and, from the data, compute estimates of the covariances
-and variance of the observable.
-
-Before generalizing this for the jth Lanczos coefficient, we consider one more coeﬂicient
-in detail. Recall that the second basis element q; in the Lanczos algorithm is given by 1;; =
-
-(Mql — oz1q1)/ﬁl. From this we can derive an expression for (12 which, though not as neat as those
-
-% 28.txt
-28
-
-for oz; and [31, is nonetheless tractable.
-
-I12 ‘IiMq2
-
-1
-[7(M91 — 0t1¢I1)tM(M¢I1 - 11191)
-1
-
-1
-= f,§(qiM3q1 - 2a1qlM2<11 + a’9iM¢11)
-1
-1
-= —2(<1iM3q1 - 2015i ‘ all (4-8)
-
-1.
-The last equality follows from equation (4.7), which can be written as q{M2q1 = off’ + (if. The
-preceding paragraphs deﬁne all but one of the quantities in the expression (4.8). The one remaining
-is q§M3q1, which can also be expressed in terms of covariances of <15, and the analysis is almost
-
-identical to that for q{M2q1:
-
-(¢ — E,¢)'IIP3(¢ — Em)
-
-"lM3"‘ (¢ — E,¢s)‘H(¢ - Em»)
-
-Cu (¢(")a ¢(n + 3))
-Varn(¢(n))
-
-To generalize the foregoing for higher order Lanczos coeﬂicients requires a recurrence
-relation between (aj+1, ﬂj) and (aj, ﬂ,~_1). As we will see, such a relation involves higher order
-Lanczos vectors q,~ — vectors we do not wish to compute. Recall that the only “vector” we can
-handle is our initial observable. Therefore, we must also establish a recurrence among the q,-’s, thus
-making them available inductively from our observable vector.
-
-To begin with the ﬂfs, recall that ﬁj E ||Mqj — ajq_.,- — ,H_,~_1q,-_;|[2. Whence,
-
-Bf (M411 — °‘j‘Ij - ﬁj-1qj—1)’(MlIj - Otjqj - 51-19141)
-
-q;-M291‘ - 0? - B,z_1 (4-9)
-
-Some simple algebra and the deﬁnitions aj E q§Mqj and ﬁJ~_1 E q;-Mq,--1 prove equality (4.9). The
-% 29.txt
-recurrence for 0114.1 is equally straight forward, though the ﬁnal form is not as neat:
-
-2
-0tj+1 ¢Ij+1M'1j+1
-
-1
-§(M¢15 - 11111;‘ - ﬂj—1<Ij—1)'M(Mqj - Otjqj - /3j—1qj—1)
-J
-
-1
-§(q;-M-“*4; — 2a,~q;M*q,- — 2x3;+’_1q;--1M*q,- + «I? + 2a,-5,51 + am (4.10)
-J
-From (4.9), q;~M2q,~ = oz? + ﬂf + ﬂ}_1 and, upon plugging this into (4.10), we arrive at
-1
-a,~+1 = E(q§»M3q_,,- — (1? — 2ozJ~ﬁJ2 — 2ﬂj_1q;»_1M2qj + ﬁ2_1) (4.11)
-J
-
-The expressions above involve the higher order Lanczos vectors qj. In the present context,
-these vectors will not be readily available. However, another recursion — this time relating forms
-involving q,-+1 to forms involving qj and qj__1 — will make these expressions accessible by induction
-on our initial observable. The general forms of interest are q;+1M"qj+1 and qj-+1M"qJ~. Again, we
-
-begin with the simpler of the two:
-
-¢I;'+1Mk9j = ij(Mqj - Otjqj - ﬂj—1t1j—1)'M"¢1j
-= i(<1§M"“¢Ij - C‘j‘I_;'Mk‘Ij - ﬂj—1q§M"qj-1)
-The second is,
-‘I_;'+1Mkqj+1 = ﬂi?(M<Ij - Otjqj - .3j—1<Ij—1)‘Mk(Mqj — Orjqj - ﬂj—1¢1j—1)
-
-1
-ﬁ(‘1§M'°"<1:‘ + °‘:2"I§M'°<J:' + ﬁ?—z<Ii—1M"<I:‘—1
-J
-—2a,-q;-M*+‘q,~ + 2am,--1q;-M*q,--1 — zza,--1q;-M*+‘q]--1) (4.12)
-
-The point of all this is that each pair of coeﬂicients (oz,-+1, /3,-) can be expressed as a
-function of the covariance between ¢(n) and ¢(n+ m), for m = 0, . . . , 21' + 1. Since the expressions
-are based on four recurrence relations, they become quite complicated as 1' gets large. However,
-deﬁning these recursions in a. computer program is trivial, and we use this fact in Section 5.4 when
-
-applying the foregoing to a particular problem.
-
-% 30.txt
-30
-
 Chapter 5
@@
 40
-
-Chapter 6
-
-Conclusion
-
-We conclude by noting a few aspects of the special cases to which the foregoing applies.
-We are interested in the subdominant eigenvalue of a large transition probability matrix. The
-
-foregoing can be used to approximate these eigenvalues when the following conditions are satisﬁed:
-o The Markov chain is reversible (detailed balance condition).
-
-0 We can easily simulate realizations of the chain as well as functions (observables) of these
-
-realizations.
-
-When these conditions hold, we can side step the traditional Lanczos algorithm, and get at the
-values of the Lanczos coeﬂicients via the variational properties of observables on the state space.
-There are many ideas in this paper that should be expanded upon in future studies. Per-
-haps most obvious is the desire to ﬁnd analogous techniques for non-reversible (i.e. non-symmetric)
-Markov chains. This is a considerable problem since the Lanczos methods described above are no
-longer applicable. Perhaps the Amoldi algorithm ([1], algorithm 6.9), or the non-symmetric Lanczos
-
-algorithm can also be usefully specialized for stochastic matrices.
-
-% 41.txt
-41
-
-When applying the Lanczos algorithm, we usually choose a single starting vector (in the
-present context, an observable), perform the algorithm for a number of steps, and gather results.
-Then we do the same with a new starting vector and compare the results to those obtained with
-the ﬁrst vector. Proceeding in this way for a number of starting vectors, we can provide statistical
-evidence that we have, indeed, located the eigenvalues of interest. However, when there are two
-observables 451, :15; of particular interest, perhaps we would beneﬁt by considering both observables
-simultaneously. Future studies might develop an algorithm which incorporates both observables,
-
-again via ¢.- = H1/2(¢.' - E,.¢.-). In this case, a Lanczos algorithm would be based on the space
-
-Ic*(M,¢1,¢»2.p) 2 ran{«/»mz»2,M«/»1,M«/:2, . . .,MP—‘¢1.M"—‘«/22}
-
-The Lanczos coeﬁicients would then involve not only the autocovariances C, (¢,<(n),¢,-(n + m)),
-
-but also the cross-covariances C, (¢.~(n), ¢j(n + m)) (deﬁned in (2.6)).
-
-% 42.txt
-Appendix: computer programs
-
-Z Hatlab code ising.m
-
-Z
-Z Written by William J. De Meo on 12/15/97
-
-2 last modified 1/10/98
-
-Z Inputs:
-
-2 d = number of nodes of the ising lattice (e.g. d=100)
-Z n = number of iterations (e.g. n=1000)
-
-Z beta = Annealing schedule (e.g. beta = 2,
-
-1 when beta -> 0 will always go to nee state
-Z vhen beta -> infty will never go to state of higher energy)
-Z
-
-X = zeros(d,1);
-Energy = zeros(n,1);
-
-Ave=0;
-
-X Initialize using hot start
-for i=1:d,
-
-if rand < .5
-
-x(i) = -1;
-
-else
-
-X(i) = 1;
-
-end
-end
-
-Z Initialize at state of high energy (not used)
-
-if 0
-X(1:2:d-1)=1;
-X(2:2:d)=—1;
-and
-
-Z Initialize energy
-
-H=0;
-
-for i=1:(d-1)
-
-H = H - X(i)*X(i+1);
-end
-
-2 Initialize observable
-phi = zeros(n+1,1);
-
-for i=1:d
-if X(i) == 1
-phi(1)= phi(1)+1;
-end
-
-end
-
-42
-
-% 43.txt
-43
-
-for j=1:n
-
-site=0;
-while site==
-
-site = round(rand#d);
-end
-Z Compute new energy (with new X(site) = —x(site)):
-if site ==
-
-Hnew = H + 2:x(1)~x(2);
-elseif site == d
-
-Hnew = H + 2*X(d)tX(d-1);
-else
-
-Hnew = H + 2*X(site)*(X(site-1) + X(site+1));
-end
-
-Z Change to new state in two ways:
-X with probability 1/2, flip the switch
-if rand < .5
-if (Hnew <= H) I (rand < exp(-beta*(Hnew-H)))
-x(site) = -x(site);
-H = Hnew;
-Z Compute new value of observable
-if X(site) -= 1
-phi(j+1) = phi(j) + 1;
-
-else
-phi(j+1) = phi(j) — 1;
-end
-else phi(j+1)=phi(j);
-end
-else phi(j+1)=phi(j);
-end
-Energy(j) = H;
-end
-p1ot(Energy)
-
-% A second observable
-phi2 = zeros(n+1,1);
-for i=1:n+1
-
-phi2(i) = phi(i)*phi(i);
-end
-
-Z Write first observable to phi.dat:
-
-fid = fopen(’/quean.a/masters/deme5025/research/nyuthesis/programs/c/phi.dat’,’w’);
-fprintf(fid,’%f\n’,phi);
-fc1ose(fid);
-
-Z Write second observable to phi2.dat:
-fid = fopen(’/quean.a/masters/deme5025/research/nyuthesis/programs/c/phi2.dat’,’w’);
-
-% 44.txt
-fprintf(fid,’%f\n’,phi2);
-fc1ose(fid);
-
-1 Compute a few Lanczos coefficients by new method
-START=1001;
-
-V1=zeros(2);
-
-C1=zeros(2);
-
-C2=zeros(2);
-
-C3=zeros(2);
-
-V1 = cov(phi(START:n+1),phi(START:n+1));
-
-V1 = V1(1,1)5
-
-C1 = cov(phi(START:n),phi(START+1:n+1));
-C1 = C1(1,2);
-
-C2 = cov(phi(START:n-1),phi(START+2:n+1));
-C2 = C2(1,2);
-
-C3 = cov(phi(START:n-2),phi(START+3:n+1));
-
-C3 = C3(1,2);
-
-a1pha1 = C1/V1;
-
-betai = sqrt(C2/V1 ' (C1/V1)‘2);
-
-a1pha2 = (1/(beta1’2))*(C3/V1 - 2*a1pha1*(C2/V1) + a1pha1‘3);
-
-T1 = Ealphal betal; betal a1pha2]
-
-v1 = cov(phi2(START:n+1),phi2(START:n+1));
-v1 = v1(1,1);
-
-c1 = cov(phi2(START:n),phi2(START+1:n+1));
-c1 = c1(1,2);
-
-c2 = cov(phi2(START:n—1),phi2(START+2:n+1));
-c2 = c2(1,2);
-
-c3 = cov(phi2(START:n-2),phi2(START+3:n+1));
-c3 = c3(1,2);
-
-alphal = C1/V1;
-betal = sqrt(C2/V1 - (C1/V1)‘2);
-a1pha2 = (1/(beta1‘2))#(C3/V1 - 2ta1pha1*(C2/V1) + a1pha1”3);
-
-T2 = [a1pha1 betai; betal a1pha2]
-
-fid = fopen(’Tmats.dat’,’v’);
-fprintf(fid,’%f\n',T1);
-fprintf(fid,’%f\n’,T2);
-fc1ose(fid);
-
-ZZZ end ising.m
-
-44
-% 45.txt
-\% Matlab code tpm.m for generating transition probability matrix
-Z and computing its eigenvalues directly
-
-1
-
-Z Written by Hilliam J. De Heo on 1/10/98
-
-1
-
-Z Inputs:
-
-Z d = number of nodes of the ising lattice (e.g. d=10)
-
-Z beta = Annealing schedule (e.g. beta - 2,
-
-1 when beta -> 0 will always go to new state
-1 when beta -> infty will never go to state of higher energy)
-2
-
-disp(’bui1ding proposition matrix...’)
-
-states = 2‘d;
-
-A = 0;
-
-for i=0:(d-1)
-
-E=eye(2”i);
-
-A = [A E;E A];
-
-end
-
-B = .5#eye(states);
-
-A = B + .5-(1/d)*A;
-
-2 our modified Glauber dynamics requires the B and the .5¥(1/d)
-
-disp(’...done’)
-
-1 display pattern of nonzero entries
-1 spy(A)
-
-Z check that all row sums are 1
-check=O;
-for i=1:states
-
-check = check + (not(sum(A(i,1:states))<.99));
-end
-1 if not all rows sum to 1, print check = (# of rows with sum=1)
-if not(check==states)
-
-check
-
-error(’Row sums are not all 1’)
-end
-
-Z construct matrix of states
-E = zeros(d,states);
-f1ip=-1;
-for i=1:d
-for j=1:2‘(i-1):states
-flip = -1*f1ip;
-for k=0:2“(i-1)
-
-45
-
-% 46.txt
-E(i,j+k) = flip;
-and
-end
-end
-E=E’;
-
-Z display first 64 states
-Z E(1:64,:)
-
-Z compute energy of each state
-H = zeros(states,1);
-for i=1:states
-
-for j=1:d-1
-H(i) = H(i) - E(i,j)tE(i,j+1);
-end
-end
-
-disp(’building Hetropolis transition matrix...')
-
-for i=1:states
-for j=i+1:states
-if not(A(i,j)==0)
-if(H(j)>H(i))
-alt = A(i,j)t(1-exp(-beta*(H(j)-H(i))));
-A(i,i) I A(i,i)+a1t;
-A(i,j) = A(i,j)-alt;
-elseif(H(j)<H(i))
-alt = A(j,i)t(1-exp(—beta#(H(i)-H(j))));
-A(j,j) = A(j,j)+a1t:
-A(j,i) = A(j,i)-alt;
-end
-end
-end
-end
-
-disp(’...done’)
-
-check=0;
-for i=1:states
-
-check = check + (not(sum(A(i,1:states))<.99));
-end
-
-Z if not all rows sum to 1, print check = (# of rows with aum=1)
-
-if not(check==states)
-
-check
-
-error(’Row sums are not all 1’)
-end
-
-46
-
-% 47.txt
-47
-
-disp(’computing eigenvalues of tpm...’)
-cput = cputime;
-
-evals = eig(A);
-
-ecput = tputime - cput;
-
-disp(’the CPU time (in secs) for computing eigenvalues of tpm: ’)
-
-ecput
-ZZZ end tpm.m
-
-/¥#*##¥###*##*#l####¥#‘###$ﬂ###ﬁi##*U3*#**#*##**##¥***ll¥#*##¥ii
-
-* 1anczos.c main program for computing Lanczos coefficients *
-3 #
-* Created by William J. De Heo #
-t on 1/7/98 *
-# Last modified 1/10/98 #
-
-*¥#######**3It¥*#¥¥¥#**##**¥##¥#####**####*#*###*##**###¥##*##*/
-
-#inc1ude <stdlib.h>
-#include <stdio.h>
-#include <math.h>
-#inc1ude "prototypes.h"
-
-/# Machine constants on dino (Sun Ultra 1 at Courant) t/
-#define HACHEPS 1.15828708535533604981e-16
-
-#define SQRTEPS 1.07623746699106147048e-08
-
-ﬂdefine HAX_NAH 100
-
-void read_name(char *);
-
-/* Prototypes from the file functions.c ¥/
-
-double a1pha(int j);
-
-double betasq(int j);
-
-double form1(int j, int k);
-
-double form2(int j, int k);
-
-double moment(doub1e ‘data, long n, double tave, double tvar, double tcov, long k);
-
-/# external variables to be used by functions */
-double *phi,tcov, tvar, tave;
-
-% 48.txt
-main()
-
-{
-
-char tfilename;
-
-FILE *ofp;
-
-double temp=O, *T;
-
-long i, j, nrou, nlanc, START;
-int f1ag=O;
-
-filename = cma11oc(HAx_NAHE);
-
-printf("\nName of file containing observed values: “);
-
-read_name(fi1ename);
-
-printf(“\nTota1 number of observations in file (iterations):
-
-scanf(“Zu”,&nrov);
-
-printf(“\nNumber of leading observations to discard: “);
-
-scanf("Zu",&START);
-
-printf("\nNumber of Lanczos coefficients desired: “);
-
-scanf("%u",&n1anc);
-
-vhi1e(2*n1anc > nrow—START)
-
-{
-printf("\nNot enough data for that many coefficients.\n“);
-printf("\nEnter a smaller number of Lanczos coefficients: “);
-scanf("Xu“,&n1anc);
-
-}
-
-II);
-
-phi = dma11oc(nrow);
-cov=dma11oc(2#n1anc);
-var=dma11oc(1);
-ave=dma11oc(1);
-
-T = dma11oc(n1anc#n1anc); for(i=0;i<n1anc;i++) T[i]=(doub1e)0;
-
-/* observable phi is stored contiguously column-wise (e.g. by HATLAB) t/
-mat1abread(phi, nrow, 1,fi1ename);
-
-/* send the observable, offset by START */
-moment(phi+START,nrou-START,ave, var, cov,2*n1anc-1);
-
-/* First column of T */
-
-T[0] = a1pha(1);
-if((temp=betasq(1))>HACHEPS*10)
-{
-
-T[1]=sqrt(temp);
-
-/* General column of T */
-
-J'=2:
-
-for(i=1;i<n1anc-1, f1ag==O;)
-{
-
-T[itn1anc+i-1]=sqrt(betasq(i));
-
-48
-
-main()
-
-{
-
-char tfilename;
-
-FILE tofp;
-
-double temp=0, #T;
-
-long i, j, nrow, nlanc, START;
-int f1ag=0;
-
-filename = cma11oc(HAx_NAH);
-
-princf("\nName of file containing observed values: “);
-read_name(fi1ename);
-
-printf("\nTota1 number of observations in file (iterations): “);
-scanf("2u“,&nrow);
-
-printf("\nNumber of leading observations to discard: “);
-scanf("Zu",&START);
-
-printf("\nNumber of Lanczos coefficients desired: “);
-scanf("%u“,&n1anc);
-
-whi1e(2*n1anc > nrou-START)
-
-{
-printf("\nNot enough data for that many coefficients.\n");
-
-printf("\nEnter a smaller number of Lanczos coefficients: “);
-scanf("Zu",&n1anc);
-}
-
-phi = dma11oc(urow);
-cov=dma1loc(2*n1anc);
-var=dma11oc(1);
-ave=dma11oc(1);
-
-T = dma11oc(n1anc*n1anc); for(i=0;i<n1anc;i++) T[i]=(doub1e)O;
-
-/* observable phi is stored contiguously column-wise (e.g. by HATLAB) */
-mat1abread(phi, nrov, 1,fi1ename);
-
-/* send the observable, offset by START #/
-moment(phi+START,nrov-START,ave, var, cov,2*n1anc-1);
-
-/# First column of T */
-
-T[0] = a1pha(1);
-if((temp=betasq(1))>HACHEPS#10)
-{
-
-T[1]=sqrt(temp);
-
-/* General column of T */
-
-J‘=2:
-
-for(i-1;i<n1anc-1, f1ag==0;)
-{
-
-T[i*nlanc+i-1]=sqrt(betasq(i));
-
-48
-
-% 49.txt
-T[i*n1anc+i] = a1pha(i+1);
-if((temp = betasq(i+1))>HACmEPS#10)
-{
-
-T[i#n1anc+i+1] = sqrt(temp);
-
-i++;
-}
-else
-
-f1ag=1;
-
-}
-
-/t Last column */
-if(f1ag!=1 it ((temp= betasq(n1anc-1))>HACHEPSt10))
-
-{
-T[n1anc#n1anc-2] = sqrt(temp);
-T[n1anctn1anc-1] = a1pha(n1anc);
-printf("\nbeta(1) = %1f\nbeta(Zd) = Zlf (last beta)",
-T[1],(nlanc-1),T[nlanc*n1anc-2]);
-}
-else
-{
-printf("\nApproximate invariant space reached at step %d.",i);
-printf("\nbeta(1) = Z1f\nbeta(Zd) = Zlf (last accurate beta)“.
-T[1],i,T[itn1anc+i—1]);
-printf("\nbeta(Zd)“2: Zlf (first spurious resu1t)“,i+1,temp);
-}
-
-printf("\nThe matrix T is: \n");
-matprint(T,n1anc,n1anc);
-
-}
-
-else
-
-{
-
-printf("\nmain(): Approximate invariant space reached at first step.“);
-
-49
-
-printf("\na1pha(1) - Z1f\nbeta(1)”2: Zlf (first spurious result)“, T[O], temp);
-
-}
-ofp = fopen("Tmat.m","v“);
-check(ofp);
-mat1aburite(T,n1anc,n1anc,ofp);
-fc1ose(ofp);
-}
-void read_name(char tname)
-{
-int c, i = 0;
-while ((c = getchar()) != EOF it c != ’ ’ it c !- ’\n’)
-name[i++] = c;
-name[i] = ’\0’;
-}
-
-/#t* end 1anczos.c *#*/
-
-% 50.txt
-50
-
-/llnkttttttttinlntint*1!!!##1##!ttitilltit###t¥#tt¥ttItt####t#ttII=##mlltttt
-
-* functions.c functions required by 1anczos.c t
-# t
-* Created by Hilliam J. De Meo #
-* on 1/7/98 *
-t Last modified 1/10/98 :
-
-¥¥##¥$¥##¥¥#$¥¥¥###¥¥$*I#****#**####*######¥#¥#¥#$#¥¥#l¥**#****/
-
-#inc1ude <math.h>
-
-#define START 1000
-
-#define ITER 10000
-
-/t Machine constants on dino (Sun Ultra 1 at Courant) */
-
-#define HACHEPS 1.15828708535533604981e-16
-
-#define SQRTEPS 1.07623746699106147048e-O8
-
-double a1pha(int j);
-
-double betasq(int j);
-
-double form1(int j, int k);
-
-double form2(int j, int k);
-
-double moment(doub1e tdata, long n, double tave, double tvar, double tcov, int k);
-
-/t external variables to be used by functions */
-double *phi,tcov, *var, tave;
-
-double alpha(int j)
-{
-
-/* alpha is never called with j < 1 */
-if(j==1)
-return
-form1(1,1);
-else if(j>1)
-return
-((doub1e)1/betasq(j-1))*
-(form1(j-1,3) — pou(a1pha(j-1),3)
-- 2 * (alpha(j-1)#betasq(j-1) + sqrt(betasq(j-2))#form2(j-1,2))
-+ betasq(j-2));
-}
-
-double betasq(int j)
-
-{
-if(j==0)
-return (doub1e)O;
-else if(j>0)
-return (form1(j,2) - pow(a1pha(j),2) - betasq(j-1));
-}
-
-double form1(int j, int k)
-{
-
-% 51.txt
-51
-
-double form14,alphal,alpha1sq,form12,form13;
-if(j==0)
-
-return (doub1e)0;
-else if(j==1)
-{
-
-/# printf("\nvar = %1f, cov(%d) = %1f \n“,*var,k,cov[k]);t/
-
-return (cov[k])/(*var); /* the only real value */
-}
-else if(j>1)
-{
-
-return ((double)1/betasq(j-1))
-
-* ( form1(j-1,k+2)
-+ pow(a1pha(j-1),2)*form1(j-1,k)
-+ 2t(a1pha(j-1) t sqrt(betasq(j-2)) * form2(j-1,k)
-- a1pha(j-1)*form1(j-1,k+1)
-- sqrt(betasq(j—2))#form2(j-1,k+1) )
-+ betasq(j-2)*form1(j-2,k)):
-}
-}
-
-double form2(int j, int k)
-{
-/* form2 is never called with j < 1 */
-if(j==1)
-return (doub1e)0;
-else if(j>1)
-return
-(pov(betasq(j-1),—.5)) 3
-(form1(j-1,k+1) - a1pha(j-1)*form1(j-1,k)
-- sqrt(betasq(j-2)) # form2(j—1,k));
-}
-
-/* moment() function for computing var and cov(k)
-arguments:
-data = a nxi array of doubles
-n = length of data[]
-ave =(on exit)= the average of data[]
-var =(on exit)= the variance of data[]
-cov =(on exit)= the covariance of data[i] and data[i+j] for i-1,...,k
-k = the max lag for cov above
-*/
-double moment(double tdata, long n, double tave, double tvar, double tcov, int k)
-{
-/* Centered about data[0] algorithm: #/
-
-long i,j;
-double avel, ave2;
-
-% 52.txt
-52
-
-*ave=O;tvar=0; ave1=ave2=O;
-for(j=0;j<=k;j++)
-cov[j]=0;
-
-for(i=1;i<n;i++)
-{
-rave += (data[i] - data[0]);
-tvar += (data[i] - data[0])#(data[i] - data[0]);
-}
-tvar /= (doub1e)(n-1);
-Ivar -= (((*ave)/(doub1e)n) * ‘(*ave)/(doub1e)(n-1)));
-/# *ave = ((*ave)/(doub1e)n) + data[0]; <- that’s the true average (not needed)*/
-
-for(j=0;j<=k;j++)
-{
-
-for(i=0;i<(n-j);i++)
-
-{
-
-avel += (data[i] - data[0]);
-ave2 += (data[i+j] - data[0]);
-cov[j] += (data[i] - data[0])*(data[i+j] — data[0]);
-
-}
-
-avel/=(doub1e)(n-j); ave2/=(doub1e)(n-j);
-
-cov[j] = ((cov[j] - (double)(n-j)tave1¢ave2)/(double)(n-j-1));
-}
-
-}
-
-/trt end funccions.c ###/

### 2013-10-18 tex/Conclusion.tex
initial commit of tex source for nyu thesis style setup

--- 2013-10-18 tex/MSThesis.tex
@@
-%\documentclass[11pt]{amsart}
-\documentclass{au}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% For completion by editorial office:
- \presentedby{\dots}
- \received{\dots}{\dots}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\usepackage[colorlinks=true,urlcolor=black,linkcolor=black,citecolor=black]{hyperref}
-
-\usepackage{amsmath,amssymb,url,mathrsfs}
-\usepackage{enumerate}
-\usepackage{tikz}
-\usetikzlibrary{calc}
-\usepackage{scalefnt}
-\numberwithin{equation}{section}
-
-\theoremstyle{plain}
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{proposition}[theorem]{Proposition}
-\newtheorem{corollary}[theorem]{Corollary}
-
-\theoremstyle{definition}
-\newtheorem{definition}[theorem]{Definition}
-\newtheorem{notation}[theorem]{Notation}
-\newtheorem{remark}[theorem]{Remark}
-\newtheorem{example}[theorem]{Example}
-\newtheorem*{remarks}{Remarks}
-
-\newcommand{\todo}[1]{~\\[5pt]{\bf TODO(wjd):} #1\\[5pt]}
-%\newcommand{\suchthat}{\ensuremath{:}}    % (old notation)
-\newcommand{\suchthat}{\ensuremath{\mid}}  % (new notation)
-%\newcommand{\defeq}{\ensuremath{:=}}      % (old notation)
-\newcommand{\defeq}{\ensuremath{=}}        % (new notation)
-\newcommand{\betai}{\ensuremath{\beta^i}}  % \beta^{\bB_i}   % (old notation)
-\newcommand{\betaj}{\ensuremath{\beta^j}}  % \beta^{\bB_j}
-\newcommand{\betak}{\ensuremath{\beta^k}}  % \beta^{\bB_k} or \beta^{\bB_{k}}
-\newcommand{\betal}{\ensuremath{\beta^\ell}}  %\beta^{\bB_\ell} or \beta^{\bB_{\ell}}
-\newcommand{\uacalc}{{\small UAC}alc}
-\newcommand{\Tuma}{T\r{u}ma}
-\newcommand{\Palfy}{P\'alfy}
-\newcommand{\Pudlak}{Pudl\'ak}
-\newcommand{\PP}{P\'alfy-Pudl\'ak}
-\newcommand{\PAP}{P\'alfy\ and Pudl\'ak}
-\newcommand{\<}{\ensuremath{\langle}}
-\renewcommand{\>}{\ensuremath{\rangle}}
-\newcommand{\beps}{\ensuremath{\boldsymbol{\varepsilon}}}
-\newcommand{\bN}{\ensuremath{\mathbf{N}}}
-\newcommand{\bA}{\ensuremath{\mathbf{A}}}
-\newcommand{\bB}{\ensuremath{\mathbf{B}}}
-\newcommand{\bBi}{\ensuremath{\mathbf{B}_i}}
-\newcommand{\sB}{\ensuremath{\mathcal{B}}}
-\newcommand{\sC}{\ensuremath{\mathcal{C}}}
-\newcommand{\sS}{\ensuremath{\mathscr{S}}}
-\newcommand{\sT}{\ensuremath{\mathscr{T}}}
-\newcommand{\sI}{\ensuremath{\mathcal{I}}}
-\newcommand{\sE}{\ensuremath{\mathcal{E}}}
-\newcommand{\sO}{\ensuremath{\mathcal{O}}}
-%\newcommand{\id}{\ensuremath{\mathrm{id}}}
-\DeclareMathOperator{\id}{id}
-%\newcommand{\Eq}{\ensuremath{\mathrm{Eq}}}
-\DeclareMathOperator{\Eq}{Eq}
-%\newcommand{\Cg}{\ensuremath{\mathrm{Cg}}}
-\DeclareMathOperator{\Cg}{Cg}
-%\newcommand{\Con}{\ensuremath{\mathrm{Con\,}}}
-\DeclareMathOperator{\Con}{Con}
-%\newcommand{\Sub}{\ensuremath{\mathrm{Sub}}}
-\DeclareMathOperator{\Sub}{Sub}
-%\newcommand{\Pol}{\ensuremath{\mathrm{Pol}}}
-\DeclareMathOperator{\Pol}{Pol}
-%\newcommand{\Clo}{\ensuremath{\mathrm{Clo}}}
-\DeclareMathOperator{\Clo}{Clo}
-\newcommand{\ps}[1]{\ensuremath{^{(#1)}}}
-\newcommand{\piB}{\ensuremath{\pi_B}}
-\newcommand{\hpsi}{\ensuremath{\widehat{\psi}}}
-\newcommand{\htheta}{\ensuremath{\widehat{\theta}}}
-\newcommand{\supi}{\ensuremath{^{i}}}
-\newcommand{\supj}{\ensuremath{^{j}}}
-\renewcommand{\leq}{\ensuremath{\leqslant}}
-\renewcommand{\nleq}{\ensuremath{\nleqslant}}
-\renewcommand{\geq}{\ensuremath{\geqslant}}
-\newcommand{\meet}{\ensuremath{\wedge}}
-\newcommand{\join}{\ensuremath{\vee}}
-\newcommand{\Meet}{\ensuremath{\bigwedge}}
-\renewcommand{\Join}{\ensuremath{\bigvee}}
-\newcommand{\resB}{\ensuremath{|_{_B}}}
-\newcommand{\resBi}{\ensuremath{|_{_{B_i}}}}
-\newcommand{\eps}{\ensuremath{\varepsilon}}
-\newcommand{\hatmap}{\ensuremath{\widehat{\phantom{x}}}}
-\newcommand{\one}{\ensuremath{\mathbf{1}}}
-\newcommand{\two}{\ensuremath{\mathbf{2}}}
-\newcommand{\three}{\ensuremath{\mathbf{3}}}
-\newcommand{\tbeta}{\ensuremath{\widetilde{\beta}}}
-\newcommand{\ttheta}{\ensuremath{\widetilde{\theta}}}
-\newcommand{\hbeta}{\ensuremath{\widehat{\beta}}}
-\newcommand{\cick}{\ensuremath{\sC_i}}
-%\newcommand{\CE}{\ensuremath{\sC_r^{\sE}}}
-\newcommand{\CE}{\ensuremath{\sB_r^n}}
-\newcommand{\CO}{\ensuremath{\sC_r^{\sO}}}
-\newcommand{\CICK}{\ensuremath{c_i/\beta \cup c^{K+1}_i/\beta^{\bB_{K+1}}}}
-\newcommand{\GAP}{{\small GAP}}
-\newcommand{\dotsize}{.8pt}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{document}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%  FRONT MATTER                                                 %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\title[Approximating Eigenvalues of Large Stochastic Matrices]{A Lanczos Procedure for Approximating Eigenvalues of Large Stochastic Matrices}
-
-%% First author (Note: The order of the items here is important!)
-\author{William DeMeo}
-\email{williamdemeo@gmail.com}
-\urladdr{http://www.math.sc.edu/~demeow}
-\address{Department of Mathematics\\
-University of South Carolina\\Columbia 29208\\USA}
-
-\thanks{The author wishes to thank a number of people who contributed to this
-  work.  In particular, the primary referee made many insightful recommendations
-  which led vast improvements, not only in the overall exposition, but also in
-  the specific algebraic constructions, theorem statements, and proofs.
-  This article is dedicated to Ralph Freese and Bill Lampe, who declined to be
-  named as co-authors despite their substantial contributions to the project.}
-
-%% Dedication (Optional)
-\dedicatory{This article is dedicated to Ralph Freese and Bill
-  Lampe.}
-
-%% AMS subject classification; see http://www.ams.org/msc
-%% Only one Primary. Possibly several Secondary.
-\subjclass[2010]{Primary: 08A30; Secondary: 08A60, 06B10.} %, 06B15.}
-
-%% Keywords and phrases
-\keywords{congruence lattice, finite algebra, finite lattice
-  representations}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%  MAIN MATTER                                                  %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{abstract}
-The rate at which a Markov chain converges to a given probability distribution
-has long been an active area of research. Well known bounds on this rate of
-convergence involve the subdominant eigenvalue of the chain’s underlying
-transition probability matrix. However, many transition probability matrices are
-so large that we are unable to store even a vector of the matrix in fast
-computer memory. Thus, traditional methods for approximating eigenvalues are
-rendered useless.
-
-In this paper we demonstrate that, if the Markov chain is reversible, and we
-understand the structure of the chain, we can derive the coefficients of the
-traditional Lanczos algorithm without storing a single vector. We do this by
-considering the variational properties of observables on the chain’s state
-space. In the process we present the classical theory which relates the
-information contained in the Lanczos coefficients to the eigenvalues of the
-Markov chain. 
-\end{abstract}
-
-\maketitle
-
-%%%%%%%%%%%%%%%%%%%%%
-\chapter{Introduction}
-The rate at which a Markov chain converges to a given probability distribution has long
-been an active area of research. This is not surprising considering this problem’s relevance to the ar-
-eas of statistics, statistical mechanics, and computer science. Markov Chain Monte Carlo (MCMC)
-algorithms provide important examples. These algorithms come in handy when we encounter a
-complicated probability distribution from which we want to draw random samples. In statistical
-mechanics, we might wish to estimate the phase average of a function on the state space. Goodman
-and Sokal [6] examine Monte Carlo methods in this context. Examples from statistics occur in the
-Bayesian paradigm when we are forced to simulate an unwieldy posterior distribution (see, e.g.,
-Geman and Geman
-
-To implement the MCMC algorithm, we invent a Markov chain that converges to the
-desired distribution (this is often accomplished using the Metropolis algorithm
-described in Chapter 5). Realizations of the chain will eventually represent
-samples from this distribution. Sometimes ``eventually'' -- meaning all but
-finitely many terms of the chain -- is just not enough. We need more practical
-results. In particular, we want to know how many terms of the chain should be 
-discarded before we are sampling from a distribution that is close (in total variation distance) to
-% 03.txt
-the distribution of interest. This is the purpose of bounding convergence rates for Markov chains.
-
-Often the Markov chains encountered in this context satisfy a condition known in the
-physics literature as detailed balance. Probabilists call chains with this property reversible. This
-simply means that the chain has the same probability law whether time moves
-forward or 
-backward.\footnote{This is not a precise definition. In particular the chain must
-  have started from its stationary distribution. Full rigor is postponed until Section 2.1.}
-In this paper, we consider the rate at which such chains converge to a \emph{stationary distribution}.\footnote{This and other italicized terms are defined in Section 2.1.}
-
-There are a number of different methods in common use for bounding convergence rates of
-Markov chains, and a good review of these methods with many references can be found in  More
-recently developed methods, employing logarithmic Sobolev inequalities, are reviewed in  Most
-of the bounds in common use involve the sub—dominant eigenvalue of the Markov chain's transition
-probability matrix, and thus require good approximations to such eigenvalues. In many applications,
-however, the transition probability matrix is so large that it becomes impossible to store even a
-single vector of the matrix in conventional computer memory. These so called out-of-core problems
-are not amenable to traditional eigenvalue algorithms\footnote{By ``traditional
-  eigenvalue algorithms'' we refer to those found, for example, in Golub and Van
-  Loan[5]. See also the book by Demmel [1] for a more recent discussion.}
-without modification. This paper develops 
-such a modification for the Markov chain eigenvalue problem. In particular it develops a method
-for approximating the first few eigenvalues of a transition probability matrix when we know the
-general structure of the underlying Markov chain. The method does not require storage of large
-matrices or vectors. Instead we need only simulate the Markov chain, and conduct a statistical
-analysis of the simulation.
-
-Here is a look at what follows. Section 2.1 contains a review of the relevant Markov chain
-theory. Readers conversant in the asymptotic theory of Markov chains might wish to at least skim
-Section 2.1, if only to become familiar with our notation. Section 2.2 describes functions on the state
-% 04.txt
-space of the Markov process. This section and Chapter 3 develop the context in which we formulate
-the new ideas of the paper. In the last section of Chapter 3, Section 3.3, we present the familiar
-Krylov subspace and explain why this represents our best approximation to a subspace containing
-extremal eigenvectors of the transition probability matrix.\footnote{or, more
-  precisely, a similarity transformation of this matrix.} The first section of
-Chapter 4 describes the \emph{Lanczos algorithm} for generating an orthonormal basis
-for the Krylov subspace. As it stands, this algorithm is useless for an
-out-of-core problem such as ours since, by definition of such problems, 
-it requires too much data movement; all the computing time is spent swapping data between slow
-and fast memory (e.g. between the hard disk and cache). Therefore, we discuss alternatives to
-Lanczos and demonstrate that the \emph{Lanczos coeficients} are readily available through simulations of
-the Markov chain, which fact allows us to avoid the standard algorithm altogether. Following this
-is a chapter describing the Metropolis algorithm used to produce a reversible stochastic matrix.
-It is here that we experiment with the procedure described in Section 4.2 and approximate the
-extremal eigenvalues of the matrix, without storing any of its vectors. Finally, Chapter 6 concludes
-the paper.
-% 05.txt
-
-\chapter{Markov Chains}
-
-\section{General Theory}
-
-This review of Markov chain theory can be found in any good probability text. The present
-discussion is most similar to that of Durrett [3], to which we refer the reader desiring greater detail.
-
-\subsection{The Basic Setup}
-
-Heuristically, a Markov chain is a stochastic process with a lack of memory property. Here
-this means that the future of the process, given its past behavior and its present state, depends only
-on its present state. This is the probabilistic analogue of a familiar property of classical particle
-systems. Given the position and velocities of all particles at time t, the equations of motion can be
-completely solved for the future evolution of the system. Thus, information describing the behavior
-of the process prior to time t is superﬂuous. To be a bit more precise, if technical, we need the
-following definitions.
-
-\bibliographystyle{spmpsci}
-\bibliography{wjd}
-
-\end{document}
-
-%%%%%%%%%%%%%%%%%%%%%  END DOCUMENT %%%%%%%%%%%%%%%%%%%%%%%%%
-
-Definition 2.1.1 Let (S, 8) be a measurable space. A sequence X,., n 2 0, of random variables
-
-taking values in S is said to be a Markov chain with respect to the filtration a(Xo, . . . ,X,,) if for
-% 06.txt
-a.llBES,
-
-p(x,,+1 e B | a(Xo, . . . ,X,.)) = P(X,.+1 e B 1 o(X,,)) (2.1)
-
-Equation (2.1) merely states that, given the present location of X,,, the past is irrelevant for
-
-predicting the location of X,,+1.
-
-Definition 2.1.2 A function p : S x S —> R is said to be a transition probability if:
-1. for each :1: E S, A —> p(a:, A) is a probability measure on (S, S).
-2. for each A E 8, :1: —> p(z, A) is a measurable function.
-
-We call X,, a Markov chain with transition probabilities 12,, if
-P(Xn+l E B l a(Xn)) =Pn(XnaB) (22)
-
-The spaces (S, S) that we encounter below are standard Borel spaces, so the existence of the
-transition probabilities follows from the existence of regular conditional probabilities on Borel spaces
-— a standard measure theory result (see e.g. [3] page 230).
-
-Suppose we are given an initial probability distribution p on (S, S) and a sequence p,, of
-
-transition probabilities. We can define a consistent set of finite dimensional distributions by
-
-P<X,- e B,-,0 s 1 s n) = / #(d1‘o) / Po($o.d-'51) . .. / Pn-l(97n—1ad$n) (2.3)
-B0 B1 B1:
-Furthermore, denote our probability space by
-(9,?) = (S{0,1,...},8{0,l,...})
-
-We call this sequence space and it is defined more explicitly by
-
-sf°»1=-~-} = {(wo,w1,...) :w.- e S}
-
-S{°*1""} = ¢7(w :40; E A,- E 3)
-
-% 07.txt
-The Markov chain that we will study on this space is simply X" (w) = w,., the coordinate maps.
-Then, by the Kolmogorov extension theorem, there exists a unique probability measure P,, on
-(9,?) so that the X" (w) have finite dimensional distributions (2.3).
-
-If instead of p, we begin with the initial distribution 6,, i.e. point mass at 2:, then we
-denote the probability measure by P3. With such measures defined for each 2:, we can in turn
-
-define distributions P,,, given any initial distribution p, by
-
-PM = / u<da.->P.(A>
-
-That the foregoing construction — which, recall, was derived merely from an initial distri-
-bution p and a sequence p,, of transition probabilities — satisfies Definition (2.2) of a Markov chain
-is not obvious, and a proof can be found in
-
-To state the converse of the foregoing, if X,, is a Markov chain with transition probabilities
-p,, and initial distribution p, then its finite dimensional distributions are given by (2.3). Proof of
-this is also found in
-
-Now that we have put the theory on a firm, if abstract, foundation, we can bring the
-discussion down to earth by making the forgoing a little more concrete. First, we specialize our
-study of Markov chains by assuming that our chain is temporally homogeneous, which means that
-the transition probabilities do not depend on time; i.e. p,,(w,,, B) E p(w,., B). (This is the stochastic
-analogue of a conservative system.) Next we assume that our state space S is finite, and suppose
-for all states 1', j E S that p(i, j) Z 0, and E, p(z', j) = 1 for all i. In this case, equation (2.2) takes
-
-a more intuitive form:
-P(Xn+l =j I X7: = 7:) =P(iaj)
-
-and our transition probabilities become
-% 08.txt
-If P is a matrix whose (i, j)th element is the transition probability p(i, 1') then P is a stochastic
-
-matriar, that is, a matrix with elements pgj satisfying
-
-Pij 20: zjpij =1: (i7j=1121"‘)d)
-
-We also refer to P as the transition probability matrix.
-
-Without loss of generality, we can further suppose our Markov chain is irreducible. This
-means that, for any states 2', 1‘, starting in state i the chain will make a transition to state 1' at
-some future time with positive probability. This state of affairs is often described by saying that
-all states communicate. We lose no generality with this assumption because any reducible Markov
-chain can be factored into irreducible classes of states which can each be studied separately.
-
-The final two conditions we place on the Markov chains considered below will cost us
-some generality. Nonetheless, there remain many examples of chains meeting these conditions,
-thus making the present study worthwhile. Furthermore, it may be the case that, with a little
-more work, we will be able to drop these conditions in future studies. The first condition is that
-the chain is aperiodic. If we let I, = {n 2 1 : p"(:c,a:) > 0}, we call a Markov chain aperiodic if,
-for any state 2:, the greatest common divisor of I, is 1. The second assumption is that our chain is
-
-reversible. This characterization is understood in terms of the following definition.
-Definition 2.1.3 A measure a is called reversible if it satisfies
-
-;4(w)p(I. y) = u(y)p(y. w)
-
-for all :c and y.
-
-We call a Markov chain reversible if its stationary distribution (defined in Section 2.1.2) is reversible.
-% 09.txt
-\subsection{A Convergence Theorem}
-
-In succeeding arguments, we use some results concerning the asymptotic behavior of
-
-Markov diains. These results require a few more definitions.
-
-Definition 2.1.4 A measure 1r is said to be a stationary measure if
-
-Z«(z)p<z.y> = we) (2.4)
-
-2
-Equation 2.4 says P,,(X1 = y) = 1r(y), and by induction that P,,(X,, = y) = 7r(y) for all n 2 1. If 1r
-is a probability measure, then we call 1r a stationary distribution. It represents an equilibrium for
-the chain in the sense that, if X0 has distribution 1r, then so does X,. for all 11..
-
-When the Markov chain is irreducible and aperiodic, the distribution of the state at time
-n converges pointwise to 1r as n —> oo, regardless of the initial state. It is convenient to state
-this convergence result in terms of the Markov chain's transition probability matrix P. Before
-doing so, we note that irreducibility of a Markov chain is equivalent to irreducibility (in the usual
-matrix theory sense) of its transition probability matrix. Furthermore, it turns out that a transition
-probability matrix of an aperiodic Markov chain falls into that class of matrices often called acyclic,
-but for simplicity we will call such stochastic matrices aperiodic. With this terminology, we can
-
-state the convergence theorem in terms of the transition probability matrix P.
-
-% Theorem 2.1.1
-\begin{theorem}
-\label{thm-2.1.1}
-Suppose P is irreducible, aperiodic, and has stationary distribution 7r. Then as n —> oo,
-p"(i,J') -> #0")
-\end{theorem}
-
-The symbol p"(i, j) is to be understood as the (ij) element of the nth power of P.
-A Markov chain whose transition probability matrix satisfies the hypotheses of
-Theorem~\ref{thm-2.1.1} is called ergodic. If we simulate an ergodic chain for
-sufficiently many steps, having 
-% 10.txt
-begun in any initial state, the final state is a sample point from a distribution which is close to 1r.
-
-To make this statement more precise requires that we define “close”.
-
-Definition 2.1.5 Let 1r be a probability measure on S. Then the total variation distance at time
-
-n with initial state I is given by
-AA") = l|P"(z, A) - 7r(-4)||Tv = 1333; I P"(I.A) - 7r(A) I
-
-In what follows, we will measure rate of convergence using the function 'r,(e), the first time after
-
-which the total variation distance is always less than 5. That is,
--r,,(e) = min{m : A,(n) 3 E for all n. 2 m}
-
-To consider the connection between convergence rates and eigenvalues, note that an ape-
-riodic stochastic matrix P has real eigenvalues 1 = /\o > A1 2 A2 3  Z /\¢_1 2 -1, where
-d = |S'] is the dimension of the state space. For an ergodic chain, the rate of convergence to the
-stationary distribution 7r is bounded by a function of the subdominant eigenvalue. By subdominant
-eigenvalue we mean that eigenvalue which is second-largest in absolute value, and we denote this
-Am“ E max{).1,|A¢_1|}. The function bounding the rate of convergence is given by the following
-
-theorem:
-
-% Theorem 2.1.2
-\begin{theorem}
-\label{thm-2.1.2}
-The quantity T1(€) satisfies
-1. 13(5) 5 (1 — /\m,_x)“(ln1r(a:)‘1 +1ne‘1)
-2. maxxes r,(e) 2 %/\,,,,x(1 — ).,,,,x)‘1ln(2e)’1
-\end{theorem}
-
-As this theorem shows, if we have an upper bound on the subdominant eigenvalue, then we have
-an upper bound on the function 11(5). In what follows, we will derive an approximation to the
-% 11.txt
-subdominant eigenvalue and supply error bounds. Together, an approximation and error bounds
-for Am“ provide enough information to make Theorem~\ref{thm-2.1.2} useful.
-
-\section{Functions on the State Space}
-
-Recall that X,.(w) = w,. 6 S denotes the state in which the Markov chain exists at time
-11. Suppose that 11> = {¢1, . . . , ¢,,} is a collection of p observables, or functions defined on the state
-space S. Furthermore, let these observables be real valued, so ¢.~ : S —> R. It is often useful, as we
-will see below, to assume that none of the observables is a constant function. Suppose now that
-the state space S is finite with at possible states. Then, since an observable is simply a map of the
-state space, we can think of each ¢,- as a vector of d real numbers — the d values that it takes on at
-the different states.
-
-Now assume the Markov chain is irreducible, and let 1r denote its stationary distribution.
-If we start the chain from its stationary distribution - i.e. suppose X0 has distribution 1r - then
-X,. is a stationary process. Furthermore, for each i, ¢,(X,.) is a stationary stochastic process with
-
-mean
-
-E.¢.- = Z) vr(x)¢.(z)
-
-165
-
-and autocovariance function
-
-Cir (¢i(Xn)v ¢i (Xn+a))
-
-E7!’  ‘ E1r¢i) (¢i(Xn+a) " E1r¢i)l
-
-Z Pn(Xn = z1X7I+S = y)  — E1r¢i)  _ E1r¢i)
-
-z,yES
-(2.5)
-
-which, by the definition of conditional probability, we can write as,
-
-2 P«(X.. = z)P.(X..+.. = y I X. = 1)(¢s($) — Em) (my) ~ Em)
-1.1165
-% 12.txt
-Z 7r($)P1y (MI) - Em.-) (¢:(1I) - E1r¢i)
-
-:c,y€S
-
-Here pgy denotes the element in row :1: and column y of P’, the sth power of the transition probability
-
-matrix. Similarly, we define the cross-covariance between the function ¢,- at time n and q5,- at time
-
-n+sas
-
-E1!’  _ E7r¢:') (¢i(Xn+a) " E1r¢i)l (2-6)
-
-C1r(¢i(Xn)a (#1 (Xn+a))
-
-Z) vr(x)p;,, (¢.-(w) — Em.-) (my) — Em)
-
-z,yES
-
-Now let (<I>) denote the matrix of mean vectors whose jth column is E,,¢,-1, where 1 = (1, . . . , 1)‘,
-and let H = diag(1r(w1), . . . , 1r(wd)), be the d X d diagonal matrix with stationary probabilities 1r(w)
-on the main diagonal, and zeros elsewhere. Finally, denoting by C(s) the p X p covariance matrix
-
-whose (i,j)th element is C, (qfi,-(X,.),¢_,< (X,.+,)), we have
-
-0(0) = E(‘1’(Xn) - (‘I>))(‘I’(Xn) - (‘PW
-= (‘P-(‘1’))‘1'1(<1’-(‘I>))
-0(8) = E(<I>(Xn) — (‘I>))(‘1’(Xn+a) - (‘1’))’
-
-(‘P - (‘1’))'1'IP‘(<1> — (<1>))
-
-Below, we will also find it useful to have at our disposal a new matrix that is similar to the
-transition probability matrix. We have in mind the matrix M E Ill/2PI'I‘1/2. As is easily verified,
-this allows us to write the covariance matrix as
-
-can = <<I>—<<I>>)°n%*M*n%<<I>—<<I>>>
-
-\II‘M’\II (2.7)
-
-where we have defined ‘II E 111/ 2(<I> —  Recall that our main motivation is that, for out-of-core
-problems, traditional eigenvalue algorithms are inadequate. With this fact, and the form (2.7) in
-% 13.txt
-mind, we will consider using the covariance of observables on the state space to implement the
-(otherwise useless) Rayleigh-Ritz procedure, which we describe below. This procedure requires
-that M be symmetric. As the next fact demonstrates, this required symmetry is the reason for our
-
-interest in the reversibility property of the Markov chain.
-
-Fact 2.2.1 The matrix M is symmetric if and only if the Markov chain is reversible (i.e. iﬂ the
-
-process satisfies the detailed balance condition).
-
-Proof:
-
-M is symmetric
-
-4:, (1-I1/2P1-I-1/2): = H1/2Pn—1/2
-<.—> r1-%*P°n%* = H1/2911-1/2
-¢> 1>'r1= = up
-~:=> (IIP)‘ = l'IP
-
-Element-wise, the final equality implies that 7|‘,'p,'j = 1rjpj.-. According to Definition 2.1.3, this states
-
-that p is a reversible measure.
-
-% 14.txt
-
-\chapter{Invariant and Approximate Invariant Subspaces}
-
-\section{Invariant Subspaces}
-Definition 3.1.1 A subspace S Q R" with the property that
-
-1:ES'=Ma:€S
-
-is said to be invariant for M.
-
-Recall that, having chosen observables <I> = ((151, . . . , qfip), we constructed the covariance
-matrix C(s) = \II‘M"\II. If the column space of \II, which we denote by ran(\II), is an invariant
-subspace for M, the definition implies Mifij E  That is, for each 111,- there exists a vector
-t of coeﬂicients such that M1/Jj = f=1t,-1/)5,-. This is true for all j and, putting each vector of
-coefficients into a matrix T, we see that M\II = \IIT. Conversely, M\II = IIIT implies that M1/Jj is a
-
-linear combination of columns of \II, so ran(\II) is invariant. We have thus proved the following
-% 15.txt
-Fact 3.1.1 The subspace ran(\II) is invariant for M if and only if there exists T E R”? such that
-
-M\II = \IIT.
-
-Consequently,
-
-Fact 3.1.2 )\(T) Q ).(M)
-
-Proof of 3.1.2:
-
-z\E)\(T) _=_ 3v€R"3Tv=).v
-<=> \IITv = A\IIv
-<=> M\II-u = /\\IIv
-< /\ E /\(M)
-
-The second ¢> follows from fact 3.1.1.
-Facts 3.1.1 and 3.1.2 are theoretically useful. To see why, write the equation of fact 3.1.1
-
-as follows:
-
-\IIT = M\II
-\II‘\IIT = \II‘M\II
-T = (\1:'~1r)-‘\1:'M\1:
-
-In this form, we recognize that T = C‘1(0)C(1). That is, using only the covariance of observables
-on the state space, we can generate a matrix T with the property A(T) (_Z ).(M). Recalling that
-M = H1/2Pl'I‘1/2, we see that M is similar to our original stochastic matrix P, and thus ).(M) =
-
-).(P).
-
-\section{Approximate Invariant Subspace}
-
-We qualiﬁed the foregoing by stating that the facts are only theoretically useful. This
-
-is because when choosing observables of the state space we aren't sure that they will satisfy the
-% 16.txt
-primary assumption underlying facts 3.1.1 and 3.1.2. Recall the assumption: ran(\II) is an invariant
-subspace for M, where ‘II = H1/2(<I> —  It may well be the case that there is a 1p; 6 ran(\II) such
-that M1/zj ¢ ran(\II), thereby violating the assumption. However, lacking an invariant subspace, we
-
-might hope that our observables at least provide an approximate invariant subspace.
-
-Deﬁnition 3.2.1 If the columns of ‘II E R“? are independent and the residual matrix E E
-
-M\II —— \IIT is small for some T 6 R7”, then ran(\II) deﬁnes an approximate invariant subspace.
-
-To see how an approximate invariant subspace can be useful in approximating eigenvalues of M,
-we extract a theorem from Golub and Van Loan 
-% Theorem 3.2.1
-\begin{theorem}
-\label{thm-3.2.1}  
-Suppose M 6 Rd” and T 6 RP"? are symmetric and let E E MQ — QT, where Q 6 Rd"? satisﬁes
-
-Q‘Q = I. Then there exist p1,... ,;z,, E A(T) and A1,. . . , A,, E A(M) such that
-ll‘k - Akl S \/§|lE"
-
-fork=1,...,p.
-\end{theorem}
-
-If the subspace ran(Q) is an approximate invariant subspace, then the deﬁnition implies that
-there is a choice T rendering the error IIEII2 small, and thus, the eigenvalues of T provide a good
-approximation to those of M.
-
-When considering the preceding ideas, it becomes apparent that their application will
-present new — but hopefully less prohibitive — problems. As these problems summarize the rt of
-the paper, now is a good time to examine them. First, the preceding theorem assumes a matrix Q
-whose columns form an orthonormal basis for the approximate invariant subspace. For our problem,
-derivation of such a Q is tricky, and we must prepare for this. Next, having an approximate invariant
-
-subspace at our disposal merely tells us that there exists some matrix T which makes the error
-% 17.txt
-"EH2 small. We must discover the form of such a T. Moreover, it is natural to seek that T which
-minimizes IIEII2 for a given approximate invariant subspace. Finally, in order to apply these ideas
-to real examples of our eigenvalue problem, we must ﬁnd a practical way to generate the most
-appropriate approximate invariant subspace. We now address each of these issues in the order
-raised.
-
-Recall the matrix I1 E III/2(<I> — (<I>)). The columns of this matrix, though independent
-(by choice of independent observables), are not necessarily orthonormal. However, consider the
-polar decomposition ‘II = QZ where Q‘Q = I and Z2 = \Il‘\II is symmetric positive semideﬁnite.1
-
-Note that Z = (\II‘\II)l/2 is nonsingular, so Q has the form
-Q = ~1:z-1 = \II(\II‘\II)“/2
-
-and it is clear that ran(Q) = ran(\II). Perhaps ra.n(Q) is a. useful approximation to the invariant
-subspace for M. If ran(Q) is not itself invariant, we have the error E = MQ — QT. It is easy to
-show, as we do in the following proof, that the T which minimizes HE“; is T = Q‘MQ. This yields
-
-the following,
-Theorem 3.2.2
-
-If \I1 = QZ is the polar decomposition of \II, then the matrix
-T = Q‘MQ = (\1:*\1:)-1/’\1:‘M\I:(\Ir‘\I:)-1/2
-minimizes IIEII2 = IIMQ - QTII2
-
-Proof:
-We prove the result by establishing the following
-
-(Hal: Given M 6 Rd” suppose Q 6 Rd"? satisﬁes Q'Q =1. Then,
-
-min ||MQ — QTII2 = H(1- QQ')MQll2
-
-TERI‘?
-
-‘Recall that the polar decomposition is derived from the SVD, \II = UZJV‘, by letting Q = UV‘ and Z = VEV‘.
-
-% 18.txt
-and T = Q’MQ is the minimizer.
-To verify the claim, observe the following application of the Pythagorean theorem:
-
-For any T E R.”"” we have
-
-IIMQ — QT||§ = IIMQ - QQ'MQ + QQ'MQ - QT|l§
-||(I - QQ’)MQ + Q(Q'MQ - T) H3
-
-||(1 - QQ‘)MQ||§ + |lQ(Q'MQ - T)||§ (3-1)
-
-2 l|(I — QQ')MQ||§
-
-Equality (3.1) holds since I — QQ' projects MQ onto the subspace orthogonal to ran(Q). Thus,
-the two terms in the expression on the right are orthogonal, and the Pythagorean theorem yields
-equality. The concluding inequality establishes that the minimizing T is that which annihilates the
-second term in (3.1); i.e. T = Q‘MQ.
-
-The second equality in Theorem 3.2.2 is a consequence of the polar decomposition, in
-
-which Q = \IIZ“ = \I:(~I:‘\I:)-1/2. Therefore,
-T = (x1z'x1:)-1/2x1:‘M~1:(\1:*~1z)—1/2
-
-is the minimizer.
-
-3.3 The Krylov Subspace
-
-Suppose the columns of a matrix Q E R“? provide an orthonormal basis for an approxi-
-
-mate invariant subspace. Then we have seen,
-1. T = Q'MQ minimizes ||E||2 = ||MQ — QT||2 and
-2. E|p.1,...,p,, E z\(T) and /\1,...,A, 6 »\(M) such that
-
-ll-‘I: - Ml S \/§|lE||2
-
-% 19.txt
-19
-fork=1,...,p.
-
-Given an approximate invariant subspace ran(Q) of dimension 1), these facts tell us what matrix we
-should use to approximate p elements of M’s spectrum. Now all we lack is a. description of ran(Q).
-That is, we have not speciﬁed which approximate invariant subspace would best suit our objective
-
-of approximating the subdominant eigenvalue A,,.a2:(M). 'I‘o do so requires the following deﬁnition:
-
-Deﬁnition 3.3.1 The Raleigh quotient of a symmetric matrix M and a nonzero vector 3 is
-
-p(2:, M) E (:z:‘Ma:)/(z‘a:).
-
-We will also denote this by p(z) when the context makes clear what matrix is involved.
-
-To ﬁnd the approximate invariant subspace most appropriate for our problem, we choose
-each dimension successively, providing justiﬁcation at each step. Thus, start with one observable
-42 on the state space, and let zpl E 111/ 2(¢ — E,,¢). That is, 1/:1 represents a centered (mean zero)
-observable whose ith coordinate is weighted by . First notice that the deﬁnition (which comes
-directly from the deﬁnition of ‘II following equation (2.7)) is such that 1/)1 is a constant vector if
-and only if :15 is oonstant on the state space, in which case 1/); E 0. This fact provides one reason
-observables that are constant on the state space are not interesting. Second, recall that the row
-sums of the matrix P are all one and, therefore, the eigenvector corresponding to the eigenvalue
-Ao(P) = 1 is the constant vector. By deﬁnition of M E 1'11/2Pl'I‘1/2, we see that the constant vector
-is also the eigenvector corresponding to A0  This will play an important role in what follows, as
-it allows us to focus primarily on the subdominant eigenvalue z\max(M) E max{}q (M), [A44 
-rather than on A0  (which we already know is 1).
-
-Now, notice that
-
-lP(¢1,M)l 5 max |P(x,M)l = »\max(M) (3-2)
-% 20.txt
-where the max is taken over all non-constant vectors. Since our interest centers on An-.3_x(M), we
-would like a. 1/11 which makes the left hand side of (3.2) large. This would be achieved if 1/)1 were to
-lie in the space spanned by, say, the first two eigenvectors of the Markov chain (sometimes referred
-to as the slowest modes of the process). However, this subspace spans only two dimensions of the
-entire d-dimensional space, and it is more likely that ¢1, at best, only comes close to lying in the
-subspace of interest. Now, given 1/11, a judicious choice for the second dimension l/)2, and hence
-W; E [1/)1 1/:2], would be that which makes max,,¢o |p(\II2a)| is as large as possible. To establish that
-
-this is indeed the pertinent objective, note the following:
-
-_ a‘\1:5Mx1:2a
-I;1;g<Ip(\I'2a,M)l = 13:3: ————a,W§q,2a
-= M
-mrggf,mIp(z. )1
-5 maxlp(z,M)l (3-3)
-= ).,,m(M)
-
-Again, the max on the right side of (3.3) is over non-constant vectors. In words, we wish to chose
-\II-2 = [1/:1 ¢2] so that there is a vector a 6 R2 making |p(\II2a,M)| close to )«max(M).
-
-Now, p(1/11) changes most rapidly in the direction of the gradient Vp(¢1).
-
-W0  (iiiiiiv---viiiléi)
-= ,,;,l (Mam — p<¢.)«m (3.4)
-
-So, to maximize the left hand side of (3.3), ‘I12 should be chosen so that the subspace ra.n(\II2)
-
-contains the gradient vector. That is, we must choose 1/22 so that
-
-VP(¢'1) E F3l1{¢1y¢2} E F-'%!1(‘I’2) (3-5)
-
-Clearly, equation (3.4) implies Vp(1/:1) E ran{z/11,M1,b1}. Therefore, if ran{1/:1,1/)2} = ra.n{¢1,M¢1},
-
-then (3.5) is satisﬁed.
-% 21.txt
-In general, having chosen \II,, E [1/)1 - -- 1/11,] so that
-
-ran{'pl9¢21' " 1¢k} = ran{¢1vM1[)la ' ‘ ' vMk—11/’k}
-
-we must chose 1/1;.“ so that, for any non-zero vector u 6 R",
-
-VP(‘I’I:a) 6 ra‘n{¢11 ‘21 - ° - 31/Jk-F1} 
-Now,
-vp<~I~.a) = gmm — P(‘I’kﬂ)‘I’k“)
-
-and therefore,
-
-Vp(‘II;,a.) E ran(M\II;,) U ra.n(\IIk) = ra.n{¢1, M1/)1, . . . , Mktﬁl}
-
-Thus, requirement (3.6) is met when ran{1/:1, 1/:2, . . . , 1/4+1} = ran{1/)1, M1/:1, . . . , Mk:/11}.
-
-In conclusion, the p-dimensional approximate invariant subspace which is most suitable
-
-for our objective is
-
-/C(Mn/mp) 2 ran{«/:1,M¢1,... ,M"“«/21}
-
-This is known as the K rylov subspace. Therefore, in answer to the problem posed at the outset
-of this section, if we take the columns of Q to be an orthonormal basis for IC(M,z/)1, p), then the
-
-eigenvalues of T = Q‘MQ should provide good estimates of p extremal eigenvalues of M.
-
-% 22.txt
-\chapter{Lanczos Procedures}
-
-The following will review the usual procedure for generating an orthonormal basis for
-lC(M, «/11, p), and conclude by showing that, for our problem, we can find the matrix Q‘MQ without
-actually carrying out this procedure. Note that this conclusion is essential since the procedure
-
-requires matrix vector multiplication - an operation we have assumed impossible for large enough :1.
-
-4.1 The General Lanczos Algorithm
-Let 21;“) E M"¢ and consider the d x :1 matrix
-«I 2 [«/»,M«/»,  ,M"“¢] 2 [«b‘°>,«/»“’,...,¢<‘-‘>1
-
-Notice that M\II = [1/:(1),... ,1/;(d'1),Md¢r] and, assuming for the moment that \I1 is non-singular,
-
-we can compute the vector h = —\II‘1M"1/J. Thus,
-
-M\II = \II[e2,. . . ,e,1, —h] E IIIH
-
-% 23.txt
-23
-
-Equivalently,
-0 0  0 -—h1
-1 0  0 —h2
-HE   3 ; =\II“M\II (4.1)
-0 "hl1—1
-1 —h,;
-
-H is a companion matrix which means that its characteristic polynomial is p(:r) = 2:‘ + 23:1 h,-:r"‘1.
-Since H is similar to M, ﬁnding the eigenvalues of M is equivalent to ﬁnding the roots of p(a:).
-However, this is of little practical use since ﬁnding h, constructing 12(1), and ﬁnding its roots is
-probably a harder problem than the one we started with. Instead, the value of decomposition (4.1)
-derives from its upper Hessenberg form. We exploit this property below.
-
-Let ‘II = QR be the QR decomposition of \II. Since ‘I! is assumed nonsingular,
-~II;‘M~r.: = (R"Q‘)M(QR) = H
-Therefore,
-Q‘MQ = RHR-1 2 T
-
-Since R and R“1 are both upper triangular and H is upper Hessenberg, T = R.HR'1 is also upper
-Hessenberg. Furthermore, since M is symmetric, it is clear that T‘ = Q‘MQ = T. That is, T is
-
-an upper Hessenberg matrix that is symmetric. Therefore, T is tridiagonal and we can write it as
-
-follows:
-
-T =    (4.2)
-
-ﬁd—1
-
-0  174-1 ad
-
-% 24.txt
-24
-
-Equating columns j on both sides of MQ = QT yields
-M113‘ = B,-—1qj—1 + 0:91 + ﬁjqj+1 (43)
-
-Since the columns of Q are orthonormal, multiplying both sides of (4.3) by q_,- and q,-+1 yields
-oz, = q§Mq,- and ﬂj = q;-+1Mq,-.
-The foregoing justifies what is called the Lanczos algorithm, performed as follows:
-
-The Lanczos algorithm for (partial) reduction to symmetric tridiaganal form
-
-91 = ¢/||¢||2, [30 = 0, 40 = 0
-
-for j=1top
-
-z=Mq_,-
-
-5:‘ = ||Z||2
-if 5, = 0, quit
-¢Ij+1 = 2/51
-
-end for
-
-The qj computed by the Lanczos algorithm are often called the Lanczos vectors. If the
-loop in the algorithm is terminated because ﬁg = 0, this indicates that an exact invariant subspace
-has been computed, and is given by ran{q1 - - - qk}. Otherwise, we usually halt the algorithm after
-p steps, in which case the algorithm converg to approximations of at most 12 eigenvalues.‘
-
-Note that equation (4.3) results from the equation MQ = QT. The latter equation holds
-since, starting with d vectors in our subspace, Q is a d Xd matrix whose columns form an orthonormal
-
-basis for all of R‘, which is clearly an invariant “sub” space. The Lanczos algorithm above, however,
-
-‘A detailed discussion of such convergence (and misconvergence) is given in
-
-% 25.txt
-25
-
-proceeds for only p steps, producing a d x p matrix Q, whose columns form an orthonormal basis
-for the approximate invariant subspace IC(M, ¢, p), and a p X p matrix T, = Q'MQ. In that case,
-IIMQ = QTp|[g =  is non-zero and gives the error bound described in Theorem 3.2.1. Now,
-
-writing the full d x :1 matrix T of (4.2) as
-
-01 51
-/71
-
-(4.4)
-
-allows us to describe the error bound ||E||-2 in terms of the submatrix Tm, and, hence, in terms of
-
-ﬁzz-
-Theorem 4.1.1
-
-HT, and T,“ are the matrices appearing in (4.4), and if the p columns of Q are computed by the
-
-Lanczos algorithm, then there exist pl, . . . ,;t,, E /\(T,) and A1, . . . , A, E A(M) such that
-
-ill» - /\I=i S iiTP'4”2 = 512
-
-fork=1,...,p.
-
-For a proof, see page 365 of [1].
-
-% 26.txt
-26
-
-4.2 A Lanczos Procedure for Markov Chains
-
-Notice that the Lanczos algorithm requires the computation of Mqj. Throughout this
-paper we have assumed that M 6 Rd“, and that d is so large as to make the matrix vector
-multiplication Mqj impossible. Now, the Lanczos algorithm constructs an orthonormal basis for
-the Krylov subspace. For our purposes, this construction is not essential. What is essential is that
-we ﬁnd a way to generate the Lanczos coejﬁcients oz,-, ﬁj, (j = 1, . . . , p), and whence the matrix
-T, without performing the operation Mqj required by the Lanczos algorithm. We now address this
-problem.
-
-Begin with the centered and weighted observable 1/1 = 111/ 2 (qS—E,,¢), and let q1 = 11;/||1/JH2.
-To simplify notation, let ¢(n) E ¢(X,.). The form of the first Lanczos coeﬂicient a1 is straight
-
-forward:
-
-01 = qiMq1
-
-I/1‘M1/I
-¢’¢
-
-and now the special properties of the Markov chain setting play an important role. For, regard the
-
-following
-
-¢‘M«/2 = (¢ — E.¢)‘II%‘MII%(¢ - Em)
-
-Cw (d>(n), ¢(" + 1)) (4-5)
-
-and
-
-W1!’ = (<15 - E«¢)'H(¢ - E«¢)
-
-Varn (¢(")) (4-6)
-
-Recall that the deﬁnition of M implies H%‘MI'I% = HP, and this justiﬁes equation (4.5). Putting
-
-% 27.txt
-27
-
-these equations together, we have a. nice form for :11:
-
-_ c, (run), an + 1))
-"‘ ' Ta:,(¢(n))
-
-The next Lanczos coeﬂicient, B1, is only slightly more complicated.
-
-[31 = ||M91-0t1tI1||2
-
-[(Mq1 — 0l1¢11)i(M¢]1 - 0101)] 1/2
-
-[t1iM2¢I1 — ail 1/2
-
-(4.7)
-
-Again recalling the deﬁnition M E 111/ 2P1'I‘1/2, a. simple manipulation veriﬁes
-
-1'li‘M’l'[% = HP’, whereby it follows that
-
-(«s — E,,¢)‘HP”(¢ — Em)
-
-"‘M2"‘ (¢ — E,¢)*n<¢ — Em)
-
-C: (¢("), ¢(n + 2))
-V3-Ur
-
-Inserting the last expression into equation (4.7) gives
-
-_ c,(¢(n).¢<n+2)) c,(¢<n>,¢(n+1)> ’ "’
-"“ jvar,<¢<n)> "(jvar,<¢<n)) ‘)
-
-Notice that we can estimate these expressions of oz; and ﬂl by running simulations of the Markov
-chain. To do so, we run the Markov chain starting from a random initial state, compute the value of
-the observable at each step of the chain and, from the data, compute estimates of the covariances
-and variance of the observable.
-
-Before generalizing this for the jth Lanczos coefficient, we consider one more coeﬂicient
-in detail. Recall that the second basis element q; in the Lanczos algorithm is given by 1;; =
-
-(Mql — oz1q1)/ﬁl. From this we can derive an expression for (12 which, though not as neat as those
-
-% 28.txt
-28
-
-for oz; and [31, is nonetheless tractable.
-
-I12 ‘IiMq2
-
-1
-[7(M91 — 0t1¢I1)tM(M¢I1 - 11191)
-1
-
-1
-= f,§(qiM3q1 - 2a1qlM2<11 + a’9iM¢11)
-1
-1
-= —2(<1iM3q1 - 2015i ‘ all (4-8)
-
-1.
-The last equality follows from equation (4.7), which can be written as q{M2q1 = off’ + (if. The
-preceding paragraphs deﬁne all but one of the quantities in the expression (4.8). The one remaining
-is q§M3q1, which can also be expressed in terms of covariances of <15, and the analysis is almost
-
-identical to that for q{M2q1:
-
-(¢ — E,¢)'IIP3(¢ — Em)
-
-"lM3"‘ (¢ — E,¢s)‘H(¢ - Em»)
-
-Cu (¢(")a ¢(n + 3))
-Varn(¢(n))
-
-To generalize the foregoing for higher order Lanczos coeﬂicients requires a recurrence
-relation between (aj+1, ﬂj) and (aj, ﬂ,~_1). As we will see, such a relation involves higher order
-Lanczos vectors q,~ — vectors we do not wish to compute. Recall that the only “vector” we can
-handle is our initial observable. Therefore, we must also establish a recurrence among the q,-’s, thus
-making them available inductively from our observable vector.
-
-To begin with the ﬂfs, recall that ﬁj E ||Mqj — ajq_.,- — ,H_,~_1q,-_;|[2. Whence,
-
-Bf (M411 — °‘j‘Ij - ﬁj-1qj—1)’(MlIj - Otjqj - 51-19141)
-
-q;-M291‘ - 0? - B,z_1 (4-9)
-
-Some simple algebra and the deﬁnitions aj E q§Mqj and ﬁJ~_1 E q;-Mq,--1 prove equality (4.9). The
-% 29.txt
-recurrence for 0114.1 is equally straight forward, though the ﬁnal form is not as neat:
-
-2
-0tj+1 ¢Ij+1M'1j+1
-
-1
-§(M¢15 - 11111;‘ - ﬂj—1<Ij—1)'M(Mqj - Otjqj - /3j—1qj—1)
-J
-
-1
-§(q;-M-“*4; — 2a,~q;M*q,- — 2x3;+’_1q;--1M*q,- + «I? + 2a,-5,51 + am (4.10)
-J
-From (4.9), q;~M2q,~ = oz? + ﬂf + ﬂ}_1 and, upon plugging this into (4.10), we arrive at
-1
-a,~+1 = E(q§»M3q_,,- — (1? — 2ozJ~ﬁJ2 — 2ﬂj_1q;»_1M2qj + ﬁ2_1) (4.11)
-J
-
-The expressions above involve the higher order Lanczos vectors qj. In the present context,
-these vectors will not be readily available. However, another recursion — this time relating forms
-involving q,-+1 to forms involving qj and qj__1 — will make these expressions accessible by induction
-on our initial observable. The general forms of interest are q;+1M"qj+1 and qj-+1M"qJ~. Again, we
-
-begin with the simpler of the two:
-
-¢I;'+1Mk9j = ij(Mqj - Otjqj - ﬂj—1t1j—1)'M"¢1j
-= i(<1§M"“¢Ij - C‘j‘I_;'Mk‘Ij - ﬂj—1q§M"qj-1)
-The second is,
-‘I_;'+1Mkqj+1 = ﬂi?(M<Ij - Otjqj - .3j—1<Ij—1)‘Mk(Mqj — Orjqj - ﬂj—1¢1j—1)
-
-1
-ﬁ(‘1§M'°"<1:‘ + °‘:2"I§M'°<J:' + ﬁ?—z<Ii—1M"<I:‘—1
-J
-—2a,-q;-M*+‘q,~ + 2am,--1q;-M*q,--1 — zza,--1q;-M*+‘q]--1) (4.12)
-
-The point of all this is that each pair of coeﬂicients (oz,-+1, /3,-) can be expressed as a
-function of the covariance between ¢(n) and ¢(n+ m), for m = 0, . . . , 21' + 1. Since the expressions
-are based on four recurrence relations, they become quite complicated as 1' gets large. However,
-deﬁning these recursions in a. computer program is trivial, and we use this fact in Section 5.4 when
-
-applying the foregoing to a particular problem.
-
-% 30.txt
-30
-
-Chapter 5
-
-An Application: Convergence to
-
-Gibbs Measure on the Ising Lattice
-
-5.1 Physical Motivation
-
-Before considering the details of the Metropolis algorithm, it helps to understand the
-setting in which it was developed. To do so we consider an typical example, the Ising model.
-Consider a model of a system consisting of 71 “sites”, each site taking values in {-1, +1}. It might
-help to imagine these sites as equally spaced points on a line or a circle, though we are not restricted
-to such cases. Then our state space 5' consists of all possible n—dimensional permutations of -1’s
-and +1’s. We reprent this as S’ = {-1, +1}", and note that |S| = 2". We can think ofeach site as
-a node existing in one of two possible positions, say “on" or “off”. Each state 1' E S is represented
-by a unique permutation a.- of —1’s and +1’s. Let a,-(k) = :|:l denote the position (on or off) of
-the kth node when the system is in state 1'. An observable ¢ = qS(a,-) on this space is a function of
-
-the permutations 47,-, 1' 6 {1,. . . , 2"}.
-
-% 31.txt
-31
-
-The energy of the system, when in state i, is deﬁned by the Hamiltonian
-
-H07-‘) = - Z Ua(i)0s(k)
-
-<j Ic>
-where the sum runs over all neart neighbor pairs. In statistical mechanics we are often concerned
-
-with the Gibbs measure of state i, deﬁned by
-1r(a,-) = Z'1 exp{—[3H(a,~)} (5.1)
-where Z ‘1 is the normalization constant, or partition function. For the Ising model, we put
-
-Z = §:9XP{~ﬁH(<7:‘)}
-i=1
-
-so that 7r is a probability measure. Now, letting 1: denote the so called Boltzman constant, if a
-classical mechanical system is in thermal equilibrium with its surroundings with absolute temper-
-ature T, and is in state i with energy H (05), then the probability density in phase-space of the
-point representing state i is given by (5.1) with [3 = (lcT)“. The central result of ergodic theory
-implies that we can also interpret 1r(o;) as the proportion of time the system spends in state i. If
-
-the system is observed at a random time, the expectation E,,¢ of any observable ¢ is thus
-
-.31 ¢(0a) 9"l>{-ﬂH(0s‘)}
-
-E” = 2?;.exp{—nH(a.-)}
-
-(5.2)
-
-Perhaps the number of sites n in our Ising model is so large that it is impractical or impossible to
-evaluate (5.2). We might instead consider importance sampling and generate states with probability
-
-density
-
-exp{—aH<a.o}
-2?;.exp{—zaH<a.->} (53)
-
-Then «I: is itself an unbiased estimator of (5.2).
-
-1r(0a') =
-
-If evaluation of (5.2) is difficult, there is no reason to believe that evaluation of (5.3) is any
-
-easier. However, Metropolis and his collaborators [9] contrived a method for producing an ergodic
-
-% 32.txt
-32
-
-Markov chain with transition probability matrix K whose elements Ic(:::, y) satisfy
-
-2«<z>k<x.y) = «<2»
-
-I
-
-By Deﬁnition 2.1.4 this means that the Markov chain has stationary distribution 7r. Since the
-drain is ergodic, we know that it will eventually converge to the desired distribution. So, our main
-problem — the rate at which such a Markov chain will converge — is clearly at issue in the application
-of the Metropolis algorithm. Moreover, as we will see below, the Metropolis algorithm ensures that
-the Markov chain has the desired stationary distribution by requiring that the chain satisfy the
-stronger condition of reversibility; i.e. detailed balance. Therefore, Markov chains produced by the
-Metropolis algorithm satisfy the assumptions of this paper, which fact conﬁrms that our theory
-
-can be used to bound convergence rates of all such chains.
-
-5.2 Description of the Metropolis Algorithm
-
-The following explains how the Metropolis Algorithm is carried out. The exposition is
-similar to that given in Hammersley and Handscomb’s book Monte Carlo Methods 
-To begin, choose an arbitrary symmetric Markov chain; that is, a Markov chain whose
-
-transition probability matrix P is symmetric. P is called the proposition matrix, and its elements,
-P071"), Satisfy
-pa, 1) 2 0. pm‘) = pm). 2,-pm‘) = 1 vi e S (5-4)
-
-Using P, we will derive a new transition probability matrix K which satisﬁes
-
-Z vr(z')k(='.:') = arm (5.5)
-
-I
-
-For 1' aé j deﬁne
-
-kw) = p<-yum")/«<¢> it nu)/1r(='> <1 (5.6)
-
-P(i,J') if 7r(J')/7r(i) Z1
-
-% 33.txt
-33
-
-For 2' = j deﬁne
-
-kw) = pm) (1—  p(-3;"), where J. = {j = 2‘ 7‘ an nu)/«(='> < 1}
-
-Notice that 1r(i) > 0 implies k(i,j) _>_ 0 and
-
-;k<«',j> = pow)  { (1 —  pm) +  +]§§p<«",:')
-= P(is'5) + 2 P(i1.1') + Z P(iy.7') (5-7)
-jel.» jug
-
-Equation (5.7) shows that, for each i, 2, k(z', j) = 1, which fact conﬁrms that K is a stochastic
-
-matrix.
-
-Next we check that K satisﬁes the detailed balance condition. First suppose that 1r(z') =
-
-1r(j). Then (5.4), (5.6), and symmetry of P imply
-k(i,J') = p(i2.7') = p(jvi) = k(j1i)
-
-and, since 1r(i) = 1r(j), the detailed balance condition, 1r(i)k(i,j) = 7r(j)k(i,j), clearly holds. Now
-
-suppose that 1r(i) > 1r(j). Then (5.4) and (5.6) imply k(j,i) =p(j, 1') and
-k(i,J') = P(iyJ')7r(J')/7r(i) = P(J3i)7r(J')/7r(i) = k(.1'» i)7r(J')/1r(i)
-
-That is, 1r(i)lc(i,j) = 1r(j)k(i,j), as we set out to prove.
-To complete the justification of the Metropolis algorithm, we note that the detailed balance
-condition, 7r(i)k(i, j) = 1r(j)k(i, j), implies that 1r is indeed the stationary distribution for a Markov
-
-chain with transition probability matrix K. To see this notice that
-
-27r(i)k(i..7') = Zﬂijlkiiajl = 1r(J') Z/€(i,J‘) = 7r(J')
-
-I i
-We now summarize the tasks performed when implementing the Metropolis algorithm.
-
-First, pick as a proposition matrix any symmetric stochastic matrix and denote it by P. Begin
-
-% 34.txt
-34
-
-the simulation of the Markov chain in state i, and then take one step of the proposition Markov
-chain to arrive in state j. That is, choose j according to the mass function p(z', -). Next, compute
-1r(j)/1r(i). If 7r(j)/1r(i) 2 1, accept j as the new state. If1r(j)/1r(i') < 1, then with probability
-1r(j)/7r(z') accept j as the new state; otherwise, with probability 1 — 1r(j)/1r(i), take i as the new
-state. Performing each of these tasks produces one step of the Markov chain represented by the
-
-transition probability matrix K.
-
-5.3 The Ising Model
-
-Returning to the Ising model considered in section 5.1, we consider how the Metropolis
-algorithm is applied to such a mode]. Again, start the simulated chain in state X0 = 1' and let
-
-X1 = j be a state chosen according to p(i, -), where p(i, j) = p(j,i). Next, compute
-
-7"(‘7j)/7"(‘7i) = eXP{-ﬁlH(<7j) - H(0i)l}
-
-Suppose the energy H (Uj) of the new state satisﬁes H (U,-) 3 H (05). Then 1r(a_,-)/1r(a,-) Z 1 and we
-
-move to state j. On the other hand, if H(aj) > H(a;), then 1r(aj)/7r(a,-) < 1 and we take
-
-j with probability e-M"
-
-X1 (5.8)
-
-i with probability 1 — e-W’
-where AH E H (0,) — H (:75) is the change in energy from state j to state 1'.
-There are many ways to choose a proposition matrix P. The simplest derives from the
-so called Glauber dynamics, which proceeds as follows: choose one of the 11. sites at random with
-probability 1/n — let .9 denote the chosen site — and change the position of this site; i.e., multiply
-
-its value by -1. The resulting state j is the permutation of -1’s and +1’s given by
-
--0,-(k) if]: = .9
-a,~(Ic)= (k=1,...,d)
-
-a,-(Ic) if]: 95 s
-
-% 35.txt
-35
-
-The procedure we employ below involves a slight variation: choose site .9 at random uniformly
-from the 11 possible sites. However, instead of changing the value of the chosen site with absolute
-certainty, as in Glauber dynamics, we “ﬂip the switch” with probability 1/2. Therefore, when k yé .9
-
-we have 0',-(k) = a.~(k), and when k = 3,
-
-—a,~(s) with probability 1/2
-01(3) =
-
-a,-(3) with probability 1/2
-The advantage of this procedure is that the proposition matrix, before altering it with the Metropo-
-lis algorithm, is a bit more interesting than that produced by traditional Glauber dynamics. For,
-it has 1/2 along the main diagonal and, thus, is aperiodic and converges to a stationary distribu-
-tion. Glauber dynamics has period 2, so its proposition matrix does not satisfy the hypotheses of
-
-our convergence theorem (Theorem 2.1.2). Keep in mind, however, that any proposition matrix
-
-modiﬁed by the Metropolis algorithm satisﬁes all of our hypotheses.
-
-5.4 Computations
-
-5.4.1 Computer Programs
-
-Constructing a computer program to simulate realizations of the Markov chain described
-above (i.e. with transition matrix K) is straight forward, and we do so using Matlab and the
-program ising.m which appears in the Appendix.
-
-When performing simulations in the present context, there are a few important aspects
-to consider. First, our program begins with a “hot start", which means that the initial state is
-picked by assigning each site to :l:1 with probability 1/2. Perhaps we should instead have picked
-our initial state from the stationary distribution. On the contrary, this would defeat the whole
-
-purpose of our analysis. Recall that the method has been developed to deal with cases for which
-
-% 36.txt
-36
-
-sampl from the stationary distribution are not quickly attainable. Furthermore, we are trying to
-ﬁgure out how long it will take for observations from the simulated chain to represent samples from
-the stationary distribution. Second, the method we have developed depends solely on covariances
-of our observable which we estimate from the data. produced by the simulations. However, such
-estimates require that the data come from the stationary distribution. Therefore, we must discard a
-number of observations before using the data to estimate covariances. The number of observations
-to discard depends on how long must we wait until we have data from the stationary distribution,
-and we are back where we started — almost. When employing our procedure, we can discard a
-conservative (very large) number of observations and perform our analysis, deriving a bound on
-the convergence time. Then, in all future studies, we will know how much data should be discarded
-before samples come (approximately) from the stationary distribution.
-
-In considering what observable to use on the state space described above, perhaps the
-most obvious is simply the number of nodes in the on position (i.e. the number of +l’s). Recalling
-that 0, (k) = i1 denotes the position (on or off) of the kth node when the system is in state j, this
-
-observable is deﬁned by,
-«M = zxca.-ac) = +1). 1 = 1.---.ISl
-I:
-
-where X is the indicator function. The program ising.m simulates the Markov chain, computes the
-values of this observable (as well as the number of 1’s squared) and writes it to a ﬁle called phi . dat.
-The program then proceeds to compute the ﬁrst few Lanczos coefficients using covariances produced
-by Matlab, so that we can check them against the rults we get from the 1anczos.c program,
-which we now describe.
-
-The lanczos .c program (listed in the Appendix along with its dependencies) implements
-the recursive relations described in section 4.2 to compute the Lanczos coefﬁcients by the new
-
-method. It accepts input from the user specifying any observable, any number of iterations, and any
-
-% 37.txt
-37
-
-TABLE 1: ESTIMATES or /\,m(K) = 0.998746 USING ¢1
-
-sim. # /32
-1 0.995107 0.006960 0.995220 0.600243
-2 0.995099 0.005891 0.994863 0.477393
-3 0.994862 0.009671 0.994925 N/ A
-4 0.995321 0.009319 0.995654 N/A
-5 0.994787 0.004872 0.405446 0.994742
-
-number of desired Lauczos coeﬁicients. That is, the observable could have come from a simulation
-of any reversible Markov chain and not just our special Metropolis chain. The output of this
-program is a T matrix containing the maximum number of Lauczos coeﬂicients (up to the number
-requested by the user) before an approximate invariant subspace was reached. The eigenvalues of
-the nonzero part of this matrix, in accordance with the foregoing theory, should well approximate
-
-eigenvalues of the Markov chain’s stochastic matrix.
-
-5.4.2 Results
-
-Table 1 shows the results of 5 simulations each of 100,000 iterations (discarding the ﬁrst
-5,000 observations) for the one dimensional Ising lattice with 10 nodes and [7 = 1. Displayed is
-the estimate of the ﬁrst Lauczos coeﬂicient a1 (which is the eigenvalue estimate after one step; i.e.
-).m,x(T1)), along with its error bound ﬁl. Appearing in the third column is the largest eigenvalue of
-the 2 x 2 matrix T2, A0 (T2), and ﬁg appears in the fourth column when it is was computable. When
-it was not computable, it was probably because an approximate invariant subspace was reached
-at the ﬁrst step, so ﬁg involved expressions which were close to machine epsilon and, thus, could
-not be computed accurately. Table 2 shows the same information when using a second observable,
-¢2 E ¢§ (i.e. the number of +1’s squared).
-
-The true value of the subdominant eigenvalue for this problem is /\max(K) = 0.998746,
-
-and the next largest is A2(K) = 0.993303. These were computed directly from the transition matrix
-
-% 38.txt
-38
-
-TABLE 2: ESTIMATES OF >.,,,,,.,,,(K) = 0.998746 USING z/>2
-
-sim. # B2
-1 0.994262 0.006142 0.994222 N / A
-2 0.994333 0.002974 0.994332 N / A
-3 0.994307 0.008892 0.994374 0.726320
-4 0.994436 0.012884 0.994685 0.361404
-5 0.993789 0.006909 0.993820 0.788710
-
-using the Matlab program tpm.m listed in the Appendix. They should be very accurate, and it
-takes the Matlab routine eig() just over 20 minutes of cpu time on a Sun Ultra to ﬁnd all the
-eigenvalues of K.
-
-As shown by Theorem 4.1.1, the [Ts provide conservative error bounds on the eigenvalue
-estimates from the previous Lanczos step. Unfortunately, for this example, ,6 appears to be too
-conservative as it does not even allow us to bound our estimates away from 1. Furthermore, it
-seems that the estimates generated by the method above fall systematically below the true value of
-the subdominant eigenvalue. In every simulation, an approximate invariant subspace (to machine
-tolerance) was reached after computing at most 2 pairs of Lanczos coefficients. For ¢1 we see that
-it was reached after computing only one pair on simulations 3 and 4, similarly for ¢2 on simulations
-1 and 2. These observations indicate that the chosen observable is close to the slowest mode (or
-eigenfunction) of the process. That is, we should observe fast convergence on an approximate
-invariant space. However, the space found might contain only the second slowest mode of the
-process and, in that case, our eigenvalue estimates would be closer to the third largest eigenvalue,
-instead of the second largest as desired.‘
-
-The matrix T was produced by the program lanczos . c and its eigenvalues were computed
-by Matlab, both operations taking a few seconds. Granted, comparing a few seconds to the 20
-
-minutes it takes Matlab is not justiﬁed since a simple call to the routine eig() is far from the
-
-‘Recall that we explained above why we will never converge to the eigenspace corresponding to the largest eigen-
-value (which is always 1). That would require a (nearly) constant observable function.
-
-% 39.txt
-39
-
-most eﬂicient traditional method for computing only a few eigenvalues of K. However, recall that
-the primary motivation for the new method are those examples where the matrix K is so large
-that we can’t store even a single vector in fast memory. Once we start swapping data to and from
-cheap memory, the time required by traditional algorithms can become years, even lifetimes. Such
-examples are not exceptional. Consider, for instance, the 1000 node problem which produces a.
-
-2“’°° x 21°°° transition matrix.
-
-% 40.txt
-40
-
 Chapter 6
@@
 but also the cross-covariances C, (¢.~(n), ¢j(n + m)) (deﬁned in (2.6)).
-
-% 42.txt
-Appendix: computer programs
-
-Z Hatlab code ising.m
-
-Z
-Z Written by William J. De Meo on 12/15/97
-
-2 last modified 1/10/98
-
-Z Inputs:
-
-2 d = number of nodes of the ising lattice (e.g. d=100)
-Z n = number of iterations (e.g. n=1000)
-
-Z beta = Annealing schedule (e.g. beta = 2,
-
-1 when beta -> 0 will always go to nee state
-Z vhen beta -> infty will never go to state of higher energy)
-Z
-
-X = zeros(d,1);
-Energy = zeros(n,1);
-
-Ave=0;
-
-X Initialize using hot start
-for i=1:d,
-
-if rand < .5
-
-x(i) = -1;
-
-else
-
-X(i) = 1;
-
-end
-end
-
-Z Initialize at state of high energy (not used)
-
-if 0
-X(1:2:d-1)=1;
-X(2:2:d)=—1;
-and
-
-Z Initialize energy
-
-H=0;
-
-for i=1:(d-1)
-
-H = H - X(i)*X(i+1);
-end
-
-2 Initialize observable
-phi = zeros(n+1,1);
-
-for i=1:d
-if X(i) == 1
-phi(1)= phi(1)+1;
-end
-
-end
-
-42
-
-% 43.txt
-43
-
-for j=1:n
-
-site=0;
-while site==
-
-site = round(rand#d);
-end
-Z Compute new energy (with new X(site) = —x(site)):
-if site ==
-
-Hnew = H + 2:x(1)~x(2);
-elseif site == d
-
-Hnew = H + 2*X(d)tX(d-1);
-else
-
-Hnew = H + 2*X(site)*(X(site-1) + X(site+1));
-end
-
-Z Change to new state in two ways:
-X with probability 1/2, flip the switch
-if rand < .5
-if (Hnew <= H) I (rand < exp(-beta*(Hnew-H)))
-x(site) = -x(site);
-H = Hnew;
-Z Compute new value of observable
-if X(site) -= 1
-phi(j+1) = phi(j) + 1;
-
-else
-phi(j+1) = phi(j) — 1;
-end
-else phi(j+1)=phi(j);
-end
-else phi(j+1)=phi(j);
-end
-Energy(j) = H;
-end
-p1ot(Energy)
-
-% A second observable
-phi2 = zeros(n+1,1);
-for i=1:n+1
-
-phi2(i) = phi(i)*phi(i);
-end
-
-Z Write first observable to phi.dat:
-
-fid = fopen(’/quean.a/masters/deme5025/research/nyuthesis/programs/c/phi.dat’,’w’);
-fprintf(fid,’%f\n’,phi);
-fc1ose(fid);
-
-Z Write second observable to phi2.dat:
-fid = fopen(’/quean.a/masters/deme5025/research/nyuthesis/programs/c/phi2.dat’,’w’);
-
-% 44.txt
-fprintf(fid,’%f\n’,phi2);
-fc1ose(fid);
-
-1 Compute a few Lanczos coefficients by new method
-START=1001;
-
-V1=zeros(2);
-
-C1=zeros(2);
-
-C2=zeros(2);
-
-C3=zeros(2);
-
-V1 = cov(phi(START:n+1),phi(START:n+1));
-
-V1 = V1(1,1)5
-
-C1 = cov(phi(START:n),phi(START+1:n+1));
-C1 = C1(1,2);
-
-C2 = cov(phi(START:n-1),phi(START+2:n+1));
-C2 = C2(1,2);
-
-C3 = cov(phi(START:n-2),phi(START+3:n+1));
-
-C3 = C3(1,2);
-
-a1pha1 = C1/V1;
-
-betai = sqrt(C2/V1 ' (C1/V1)‘2);
-
-a1pha2 = (1/(beta1’2))*(C3/V1 - 2*a1pha1*(C2/V1) + a1pha1‘3);
-
-T1 = Ealphal betal; betal a1pha2]
-
-v1 = cov(phi2(START:n+1),phi2(START:n+1));
-v1 = v1(1,1);
-
-c1 = cov(phi2(START:n),phi2(START+1:n+1));
-c1 = c1(1,2);
-
-c2 = cov(phi2(START:n—1),phi2(START+2:n+1));
-c2 = c2(1,2);
-
-c3 = cov(phi2(START:n-2),phi2(START+3:n+1));
-c3 = c3(1,2);
-
-alphal = C1/V1;
-betal = sqrt(C2/V1 - (C1/V1)‘2);
-a1pha2 = (1/(beta1‘2))#(C3/V1 - 2ta1pha1*(C2/V1) + a1pha1”3);
-
-T2 = [a1pha1 betai; betal a1pha2]
-
-fid = fopen(’Tmats.dat’,’v’);
-fprintf(fid,’%f\n',T1);
-fprintf(fid,’%f\n’,T2);
-fc1ose(fid);
-
-ZZZ end ising.m
-
-44
-% 45.txt
-\% Matlab code tpm.m for generating transition probability matrix
-Z and computing its eigenvalues directly
-
-1
-
-Z Written by Hilliam J. De Heo on 1/10/98
-
-1
-
-Z Inputs:
-
-Z d = number of nodes of the ising lattice (e.g. d=10)
-
-Z beta = Annealing schedule (e.g. beta - 2,
-
-1 when beta -> 0 will always go to new state
-1 when beta -> infty will never go to state of higher energy)
-2
-
-disp(’bui1ding proposition matrix...’)
-
-states = 2‘d;
-
-A = 0;
-
-for i=0:(d-1)
-
-E=eye(2”i);
-
-A = [A E;E A];
-
-end
-
-B = .5#eye(states);
-
-A = B + .5-(1/d)*A;
-
-2 our modified Glauber dynamics requires the B and the .5¥(1/d)
-
-disp(’...done’)
-
-1 display pattern of nonzero entries
-1 spy(A)
-
-Z check that all row sums are 1
-check=O;
-for i=1:states
-
-check = check + (not(sum(A(i,1:states))<.99));
-end
-1 if not all rows sum to 1, print check = (# of rows with sum=1)
-if not(check==states)
-
-check
-
-error(’Row sums are not all 1’)
-end
-
-Z construct matrix of states
-E = zeros(d,states);
-f1ip=-1;
-for i=1:d
-for j=1:2‘(i-1):states
-flip = -1*f1ip;
-for k=0:2“(i-1)
-
-45
-
-% 46.txt
-E(i,j+k) = flip;
-and
-end
-end
-E=E’;
-
-Z display first 64 states
-Z E(1:64,:)
-
-Z compute energy of each state
-H = zeros(states,1);
-for i=1:states
-
-for j=1:d-1
-H(i) = H(i) - E(i,j)tE(i,j+1);
-end
-end
-
-disp(’building Hetropolis transition matrix...')
-
-for i=1:states
-for j=i+1:states
-if not(A(i,j)==0)
-if(H(j)>H(i))
-alt = A(i,j)t(1-exp(-beta*(H(j)-H(i))));
-A(i,i) I A(i,i)+a1t;
-A(i,j) = A(i,j)-alt;
-elseif(H(j)<H(i))
-alt = A(j,i)t(1-exp(—beta#(H(i)-H(j))));
-A(j,j) = A(j,j)+a1t:
-A(j,i) = A(j,i)-alt;
-end
-end
-end
-end
-
-disp(’...done’)
-
-check=0;
-for i=1:states
-
-check = check + (not(sum(A(i,1:states))<.99));
-end
-
-Z if not all rows sum to 1, print check = (# of rows with aum=1)
-
-if not(check==states)
-
-check
-
-error(’Row sums are not all 1’)
-end
-
-46
-
-% 47.txt
-47
-
-disp(’computing eigenvalues of tpm...’)
-cput = cputime;
-
-evals = eig(A);
-
-ecput = tputime - cput;
-
-disp(’the CPU time (in secs) for computing eigenvalues of tpm: ’)
-
-ecput
-ZZZ end tpm.m
-
-/¥#*##¥###*##*#l####¥#‘###$ﬂ###ﬁi##*U3*#**#*##**##¥***ll¥#*##¥ii
-
-* 1anczos.c main program for computing Lanczos coefficients *
-3 #
-* Created by William J. De Heo #
-t on 1/7/98 *
-# Last modified 1/10/98 #
-
-*¥#######**3It¥*#¥¥¥#**##**¥##¥#####**####*#*###*##**###¥##*##*/
-
-#inc1ude <stdlib.h>
-#include <stdio.h>
-#include <math.h>
-#inc1ude "prototypes.h"
-
-/# Machine constants on dino (Sun Ultra 1 at Courant) t/
-#define HACHEPS 1.15828708535533604981e-16
-
-#define SQRTEPS 1.07623746699106147048e-08
-
-ﬂdefine HAX_NAH 100
-
-void read_name(char *);
-
-/* Prototypes from the file functions.c ¥/
-
-double a1pha(int j);
-
-double betasq(int j);
-
-double form1(int j, int k);
-
-double form2(int j, int k);
-
-double moment(doub1e ‘data, long n, double tave, double tvar, double tcov, long k);
-
-/# external variables to be used by functions */
-double *phi,tcov, tvar, tave;
-
-% 48.txt
-main()
-
-{
-
-char tfilename;
-
-FILE *ofp;
-
-double temp=O, *T;
-
-long i, j, nrou, nlanc, START;
-int f1ag=O;
-
-filename = cma11oc(HAx_NAHE);
-
-printf("\nName of file containing observed values: “);
-
-read_name(fi1ename);
-
-printf(“\nTota1 number of observations in file (iterations):
-
-scanf(“Zu”,&nrov);
-
-printf(“\nNumber of leading observations to discard: “);
-
-scanf("Zu",&START);
-
-printf("\nNumber of Lanczos coefficients desired: “);
-
-scanf("%u",&n1anc);
-
-vhi1e(2*n1anc > nrow—START)
-
-{
-printf("\nNot enough data for that many coefficients.\n“);
-printf("\nEnter a smaller number of Lanczos coefficients: “);
-scanf("Xu“,&n1anc);
-
-}
-
-II);
-
-phi = dma11oc(nrow);
-cov=dma11oc(2#n1anc);
-var=dma11oc(1);
-ave=dma11oc(1);
-
-T = dma11oc(n1anc#n1anc); for(i=0;i<n1anc;i++) T[i]=(doub1e)0;
-
-/* observable phi is stored contiguously column-wise (e.g. by HATLAB) t/
-mat1abread(phi, nrow, 1,fi1ename);
-
-/* send the observable, offset by START */
-moment(phi+START,nrou-START,ave, var, cov,2*n1anc-1);
-
-/* First column of T */
-
-T[0] = a1pha(1);
-if((temp=betasq(1))>HACHEPS*10)
-{
-
-T[1]=sqrt(temp);
-
-/* General column of T */
-
-J'=2:
-
-for(i=1;i<n1anc-1, f1ag==O;)
-{
-
-T[itn1anc+i-1]=sqrt(betasq(i));
-
-48
-
-main()
-
-{
-
-char tfilename;
-
-FILE tofp;
-
-double temp=0, #T;
-
-long i, j, nrow, nlanc, START;
-int f1ag=0;
-
-filename = cma11oc(HAx_NAH);
-
-princf("\nName of file containing observed values: “);
-read_name(fi1ename);
-
-printf("\nTota1 number of observations in file (iterations): “);
-scanf("2u“,&nrow);
-
-printf("\nNumber of leading observations to discard: “);
-scanf("Zu",&START);
-
-printf("\nNumber of Lanczos coefficients desired: “);
-scanf("%u“,&n1anc);
-
-whi1e(2*n1anc > nrou-START)
-
-{
-printf("\nNot enough data for that many coefficients.\n");
-
-printf("\nEnter a smaller number of Lanczos coefficients: “);
-scanf("Zu",&n1anc);
-}
-
-phi = dma11oc(urow);
-cov=dma1loc(2*n1anc);
-var=dma11oc(1);
-ave=dma11oc(1);
-
-T = dma11oc(n1anc*n1anc); for(i=0;i<n1anc;i++) T[i]=(doub1e)O;
-
-/* observable phi is stored contiguously column-wise (e.g. by HATLAB) */
-mat1abread(phi, nrov, 1,fi1ename);
-
-/* send the observable, offset by START #/
-moment(phi+START,nrov-START,ave, var, cov,2*n1anc-1);
-
-/# First column of T */
-
-T[0] = a1pha(1);
-if((temp=betasq(1))>HACHEPS#10)
-{
-
-T[1]=sqrt(temp);
-
-/* General column of T */
-
-J‘=2:
-
-for(i-1;i<n1anc-1, f1ag==0;)
-{
-
-T[i*nlanc+i-1]=sqrt(betasq(i));
-
-48
-
-% 49.txt
-T[i*n1anc+i] = a1pha(i+1);
-if((temp = betasq(i+1))>HACmEPS#10)
-{
-
-T[i#n1anc+i+1] = sqrt(temp);
-
-i++;
-}
-else
-
-f1ag=1;
-
-}
-
-/t Last column */
-if(f1ag!=1 it ((temp= betasq(n1anc-1))>HACHEPSt10))
-
-{
-T[n1anc#n1anc-2] = sqrt(temp);
-T[n1anctn1anc-1] = a1pha(n1anc);
-printf("\nbeta(1) = %1f\nbeta(Zd) = Zlf (last beta)",
-T[1],(nlanc-1),T[nlanc*n1anc-2]);
-}
-else
-{
-printf("\nApproximate invariant space reached at step %d.",i);
-printf("\nbeta(1) = Z1f\nbeta(Zd) = Zlf (last accurate beta)“.
-T[1],i,T[itn1anc+i—1]);
-printf("\nbeta(Zd)“2: Zlf (first spurious resu1t)“,i+1,temp);
-}
-
-printf("\nThe matrix T is: \n");
-matprint(T,n1anc,n1anc);
-
-}
-
-else
-
-{
-
-printf("\nmain(): Approximate invariant space reached at first step.“);
-
-49
-
-printf("\na1pha(1) - Z1f\nbeta(1)”2: Zlf (first spurious result)“, T[O], temp);
-
-}
-ofp = fopen("Tmat.m","v“);
-check(ofp);
-mat1aburite(T,n1anc,n1anc,ofp);
-fc1ose(ofp);
-}
-void read_name(char tname)
-{
-int c, i = 0;
-while ((c = getchar()) != EOF it c != ’ ’ it c !- ’\n’)
-name[i++] = c;
-name[i] = ’\0’;
-}
-
-/#t* end 1anczos.c *#*/
-
-% 50.txt
-50
-
-/llnkttttttttinlntint*1!!!##1##!ttitilltit###t¥#tt¥ttItt####t#ttII=##mlltttt
-
-* functions.c functions required by 1anczos.c t
-# t
-* Created by Hilliam J. De Meo #
-* on 1/7/98 *
-t Last modified 1/10/98 :
-
-¥¥##¥$¥##¥¥#$¥¥¥###¥¥$*I#****#**####*######¥#¥#¥#$#¥¥#l¥**#****/
-
-#inc1ude <math.h>
-
-#define START 1000
-
-#define ITER 10000
-
-/t Machine constants on dino (Sun Ultra 1 at Courant) */
-
-#define HACHEPS 1.15828708535533604981e-16
-
-#define SQRTEPS 1.07623746699106147048e-O8
-
-double a1pha(int j);
-
-double betasq(int j);
-
-double form1(int j, int k);
-
-double form2(int j, int k);
-
-double moment(doub1e tdata, long n, double tave, double tvar, double tcov, int k);
-
-/t external variables to be used by functions */
-double *phi,tcov, *var, tave;
-
-double alpha(int j)
-{
-
-/* alpha is never called with j < 1 */
-if(j==1)
-return
-form1(1,1);
-else if(j>1)
-return
-((doub1e)1/betasq(j-1))*
-(form1(j-1,3) — pou(a1pha(j-1),3)
-- 2 * (alpha(j-1)#betasq(j-1) + sqrt(betasq(j-2))#form2(j-1,2))
-+ betasq(j-2));
-}
-
-double betasq(int j)
-
-{
-if(j==0)
-return (doub1e)O;
-else if(j>0)
-return (form1(j,2) - pow(a1pha(j),2) - betasq(j-1));
-}
-
-double form1(int j, int k)
-{
-
-% 51.txt
-51
-
-double form14,alphal,alpha1sq,form12,form13;
-if(j==0)
-
-return (doub1e)0;
-else if(j==1)
-{
-
-/# printf("\nvar = %1f, cov(%d) = %1f \n“,*var,k,cov[k]);t/
-
-return (cov[k])/(*var); /* the only real value */
-}
-else if(j>1)
-{
-
-return ((double)1/betasq(j-1))
-
-* ( form1(j-1,k+2)
-+ pow(a1pha(j-1),2)*form1(j-1,k)
-+ 2t(a1pha(j-1) t sqrt(betasq(j-2)) * form2(j-1,k)
-- a1pha(j-1)*form1(j-1,k+1)
-- sqrt(betasq(j—2))#form2(j-1,k+1) )
-+ betasq(j-2)*form1(j-2,k)):
-}
-}
-
-double form2(int j, int k)
-{
-/* form2 is never called with j < 1 */
-if(j==1)
-return (doub1e)0;
-else if(j>1)
-return
-(pov(betasq(j-1),—.5)) 3
-(form1(j-1,k+1) - a1pha(j-1)*form1(j-1,k)
-- sqrt(betasq(j-2)) # form2(j—1,k));
-}
-
-/* moment() function for computing var and cov(k)
-arguments:
-data = a nxi array of doubles
-n = length of data[]
-ave =(on exit)= the average of data[]
-var =(on exit)= the variance of data[]
-cov =(on exit)= the covariance of data[i] and data[i+j] for i-1,...,k
-k = the max lag for cov above
-*/
-double moment(double tdata, long n, double tave, double tvar, double tcov, int k)
-{
-/* Centered about data[0] algorithm: #/
-
-long i,j;
-double avel, ave2;
-
-% 52.txt
-52
-
-*ave=O;tvar=0; ave1=ave2=O;
-for(j=0;j<=k;j++)
-cov[j]=0;
-
-for(i=1;i<n;i++)
-{
-rave += (data[i] - data[0]);
-tvar += (data[i] - data[0])#(data[i] - data[0]);
-}
-tvar /= (doub1e)(n-1);
-Ivar -= (((*ave)/(doub1e)n) * ‘(*ave)/(doub1e)(n-1)));
-/# *ave = ((*ave)/(doub1e)n) + data[0]; <- that’s the true average (not needed)*/
-
-for(j=0;j<=k;j++)
-{
-
-for(i=0;i<(n-j);i++)
-
-{
-
-avel += (data[i] - data[0]);
-ave2 += (data[i+j] - data[0]);
-cov[j] += (data[i] - data[0])*(data[i+j] — data[0]);
-
-}
-
-avel/=(doub1e)(n-j); ave2/=(doub1e)(n-j);
-
-cov[j] = ((cov[j] - (double)(n-j)tave1¢ave2)/(double)(n-j-1));
-}
-
-}
-
-/trt end funccions.c ###/

### 2013-10-18 tex/InvariantSubspaces.tex
initial commit of tex source for nyu thesis style setup

--- 2013-10-18 tex/MSThesis.tex
@@
-%\documentclass[11pt]{amsart}
-\documentclass{au}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% For completion by editorial office:
- \presentedby{\dots}
- \received{\dots}{\dots}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\usepackage[colorlinks=true,urlcolor=black,linkcolor=black,citecolor=black]{hyperref}
-
-\usepackage{amsmath,amssymb,url,mathrsfs}
-\usepackage{enumerate}
-\usepackage{tikz}
-\usetikzlibrary{calc}
-\usepackage{scalefnt}
-\numberwithin{equation}{section}
-
-\theoremstyle{plain}
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{proposition}[theorem]{Proposition}
-\newtheorem{corollary}[theorem]{Corollary}
-
-\theoremstyle{definition}
-\newtheorem{definition}[theorem]{Definition}
-\newtheorem{notation}[theorem]{Notation}
-\newtheorem{remark}[theorem]{Remark}
-\newtheorem{example}[theorem]{Example}
-\newtheorem*{remarks}{Remarks}
-
-\newcommand{\todo}[1]{~\\[5pt]{\bf TODO(wjd):} #1\\[5pt]}
-%\newcommand{\suchthat}{\ensuremath{:}}    % (old notation)
-\newcommand{\suchthat}{\ensuremath{\mid}}  % (new notation)
-%\newcommand{\defeq}{\ensuremath{:=}}      % (old notation)
-\newcommand{\defeq}{\ensuremath{=}}        % (new notation)
-\newcommand{\betai}{\ensuremath{\beta^i}}  % \beta^{\bB_i}   % (old notation)
-\newcommand{\betaj}{\ensuremath{\beta^j}}  % \beta^{\bB_j}
-\newcommand{\betak}{\ensuremath{\beta^k}}  % \beta^{\bB_k} or \beta^{\bB_{k}}
-\newcommand{\betal}{\ensuremath{\beta^\ell}}  %\beta^{\bB_\ell} or \beta^{\bB_{\ell}}
-\newcommand{\uacalc}{{\small UAC}alc}
-\newcommand{\Tuma}{T\r{u}ma}
-\newcommand{\Palfy}{P\'alfy}
-\newcommand{\Pudlak}{Pudl\'ak}
-\newcommand{\PP}{P\'alfy-Pudl\'ak}
-\newcommand{\PAP}{P\'alfy\ and Pudl\'ak}
-\newcommand{\<}{\ensuremath{\langle}}
-\renewcommand{\>}{\ensuremath{\rangle}}
-\newcommand{\beps}{\ensuremath{\boldsymbol{\varepsilon}}}
-\newcommand{\bN}{\ensuremath{\mathbf{N}}}
-\newcommand{\bA}{\ensuremath{\mathbf{A}}}
-\newcommand{\bB}{\ensuremath{\mathbf{B}}}
-\newcommand{\bBi}{\ensuremath{\mathbf{B}_i}}
-\newcommand{\sB}{\ensuremath{\mathcal{B}}}
-\newcommand{\sC}{\ensuremath{\mathcal{C}}}
-\newcommand{\sS}{\ensuremath{\mathscr{S}}}
-\newcommand{\sT}{\ensuremath{\mathscr{T}}}
-\newcommand{\sI}{\ensuremath{\mathcal{I}}}
-\newcommand{\sE}{\ensuremath{\mathcal{E}}}
-\newcommand{\sO}{\ensuremath{\mathcal{O}}}
-%\newcommand{\id}{\ensuremath{\mathrm{id}}}
-\DeclareMathOperator{\id}{id}
-%\newcommand{\Eq}{\ensuremath{\mathrm{Eq}}}
-\DeclareMathOperator{\Eq}{Eq}
-%\newcommand{\Cg}{\ensuremath{\mathrm{Cg}}}
-\DeclareMathOperator{\Cg}{Cg}
-%\newcommand{\Con}{\ensuremath{\mathrm{Con\,}}}
-\DeclareMathOperator{\Con}{Con}
-%\newcommand{\Sub}{\ensuremath{\mathrm{Sub}}}
-\DeclareMathOperator{\Sub}{Sub}
-%\newcommand{\Pol}{\ensuremath{\mathrm{Pol}}}
-\DeclareMathOperator{\Pol}{Pol}
-%\newcommand{\Clo}{\ensuremath{\mathrm{Clo}}}
-\DeclareMathOperator{\Clo}{Clo}
-\newcommand{\ps}[1]{\ensuremath{^{(#1)}}}
-\newcommand{\piB}{\ensuremath{\pi_B}}
-\newcommand{\hpsi}{\ensuremath{\widehat{\psi}}}
-\newcommand{\htheta}{\ensuremath{\widehat{\theta}}}
-\newcommand{\supi}{\ensuremath{^{i}}}
-\newcommand{\supj}{\ensuremath{^{j}}}
-\renewcommand{\leq}{\ensuremath{\leqslant}}
-\renewcommand{\nleq}{\ensuremath{\nleqslant}}
-\renewcommand{\geq}{\ensuremath{\geqslant}}
-\newcommand{\meet}{\ensuremath{\wedge}}
-\newcommand{\join}{\ensuremath{\vee}}
-\newcommand{\Meet}{\ensuremath{\bigwedge}}
-\renewcommand{\Join}{\ensuremath{\bigvee}}
-\newcommand{\resB}{\ensuremath{|_{_B}}}
-\newcommand{\resBi}{\ensuremath{|_{_{B_i}}}}
-\newcommand{\eps}{\ensuremath{\varepsilon}}
-\newcommand{\hatmap}{\ensuremath{\widehat{\phantom{x}}}}
-\newcommand{\one}{\ensuremath{\mathbf{1}}}
-\newcommand{\two}{\ensuremath{\mathbf{2}}}
-\newcommand{\three}{\ensuremath{\mathbf{3}}}
-\newcommand{\tbeta}{\ensuremath{\widetilde{\beta}}}
-\newcommand{\ttheta}{\ensuremath{\widetilde{\theta}}}
-\newcommand{\hbeta}{\ensuremath{\widehat{\beta}}}
-\newcommand{\cick}{\ensuremath{\sC_i}}
-%\newcommand{\CE}{\ensuremath{\sC_r^{\sE}}}
-\newcommand{\CE}{\ensuremath{\sB_r^n}}
-\newcommand{\CO}{\ensuremath{\sC_r^{\sO}}}
-\newcommand{\CICK}{\ensuremath{c_i/\beta \cup c^{K+1}_i/\beta^{\bB_{K+1}}}}
-\newcommand{\GAP}{{\small GAP}}
-\newcommand{\dotsize}{.8pt}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{document}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%  FRONT MATTER                                                 %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\title[Approximating Eigenvalues of Large Stochastic Matrices]{A Lanczos Procedure for Approximating Eigenvalues of Large Stochastic Matrices}
-
-%% First author (Note: The order of the items here is important!)
-\author{William DeMeo}
-\email{williamdemeo@gmail.com}
-\urladdr{http://www.math.sc.edu/~demeow}
-\address{Department of Mathematics\\
-University of South Carolina\\Columbia 29208\\USA}
-
-\thanks{The author wishes to thank a number of people who contributed to this
-  work.  In particular, the primary referee made many insightful recommendations
-  which led vast improvements, not only in the overall exposition, but also in
-  the specific algebraic constructions, theorem statements, and proofs.
-  This article is dedicated to Ralph Freese and Bill Lampe, who declined to be
-  named as co-authors despite their substantial contributions to the project.}
-
-%% Dedication (Optional)
-\dedicatory{This article is dedicated to Ralph Freese and Bill
-  Lampe.}
-
-%% AMS subject classification; see http://www.ams.org/msc
-%% Only one Primary. Possibly several Secondary.
-\subjclass[2010]{Primary: 08A30; Secondary: 08A60, 06B10.} %, 06B15.}
-
-%% Keywords and phrases
-\keywords{congruence lattice, finite algebra, finite lattice
-  representations}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%  MAIN MATTER                                                  %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{abstract}
-The rate at which a Markov chain converges to a given probability distribution
-has long been an active area of research. Well known bounds on this rate of
-convergence involve the subdominant eigenvalue of the chain’s underlying
-transition probability matrix. However, many transition probability matrices are
-so large that we are unable to store even a vector of the matrix in fast
-computer memory. Thus, traditional methods for approximating eigenvalues are
-rendered useless.
-
-In this paper we demonstrate that, if the Markov chain is reversible, and we
-understand the structure of the chain, we can derive the coefficients of the
-traditional Lanczos algorithm without storing a single vector. We do this by
-considering the variational properties of observables on the chain’s state
-space. In the process we present the classical theory which relates the
-information contained in the Lanczos coefficients to the eigenvalues of the
-Markov chain. 
-\end{abstract}
-
-\maketitle
-
-%%%%%%%%%%%%%%%%%%%%%
-\chapter{Introduction}
-The rate at which a Markov chain converges to a given probability distribution has long
-been an active area of research. This is not surprising considering this problem’s relevance to the ar-
-eas of statistics, statistical mechanics, and computer science. Markov Chain Monte Carlo (MCMC)
-algorithms provide important examples. These algorithms come in handy when we encounter a
-complicated probability distribution from which we want to draw random samples. In statistical
-mechanics, we might wish to estimate the phase average of a function on the state space. Goodman
-and Sokal [6] examine Monte Carlo methods in this context. Examples from statistics occur in the
-Bayesian paradigm when we are forced to simulate an unwieldy posterior distribution (see, e.g.,
-Geman and Geman
-
-To implement the MCMC algorithm, we invent a Markov chain that converges to the
-desired distribution (this is often accomplished using the Metropolis algorithm
-described in Chapter 5). Realizations of the chain will eventually represent
-samples from this distribution. Sometimes ``eventually'' -- meaning all but
-finitely many terms of the chain -- is just not enough. We need more practical
-results. In particular, we want to know how many terms of the chain should be 
-discarded before we are sampling from a distribution that is close (in total variation distance) to
-% 03.txt
-the distribution of interest. This is the purpose of bounding convergence rates for Markov chains.
-
-Often the Markov chains encountered in this context satisfy a condition known in the
-physics literature as detailed balance. Probabilists call chains with this property reversible. This
-simply means that the chain has the same probability law whether time moves
-forward or 
-backward.\footnote{This is not a precise definition. In particular the chain must
-  have started from its stationary distribution. Full rigor is postponed until Section 2.1.}
-In this paper, we consider the rate at which such chains converge to a \emph{stationary distribution}.\footnote{This and other italicized terms are defined in Section 2.1.}
-
-There are a number of different methods in common use for bounding convergence rates of
-Markov chains, and a good review of these methods with many references can be found in  More
-recently developed methods, employing logarithmic Sobolev inequalities, are reviewed in  Most
-of the bounds in common use involve the sub—dominant eigenvalue of the Markov chain's transition
-probability matrix, and thus require good approximations to such eigenvalues. In many applications,
-however, the transition probability matrix is so large that it becomes impossible to store even a
-single vector of the matrix in conventional computer memory. These so called out-of-core problems
-are not amenable to traditional eigenvalue algorithms\footnote{By ``traditional
-  eigenvalue algorithms'' we refer to those found, for example, in Golub and Van
-  Loan[5]. See also the book by Demmel [1] for a more recent discussion.}
-without modification. This paper develops 
-such a modification for the Markov chain eigenvalue problem. In particular it develops a method
-for approximating the first few eigenvalues of a transition probability matrix when we know the
-general structure of the underlying Markov chain. The method does not require storage of large
-matrices or vectors. Instead we need only simulate the Markov chain, and conduct a statistical
-analysis of the simulation.
-
-Here is a look at what follows. Section 2.1 contains a review of the relevant Markov chain
-theory. Readers conversant in the asymptotic theory of Markov chains might wish to at least skim
-Section 2.1, if only to become familiar with our notation. Section 2.2 describes functions on the state
-% 04.txt
-space of the Markov process. This section and Chapter 3 develop the context in which we formulate
-the new ideas of the paper. In the last section of Chapter 3, Section 3.3, we present the familiar
-Krylov subspace and explain why this represents our best approximation to a subspace containing
-extremal eigenvectors of the transition probability matrix.\footnote{or, more
-  precisely, a similarity transformation of this matrix.} The first section of
-Chapter 4 describes the \emph{Lanczos algorithm} for generating an orthonormal basis
-for the Krylov subspace. As it stands, this algorithm is useless for an
-out-of-core problem such as ours since, by definition of such problems, 
-it requires too much data movement; all the computing time is spent swapping data between slow
-and fast memory (e.g. between the hard disk and cache). Therefore, we discuss alternatives to
-Lanczos and demonstrate that the \emph{Lanczos coeficients} are readily available through simulations of
-the Markov chain, which fact allows us to avoid the standard algorithm altogether. Following this
-is a chapter describing the Metropolis algorithm used to produce a reversible stochastic matrix.
-It is here that we experiment with the procedure described in Section 4.2 and approximate the
-extremal eigenvalues of the matrix, without storing any of its vectors. Finally, Chapter 6 concludes
-the paper.
-% 05.txt
-
-\chapter{Markov Chains}
-
-\section{General Theory}
-
-This review of Markov chain theory can be found in any good probability text. The present
-discussion is most similar to that of Durrett [3], to which we refer the reader desiring greater detail.
-
-\subsection{The Basic Setup}
-
-Heuristically, a Markov chain is a stochastic process with a lack of memory property. Here
-this means that the future of the process, given its past behavior and its present state, depends only
-on its present state. This is the probabilistic analogue of a familiar property of classical particle
-systems. Given the position and velocities of all particles at time t, the equations of motion can be
-completely solved for the future evolution of the system. Thus, information describing the behavior
-of the process prior to time t is superﬂuous. To be a bit more precise, if technical, we need the
-following definitions.
-
-\bibliographystyle{spmpsci}
-\bibliography{wjd}
-
-\end{document}
-
-%%%%%%%%%%%%%%%%%%%%%  END DOCUMENT %%%%%%%%%%%%%%%%%%%%%%%%%
-
-Definition 2.1.1 Let (S, 8) be a measurable space. A sequence X,., n 2 0, of random variables
-
-taking values in S is said to be a Markov chain with respect to the filtration a(Xo, . . . ,X,,) if for
-% 06.txt
-a.llBES,
-
-p(x,,+1 e B | a(Xo, . . . ,X,.)) = P(X,.+1 e B 1 o(X,,)) (2.1)
-
-Equation (2.1) merely states that, given the present location of X,,, the past is irrelevant for
-
-predicting the location of X,,+1.
-
-Definition 2.1.2 A function p : S x S —> R is said to be a transition probability if:
-1. for each :1: E S, A —> p(a:, A) is a probability measure on (S, S).
-2. for each A E 8, :1: —> p(z, A) is a measurable function.
-
-We call X,, a Markov chain with transition probabilities 12,, if
-P(Xn+l E B l a(Xn)) =Pn(XnaB) (22)
-
-The spaces (S, S) that we encounter below are standard Borel spaces, so the existence of the
-transition probabilities follows from the existence of regular conditional probabilities on Borel spaces
-— a standard measure theory result (see e.g. [3] page 230).
-
-Suppose we are given an initial probability distribution p on (S, S) and a sequence p,, of
-
-transition probabilities. We can define a consistent set of finite dimensional distributions by
-
-P<X,- e B,-,0 s 1 s n) = / #(d1‘o) / Po($o.d-'51) . .. / Pn-l(97n—1ad$n) (2.3)
-B0 B1 B1:
-Furthermore, denote our probability space by
-(9,?) = (S{0,1,...},8{0,l,...})
-
-We call this sequence space and it is defined more explicitly by
-
-sf°»1=-~-} = {(wo,w1,...) :w.- e S}
-
-S{°*1""} = ¢7(w :40; E A,- E 3)
-
-% 07.txt
-The Markov chain that we will study on this space is simply X" (w) = w,., the coordinate maps.
-Then, by the Kolmogorov extension theorem, there exists a unique probability measure P,, on
-(9,?) so that the X" (w) have finite dimensional distributions (2.3).
-
-If instead of p, we begin with the initial distribution 6,, i.e. point mass at 2:, then we
-denote the probability measure by P3. With such measures defined for each 2:, we can in turn
-
-define distributions P,,, given any initial distribution p, by
-
-PM = / u<da.->P.(A>
-
-That the foregoing construction — which, recall, was derived merely from an initial distri-
-bution p and a sequence p,, of transition probabilities — satisfies Definition (2.2) of a Markov chain
-is not obvious, and a proof can be found in
-
-To state the converse of the foregoing, if X,, is a Markov chain with transition probabilities
-p,, and initial distribution p, then its finite dimensional distributions are given by (2.3). Proof of
-this is also found in
-
-Now that we have put the theory on a firm, if abstract, foundation, we can bring the
-discussion down to earth by making the forgoing a little more concrete. First, we specialize our
-study of Markov chains by assuming that our chain is temporally homogeneous, which means that
-the transition probabilities do not depend on time; i.e. p,,(w,,, B) E p(w,., B). (This is the stochastic
-analogue of a conservative system.) Next we assume that our state space S is finite, and suppose
-for all states 1', j E S that p(i, j) Z 0, and E, p(z', j) = 1 for all i. In this case, equation (2.2) takes
-
-a more intuitive form:
-P(Xn+l =j I X7: = 7:) =P(iaj)
-
-and our transition probabilities become
-% 08.txt
-If P is a matrix whose (i, j)th element is the transition probability p(i, 1') then P is a stochastic
-
-matriar, that is, a matrix with elements pgj satisfying
-
-Pij 20: zjpij =1: (i7j=1121"‘)d)
-
-We also refer to P as the transition probability matrix.
-
-Without loss of generality, we can further suppose our Markov chain is irreducible. This
-means that, for any states 2', 1‘, starting in state i the chain will make a transition to state 1' at
-some future time with positive probability. This state of affairs is often described by saying that
-all states communicate. We lose no generality with this assumption because any reducible Markov
-chain can be factored into irreducible classes of states which can each be studied separately.
-
-The final two conditions we place on the Markov chains considered below will cost us
-some generality. Nonetheless, there remain many examples of chains meeting these conditions,
-thus making the present study worthwhile. Furthermore, it may be the case that, with a little
-more work, we will be able to drop these conditions in future studies. The first condition is that
-the chain is aperiodic. If we let I, = {n 2 1 : p"(:c,a:) > 0}, we call a Markov chain aperiodic if,
-for any state 2:, the greatest common divisor of I, is 1. The second assumption is that our chain is
-
-reversible. This characterization is understood in terms of the following definition.
-Definition 2.1.3 A measure a is called reversible if it satisfies
-
-;4(w)p(I. y) = u(y)p(y. w)
-
-for all :c and y.
-
-We call a Markov chain reversible if its stationary distribution (defined in Section 2.1.2) is reversible.
-% 09.txt
-\subsection{A Convergence Theorem}
-
-In succeeding arguments, we use some results concerning the asymptotic behavior of
-
-Markov diains. These results require a few more definitions.
-
-Definition 2.1.4 A measure 1r is said to be a stationary measure if
-
-Z«(z)p<z.y> = we) (2.4)
-
-2
-Equation 2.4 says P,,(X1 = y) = 1r(y), and by induction that P,,(X,, = y) = 7r(y) for all n 2 1. If 1r
-is a probability measure, then we call 1r a stationary distribution. It represents an equilibrium for
-the chain in the sense that, if X0 has distribution 1r, then so does X,. for all 11..
-
-When the Markov chain is irreducible and aperiodic, the distribution of the state at time
-n converges pointwise to 1r as n —> oo, regardless of the initial state. It is convenient to state
-this convergence result in terms of the Markov chain's transition probability matrix P. Before
-doing so, we note that irreducibility of a Markov chain is equivalent to irreducibility (in the usual
-matrix theory sense) of its transition probability matrix. Furthermore, it turns out that a transition
-probability matrix of an aperiodic Markov chain falls into that class of matrices often called acyclic,
-but for simplicity we will call such stochastic matrices aperiodic. With this terminology, we can
-
-state the convergence theorem in terms of the transition probability matrix P.
-
-% Theorem 2.1.1
-\begin{theorem}
-\label{thm-2.1.1}
-Suppose P is irreducible, aperiodic, and has stationary distribution 7r. Then as n —> oo,
-p"(i,J') -> #0")
-\end{theorem}
-
-The symbol p"(i, j) is to be understood as the (ij) element of the nth power of P.
-A Markov chain whose transition probability matrix satisfies the hypotheses of
-Theorem~\ref{thm-2.1.1} is called ergodic. If we simulate an ergodic chain for
-sufficiently many steps, having 
-% 10.txt
-begun in any initial state, the final state is a sample point from a distribution which is close to 1r.
-
-To make this statement more precise requires that we define “close”.
-
-Definition 2.1.5 Let 1r be a probability measure on S. Then the total variation distance at time
-
-n with initial state I is given by
-AA") = l|P"(z, A) - 7r(-4)||Tv = 1333; I P"(I.A) - 7r(A) I
-
-In what follows, we will measure rate of convergence using the function 'r,(e), the first time after
-
-which the total variation distance is always less than 5. That is,
--r,,(e) = min{m : A,(n) 3 E for all n. 2 m}
-
-To consider the connection between convergence rates and eigenvalues, note that an ape-
-riodic stochastic matrix P has real eigenvalues 1 = /\o > A1 2 A2 3  Z /\¢_1 2 -1, where
-d = |S'] is the dimension of the state space. For an ergodic chain, the rate of convergence to the
-stationary distribution 7r is bounded by a function of the subdominant eigenvalue. By subdominant
-eigenvalue we mean that eigenvalue which is second-largest in absolute value, and we denote this
-Am“ E max{).1,|A¢_1|}. The function bounding the rate of convergence is given by the following
-
-theorem:
-
-% Theorem 2.1.2
-\begin{theorem}
-\label{thm-2.1.2}
-The quantity T1(€) satisfies
-1. 13(5) 5 (1 — /\m,_x)“(ln1r(a:)‘1 +1ne‘1)
-2. maxxes r,(e) 2 %/\,,,,x(1 — ).,,,,x)‘1ln(2e)’1
-\end{theorem}
-
-As this theorem shows, if we have an upper bound on the subdominant eigenvalue, then we have
-an upper bound on the function 11(5). In what follows, we will derive an approximation to the
-% 11.txt
-subdominant eigenvalue and supply error bounds. Together, an approximation and error bounds
-for Am“ provide enough information to make Theorem~\ref{thm-2.1.2} useful.
-
-\section{Functions on the State Space}
-
-Recall that X,.(w) = w,. 6 S denotes the state in which the Markov chain exists at time
-11. Suppose that 11> = {¢1, . . . , ¢,,} is a collection of p observables, or functions defined on the state
-space S. Furthermore, let these observables be real valued, so ¢.~ : S —> R. It is often useful, as we
-will see below, to assume that none of the observables is a constant function. Suppose now that
-the state space S is finite with at possible states. Then, since an observable is simply a map of the
-state space, we can think of each ¢,- as a vector of d real numbers — the d values that it takes on at
-the different states.
-
-Now assume the Markov chain is irreducible, and let 1r denote its stationary distribution.
-If we start the chain from its stationary distribution - i.e. suppose X0 has distribution 1r - then
-X,. is a stationary process. Furthermore, for each i, ¢,(X,.) is a stationary stochastic process with
-
-mean
-
-E.¢.- = Z) vr(x)¢.(z)
-
-165
-
-and autocovariance function
-
-Cir (¢i(Xn)v ¢i (Xn+a))
-
-E7!’  ‘ E1r¢i) (¢i(Xn+a) " E1r¢i)l
-
-Z Pn(Xn = z1X7I+S = y)  — E1r¢i)  _ E1r¢i)
-
-z,yES
-(2.5)
-
-which, by the definition of conditional probability, we can write as,
-
-2 P«(X.. = z)P.(X..+.. = y I X. = 1)(¢s($) — Em) (my) ~ Em)
-1.1165
-% 12.txt
-Z 7r($)P1y (MI) - Em.-) (¢:(1I) - E1r¢i)
-
-:c,y€S
-
-Here pgy denotes the element in row :1: and column y of P’, the sth power of the transition probability
-
-matrix. Similarly, we define the cross-covariance between the function ¢,- at time n and q5,- at time
-
-n+sas
-
-E1!’  _ E7r¢:') (¢i(Xn+a) " E1r¢i)l (2-6)
-
-C1r(¢i(Xn)a (#1 (Xn+a))
-
-Z) vr(x)p;,, (¢.-(w) — Em.-) (my) — Em)
-
-z,yES
-
-Now let (<I>) denote the matrix of mean vectors whose jth column is E,,¢,-1, where 1 = (1, . . . , 1)‘,
-and let H = diag(1r(w1), . . . , 1r(wd)), be the d X d diagonal matrix with stationary probabilities 1r(w)
-on the main diagonal, and zeros elsewhere. Finally, denoting by C(s) the p X p covariance matrix
-
-whose (i,j)th element is C, (qfi,-(X,.),¢_,< (X,.+,)), we have
-
-0(0) = E(‘1’(Xn) - (‘I>))(‘I’(Xn) - (‘PW
-= (‘P-(‘1’))‘1'1(<1’-(‘I>))
-0(8) = E(<I>(Xn) — (‘I>))(‘1’(Xn+a) - (‘1’))’
-
-(‘P - (‘1’))'1'IP‘(<1> — (<1>))
-
-Below, we will also find it useful to have at our disposal a new matrix that is similar to the
-transition probability matrix. We have in mind the matrix M E Ill/2PI'I‘1/2. As is easily verified,
-this allows us to write the covariance matrix as
-
-can = <<I>—<<I>>)°n%*M*n%<<I>—<<I>>>
-
-\II‘M’\II (2.7)
-
-where we have defined ‘II E 111/ 2(<I> —  Recall that our main motivation is that, for out-of-core
-problems, traditional eigenvalue algorithms are inadequate. With this fact, and the form (2.7) in
-% 13.txt
-mind, we will consider using the covariance of observables on the state space to implement the
-(otherwise useless) Rayleigh-Ritz procedure, which we describe below. This procedure requires
-that M be symmetric. As the next fact demonstrates, this required symmetry is the reason for our
-
-interest in the reversibility property of the Markov chain.
-
-Fact 2.2.1 The matrix M is symmetric if and only if the Markov chain is reversible (i.e. iﬂ the
-
-process satisfies the detailed balance condition).
-
-Proof:
-
-M is symmetric
-
-4:, (1-I1/2P1-I-1/2): = H1/2Pn—1/2
-<.—> r1-%*P°n%* = H1/2911-1/2
-¢> 1>'r1= = up
-~:=> (IIP)‘ = l'IP
-
-Element-wise, the final equality implies that 7|‘,'p,'j = 1rjpj.-. According to Definition 2.1.3, this states
-
-that p is a reversible measure.
-
-% 14.txt
-
 \chapter{Invariant and Approximate Invariant Subspaces}
@@
 % 22.txt
-\chapter{Lanczos Procedures}
-
-The following will review the usual procedure for generating an orthonormal basis for
-lC(M, «/11, p), and conclude by showing that, for our problem, we can find the matrix Q‘MQ without
-actually carrying out this procedure. Note that this conclusion is essential since the procedure
-
-requires matrix vector multiplication - an operation we have assumed impossible for large enough :1.
-
-4.1 The General Lanczos Algorithm
-Let 21;“) E M"¢ and consider the d x :1 matrix
-«I 2 [«/»,M«/»,  ,M"“¢] 2 [«b‘°>,«/»“’,...,¢<‘-‘>1
-
-Notice that M\II = [1/:(1),... ,1/;(d'1),Md¢r] and, assuming for the moment that \I1 is non-singular,
-
-we can compute the vector h = —\II‘1M"1/J. Thus,
-
-M\II = \II[e2,. . . ,e,1, —h] E IIIH
-
-% 23.txt
-23
-
-Equivalently,
-0 0  0 -—h1
-1 0  0 —h2
-HE   3 ; =\II“M\II (4.1)
-0 "hl1—1
-1 —h,;
-
-H is a companion matrix which means that its characteristic polynomial is p(:r) = 2:‘ + 23:1 h,-:r"‘1.
-Since H is similar to M, ﬁnding the eigenvalues of M is equivalent to ﬁnding the roots of p(a:).
-However, this is of little practical use since ﬁnding h, constructing 12(1), and ﬁnding its roots is
-probably a harder problem than the one we started with. Instead, the value of decomposition (4.1)
-derives from its upper Hessenberg form. We exploit this property below.
-
-Let ‘II = QR be the QR decomposition of \II. Since ‘I! is assumed nonsingular,
-~II;‘M~r.: = (R"Q‘)M(QR) = H
-Therefore,
-Q‘MQ = RHR-1 2 T
-
-Since R and R“1 are both upper triangular and H is upper Hessenberg, T = R.HR'1 is also upper
-Hessenberg. Furthermore, since M is symmetric, it is clear that T‘ = Q‘MQ = T. That is, T is
-
-an upper Hessenberg matrix that is symmetric. Therefore, T is tridiagonal and we can write it as
-
-follows:
-
-T =    (4.2)
-
-ﬁd—1
-
-0  174-1 ad
-
-% 24.txt
-24
-
-Equating columns j on both sides of MQ = QT yields
-M113‘ = B,-—1qj—1 + 0:91 + ﬁjqj+1 (43)
-
-Since the columns of Q are orthonormal, multiplying both sides of (4.3) by q_,- and q,-+1 yields
-oz, = q§Mq,- and ﬂj = q;-+1Mq,-.
-The foregoing justifies what is called the Lanczos algorithm, performed as follows:
-
-The Lanczos algorithm for (partial) reduction to symmetric tridiaganal form
-
-91 = ¢/||¢||2, [30 = 0, 40 = 0
-
-for j=1top
-
-z=Mq_,-
-
-5:‘ = ||Z||2
-if 5, = 0, quit
-¢Ij+1 = 2/51
-
-end for
-
-The qj computed by the Lanczos algorithm are often called the Lanczos vectors. If the
-loop in the algorithm is terminated because ﬁg = 0, this indicates that an exact invariant subspace
-has been computed, and is given by ran{q1 - - - qk}. Otherwise, we usually halt the algorithm after
-p steps, in which case the algorithm converg to approximations of at most 12 eigenvalues.‘
-
-Note that equation (4.3) results from the equation MQ = QT. The latter equation holds
-since, starting with d vectors in our subspace, Q is a d Xd matrix whose columns form an orthonormal
-
-basis for all of R‘, which is clearly an invariant “sub” space. The Lanczos algorithm above, however,
-
-‘A detailed discussion of such convergence (and misconvergence) is given in
-
-% 25.txt
-25
-
-proceeds for only p steps, producing a d x p matrix Q, whose columns form an orthonormal basis
-for the approximate invariant subspace IC(M, ¢, p), and a p X p matrix T, = Q'MQ. In that case,
-IIMQ = QTp|[g =  is non-zero and gives the error bound described in Theorem 3.2.1. Now,
-
-writing the full d x :1 matrix T of (4.2) as
-
-01 51
-/71
-
-(4.4)
-
-allows us to describe the error bound ||E||-2 in terms of the submatrix Tm, and, hence, in terms of
-
-ﬁzz-
-Theorem 4.1.1
-
-HT, and T,“ are the matrices appearing in (4.4), and if the p columns of Q are computed by the
-
-Lanczos algorithm, then there exist pl, . . . ,;t,, E /\(T,) and A1, . . . , A, E A(M) such that
-
-ill» - /\I=i S iiTP'4”2 = 512
-
-fork=1,...,p.
-
-For a proof, see page 365 of [1].
-
-% 26.txt
-26
-
-4.2 A Lanczos Procedure for Markov Chains
-
-Notice that the Lanczos algorithm requires the computation of Mqj. Throughout this
-paper we have assumed that M 6 Rd“, and that d is so large as to make the matrix vector
-multiplication Mqj impossible. Now, the Lanczos algorithm constructs an orthonormal basis for
-the Krylov subspace. For our purposes, this construction is not essential. What is essential is that
-we ﬁnd a way to generate the Lanczos coejﬁcients oz,-, ﬁj, (j = 1, . . . , p), and whence the matrix
-T, without performing the operation Mqj required by the Lanczos algorithm. We now address this
-problem.
-
-Begin with the centered and weighted observable 1/1 = 111/ 2 (qS—E,,¢), and let q1 = 11;/||1/JH2.
-To simplify notation, let ¢(n) E ¢(X,.). The form of the first Lanczos coeﬂicient a1 is straight
-
-forward:
-
-01 = qiMq1
-
-I/1‘M1/I
-¢’¢
-
-and now the special properties of the Markov chain setting play an important role. For, regard the
-
-following
-
-¢‘M«/2 = (¢ — E.¢)‘II%‘MII%(¢ - Em)
-
-Cw (d>(n), ¢(" + 1)) (4-5)
-
-and
-
-W1!’ = (<15 - E«¢)'H(¢ - E«¢)
-
-Varn (¢(")) (4-6)
-
-Recall that the deﬁnition of M implies H%‘MI'I% = HP, and this justiﬁes equation (4.5). Putting
-
-% 27.txt
-27
-
-these equations together, we have a. nice form for :11:
-
-_ c, (run), an + 1))
-"‘ ' Ta:,(¢(n))
-
-The next Lanczos coeﬂicient, B1, is only slightly more complicated.
-
-[31 = ||M91-0t1tI1||2
-
-[(Mq1 — 0l1¢11)i(M¢]1 - 0101)] 1/2
-
-[t1iM2¢I1 — ail 1/2
-
-(4.7)
-
-Again recalling the deﬁnition M E 111/ 2P1'I‘1/2, a. simple manipulation veriﬁes
-
-1'li‘M’l'[% = HP’, whereby it follows that
-
-(«s — E,,¢)‘HP”(¢ — Em)
-
-"‘M2"‘ (¢ — E,¢)*n<¢ — Em)
-
-C: (¢("), ¢(n + 2))
-V3-Ur
-
-Inserting the last expression into equation (4.7) gives
-
-_ c,(¢(n).¢<n+2)) c,(¢<n>,¢(n+1)> ’ "’
-"“ jvar,<¢<n)> "(jvar,<¢<n)) ‘)
-
-Notice that we can estimate these expressions of oz; and ﬂl by running simulations of the Markov
-chain. To do so, we run the Markov chain starting from a random initial state, compute the value of
-the observable at each step of the chain and, from the data, compute estimates of the covariances
-and variance of the observable.
-
-Before generalizing this for the jth Lanczos coefficient, we consider one more coeﬂicient
-in detail. Recall that the second basis element q; in the Lanczos algorithm is given by 1;; =
-
-(Mql — oz1q1)/ﬁl. From this we can derive an expression for (12 which, though not as neat as those
-
-% 28.txt
-28
-
-for oz; and [31, is nonetheless tractable.
-
-I12 ‘IiMq2
-
-1
-[7(M91 — 0t1¢I1)tM(M¢I1 - 11191)
-1
-
-1
-= f,§(qiM3q1 - 2a1qlM2<11 + a’9iM¢11)
-1
-1
-= —2(<1iM3q1 - 2015i ‘ all (4-8)
-
-1.
-The last equality follows from equation (4.7), which can be written as q{M2q1 = off’ + (if. The
-preceding paragraphs deﬁne all but one of the quantities in the expression (4.8). The one remaining
-is q§M3q1, which can also be expressed in terms of covariances of <15, and the analysis is almost
-
-identical to that for q{M2q1:
-
-(¢ — E,¢)'IIP3(¢ — Em)
-
-"lM3"‘ (¢ — E,¢s)‘H(¢ - Em»)
-
-Cu (¢(")a ¢(n + 3))
-Varn(¢(n))
-
-To generalize the foregoing for higher order Lanczos coeﬂicients requires a recurrence
-relation between (aj+1, ﬂj) and (aj, ﬂ,~_1). As we will see, such a relation involves higher order
-Lanczos vectors q,~ — vectors we do not wish to compute. Recall that the only “vector” we can
-handle is our initial observable. Therefore, we must also establish a recurrence among the q,-’s, thus
-making them available inductively from our observable vector.
-
-To begin with the ﬂfs, recall that ﬁj E ||Mqj — ajq_.,- — ,H_,~_1q,-_;|[2. Whence,
-
-Bf (M411 — °‘j‘Ij - ﬁj-1qj—1)’(MlIj - Otjqj - 51-19141)
-
-q;-M291‘ - 0? - B,z_1 (4-9)
-
-Some simple algebra and the deﬁnitions aj E q§Mqj and ﬁJ~_1 E q;-Mq,--1 prove equality (4.9). The
-% 29.txt
-recurrence for 0114.1 is equally straight forward, though the ﬁnal form is not as neat:
-
-2
-0tj+1 ¢Ij+1M'1j+1
-
-1
-§(M¢15 - 11111;‘ - ﬂj—1<Ij—1)'M(Mqj - Otjqj - /3j—1qj—1)
-J
-
-1
-§(q;-M-“*4; — 2a,~q;M*q,- — 2x3;+’_1q;--1M*q,- + «I? + 2a,-5,51 + am (4.10)
-J
-From (4.9), q;~M2q,~ = oz? + ﬂf + ﬂ}_1 and, upon plugging this into (4.10), we arrive at
-1
-a,~+1 = E(q§»M3q_,,- — (1? — 2ozJ~ﬁJ2 — 2ﬂj_1q;»_1M2qj + ﬁ2_1) (4.11)
-J
-
-The expressions above involve the higher order Lanczos vectors qj. In the present context,
-these vectors will not be readily available. However, another recursion — this time relating forms
-involving q,-+1 to forms involving qj and qj__1 — will make these expressions accessible by induction
-on our initial observable. The general forms of interest are q;+1M"qj+1 and qj-+1M"qJ~. Again, we
-
-begin with the simpler of the two:
-
-¢I;'+1Mk9j = ij(Mqj - Otjqj - ﬂj—1t1j—1)'M"¢1j
-= i(<1§M"“¢Ij - C‘j‘I_;'Mk‘Ij - ﬂj—1q§M"qj-1)
-The second is,
-‘I_;'+1Mkqj+1 = ﬂi?(M<Ij - Otjqj - .3j—1<Ij—1)‘Mk(Mqj — Orjqj - ﬂj—1¢1j—1)
-
-1
-ﬁ(‘1§M'°"<1:‘ + °‘:2"I§M'°<J:' + ﬁ?—z<Ii—1M"<I:‘—1
-J
-—2a,-q;-M*+‘q,~ + 2am,--1q;-M*q,--1 — zza,--1q;-M*+‘q]--1) (4.12)
-
-The point of all this is that each pair of coeﬂicients (oz,-+1, /3,-) can be expressed as a
-function of the covariance between ¢(n) and ¢(n+ m), for m = 0, . . . , 21' + 1. Since the expressions
-are based on four recurrence relations, they become quite complicated as 1' gets large. However,
-deﬁning these recursions in a. computer program is trivial, and we use this fact in Section 5.4 when
-
-applying the foregoing to a particular problem.
-
-% 30.txt
-30
-
-Chapter 5
-
-An Application: Convergence to
-
-Gibbs Measure on the Ising Lattice
-
-5.1 Physical Motivation
-
-Before considering the details of the Metropolis algorithm, it helps to understand the
-setting in which it was developed. To do so we consider an typical example, the Ising model.
-Consider a model of a system consisting of 71 “sites”, each site taking values in {-1, +1}. It might
-help to imagine these sites as equally spaced points on a line or a circle, though we are not restricted
-to such cases. Then our state space 5' consists of all possible n—dimensional permutations of -1’s
-and +1’s. We reprent this as S’ = {-1, +1}", and note that |S| = 2". We can think ofeach site as
-a node existing in one of two possible positions, say “on" or “off”. Each state 1' E S is represented
-by a unique permutation a.- of —1’s and +1’s. Let a,-(k) = :|:l denote the position (on or off) of
-the kth node when the system is in state 1'. An observable ¢ = qS(a,-) on this space is a function of
-
-the permutations 47,-, 1' 6 {1,. . . , 2"}.
-
-% 31.txt
-31
-
-The energy of the system, when in state i, is deﬁned by the Hamiltonian
-
-H07-‘) = - Z Ua(i)0s(k)
-
-<j Ic>
-where the sum runs over all neart neighbor pairs. In statistical mechanics we are often concerned
-
-with the Gibbs measure of state i, deﬁned by
-1r(a,-) = Z'1 exp{—[3H(a,~)} (5.1)
-where Z ‘1 is the normalization constant, or partition function. For the Ising model, we put
-
-Z = §:9XP{~ﬁH(<7:‘)}
-i=1
-
-so that 7r is a probability measure. Now, letting 1: denote the so called Boltzman constant, if a
-classical mechanical system is in thermal equilibrium with its surroundings with absolute temper-
-ature T, and is in state i with energy H (05), then the probability density in phase-space of the
-point representing state i is given by (5.1) with [3 = (lcT)“. The central result of ergodic theory
-implies that we can also interpret 1r(o;) as the proportion of time the system spends in state i. If
-
-the system is observed at a random time, the expectation E,,¢ of any observable ¢ is thus
-
-.31 ¢(0a) 9"l>{-ﬂH(0s‘)}
-
-E” = 2?;.exp{—nH(a.-)}
-
-(5.2)
-
-Perhaps the number of sites n in our Ising model is so large that it is impractical or impossible to
-evaluate (5.2). We might instead consider importance sampling and generate states with probability
-
-density
-
-exp{—aH<a.o}
-2?;.exp{—zaH<a.->} (53)
-
-Then «I: is itself an unbiased estimator of (5.2).
-
-1r(0a') =
-
-If evaluation of (5.2) is difficult, there is no reason to believe that evaluation of (5.3) is any
-
-easier. However, Metropolis and his collaborators [9] contrived a method for producing an ergodic
-
-% 32.txt
-32
-
-Markov chain with transition probability matrix K whose elements Ic(:::, y) satisfy
-
-2«<z>k<x.y) = «<2»
-
-I
-
-By Deﬁnition 2.1.4 this means that the Markov chain has stationary distribution 7r. Since the
-drain is ergodic, we know that it will eventually converge to the desired distribution. So, our main
-problem — the rate at which such a Markov chain will converge — is clearly at issue in the application
-of the Metropolis algorithm. Moreover, as we will see below, the Metropolis algorithm ensures that
-the Markov chain has the desired stationary distribution by requiring that the chain satisfy the
-stronger condition of reversibility; i.e. detailed balance. Therefore, Markov chains produced by the
-Metropolis algorithm satisfy the assumptions of this paper, which fact conﬁrms that our theory
-
-can be used to bound convergence rates of all such chains.
-
-5.2 Description of the Metropolis Algorithm
-
-The following explains how the Metropolis Algorithm is carried out. The exposition is
-similar to that given in Hammersley and Handscomb’s book Monte Carlo Methods 
-To begin, choose an arbitrary symmetric Markov chain; that is, a Markov chain whose
-
-transition probability matrix P is symmetric. P is called the proposition matrix, and its elements,
-P071"), Satisfy
-pa, 1) 2 0. pm‘) = pm). 2,-pm‘) = 1 vi e S (5-4)
-
-Using P, we will derive a new transition probability matrix K which satisﬁes
-
-Z vr(z')k(='.:') = arm (5.5)
-
-I
-
-For 1' aé j deﬁne
-
-kw) = p<-yum")/«<¢> it nu)/1r(='> <1 (5.6)
-
-P(i,J') if 7r(J')/7r(i) Z1
-
-% 33.txt
-33
-
-For 2' = j deﬁne
-
-kw) = pm) (1—  p(-3;"), where J. = {j = 2‘ 7‘ an nu)/«(='> < 1}
-
-Notice that 1r(i) > 0 implies k(i,j) _>_ 0 and
-
-;k<«',j> = pow)  { (1 —  pm) +  +]§§p<«",:')
-= P(is'5) + 2 P(i1.1') + Z P(iy.7') (5-7)
-jel.» jug
-
-Equation (5.7) shows that, for each i, 2, k(z', j) = 1, which fact conﬁrms that K is a stochastic
-
-matrix.
-
-Next we check that K satisﬁes the detailed balance condition. First suppose that 1r(z') =
-
-1r(j). Then (5.4), (5.6), and symmetry of P imply
-k(i,J') = p(i2.7') = p(jvi) = k(j1i)
-
-and, since 1r(i) = 1r(j), the detailed balance condition, 1r(i)k(i,j) = 7r(j)k(i,j), clearly holds. Now
-
-suppose that 1r(i) > 1r(j). Then (5.4) and (5.6) imply k(j,i) =p(j, 1') and
-k(i,J') = P(iyJ')7r(J')/7r(i) = P(J3i)7r(J')/7r(i) = k(.1'» i)7r(J')/1r(i)
-
-That is, 1r(i)lc(i,j) = 1r(j)k(i,j), as we set out to prove.
-To complete the justification of the Metropolis algorithm, we note that the detailed balance
-condition, 7r(i)k(i, j) = 1r(j)k(i, j), implies that 1r is indeed the stationary distribution for a Markov
-
-chain with transition probability matrix K. To see this notice that
-
-27r(i)k(i..7') = Zﬂijlkiiajl = 1r(J') Z/€(i,J‘) = 7r(J')
-
-I i
-We now summarize the tasks performed when implementing the Metropolis algorithm.
-
-First, pick as a proposition matrix any symmetric stochastic matrix and denote it by P. Begin
-
-% 34.txt
-34
-
-the simulation of the Markov chain in state i, and then take one step of the proposition Markov
-chain to arrive in state j. That is, choose j according to the mass function p(z', -). Next, compute
-1r(j)/1r(i). If 7r(j)/1r(i) 2 1, accept j as the new state. If1r(j)/1r(i') < 1, then with probability
-1r(j)/7r(z') accept j as the new state; otherwise, with probability 1 — 1r(j)/1r(i), take i as the new
-state. Performing each of these tasks produces one step of the Markov chain represented by the
-
-transition probability matrix K.
-
-5.3 The Ising Model
-
-Returning to the Ising model considered in section 5.1, we consider how the Metropolis
-algorithm is applied to such a mode]. Again, start the simulated chain in state X0 = 1' and let
-
-X1 = j be a state chosen according to p(i, -), where p(i, j) = p(j,i). Next, compute
-
-7"(‘7j)/7"(‘7i) = eXP{-ﬁlH(<7j) - H(0i)l}
-
-Suppose the energy H (Uj) of the new state satisﬁes H (U,-) 3 H (05). Then 1r(a_,-)/1r(a,-) Z 1 and we
-
-move to state j. On the other hand, if H(aj) > H(a;), then 1r(aj)/7r(a,-) < 1 and we take
-
-j with probability e-M"
-
-X1 (5.8)
-
-i with probability 1 — e-W’
-where AH E H (0,) — H (:75) is the change in energy from state j to state 1'.
-There are many ways to choose a proposition matrix P. The simplest derives from the
-so called Glauber dynamics, which proceeds as follows: choose one of the 11. sites at random with
-probability 1/n — let .9 denote the chosen site — and change the position of this site; i.e., multiply
-
-its value by -1. The resulting state j is the permutation of -1’s and +1’s given by
-
--0,-(k) if]: = .9
-a,~(Ic)= (k=1,...,d)
-
-a,-(Ic) if]: 95 s
-
-% 35.txt
-35
-
-The procedure we employ below involves a slight variation: choose site .9 at random uniformly
-from the 11 possible sites. However, instead of changing the value of the chosen site with absolute
-certainty, as in Glauber dynamics, we “ﬂip the switch” with probability 1/2. Therefore, when k yé .9
-
-we have 0',-(k) = a.~(k), and when k = 3,
-
-—a,~(s) with probability 1/2
-01(3) =
-
-a,-(3) with probability 1/2
-The advantage of this procedure is that the proposition matrix, before altering it with the Metropo-
-lis algorithm, is a bit more interesting than that produced by traditional Glauber dynamics. For,
-it has 1/2 along the main diagonal and, thus, is aperiodic and converges to a stationary distribu-
-tion. Glauber dynamics has period 2, so its proposition matrix does not satisfy the hypotheses of
-
-our convergence theorem (Theorem 2.1.2). Keep in mind, however, that any proposition matrix
-
-modiﬁed by the Metropolis algorithm satisﬁes all of our hypotheses.
-
-5.4 Computations
-
-5.4.1 Computer Programs
-
-Constructing a computer program to simulate realizations of the Markov chain described
-above (i.e. with transition matrix K) is straight forward, and we do so using Matlab and the
-program ising.m which appears in the Appendix.
-
-When performing simulations in the present context, there are a few important aspects
-to consider. First, our program begins with a “hot start", which means that the initial state is
-picked by assigning each site to :l:1 with probability 1/2. Perhaps we should instead have picked
-our initial state from the stationary distribution. On the contrary, this would defeat the whole
-
-purpose of our analysis. Recall that the method has been developed to deal with cases for which
-
-% 36.txt
-36
-
-sampl from the stationary distribution are not quickly attainable. Furthermore, we are trying to
-ﬁgure out how long it will take for observations from the simulated chain to represent samples from
-the stationary distribution. Second, the method we have developed depends solely on covariances
-of our observable which we estimate from the data. produced by the simulations. However, such
-estimates require that the data come from the stationary distribution. Therefore, we must discard a
-number of observations before using the data to estimate covariances. The number of observations
-to discard depends on how long must we wait until we have data from the stationary distribution,
-and we are back where we started — almost. When employing our procedure, we can discard a
-conservative (very large) number of observations and perform our analysis, deriving a bound on
-the convergence time. Then, in all future studies, we will know how much data should be discarded
-before samples come (approximately) from the stationary distribution.
-
-In considering what observable to use on the state space described above, perhaps the
-most obvious is simply the number of nodes in the on position (i.e. the number of +l’s). Recalling
-that 0, (k) = i1 denotes the position (on or off) of the kth node when the system is in state j, this
-
-observable is deﬁned by,
-«M = zxca.-ac) = +1). 1 = 1.---.ISl
-I:
-
-where X is the indicator function. The program ising.m simulates the Markov chain, computes the
-values of this observable (as well as the number of 1’s squared) and writes it to a ﬁle called phi . dat.
-The program then proceeds to compute the ﬁrst few Lanczos coefficients using covariances produced
-by Matlab, so that we can check them against the rults we get from the 1anczos.c program,
-which we now describe.
-
-The lanczos .c program (listed in the Appendix along with its dependencies) implements
-the recursive relations described in section 4.2 to compute the Lanczos coefﬁcients by the new
-
-method. It accepts input from the user specifying any observable, any number of iterations, and any
-
-% 37.txt
-37
-
-TABLE 1: ESTIMATES or /\,m(K) = 0.998746 USING ¢1
-
-sim. # /32
-1 0.995107 0.006960 0.995220 0.600243
-2 0.995099 0.005891 0.994863 0.477393
-3 0.994862 0.009671 0.994925 N/ A
-4 0.995321 0.009319 0.995654 N/A
-5 0.994787 0.004872 0.405446 0.994742
-
-number of desired Lauczos coeﬁicients. That is, the observable could have come from a simulation
-of any reversible Markov chain and not just our special Metropolis chain. The output of this
-program is a T matrix containing the maximum number of Lauczos coeﬂicients (up to the number
-requested by the user) before an approximate invariant subspace was reached. The eigenvalues of
-the nonzero part of this matrix, in accordance with the foregoing theory, should well approximate
-
-eigenvalues of the Markov chain’s stochastic matrix.
-
-5.4.2 Results
-
-Table 1 shows the results of 5 simulations each of 100,000 iterations (discarding the ﬁrst
-5,000 observations) for the one dimensional Ising lattice with 10 nodes and [7 = 1. Displayed is
-the estimate of the ﬁrst Lauczos coeﬂicient a1 (which is the eigenvalue estimate after one step; i.e.
-).m,x(T1)), along with its error bound ﬁl. Appearing in the third column is the largest eigenvalue of
-the 2 x 2 matrix T2, A0 (T2), and ﬁg appears in the fourth column when it is was computable. When
-it was not computable, it was probably because an approximate invariant subspace was reached
-at the ﬁrst step, so ﬁg involved expressions which were close to machine epsilon and, thus, could
-not be computed accurately. Table 2 shows the same information when using a second observable,
-¢2 E ¢§ (i.e. the number of +1’s squared).
-
-The true value of the subdominant eigenvalue for this problem is /\max(K) = 0.998746,
-
-and the next largest is A2(K) = 0.993303. These were computed directly from the transition matrix
-
-% 38.txt
-38
-
-TABLE 2: ESTIMATES OF >.,,,,,.,,,(K) = 0.998746 USING z/>2
-
-sim. # B2
-1 0.994262 0.006142 0.994222 N / A
-2 0.994333 0.002974 0.994332 N / A
-3 0.994307 0.008892 0.994374 0.726320
-4 0.994436 0.012884 0.994685 0.361404
-5 0.993789 0.006909 0.993820 0.788710
-
-using the Matlab program tpm.m listed in the Appendix. They should be very accurate, and it
-takes the Matlab routine eig() just over 20 minutes of cpu time on a Sun Ultra to ﬁnd all the
-eigenvalues of K.
-
-As shown by Theorem 4.1.1, the [Ts provide conservative error bounds on the eigenvalue
-estimates from the previous Lanczos step. Unfortunately, for this example, ,6 appears to be too
-conservative as it does not even allow us to bound our estimates away from 1. Furthermore, it
-seems that the estimates generated by the method above fall systematically below the true value of
-the subdominant eigenvalue. In every simulation, an approximate invariant subspace (to machine
-tolerance) was reached after computing at most 2 pairs of Lanczos coefficients. For ¢1 we see that
-it was reached after computing only one pair on simulations 3 and 4, similarly for ¢2 on simulations
-1 and 2. These observations indicate that the chosen observable is close to the slowest mode (or
-eigenfunction) of the process. That is, we should observe fast convergence on an approximate
-invariant space. However, the space found might contain only the second slowest mode of the
-process and, in that case, our eigenvalue estimates would be closer to the third largest eigenvalue,
-instead of the second largest as desired.‘
-
-The matrix T was produced by the program lanczos . c and its eigenvalues were computed
-by Matlab, both operations taking a few seconds. Granted, comparing a few seconds to the 20
-
-minutes it takes Matlab is not justiﬁed since a simple call to the routine eig() is far from the
-
-‘Recall that we explained above why we will never converge to the eigenspace corresponding to the largest eigen-
-value (which is always 1). That would require a (nearly) constant observable function.
-
-% 39.txt
-39
-
-most eﬂicient traditional method for computing only a few eigenvalues of K. However, recall that
-the primary motivation for the new method are those examples where the matrix K is so large
-that we can’t store even a single vector in fast memory. Once we start swapping data to and from
-cheap memory, the time required by traditional algorithms can become years, even lifetimes. Such
-examples are not exceptional. Consider, for instance, the 1000 node problem which produces a.
-
-2“’°° x 21°°° transition matrix.
-
-% 40.txt
-40
-
-Chapter 6
-
-Conclusion
-
-We conclude by noting a few aspects of the special cases to which the foregoing applies.
-We are interested in the subdominant eigenvalue of a large transition probability matrix. The
-
-foregoing can be used to approximate these eigenvalues when the following conditions are satisﬁed:
-o The Markov chain is reversible (detailed balance condition).
-
-0 We can easily simulate realizations of the chain as well as functions (observables) of these
-
-realizations.
-
-When these conditions hold, we can side step the traditional Lanczos algorithm, and get at the
-values of the Lanczos coeﬂicients via the variational properties of observables on the state space.
-There are many ideas in this paper that should be expanded upon in future studies. Per-
-haps most obvious is the desire to ﬁnd analogous techniques for non-reversible (i.e. non-symmetric)
-Markov chains. This is a considerable problem since the Lanczos methods described above are no
-longer applicable. Perhaps the Amoldi algorithm ([1], algorithm 6.9), or the non-symmetric Lanczos
-
-algorithm can also be usefully specialized for stochastic matrices.
-
-% 41.txt
-41
-
-When applying the Lanczos algorithm, we usually choose a single starting vector (in the
-present context, an observable), perform the algorithm for a number of steps, and gather results.
-Then we do the same with a new starting vector and compare the results to those obtained with
-the ﬁrst vector. Proceeding in this way for a number of starting vectors, we can provide statistical
-evidence that we have, indeed, located the eigenvalues of interest. However, when there are two
-observables 451, :15; of particular interest, perhaps we would beneﬁt by considering both observables
-simultaneously. Future studies might develop an algorithm which incorporates both observables,
-
-again via ¢.- = H1/2(¢.' - E,.¢.-). In this case, a Lanczos algorithm would be based on the space
-
-Ic*(M,¢1,¢»2.p) 2 ran{«/»mz»2,M«/»1,M«/:2, . . .,MP—‘¢1.M"—‘«/22}
-
-The Lanczos coeﬁicients would then involve not only the autocovariances C, (¢,<(n),¢,-(n + m)),
-
-but also the cross-covariances C, (¢.~(n), ¢j(n + m)) (deﬁned in (2.6)).
-
-% 42.txt
-Appendix: computer programs
-
-Z Hatlab code ising.m
-
-Z
-Z Written by William J. De Meo on 12/15/97
-
-2 last modified 1/10/98
-
-Z Inputs:
-
-2 d = number of nodes of the ising lattice (e.g. d=100)
-Z n = number of iterations (e.g. n=1000)
-
-Z beta = Annealing schedule (e.g. beta = 2,
-
-1 when beta -> 0 will always go to nee state
-Z vhen beta -> infty will never go to state of higher energy)
-Z
-
-X = zeros(d,1);
-Energy = zeros(n,1);
-
-Ave=0;
-
-X Initialize using hot start
-for i=1:d,
-
-if rand < .5
-
-x(i) = -1;
-
-else
-
-X(i) = 1;
-
-end
-end
-
-Z Initialize at state of high energy (not used)
-
-if 0
-X(1:2:d-1)=1;
-X(2:2:d)=—1;
-and
-
-Z Initialize energy
-
-H=0;
-
-for i=1:(d-1)
-
-H = H - X(i)*X(i+1);
-end
-
-2 Initialize observable
-phi = zeros(n+1,1);
-
-for i=1:d
-if X(i) == 1
-phi(1)= phi(1)+1;
-end
-
-end
-
-42
-
-% 43.txt
-43
-
-for j=1:n
-
-site=0;
-while site==
-
-site = round(rand#d);
-end
-Z Compute new energy (with new X(site) = —x(site)):
-if site ==
-
-Hnew = H + 2:x(1)~x(2);
-elseif site == d
-
-Hnew = H + 2*X(d)tX(d-1);
-else
-
-Hnew = H + 2*X(site)*(X(site-1) + X(site+1));
-end
-
-Z Change to new state in two ways:
-X with probability 1/2, flip the switch
-if rand < .5
-if (Hnew <= H) I (rand < exp(-beta*(Hnew-H)))
-x(site) = -x(site);
-H = Hnew;
-Z Compute new value of observable
-if X(site) -= 1
-phi(j+1) = phi(j) + 1;
-
-else
-phi(j+1) = phi(j) — 1;
-end
-else phi(j+1)=phi(j);
-end
-else phi(j+1)=phi(j);
-end
-Energy(j) = H;
-end
-p1ot(Energy)
-
-% A second observable
-phi2 = zeros(n+1,1);
-for i=1:n+1
-
-phi2(i) = phi(i)*phi(i);
-end
-
-Z Write first observable to phi.dat:
-
-fid = fopen(’/quean.a/masters/deme5025/research/nyuthesis/programs/c/phi.dat’,’w’);
-fprintf(fid,’%f\n’,phi);
-fc1ose(fid);
-
-Z Write second observable to phi2.dat:
-fid = fopen(’/quean.a/masters/deme5025/research/nyuthesis/programs/c/phi2.dat’,’w’);
-
-% 44.txt
-fprintf(fid,’%f\n’,phi2);
-fc1ose(fid);
-
-1 Compute a few Lanczos coefficients by new method
-START=1001;
-
-V1=zeros(2);
-
-C1=zeros(2);
-
-C2=zeros(2);
-
-C3=zeros(2);
-
-V1 = cov(phi(START:n+1),phi(START:n+1));
-
-V1 = V1(1,1)5
-
-C1 = cov(phi(START:n),phi(START+1:n+1));
-C1 = C1(1,2);
-
-C2 = cov(phi(START:n-1),phi(START+2:n+1));
-C2 = C2(1,2);
-
-C3 = cov(phi(START:n-2),phi(START+3:n+1));
-
-C3 = C3(1,2);
-
-a1pha1 = C1/V1;
-
-betai = sqrt(C2/V1 ' (C1/V1)‘2);
-
-a1pha2 = (1/(beta1’2))*(C3/V1 - 2*a1pha1*(C2/V1) + a1pha1‘3);
-
-T1 = Ealphal betal; betal a1pha2]
-
-v1 = cov(phi2(START:n+1),phi2(START:n+1));
-v1 = v1(1,1);
-
-c1 = cov(phi2(START:n),phi2(START+1:n+1));
-c1 = c1(1,2);
-
-c2 = cov(phi2(START:n—1),phi2(START+2:n+1));
-c2 = c2(1,2);
-
-c3 = cov(phi2(START:n-2),phi2(START+3:n+1));
-c3 = c3(1,2);
-
-alphal = C1/V1;
-betal = sqrt(C2/V1 - (C1/V1)‘2);
-a1pha2 = (1/(beta1‘2))#(C3/V1 - 2ta1pha1*(C2/V1) + a1pha1”3);
-
-T2 = [a1pha1 betai; betal a1pha2]
-
-fid = fopen(’Tmats.dat’,’v’);
-fprintf(fid,’%f\n',T1);
-fprintf(fid,’%f\n’,T2);
-fc1ose(fid);
-
-ZZZ end ising.m
-
-44
-% 45.txt
-\% Matlab code tpm.m for generating transition probability matrix
-Z and computing its eigenvalues directly
-
-1
-
-Z Written by Hilliam J. De Heo on 1/10/98
-
-1
-
-Z Inputs:
-
-Z d = number of nodes of the ising lattice (e.g. d=10)
-
-Z beta = Annealing schedule (e.g. beta - 2,
-
-1 when beta -> 0 will always go to new state
-1 when beta -> infty will never go to state of higher energy)
-2
-
-disp(’bui1ding proposition matrix...’)
-
-states = 2‘d;
-
-A = 0;
-
-for i=0:(d-1)
-
-E=eye(2”i);
-
-A = [A E;E A];
-
-end
-
-B = .5#eye(states);
-
-A = B + .5-(1/d)*A;
-
-2 our modified Glauber dynamics requires the B and the .5¥(1/d)
-
-disp(’...done’)
-
-1 display pattern of nonzero entries
-1 spy(A)
-
-Z check that all row sums are 1
-check=O;
-for i=1:states
-
-check = check + (not(sum(A(i,1:states))<.99));
-end
-1 if not all rows sum to 1, print check = (# of rows with sum=1)
-if not(check==states)
-
-check
-
-error(’Row sums are not all 1’)
-end
-
-Z construct matrix of states
-E = zeros(d,states);
-f1ip=-1;
-for i=1:d
-for j=1:2‘(i-1):states
-flip = -1*f1ip;
-for k=0:2“(i-1)
-
-45
-
-% 46.txt
-E(i,j+k) = flip;
-and
-end
-end
-E=E’;
-
-Z display first 64 states
-Z E(1:64,:)
-
-Z compute energy of each state
-H = zeros(states,1);
-for i=1:states
-
-for j=1:d-1
-H(i) = H(i) - E(i,j)tE(i,j+1);
-end
-end
-
-disp(’building Hetropolis transition matrix...')
-
-for i=1:states
-for j=i+1:states
-if not(A(i,j)==0)
-if(H(j)>H(i))
-alt = A(i,j)t(1-exp(-beta*(H(j)-H(i))));
-A(i,i) I A(i,i)+a1t;
-A(i,j) = A(i,j)-alt;
-elseif(H(j)<H(i))
-alt = A(j,i)t(1-exp(—beta#(H(i)-H(j))));
-A(j,j) = A(j,j)+a1t:
-A(j,i) = A(j,i)-alt;
-end
-end
-end
-end
-
-disp(’...done’)
-
-check=0;
-for i=1:states
-
-check = check + (not(sum(A(i,1:states))<.99));
-end
-
-Z if not all rows sum to 1, print check = (# of rows with aum=1)
-
-if not(check==states)
-
-check
-
-error(’Row sums are not all 1’)
-end
-
-46
-
-% 47.txt
-47
-
-disp(’computing eigenvalues of tpm...’)
-cput = cputime;
-
-evals = eig(A);
-
-ecput = tputime - cput;
-
-disp(’the CPU time (in secs) for computing eigenvalues of tpm: ’)
-
-ecput
-ZZZ end tpm.m
-
-/¥#*##¥###*##*#l####¥#‘###$ﬂ###ﬁi##*U3*#**#*##**##¥***ll¥#*##¥ii
-
-* 1anczos.c main program for computing Lanczos coefficients *
-3 #
-* Created by William J. De Heo #
-t on 1/7/98 *
-# Last modified 1/10/98 #
-
-*¥#######**3It¥*#¥¥¥#**##**¥##¥#####**####*#*###*##**###¥##*##*/
-
-#inc1ude <stdlib.h>
-#include <stdio.h>
-#include <math.h>
-#inc1ude "prototypes.h"
-
-/# Machine constants on dino (Sun Ultra 1 at Courant) t/
-#define HACHEPS 1.15828708535533604981e-16
-
-#define SQRTEPS 1.07623746699106147048e-08
-
-ﬂdefine HAX_NAH 100
-
-void read_name(char *);
-
-/* Prototypes from the file functions.c ¥/
-
-double a1pha(int j);
-
-double betasq(int j);
-
-double form1(int j, int k);
-
-double form2(int j, int k);
-
-double moment(doub1e ‘data, long n, double tave, double tvar, double tcov, long k);
-
-/# external variables to be used by functions */
-double *phi,tcov, tvar, tave;
-
-% 48.txt
-main()
-
-{
-
-char tfilename;
-
-FILE *ofp;
-
-double temp=O, *T;
-
-long i, j, nrou, nlanc, START;
-int f1ag=O;
-
-filename = cma11oc(HAx_NAHE);
-
-printf("\nName of file containing observed values: “);
-
-read_name(fi1ename);
-
-printf(“\nTota1 number of observations in file (iterations):
-
-scanf(“Zu”,&nrov);
-
-printf(“\nNumber of leading observations to discard: “);
-
-scanf("Zu",&START);
-
-printf("\nNumber of Lanczos coefficients desired: “);
-
-scanf("%u",&n1anc);
-
-vhi1e(2*n1anc > nrow—START)
-
-{
-printf("\nNot enough data for that many coefficients.\n“);
-printf("\nEnter a smaller number of Lanczos coefficients: “);
-scanf("Xu“,&n1anc);
-
-}
-
-II);
-
-phi = dma11oc(nrow);
-cov=dma11oc(2#n1anc);
-var=dma11oc(1);
-ave=dma11oc(1);
-
-T = dma11oc(n1anc#n1anc); for(i=0;i<n1anc;i++) T[i]=(doub1e)0;
-
-/* observable phi is stored contiguously column-wise (e.g. by HATLAB) t/
-mat1abread(phi, nrow, 1,fi1ename);
-
-/* send the observable, offset by START */
-moment(phi+START,nrou-START,ave, var, cov,2*n1anc-1);
-
-/* First column of T */
-
-T[0] = a1pha(1);
-if((temp=betasq(1))>HACHEPS*10)
-{
-
-T[1]=sqrt(temp);
-
-/* General column of T */
-
-J'=2:
-
-for(i=1;i<n1anc-1, f1ag==O;)
-{
-
-T[itn1anc+i-1]=sqrt(betasq(i));
-
-48
-
-main()
-
-{
-
-char tfilename;
-
-FILE tofp;
-
-double temp=0, #T;
-
-long i, j, nrow, nlanc, START;
-int f1ag=0;
-
-filename = cma11oc(HAx_NAH);
-
-princf("\nName of file containing observed values: “);
-read_name(fi1ename);
-
-printf("\nTota1 number of observations in file (iterations): “);
-scanf("2u“,&nrow);
-
-printf("\nNumber of leading observations to discard: “);
-scanf("Zu",&START);
-
-printf("\nNumber of Lanczos coefficients desired: “);
-scanf("%u“,&n1anc);
-
-whi1e(2*n1anc > nrou-START)
-
-{
-printf("\nNot enough data for that many coefficients.\n");
-
-printf("\nEnter a smaller number of Lanczos coefficients: “);
-scanf("Zu",&n1anc);
-}
-
-phi = dma11oc(urow);
-cov=dma1loc(2*n1anc);
-var=dma11oc(1);
-ave=dma11oc(1);
-
-T = dma11oc(n1anc*n1anc); for(i=0;i<n1anc;i++) T[i]=(doub1e)O;
-
-/* observable phi is stored contiguously column-wise (e.g. by HATLAB) */
-mat1abread(phi, nrov, 1,fi1ename);
-
-/* send the observable, offset by START #/
-moment(phi+START,nrov-START,ave, var, cov,2*n1anc-1);
-
-/# First column of T */
-
-T[0] = a1pha(1);
-if((temp=betasq(1))>HACHEPS#10)
-{
-
-T[1]=sqrt(temp);
-
-/* General column of T */
-
-J‘=2:
-
-for(i-1;i<n1anc-1, f1ag==0;)
-{
-
-T[i*nlanc+i-1]=sqrt(betasq(i));
-
-48
-
-% 49.txt
-T[i*n1anc+i] = a1pha(i+1);
-if((temp = betasq(i+1))>HACmEPS#10)
-{
-
-T[i#n1anc+i+1] = sqrt(temp);
-
-i++;
-}
-else
-
-f1ag=1;
-
-}
-
-/t Last column */
-if(f1ag!=1 it ((temp= betasq(n1anc-1))>HACHEPSt10))
-
-{
-T[n1anc#n1anc-2] = sqrt(temp);
-T[n1anctn1anc-1] = a1pha(n1anc);
-printf("\nbeta(1) = %1f\nbeta(Zd) = Zlf (last beta)",
-T[1],(nlanc-1),T[nlanc*n1anc-2]);
-}
-else
-{
-printf("\nApproximate invariant space reached at step %d.",i);
-printf("\nbeta(1) = Z1f\nbeta(Zd) = Zlf (last accurate beta)“.
-T[1],i,T[itn1anc+i—1]);
-printf("\nbeta(Zd)“2: Zlf (first spurious resu1t)“,i+1,temp);
-}
-
-printf("\nThe matrix T is: \n");
-matprint(T,n1anc,n1anc);
-
-}
-
-else
-
-{
-
-printf("\nmain(): Approximate invariant space reached at first step.“);
-
-49
-
-printf("\na1pha(1) - Z1f\nbeta(1)”2: Zlf (first spurious result)“, T[O], temp);
-
-}
-ofp = fopen("Tmat.m","v“);
-check(ofp);
-mat1aburite(T,n1anc,n1anc,ofp);
-fc1ose(ofp);
-}
-void read_name(char tname)
-{
-int c, i = 0;
-while ((c = getchar()) != EOF it c != ’ ’ it c !- ’\n’)
-name[i++] = c;
-name[i] = ’\0’;
-}
-
-/#t* end 1anczos.c *#*/
-
-% 50.txt
-50
-
-/llnkttttttttinlntint*1!!!##1##!ttitilltit###t¥#tt¥ttItt####t#ttII=##mlltttt
-
-* functions.c functions required by 1anczos.c t
-# t
-* Created by Hilliam J. De Meo #
-* on 1/7/98 *
-t Last modified 1/10/98 :
-
-¥¥##¥$¥##¥¥#$¥¥¥###¥¥$*I#****#**####*######¥#¥#¥#$#¥¥#l¥**#****/
-
-#inc1ude <math.h>
-
-#define START 1000
-
-#define ITER 10000
-
-/t Machine constants on dino (Sun Ultra 1 at Courant) */
-
-#define HACHEPS 1.15828708535533604981e-16
-
-#define SQRTEPS 1.07623746699106147048e-O8
-
-double a1pha(int j);
-
-double betasq(int j);
-
-double form1(int j, int k);
-
-double form2(int j, int k);
-
-double moment(doub1e tdata, long n, double tave, double tvar, double tcov, int k);
-
-/t external variables to be used by functions */
-double *phi,tcov, *var, tave;
-
-double alpha(int j)
-{
-
-/* alpha is never called with j < 1 */
-if(j==1)
-return
-form1(1,1);
-else if(j>1)
-return
-((doub1e)1/betasq(j-1))*
-(form1(j-1,3) — pou(a1pha(j-1),3)
-- 2 * (alpha(j-1)#betasq(j-1) + sqrt(betasq(j-2))#form2(j-1,2))
-+ betasq(j-2));
-}
-
-double betasq(int j)
-
-{
-if(j==0)
-return (doub1e)O;
-else if(j>0)
-return (form1(j,2) - pow(a1pha(j),2) - betasq(j-1));
-}
-
-double form1(int j, int k)
-{
-
-% 51.txt
-51
-
-double form14,alphal,alpha1sq,form12,form13;
-if(j==0)
-
-return (doub1e)0;
-else if(j==1)
-{
-
-/# printf("\nvar = %1f, cov(%d) = %1f \n“,*var,k,cov[k]);t/
-
-return (cov[k])/(*var); /* the only real value */
-}
-else if(j>1)
-{
-
-return ((double)1/betasq(j-1))
-
-* ( form1(j-1,k+2)
-+ pow(a1pha(j-1),2)*form1(j-1,k)
-+ 2t(a1pha(j-1) t sqrt(betasq(j-2)) * form2(j-1,k)
-- a1pha(j-1)*form1(j-1,k+1)
-- sqrt(betasq(j—2))#form2(j-1,k+1) )
-+ betasq(j-2)*form1(j-2,k)):
-}
-}
-
-double form2(int j, int k)
-{
-/* form2 is never called with j < 1 */
-if(j==1)
-return (doub1e)0;
-else if(j>1)
-return
-(pov(betasq(j-1),—.5)) 3
-(form1(j-1,k+1) - a1pha(j-1)*form1(j-1,k)
-- sqrt(betasq(j-2)) # form2(j—1,k));
-}
-
-/* moment() function for computing var and cov(k)
-arguments:
-data = a nxi array of doubles
-n = length of data[]
-ave =(on exit)= the average of data[]
-var =(on exit)= the variance of data[]
-cov =(on exit)= the covariance of data[i] and data[i+j] for i-1,...,k
-k = the max lag for cov above
-*/
-double moment(double tdata, long n, double tave, double tvar, double tcov, int k)
-{
-/* Centered about data[0] algorithm: #/
-
-long i,j;
-double avel, ave2;
-
-% 52.txt
-52
-
-*ave=O;tvar=0; ave1=ave2=O;
-for(j=0;j<=k;j++)
-cov[j]=0;
-
-for(i=1;i<n;i++)
-{
-rave += (data[i] - data[0]);
-tvar += (data[i] - data[0])#(data[i] - data[0]);
-}
-tvar /= (doub1e)(n-1);
-Ivar -= (((*ave)/(doub1e)n) * ‘(*ave)/(doub1e)(n-1)));
-/# *ave = ((*ave)/(doub1e)n) + data[0]; <- that’s the true average (not needed)*/
-
-for(j=0;j<=k;j++)
-{
-
-for(i=0;i<(n-j);i++)
-
-{
-
-avel += (data[i] - data[0]);
-ave2 += (data[i+j] - data[0]);
-cov[j] += (data[i] - data[0])*(data[i+j] — data[0]);
-
-}
-
-avel/=(doub1e)(n-j); ave2/=(doub1e)(n-j);
-
-cov[j] = ((cov[j] - (double)(n-j)tave1¢ave2)/(double)(n-j-1));
-}
-
-}
-
-/trt end funccions.c ###/

### 2013-10-18 tex/LanczosProcedures.tex
initial commit of tex source for nyu thesis style setup

--- 2013-10-18 tex/MSThesis.tex
@@
-%\documentclass[11pt]{amsart}
-\documentclass{au}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% For completion by editorial office:
- \presentedby{\dots}
- \received{\dots}{\dots}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\usepackage[colorlinks=true,urlcolor=black,linkcolor=black,citecolor=black]{hyperref}
-
-\usepackage{amsmath,amssymb,url,mathrsfs}
-\usepackage{enumerate}
-\usepackage{tikz}
-\usetikzlibrary{calc}
-\usepackage{scalefnt}
-\numberwithin{equation}{section}
-
-\theoremstyle{plain}
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{proposition}[theorem]{Proposition}
-\newtheorem{corollary}[theorem]{Corollary}
-
-\theoremstyle{definition}
-\newtheorem{definition}[theorem]{Definition}
-\newtheorem{notation}[theorem]{Notation}
-\newtheorem{remark}[theorem]{Remark}
-\newtheorem{example}[theorem]{Example}
-\newtheorem*{remarks}{Remarks}
-
-\newcommand{\todo}[1]{~\\[5pt]{\bf TODO(wjd):} #1\\[5pt]}
-%\newcommand{\suchthat}{\ensuremath{:}}    % (old notation)
-\newcommand{\suchthat}{\ensuremath{\mid}}  % (new notation)
-%\newcommand{\defeq}{\ensuremath{:=}}      % (old notation)
-\newcommand{\defeq}{\ensuremath{=}}        % (new notation)
-\newcommand{\betai}{\ensuremath{\beta^i}}  % \beta^{\bB_i}   % (old notation)
-\newcommand{\betaj}{\ensuremath{\beta^j}}  % \beta^{\bB_j}
-\newcommand{\betak}{\ensuremath{\beta^k}}  % \beta^{\bB_k} or \beta^{\bB_{k}}
-\newcommand{\betal}{\ensuremath{\beta^\ell}}  %\beta^{\bB_\ell} or \beta^{\bB_{\ell}}
-\newcommand{\uacalc}{{\small UAC}alc}
-\newcommand{\Tuma}{T\r{u}ma}
-\newcommand{\Palfy}{P\'alfy}
-\newcommand{\Pudlak}{Pudl\'ak}
-\newcommand{\PP}{P\'alfy-Pudl\'ak}
-\newcommand{\PAP}{P\'alfy\ and Pudl\'ak}
-\newcommand{\<}{\ensuremath{\langle}}
-\renewcommand{\>}{\ensuremath{\rangle}}
-\newcommand{\beps}{\ensuremath{\boldsymbol{\varepsilon}}}
-\newcommand{\bN}{\ensuremath{\mathbf{N}}}
-\newcommand{\bA}{\ensuremath{\mathbf{A}}}
-\newcommand{\bB}{\ensuremath{\mathbf{B}}}
-\newcommand{\bBi}{\ensuremath{\mathbf{B}_i}}
-\newcommand{\sB}{\ensuremath{\mathcal{B}}}
-\newcommand{\sC}{\ensuremath{\mathcal{C}}}
-\newcommand{\sS}{\ensuremath{\mathscr{S}}}
-\newcommand{\sT}{\ensuremath{\mathscr{T}}}
-\newcommand{\sI}{\ensuremath{\mathcal{I}}}
-\newcommand{\sE}{\ensuremath{\mathcal{E}}}
-\newcommand{\sO}{\ensuremath{\mathcal{O}}}
-%\newcommand{\id}{\ensuremath{\mathrm{id}}}
-\DeclareMathOperator{\id}{id}
-%\newcommand{\Eq}{\ensuremath{\mathrm{Eq}}}
-\DeclareMathOperator{\Eq}{Eq}
-%\newcommand{\Cg}{\ensuremath{\mathrm{Cg}}}
-\DeclareMathOperator{\Cg}{Cg}
-%\newcommand{\Con}{\ensuremath{\mathrm{Con\,}}}
-\DeclareMathOperator{\Con}{Con}
-%\newcommand{\Sub}{\ensuremath{\mathrm{Sub}}}
-\DeclareMathOperator{\Sub}{Sub}
-%\newcommand{\Pol}{\ensuremath{\mathrm{Pol}}}
-\DeclareMathOperator{\Pol}{Pol}
-%\newcommand{\Clo}{\ensuremath{\mathrm{Clo}}}
-\DeclareMathOperator{\Clo}{Clo}
-\newcommand{\ps}[1]{\ensuremath{^{(#1)}}}
-\newcommand{\piB}{\ensuremath{\pi_B}}
-\newcommand{\hpsi}{\ensuremath{\widehat{\psi}}}
-\newcommand{\htheta}{\ensuremath{\widehat{\theta}}}
-\newcommand{\supi}{\ensuremath{^{i}}}
-\newcommand{\supj}{\ensuremath{^{j}}}
-\renewcommand{\leq}{\ensuremath{\leqslant}}
-\renewcommand{\nleq}{\ensuremath{\nleqslant}}
-\renewcommand{\geq}{\ensuremath{\geqslant}}
-\newcommand{\meet}{\ensuremath{\wedge}}
-\newcommand{\join}{\ensuremath{\vee}}
-\newcommand{\Meet}{\ensuremath{\bigwedge}}
-\renewcommand{\Join}{\ensuremath{\bigvee}}
-\newcommand{\resB}{\ensuremath{|_{_B}}}
-\newcommand{\resBi}{\ensuremath{|_{_{B_i}}}}
-\newcommand{\eps}{\ensuremath{\varepsilon}}
-\newcommand{\hatmap}{\ensuremath{\widehat{\phantom{x}}}}
-\newcommand{\one}{\ensuremath{\mathbf{1}}}
-\newcommand{\two}{\ensuremath{\mathbf{2}}}
-\newcommand{\three}{\ensuremath{\mathbf{3}}}
-\newcommand{\tbeta}{\ensuremath{\widetilde{\beta}}}
-\newcommand{\ttheta}{\ensuremath{\widetilde{\theta}}}
-\newcommand{\hbeta}{\ensuremath{\widehat{\beta}}}
-\newcommand{\cick}{\ensuremath{\sC_i}}
-%\newcommand{\CE}{\ensuremath{\sC_r^{\sE}}}
-\newcommand{\CE}{\ensuremath{\sB_r^n}}
-\newcommand{\CO}{\ensuremath{\sC_r^{\sO}}}
-\newcommand{\CICK}{\ensuremath{c_i/\beta \cup c^{K+1}_i/\beta^{\bB_{K+1}}}}
-\newcommand{\GAP}{{\small GAP}}
-\newcommand{\dotsize}{.8pt}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{document}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%  FRONT MATTER                                                 %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\title[Approximating Eigenvalues of Large Stochastic Matrices]{A Lanczos Procedure for Approximating Eigenvalues of Large Stochastic Matrices}
-
-%% First author (Note: The order of the items here is important!)
-\author{William DeMeo}
-\email{williamdemeo@gmail.com}
-\urladdr{http://www.math.sc.edu/~demeow}
-\address{Department of Mathematics\\
-University of South Carolina\\Columbia 29208\\USA}
-
-\thanks{The author wishes to thank a number of people who contributed to this
-  work.  In particular, the primary referee made many insightful recommendations
-  which led vast improvements, not only in the overall exposition, but also in
-  the specific algebraic constructions, theorem statements, and proofs.
-  This article is dedicated to Ralph Freese and Bill Lampe, who declined to be
-  named as co-authors despite their substantial contributions to the project.}
-
-%% Dedication (Optional)
-\dedicatory{This article is dedicated to Ralph Freese and Bill
-  Lampe.}
-
-%% AMS subject classification; see http://www.ams.org/msc
-%% Only one Primary. Possibly several Secondary.
-\subjclass[2010]{Primary: 08A30; Secondary: 08A60, 06B10.} %, 06B15.}
-
-%% Keywords and phrases
-\keywords{congruence lattice, finite algebra, finite lattice
-  representations}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%  MAIN MATTER                                                  %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{abstract}
-The rate at which a Markov chain converges to a given probability distribution
-has long been an active area of research. Well known bounds on this rate of
-convergence involve the subdominant eigenvalue of the chain’s underlying
-transition probability matrix. However, many transition probability matrices are
-so large that we are unable to store even a vector of the matrix in fast
-computer memory. Thus, traditional methods for approximating eigenvalues are
-rendered useless.
-
-In this paper we demonstrate that, if the Markov chain is reversible, and we
-understand the structure of the chain, we can derive the coefficients of the
-traditional Lanczos algorithm without storing a single vector. We do this by
-considering the variational properties of observables on the chain’s state
-space. In the process we present the classical theory which relates the
-information contained in the Lanczos coefficients to the eigenvalues of the
-Markov chain. 
-\end{abstract}
-
-\maketitle
-
-%%%%%%%%%%%%%%%%%%%%%
-\chapter{Introduction}
-The rate at which a Markov chain converges to a given probability distribution has long
-been an active area of research. This is not surprising considering this problem’s relevance to the ar-
-eas of statistics, statistical mechanics, and computer science. Markov Chain Monte Carlo (MCMC)
-algorithms provide important examples. These algorithms come in handy when we encounter a
-complicated probability distribution from which we want to draw random samples. In statistical
-mechanics, we might wish to estimate the phase average of a function on the state space. Goodman
-and Sokal [6] examine Monte Carlo methods in this context. Examples from statistics occur in the
-Bayesian paradigm when we are forced to simulate an unwieldy posterior distribution (see, e.g.,
-Geman and Geman
-
-To implement the MCMC algorithm, we invent a Markov chain that converges to the
-desired distribution (this is often accomplished using the Metropolis algorithm
-described in Chapter 5). Realizations of the chain will eventually represent
-samples from this distribution. Sometimes ``eventually'' -- meaning all but
-finitely many terms of the chain -- is just not enough. We need more practical
-results. In particular, we want to know how many terms of the chain should be 
-discarded before we are sampling from a distribution that is close (in total variation distance) to
-% 03.txt
-the distribution of interest. This is the purpose of bounding convergence rates for Markov chains.
-
-Often the Markov chains encountered in this context satisfy a condition known in the
-physics literature as detailed balance. Probabilists call chains with this property reversible. This
-simply means that the chain has the same probability law whether time moves
-forward or 
-backward.\footnote{This is not a precise definition. In particular the chain must
-  have started from its stationary distribution. Full rigor is postponed until Section 2.1.}
-In this paper, we consider the rate at which such chains converge to a \emph{stationary distribution}.\footnote{This and other italicized terms are defined in Section 2.1.}
-
-There are a number of different methods in common use for bounding convergence rates of
-Markov chains, and a good review of these methods with many references can be found in  More
-recently developed methods, employing logarithmic Sobolev inequalities, are reviewed in  Most
-of the bounds in common use involve the sub—dominant eigenvalue of the Markov chain's transition
-probability matrix, and thus require good approximations to such eigenvalues. In many applications,
-however, the transition probability matrix is so large that it becomes impossible to store even a
-single vector of the matrix in conventional computer memory. These so called out-of-core problems
-are not amenable to traditional eigenvalue algorithms\footnote{By ``traditional
-  eigenvalue algorithms'' we refer to those found, for example, in Golub and Van
-  Loan[5]. See also the book by Demmel [1] for a more recent discussion.}
-without modification. This paper develops 
-such a modification for the Markov chain eigenvalue problem. In particular it develops a method
-for approximating the first few eigenvalues of a transition probability matrix when we know the
-general structure of the underlying Markov chain. The method does not require storage of large
-matrices or vectors. Instead we need only simulate the Markov chain, and conduct a statistical
-analysis of the simulation.
-
-Here is a look at what follows. Section 2.1 contains a review of the relevant Markov chain
-theory. Readers conversant in the asymptotic theory of Markov chains might wish to at least skim
-Section 2.1, if only to become familiar with our notation. Section 2.2 describes functions on the state
-% 04.txt
-space of the Markov process. This section and Chapter 3 develop the context in which we formulate
-the new ideas of the paper. In the last section of Chapter 3, Section 3.3, we present the familiar
-Krylov subspace and explain why this represents our best approximation to a subspace containing
-extremal eigenvectors of the transition probability matrix.\footnote{or, more
-  precisely, a similarity transformation of this matrix.} The first section of
-Chapter 4 describes the \emph{Lanczos algorithm} for generating an orthonormal basis
-for the Krylov subspace. As it stands, this algorithm is useless for an
-out-of-core problem such as ours since, by definition of such problems, 
-it requires too much data movement; all the computing time is spent swapping data between slow
-and fast memory (e.g. between the hard disk and cache). Therefore, we discuss alternatives to
-Lanczos and demonstrate that the \emph{Lanczos coeficients} are readily available through simulations of
-the Markov chain, which fact allows us to avoid the standard algorithm altogether. Following this
-is a chapter describing the Metropolis algorithm used to produce a reversible stochastic matrix.
-It is here that we experiment with the procedure described in Section 4.2 and approximate the
-extremal eigenvalues of the matrix, without storing any of its vectors. Finally, Chapter 6 concludes
-the paper.
-% 05.txt
-
-\chapter{Markov Chains}
-
-\section{General Theory}
-
-This review of Markov chain theory can be found in any good probability text. The present
-discussion is most similar to that of Durrett [3], to which we refer the reader desiring greater detail.
-
-\subsection{The Basic Setup}
-
-Heuristically, a Markov chain is a stochastic process with a lack of memory property. Here
-this means that the future of the process, given its past behavior and its present state, depends only
-on its present state. This is the probabilistic analogue of a familiar property of classical particle
-systems. Given the position and velocities of all particles at time t, the equations of motion can be
-completely solved for the future evolution of the system. Thus, information describing the behavior
-of the process prior to time t is superﬂuous. To be a bit more precise, if technical, we need the
-following definitions.
-
-\bibliographystyle{spmpsci}
-\bibliography{wjd}
-
-\end{document}
-
-%%%%%%%%%%%%%%%%%%%%%  END DOCUMENT %%%%%%%%%%%%%%%%%%%%%%%%%
-
-Definition 2.1.1 Let (S, 8) be a measurable space. A sequence X,., n 2 0, of random variables
-
-taking values in S is said to be a Markov chain with respect to the filtration a(Xo, . . . ,X,,) if for
-% 06.txt
-a.llBES,
-
-p(x,,+1 e B | a(Xo, . . . ,X,.)) = P(X,.+1 e B 1 o(X,,)) (2.1)
-
-Equation (2.1) merely states that, given the present location of X,,, the past is irrelevant for
-
-predicting the location of X,,+1.
-
-Definition 2.1.2 A function p : S x S —> R is said to be a transition probability if:
-1. for each :1: E S, A —> p(a:, A) is a probability measure on (S, S).
-2. for each A E 8, :1: —> p(z, A) is a measurable function.
-
-We call X,, a Markov chain with transition probabilities 12,, if
-P(Xn+l E B l a(Xn)) =Pn(XnaB) (22)
-
-The spaces (S, S) that we encounter below are standard Borel spaces, so the existence of the
-transition probabilities follows from the existence of regular conditional probabilities on Borel spaces
-— a standard measure theory result (see e.g. [3] page 230).
-
-Suppose we are given an initial probability distribution p on (S, S) and a sequence p,, of
-
-transition probabilities. We can define a consistent set of finite dimensional distributions by
-
-P<X,- e B,-,0 s 1 s n) = / #(d1‘o) / Po($o.d-'51) . .. / Pn-l(97n—1ad$n) (2.3)
-B0 B1 B1:
-Furthermore, denote our probability space by
-(9,?) = (S{0,1,...},8{0,l,...})
-
-We call this sequence space and it is defined more explicitly by
-
-sf°»1=-~-} = {(wo,w1,...) :w.- e S}
-
-S{°*1""} = ¢7(w :40; E A,- E 3)
-
-% 07.txt
-The Markov chain that we will study on this space is simply X" (w) = w,., the coordinate maps.
-Then, by the Kolmogorov extension theorem, there exists a unique probability measure P,, on
-(9,?) so that the X" (w) have finite dimensional distributions (2.3).
-
-If instead of p, we begin with the initial distribution 6,, i.e. point mass at 2:, then we
-denote the probability measure by P3. With such measures defined for each 2:, we can in turn
-
-define distributions P,,, given any initial distribution p, by
-
-PM = / u<da.->P.(A>
-
-That the foregoing construction — which, recall, was derived merely from an initial distri-
-bution p and a sequence p,, of transition probabilities — satisfies Definition (2.2) of a Markov chain
-is not obvious, and a proof can be found in
-
-To state the converse of the foregoing, if X,, is a Markov chain with transition probabilities
-p,, and initial distribution p, then its finite dimensional distributions are given by (2.3). Proof of
-this is also found in
-
-Now that we have put the theory on a firm, if abstract, foundation, we can bring the
-discussion down to earth by making the forgoing a little more concrete. First, we specialize our
-study of Markov chains by assuming that our chain is temporally homogeneous, which means that
-the transition probabilities do not depend on time; i.e. p,,(w,,, B) E p(w,., B). (This is the stochastic
-analogue of a conservative system.) Next we assume that our state space S is finite, and suppose
-for all states 1', j E S that p(i, j) Z 0, and E, p(z', j) = 1 for all i. In this case, equation (2.2) takes
-
-a more intuitive form:
-P(Xn+l =j I X7: = 7:) =P(iaj)
-
-and our transition probabilities become
-% 08.txt
-If P is a matrix whose (i, j)th element is the transition probability p(i, 1') then P is a stochastic
-
-matriar, that is, a matrix with elements pgj satisfying
-
-Pij 20: zjpij =1: (i7j=1121"‘)d)
-
-We also refer to P as the transition probability matrix.
-
-Without loss of generality, we can further suppose our Markov chain is irreducible. This
-means that, for any states 2', 1‘, starting in state i the chain will make a transition to state 1' at
-some future time with positive probability. This state of affairs is often described by saying that
-all states communicate. We lose no generality with this assumption because any reducible Markov
-chain can be factored into irreducible classes of states which can each be studied separately.
-
-The final two conditions we place on the Markov chains considered below will cost us
-some generality. Nonetheless, there remain many examples of chains meeting these conditions,
-thus making the present study worthwhile. Furthermore, it may be the case that, with a little
-more work, we will be able to drop these conditions in future studies. The first condition is that
-the chain is aperiodic. If we let I, = {n 2 1 : p"(:c,a:) > 0}, we call a Markov chain aperiodic if,
-for any state 2:, the greatest common divisor of I, is 1. The second assumption is that our chain is
-
-reversible. This characterization is understood in terms of the following definition.
-Definition 2.1.3 A measure a is called reversible if it satisfies
-
-;4(w)p(I. y) = u(y)p(y. w)
-
-for all :c and y.
-
-We call a Markov chain reversible if its stationary distribution (defined in Section 2.1.2) is reversible.
-% 09.txt
-\subsection{A Convergence Theorem}
-
-In succeeding arguments, we use some results concerning the asymptotic behavior of
-
-Markov diains. These results require a few more definitions.
-
-Definition 2.1.4 A measure 1r is said to be a stationary measure if
-
-Z«(z)p<z.y> = we) (2.4)
-
-2
-Equation 2.4 says P,,(X1 = y) = 1r(y), and by induction that P,,(X,, = y) = 7r(y) for all n 2 1. If 1r
-is a probability measure, then we call 1r a stationary distribution. It represents an equilibrium for
-the chain in the sense that, if X0 has distribution 1r, then so does X,. for all 11..
-
-When the Markov chain is irreducible and aperiodic, the distribution of the state at time
-n converges pointwise to 1r as n —> oo, regardless of the initial state. It is convenient to state
-this convergence result in terms of the Markov chain's transition probability matrix P. Before
-doing so, we note that irreducibility of a Markov chain is equivalent to irreducibility (in the usual
-matrix theory sense) of its transition probability matrix. Furthermore, it turns out that a transition
-probability matrix of an aperiodic Markov chain falls into that class of matrices often called acyclic,
-but for simplicity we will call such stochastic matrices aperiodic. With this terminology, we can
-
-state the convergence theorem in terms of the transition probability matrix P.
-
-% Theorem 2.1.1
-\begin{theorem}
-\label{thm-2.1.1}
-Suppose P is irreducible, aperiodic, and has stationary distribution 7r. Then as n —> oo,
-p"(i,J') -> #0")
-\end{theorem}
-
-The symbol p"(i, j) is to be understood as the (ij) element of the nth power of P.
-A Markov chain whose transition probability matrix satisfies the hypotheses of
-Theorem~\ref{thm-2.1.1} is called ergodic. If we simulate an ergodic chain for
-sufficiently many steps, having 
-% 10.txt
-begun in any initial state, the final state is a sample point from a distribution which is close to 1r.
-
-To make this statement more precise requires that we define “close”.
-
-Definition 2.1.5 Let 1r be a probability measure on S. Then the total variation distance at time
-
-n with initial state I is given by
-AA") = l|P"(z, A) - 7r(-4)||Tv = 1333; I P"(I.A) - 7r(A) I
-
-In what follows, we will measure rate of convergence using the function 'r,(e), the first time after
-
-which the total variation distance is always less than 5. That is,
--r,,(e) = min{m : A,(n) 3 E for all n. 2 m}
-
-To consider the connection between convergence rates and eigenvalues, note that an ape-
-riodic stochastic matrix P has real eigenvalues 1 = /\o > A1 2 A2 3  Z /\¢_1 2 -1, where
-d = |S'] is the dimension of the state space. For an ergodic chain, the rate of convergence to the
-stationary distribution 7r is bounded by a function of the subdominant eigenvalue. By subdominant
-eigenvalue we mean that eigenvalue which is second-largest in absolute value, and we denote this
-Am“ E max{).1,|A¢_1|}. The function bounding the rate of convergence is given by the following
-
-theorem:
-
-% Theorem 2.1.2
-\begin{theorem}
-\label{thm-2.1.2}
-The quantity T1(€) satisfies
-1. 13(5) 5 (1 — /\m,_x)“(ln1r(a:)‘1 +1ne‘1)
-2. maxxes r,(e) 2 %/\,,,,x(1 — ).,,,,x)‘1ln(2e)’1
-\end{theorem}
-
-As this theorem shows, if we have an upper bound on the subdominant eigenvalue, then we have
-an upper bound on the function 11(5). In what follows, we will derive an approximation to the
-% 11.txt
-subdominant eigenvalue and supply error bounds. Together, an approximation and error bounds
-for Am“ provide enough information to make Theorem~\ref{thm-2.1.2} useful.
-
-\section{Functions on the State Space}
-
-Recall that X,.(w) = w,. 6 S denotes the state in which the Markov chain exists at time
-11. Suppose that 11> = {¢1, . . . , ¢,,} is a collection of p observables, or functions defined on the state
-space S. Furthermore, let these observables be real valued, so ¢.~ : S —> R. It is often useful, as we
-will see below, to assume that none of the observables is a constant function. Suppose now that
-the state space S is finite with at possible states. Then, since an observable is simply a map of the
-state space, we can think of each ¢,- as a vector of d real numbers — the d values that it takes on at
-the different states.
-
-Now assume the Markov chain is irreducible, and let 1r denote its stationary distribution.
-If we start the chain from its stationary distribution - i.e. suppose X0 has distribution 1r - then
-X,. is a stationary process. Furthermore, for each i, ¢,(X,.) is a stationary stochastic process with
-
-mean
-
-E.¢.- = Z) vr(x)¢.(z)
-
-165
-
-and autocovariance function
-
-Cir (¢i(Xn)v ¢i (Xn+a))
-
-E7!’  ‘ E1r¢i) (¢i(Xn+a) " E1r¢i)l
-
-Z Pn(Xn = z1X7I+S = y)  — E1r¢i)  _ E1r¢i)
-
-z,yES
-(2.5)
-
-which, by the definition of conditional probability, we can write as,
-
-2 P«(X.. = z)P.(X..+.. = y I X. = 1)(¢s($) — Em) (my) ~ Em)
-1.1165
-% 12.txt
-Z 7r($)P1y (MI) - Em.-) (¢:(1I) - E1r¢i)
-
-:c,y€S
-
-Here pgy denotes the element in row :1: and column y of P’, the sth power of the transition probability
-
-matrix. Similarly, we define the cross-covariance between the function ¢,- at time n and q5,- at time
-
-n+sas
-
-E1!’  _ E7r¢:') (¢i(Xn+a) " E1r¢i)l (2-6)
-
-C1r(¢i(Xn)a (#1 (Xn+a))
-
-Z) vr(x)p;,, (¢.-(w) — Em.-) (my) — Em)
-
-z,yES
-
-Now let (<I>) denote the matrix of mean vectors whose jth column is E,,¢,-1, where 1 = (1, . . . , 1)‘,
-and let H = diag(1r(w1), . . . , 1r(wd)), be the d X d diagonal matrix with stationary probabilities 1r(w)
-on the main diagonal, and zeros elsewhere. Finally, denoting by C(s) the p X p covariance matrix
-
-whose (i,j)th element is C, (qfi,-(X,.),¢_,< (X,.+,)), we have
-
-0(0) = E(‘1’(Xn) - (‘I>))(‘I’(Xn) - (‘PW
-= (‘P-(‘1’))‘1'1(<1’-(‘I>))
-0(8) = E(<I>(Xn) — (‘I>))(‘1’(Xn+a) - (‘1’))’
-
-(‘P - (‘1’))'1'IP‘(<1> — (<1>))
-
-Below, we will also find it useful to have at our disposal a new matrix that is similar to the
-transition probability matrix. We have in mind the matrix M E Ill/2PI'I‘1/2. As is easily verified,
-this allows us to write the covariance matrix as
-
-can = <<I>—<<I>>)°n%*M*n%<<I>—<<I>>>
-
-\II‘M’\II (2.7)
-
-where we have defined ‘II E 111/ 2(<I> —  Recall that our main motivation is that, for out-of-core
-problems, traditional eigenvalue algorithms are inadequate. With this fact, and the form (2.7) in
-% 13.txt
-mind, we will consider using the covariance of observables on the state space to implement the
-(otherwise useless) Rayleigh-Ritz procedure, which we describe below. This procedure requires
-that M be symmetric. As the next fact demonstrates, this required symmetry is the reason for our
-
-interest in the reversibility property of the Markov chain.
-
-Fact 2.2.1 The matrix M is symmetric if and only if the Markov chain is reversible (i.e. iﬂ the
-
-process satisfies the detailed balance condition).
-
-Proof:
-
-M is symmetric
-
-4:, (1-I1/2P1-I-1/2): = H1/2Pn—1/2
-<.—> r1-%*P°n%* = H1/2911-1/2
-¢> 1>'r1= = up
-~:=> (IIP)‘ = l'IP
-
-Element-wise, the final equality implies that 7|‘,'p,'j = 1rjpj.-. According to Definition 2.1.3, this states
-
-that p is a reversible measure.
-
-% 14.txt
-
-\chapter{Invariant and Approximate Invariant Subspaces}
-
-\section{Invariant Subspaces}
-Definition 3.1.1 A subspace S Q R" with the property that
-
-1:ES'=Ma:€S
-
-is said to be invariant for M.
-
-Recall that, having chosen observables <I> = ((151, . . . , qfip), we constructed the covariance
-matrix C(s) = \II‘M"\II. If the column space of \II, which we denote by ran(\II), is an invariant
-subspace for M, the definition implies Mifij E  That is, for each 111,- there exists a vector
-t of coeﬂicients such that M1/Jj = f=1t,-1/)5,-. This is true for all j and, putting each vector of
-coefficients into a matrix T, we see that M\II = \IIT. Conversely, M\II = IIIT implies that M1/Jj is a
-
-linear combination of columns of \II, so ran(\II) is invariant. We have thus proved the following
-% 15.txt
-Fact 3.1.1 The subspace ran(\II) is invariant for M if and only if there exists T E R”? such that
-
-M\II = \IIT.
-
-Consequently,
-
-Fact 3.1.2 )\(T) Q ).(M)
-
-Proof of 3.1.2:
-
-z\E)\(T) _=_ 3v€R"3Tv=).v
-<=> \IITv = A\IIv
-<=> M\II-u = /\\IIv
-< /\ E /\(M)
-
-The second ¢> follows from fact 3.1.1.
-Facts 3.1.1 and 3.1.2 are theoretically useful. To see why, write the equation of fact 3.1.1
-
-as follows:
-
-\IIT = M\II
-\II‘\IIT = \II‘M\II
-T = (\1:'~1r)-‘\1:'M\1:
-
-In this form, we recognize that T = C‘1(0)C(1). That is, using only the covariance of observables
-on the state space, we can generate a matrix T with the property A(T) (_Z ).(M). Recalling that
-M = H1/2Pl'I‘1/2, we see that M is similar to our original stochastic matrix P, and thus ).(M) =
-
-).(P).
-
-\section{Approximate Invariant Subspace}
-
-We qualiﬁed the foregoing by stating that the facts are only theoretically useful. This
-
-is because when choosing observables of the state space we aren't sure that they will satisfy the
-% 16.txt
-primary assumption underlying facts 3.1.1 and 3.1.2. Recall the assumption: ran(\II) is an invariant
-subspace for M, where ‘II = H1/2(<I> —  It may well be the case that there is a 1p; 6 ran(\II) such
-that M1/zj ¢ ran(\II), thereby violating the assumption. However, lacking an invariant subspace, we
-
-might hope that our observables at least provide an approximate invariant subspace.
-
-Deﬁnition 3.2.1 If the columns of ‘II E R“? are independent and the residual matrix E E
-
-M\II —— \IIT is small for some T 6 R7”, then ran(\II) deﬁnes an approximate invariant subspace.
-
-To see how an approximate invariant subspace can be useful in approximating eigenvalues of M,
-we extract a theorem from Golub and Van Loan 
-% Theorem 3.2.1
-\begin{theorem}
-\label{thm-3.2.1}  
-Suppose M 6 Rd” and T 6 RP"? are symmetric and let E E MQ — QT, where Q 6 Rd"? satisﬁes
-
-Q‘Q = I. Then there exist p1,... ,;z,, E A(T) and A1,. . . , A,, E A(M) such that
-ll‘k - Akl S \/§|lE"
-
-fork=1,...,p.
-\end{theorem}
-
-If the subspace ran(Q) is an approximate invariant subspace, then the deﬁnition implies that
-there is a choice T rendering the error IIEII2 small, and thus, the eigenvalues of T provide a good
-approximation to those of M.
-
-When considering the preceding ideas, it becomes apparent that their application will
-present new — but hopefully less prohibitive — problems. As these problems summarize the rt of
-the paper, now is a good time to examine them. First, the preceding theorem assumes a matrix Q
-whose columns form an orthonormal basis for the approximate invariant subspace. For our problem,
-derivation of such a Q is tricky, and we must prepare for this. Next, having an approximate invariant
-
-subspace at our disposal merely tells us that there exists some matrix T which makes the error
-% 17.txt
-"EH2 small. We must discover the form of such a T. Moreover, it is natural to seek that T which
-minimizes IIEII2 for a given approximate invariant subspace. Finally, in order to apply these ideas
-to real examples of our eigenvalue problem, we must ﬁnd a practical way to generate the most
-appropriate approximate invariant subspace. We now address each of these issues in the order
-raised.
-
-Recall the matrix I1 E III/2(<I> — (<I>)). The columns of this matrix, though independent
-(by choice of independent observables), are not necessarily orthonormal. However, consider the
-polar decomposition ‘II = QZ where Q‘Q = I and Z2 = \Il‘\II is symmetric positive semideﬁnite.1
-
-Note that Z = (\II‘\II)l/2 is nonsingular, so Q has the form
-Q = ~1:z-1 = \II(\II‘\II)“/2
-
-and it is clear that ran(Q) = ran(\II). Perhaps ra.n(Q) is a. useful approximation to the invariant
-subspace for M. If ran(Q) is not itself invariant, we have the error E = MQ — QT. It is easy to
-show, as we do in the following proof, that the T which minimizes HE“; is T = Q‘MQ. This yields
-
-the following,
-Theorem 3.2.2
-
-If \I1 = QZ is the polar decomposition of \II, then the matrix
-T = Q‘MQ = (\1:*\1:)-1/’\1:‘M\I:(\Ir‘\I:)-1/2
-minimizes IIEII2 = IIMQ - QTII2
-
-Proof:
-We prove the result by establishing the following
-
-(Hal: Given M 6 Rd” suppose Q 6 Rd"? satisﬁes Q'Q =1. Then,
-
-min ||MQ — QTII2 = H(1- QQ')MQll2
-
-TERI‘?
-
-‘Recall that the polar decomposition is derived from the SVD, \II = UZJV‘, by letting Q = UV‘ and Z = VEV‘.
-
-% 18.txt
-and T = Q’MQ is the minimizer.
-To verify the claim, observe the following application of the Pythagorean theorem:
-
-For any T E R.”"” we have
-
-IIMQ — QT||§ = IIMQ - QQ'MQ + QQ'MQ - QT|l§
-||(I - QQ’)MQ + Q(Q'MQ - T) H3
-
-||(1 - QQ‘)MQ||§ + |lQ(Q'MQ - T)||§ (3-1)
-
-2 l|(I — QQ')MQ||§
-
-Equality (3.1) holds since I — QQ' projects MQ onto the subspace orthogonal to ran(Q). Thus,
-the two terms in the expression on the right are orthogonal, and the Pythagorean theorem yields
-equality. The concluding inequality establishes that the minimizing T is that which annihilates the
-second term in (3.1); i.e. T = Q‘MQ.
-
-The second equality in Theorem 3.2.2 is a consequence of the polar decomposition, in
-
-which Q = \IIZ“ = \I:(~I:‘\I:)-1/2. Therefore,
-T = (x1z'x1:)-1/2x1:‘M~1:(\1:*~1z)—1/2
-
-is the minimizer.
-
-3.3 The Krylov Subspace
-
-Suppose the columns of a matrix Q E R“? provide an orthonormal basis for an approxi-
-
-mate invariant subspace. Then we have seen,
-1. T = Q'MQ minimizes ||E||2 = ||MQ — QT||2 and
-2. E|p.1,...,p,, E z\(T) and /\1,...,A, 6 »\(M) such that
-
-ll-‘I: - Ml S \/§|lE||2
-
-% 19.txt
-19
-fork=1,...,p.
-
-Given an approximate invariant subspace ran(Q) of dimension 1), these facts tell us what matrix we
-should use to approximate p elements of M’s spectrum. Now all we lack is a. description of ran(Q).
-That is, we have not speciﬁed which approximate invariant subspace would best suit our objective
-
-of approximating the subdominant eigenvalue A,,.a2:(M). 'I‘o do so requires the following deﬁnition:
-
-Deﬁnition 3.3.1 The Raleigh quotient of a symmetric matrix M and a nonzero vector 3 is
-
-p(2:, M) E (:z:‘Ma:)/(z‘a:).
-
-We will also denote this by p(z) when the context makes clear what matrix is involved.
-
-To ﬁnd the approximate invariant subspace most appropriate for our problem, we choose
-each dimension successively, providing justiﬁcation at each step. Thus, start with one observable
-42 on the state space, and let zpl E 111/ 2(¢ — E,,¢). That is, 1/:1 represents a centered (mean zero)
-observable whose ith coordinate is weighted by . First notice that the deﬁnition (which comes
-directly from the deﬁnition of ‘II following equation (2.7)) is such that 1/)1 is a constant vector if
-and only if :15 is oonstant on the state space, in which case 1/); E 0. This fact provides one reason
-observables that are constant on the state space are not interesting. Second, recall that the row
-sums of the matrix P are all one and, therefore, the eigenvector corresponding to the eigenvalue
-Ao(P) = 1 is the constant vector. By deﬁnition of M E 1'11/2Pl'I‘1/2, we see that the constant vector
-is also the eigenvector corresponding to A0  This will play an important role in what follows, as
-it allows us to focus primarily on the subdominant eigenvalue z\max(M) E max{}q (M), [A44 
-rather than on A0  (which we already know is 1).
-
-Now, notice that
-
-lP(¢1,M)l 5 max |P(x,M)l = »\max(M) (3-2)
-% 20.txt
-where the max is taken over all non-constant vectors. Since our interest centers on An-.3_x(M), we
-would like a. 1/11 which makes the left hand side of (3.2) large. This would be achieved if 1/)1 were to
-lie in the space spanned by, say, the first two eigenvectors of the Markov chain (sometimes referred
-to as the slowest modes of the process). However, this subspace spans only two dimensions of the
-entire d-dimensional space, and it is more likely that ¢1, at best, only comes close to lying in the
-subspace of interest. Now, given 1/11, a judicious choice for the second dimension l/)2, and hence
-W; E [1/)1 1/:2], would be that which makes max,,¢o |p(\II2a)| is as large as possible. To establish that
-
-this is indeed the pertinent objective, note the following:
-
-_ a‘\1:5Mx1:2a
-I;1;g<Ip(\I'2a,M)l = 13:3: ————a,W§q,2a
-= M
-mrggf,mIp(z. )1
-5 maxlp(z,M)l (3-3)
-= ).,,m(M)
-
-Again, the max on the right side of (3.3) is over non-constant vectors. In words, we wish to chose
-\II-2 = [1/:1 ¢2] so that there is a vector a 6 R2 making |p(\II2a,M)| close to )«max(M).
-
-Now, p(1/11) changes most rapidly in the direction of the gradient Vp(¢1).
-
-W0  (iiiiiiv---viiiléi)
-= ,,;,l (Mam — p<¢.)«m (3.4)
-
-So, to maximize the left hand side of (3.3), ‘I12 should be chosen so that the subspace ra.n(\II2)
-
-contains the gradient vector. That is, we must choose 1/22 so that
-
-VP(¢'1) E F3l1{¢1y¢2} E F-'%!1(‘I’2) (3-5)
-
-Clearly, equation (3.4) implies Vp(1/:1) E ran{z/11,M1,b1}. Therefore, if ran{1/:1,1/)2} = ra.n{¢1,M¢1},
-
-then (3.5) is satisﬁed.
-% 21.txt
-In general, having chosen \II,, E [1/)1 - -- 1/11,] so that
-
-ran{'pl9¢21' " 1¢k} = ran{¢1vM1[)la ' ‘ ' vMk—11/’k}
-
-we must chose 1/1;.“ so that, for any non-zero vector u 6 R",
-
-VP(‘I’I:a) 6 ra‘n{¢11 ‘21 - ° - 31/Jk-F1} 
-Now,
-vp<~I~.a) = gmm — P(‘I’kﬂ)‘I’k“)
-
-and therefore,
-
-Vp(‘II;,a.) E ran(M\II;,) U ra.n(\IIk) = ra.n{¢1, M1/)1, . . . , Mktﬁl}
-
-Thus, requirement (3.6) is met when ran{1/:1, 1/:2, . . . , 1/4+1} = ran{1/)1, M1/:1, . . . , Mk:/11}.
-
-In conclusion, the p-dimensional approximate invariant subspace which is most suitable
-
-for our objective is
-
-/C(Mn/mp) 2 ran{«/:1,M¢1,... ,M"“«/21}
-
-This is known as the K rylov subspace. Therefore, in answer to the problem posed at the outset
-of this section, if we take the columns of Q to be an orthonormal basis for IC(M,z/)1, p), then the
-
-eigenvalues of T = Q‘MQ should provide good estimates of p extremal eigenvalues of M.
-
-% 22.txt
 \chapter{Lanczos Procedures}
@@
 30
-
-Chapter 5
-
-An Application: Convergence to
-
-Gibbs Measure on the Ising Lattice
-
-5.1 Physical Motivation
-
-Before considering the details of the Metropolis algorithm, it helps to understand the
-setting in which it was developed. To do so we consider an typical example, the Ising model.
-Consider a model of a system consisting of 71 “sites”, each site taking values in {-1, +1}. It might
-help to imagine these sites as equally spaced points on a line or a circle, though we are not restricted
-to such cases. Then our state space 5' consists of all possible n—dimensional permutations of -1’s
-and +1’s. We reprent this as S’ = {-1, +1}", and note that |S| = 2". We can think ofeach site as
-a node existing in one of two possible positions, say “on" or “off”. Each state 1' E S is represented
-by a unique permutation a.- of —1’s and +1’s. Let a,-(k) = :|:l denote the position (on or off) of
-the kth node when the system is in state 1'. An observable ¢ = qS(a,-) on this space is a function of
-
-the permutations 47,-, 1' 6 {1,. . . , 2"}.
-
-% 31.txt
-31
-
-The energy of the system, when in state i, is deﬁned by the Hamiltonian
-
-H07-‘) = - Z Ua(i)0s(k)
-
-<j Ic>
-where the sum runs over all neart neighbor pairs. In statistical mechanics we are often concerned
-
-with the Gibbs measure of state i, deﬁned by
-1r(a,-) = Z'1 exp{—[3H(a,~)} (5.1)
-where Z ‘1 is the normalization constant, or partition function. For the Ising model, we put
-
-Z = §:9XP{~ﬁH(<7:‘)}
-i=1
-
-so that 7r is a probability measure. Now, letting 1: denote the so called Boltzman constant, if a
-classical mechanical system is in thermal equilibrium with its surroundings with absolute temper-
-ature T, and is in state i with energy H (05), then the probability density in phase-space of the
-point representing state i is given by (5.1) with [3 = (lcT)“. The central result of ergodic theory
-implies that we can also interpret 1r(o;) as the proportion of time the system spends in state i. If
-
-the system is observed at a random time, the expectation E,,¢ of any observable ¢ is thus
-
-.31 ¢(0a) 9"l>{-ﬂH(0s‘)}
-
-E” = 2?;.exp{—nH(a.-)}
-
-(5.2)
-
-Perhaps the number of sites n in our Ising model is so large that it is impractical or impossible to
-evaluate (5.2). We might instead consider importance sampling and generate states with probability
-
-density
-
-exp{—aH<a.o}
-2?;.exp{—zaH<a.->} (53)
-
-Then «I: is itself an unbiased estimator of (5.2).
-
-1r(0a') =
-
-If evaluation of (5.2) is difficult, there is no reason to believe that evaluation of (5.3) is any
-
-easier. However, Metropolis and his collaborators [9] contrived a method for producing an ergodic
-
-% 32.txt
-32
-
-Markov chain with transition probability matrix K whose elements Ic(:::, y) satisfy
-
-2«<z>k<x.y) = «<2»
-
-I
-
-By Deﬁnition 2.1.4 this means that the Markov chain has stationary distribution 7r. Since the
-drain is ergodic, we know that it will eventually converge to the desired distribution. So, our main
-problem — the rate at which such a Markov chain will converge — is clearly at issue in the application
-of the Metropolis algorithm. Moreover, as we will see below, the Metropolis algorithm ensures that
-the Markov chain has the desired stationary distribution by requiring that the chain satisfy the
-stronger condition of reversibility; i.e. detailed balance. Therefore, Markov chains produced by the
-Metropolis algorithm satisfy the assumptions of this paper, which fact conﬁrms that our theory
-
-can be used to bound convergence rates of all such chains.
-
-5.2 Description of the Metropolis Algorithm
-
-The following explains how the Metropolis Algorithm is carried out. The exposition is
-similar to that given in Hammersley and Handscomb’s book Monte Carlo Methods 
-To begin, choose an arbitrary symmetric Markov chain; that is, a Markov chain whose
-
-transition probability matrix P is symmetric. P is called the proposition matrix, and its elements,
-P071"), Satisfy
-pa, 1) 2 0. pm‘) = pm). 2,-pm‘) = 1 vi e S (5-4)
-
-Using P, we will derive a new transition probability matrix K which satisﬁes
-
-Z vr(z')k(='.:') = arm (5.5)
-
-I
-
-For 1' aé j deﬁne
-
-kw) = p<-yum")/«<¢> it nu)/1r(='> <1 (5.6)
-
-P(i,J') if 7r(J')/7r(i) Z1
-
-% 33.txt
-33
-
-For 2' = j deﬁne
-
-kw) = pm) (1—  p(-3;"), where J. = {j = 2‘ 7‘ an nu)/«(='> < 1}
-
-Notice that 1r(i) > 0 implies k(i,j) _>_ 0 and
-
-;k<«',j> = pow)  { (1 —  pm) +  +]§§p<«",:')
-= P(is'5) + 2 P(i1.1') + Z P(iy.7') (5-7)
-jel.» jug
-
-Equation (5.7) shows that, for each i, 2, k(z', j) = 1, which fact conﬁrms that K is a stochastic
-
-matrix.
-
-Next we check that K satisﬁes the detailed balance condition. First suppose that 1r(z') =
-
-1r(j). Then (5.4), (5.6), and symmetry of P imply
-k(i,J') = p(i2.7') = p(jvi) = k(j1i)
-
-and, since 1r(i) = 1r(j), the detailed balance condition, 1r(i)k(i,j) = 7r(j)k(i,j), clearly holds. Now
-
-suppose that 1r(i) > 1r(j). Then (5.4) and (5.6) imply k(j,i) =p(j, 1') and
-k(i,J') = P(iyJ')7r(J')/7r(i) = P(J3i)7r(J')/7r(i) = k(.1'» i)7r(J')/1r(i)
-
-That is, 1r(i)lc(i,j) = 1r(j)k(i,j), as we set out to prove.
-To complete the justification of the Metropolis algorithm, we note that the detailed balance
-condition, 7r(i)k(i, j) = 1r(j)k(i, j), implies that 1r is indeed the stationary distribution for a Markov
-
-chain with transition probability matrix K. To see this notice that
-
-27r(i)k(i..7') = Zﬂijlkiiajl = 1r(J') Z/€(i,J‘) = 7r(J')
-
-I i
-We now summarize the tasks performed when implementing the Metropolis algorithm.
-
-First, pick as a proposition matrix any symmetric stochastic matrix and denote it by P. Begin
-
-% 34.txt
-34
-
-the simulation of the Markov chain in state i, and then take one step of the proposition Markov
-chain to arrive in state j. That is, choose j according to the mass function p(z', -). Next, compute
-1r(j)/1r(i). If 7r(j)/1r(i) 2 1, accept j as the new state. If1r(j)/1r(i') < 1, then with probability
-1r(j)/7r(z') accept j as the new state; otherwise, with probability 1 — 1r(j)/1r(i), take i as the new
-state. Performing each of these tasks produces one step of the Markov chain represented by the
-
-transition probability matrix K.
-
-5.3 The Ising Model
-
-Returning to the Ising model considered in section 5.1, we consider how the Metropolis
-algorithm is applied to such a mode]. Again, start the simulated chain in state X0 = 1' and let
-
-X1 = j be a state chosen according to p(i, -), where p(i, j) = p(j,i). Next, compute
-
-7"(‘7j)/7"(‘7i) = eXP{-ﬁlH(<7j) - H(0i)l}
-
-Suppose the energy H (Uj) of the new state satisﬁes H (U,-) 3 H (05). Then 1r(a_,-)/1r(a,-) Z 1 and we
-
-move to state j. On the other hand, if H(aj) > H(a;), then 1r(aj)/7r(a,-) < 1 and we take
-
-j with probability e-M"
-
-X1 (5.8)
-
-i with probability 1 — e-W’
-where AH E H (0,) — H (:75) is the change in energy from state j to state 1'.
-There are many ways to choose a proposition matrix P. The simplest derives from the
-so called Glauber dynamics, which proceeds as follows: choose one of the 11. sites at random with
-probability 1/n — let .9 denote the chosen site — and change the position of this site; i.e., multiply
-
-its value by -1. The resulting state j is the permutation of -1’s and +1’s given by
-
--0,-(k) if]: = .9
-a,~(Ic)= (k=1,...,d)
-
-a,-(Ic) if]: 95 s
-
-% 35.txt
-35
-
-The procedure we employ below involves a slight variation: choose site .9 at random uniformly
-from the 11 possible sites. However, instead of changing the value of the chosen site with absolute
-certainty, as in Glauber dynamics, we “ﬂip the switch” with probability 1/2. Therefore, when k yé .9
-
-we have 0',-(k) = a.~(k), and when k = 3,
-
-—a,~(s) with probability 1/2
-01(3) =
-
-a,-(3) with probability 1/2
-The advantage of this procedure is that the proposition matrix, before altering it with the Metropo-
-lis algorithm, is a bit more interesting than that produced by traditional Glauber dynamics. For,
-it has 1/2 along the main diagonal and, thus, is aperiodic and converges to a stationary distribu-
-tion. Glauber dynamics has period 2, so its proposition matrix does not satisfy the hypotheses of
-
-our convergence theorem (Theorem 2.1.2). Keep in mind, however, that any proposition matrix
-
-modiﬁed by the Metropolis algorithm satisﬁes all of our hypotheses.
-
-5.4 Computations
-
-5.4.1 Computer Programs
-
-Constructing a computer program to simulate realizations of the Markov chain described
-above (i.e. with transition matrix K) is straight forward, and we do so using Matlab and the
-program ising.m which appears in the Appendix.
-
-When performing simulations in the present context, there are a few important aspects
-to consider. First, our program begins with a “hot start", which means that the initial state is
-picked by assigning each site to :l:1 with probability 1/2. Perhaps we should instead have picked
-our initial state from the stationary distribution. On the contrary, this would defeat the whole
-
-purpose of our analysis. Recall that the method has been developed to deal with cases for which
-
-% 36.txt
-36
-
-sampl from the stationary distribution are not quickly attainable. Furthermore, we are trying to
-ﬁgure out how long it will take for observations from the simulated chain to represent samples from
-the stationary distribution. Second, the method we have developed depends solely on covariances
-of our observable which we estimate from the data. produced by the simulations. However, such
-estimates require that the data come from the stationary distribution. Therefore, we must discard a
-number of observations before using the data to estimate covariances. The number of observations
-to discard depends on how long must we wait until we have data from the stationary distribution,
-and we are back where we started — almost. When employing our procedure, we can discard a
-conservative (very large) number of observations and perform our analysis, deriving a bound on
-the convergence time. Then, in all future studies, we will know how much data should be discarded
-before samples come (approximately) from the stationary distribution.
-
-In considering what observable to use on the state space described above, perhaps the
-most obvious is simply the number of nodes in the on position (i.e. the number of +l’s). Recalling
-that 0, (k) = i1 denotes the position (on or off) of the kth node when the system is in state j, this
-
-observable is deﬁned by,
-«M = zxca.-ac) = +1). 1 = 1.---.ISl
-I:
-
-where X is the indicator function. The program ising.m simulates the Markov chain, computes the
-values of this observable (as well as the number of 1’s squared) and writes it to a ﬁle called phi . dat.
-The program then proceeds to compute the ﬁrst few Lanczos coefficients using covariances produced
-by Matlab, so that we can check them against the rults we get from the 1anczos.c program,
-which we now describe.
-
-The lanczos .c program (listed in the Appendix along with its dependencies) implements
-the recursive relations described in section 4.2 to compute the Lanczos coefﬁcients by the new
-
-method. It accepts input from the user specifying any observable, any number of iterations, and any
-
-% 37.txt
-37
-
-TABLE 1: ESTIMATES or /\,m(K) = 0.998746 USING ¢1
-
-sim. # /32
-1 0.995107 0.006960 0.995220 0.600243
-2 0.995099 0.005891 0.994863 0.477393
-3 0.994862 0.009671 0.994925 N/ A
-4 0.995321 0.009319 0.995654 N/A
-5 0.994787 0.004872 0.405446 0.994742
-
-number of desired Lauczos coeﬁicients. That is, the observable could have come from a simulation
-of any reversible Markov chain and not just our special Metropolis chain. The output of this
-program is a T matrix containing the maximum number of Lauczos coeﬂicients (up to the number
-requested by the user) before an approximate invariant subspace was reached. The eigenvalues of
-the nonzero part of this matrix, in accordance with the foregoing theory, should well approximate
-
-eigenvalues of the Markov chain’s stochastic matrix.
-
-5.4.2 Results
-
-Table 1 shows the results of 5 simulations each of 100,000 iterations (discarding the ﬁrst
-5,000 observations) for the one dimensional Ising lattice with 10 nodes and [7 = 1. Displayed is
-the estimate of the ﬁrst Lauczos coeﬂicient a1 (which is the eigenvalue estimate after one step; i.e.
-).m,x(T1)), along with its error bound ﬁl. Appearing in the third column is the largest eigenvalue of
-the 2 x 2 matrix T2, A0 (T2), and ﬁg appears in the fourth column when it is was computable. When
-it was not computable, it was probably because an approximate invariant subspace was reached
-at the ﬁrst step, so ﬁg involved expressions which were close to machine epsilon and, thus, could
-not be computed accurately. Table 2 shows the same information when using a second observable,
-¢2 E ¢§ (i.e. the number of +1’s squared).
-
-The true value of the subdominant eigenvalue for this problem is /\max(K) = 0.998746,
-
-and the next largest is A2(K) = 0.993303. These were computed directly from the transition matrix
-
-% 38.txt
-38
-
-TABLE 2: ESTIMATES OF >.,,,,,.,,,(K) = 0.998746 USING z/>2
-
-sim. # B2
-1 0.994262 0.006142 0.994222 N / A
-2 0.994333 0.002974 0.994332 N / A
-3 0.994307 0.008892 0.994374 0.726320
-4 0.994436 0.012884 0.994685 0.361404
-5 0.993789 0.006909 0.993820 0.788710
-
-using the Matlab program tpm.m listed in the Appendix. They should be very accurate, and it
-takes the Matlab routine eig() just over 20 minutes of cpu time on a Sun Ultra to ﬁnd all the
-eigenvalues of K.
-
-As shown by Theorem 4.1.1, the [Ts provide conservative error bounds on the eigenvalue
-estimates from the previous Lanczos step. Unfortunately, for this example, ,6 appears to be too
-conservative as it does not even allow us to bound our estimates away from 1. Furthermore, it
-seems that the estimates generated by the method above fall systematically below the true value of
-the subdominant eigenvalue. In every simulation, an approximate invariant subspace (to machine
-tolerance) was reached after computing at most 2 pairs of Lanczos coefficients. For ¢1 we see that
-it was reached after computing only one pair on simulations 3 and 4, similarly for ¢2 on simulations
-1 and 2. These observations indicate that the chosen observable is close to the slowest mode (or
-eigenfunction) of the process. That is, we should observe fast convergence on an approximate
-invariant space. However, the space found might contain only the second slowest mode of the
-process and, in that case, our eigenvalue estimates would be closer to the third largest eigenvalue,
-instead of the second largest as desired.‘
-
-The matrix T was produced by the program lanczos . c and its eigenvalues were computed
-by Matlab, both operations taking a few seconds. Granted, comparing a few seconds to the 20
-
-minutes it takes Matlab is not justiﬁed since a simple call to the routine eig() is far from the
-
-‘Recall that we explained above why we will never converge to the eigenspace corresponding to the largest eigen-
-value (which is always 1). That would require a (nearly) constant observable function.
-
-% 39.txt
-39
-
-most eﬂicient traditional method for computing only a few eigenvalues of K. However, recall that
-the primary motivation for the new method are those examples where the matrix K is so large
-that we can’t store even a single vector in fast memory. Once we start swapping data to and from
-cheap memory, the time required by traditional algorithms can become years, even lifetimes. Such
-examples are not exceptional. Consider, for instance, the 1000 node problem which produces a.
-
-2“’°° x 21°°° transition matrix.
-
-% 40.txt
-40
-
-Chapter 6
-
-Conclusion
-
-We conclude by noting a few aspects of the special cases to which the foregoing applies.
-We are interested in the subdominant eigenvalue of a large transition probability matrix. The
-
-foregoing can be used to approximate these eigenvalues when the following conditions are satisﬁed:
-o The Markov chain is reversible (detailed balance condition).
-
-0 We can easily simulate realizations of the chain as well as functions (observables) of these
-
-realizations.
-
-When these conditions hold, we can side step the traditional Lanczos algorithm, and get at the
-values of the Lanczos coeﬂicients via the variational properties of observables on the state space.
-There are many ideas in this paper that should be expanded upon in future studies. Per-
-haps most obvious is the desire to ﬁnd analogous techniques for non-reversible (i.e. non-symmetric)
-Markov chains. This is a considerable problem since the Lanczos methods described above are no
-longer applicable. Perhaps the Amoldi algorithm ([1], algorithm 6.9), or the non-symmetric Lanczos
-
-algorithm can also be usefully specialized for stochastic matrices.
-
-% 41.txt
-41
-
-When applying the Lanczos algorithm, we usually choose a single starting vector (in the
-present context, an observable), perform the algorithm for a number of steps, and gather results.
-Then we do the same with a new starting vector and compare the results to those obtained with
-the ﬁrst vector. Proceeding in this way for a number of starting vectors, we can provide statistical
-evidence that we have, indeed, located the eigenvalues of interest. However, when there are two
-observables 451, :15; of particular interest, perhaps we would beneﬁt by considering both observables
-simultaneously. Future studies might develop an algorithm which incorporates both observables,
-
-again via ¢.- = H1/2(¢.' - E,.¢.-). In this case, a Lanczos algorithm would be based on the space
-
-Ic*(M,¢1,¢»2.p) 2 ran{«/»mz»2,M«/»1,M«/:2, . . .,MP—‘¢1.M"—‘«/22}
-
-The Lanczos coeﬁicients would then involve not only the autocovariances C, (¢,<(n),¢,-(n + m)),
-
-but also the cross-covariances C, (¢.~(n), ¢j(n + m)) (deﬁned in (2.6)).
-
-% 42.txt
-Appendix: computer programs
-
-Z Hatlab code ising.m
-
-Z
-Z Written by William J. De Meo on 12/15/97
-
-2 last modified 1/10/98
-
-Z Inputs:
-
-2 d = number of nodes of the ising lattice (e.g. d=100)
-Z n = number of iterations (e.g. n=1000)
-
-Z beta = Annealing schedule (e.g. beta = 2,
-
-1 when beta -> 0 will always go to nee state
-Z vhen beta -> infty will never go to state of higher energy)
-Z
-
-X = zeros(d,1);
-Energy = zeros(n,1);
-
-Ave=0;
-
-X Initialize using hot start
-for i=1:d,
-
-if rand < .5
-
-x(i) = -1;
-
-else
-
-X(i) = 1;
-
-end
-end
-
-Z Initialize at state of high energy (not used)
-
-if 0
-X(1:2:d-1)=1;
-X(2:2:d)=—1;
-and
-
-Z Initialize energy
-
-H=0;
-
-for i=1:(d-1)
-
-H = H - X(i)*X(i+1);
-end
-
-2 Initialize observable
-phi = zeros(n+1,1);
-
-for i=1:d
-if X(i) == 1
-phi(1)= phi(1)+1;
-end
-
-end
-
-42
-
-% 43.txt
-43
-
-for j=1:n
-
-site=0;
-while site==
-
-site = round(rand#d);
-end
-Z Compute new energy (with new X(site) = —x(site)):
-if site ==
-
-Hnew = H + 2:x(1)~x(2);
-elseif site == d
-
-Hnew = H + 2*X(d)tX(d-1);
-else
-
-Hnew = H + 2*X(site)*(X(site-1) + X(site+1));
-end
-
-Z Change to new state in two ways:
-X with probability 1/2, flip the switch
-if rand < .5
-if (Hnew <= H) I (rand < exp(-beta*(Hnew-H)))
-x(site) = -x(site);
-H = Hnew;
-Z Compute new value of observable
-if X(site) -= 1
-phi(j+1) = phi(j) + 1;
-
-else
-phi(j+1) = phi(j) — 1;
-end
-else phi(j+1)=phi(j);
-end
-else phi(j+1)=phi(j);
-end
-Energy(j) = H;
-end
-p1ot(Energy)
-
-% A second observable
-phi2 = zeros(n+1,1);
-for i=1:n+1
-
-phi2(i) = phi(i)*phi(i);
-end
-
-Z Write first observable to phi.dat:
-
-fid = fopen(’/quean.a/masters/deme5025/research/nyuthesis/programs/c/phi.dat’,’w’);
-fprintf(fid,’%f\n’,phi);
-fc1ose(fid);
-
-Z Write second observable to phi2.dat:
-fid = fopen(’/quean.a/masters/deme5025/research/nyuthesis/programs/c/phi2.dat’,’w’);
-
-% 44.txt
-fprintf(fid,’%f\n’,phi2);
-fc1ose(fid);
-
-1 Compute a few Lanczos coefficients by new method
-START=1001;
-
-V1=zeros(2);
-
-C1=zeros(2);
-
-C2=zeros(2);
-
-C3=zeros(2);
-
-V1 = cov(phi(START:n+1),phi(START:n+1));
-
-V1 = V1(1,1)5
-
-C1 = cov(phi(START:n),phi(START+1:n+1));
-C1 = C1(1,2);
-
-C2 = cov(phi(START:n-1),phi(START+2:n+1));
-C2 = C2(1,2);
-
-C3 = cov(phi(START:n-2),phi(START+3:n+1));
-
-C3 = C3(1,2);
-
-a1pha1 = C1/V1;
-
-betai = sqrt(C2/V1 ' (C1/V1)‘2);
-
-a1pha2 = (1/(beta1’2))*(C3/V1 - 2*a1pha1*(C2/V1) + a1pha1‘3);
-
-T1 = Ealphal betal; betal a1pha2]
-
-v1 = cov(phi2(START:n+1),phi2(START:n+1));
-v1 = v1(1,1);
-
-c1 = cov(phi2(START:n),phi2(START+1:n+1));
-c1 = c1(1,2);
-
-c2 = cov(phi2(START:n—1),phi2(START+2:n+1));
-c2 = c2(1,2);
-
-c3 = cov(phi2(START:n-2),phi2(START+3:n+1));
-c3 = c3(1,2);
-
-alphal = C1/V1;
-betal = sqrt(C2/V1 - (C1/V1)‘2);
-a1pha2 = (1/(beta1‘2))#(C3/V1 - 2ta1pha1*(C2/V1) + a1pha1”3);
-
-T2 = [a1pha1 betai; betal a1pha2]
-
-fid = fopen(’Tmats.dat’,’v’);
-fprintf(fid,’%f\n',T1);
-fprintf(fid,’%f\n’,T2);
-fc1ose(fid);
-
-ZZZ end ising.m
-
-44
-% 45.txt
-\% Matlab code tpm.m for generating transition probability matrix
-Z and computing its eigenvalues directly
-
-1
-
-Z Written by Hilliam J. De Heo on 1/10/98
-
-1
-
-Z Inputs:
-
-Z d = number of nodes of the ising lattice (e.g. d=10)
-
-Z beta = Annealing schedule (e.g. beta - 2,
-
-1 when beta -> 0 will always go to new state
-1 when beta -> infty will never go to state of higher energy)
-2
-
-disp(’bui1ding proposition matrix...’)
-
-states = 2‘d;
-
-A = 0;
-
-for i=0:(d-1)
-
-E=eye(2”i);
-
-A = [A E;E A];
-
-end
-
-B = .5#eye(states);
-
-A = B + .5-(1/d)*A;
-
-2 our modified Glauber dynamics requires the B and the .5¥(1/d)
-
-disp(’...done’)
-
-1 display pattern of nonzero entries
-1 spy(A)
-
-Z check that all row sums are 1
-check=O;
-for i=1:states
-
-check = check + (not(sum(A(i,1:states))<.99));
-end
-1 if not all rows sum to 1, print check = (# of rows with sum=1)
-if not(check==states)
-
-check
-
-error(’Row sums are not all 1’)
-end
-
-Z construct matrix of states
-E = zeros(d,states);
-f1ip=-1;
-for i=1:d
-for j=1:2‘(i-1):states
-flip = -1*f1ip;
-for k=0:2“(i-1)
-
-45
-
-% 46.txt
-E(i,j+k) = flip;
-and
-end
-end
-E=E’;
-
-Z display first 64 states
-Z E(1:64,:)
-
-Z compute energy of each state
-H = zeros(states,1);
-for i=1:states
-
-for j=1:d-1
-H(i) = H(i) - E(i,j)tE(i,j+1);
-end
-end
-
-disp(’building Hetropolis transition matrix...')
-
-for i=1:states
-for j=i+1:states
-if not(A(i,j)==0)
-if(H(j)>H(i))
-alt = A(i,j)t(1-exp(-beta*(H(j)-H(i))));
-A(i,i) I A(i,i)+a1t;
-A(i,j) = A(i,j)-alt;
-elseif(H(j)<H(i))
-alt = A(j,i)t(1-exp(—beta#(H(i)-H(j))));
-A(j,j) = A(j,j)+a1t:
-A(j,i) = A(j,i)-alt;
-end
-end
-end
-end
-
-disp(’...done’)
-
-check=0;
-for i=1:states
-
-check = check + (not(sum(A(i,1:states))<.99));
-end
-
-Z if not all rows sum to 1, print check = (# of rows with aum=1)
-
-if not(check==states)
-
-check
-
-error(’Row sums are not all 1’)
-end
-
-46
-
-% 47.txt
-47
-
-disp(’computing eigenvalues of tpm...’)
-cput = cputime;
-
-evals = eig(A);
-
-ecput = tputime - cput;
-
-disp(’the CPU time (in secs) for computing eigenvalues of tpm: ’)
-
-ecput
-ZZZ end tpm.m
-
-/¥#*##¥###*##*#l####¥#‘###$ﬂ###ﬁi##*U3*#**#*##**##¥***ll¥#*##¥ii
-
-* 1anczos.c main program for computing Lanczos coefficients *
-3 #
-* Created by William J. De Heo #
-t on 1/7/98 *
-# Last modified 1/10/98 #
-
-*¥#######**3It¥*#¥¥¥#**##**¥##¥#####**####*#*###*##**###¥##*##*/
-
-#inc1ude <stdlib.h>
-#include <stdio.h>
-#include <math.h>
-#inc1ude "prototypes.h"
-
-/# Machine constants on dino (Sun Ultra 1 at Courant) t/
-#define HACHEPS 1.15828708535533604981e-16
-
-#define SQRTEPS 1.07623746699106147048e-08
-
-ﬂdefine HAX_NAH 100
-
-void read_name(char *);
-
-/* Prototypes from the file functions.c ¥/
-
-double a1pha(int j);
-
-double betasq(int j);
-
-double form1(int j, int k);
-
-double form2(int j, int k);
-
-double moment(doub1e ‘data, long n, double tave, double tvar, double tcov, long k);
-
-/# external variables to be used by functions */
-double *phi,tcov, tvar, tave;
-
-% 48.txt
-main()
-
-{
-
-char tfilename;
-
-FILE *ofp;
-
-double temp=O, *T;
-
-long i, j, nrou, nlanc, START;
-int f1ag=O;
-
-filename = cma11oc(HAx_NAHE);
-
-printf("\nName of file containing observed values: “);
-
-read_name(fi1ename);
-
-printf(“\nTota1 number of observations in file (iterations):
-
-scanf(“Zu”,&nrov);
-
-printf(“\nNumber of leading observations to discard: “);
-
-scanf("Zu",&START);
-
-printf("\nNumber of Lanczos coefficients desired: “);
-
-scanf("%u",&n1anc);
-
-vhi1e(2*n1anc > nrow—START)
-
-{
-printf("\nNot enough data for that many coefficients.\n“);
-printf("\nEnter a smaller number of Lanczos coefficients: “);
-scanf("Xu“,&n1anc);
-
-}
-
-II);
-
-phi = dma11oc(nrow);
-cov=dma11oc(2#n1anc);
-var=dma11oc(1);
-ave=dma11oc(1);
-
-T = dma11oc(n1anc#n1anc); for(i=0;i<n1anc;i++) T[i]=(doub1e)0;
-
-/* observable phi is stored contiguously column-wise (e.g. by HATLAB) t/
-mat1abread(phi, nrow, 1,fi1ename);
-
-/* send the observable, offset by START */
-moment(phi+START,nrou-START,ave, var, cov,2*n1anc-1);
-
-/* First column of T */
-
-T[0] = a1pha(1);
-if((temp=betasq(1))>HACHEPS*10)
-{
-
-T[1]=sqrt(temp);
-
-/* General column of T */
-
-J'=2:
-
-for(i=1;i<n1anc-1, f1ag==O;)
-{
-
-T[itn1anc+i-1]=sqrt(betasq(i));
-
-48
-
-main()
-
-{
-
-char tfilename;
-
-FILE tofp;
-
-double temp=0, #T;
-
-long i, j, nrow, nlanc, START;
-int f1ag=0;
-
-filename = cma11oc(HAx_NAH);
-
-princf("\nName of file containing observed values: “);
-read_name(fi1ename);
-
-printf("\nTota1 number of observations in file (iterations): “);
-scanf("2u“,&nrow);
-
-printf("\nNumber of leading observations to discard: “);
-scanf("Zu",&START);
-
-printf("\nNumber of Lanczos coefficients desired: “);
-scanf("%u“,&n1anc);
-
-whi1e(2*n1anc > nrou-START)
-
-{
-printf("\nNot enough data for that many coefficients.\n");
-
-printf("\nEnter a smaller number of Lanczos coefficients: “);
-scanf("Zu",&n1anc);
-}
-
-phi = dma11oc(urow);
-cov=dma1loc(2*n1anc);
-var=dma11oc(1);
-ave=dma11oc(1);
-
-T = dma11oc(n1anc*n1anc); for(i=0;i<n1anc;i++) T[i]=(doub1e)O;
-
-/* observable phi is stored contiguously column-wise (e.g. by HATLAB) */
-mat1abread(phi, nrov, 1,fi1ename);
-
-/* send the observable, offset by START #/
-moment(phi+START,nrov-START,ave, var, cov,2*n1anc-1);
-
-/# First column of T */
-
-T[0] = a1pha(1);
-if((temp=betasq(1))>HACHEPS#10)
-{
-
-T[1]=sqrt(temp);
-
-/* General column of T */
-
-J‘=2:
-
-for(i-1;i<n1anc-1, f1ag==0;)
-{
-
-T[i*nlanc+i-1]=sqrt(betasq(i));
-
-48
-
-% 49.txt
-T[i*n1anc+i] = a1pha(i+1);
-if((temp = betasq(i+1))>HACmEPS#10)
-{
-
-T[i#n1anc+i+1] = sqrt(temp);
-
-i++;
-}
-else
-
-f1ag=1;
-
-}
-
-/t Last column */
-if(f1ag!=1 it ((temp= betasq(n1anc-1))>HACHEPSt10))
-
-{
-T[n1anc#n1anc-2] = sqrt(temp);
-T[n1anctn1anc-1] = a1pha(n1anc);
-printf("\nbeta(1) = %1f\nbeta(Zd) = Zlf (last beta)",
-T[1],(nlanc-1),T[nlanc*n1anc-2]);
-}
-else
-{
-printf("\nApproximate invariant space reached at step %d.",i);
-printf("\nbeta(1) = Z1f\nbeta(Zd) = Zlf (last accurate beta)“.
-T[1],i,T[itn1anc+i—1]);
-printf("\nbeta(Zd)“2: Zlf (first spurious resu1t)“,i+1,temp);
-}
-
-printf("\nThe matrix T is: \n");
-matprint(T,n1anc,n1anc);
-
-}
-
-else
-
-{
-
-printf("\nmain(): Approximate invariant space reached at first step.“);
-
-49
-
-printf("\na1pha(1) - Z1f\nbeta(1)”2: Zlf (first spurious result)“, T[O], temp);
-
-}
-ofp = fopen("Tmat.m","v“);
-check(ofp);
-mat1aburite(T,n1anc,n1anc,ofp);
-fc1ose(ofp);
-}
-void read_name(char tname)
-{
-int c, i = 0;
-while ((c = getchar()) != EOF it c != ’ ’ it c !- ’\n’)
-name[i++] = c;
-name[i] = ’\0’;
-}
-
-/#t* end 1anczos.c *#*/
-
-% 50.txt
-50
-
-/llnkttttttttinlntint*1!!!##1##!ttitilltit###t¥#tt¥ttItt####t#ttII=##mlltttt
-
-* functions.c functions required by 1anczos.c t
-# t
-* Created by Hilliam J. De Meo #
-* on 1/7/98 *
-t Last modified 1/10/98 :
-
-¥¥##¥$¥##¥¥#$¥¥¥###¥¥$*I#****#**####*######¥#¥#¥#$#¥¥#l¥**#****/
-
-#inc1ude <math.h>
-
-#define START 1000
-
-#define ITER 10000
-
-/t Machine constants on dino (Sun Ultra 1 at Courant) */
-
-#define HACHEPS 1.15828708535533604981e-16
-
-#define SQRTEPS 1.07623746699106147048e-O8
-
-double a1pha(int j);
-
-double betasq(int j);
-
-double form1(int j, int k);
-
-double form2(int j, int k);
-
-double moment(doub1e tdata, long n, double tave, double tvar, double tcov, int k);
-
-/t external variables to be used by functions */
-double *phi,tcov, *var, tave;
-
-double alpha(int j)
-{
-
-/* alpha is never called with j < 1 */
-if(j==1)
-return
-form1(1,1);
-else if(j>1)
-return
-((doub1e)1/betasq(j-1))*
-(form1(j-1,3) — pou(a1pha(j-1),3)
-- 2 * (alpha(j-1)#betasq(j-1) + sqrt(betasq(j-2))#form2(j-1,2))
-+ betasq(j-2));
-}
-
-double betasq(int j)
-
-{
-if(j==0)
-return (doub1e)O;
-else if(j>0)
-return (form1(j,2) - pow(a1pha(j),2) - betasq(j-1));
-}
-
-double form1(int j, int k)
-{
-
-% 51.txt
-51
-
-double form14,alphal,alpha1sq,form12,form13;
-if(j==0)
-
-return (doub1e)0;
-else if(j==1)
-{
-
-/# printf("\nvar = %1f, cov(%d) = %1f \n“,*var,k,cov[k]);t/
-
-return (cov[k])/(*var); /* the only real value */
-}
-else if(j>1)
-{
-
-return ((double)1/betasq(j-1))
-
-* ( form1(j-1,k+2)
-+ pow(a1pha(j-1),2)*form1(j-1,k)
-+ 2t(a1pha(j-1) t sqrt(betasq(j-2)) * form2(j-1,k)
-- a1pha(j-1)*form1(j-1,k+1)
-- sqrt(betasq(j—2))#form2(j-1,k+1) )
-+ betasq(j-2)*form1(j-2,k)):
-}
-}
-
-double form2(int j, int k)
-{
-/* form2 is never called with j < 1 */
-if(j==1)
-return (doub1e)0;
-else if(j>1)
-return
-(pov(betasq(j-1),—.5)) 3
-(form1(j-1,k+1) - a1pha(j-1)*form1(j-1,k)
-- sqrt(betasq(j-2)) # form2(j—1,k));
-}
-
-/* moment() function for computing var and cov(k)
-arguments:
-data = a nxi array of doubles
-n = length of data[]
-ave =(on exit)= the average of data[]
-var =(on exit)= the variance of data[]
-cov =(on exit)= the covariance of data[i] and data[i+j] for i-1,...,k
-k = the max lag for cov above
-*/
-double moment(double tdata, long n, double tave, double tvar, double tcov, int k)
-{
-/* Centered about data[0] algorithm: #/
-
-long i,j;
-double avel, ave2;
-
-% 52.txt
-52
-
-*ave=O;tvar=0; ave1=ave2=O;
-for(j=0;j<=k;j++)
-cov[j]=0;
-
-for(i=1;i<n;i++)
-{
-rave += (data[i] - data[0]);
-tvar += (data[i] - data[0])#(data[i] - data[0]);
-}
-tvar /= (doub1e)(n-1);
-Ivar -= (((*ave)/(doub1e)n) * ‘(*ave)/(doub1e)(n-1)));
-/# *ave = ((*ave)/(doub1e)n) + data[0]; <- that’s the true average (not needed)*/
-
-for(j=0;j<=k;j++)
-{
-
-for(i=0;i<(n-j);i++)
-
-{
-
-avel += (data[i] - data[0]);
-ave2 += (data[i+j] - data[0]);
-cov[j] += (data[i] - data[0])*(data[i+j] — data[0]);
-
-}
-
-avel/=(doub1e)(n-j); ave2/=(doub1e)(n-j);
-
-cov[j] = ((cov[j] - (double)(n-j)tave1¢ave2)/(double)(n-j-1));
-}
-
-}
-
-/trt end funccions.c ###/

### 2013-10-18 tex/MarkovChains.tex
initial commit of tex source for nyu thesis style setup

\chapter{Markov Chains}

\section{General Theory}

This review of Markov chain theory can be found in any good probability text. The present
discussion is most similar to that of Durrett~\cite{Durret:1996}, to which we
refer the reader desiring greater detail.

\subsection{The Basic Setup}

Heuristically, a Markov chain is a stochastic process with a lack of memory property. Here
this means that the future of the process, given its past behavior and its present state, depends only
on its present state. This is the probabilistic analogue of a familiar property of classical particle
systems. Given the position and velocities of all particles at time t, the equations of motion can be
completely solved for the future evolution of the system. Thus, information describing the behavior
of the process prior to time t is superﬂuous. To be a bit more precise, if technical, we need the
following definitions.

\begin{definition}
%Definition 2.1.1 
Let $(\S, \S)$ be a measurable space. A sequence $X_n$, $n\geq 0$, of random variables
taking values in \S\ is said to be a Markov chain with respect to the filtration 
$\sigma(X_0, \dots, X_{n})$ if for all $B \in \S$,
\begin{equation}
% Eqn 2.1
\label{eq:1}
P(X_{n+1} \in B \mid \sigma(X_0, \dots, X_{n}))
=P(X_{n+1} \in B \mid \sigma(X_{n})).
\end{equation}
\end{definition}
Equation~(\ref{eq:1}) merely states that if we know the present location or state of $X_{n}$,
then information about earlier locations or states is irrelevant for predicting $X_{n+1}$.

\begin{definition}
%% Definition 2.1.2 
A function $p : S \times \S \rightarrow \R$ is said to be a \emph{transition probability} if:
\begin{enumerate}
\item for each $x \in S$, $A \mapsto p(x, A)$ is a probability measure on $(S, \S)$.
\item for each $A \in \S$, $x \mapsto p(x, A)$ is a measurable function.
\end{enumerate}
\end{definition}
We call $X_n$ a Markov chain with transition probabilities $p_n$ if
\begin{equation}
% Eqn 2.2
\label{eq:2.2}
P(X_{n+1} \in B \mid  \sigma(X_n)) = p_n(X_n, B)
\end{equation}
The spaces $(S, \S)$ that we encounter below are standard Borel spaces, so the existence of the
transition probabilities follows from the existence of regular conditional
probabilities on Borel spaces---a standard measure theory result 
(see e.g.\cite{Durret:1996}). %[3] page 230

Suppose we are given an initial probability distribution $\mu$ on $(S, \S)$ and a sequence $p_n$ of
transition probabilities. We can define a consistent set of finite dimensional distributions by
\begin{equation}
% Eqn 2.3
\label{eq:2.3}
P(X_j\in B_j ,0 \leq j \leq n) = \int_{B_0} \mu(dx_0) \int_{B_1} p_0(x_0, dx_1) 
\int_{B_2} p_1(x_1, dx_2) \cdots 
\int_{B_n} p_{n-1}(x_{n-1}, dx_n).
\end{equation}
Furthermore, denote our probability space by
\[
(\Omega, \F) = (S^\omega,\S^\omega), \quad \text{where} \omega = \{0,l,...\}.
\]
We call this \emph{sequence space} and it is defined more explicitly by
\[
S^\omega  = \{(\omega_0, \omega_1, \dots) : \omega_i in S\}\quad \text{ and }
\quad
\S^\omega  = \sigma(\omega : \omega_i \in A_i \in \S).
\]
% 07.txt
The Markov chain that we will study on this space is simply $X_n(\omega) = \omega$, the coordinate maps.
Then, by the Kolmogorov extension theorem, there exists a unique probability measure $P_\mu$ on
$(\Omega, \F)$ so that the $X_n(\omega)$ have finite dimensional distributions~(\ref{eq:2.3}).

If instead of $\mu$, we begin with the initial distribution $\delta_x$, i.e., point mass at $x$, then we
denote the probability measure by $P_x$. With such measures defined for each $x$, we can in turn
define distributions $P_\mu$, given any initial distribution $\mu$, by
\[
P_\mu(A) = \int \mu(dx)P_x(A).
\]

That the foregoing construction---which, recall, was derived merely from an
initial distribution $\mu$ and a sequence $p_n$ of transition
probabilities---satisfies Definition~(\ref{eq:2.2}) of a Markov chain is not
obvious, and a proof can be found in~\cite{Durret:1996}.

To state the converse of the foregoing, if $X_n$ is a Markov chain with
transition probabilities $p_n$ and initial distribution $\mu$, then its finite
dimensional distributions are given by~(\ref{eq:2.3}). Proof of this is also found
in~\cite{Durret:1996}.

Now that we have put the theory on a firm, if abstract, foundation, we can bring
the discussion down to earth by making the forgoing a little more
concrete. First, we specialize our study of Markov chains by assuming that our
chain is \emph{temporally homogeneous}, which means that the transition
probabilities do not depend on time; i.e., 
$p_n(\omega_n, B) = p(\omega_n, B)$. (This is the stochastic analogue of a
conservative system.) 
Next we assume that our state space $S$ is finite, and suppose for all states
$i,j \in S$ that $p(i, j) \geq 0$, and $\Sigma_j p(i, j) = 1$ for all $i$. In
this case, equation~(\ref{eq:2.2}) takes a more intuitive form,
\[
P(X_{n+1}=j \mid X_n = i) =p(i,j),
\]
and our transition probabilities become
\[
p(i,A) = \sigma_{j\in A} p(i,j).
\]
% 08.txt
If $\P$ is a matrix whose $(i, j)$ element is the transition probability 
$p(i, j)$ then P is a \emph{stochastic matrix}; that is, a matrix with elements
$p_{ij}$ satisfying 
\[
\pij \geq 0, \quad \Sigma_j \pij =1, \quad (i, j =1, 2, \dots, d)
We also refer to $\P$ as the transition probability matrix.

Without loss of generality, we can further suppose our Markov chain is \emph{irreducible}. This
means that, for any states $i, j$, starting in state $i$ the chain will make a
transition to state $j$ at some future time with positive probability. This
state of affairs is often described by saying that all states \emph{communicate}. We
lose no generality with this assumption because any \emph{reducible} Markov chain can
be factored into irreducible classes of states which can each be studied
separately.

The final two conditions we place on the Markov chains considered below will cost us
some generality. Nonetheless, there remain many examples of chains meeting these conditions
and making the present study worthwhile. Furthermore, it may be the case that, with a little
more work, we will be able to drop these conditions in future studies. The first condition is that
the chain is \emph{aperiodic}. If we let $I_x = \{n \geq 1: p^n(x,x) > 0\}$, we
call a Markov chain \emph{aperiodic} if, for any state $x$, the greatest common
divisor of $I_x$ is 1. The second assumption is that our chain is 
\emph{reversible}. This characterization is understood in terms of the following definition.

\begin{definition}
%% Definition 2.1.3 
\label{def:2.1.3}
A measure $\mu$ is called \emph{reversible} if it satisfies
\[
\mu(x)p(x,y) = \mu(y)p(y,x), \quad  \text{ for all $x$ and $y$.}
\]
We call a Markov chain \emph{reversible} if its stationary distribution (defined
in Section~\ref{sec:convergence-theorem}) is reversible. 
\end{definition}

% ---------- 09.txt ------------------------------------------------------------

\subsection{A Convergence Theorem}
% Section 2.1.2
\label{sec:convergence-theorem}
In succeeding arguments, we use some results concerning the asymptotic behavior of
Markov chains. These results require a few more definitions.

\begin{definition}
%% Definition 2.1.4
A measure $\pi$ is said to be a \emph{stationary measure} if
\begin{equation}
% Eqn 2.4
  \label{eq:2.4}
\Sigma_x \pi(x) p(x,y) = \pi(x).
\end{equation}
\end{definition}

Equation~(\ref{eq:2.4}) says $P_\pi(X_1 = y) = \pi(y)$, 
and by induction that $P_\pi(X_n = y) = \pi(y)$ for all $n \geq 1$. 
If $\pi$ is a probability measure, then we call $\pi$ a stationary distribution. It
represents an equilibrium for the chain in the sense that, if $X_0$ has
distribution $\pi$, then so does $X_n$ for all $n$.

When the Markov chain is irreducible and aperiodic, the distribution of the state at time
$n$ converges pointwise to $\pi$ as $n \rightarrow \infty$, regardless of the
initial state. It is convenient to state this convergence result in terms of the
Markov chain's transition probability matrix $\P$. Before doing so, we note that
irreducibility of a Markov chain is equivalent to irreducibility (in the usual 
matrix theory sense) of its transition probability matrix. Furthermore, it turns
out that a transition probability matrix of an aperiodic Markov chain falls into
that class of matrices often called acyclic, but for simplicity we will call
such stochastic matrices aperiodic. With this terminology, we can state the
convergence theorem in terms of the transition probability matrix \P.

% Theorem 2.1.1
\begin{theorem}
\label{thm-2.1.1}
Suppose $\P$ is irreducible, aperiodic, and has stationary distribution
$\pi$. Then as $n\rightarrow \infty$, $p^n(i,j) \rightarrow \pi(j)$.
\end{theorem}

The notation $p^n(i,j)$ means the $(i,j)$ element of the $n$th power of $\P$.

A Markov chain whose transition probability matrix satisfies the hypotheses of
Theorem~\ref{thm-2.1.1} is called \emph{ergodic}. If we simulate an ergodic chain for
sufficiently many steps, having 
%
%
%
% ---------------10.txt --------------------------------------------------
%
%
%
begun in any initial state, the final state is a sample point from a
distribution that is close to $\pi$.

To make this statement more precise requires that we define ``close.''
\begin{definition}
%% Definition 2.1.5 
Let $\pi$ be a probability measure on $S$. Then the \emph{total variation distance} at time
$n$ with initial state $x$ is given by 
\[
\Delta_x(n) = \|\P^n(x, A) - \pi(A)\|_{TV} = \max_{A\in \S} | \P^n(x,A)-\pi(A)|.
\]
\end{definition}
In what follows, we will measure rate of convergence using the function
$\tau_x(\epsilon)$, defined as the first time after which the total variation
distance is always less than $\epsilon$. That is, 
\[
\tau_x(\epsilon) = \min\{m : \Delta_x(n) \leq \epsilon \text{ for all } n\geq m\}.

To begin our consideration of the connection between convergence rates of Markov
chains and eigenvalues, we first note that an aperiodic stochastic matrix $\P$ has real
eigenvalues 
$1 = \lambda_0 > \lambda_1 \geq \lambda_2 \geq \cdots \geq \lambda_{d-1} \geq -1$, 
where $d = |S|$ is the dimension of the state space. For an ergodic chain, the
rate of convergence to the stationary distribution $\pi$ is bounded by a
function of the \emph{subdominant} eigenvalue. By subdominant eigenvalue we mean that
eigenvalue which is second largest in absolute value, and we denote this
eigenvalue by $\lambda_{\max} = \max{\lambda_1, |\lambda_{d-1}|}. 
The function bounding the rate of convergence of a Markov chain is given by the
following theorem:
\begin{theorem}
% Theorem 2.1.2
\label{thm-2.1.2}
The quantity $\tau_x(\epsilon)$ satisfies
1. $\tau_x(\epsilon) \leq (1-\lambda_\max)^{-1}(\log \pi(x)^{-1} + \log \epsilon^{-1})$;
2. $\max_{x \in S} \tau_x(\epsilon) \geq \frac{1}{2} \lambda_\max(1-\lambda_\max)^{-1} \log(2\epsilon)^{-1}$.
\end{theorem}
As this theorem shows, if we have an upper bound on the subdominant eigenvalue, then we have
an upper bound on the function $\tau_x(\epsilon)$. In what follows, we will
derive an approximation to the 
% 11.txt
subdominant eigenvalue and supply error bounds. Together, an approximation and error bounds
for $\lambda_\max$  provide enough information to make Theorem~\ref{thm-2.1.2} useful.

\section{Functions on the State Space}
% Sec 2.2
\label{sec:funct-state-space}
Recall that $X_n(\omega) = \omega_n\in S$ denotes the state in which the Markov
chain exists at time $n$. 
Suppose that 
$\Phi = \{\phi_1, \dots, \phi_p\}$ 
is a collection of $p$ \emph{observables}, or functions defined on the state
space $S$. Furthermore, let these observables be real valued, 
$\phi_i : S \rightarrow \R$. 
It is often useful to assume that none of the observables is a constant function. Suppose now that
the state space $S$ is finite with $d$ possible states. Then, since an
observable is simply a map of the state space, we can think of each $\phi_i$ as a
vector of $d$ real numbers---the $d$ values that it takes on at the different states.

Now assume the Markov chain is irreducible, and let $\pi$ denote its stationary distribution.
If we start the chain from its stationary distribution---i.e., suppose $X_0$ has
distribution $\pi$---then $X_n$ is a stationary process. 
Furthermore, for each $i$, $\phi_i(X_n)$ is a stationary stochastic process with
\emph{mean}
\[
\E_\pi\phi_i = \Sigma_{x\in S} \pi(x)\phi_i(x)
\]
and \emph{autocovariance function}
\begin{align}
% Eqn 2.5
\C_\pi (\phi_i(X_n),  \phi_i (X_{n+s})) 
&= \E_\pi[(\phi_i(X_n) - \E_\pi\phi_i)(\phi_i(X_{n+s}) - \E_\pi\phi_i)]\nonumber\\
&= \Sigma_{x, y\in S} \P_\pi(X_n = x, X_{n+s} = y) (\phi_i(x)- \E_\pi\phi_i)(\phi_i(y) - \E_\pi\phi_i)
\end{align}
which, by the definition of conditional probability, we can write as,
\[
\Sigma_{x, y\in S} \P_\pi(X_n = x) \P_\pi(X_{n+s} = y\mid X_n = x) (\phi_i(x)-
\E_\pi\phi_i)(\phi_i(y) - \E_\pi\phi_i).
\]
Equivalently,
\[
\Sigma_{x, y\in S} \pi(x)p^s_{xy} (\phi_i(x)-\E_\pi\phi_i)(\phi_i(y) - \E_\pi\phi_i).
\]
Here $p^s_{xy}$ denotes the element in row $x$ and column $y$ of $\P^s$, the
$s$th power of the transition probability matrix. 
Similarly, we define the \emph{cross-covariance} between the function 
$\phi_i$ at time $n$ and $\phi_j$ at time $n+s$ as
\begin{align}
% Eqn 2.6
\C_\pi (\phi_i(X_n),  \phi_j (X_{n+s})) 
&= \E_\pi[(\phi_i(X_n) - \E_\pi\phi_i)(\phi_j(X_{n+s}) - \E_\pi\phi_j)]\nonumber\\
&= \Sigma_{x, y\in S} \pi(x)p^s_{xy} (\phi_i(x)-\E_\pi\phi_i)(\phi_j(y) - \E_\pi\phi_j).
\end{align}
Now let $\<\Phi\>$ denote the matrix of mean vectors whose $j$th column is 
$\E_\pi\phi_j\one$, where $\one = (1,\dots, 1)^t$,
and let $\Pi = \diag(\pi(\omega_1),\dots, \pi(\omega_d))$ be the $d \times d$
diagonal matrix with stationary probabilities $\pi(\omega)$
on the main diagonal and zeros elsewhere. 
Finally, denoting by $C(s)$ the $p \times p$ covariance matrix
whose $(i,j)$ element is $C_\pi(\phi_i(X_n), \phi_j(X_{n+s}))$, we have
\begin{align*}
C(0) &= \E(\Phi(X_n) - \<\Phi\>)(\Phi(X_n) - \<\Phi\>)^t\\
&= (\Phi - \<\Phi\>)^t\Pi (\Phi-\<\Phi\>), \\
C(s) &= \E(\Phi(X_n) - \<\Phi\>)(\Phi(X_{n+s}) - \<\Phi\>)^t\\
&= (\Phi - \<\Phi\>)^t \Pi \P^s (\Phi - \<\Phi\>).
\end{align*}

Below, we will also find it useful to have at our disposal a new matrix that is \emph{similar} to the
transition probability matrix. We have in mind the matrix 
$\M = \Pi^{1/2}\P\Pi^{-1/2}$. 
As is easily verified, this allows us to write the covariance matrix as
\begin{align}
\label{eq:2.7}
C(s) &= 
(\Phi - \<\Phi\>)^t \Pi^{\frac{1}{2}t}\M^s\Pi^{\frac{1}{2}}(\Phi - \<\Phi\>)\nonumber\\
&=\Psi^t \M^s \Psi,  % Eqn 2.7
\end{align*}
%% \\Phi\Phi‘M’\\Phi\Phi (2.7)
where we have defined $\Psi = \Pi^{\frac{1}{2}}(\Phi - \<\Phi\>)$. Recall that
our main motivation is that for out-of-core problems traditional eigenvalue
algorithms are inadequate. With this fact and the form~(\ref{eq:2.7}) % (2.7) 
in 
%
%
% ------------------ 13.txt ---------------------------------------------------
%
%
mind, we will consider using the covariance of observables on the state space to implement the
%% (otherwise useless) 
Rayleigh-Ritz procedure, which we describe below. This procedure requires
that $\M$ be symmetric. As the next fact demonstrates, 
%% this required symmetry is the reason for our
%% interest in the reversibility property of the Markov chain.
this need for symmetry is the reason we insist that the Markov chain be reversible.
\begin{fact}
%% Fact 2.2.1 
The matrix $\M$ is symmetric if and only if the Markov chain is reversible (i.e., iff the
process satisfies the \emph{detailed balance} condition).
\end{fact}
\begin{proof}
  \begin{align*}
  \M \text{ is symmetric } 
& \Leftrightarrow 
 (\Pi^{1/2}\P\Pi^{-1/2})^t = \Pi^{1/2}\P\Pi^{-1/2}\\
& \Leftrightarrow 
 \Pi^{-\frac{1}{2}t}\P^t\Pi^{\frac{1}{2}t} = \Pi^{1/2}\P\Pi^{-1/2}\\
& \Leftrightarrow 
 \P^t\Pi^t = \Pi\P.
  \end{align*}
Elementwise, the final equality is $\pi_i\pij = \pi_j\pji$. 
According to Definition~\ref{def:2.1.3}, this states
that $p$ is a reversible measure.
\end{proof}

% 14.txt

### 2013-10-18 tex/abstract.tex
initial commit of tex source for nyu thesis style setup

--- 2013-10-18 tex/MSThesis.tex
@@
-%\documentclass[11pt]{amsart}
-\documentclass{au}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% For completion by editorial office:
- \presentedby{\dots}
- \received{\dots}{\dots}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{center}
+{\bf Abstract}\\
+\\
+A Lanczos Procedure for Approximating Eigenvalues of Large Stochastic Matrices\\
+\\
+by\\
+\\
+William J. DeMeo\\
+\\
+Master of Science in Mathematics\\
+\\
+New York University\\
+\\
+Professor Jonathan Goodman, Chair
+\end{center}
 
-\usepackage[colorlinks=true,urlcolor=black,linkcolor=black,citecolor=black]{hyperref}
-
-\usepackage{amsmath,amssymb,url,mathrsfs}
-\usepackage{enumerate}
-\usepackage{tikz}
-\usetikzlibrary{calc}
-\usepackage{scalefnt}
-\numberwithin{equation}{section}
-
-\theoremstyle{plain}
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{proposition}[theorem]{Proposition}
-\newtheorem{corollary}[theorem]{Corollary}
-
-\theoremstyle{definition}
-\newtheorem{definition}[theorem]{Definition}
-\newtheorem{notation}[theorem]{Notation}
-\newtheorem{remark}[theorem]{Remark}
-\newtheorem{example}[theorem]{Example}
-\newtheorem*{remarks}{Remarks}
-
-\newcommand{\todo}[1]{~\\[5pt]{\bf TODO(wjd):} #1\\[5pt]}
-%\newcommand{\suchthat}{\ensuremath{:}}    % (old notation)
-\newcommand{\suchthat}{\ensuremath{\mid}}  % (new notation)
-%\newcommand{\defeq}{\ensuremath{:=}}      % (old notation)
-\newcommand{\defeq}{\ensuremath{=}}        % (new notation)
-\newcommand{\betai}{\ensuremath{\beta^i}}  % \beta^{\bB_i}   % (old notation)
-\newcommand{\betaj}{\ensuremath{\beta^j}}  % \beta^{\bB_j}
-\newcommand{\betak}{\ensuremath{\beta^k}}  % \beta^{\bB_k} or \beta^{\bB_{k}}
-\newcommand{\betal}{\ensuremath{\beta^\ell}}  %\beta^{\bB_\ell} or \beta^{\bB_{\ell}}
-\newcommand{\uacalc}{{\small UAC}alc}
-\newcommand{\Tuma}{T\r{u}ma}
-\newcommand{\Palfy}{P\'alfy}
-\newcommand{\Pudlak}{Pudl\'ak}
-\newcommand{\PP}{P\'alfy-Pudl\'ak}
-\newcommand{\PAP}{P\'alfy\ and Pudl\'ak}
-\newcommand{\<}{\ensuremath{\langle}}
-\renewcommand{\>}{\ensuremath{\rangle}}
-\newcommand{\beps}{\ensuremath{\boldsymbol{\varepsilon}}}
-\newcommand{\bN}{\ensuremath{\mathbf{N}}}
-\newcommand{\bA}{\ensuremath{\mathbf{A}}}
-\newcommand{\bB}{\ensuremath{\mathbf{B}}}
-\newcommand{\bBi}{\ensuremath{\mathbf{B}_i}}
-\newcommand{\sB}{\ensuremath{\mathcal{B}}}
-\newcommand{\sC}{\ensuremath{\mathcal{C}}}
-\newcommand{\sS}{\ensuremath{\mathscr{S}}}
-\newcommand{\sT}{\ensuremath{\mathscr{T}}}
-\newcommand{\sI}{\ensuremath{\mathcal{I}}}
-\newcommand{\sE}{\ensuremath{\mathcal{E}}}
-\newcommand{\sO}{\ensuremath{\mathcal{O}}}
-%\newcommand{\id}{\ensuremath{\mathrm{id}}}
-\DeclareMathOperator{\id}{id}
-%\newcommand{\Eq}{\ensuremath{\mathrm{Eq}}}
-\DeclareMathOperator{\Eq}{Eq}
-%\newcommand{\Cg}{\ensuremath{\mathrm{Cg}}}
-\DeclareMathOperator{\Cg}{Cg}
-%\newcommand{\Con}{\ensuremath{\mathrm{Con\,}}}
-\DeclareMathOperator{\Con}{Con}
-%\newcommand{\Sub}{\ensuremath{\mathrm{Sub}}}
-\DeclareMathOperator{\Sub}{Sub}
-%\newcommand{\Pol}{\ensuremath{\mathrm{Pol}}}
-\DeclareMathOperator{\Pol}{Pol}
-%\newcommand{\Clo}{\ensuremath{\mathrm{Clo}}}
-\DeclareMathOperator{\Clo}{Clo}
-\newcommand{\ps}[1]{\ensuremath{^{(#1)}}}
-\newcommand{\piB}{\ensuremath{\pi_B}}
-\newcommand{\hpsi}{\ensuremath{\widehat{\psi}}}
-\newcommand{\htheta}{\ensuremath{\widehat{\theta}}}
-\newcommand{\supi}{\ensuremath{^{i}}}
-\newcommand{\supj}{\ensuremath{^{j}}}
-\renewcommand{\leq}{\ensuremath{\leqslant}}
-\renewcommand{\nleq}{\ensuremath{\nleqslant}}
-\renewcommand{\geq}{\ensuremath{\geqslant}}
-\newcommand{\meet}{\ensuremath{\wedge}}
-\newcommand{\join}{\ensuremath{\vee}}
-\newcommand{\Meet}{\ensuremath{\bigwedge}}
-\renewcommand{\Join}{\ensuremath{\bigvee}}
-\newcommand{\resB}{\ensuremath{|_{_B}}}
-\newcommand{\resBi}{\ensuremath{|_{_{B_i}}}}
-\newcommand{\eps}{\ensuremath{\varepsilon}}
-\newcommand{\hatmap}{\ensuremath{\widehat{\phantom{x}}}}
-\newcommand{\one}{\ensuremath{\mathbf{1}}}
-\newcommand{\two}{\ensuremath{\mathbf{2}}}
-\newcommand{\three}{\ensuremath{\mathbf{3}}}
-\newcommand{\tbeta}{\ensuremath{\widetilde{\beta}}}
-\newcommand{\ttheta}{\ensuremath{\widetilde{\theta}}}
-\newcommand{\hbeta}{\ensuremath{\widehat{\beta}}}
-\newcommand{\cick}{\ensuremath{\sC_i}}
-%\newcommand{\CE}{\ensuremath{\sC_r^{\sE}}}
-\newcommand{\CE}{\ensuremath{\sB_r^n}}
-\newcommand{\CO}{\ensuremath{\sC_r^{\sO}}}
-\newcommand{\CICK}{\ensuremath{c_i/\beta \cup c^{K+1}_i/\beta^{\bB_{K+1}}}}
-\newcommand{\GAP}{{\small GAP}}
-\newcommand{\dotsize}{.8pt}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{document}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%  FRONT MATTER                                                 %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\title[Approximating Eigenvalues of Large Stochastic Matrices]{A Lanczos Procedure for Approximating Eigenvalues of Large Stochastic Matrices}
-
-%% First author (Note: The order of the items here is important!)
-\author{William DeMeo}
-\email{williamdemeo@gmail.com}
-\urladdr{http://www.math.sc.edu/~demeow}
-\address{Department of Mathematics\\
-University of South Carolina\\Columbia 29208\\USA}
-
-\thanks{The author wishes to thank a number of people who contributed to this
-  work.  In particular, the primary referee made many insightful recommendations
-  which led vast improvements, not only in the overall exposition, but also in
-  the specific algebraic constructions, theorem statements, and proofs.
-  This article is dedicated to Ralph Freese and Bill Lampe, who declined to be
-  named as co-authors despite their substantial contributions to the project.}
-
-%% Dedication (Optional)
-\dedicatory{This article is dedicated to Ralph Freese and Bill
-  Lampe.}
-
-%% AMS subject classification; see http://www.ams.org/msc
-%% Only one Primary. Possibly several Secondary.
-\subjclass[2010]{Primary: 08A30; Secondary: 08A60, 06B10.} %, 06B15.}
-
-%% Keywords and phrases
-\keywords{congruence lattice, finite algebra, finite lattice
-  representations}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%  MAIN MATTER                                                  %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{abstract}
 The rate at which a Markov chain converges to a given probability distribution
 has long been an active area of research. Well known bounds on this rate of
-convergence involve the subdominant eigenvalue of the chain’s underlying
+convergence involve the subdominant eigenvalue of the chain's underlying
 transition probability matrix. However, many transition probability matrices are
@@
 traditional Lanczos algorithm without storing a single vector. We do this by
-considering the variational properties of observables on the chain’s state
+considering the variational properties of observables on the chain's state
 space. In the process we present the classical theory which relates the
@@
 Markov chain.
-\end{abstract}
-
-\maketitle
-
-%%%%%%%%%%%%%%%%%%%%%
-\chapter{Introduction}
-The rate at which a Markov chain converges to a given probability distribution has long
-been an active area of research. This is not surprising considering this problem’s relevance to the ar-
-eas of statistics, statistical mechanics, and computer science. Markov Chain Monte Carlo (MCMC)
-algorithms provide important examples. These algorithms come in handy when we encounter a
-complicated probability distribution from which we want to draw random samples. In statistical
-mechanics, we might wish to estimate the phase average of a function on the state space. Goodman
-and Sokal [6] examine Monte Carlo methods in this context. Examples from statistics occur in the
-Bayesian paradigm when we are forced to simulate an unwieldy posterior distribution (see, e.g.,
-Geman and Geman
-
-To implement the MCMC algorithm, we invent a Markov chain that converges to the
-desired distribution (this is often accomplished using the Metropolis algorithm
-described in Chapter 5). Realizations of the chain will eventually represent
-samples from this distribution. Sometimes ``eventually'' -- meaning all but
-finitely many terms of the chain -- is just not enough. We need more practical
-results. In particular, we want to know how many terms of the chain should be 
-discarded before we are sampling from a distribution that is close (in total variation distance) to
-% 03.txt
-the distribution of interest. This is the purpose of bounding convergence rates for Markov chains.
-
-Often the Markov chains encountered in this context satisfy a condition known in the
-physics literature as detailed balance. Probabilists call chains with this property reversible. This
-simply means that the chain has the same probability law whether time moves
-forward or 
-backward.\footnote{This is not a precise definition. In particular the chain must
-  have started from its stationary distribution. Full rigor is postponed until Section 2.1.}
-In this paper, we consider the rate at which such chains converge to a \emph{stationary distribution}.\footnote{This and other italicized terms are defined in Section 2.1.}
-
-There are a number of different methods in common use for bounding convergence rates of
-Markov chains, and a good review of these methods with many references can be found in  More
-recently developed methods, employing logarithmic Sobolev inequalities, are reviewed in  Most
-of the bounds in common use involve the sub—dominant eigenvalue of the Markov chain's transition
-probability matrix, and thus require good approximations to such eigenvalues. In many applications,
-however, the transition probability matrix is so large that it becomes impossible to store even a
-single vector of the matrix in conventional computer memory. These so called out-of-core problems
-are not amenable to traditional eigenvalue algorithms\footnote{By ``traditional
-  eigenvalue algorithms'' we refer to those found, for example, in Golub and Van
-  Loan[5]. See also the book by Demmel [1] for a more recent discussion.}
-without modification. This paper develops 
-such a modification for the Markov chain eigenvalue problem. In particular it develops a method
-for approximating the first few eigenvalues of a transition probability matrix when we know the
-general structure of the underlying Markov chain. The method does not require storage of large
-matrices or vectors. Instead we need only simulate the Markov chain, and conduct a statistical
-analysis of the simulation.
-
-Here is a look at what follows. Section 2.1 contains a review of the relevant Markov chain
-theory. Readers conversant in the asymptotic theory of Markov chains might wish to at least skim
-Section 2.1, if only to become familiar with our notation. Section 2.2 describes functions on the state
-% 04.txt
-space of the Markov process. This section and Chapter 3 develop the context in which we formulate
-the new ideas of the paper. In the last section of Chapter 3, Section 3.3, we present the familiar
-Krylov subspace and explain why this represents our best approximation to a subspace containing
-extremal eigenvectors of the transition probability matrix.\footnote{or, more
-  precisely, a similarity transformation of this matrix.} The first section of
-Chapter 4 describes the \emph{Lanczos algorithm} for generating an orthonormal basis
-for the Krylov subspace. As it stands, this algorithm is useless for an
-out-of-core problem such as ours since, by definition of such problems, 
-it requires too much data movement; all the computing time is spent swapping data between slow
-and fast memory (e.g. between the hard disk and cache). Therefore, we discuss alternatives to
-Lanczos and demonstrate that the \emph{Lanczos coeficients} are readily available through simulations of
-the Markov chain, which fact allows us to avoid the standard algorithm altogether. Following this
-is a chapter describing the Metropolis algorithm used to produce a reversible stochastic matrix.
-It is here that we experiment with the procedure described in Section 4.2 and approximate the
-extremal eigenvalues of the matrix, without storing any of its vectors. Finally, Chapter 6 concludes
-the paper.
-% 05.txt
-
-\chapter{Markov Chains}
-
-\section{General Theory}
-
-This review of Markov chain theory can be found in any good probability text. The present
-discussion is most similar to that of Durrett [3], to which we refer the reader desiring greater detail.
-
-\subsection{The Basic Setup}
-
-Heuristically, a Markov chain is a stochastic process with a lack of memory property. Here
-this means that the future of the process, given its past behavior and its present state, depends only
-on its present state. This is the probabilistic analogue of a familiar property of classical particle
-systems. Given the position and velocities of all particles at time t, the equations of motion can be
-completely solved for the future evolution of the system. Thus, information describing the behavior
-of the process prior to time t is superﬂuous. To be a bit more precise, if technical, we need the
-following definitions.
-
-\bibliographystyle{spmpsci}
-\bibliography{wjd}
-
-\end{document}
-
-%%%%%%%%%%%%%%%%%%%%%  END DOCUMENT %%%%%%%%%%%%%%%%%%%%%%%%%
-
-Definition 2.1.1 Let (S, 8) be a measurable space. A sequence X,., n 2 0, of random variables
-
-taking values in S is said to be a Markov chain with respect to the filtration a(Xo, . . . ,X,,) if for
-% 06.txt
-a.llBES,
-
-p(x,,+1 e B | a(Xo, . . . ,X,.)) = P(X,.+1 e B 1 o(X,,)) (2.1)
-
-Equation (2.1) merely states that, given the present location of X,,, the past is irrelevant for
-
-predicting the location of X,,+1.
-
-Definition 2.1.2 A function p : S x S —> R is said to be a transition probability if:
-1. for each :1: E S, A —> p(a:, A) is a probability measure on (S, S).
-2. for each A E 8, :1: —> p(z, A) is a measurable function.
-
-We call X,, a Markov chain with transition probabilities 12,, if
-P(Xn+l E B l a(Xn)) =Pn(XnaB) (22)
-
-The spaces (S, S) that we encounter below are standard Borel spaces, so the existence of the
-transition probabilities follows from the existence of regular conditional probabilities on Borel spaces
-— a standard measure theory result (see e.g. [3] page 230).
-
-Suppose we are given an initial probability distribution p on (S, S) and a sequence p,, of
-
-transition probabilities. We can define a consistent set of finite dimensional distributions by
-
-P<X,- e B,-,0 s 1 s n) = / #(d1‘o) / Po($o.d-'51) . .. / Pn-l(97n—1ad$n) (2.3)
-B0 B1 B1:
-Furthermore, denote our probability space by
-(9,?) = (S{0,1,...},8{0,l,...})
-
-We call this sequence space and it is defined more explicitly by
-
-sf°»1=-~-} = {(wo,w1,...) :w.- e S}
-
-S{°*1""} = ¢7(w :40; E A,- E 3)
-
-% 07.txt
-The Markov chain that we will study on this space is simply X" (w) = w,., the coordinate maps.
-Then, by the Kolmogorov extension theorem, there exists a unique probability measure P,, on
-(9,?) so that the X" (w) have finite dimensional distributions (2.3).
-
-If instead of p, we begin with the initial distribution 6,, i.e. point mass at 2:, then we
-denote the probability measure by P3. With such measures defined for each 2:, we can in turn
-
-define distributions P,,, given any initial distribution p, by
-
-PM = / u<da.->P.(A>
-
-That the foregoing construction — which, recall, was derived merely from an initial distri-
-bution p and a sequence p,, of transition probabilities — satisfies Definition (2.2) of a Markov chain
-is not obvious, and a proof can be found in
-
-To state the converse of the foregoing, if X,, is a Markov chain with transition probabilities
-p,, and initial distribution p, then its finite dimensional distributions are given by (2.3). Proof of
-this is also found in
-
-Now that we have put the theory on a firm, if abstract, foundation, we can bring the
-discussion down to earth by making the forgoing a little more concrete. First, we specialize our
-study of Markov chains by assuming that our chain is temporally homogeneous, which means that
-the transition probabilities do not depend on time; i.e. p,,(w,,, B) E p(w,., B). (This is the stochastic
-analogue of a conservative system.) Next we assume that our state space S is finite, and suppose
-for all states 1', j E S that p(i, j) Z 0, and E, p(z', j) = 1 for all i. In this case, equation (2.2) takes
-
-a more intuitive form:
-P(Xn+l =j I X7: = 7:) =P(iaj)
-
-and our transition probabilities become
-% 08.txt
-If P is a matrix whose (i, j)th element is the transition probability p(i, 1') then P is a stochastic
-
-matriar, that is, a matrix with elements pgj satisfying
-
-Pij 20: zjpij =1: (i7j=1121"‘)d)
-
-We also refer to P as the transition probability matrix.
-
-Without loss of generality, we can further suppose our Markov chain is irreducible. This
-means that, for any states 2', 1‘, starting in state i the chain will make a transition to state 1' at
-some future time with positive probability. This state of affairs is often described by saying that
-all states communicate. We lose no generality with this assumption because any reducible Markov
-chain can be factored into irreducible classes of states which can each be studied separately.
-
-The final two conditions we place on the Markov chains considered below will cost us
-some generality. Nonetheless, there remain many examples of chains meeting these conditions,
-thus making the present study worthwhile. Furthermore, it may be the case that, with a little
-more work, we will be able to drop these conditions in future studies. The first condition is that
-the chain is aperiodic. If we let I, = {n 2 1 : p"(:c,a:) > 0}, we call a Markov chain aperiodic if,
-for any state 2:, the greatest common divisor of I, is 1. The second assumption is that our chain is
-
-reversible. This characterization is understood in terms of the following definition.
-Definition 2.1.3 A measure a is called reversible if it satisfies
-
-;4(w)p(I. y) = u(y)p(y. w)
-
-for all :c and y.
-
-We call a Markov chain reversible if its stationary distribution (defined in Section 2.1.2) is reversible.
-% 09.txt
-\subsection{A Convergence Theorem}
-
-In succeeding arguments, we use some results concerning the asymptotic behavior of
-
-Markov diains. These results require a few more definitions.
-
-Definition 2.1.4 A measure 1r is said to be a stationary measure if
-
-Z«(z)p<z.y> = we) (2.4)
-
-2
-Equation 2.4 says P,,(X1 = y) = 1r(y), and by induction that P,,(X,, = y) = 7r(y) for all n 2 1. If 1r
-is a probability measure, then we call 1r a stationary distribution. It represents an equilibrium for
-the chain in the sense that, if X0 has distribution 1r, then so does X,. for all 11..
-
-When the Markov chain is irreducible and aperiodic, the distribution of the state at time
-n converges pointwise to 1r as n —> oo, regardless of the initial state. It is convenient to state
-this convergence result in terms of the Markov chain's transition probability matrix P. Before
-doing so, we note that irreducibility of a Markov chain is equivalent to irreducibility (in the usual
-matrix theory sense) of its transition probability matrix. Furthermore, it turns out that a transition
-probability matrix of an aperiodic Markov chain falls into that class of matrices often called acyclic,
-but for simplicity we will call such stochastic matrices aperiodic. With this terminology, we can
-
-state the convergence theorem in terms of the transition probability matrix P.
-
-% Theorem 2.1.1
-\begin{theorem}
-\label{thm-2.1.1}
-Suppose P is irreducible, aperiodic, and has stationary distribution 7r. Then as n —> oo,
-p"(i,J') -> #0")
-\end{theorem}
-
-The symbol p"(i, j) is to be understood as the (ij) element of the nth power of P.
-A Markov chain whose transition probability matrix satisfies the hypotheses of
-Theorem~\ref{thm-2.1.1} is called ergodic. If we simulate an ergodic chain for
-sufficiently many steps, having 
-% 10.txt
-begun in any initial state, the final state is a sample point from a distribution which is close to 1r.
-
-To make this statement more precise requires that we define “close”.
-
-Definition 2.1.5 Let 1r be a probability measure on S. Then the total variation distance at time
-
-n with initial state I is given by
-AA") = l|P"(z, A) - 7r(-4)||Tv = 1333; I P"(I.A) - 7r(A) I
-
-In what follows, we will measure rate of convergence using the function 'r,(e), the first time after
-
-which the total variation distance is always less than 5. That is,
--r,,(e) = min{m : A,(n) 3 E for all n. 2 m}
-
-To consider the connection between convergence rates and eigenvalues, note that an ape-
-riodic stochastic matrix P has real eigenvalues 1 = /\o > A1 2 A2 3  Z /\¢_1 2 -1, where
-d = |S'] is the dimension of the state space. For an ergodic chain, the rate of convergence to the
-stationary distribution 7r is bounded by a function of the subdominant eigenvalue. By subdominant
-eigenvalue we mean that eigenvalue which is second-largest in absolute value, and we denote this
-Am“ E max{).1,|A¢_1|}. The function bounding the rate of convergence is given by the following
-
-theorem:
-
-% Theorem 2.1.2
-\begin{theorem}
-\label{thm-2.1.2}
-The quantity T1(€) satisfies
-1. 13(5) 5 (1 — /\m,_x)“(ln1r(a:)‘1 +1ne‘1)
-2. maxxes r,(e) 2 %/\,,,,x(1 — ).,,,,x)‘1ln(2e)’1
-\end{theorem}
-
-As this theorem shows, if we have an upper bound on the subdominant eigenvalue, then we have
-an upper bound on the function 11(5). In what follows, we will derive an approximation to the
-% 11.txt
-subdominant eigenvalue and supply error bounds. Together, an approximation and error bounds
-for Am“ provide enough information to make Theorem~\ref{thm-2.1.2} useful.
-
-\section{Functions on the State Space}
-
-Recall that X,.(w) = w,. 6 S denotes the state in which the Markov chain exists at time
-11. Suppose that 11> = {¢1, . . . , ¢,,} is a collection of p observables, or functions defined on the state
-space S. Furthermore, let these observables be real valued, so ¢.~ : S —> R. It is often useful, as we
-will see below, to assume that none of the observables is a constant function. Suppose now that
-the state space S is finite with at possible states. Then, since an observable is simply a map of the
-state space, we can think of each ¢,- as a vector of d real numbers — the d values that it takes on at
-the different states.
-
-Now assume the Markov chain is irreducible, and let 1r denote its stationary distribution.
-If we start the chain from its stationary distribution - i.e. suppose X0 has distribution 1r - then
-X,. is a stationary process. Furthermore, for each i, ¢,(X,.) is a stationary stochastic process with
-
-mean
-
-E.¢.- = Z) vr(x)¢.(z)
-
-165
-
-and autocovariance function
-
-Cir (¢i(Xn)v ¢i (Xn+a))
-
-E7!’  ‘ E1r¢i) (¢i(Xn+a) " E1r¢i)l
-
-Z Pn(Xn = z1X7I+S = y)  — E1r¢i)  _ E1r¢i)
-
-z,yES
-(2.5)
-
-which, by the definition of conditional probability, we can write as,
-
-2 P«(X.. = z)P.(X..+.. = y I X. = 1)(¢s($) — Em) (my) ~ Em)
-1.1165
-% 12.txt
-Z 7r($)P1y (MI) - Em.-) (¢:(1I) - E1r¢i)
-
-:c,y€S
-
-Here pgy denotes the element in row :1: and column y of P’, the sth power of the transition probability
-
-matrix. Similarly, we define the cross-covariance between the function ¢,- at time n and q5,- at time
-
-n+sas
-
-E1!’  _ E7r¢:') (¢i(Xn+a) " E1r¢i)l (2-6)
-
-C1r(¢i(Xn)a (#1 (Xn+a))
-
-Z) vr(x)p;,, (¢.-(w) — Em.-) (my) — Em)
-
-z,yES
-
-Now let (<I>) denote the matrix of mean vectors whose jth column is E,,¢,-1, where 1 = (1, . . . , 1)‘,
-and let H = diag(1r(w1), . . . , 1r(wd)), be the d X d diagonal matrix with stationary probabilities 1r(w)
-on the main diagonal, and zeros elsewhere. Finally, denoting by C(s) the p X p covariance matrix
-
-whose (i,j)th element is C, (qfi,-(X,.),¢_,< (X,.+,)), we have
-
-0(0) = E(‘1’(Xn) - (‘I>))(‘I’(Xn) - (‘PW
-= (‘P-(‘1’))‘1'1(<1’-(‘I>))
-0(8) = E(<I>(Xn) — (‘I>))(‘1’(Xn+a) - (‘1’))’
-
-(‘P - (‘1’))'1'IP‘(<1> — (<1>))
-
-Below, we will also find it useful to have at our disposal a new matrix that is similar to the
-transition probability matrix. We have in mind the matrix M E Ill/2PI'I‘1/2. As is easily verified,
-this allows us to write the covariance matrix as
-
-can = <<I>—<<I>>)°n%*M*n%<<I>—<<I>>>
-
-\II‘M’\II (2.7)
-
-where we have defined ‘II E 111/ 2(<I> —  Recall that our main motivation is that, for out-of-core
-problems, traditional eigenvalue algorithms are inadequate. With this fact, and the form (2.7) in
-% 13.txt
-mind, we will consider using the covariance of observables on the state space to implement the
-(otherwise useless) Rayleigh-Ritz procedure, which we describe below. This procedure requires
-that M be symmetric. As the next fact demonstrates, this required symmetry is the reason for our
-
-interest in the reversibility property of the Markov chain.
-
-Fact 2.2.1 The matrix M is symmetric if and only if the Markov chain is reversible (i.e. iﬂ the
-
-process satisfies the detailed balance condition).
-
-Proof:
-
-M is symmetric
-
-4:, (1-I1/2P1-I-1/2): = H1/2Pn—1/2
-<.—> r1-%*P°n%* = H1/2911-1/2
-¢> 1>'r1= = up
-~:=> (IIP)‘ = l'IP
-
-Element-wise, the final equality implies that 7|‘,'p,'j = 1rjpj.-. According to Definition 2.1.3, this states
-
-that p is a reversible measure.
-
-% 14.txt
-
-\chapter{Invariant and Approximate Invariant Subspaces}
-
-\section{Invariant Subspaces}
-Definition 3.1.1 A subspace S Q R" with the property that
-
-1:ES'=Ma:€S
-
-is said to be invariant for M.
-
-Recall that, having chosen observables <I> = ((151, . . . , qfip), we constructed the covariance
-matrix C(s) = \II‘M"\II. If the column space of \II, which we denote by ran(\II), is an invariant
-subspace for M, the definition implies Mifij E  That is, for each 111,- there exists a vector
-t of coeﬂicients such that M1/Jj = f=1t,-1/)5,-. This is true for all j and, putting each vector of
-coefficients into a matrix T, we see that M\II = \IIT. Conversely, M\II = IIIT implies that M1/Jj is a
-
-linear combination of columns of \II, so ran(\II) is invariant. We have thus proved the following
-% 15.txt
-Fact 3.1.1 The subspace ran(\II) is invariant for M if and only if there exists T E R”? such that
-
-M\II = \IIT.
-
-Consequently,
-
-Fact 3.1.2 )\(T) Q ).(M)
-
-Proof of 3.1.2:
-
-z\E)\(T) _=_ 3v€R"3Tv=).v
-<=> \IITv = A\IIv
-<=> M\II-u = /\\IIv
-< /\ E /\(M)
-
-The second ¢> follows from fact 3.1.1.
-Facts 3.1.1 and 3.1.2 are theoretically useful. To see why, write the equation of fact 3.1.1
-
-as follows:
-
-\IIT = M\II
-\II‘\IIT = \II‘M\II
-T = (\1:'~1r)-‘\1:'M\1:
-
-In this form, we recognize that T = C‘1(0)C(1). That is, using only the covariance of observables
-on the state space, we can generate a matrix T with the property A(T) (_Z ).(M). Recalling that
-M = H1/2Pl'I‘1/2, we see that M is similar to our original stochastic matrix P, and thus ).(M) =
-
-).(P).
-
-\section{Approximate Invariant Subspace}
-
-We qualiﬁed the foregoing by stating that the facts are only theoretically useful. This
-
-is because when choosing observables of the state space we aren't sure that they will satisfy the
-% 16.txt
-primary assumption underlying facts 3.1.1 and 3.1.2. Recall the assumption: ran(\II) is an invariant
-subspace for M, where ‘II = H1/2(<I> —  It may well be the case that there is a 1p; 6 ran(\II) such
-that M1/zj ¢ ran(\II), thereby violating the assumption. However, lacking an invariant subspace, we
-
-might hope that our observables at least provide an approximate invariant subspace.
-
-Deﬁnition 3.2.1 If the columns of ‘II E R“? are independent and the residual matrix E E
-
-M\II —— \IIT is small for some T 6 R7”, then ran(\II) deﬁnes an approximate invariant subspace.
-
-To see how an approximate invariant subspace can be useful in approximating eigenvalues of M,
-we extract a theorem from Golub and Van Loan 
-% Theorem 3.2.1
-\begin{theorem}
-\label{thm-3.2.1}  
-Suppose M 6 Rd” and T 6 RP"? are symmetric and let E E MQ — QT, where Q 6 Rd"? satisﬁes
-
-Q‘Q = I. Then there exist p1,... ,;z,, E A(T) and A1,. . . , A,, E A(M) such that
-ll‘k - Akl S \/§|lE"
-
-fork=1,...,p.
-\end{theorem}
-
-If the subspace ran(Q) is an approximate invariant subspace, then the deﬁnition implies that
-there is a choice T rendering the error IIEII2 small, and thus, the eigenvalues of T provide a good
-approximation to those of M.
-
-When considering the preceding ideas, it becomes apparent that their application will
-present new — but hopefully less prohibitive — problems. As these problems summarize the rt of
-the paper, now is a good time to examine them. First, the preceding theorem assumes a matrix Q
-whose columns form an orthonormal basis for the approximate invariant subspace. For our problem,
-derivation of such a Q is tricky, and we must prepare for this. Next, having an approximate invariant
-
-subspace at our disposal merely tells us that there exists some matrix T which makes the error
-% 17.txt
-"EH2 small. We must discover the form of such a T. Moreover, it is natural to seek that T which
-minimizes IIEII2 for a given approximate invariant subspace. Finally, in order to apply these ideas
-to real examples of our eigenvalue problem, we must ﬁnd a practical way to generate the most
-appropriate approximate invariant subspace. We now address each of these issues in the order
-raised.
-
-Recall the matrix I1 E III/2(<I> — (<I>)). The columns of this matrix, though independent
-(by choice of independent observables), are not necessarily orthonormal. However, consider the
-polar decomposition ‘II = QZ where Q‘Q = I and Z2 = \Il‘\II is symmetric positive semideﬁnite.1
-
-Note that Z = (\II‘\II)l/2 is nonsingular, so Q has the form
-Q = ~1:z-1 = \II(\II‘\II)“/2
-
-and it is clear that ran(Q) = ran(\II). Perhaps ra.n(Q) is a. useful approximation to the invariant
-subspace for M. If ran(Q) is not itself invariant, we have the error E = MQ — QT. It is easy to
-show, as we do in the following proof, that the T which minimizes HE“; is T = Q‘MQ. This yields
-
-the following,
-Theorem 3.2.2
-
-If \I1 = QZ is the polar decomposition of \II, then the matrix
-T = Q‘MQ = (\1:*\1:)-1/’\1:‘M\I:(\Ir‘\I:)-1/2
-minimizes IIEII2 = IIMQ - QTII2
-
-Proof:
-We prove the result by establishing the following
-
-(Hal: Given M 6 Rd” suppose Q 6 Rd"? satisﬁes Q'Q =1. Then,
-
-min ||MQ — QTII2 = H(1- QQ')MQll2
-
-TERI‘?
-
-‘Recall that the polar decomposition is derived from the SVD, \II = UZJV‘, by letting Q = UV‘ and Z = VEV‘.
-
-% 18.txt
-and T = Q’MQ is the minimizer.
-To verify the claim, observe the following application of the Pythagorean theorem:
-
-For any T E R.”"” we have
-
-IIMQ — QT||§ = IIMQ - QQ'MQ + QQ'MQ - QT|l§
-||(I - QQ’)MQ + Q(Q'MQ - T) H3
-
-||(1 - QQ‘)MQ||§ + |lQ(Q'MQ - T)||§ (3-1)
-
-2 l|(I — QQ')MQ||§
-
-Equality (3.1) holds since I — QQ' projects MQ onto the subspace orthogonal to ran(Q). Thus,
-the two terms in the expression on the right are orthogonal, and the Pythagorean theorem yields
-equality. The concluding inequality establishes that the minimizing T is that which annihilates the
-second term in (3.1); i.e. T = Q‘MQ.
-
-The second equality in Theorem 3.2.2 is a consequence of the polar decomposition, in
-
-which Q = \IIZ“ = \I:(~I:‘\I:)-1/2. Therefore,
-T = (x1z'x1:)-1/2x1:‘M~1:(\1:*~1z)—1/2
-
-is the minimizer.
-
-3.3 The Krylov Subspace
-
-Suppose the columns of a matrix Q E R“? provide an orthonormal basis for an approxi-
-
-mate invariant subspace. Then we have seen,
-1. T = Q'MQ minimizes ||E||2 = ||MQ — QT||2 and
-2. E|p.1,...,p,, E z\(T) and /\1,...,A, 6 »\(M) such that
-
-ll-‘I: - Ml S \/§|lE||2
-
-% 19.txt
-19
-fork=1,...,p.
-
-Given an approximate invariant subspace ran(Q) of dimension 1), these facts tell us what matrix we
-should use to approximate p elements of M’s spectrum. Now all we lack is a. description of ran(Q).
-That is, we have not speciﬁed which approximate invariant subspace would best suit our objective
-
-of approximating the subdominant eigenvalue A,,.a2:(M). 'I‘o do so requires the following deﬁnition:
-
-Deﬁnition 3.3.1 The Raleigh quotient of a symmetric matrix M and a nonzero vector 3 is
-
-p(2:, M) E (:z:‘Ma:)/(z‘a:).
-
-We will also denote this by p(z) when the context makes clear what matrix is involved.
-
-To ﬁnd the approximate invariant subspace most appropriate for our problem, we choose
-each dimension successively, providing justiﬁcation at each step. Thus, start with one observable
-42 on the state space, and let zpl E 111/ 2(¢ — E,,¢). That is, 1/:1 represents a centered (mean zero)
-observable whose ith coordinate is weighted by . First notice that the deﬁnition (which comes
-directly from the deﬁnition of ‘II following equation (2.7)) is such that 1/)1 is a constant vector if
-and only if :15 is oonstant on the state space, in which case 1/); E 0. This fact provides one reason
-observables that are constant on the state space are not interesting. Second, recall that the row
-sums of the matrix P are all one and, therefore, the eigenvector corresponding to the eigenvalue
-Ao(P) = 1 is the constant vector. By deﬁnition of M E 1'11/2Pl'I‘1/2, we see that the constant vector
-is also the eigenvector corresponding to A0  This will play an important role in what follows, as
-it allows us to focus primarily on the subdominant eigenvalue z\max(M) E max{}q (M), [A44 
-rather than on A0  (which we already know is 1).
-
-Now, notice that
-
-lP(¢1,M)l 5 max |P(x,M)l = »\max(M) (3-2)
-% 20.txt
-where the max is taken over all non-constant vectors. Since our interest centers on An-.3_x(M), we
-would like a. 1/11 which makes the left hand side of (3.2) large. This would be achieved if 1/)1 were to
-lie in the space spanned by, say, the first two eigenvectors of the Markov chain (sometimes referred
-to as the slowest modes of the process). However, this subspace spans only two dimensions of the
-entire d-dimensional space, and it is more likely that ¢1, at best, only comes close to lying in the
-subspace of interest. Now, given 1/11, a judicious choice for the second dimension l/)2, and hence
-W; E [1/)1 1/:2], would be that which makes max,,¢o |p(\II2a)| is as large as possible. To establish that
-
-this is indeed the pertinent objective, note the following:
-
-_ a‘\1:5Mx1:2a
-I;1;g<Ip(\I'2a,M)l = 13:3: ————a,W§q,2a
-= M
-mrggf,mIp(z. )1
-5 maxlp(z,M)l (3-3)
-= ).,,m(M)
-
-Again, the max on the right side of (3.3) is over non-constant vectors. In words, we wish to chose
-\II-2 = [1/:1 ¢2] so that there is a vector a 6 R2 making |p(\II2a,M)| close to )«max(M).
-
-Now, p(1/11) changes most rapidly in the direction of the gradient Vp(¢1).
-
-W0  (iiiiiiv---viiiléi)
-= ,,;,l (Mam — p<¢.)«m (3.4)
-
-So, to maximize the left hand side of (3.3), ‘I12 should be chosen so that the subspace ra.n(\II2)
-
-contains the gradient vector. That is, we must choose 1/22 so that
-
-VP(¢'1) E F3l1{¢1y¢2} E F-'%!1(‘I’2) (3-5)
-
-Clearly, equation (3.4) implies Vp(1/:1) E ran{z/11,M1,b1}. Therefore, if ran{1/:1,1/)2} = ra.n{¢1,M¢1},
-
-then (3.5) is satisﬁed.
-% 21.txt
-In general, having chosen \II,, E [1/)1 - -- 1/11,] so that
-
-ran{'pl9¢21' " 1¢k} = ran{¢1vM1[)la ' ‘ ' vMk—11/’k}
-
-we must chose 1/1;.“ so that, for any non-zero vector u 6 R",
-
-VP(‘I’I:a) 6 ra‘n{¢11 ‘21 - ° - 31/Jk-F1} 
-Now,
-vp<~I~.a) = gmm — P(‘I’kﬂ)‘I’k“)
-
-and therefore,
-
-Vp(‘II;,a.) E ran(M\II;,) U ra.n(\IIk) = ra.n{¢1, M1/)1, . . . , Mktﬁl}
-
-Thus, requirement (3.6) is met when ran{1/:1, 1/:2, . . . , 1/4+1} = ran{1/)1, M1/:1, . . . , Mk:/11}.
-
-In conclusion, the p-dimensional approximate invariant subspace which is most suitable
-
-for our objective is
-
-/C(Mn/mp) 2 ran{«/:1,M¢1,... ,M"“«/21}
-
-This is known as the K rylov subspace. Therefore, in answer to the problem posed at the outset
-of this section, if we take the columns of Q to be an orthonormal basis for IC(M,z/)1, p), then the
-
-eigenvalues of T = Q‘MQ should provide good estimates of p extremal eigenvalues of M.
-
-% 22.txt
-\chapter{Lanczos Procedures}
-
-The following will review the usual procedure for generating an orthonormal basis for
-lC(M, «/11, p), and conclude by showing that, for our problem, we can find the matrix Q‘MQ without
-actually carrying out this procedure. Note that this conclusion is essential since the procedure
-
-requires matrix vector multiplication - an operation we have assumed impossible for large enough :1.
-
-4.1 The General Lanczos Algorithm
-Let 21;“) E M"¢ and consider the d x :1 matrix
-«I 2 [«/»,M«/»,  ,M"“¢] 2 [«b‘°>,«/»“’,...,¢<‘-‘>1
-
-Notice that M\II = [1/:(1),... ,1/;(d'1),Md¢r] and, assuming for the moment that \I1 is non-singular,
-
-we can compute the vector h = —\II‘1M"1/J. Thus,
-
-M\II = \II[e2,. . . ,e,1, —h] E IIIH
-
-% 23.txt
-23
-
-Equivalently,
-0 0  0 -—h1
-1 0  0 —h2
-HE   3 ; =\II“M\II (4.1)
-0 "hl1—1
-1 —h,;
-
-H is a companion matrix which means that its characteristic polynomial is p(:r) = 2:‘ + 23:1 h,-:r"‘1.
-Since H is similar to M, ﬁnding the eigenvalues of M is equivalent to ﬁnding the roots of p(a:).
-However, this is of little practical use since ﬁnding h, constructing 12(1), and ﬁnding its roots is
-probably a harder problem than the one we started with. Instead, the value of decomposition (4.1)
-derives from its upper Hessenberg form. We exploit this property below.
-
-Let ‘II = QR be the QR decomposition of \II. Since ‘I! is assumed nonsingular,
-~II;‘M~r.: = (R"Q‘)M(QR) = H
-Therefore,
-Q‘MQ = RHR-1 2 T
-
-Since R and R“1 are both upper triangular and H is upper Hessenberg, T = R.HR'1 is also upper
-Hessenberg. Furthermore, since M is symmetric, it is clear that T‘ = Q‘MQ = T. That is, T is
-
-an upper Hessenberg matrix that is symmetric. Therefore, T is tridiagonal and we can write it as
-
-follows:
-
-T =    (4.2)
-
-ﬁd—1
-
-0  174-1 ad
-
-% 24.txt
-24
-
-Equating columns j on both sides of MQ = QT yields
-M113‘ = B,-—1qj—1 + 0:91 + ﬁjqj+1 (43)
-
-Since the columns of Q are orthonormal, multiplying both sides of (4.3) by q_,- and q,-+1 yields
-oz, = q§Mq,- and ﬂj = q;-+1Mq,-.
-The foregoing justifies what is called the Lanczos algorithm, performed as follows:
-
-The Lanczos algorithm for (partial) reduction to symmetric tridiaganal form
-
-91 = ¢/||¢||2, [30 = 0, 40 = 0
-
-for j=1top
-
-z=Mq_,-
-
-5:‘ = ||Z||2
-if 5, = 0, quit
-¢Ij+1 = 2/51
-
-end for
-
-The qj computed by the Lanczos algorithm are often called the Lanczos vectors. If the
-loop in the algorithm is terminated because ﬁg = 0, this indicates that an exact invariant subspace
-has been computed, and is given by ran{q1 - - - qk}. Otherwise, we usually halt the algorithm after
-p steps, in which case the algorithm converg to approximations of at most 12 eigenvalues.‘
-
-Note that equation (4.3) results from the equation MQ = QT. The latter equation holds
-since, starting with d vectors in our subspace, Q is a d Xd matrix whose columns form an orthonormal
-
-basis for all of R‘, which is clearly an invariant “sub” space. The Lanczos algorithm above, however,
-
-‘A detailed discussion of such convergence (and misconvergence) is given in
-
-% 25.txt
-25
-
-proceeds for only p steps, producing a d x p matrix Q, whose columns form an orthonormal basis
-for the approximate invariant subspace IC(M, ¢, p), and a p X p matrix T, = Q'MQ. In that case,
-IIMQ = QTp|[g =  is non-zero and gives the error bound described in Theorem 3.2.1. Now,
-
-writing the full d x :1 matrix T of (4.2) as
-
-01 51
-/71
-
-(4.4)
-
-allows us to describe the error bound ||E||-2 in terms of the submatrix Tm, and, hence, in terms of
-
-ﬁzz-
-Theorem 4.1.1
-
-HT, and T,“ are the matrices appearing in (4.4), and if the p columns of Q are computed by the
-
-Lanczos algorithm, then there exist pl, . . . ,;t,, E /\(T,) and A1, . . . , A, E A(M) such that
-
-ill» - /\I=i S iiTP'4”2 = 512
-
-fork=1,...,p.
-
-For a proof, see page 365 of [1].
-
-% 26.txt
-26
-
-4.2 A Lanczos Procedure for Markov Chains
-
-Notice that the Lanczos algorithm requires the computation of Mqj. Throughout this
-paper we have assumed that M 6 Rd“, and that d is so large as to make the matrix vector
-multiplication Mqj impossible. Now, the Lanczos algorithm constructs an orthonormal basis for
-the Krylov subspace. For our purposes, this construction is not essential. What is essential is that
-we ﬁnd a way to generate the Lanczos coejﬁcients oz,-, ﬁj, (j = 1, . . . , p), and whence the matrix
-T, without performing the operation Mqj required by the Lanczos algorithm. We now address this
-problem.
-
-Begin with the centered and weighted observable 1/1 = 111/ 2 (qS—E,,¢), and let q1 = 11;/||1/JH2.
-To simplify notation, let ¢(n) E ¢(X,.). The form of the first Lanczos coeﬂicient a1 is straight
-
-forward:
-
-01 = qiMq1
-
-I/1‘M1/I
-¢’¢
-
-and now the special properties of the Markov chain setting play an important role. For, regard the
-
-following
-
-¢‘M«/2 = (¢ — E.¢)‘II%‘MII%(¢ - Em)
-
-Cw (d>(n), ¢(" + 1)) (4-5)
-
-and
-
-W1!’ = (<15 - E«¢)'H(¢ - E«¢)
-
-Varn (¢(")) (4-6)
-
-Recall that the deﬁnition of M implies H%‘MI'I% = HP, and this justiﬁes equation (4.5). Putting
-
-% 27.txt
-27
-
-these equations together, we have a. nice form for :11:
-
-_ c, (run), an + 1))
-"‘ ' Ta:,(¢(n))
-
-The next Lanczos coeﬂicient, B1, is only slightly more complicated.
-
-[31 = ||M91-0t1tI1||2
-
-[(Mq1 — 0l1¢11)i(M¢]1 - 0101)] 1/2
-
-[t1iM2¢I1 — ail 1/2
-
-(4.7)
-
-Again recalling the deﬁnition M E 111/ 2P1'I‘1/2, a. simple manipulation veriﬁes
-
-1'li‘M’l'[% = HP’, whereby it follows that
-
-(«s — E,,¢)‘HP”(¢ — Em)
-
-"‘M2"‘ (¢ — E,¢)*n<¢ — Em)
-
-C: (¢("), ¢(n + 2))
-V3-Ur
-
-Inserting the last expression into equation (4.7) gives
-
-_ c,(¢(n).¢<n+2)) c,(¢<n>,¢(n+1)> ’ "’
-"“ jvar,<¢<n)> "(jvar,<¢<n)) ‘)
-
-Notice that we can estimate these expressions of oz; and ﬂl by running simulations of the Markov
-chain. To do so, we run the Markov chain starting from a random initial state, compute the value of
-the observable at each step of the chain and, from the data, compute estimates of the covariances
-and variance of the observable.
-
-Before generalizing this for the jth Lanczos coefficient, we consider one more coeﬂicient
-in detail. Recall that the second basis element q; in the Lanczos algorithm is given by 1;; =
-
-(Mql — oz1q1)/ﬁl. From this we can derive an expression for (12 which, though not as neat as those
-
-% 28.txt
-28
-
-for oz; and [31, is nonetheless tractable.
-
-I12 ‘IiMq2
-
-1
-[7(M91 — 0t1¢I1)tM(M¢I1 - 11191)
-1
-
-1
-= f,§(qiM3q1 - 2a1qlM2<11 + a’9iM¢11)
-1
-1
-= —2(<1iM3q1 - 2015i ‘ all (4-8)
-
-1.
-The last equality follows from equation (4.7), which can be written as q{M2q1 = off’ + (if. The
-preceding paragraphs deﬁne all but one of the quantities in the expression (4.8). The one remaining
-is q§M3q1, which can also be expressed in terms of covariances of <15, and the analysis is almost
-
-identical to that for q{M2q1:
-
-(¢ — E,¢)'IIP3(¢ — Em)
-
-"lM3"‘ (¢ — E,¢s)‘H(¢ - Em»)
-
-Cu (¢(")a ¢(n + 3))
-Varn(¢(n))
-
-To generalize the foregoing for higher order Lanczos coeﬂicients requires a recurrence
-relation between (aj+1, ﬂj) and (aj, ﬂ,~_1). As we will see, such a relation involves higher order
-Lanczos vectors q,~ — vectors we do not wish to compute. Recall that the only “vector” we can
-handle is our initial observable. Therefore, we must also establish a recurrence among the q,-’s, thus
-making them available inductively from our observable vector.
-
-To begin with the ﬂfs, recall that ﬁj E ||Mqj — ajq_.,- — ,H_,~_1q,-_;|[2. Whence,
-
-Bf (M411 — °‘j‘Ij - ﬁj-1qj—1)’(MlIj - Otjqj - 51-19141)
-
-q;-M291‘ - 0? - B,z_1 (4-9)
-
-Some simple algebra and the deﬁnitions aj E q§Mqj and ﬁJ~_1 E q;-Mq,--1 prove equality (4.9). The
-% 29.txt
-recurrence for 0114.1 is equally straight forward, though the ﬁnal form is not as neat:
-
-2
-0tj+1 ¢Ij+1M'1j+1
-
-1
-§(M¢15 - 11111;‘ - ﬂj—1<Ij—1)'M(Mqj - Otjqj - /3j—1qj—1)
-J
-
-1
-§(q;-M-“*4; — 2a,~q;M*q,- — 2x3;+’_1q;--1M*q,- + «I? + 2a,-5,51 + am (4.10)
-J
-From (4.9), q;~M2q,~ = oz? + ﬂf + ﬂ}_1 and, upon plugging this into (4.10), we arrive at
-1
-a,~+1 = E(q§»M3q_,,- — (1? — 2ozJ~ﬁJ2 — 2ﬂj_1q;»_1M2qj + ﬁ2_1) (4.11)
-J
-
-The expressions above involve the higher order Lanczos vectors qj. In the present context,
-these vectors will not be readily available. However, another recursion — this time relating forms
-involving q,-+1 to forms involving qj and qj__1 — will make these expressions accessible by induction
-on our initial observable. The general forms of interest are q;+1M"qj+1 and qj-+1M"qJ~. Again, we
-
-begin with the simpler of the two:
-
-¢I;'+1Mk9j = ij(Mqj - Otjqj - ﬂj—1t1j—1)'M"¢1j
-= i(<1§M"“¢Ij - C‘j‘I_;'Mk‘Ij - ﬂj—1q§M"qj-1)
-The second is,
-‘I_;'+1Mkqj+1 = ﬂi?(M<Ij - Otjqj - .3j—1<Ij—1)‘Mk(Mqj — Orjqj - ﬂj—1¢1j—1)
-
-1
-ﬁ(‘1§M'°"<1:‘ + °‘:2"I§M'°<J:' + ﬁ?—z<Ii—1M"<I:‘—1
-J
-—2a,-q;-M*+‘q,~ + 2am,--1q;-M*q,--1 — zza,--1q;-M*+‘q]--1) (4.12)
-
-The point of all this is that each pair of coeﬂicients (oz,-+1, /3,-) can be expressed as a
-function of the covariance between ¢(n) and ¢(n+ m), for m = 0, . . . , 21' + 1. Since the expressions
-are based on four recurrence relations, they become quite complicated as 1' gets large. However,
-deﬁning these recursions in a. computer program is trivial, and we use this fact in Section 5.4 when
-
-applying the foregoing to a particular problem.
-
-% 30.txt
-30
-
-Chapter 5
-
-An Application: Convergence to
-
-Gibbs Measure on the Ising Lattice
-
-5.1 Physical Motivation
-
-Before considering the details of the Metropolis algorithm, it helps to understand the
-setting in which it was developed. To do so we consider an typical example, the Ising model.
-Consider a model of a system consisting of 71 “sites”, each site taking values in {-1, +1}. It might
-help to imagine these sites as equally spaced points on a line or a circle, though we are not restricted
-to such cases. Then our state space 5' consists of all possible n—dimensional permutations of -1’s
-and +1’s. We reprent this as S’ = {-1, +1}", and note that |S| = 2". We can think ofeach site as
-a node existing in one of two possible positions, say “on" or “off”. Each state 1' E S is represented
-by a unique permutation a.- of —1’s and +1’s. Let a,-(k) = :|:l denote the position (on or off) of
-the kth node when the system is in state 1'. An observable ¢ = qS(a,-) on this space is a function of
-
-the permutations 47,-, 1' 6 {1,. . . , 2"}.
-
-% 31.txt
-31
-
-The energy of the system, when in state i, is deﬁned by the Hamiltonian
-
-H07-‘) = - Z Ua(i)0s(k)
-
-<j Ic>
-where the sum runs over all neart neighbor pairs. In statistical mechanics we are often concerned
-
-with the Gibbs measure of state i, deﬁned by
-1r(a,-) = Z'1 exp{—[3H(a,~)} (5.1)
-where Z ‘1 is the normalization constant, or partition function. For the Ising model, we put
-
-Z = §:9XP{~ﬁH(<7:‘)}
-i=1
-
-so that 7r is a probability measure. Now, letting 1: denote the so called Boltzman constant, if a
-classical mechanical system is in thermal equilibrium with its surroundings with absolute temper-
-ature T, and is in state i with energy H (05), then the probability density in phase-space of the
-point representing state i is given by (5.1) with [3 = (lcT)“. The central result of ergodic theory
-implies that we can also interpret 1r(o;) as the proportion of time the system spends in state i. If
-
-the system is observed at a random time, the expectation E,,¢ of any observable ¢ is thus
-
-.31 ¢(0a) 9"l>{-ﬂH(0s‘)}
-
-E” = 2?;.exp{—nH(a.-)}
-
-(5.2)
-
-Perhaps the number of sites n in our Ising model is so large that it is impractical or impossible to
-evaluate (5.2). We might instead consider importance sampling and generate states with probability
-
-density
-
-exp{—aH<a.o}
-2?;.exp{—zaH<a.->} (53)
-
-Then «I: is itself an unbiased estimator of (5.2).
-
-1r(0a') =
-
-If evaluation of (5.2) is difficult, there is no reason to believe that evaluation of (5.3) is any
-
-easier. However, Metropolis and his collaborators [9] contrived a method for producing an ergodic
-
-% 32.txt
-32
-
-Markov chain with transition probability matrix K whose elements Ic(:::, y) satisfy
-
-2«<z>k<x.y) = «<2»
-
-I
-
-By Deﬁnition 2.1.4 this means that the Markov chain has stationary distribution 7r. Since the
-drain is ergodic, we know that it will eventually converge to the desired distribution. So, our main
-problem — the rate at which such a Markov chain will converge — is clearly at issue in the application
-of the Metropolis algorithm. Moreover, as we will see below, the Metropolis algorithm ensures that
-the Markov chain has the desired stationary distribution by requiring that the chain satisfy the
-stronger condition of reversibility; i.e. detailed balance. Therefore, Markov chains produced by the
-Metropolis algorithm satisfy the assumptions of this paper, which fact conﬁrms that our theory
-
-can be used to bound convergence rates of all such chains.
-
-5.2 Description of the Metropolis Algorithm
-
-The following explains how the Metropolis Algorithm is carried out. The exposition is
-similar to that given in Hammersley and Handscomb’s book Monte Carlo Methods 
-To begin, choose an arbitrary symmetric Markov chain; that is, a Markov chain whose
-
-transition probability matrix P is symmetric. P is called the proposition matrix, and its elements,
-P071"), Satisfy
-pa, 1) 2 0. pm‘) = pm). 2,-pm‘) = 1 vi e S (5-4)
-
-Using P, we will derive a new transition probability matrix K which satisﬁes
-
-Z vr(z')k(='.:') = arm (5.5)
-
-I
-
-For 1' aé j deﬁne
-
-kw) = p<-yum")/«<¢> it nu)/1r(='> <1 (5.6)
-
-P(i,J') if 7r(J')/7r(i) Z1
-
-% 33.txt
-33
-
-For 2' = j deﬁne
-
-kw) = pm) (1—  p(-3;"), where J. = {j = 2‘ 7‘ an nu)/«(='> < 1}
-
-Notice that 1r(i) > 0 implies k(i,j) _>_ 0 and
-
-;k<«',j> = pow)  { (1 —  pm) +  +]§§p<«",:')
-= P(is'5) + 2 P(i1.1') + Z P(iy.7') (5-7)
-jel.» jug
-
-Equation (5.7) shows that, for each i, 2, k(z', j) = 1, which fact conﬁrms that K is a stochastic
-
-matrix.
-
-Next we check that K satisﬁes the detailed balance condition. First suppose that 1r(z') =
-
-1r(j). Then (5.4), (5.6), and symmetry of P imply
-k(i,J') = p(i2.7') = p(jvi) = k(j1i)
-
-and, since 1r(i) = 1r(j), the detailed balance condition, 1r(i)k(i,j) = 7r(j)k(i,j), clearly holds. Now
-
-suppose that 1r(i) > 1r(j). Then (5.4) and (5.6) imply k(j,i) =p(j, 1') and
-k(i,J') = P(iyJ')7r(J')/7r(i) = P(J3i)7r(J')/7r(i) = k(.1'» i)7r(J')/1r(i)
-
-That is, 1r(i)lc(i,j) = 1r(j)k(i,j), as we set out to prove.
-To complete the justification of the Metropolis algorithm, we note that the detailed balance
-condition, 7r(i)k(i, j) = 1r(j)k(i, j), implies that 1r is indeed the stationary distribution for a Markov
-
-chain with transition probability matrix K. To see this notice that
-
-27r(i)k(i..7') = Zﬂijlkiiajl = 1r(J') Z/€(i,J‘) = 7r(J')
-
-I i
-We now summarize the tasks performed when implementing the Metropolis algorithm.
-
-First, pick as a proposition matrix any symmetric stochastic matrix and denote it by P. Begin
-
-% 34.txt
-34
-
-the simulation of the Markov chain in state i, and then take one step of the proposition Markov
-chain to arrive in state j. That is, choose j according to the mass function p(z', -). Next, compute
-1r(j)/1r(i). If 7r(j)/1r(i) 2 1, accept j as the new state. If1r(j)/1r(i') < 1, then with probability
-1r(j)/7r(z') accept j as the new state; otherwise, with probability 1 — 1r(j)/1r(i), take i as the new
-state. Performing each of these tasks produces one step of the Markov chain represented by the
-
-transition probability matrix K.
-
-5.3 The Ising Model
-
-Returning to the Ising model considered in section 5.1, we consider how the Metropolis
-algorithm is applied to such a mode]. Again, start the simulated chain in state X0 = 1' and let
-
-X1 = j be a state chosen according to p(i, -), where p(i, j) = p(j,i). Next, compute
-
-7"(‘7j)/7"(‘7i) = eXP{-ﬁlH(<7j) - H(0i)l}
-
-Suppose the energy H (Uj) of the new state satisﬁes H (U,-) 3 H (05). Then 1r(a_,-)/1r(a,-) Z 1 and we
-
-move to state j. On the other hand, if H(aj) > H(a;), then 1r(aj)/7r(a,-) < 1 and we take
-
-j with probability e-M"
-
-X1 (5.8)
-
-i with probability 1 — e-W’
-where AH E H (0,) — H (:75) is the change in energy from state j to state 1'.
-There are many ways to choose a proposition matrix P. The simplest derives from the
-so called Glauber dynamics, which proceeds as follows: choose one of the 11. sites at random with
-probability 1/n — let .9 denote the chosen site — and change the position of this site; i.e., multiply
-
-its value by -1. The resulting state j is the permutation of -1’s and +1’s given by
-
--0,-(k) if]: = .9
-a,~(Ic)= (k=1,...,d)
-
-a,-(Ic) if]: 95 s
-
-% 35.txt
-35
-
-The procedure we employ below involves a slight variation: choose site .9 at random uniformly
-from the 11 possible sites. However, instead of changing the value of the chosen site with absolute
-certainty, as in Glauber dynamics, we “ﬂip the switch” with probability 1/2. Therefore, when k yé .9
-
-we have 0',-(k) = a.~(k), and when k = 3,
-
-—a,~(s) with probability 1/2
-01(3) =
-
-a,-(3) with probability 1/2
-The advantage of this procedure is that the proposition matrix, before altering it with the Metropo-
-lis algorithm, is a bit more interesting than that produced by traditional Glauber dynamics. For,
-it has 1/2 along the main diagonal and, thus, is aperiodic and converges to a stationary distribu-
-tion. Glauber dynamics has period 2, so its proposition matrix does not satisfy the hypotheses of
-
-our convergence theorem (Theorem 2.1.2). Keep in mind, however, that any proposition matrix
-
-modiﬁed by the Metropolis algorithm satisﬁes all of our hypotheses.
-
-5.4 Computations
-
-5.4.1 Computer Programs
-
-Constructing a computer program to simulate realizations of the Markov chain described
-above (i.e. with transition matrix K) is straight forward, and we do so using Matlab and the
-program ising.m which appears in the Appendix.
-
-When performing simulations in the present context, there are a few important aspects
-to consider. First, our program begins with a “hot start", which means that the initial state is
-picked by assigning each site to :l:1 with probability 1/2. Perhaps we should instead have picked
-our initial state from the stationary distribution. On the contrary, this would defeat the whole
-
-purpose of our analysis. Recall that the method has been developed to deal with cases for which
-
-% 36.txt
-36
-
-sampl from the stationary distribution are not quickly attainable. Furthermore, we are trying to
-ﬁgure out how long it will take for observations from the simulated chain to represent samples from
-the stationary distribution. Second, the method we have developed depends solely on covariances
-of our observable which we estimate from the data. produced by the simulations. However, such
-estimates require that the data come from the stationary distribution. Therefore, we must discard a
-number of observations before using the data to estimate covariances. The number of observations
-to discard depends on how long must we wait until we have data from the stationary distribution,
-and we are back where we started — almost. When employing our procedure, we can discard a
-conservative (very large) number of observations and perform our analysis, deriving a bound on
-the convergence time. Then, in all future studies, we will know how much data should be discarded
-before samples come (approximately) from the stationary distribution.
-
-In considering what observable to use on the state space described above, perhaps the
-most obvious is simply the number of nodes in the on position (i.e. the number of +l’s). Recalling
-that 0, (k) = i1 denotes the position (on or off) of the kth node when the system is in state j, this
-
-observable is deﬁned by,
-«M = zxca.-ac) = +1). 1 = 1.---.ISl
-I:
-
-where X is the indicator function. The program ising.m simulates the Markov chain, computes the
-values of this observable (as well as the number of 1’s squared) and writes it to a ﬁle called phi . dat.
-The program then proceeds to compute the ﬁrst few Lanczos coefficients using covariances produced
-by Matlab, so that we can check them against the rults we get from the 1anczos.c program,
-which we now describe.
-
-The lanczos .c program (listed in the Appendix along with its dependencies) implements
-the recursive relations described in section 4.2 to compute the Lanczos coefﬁcients by the new
-
-method. It accepts input from the user specifying any observable, any number of iterations, and any
-
-% 37.txt
-37
-
-TABLE 1: ESTIMATES or /\,m(K) = 0.998746 USING ¢1
-
-sim. # /32
-1 0.995107 0.006960 0.995220 0.600243
-2 0.995099 0.005891 0.994863 0.477393
-3 0.994862 0.009671 0.994925 N/ A
-4 0.995321 0.009319 0.995654 N/A
-5 0.994787 0.004872 0.405446 0.994742
-
-number of desired Lauczos coeﬁicients. That is, the observable could have come from a simulation
-of any reversible Markov chain and not just our special Metropolis chain. The output of this
-program is a T matrix containing the maximum number of Lauczos coeﬂicients (up to the number
-requested by the user) before an approximate invariant subspace was reached. The eigenvalues of
-the nonzero part of this matrix, in accordance with the foregoing theory, should well approximate
-
-eigenvalues of the Markov chain’s stochastic matrix.
-
-5.4.2 Results
-
-Table 1 shows the results of 5 simulations each of 100,000 iterations (discarding the ﬁrst
-5,000 observations) for the one dimensional Ising lattice with 10 nodes and [7 = 1. Displayed is
-the estimate of the ﬁrst Lauczos coeﬂicient a1 (which is the eigenvalue estimate after one step; i.e.
-).m,x(T1)), along with its error bound ﬁl. Appearing in the third column is the largest eigenvalue of
-the 2 x 2 matrix T2, A0 (T2), and ﬁg appears in the fourth column when it is was computable. When
-it was not computable, it was probably because an approximate invariant subspace was reached
-at the ﬁrst step, so ﬁg involved expressions which were close to machine epsilon and, thus, could
-not be computed accurately. Table 2 shows the same information when using a second observable,
-¢2 E ¢§ (i.e. the number of +1’s squared).
-
-The true value of the subdominant eigenvalue for this problem is /\max(K) = 0.998746,
-
-and the next largest is A2(K) = 0.993303. These were computed directly from the transition matrix
-
-% 38.txt
-38
-
-TABLE 2: ESTIMATES OF >.,,,,,.,,,(K) = 0.998746 USING z/>2
-
-sim. # B2
-1 0.994262 0.006142 0.994222 N / A
-2 0.994333 0.002974 0.994332 N / A
-3 0.994307 0.008892 0.994374 0.726320
-4 0.994436 0.012884 0.994685 0.361404
-5 0.993789 0.006909 0.993820 0.788710
-
-using the Matlab program tpm.m listed in the Appendix. They should be very accurate, and it
-takes the Matlab routine eig() just over 20 minutes of cpu time on a Sun Ultra to ﬁnd all the
-eigenvalues of K.
-
-As shown by Theorem 4.1.1, the [Ts provide conservative error bounds on the eigenvalue
-estimates from the previous Lanczos step. Unfortunately, for this example, ,6 appears to be too
-conservative as it does not even allow us to bound our estimates away from 1. Furthermore, it
-seems that the estimates generated by the method above fall systematically below the true value of
-the subdominant eigenvalue. In every simulation, an approximate invariant subspace (to machine
-tolerance) was reached after computing at most 2 pairs of Lanczos coefficients. For ¢1 we see that
-it was reached after computing only one pair on simulations 3 and 4, similarly for ¢2 on simulations
-1 and 2. These observations indicate that the chosen observable is close to the slowest mode (or
-eigenfunction) of the process. That is, we should observe fast convergence on an approximate
-invariant space. However, the space found might contain only the second slowest mode of the
-process and, in that case, our eigenvalue estimates would be closer to the third largest eigenvalue,
-instead of the second largest as desired.‘
-
-The matrix T was produced by the program lanczos . c and its eigenvalues were computed
-by Matlab, both operations taking a few seconds. Granted, comparing a few seconds to the 20
-
-minutes it takes Matlab is not justiﬁed since a simple call to the routine eig() is far from the
-
-‘Recall that we explained above why we will never converge to the eigenspace corresponding to the largest eigen-
-value (which is always 1). That would require a (nearly) constant observable function.
-
-% 39.txt
-39
-
-most eﬂicient traditional method for computing only a few eigenvalues of K. However, recall that
-the primary motivation for the new method are those examples where the matrix K is so large
-that we can’t store even a single vector in fast memory. Once we start swapping data to and from
-cheap memory, the time required by traditional algorithms can become years, even lifetimes. Such
-examples are not exceptional. Consider, for instance, the 1000 node problem which produces a.
-
-2“’°° x 21°°° transition matrix.
-
-% 40.txt
-40
-
-Chapter 6
-
-Conclusion
-
-We conclude by noting a few aspects of the special cases to which the foregoing applies.
-We are interested in the subdominant eigenvalue of a large transition probability matrix. The
-
-foregoing can be used to approximate these eigenvalues when the following conditions are satisﬁed:
-o The Markov chain is reversible (detailed balance condition).
-
-0 We can easily simulate realizations of the chain as well as functions (observables) of these
-
-realizations.
-
-When these conditions hold, we can side step the traditional Lanczos algorithm, and get at the
-values of the Lanczos coeﬂicients via the variational properties of observables on the state space.
-There are many ideas in this paper that should be expanded upon in future studies. Per-
-haps most obvious is the desire to ﬁnd analogous techniques for non-reversible (i.e. non-symmetric)
-Markov chains. This is a considerable problem since the Lanczos methods described above are no
-longer applicable. Perhaps the Amoldi algorithm ([1], algorithm 6.9), or the non-symmetric Lanczos
-
-algorithm can also be usefully specialized for stochastic matrices.
-
-% 41.txt
-41
-
-When applying the Lanczos algorithm, we usually choose a single starting vector (in the
-present context, an observable), perform the algorithm for a number of steps, and gather results.
-Then we do the same with a new starting vector and compare the results to those obtained with
-the ﬁrst vector. Proceeding in this way for a number of starting vectors, we can provide statistical
-evidence that we have, indeed, located the eigenvalues of interest. However, when there are two
-observables 451, :15; of particular interest, perhaps we would beneﬁt by considering both observables
-simultaneously. Future studies might develop an algorithm which incorporates both observables,
-
-again via ¢.- = H1/2(¢.' - E,.¢.-). In this case, a Lanczos algorithm would be based on the space
-
-Ic*(M,¢1,¢»2.p) 2 ran{«/»mz»2,M«/»1,M«/:2, . . .,MP—‘¢1.M"—‘«/22}
-
-The Lanczos coeﬁicients would then involve not only the autocovariances C, (¢,<(n),¢,-(n + m)),
-
-but also the cross-covariances C, (¢.~(n), ¢j(n + m)) (deﬁned in (2.6)).
-
-% 42.txt
-Appendix: computer programs
-
-Z Hatlab code ising.m
-
-Z
-Z Written by William J. De Meo on 12/15/97
-
-2 last modified 1/10/98
-
-Z Inputs:
-
-2 d = number of nodes of the ising lattice (e.g. d=100)
-Z n = number of iterations (e.g. n=1000)
-
-Z beta = Annealing schedule (e.g. beta = 2,
-
-1 when beta -> 0 will always go to nee state
-Z vhen beta -> infty will never go to state of higher energy)
-Z
-
-X = zeros(d,1);
-Energy = zeros(n,1);
-
-Ave=0;
-
-X Initialize using hot start
-for i=1:d,
-
-if rand < .5
-
-x(i) = -1;
-
-else
-
-X(i) = 1;
-
-end
-end
-
-Z Initialize at state of high energy (not used)
-
-if 0
-X(1:2:d-1)=1;
-X(2:2:d)=—1;
-and
-
-Z Initialize energy
-
-H=0;
-
-for i=1:(d-1)
-
-H = H - X(i)*X(i+1);
-end
-
-2 Initialize observable
-phi = zeros(n+1,1);
-
-for i=1:d
-if X(i) == 1
-phi(1)= phi(1)+1;
-end
-
-end
-
-42
-
-% 43.txt
-43
-
-for j=1:n
-
-site=0;
-while site==
-
-site = round(rand#d);
-end
-Z Compute new energy (with new X(site) = —x(site)):
-if site ==
-
-Hnew = H + 2:x(1)~x(2);
-elseif site == d
-
-Hnew = H + 2*X(d)tX(d-1);
-else
-
-Hnew = H + 2*X(site)*(X(site-1) + X(site+1));
-end
-
-Z Change to new state in two ways:
-X with probability 1/2, flip the switch
-if rand < .5
-if (Hnew <= H) I (rand < exp(-beta*(Hnew-H)))
-x(site) = -x(site);
-H = Hnew;
-Z Compute new value of observable
-if X(site) -= 1
-phi(j+1) = phi(j) + 1;
-
-else
-phi(j+1) = phi(j) — 1;
-end
-else phi(j+1)=phi(j);
-end
-else phi(j+1)=phi(j);
-end
-Energy(j) = H;
-end
-p1ot(Energy)
-
-% A second observable
-phi2 = zeros(n+1,1);
-for i=1:n+1
-
-phi2(i) = phi(i)*phi(i);
-end
-
-Z Write first observable to phi.dat:
-
-fid = fopen(’/quean.a/masters/deme5025/research/nyuthesis/programs/c/phi.dat’,’w’);
-fprintf(fid,’%f\n’,phi);
-fc1ose(fid);
-
-Z Write second observable to phi2.dat:
-fid = fopen(’/quean.a/masters/deme5025/research/nyuthesis/programs/c/phi2.dat’,’w’);
-
-% 44.txt
-fprintf(fid,’%f\n’,phi2);
-fc1ose(fid);
-
-1 Compute a few Lanczos coefficients by new method
-START=1001;
-
-V1=zeros(2);
-
-C1=zeros(2);
-
-C2=zeros(2);
-
-C3=zeros(2);
-
-V1 = cov(phi(START:n+1),phi(START:n+1));
-
-V1 = V1(1,1)5
-
-C1 = cov(phi(START:n),phi(START+1:n+1));
-C1 = C1(1,2);
-
-C2 = cov(phi(START:n-1),phi(START+2:n+1));
-C2 = C2(1,2);
-
-C3 = cov(phi(START:n-2),phi(START+3:n+1));
-
-C3 = C3(1,2);
-
-a1pha1 = C1/V1;
-
-betai = sqrt(C2/V1 ' (C1/V1)‘2);
-
-a1pha2 = (1/(beta1’2))*(C3/V1 - 2*a1pha1*(C2/V1) + a1pha1‘3);
-
-T1 = Ealphal betal; betal a1pha2]
-
-v1 = cov(phi2(START:n+1),phi2(START:n+1));
-v1 = v1(1,1);
-
-c1 = cov(phi2(START:n),phi2(START+1:n+1));
-c1 = c1(1,2);
-
-c2 = cov(phi2(START:n—1),phi2(START+2:n+1));
-c2 = c2(1,2);
-
-c3 = cov(phi2(START:n-2),phi2(START+3:n+1));
-c3 = c3(1,2);
-
-alphal = C1/V1;
-betal = sqrt(C2/V1 - (C1/V1)‘2);
-a1pha2 = (1/(beta1‘2))#(C3/V1 - 2ta1pha1*(C2/V1) + a1pha1”3);
-
-T2 = [a1pha1 betai; betal a1pha2]
-
-fid = fopen(’Tmats.dat’,’v’);
-fprintf(fid,’%f\n',T1);
-fprintf(fid,’%f\n’,T2);
-fc1ose(fid);
-
-ZZZ end ising.m
-
-44
-% 45.txt
-\% Matlab code tpm.m for generating transition probability matrix
-Z and computing its eigenvalues directly
-
-1
-
-Z Written by Hilliam J. De Heo on 1/10/98
-
-1
-
-Z Inputs:
-
-Z d = number of nodes of the ising lattice (e.g. d=10)
-
-Z beta = Annealing schedule (e.g. beta - 2,
-
-1 when beta -> 0 will always go to new state
-1 when beta -> infty will never go to state of higher energy)
-2
-
-disp(’bui1ding proposition matrix...’)
-
-states = 2‘d;
-
-A = 0;
-
-for i=0:(d-1)
-
-E=eye(2”i);
-
-A = [A E;E A];
-
-end
-
-B = .5#eye(states);
-
-A = B + .5-(1/d)*A;
-
-2 our modified Glauber dynamics requires the B and the .5¥(1/d)
-
-disp(’...done’)
-
-1 display pattern of nonzero entries
-1 spy(A)
-
-Z check that all row sums are 1
-check=O;
-for i=1:states
-
-check = check + (not(sum(A(i,1:states))<.99));
-end
-1 if not all rows sum to 1, print check = (# of rows with sum=1)
-if not(check==states)
-
-check
-
-error(’Row sums are not all 1’)
-end
-
-Z construct matrix of states
-E = zeros(d,states);
-f1ip=-1;
-for i=1:d
-for j=1:2‘(i-1):states
-flip = -1*f1ip;
-for k=0:2“(i-1)
-
-45
-
-% 46.txt
-E(i,j+k) = flip;
-and
-end
-end
-E=E’;
-
-Z display first 64 states
-Z E(1:64,:)
-
-Z compute energy of each state
-H = zeros(states,1);
-for i=1:states
-
-for j=1:d-1
-H(i) = H(i) - E(i,j)tE(i,j+1);
-end
-end
-
-disp(’building Hetropolis transition matrix...')
-
-for i=1:states
-for j=i+1:states
-if not(A(i,j)==0)
-if(H(j)>H(i))
-alt = A(i,j)t(1-exp(-beta*(H(j)-H(i))));
-A(i,i) I A(i,i)+a1t;
-A(i,j) = A(i,j)-alt;
-elseif(H(j)<H(i))
-alt = A(j,i)t(1-exp(—beta#(H(i)-H(j))));
-A(j,j) = A(j,j)+a1t:
-A(j,i) = A(j,i)-alt;
-end
-end
-end
-end
-
-disp(’...done’)
-
-check=0;
-for i=1:states
-
-check = check + (not(sum(A(i,1:states))<.99));
-end
-
-Z if not all rows sum to 1, print check = (# of rows with aum=1)
-
-if not(check==states)
-
-check
-
-error(’Row sums are not all 1’)
-end
-
-46
-
-% 47.txt
-47
-
-disp(’computing eigenvalues of tpm...’)
-cput = cputime;
-
-evals = eig(A);
-
-ecput = tputime - cput;
-
-disp(’the CPU time (in secs) for computing eigenvalues of tpm: ’)
-
-ecput
-ZZZ end tpm.m
-
-/¥#*##¥###*##*#l####¥#‘###$ﬂ###ﬁi##*U3*#**#*##**##¥***ll¥#*##¥ii
-
-* 1anczos.c main program for computing Lanczos coefficients *
-3 #
-* Created by William J. De Heo #
-t on 1/7/98 *
-# Last modified 1/10/98 #
-
-*¥#######**3It¥*#¥¥¥#**##**¥##¥#####**####*#*###*##**###¥##*##*/
-
-#inc1ude <stdlib.h>
-#include <stdio.h>
-#include <math.h>
-#inc1ude "prototypes.h"
-
-/# Machine constants on dino (Sun Ultra 1 at Courant) t/
-#define HACHEPS 1.15828708535533604981e-16
-
-#define SQRTEPS 1.07623746699106147048e-08
-
-ﬂdefine HAX_NAH 100
-
-void read_name(char *);
-
-/* Prototypes from the file functions.c ¥/
-
-double a1pha(int j);
-
-double betasq(int j);
-
-double form1(int j, int k);
-
-double form2(int j, int k);
-
-double moment(doub1e ‘data, long n, double tave, double tvar, double tcov, long k);
-
-/# external variables to be used by functions */
-double *phi,tcov, tvar, tave;
-
-% 48.txt
-main()
-
-{
-
-char tfilename;
-
-FILE *ofp;
-
-double temp=O, *T;
-
-long i, j, nrou, nlanc, START;
-int f1ag=O;
-
-filename = cma11oc(HAx_NAHE);
-
-printf("\nName of file containing observed values: “);
-
-read_name(fi1ename);
-
-printf(“\nTota1 number of observations in file (iterations):
-
-scanf(“Zu”,&nrov);
-
-printf(“\nNumber of leading observations to discard: “);
-
-scanf("Zu",&START);
-
-printf("\nNumber of Lanczos coefficients desired: “);
-
-scanf("%u",&n1anc);
-
-vhi1e(2*n1anc > nrow—START)
-
-{
-printf("\nNot enough data for that many coefficients.\n“);
-printf("\nEnter a smaller number of Lanczos coefficients: “);
-scanf("Xu“,&n1anc);
-
-}
-
-II);
-
-phi = dma11oc(nrow);
-cov=dma11oc(2#n1anc);
-var=dma11oc(1);
-ave=dma11oc(1);
-
-T = dma11oc(n1anc#n1anc); for(i=0;i<n1anc;i++) T[i]=(doub1e)0;
-
-/* observable phi is stored contiguously column-wise (e.g. by HATLAB) t/
-mat1abread(phi, nrow, 1,fi1ename);
-
-/* send the observable, offset by START */
-moment(phi+START,nrou-START,ave, var, cov,2*n1anc-1);
-
-/* First column of T */
-
-T[0] = a1pha(1);
-if((temp=betasq(1))>HACHEPS*10)
-{
-
-T[1]=sqrt(temp);
-
-/* General column of T */
-
-J'=2:
-
-for(i=1;i<n1anc-1, f1ag==O;)
-{
-
-T[itn1anc+i-1]=sqrt(betasq(i));
-
-48
-
-main()
-
-{
-
-char tfilename;
-
-FILE tofp;
-
-double temp=0, #T;
-
-long i, j, nrow, nlanc, START;
-int f1ag=0;
-
-filename = cma11oc(HAx_NAH);
-
-princf("\nName of file containing observed values: “);
-read_name(fi1ename);
-
-printf("\nTota1 number of observations in file (iterations): “);
-scanf("2u“,&nrow);
-
-printf("\nNumber of leading observations to discard: “);
-scanf("Zu",&START);
-
-printf("\nNumber of Lanczos coefficients desired: “);
-scanf("%u“,&n1anc);
-
-whi1e(2*n1anc > nrou-START)
-
-{
-printf("\nNot enough data for that many coefficients.\n");
-
-printf("\nEnter a smaller number of Lanczos coefficients: “);
-scanf("Zu",&n1anc);
-}
-
-phi = dma11oc(urow);
-cov=dma1loc(2*n1anc);
-var=dma11oc(1);
-ave=dma11oc(1);
-
-T = dma11oc(n1anc*n1anc); for(i=0;i<n1anc;i++) T[i]=(doub1e)O;
-
-/* observable phi is stored contiguously column-wise (e.g. by HATLAB) */
-mat1abread(phi, nrov, 1,fi1ename);
-
-/* send the observable, offset by START #/
-moment(phi+START,nrov-START,ave, var, cov,2*n1anc-1);
-
-/# First column of T */
-
-T[0] = a1pha(1);
-if((temp=betasq(1))>HACHEPS#10)
-{
-
-T[1]=sqrt(temp);
-
-/* General column of T */
-
-J‘=2:
-
-for(i-1;i<n1anc-1, f1ag==0;)
-{
-
-T[i*nlanc+i-1]=sqrt(betasq(i));
-
-48
-
-% 49.txt
-T[i*n1anc+i] = a1pha(i+1);
-if((temp = betasq(i+1))>HACmEPS#10)
-{
-
-T[i#n1anc+i+1] = sqrt(temp);
-
-i++;
-}
-else
-
-f1ag=1;
-
-}
-
-/t Last column */
-if(f1ag!=1 it ((temp= betasq(n1anc-1))>HACHEPSt10))
-
-{
-T[n1anc#n1anc-2] = sqrt(temp);
-T[n1anctn1anc-1] = a1pha(n1anc);
-printf("\nbeta(1) = %1f\nbeta(Zd) = Zlf (last beta)",
-T[1],(nlanc-1),T[nlanc*n1anc-2]);
-}
-else
-{
-printf("\nApproximate invariant space reached at step %d.",i);
-printf("\nbeta(1) = Z1f\nbeta(Zd) = Zlf (last accurate beta)“.
-T[1],i,T[itn1anc+i—1]);
-printf("\nbeta(Zd)“2: Zlf (first spurious resu1t)“,i+1,temp);
-}
-
-printf("\nThe matrix T is: \n");
-matprint(T,n1anc,n1anc);
-
-}
-
-else
-
-{
-
-printf("\nmain(): Approximate invariant space reached at first step.“);
-
-49
-
-printf("\na1pha(1) - Z1f\nbeta(1)”2: Zlf (first spurious result)“, T[O], temp);
-
-}
-ofp = fopen("Tmat.m","v“);
-check(ofp);
-mat1aburite(T,n1anc,n1anc,ofp);
-fc1ose(ofp);
-}
-void read_name(char tname)
-{
-int c, i = 0;
-while ((c = getchar()) != EOF it c != ’ ’ it c !- ’\n’)
-name[i++] = c;
-name[i] = ’\0’;
-}
-
-/#t* end 1anczos.c *#*/
-
-% 50.txt
-50
-
-/llnkttttttttinlntint*1!!!##1##!ttitilltit###t¥#tt¥ttItt####t#ttII=##mlltttt
-
-* functions.c functions required by 1anczos.c t
-# t
-* Created by Hilliam J. De Meo #
-* on 1/7/98 *
-t Last modified 1/10/98 :
-
-¥¥##¥$¥##¥¥#$¥¥¥###¥¥$*I#****#**####*######¥#¥#¥#$#¥¥#l¥**#****/
-
-#inc1ude <math.h>
-
-#define START 1000
-
-#define ITER 10000
-
-/t Machine constants on dino (Sun Ultra 1 at Courant) */
-
-#define HACHEPS 1.15828708535533604981e-16
-
-#define SQRTEPS 1.07623746699106147048e-O8
-
-double a1pha(int j);
-
-double betasq(int j);
-
-double form1(int j, int k);
-
-double form2(int j, int k);
-
-double moment(doub1e tdata, long n, double tave, double tvar, double tcov, int k);
-
-/t external variables to be used by functions */
-double *phi,tcov, *var, tave;
-
-double alpha(int j)
-{
-
-/* alpha is never called with j < 1 */
-if(j==1)
-return
-form1(1,1);
-else if(j>1)
-return
-((doub1e)1/betasq(j-1))*
-(form1(j-1,3) — pou(a1pha(j-1),3)
-- 2 * (alpha(j-1)#betasq(j-1) + sqrt(betasq(j-2))#form2(j-1,2))
-+ betasq(j-2));
-}
-
-double betasq(int j)
-
-{
-if(j==0)
-return (doub1e)O;
-else if(j>0)
-return (form1(j,2) - pow(a1pha(j),2) - betasq(j-1));
-}
-
-double form1(int j, int k)
-{
-
-% 51.txt
-51
-
-double form14,alphal,alpha1sq,form12,form13;
-if(j==0)
-
-return (doub1e)0;
-else if(j==1)
-{
-
-/# printf("\nvar = %1f, cov(%d) = %1f \n“,*var,k,cov[k]);t/
-
-return (cov[k])/(*var); /* the only real value */
-}
-else if(j>1)
-{
-
-return ((double)1/betasq(j-1))
-
-* ( form1(j-1,k+2)
-+ pow(a1pha(j-1),2)*form1(j-1,k)
-+ 2t(a1pha(j-1) t sqrt(betasq(j-2)) * form2(j-1,k)
-- a1pha(j-1)*form1(j-1,k+1)
-- sqrt(betasq(j—2))#form2(j-1,k+1) )
-+ betasq(j-2)*form1(j-2,k)):
-}
-}
-
-double form2(int j, int k)
-{
-/* form2 is never called with j < 1 */
-if(j==1)
-return (doub1e)0;
-else if(j>1)
-return
-(pov(betasq(j-1),—.5)) 3
-(form1(j-1,k+1) - a1pha(j-1)*form1(j-1,k)
-- sqrt(betasq(j-2)) # form2(j—1,k));
-}
-
-/* moment() function for computing var and cov(k)
-arguments:
-data = a nxi array of doubles
-n = length of data[]
-ave =(on exit)= the average of data[]
-var =(on exit)= the variance of data[]
-cov =(on exit)= the covariance of data[i] and data[i+j] for i-1,...,k
-k = the max lag for cov above
-*/
-double moment(double tdata, long n, double tave, double tvar, double tcov, int k)
-{
-/* Centered about data[0] algorithm: #/
-
-long i,j;
-double avel, ave2;
-
-% 52.txt
-52
-
-*ave=O;tvar=0; ave1=ave2=O;
-for(j=0;j<=k;j++)
-cov[j]=0;
-
-for(i=1;i<n;i++)
-{
-rave += (data[i] - data[0]);
-tvar += (data[i] - data[0])#(data[i] - data[0]);
-}
-tvar /= (doub1e)(n-1);
-Ivar -= (((*ave)/(doub1e)n) * ‘(*ave)/(doub1e)(n-1)));
-/# *ave = ((*ave)/(doub1e)n) + data[0]; <- that’s the true average (not needed)*/
-
-for(j=0;j<=k;j++)
-{
-
-for(i=0;i<(n-j);i++)
-
-{
-
-avel += (data[i] - data[0]);
-ave2 += (data[i+j] - data[0]);
-cov[j] += (data[i] - data[0])*(data[i+j] — data[0]);
-
-}
-
-avel/=(doub1e)(n-j); ave2/=(doub1e)(n-j);
-
-cov[j] = ((cov[j] - (double)(n-j)tave1¢ave2)/(double)(n-j-1));
-}
-
-}
-
-/trt end funccions.c ###/

### 2013-10-18 tex/acknowledge.tex
initial commit of tex source for nyu thesis style setup

First, I thank my advisor, Professor Jonathan Goodman, for giving me the opportunity to work on
this problem, and helping me arrive at the following exposition. Next, I wish to thank Professor
Helena Frydman for ﬁrst sparking my interest in Markov chains by giving lucid descriptions of their
many interesting properties and applications. I thank Professors James Demmel and Beresford
Parlett, for answering many questions pertaining to this problem, and Professor Leslie Greengard
for agreeing to review this paper.

Finally, I would like to thank my family, for their support of my interests in research (and
everything else!), and especially my parents, Barbara and Ted Terry, and Benita and Bill De Meo.

Without them, this paper would not have been written.

### 2013-10-18 tex/intro.tex
initial commit of tex source for nyu thesis style setup

--- 2013-10-18 tex/MSThesis.tex
@@
-%\documentclass[11pt]{amsart}
-\documentclass{au}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% For completion by editorial office:
- \presentedby{\dots}
- \received{\dots}{\dots}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\usepackage[colorlinks=true,urlcolor=black,linkcolor=black,citecolor=black]{hyperref}
-
-\usepackage{amsmath,amssymb,url,mathrsfs}
-\usepackage{enumerate}
-\usepackage{tikz}
-\usetikzlibrary{calc}
-\usepackage{scalefnt}
-\numberwithin{equation}{section}
-
-\theoremstyle{plain}
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{proposition}[theorem]{Proposition}
-\newtheorem{corollary}[theorem]{Corollary}
-
-\theoremstyle{definition}
-\newtheorem{definition}[theorem]{Definition}
-\newtheorem{notation}[theorem]{Notation}
-\newtheorem{remark}[theorem]{Remark}
-\newtheorem{example}[theorem]{Example}
-\newtheorem*{remarks}{Remarks}
-
-\newcommand{\todo}[1]{~\\[5pt]{\bf TODO(wjd):} #1\\[5pt]}
-%\newcommand{\suchthat}{\ensuremath{:}}    % (old notation)
-\newcommand{\suchthat}{\ensuremath{\mid}}  % (new notation)
-%\newcommand{\defeq}{\ensuremath{:=}}      % (old notation)
-\newcommand{\defeq}{\ensuremath{=}}        % (new notation)
-\newcommand{\betai}{\ensuremath{\beta^i}}  % \beta^{\bB_i}   % (old notation)
-\newcommand{\betaj}{\ensuremath{\beta^j}}  % \beta^{\bB_j}
-\newcommand{\betak}{\ensuremath{\beta^k}}  % \beta^{\bB_k} or \beta^{\bB_{k}}
-\newcommand{\betal}{\ensuremath{\beta^\ell}}  %\beta^{\bB_\ell} or \beta^{\bB_{\ell}}
-\newcommand{\uacalc}{{\small UAC}alc}
-\newcommand{\Tuma}{T\r{u}ma}
-\newcommand{\Palfy}{P\'alfy}
-\newcommand{\Pudlak}{Pudl\'ak}
-\newcommand{\PP}{P\'alfy-Pudl\'ak}
-\newcommand{\PAP}{P\'alfy\ and Pudl\'ak}
-\newcommand{\<}{\ensuremath{\langle}}
-\renewcommand{\>}{\ensuremath{\rangle}}
-\newcommand{\beps}{\ensuremath{\boldsymbol{\varepsilon}}}
-\newcommand{\bN}{\ensuremath{\mathbf{N}}}
-\newcommand{\bA}{\ensuremath{\mathbf{A}}}
-\newcommand{\bB}{\ensuremath{\mathbf{B}}}
-\newcommand{\bBi}{\ensuremath{\mathbf{B}_i}}
-\newcommand{\sB}{\ensuremath{\mathcal{B}}}
-\newcommand{\sC}{\ensuremath{\mathcal{C}}}
-\newcommand{\sS}{\ensuremath{\mathscr{S}}}
-\newcommand{\sT}{\ensuremath{\mathscr{T}}}
-\newcommand{\sI}{\ensuremath{\mathcal{I}}}
-\newcommand{\sE}{\ensuremath{\mathcal{E}}}
-\newcommand{\sO}{\ensuremath{\mathcal{O}}}
-%\newcommand{\id}{\ensuremath{\mathrm{id}}}
-\DeclareMathOperator{\id}{id}
-%\newcommand{\Eq}{\ensuremath{\mathrm{Eq}}}
-\DeclareMathOperator{\Eq}{Eq}
-%\newcommand{\Cg}{\ensuremath{\mathrm{Cg}}}
-\DeclareMathOperator{\Cg}{Cg}
-%\newcommand{\Con}{\ensuremath{\mathrm{Con\,}}}
-\DeclareMathOperator{\Con}{Con}
-%\newcommand{\Sub}{\ensuremath{\mathrm{Sub}}}
-\DeclareMathOperator{\Sub}{Sub}
-%\newcommand{\Pol}{\ensuremath{\mathrm{Pol}}}
-\DeclareMathOperator{\Pol}{Pol}
-%\newcommand{\Clo}{\ensuremath{\mathrm{Clo}}}
-\DeclareMathOperator{\Clo}{Clo}
-\newcommand{\ps}[1]{\ensuremath{^{(#1)}}}
-\newcommand{\piB}{\ensuremath{\pi_B}}
-\newcommand{\hpsi}{\ensuremath{\widehat{\psi}}}
-\newcommand{\htheta}{\ensuremath{\widehat{\theta}}}
-\newcommand{\supi}{\ensuremath{^{i}}}
-\newcommand{\supj}{\ensuremath{^{j}}}
-\renewcommand{\leq}{\ensuremath{\leqslant}}
-\renewcommand{\nleq}{\ensuremath{\nleqslant}}
-\renewcommand{\geq}{\ensuremath{\geqslant}}
-\newcommand{\meet}{\ensuremath{\wedge}}
-\newcommand{\join}{\ensuremath{\vee}}
-\newcommand{\Meet}{\ensuremath{\bigwedge}}
-\renewcommand{\Join}{\ensuremath{\bigvee}}
-\newcommand{\resB}{\ensuremath{|_{_B}}}
-\newcommand{\resBi}{\ensuremath{|_{_{B_i}}}}
-\newcommand{\eps}{\ensuremath{\varepsilon}}
-\newcommand{\hatmap}{\ensuremath{\widehat{\phantom{x}}}}
-\newcommand{\one}{\ensuremath{\mathbf{1}}}
-\newcommand{\two}{\ensuremath{\mathbf{2}}}
-\newcommand{\three}{\ensuremath{\mathbf{3}}}
-\newcommand{\tbeta}{\ensuremath{\widetilde{\beta}}}
-\newcommand{\ttheta}{\ensuremath{\widetilde{\theta}}}
-\newcommand{\hbeta}{\ensuremath{\widehat{\beta}}}
-\newcommand{\cick}{\ensuremath{\sC_i}}
-%\newcommand{\CE}{\ensuremath{\sC_r^{\sE}}}
-\newcommand{\CE}{\ensuremath{\sB_r^n}}
-\newcommand{\CO}{\ensuremath{\sC_r^{\sO}}}
-\newcommand{\CICK}{\ensuremath{c_i/\beta \cup c^{K+1}_i/\beta^{\bB_{K+1}}}}
-\newcommand{\GAP}{{\small GAP}}
-\newcommand{\dotsize}{.8pt}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{document}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%  FRONT MATTER                                                 %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\title[Approximating Eigenvalues of Large Stochastic Matrices]{A Lanczos Procedure for Approximating Eigenvalues of Large Stochastic Matrices}
-
-%% First author (Note: The order of the items here is important!)
-\author{William DeMeo}
-\email{williamdemeo@gmail.com}
-\urladdr{http://www.math.sc.edu/~demeow}
-\address{Department of Mathematics\\
-University of South Carolina\\Columbia 29208\\USA}
-
-\thanks{The author wishes to thank a number of people who contributed to this
-  work.  In particular, the primary referee made many insightful recommendations
-  which led vast improvements, not only in the overall exposition, but also in
-  the specific algebraic constructions, theorem statements, and proofs.
-  This article is dedicated to Ralph Freese and Bill Lampe, who declined to be
-  named as co-authors despite their substantial contributions to the project.}
-
-%% Dedication (Optional)
-\dedicatory{This article is dedicated to Ralph Freese and Bill
-  Lampe.}
-
-%% AMS subject classification; see http://www.ams.org/msc
-%% Only one Primary. Possibly several Secondary.
-\subjclass[2010]{Primary: 08A30; Secondary: 08A60, 06B10.} %, 06B15.}
-
-%% Keywords and phrases
-\keywords{congruence lattice, finite algebra, finite lattice
-  representations}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%  MAIN MATTER                                                  %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{abstract}
-The rate at which a Markov chain converges to a given probability distribution
-has long been an active area of research. Well known bounds on this rate of
-convergence involve the subdominant eigenvalue of the chain’s underlying
-transition probability matrix. However, many transition probability matrices are
-so large that we are unable to store even a vector of the matrix in fast
-computer memory. Thus, traditional methods for approximating eigenvalues are
-rendered useless.
-
-In this paper we demonstrate that, if the Markov chain is reversible, and we
-understand the structure of the chain, we can derive the coefficients of the
-traditional Lanczos algorithm without storing a single vector. We do this by
-considering the variational properties of observables on the chain’s state
-space. In the process we present the classical theory which relates the
-information contained in the Lanczos coefficients to the eigenvalues of the
-Markov chain. 
-\end{abstract}
-
-\maketitle
-
-%%%%%%%%%%%%%%%%%%%%%
-\chapter{Introduction}
 The rate at which a Markov chain converges to a given probability distribution has long
@@
 % 05.txt
-
-\chapter{Markov Chains}
-
-\section{General Theory}
-
-This review of Markov chain theory can be found in any good probability text. The present
-discussion is most similar to that of Durrett [3], to which we refer the reader desiring greater detail.
-
-\subsection{The Basic Setup}
-
-Heuristically, a Markov chain is a stochastic process with a lack of memory property. Here
-this means that the future of the process, given its past behavior and its present state, depends only
-on its present state. This is the probabilistic analogue of a familiar property of classical particle
-systems. Given the position and velocities of all particles at time t, the equations of motion can be
-completely solved for the future evolution of the system. Thus, information describing the behavior
-of the process prior to time t is superﬂuous. To be a bit more precise, if technical, we need the
-following definitions.
-
-\bibliographystyle{spmpsci}
-\bibliography{wjd}
-
-\end{document}
-
-%%%%%%%%%%%%%%%%%%%%%  END DOCUMENT %%%%%%%%%%%%%%%%%%%%%%%%%
-
-Definition 2.1.1 Let (S, 8) be a measurable space. A sequence X,., n 2 0, of random variables
-
-taking values in S is said to be a Markov chain with respect to the filtration a(Xo, . . . ,X,,) if for
-% 06.txt
-a.llBES,
-
-p(x,,+1 e B | a(Xo, . . . ,X,.)) = P(X,.+1 e B 1 o(X,,)) (2.1)
-
-Equation (2.1) merely states that, given the present location of X,,, the past is irrelevant for
-
-predicting the location of X,,+1.
-
-Definition 2.1.2 A function p : S x S —> R is said to be a transition probability if:
-1. for each :1: E S, A —> p(a:, A) is a probability measure on (S, S).
-2. for each A E 8, :1: —> p(z, A) is a measurable function.
-
-We call X,, a Markov chain with transition probabilities 12,, if
-P(Xn+l E B l a(Xn)) =Pn(XnaB) (22)
-
-The spaces (S, S) that we encounter below are standard Borel spaces, so the existence of the
-transition probabilities follows from the existence of regular conditional probabilities on Borel spaces
-— a standard measure theory result (see e.g. [3] page 230).
-
-Suppose we are given an initial probability distribution p on (S, S) and a sequence p,, of
-
-transition probabilities. We can define a consistent set of finite dimensional distributions by
-
-P<X,- e B,-,0 s 1 s n) = / #(d1‘o) / Po($o.d-'51) . .. / Pn-l(97n—1ad$n) (2.3)
-B0 B1 B1:
-Furthermore, denote our probability space by
-(9,?) = (S{0,1,...},8{0,l,...})
-
-We call this sequence space and it is defined more explicitly by
-
-sf°»1=-~-} = {(wo,w1,...) :w.- e S}
-
-S{°*1""} = ¢7(w :40; E A,- E 3)
-
-% 07.txt
-The Markov chain that we will study on this space is simply X" (w) = w,., the coordinate maps.
-Then, by the Kolmogorov extension theorem, there exists a unique probability measure P,, on
-(9,?) so that the X" (w) have finite dimensional distributions (2.3).
-
-If instead of p, we begin with the initial distribution 6,, i.e. point mass at 2:, then we
-denote the probability measure by P3. With such measures defined for each 2:, we can in turn
-
-define distributions P,,, given any initial distribution p, by
-
-PM = / u<da.->P.(A>
-
-That the foregoing construction — which, recall, was derived merely from an initial distri-
-bution p and a sequence p,, of transition probabilities — satisfies Definition (2.2) of a Markov chain
-is not obvious, and a proof can be found in
-
-To state the converse of the foregoing, if X,, is a Markov chain with transition probabilities
-p,, and initial distribution p, then its finite dimensional distributions are given by (2.3). Proof of
-this is also found in
-
-Now that we have put the theory on a firm, if abstract, foundation, we can bring the
-discussion down to earth by making the forgoing a little more concrete. First, we specialize our
-study of Markov chains by assuming that our chain is temporally homogeneous, which means that
-the transition probabilities do not depend on time; i.e. p,,(w,,, B) E p(w,., B). (This is the stochastic
-analogue of a conservative system.) Next we assume that our state space S is finite, and suppose
-for all states 1', j E S that p(i, j) Z 0, and E, p(z', j) = 1 for all i. In this case, equation (2.2) takes
-
-a more intuitive form:
-P(Xn+l =j I X7: = 7:) =P(iaj)
-
-and our transition probabilities become
-% 08.txt
-If P is a matrix whose (i, j)th element is the transition probability p(i, 1') then P is a stochastic
-
-matriar, that is, a matrix with elements pgj satisfying
-
-Pij 20: zjpij =1: (i7j=1121"‘)d)
-
-We also refer to P as the transition probability matrix.
-
-Without loss of generality, we can further suppose our Markov chain is irreducible. This
-means that, for any states 2', 1‘, starting in state i the chain will make a transition to state 1' at
-some future time with positive probability. This state of affairs is often described by saying that
-all states communicate. We lose no generality with this assumption because any reducible Markov
-chain can be factored into irreducible classes of states which can each be studied separately.
-
-The final two conditions we place on the Markov chains considered below will cost us
-some generality. Nonetheless, there remain many examples of chains meeting these conditions,
-thus making the present study worthwhile. Furthermore, it may be the case that, with a little
-more work, we will be able to drop these conditions in future studies. The first condition is that
-the chain is aperiodic. If we let I, = {n 2 1 : p"(:c,a:) > 0}, we call a Markov chain aperiodic if,
-for any state 2:, the greatest common divisor of I, is 1. The second assumption is that our chain is
-
-reversible. This characterization is understood in terms of the following definition.
-Definition 2.1.3 A measure a is called reversible if it satisfies
-
-;4(w)p(I. y) = u(y)p(y. w)
-
-for all :c and y.
-
-We call a Markov chain reversible if its stationary distribution (defined in Section 2.1.2) is reversible.
-% 09.txt
-\subsection{A Convergence Theorem}
-
-In succeeding arguments, we use some results concerning the asymptotic behavior of
-
-Markov diains. These results require a few more definitions.
-
-Definition 2.1.4 A measure 1r is said to be a stationary measure if
-
-Z«(z)p<z.y> = we) (2.4)
-
-2
-Equation 2.4 says P,,(X1 = y) = 1r(y), and by induction that P,,(X,, = y) = 7r(y) for all n 2 1. If 1r
-is a probability measure, then we call 1r a stationary distribution. It represents an equilibrium for
-the chain in the sense that, if X0 has distribution 1r, then so does X,. for all 11..
-
-When the Markov chain is irreducible and aperiodic, the distribution of the state at time
-n converges pointwise to 1r as n —> oo, regardless of the initial state. It is convenient to state
-this convergence result in terms of the Markov chain's transition probability matrix P. Before
-doing so, we note that irreducibility of a Markov chain is equivalent to irreducibility (in the usual
-matrix theory sense) of its transition probability matrix. Furthermore, it turns out that a transition
-probability matrix of an aperiodic Markov chain falls into that class of matrices often called acyclic,
-but for simplicity we will call such stochastic matrices aperiodic. With this terminology, we can
-
-state the convergence theorem in terms of the transition probability matrix P.
-
-% Theorem 2.1.1
-\begin{theorem}
-\label{thm-2.1.1}
-Suppose P is irreducible, aperiodic, and has stationary distribution 7r. Then as n —> oo,
-p"(i,J') -> #0")
-\end{theorem}
-
-The symbol p"(i, j) is to be understood as the (ij) element of the nth power of P.
-A Markov chain whose transition probability matrix satisfies the hypotheses of
-Theorem~\ref{thm-2.1.1} is called ergodic. If we simulate an ergodic chain for
-sufficiently many steps, having 
-% 10.txt
-begun in any initial state, the final state is a sample point from a distribution which is close to 1r.
-
-To make this statement more precise requires that we define “close”.
-
-Definition 2.1.5 Let 1r be a probability measure on S. Then the total variation distance at time
-
-n with initial state I is given by
-AA") = l|P"(z, A) - 7r(-4)||Tv = 1333; I P"(I.A) - 7r(A) I
-
-In what follows, we will measure rate of convergence using the function 'r,(e), the first time after
-
-which the total variation distance is always less than 5. That is,
--r,,(e) = min{m : A,(n) 3 E for all n. 2 m}
-
-To consider the connection between convergence rates and eigenvalues, note that an ape-
-riodic stochastic matrix P has real eigenvalues 1 = /\o > A1 2 A2 3  Z /\¢_1 2 -1, where
-d = |S'] is the dimension of the state space. For an ergodic chain, the rate of convergence to the
-stationary distribution 7r is bounded by a function of the subdominant eigenvalue. By subdominant
-eigenvalue we mean that eigenvalue which is second-largest in absolute value, and we denote this
-Am“ E max{).1,|A¢_1|}. The function bounding the rate of convergence is given by the following
-
-theorem:
-
-% Theorem 2.1.2
-\begin{theorem}
-\label{thm-2.1.2}
-The quantity T1(€) satisfies
-1. 13(5) 5 (1 — /\m,_x)“(ln1r(a:)‘1 +1ne‘1)
-2. maxxes r,(e) 2 %/\,,,,x(1 — ).,,,,x)‘1ln(2e)’1
-\end{theorem}
-
-As this theorem shows, if we have an upper bound on the subdominant eigenvalue, then we have
-an upper bound on the function 11(5). In what follows, we will derive an approximation to the
-% 11.txt
-subdominant eigenvalue and supply error bounds. Together, an approximation and error bounds
-for Am“ provide enough information to make Theorem~\ref{thm-2.1.2} useful.
-
-\section{Functions on the State Space}
-
-Recall that X,.(w) = w,. 6 S denotes the state in which the Markov chain exists at time
-11. Suppose that 11> = {¢1, . . . , ¢,,} is a collection of p observables, or functions defined on the state
-space S. Furthermore, let these observables be real valued, so ¢.~ : S —> R. It is often useful, as we
-will see below, to assume that none of the observables is a constant function. Suppose now that
-the state space S is finite with at possible states. Then, since an observable is simply a map of the
-state space, we can think of each ¢,- as a vector of d real numbers — the d values that it takes on at
-the different states.
-
-Now assume the Markov chain is irreducible, and let 1r denote its stationary distribution.
-If we start the chain from its stationary distribution - i.e. suppose X0 has distribution 1r - then
-X,. is a stationary process. Furthermore, for each i, ¢,(X,.) is a stationary stochastic process with
-
-mean
-
-E.¢.- = Z) vr(x)¢.(z)
-
-165
-
-and autocovariance function
-
-Cir (¢i(Xn)v ¢i (Xn+a))
-
-E7!’  ‘ E1r¢i) (¢i(Xn+a) " E1r¢i)l
-
-Z Pn(Xn = z1X7I+S = y)  — E1r¢i)  _ E1r¢i)
-
-z,yES
-(2.5)
-
-which, by the definition of conditional probability, we can write as,
-
-2 P«(X.. = z)P.(X..+.. = y I X. = 1)(¢s($) — Em) (my) ~ Em)
-1.1165
-% 12.txt
-Z 7r($)P1y (MI) - Em.-) (¢:(1I) - E1r¢i)
-
-:c,y€S
-
-Here pgy denotes the element in row :1: and column y of P’, the sth power of the transition probability
-
-matrix. Similarly, we define the cross-covariance between the function ¢,- at time n and q5,- at time
-
-n+sas
-
-E1!’  _ E7r¢:') (¢i(Xn+a) " E1r¢i)l (2-6)
-
-C1r(¢i(Xn)a (#1 (Xn+a))
-
-Z) vr(x)p;,, (¢.-(w) — Em.-) (my) — Em)
-
-z,yES
-
-Now let (<I>) denote the matrix of mean vectors whose jth column is E,,¢,-1, where 1 = (1, . . . , 1)‘,
-and let H = diag(1r(w1), . . . , 1r(wd)), be the d X d diagonal matrix with stationary probabilities 1r(w)
-on the main diagonal, and zeros elsewhere. Finally, denoting by C(s) the p X p covariance matrix
-
-whose (i,j)th element is C, (qfi,-(X,.),¢_,< (X,.+,)), we have
-
-0(0) = E(‘1’(Xn) - (‘I>))(‘I’(Xn) - (‘PW
-= (‘P-(‘1’))‘1'1(<1’-(‘I>))
-0(8) = E(<I>(Xn) — (‘I>))(‘1’(Xn+a) - (‘1’))’
-
-(‘P - (‘1’))'1'IP‘(<1> — (<1>))
-
-Below, we will also find it useful to have at our disposal a new matrix that is similar to the
-transition probability matrix. We have in mind the matrix M E Ill/2PI'I‘1/2. As is easily verified,
-this allows us to write the covariance matrix as
-
-can = <<I>—<<I>>)°n%*M*n%<<I>—<<I>>>
-
-\II‘M’\II (2.7)
-
-where we have defined ‘II E 111/ 2(<I> —  Recall that our main motivation is that, for out-of-core
-problems, traditional eigenvalue algorithms are inadequate. With this fact, and the form (2.7) in
-% 13.txt
-mind, we will consider using the covariance of observables on the state space to implement the
-(otherwise useless) Rayleigh-Ritz procedure, which we describe below. This procedure requires
-that M be symmetric. As the next fact demonstrates, this required symmetry is the reason for our
-
-interest in the reversibility property of the Markov chain.
-
-Fact 2.2.1 The matrix M is symmetric if and only if the Markov chain is reversible (i.e. iﬂ the
-
-process satisfies the detailed balance condition).
-
-Proof:
-
-M is symmetric
-
-4:, (1-I1/2P1-I-1/2): = H1/2Pn—1/2
-<.—> r1-%*P°n%* = H1/2911-1/2
-¢> 1>'r1= = up
-~:=> (IIP)‘ = l'IP
-
-Element-wise, the final equality implies that 7|‘,'p,'j = 1rjpj.-. According to Definition 2.1.3, this states
-
-that p is a reversible measure.
-
-% 14.txt
-
-\chapter{Invariant and Approximate Invariant Subspaces}
-
-\section{Invariant Subspaces}
-Definition 3.1.1 A subspace S Q R" with the property that
-
-1:ES'=Ma:€S
-
-is said to be invariant for M.
-
-Recall that, having chosen observables <I> = ((151, . . . , qfip), we constructed the covariance
-matrix C(s) = \II‘M"\II. If the column space of \II, which we denote by ran(\II), is an invariant
-subspace for M, the definition implies Mifij E  That is, for each 111,- there exists a vector
-t of coeﬂicients such that M1/Jj = f=1t,-1/)5,-. This is true for all j and, putting each vector of
-coefficients into a matrix T, we see that M\II = \IIT. Conversely, M\II = IIIT implies that M1/Jj is a
-
-linear combination of columns of \II, so ran(\II) is invariant. We have thus proved the following
-% 15.txt
-Fact 3.1.1 The subspace ran(\II) is invariant for M if and only if there exists T E R”? such that
-
-M\II = \IIT.
-
-Consequently,
-
-Fact 3.1.2 )\(T) Q ).(M)
-
-Proof of 3.1.2:
-
-z\E)\(T) _=_ 3v€R"3Tv=).v
-<=> \IITv = A\IIv
-<=> M\II-u = /\\IIv
-< /\ E /\(M)
-
-The second ¢> follows from fact 3.1.1.
-Facts 3.1.1 and 3.1.2 are theoretically useful. To see why, write the equation of fact 3.1.1
-
-as follows:
-
-\IIT = M\II
-\II‘\IIT = \II‘M\II
-T = (\1:'~1r)-‘\1:'M\1:
-
-In this form, we recognize that T = C‘1(0)C(1). That is, using only the covariance of observables
-on the state space, we can generate a matrix T with the property A(T) (_Z ).(M). Recalling that
-M = H1/2Pl'I‘1/2, we see that M is similar to our original stochastic matrix P, and thus ).(M) =
-
-).(P).
-
-\section{Approximate Invariant Subspace}
-
-We qualiﬁed the foregoing by stating that the facts are only theoretically useful. This
-
-is because when choosing observables of the state space we aren't sure that they will satisfy the
-% 16.txt
-primary assumption underlying facts 3.1.1 and 3.1.2. Recall the assumption: ran(\II) is an invariant
-subspace for M, where ‘II = H1/2(<I> —  It may well be the case that there is a 1p; 6 ran(\II) such
-that M1/zj ¢ ran(\II), thereby violating the assumption. However, lacking an invariant subspace, we
-
-might hope that our observables at least provide an approximate invariant subspace.
-
-Deﬁnition 3.2.1 If the columns of ‘II E R“? are independent and the residual matrix E E
-
-M\II —— \IIT is small for some T 6 R7”, then ran(\II) deﬁnes an approximate invariant subspace.
-
-To see how an approximate invariant subspace can be useful in approximating eigenvalues of M,
-we extract a theorem from Golub and Van Loan 
-% Theorem 3.2.1
-\begin{theorem}
-\label{thm-3.2.1}  
-Suppose M 6 Rd” and T 6 RP"? are symmetric and let E E MQ — QT, where Q 6 Rd"? satisﬁes
-
-Q‘Q = I. Then there exist p1,... ,;z,, E A(T) and A1,. . . , A,, E A(M) such that
-ll‘k - Akl S \/§|lE"
-
-fork=1,...,p.
-\end{theorem}
-
-If the subspace ran(Q) is an approximate invariant subspace, then the deﬁnition implies that
-there is a choice T rendering the error IIEII2 small, and thus, the eigenvalues of T provide a good
-approximation to those of M.
-
-When considering the preceding ideas, it becomes apparent that their application will
-present new — but hopefully less prohibitive — problems. As these problems summarize the rt of
-the paper, now is a good time to examine them. First, the preceding theorem assumes a matrix Q
-whose columns form an orthonormal basis for the approximate invariant subspace. For our problem,
-derivation of such a Q is tricky, and we must prepare for this. Next, having an approximate invariant
-
-subspace at our disposal merely tells us that there exists some matrix T which makes the error
-% 17.txt
-"EH2 small. We must discover the form of such a T. Moreover, it is natural to seek that T which
-minimizes IIEII2 for a given approximate invariant subspace. Finally, in order to apply these ideas
-to real examples of our eigenvalue problem, we must ﬁnd a practical way to generate the most
-appropriate approximate invariant subspace. We now address each of these issues in the order
-raised.
-
-Recall the matrix I1 E III/2(<I> — (<I>)). The columns of this matrix, though independent
-(by choice of independent observables), are not necessarily orthonormal. However, consider the
-polar decomposition ‘II = QZ where Q‘Q = I and Z2 = \Il‘\II is symmetric positive semideﬁnite.1
-
-Note that Z = (\II‘\II)l/2 is nonsingular, so Q has the form
-Q = ~1:z-1 = \II(\II‘\II)“/2
-
-and it is clear that ran(Q) = ran(\II). Perhaps ra.n(Q) is a. useful approximation to the invariant
-subspace for M. If ran(Q) is not itself invariant, we have the error E = MQ — QT. It is easy to
-show, as we do in the following proof, that the T which minimizes HE“; is T = Q‘MQ. This yields
-
-the following,
-Theorem 3.2.2
-
-If \I1 = QZ is the polar decomposition of \II, then the matrix
-T = Q‘MQ = (\1:*\1:)-1/’\1:‘M\I:(\Ir‘\I:)-1/2
-minimizes IIEII2 = IIMQ - QTII2
-
-Proof:
-We prove the result by establishing the following
-
-(Hal: Given M 6 Rd” suppose Q 6 Rd"? satisﬁes Q'Q =1. Then,
-
-min ||MQ — QTII2 = H(1- QQ')MQll2
-
-TERI‘?
-
-‘Recall that the polar decomposition is derived from the SVD, \II = UZJV‘, by letting Q = UV‘ and Z = VEV‘.
-
-% 18.txt
-and T = Q’MQ is the minimizer.
-To verify the claim, observe the following application of the Pythagorean theorem:
-
-For any T E R.”"” we have
-
-IIMQ — QT||§ = IIMQ - QQ'MQ + QQ'MQ - QT|l§
-||(I - QQ’)MQ + Q(Q'MQ - T) H3
-
-||(1 - QQ‘)MQ||§ + |lQ(Q'MQ - T)||§ (3-1)
-
-2 l|(I — QQ')MQ||§
-
-Equality (3.1) holds since I — QQ' projects MQ onto the subspace orthogonal to ran(Q). Thus,
-the two terms in the expression on the right are orthogonal, and the Pythagorean theorem yields
-equality. The concluding inequality establishes that the minimizing T is that which annihilates the
-second term in (3.1); i.e. T = Q‘MQ.
-
-The second equality in Theorem 3.2.2 is a consequence of the polar decomposition, in
-
-which Q = \IIZ“ = \I:(~I:‘\I:)-1/2. Therefore,
-T = (x1z'x1:)-1/2x1:‘M~1:(\1:*~1z)—1/2
-
-is the minimizer.
-
-3.3 The Krylov Subspace
-
-Suppose the columns of a matrix Q E R“? provide an orthonormal basis for an approxi-
-
-mate invariant subspace. Then we have seen,
-1. T = Q'MQ minimizes ||E||2 = ||MQ — QT||2 and
-2. E|p.1,...,p,, E z\(T) and /\1,...,A, 6 »\(M) such that
-
-ll-‘I: - Ml S \/§|lE||2
-
-% 19.txt
-19
-fork=1,...,p.
-
-Given an approximate invariant subspace ran(Q) of dimension 1), these facts tell us what matrix we
-should use to approximate p elements of M’s spectrum. Now all we lack is a. description of ran(Q).
-That is, we have not speciﬁed which approximate invariant subspace would best suit our objective
-
-of approximating the subdominant eigenvalue A,,.a2:(M). 'I‘o do so requires the following deﬁnition:
-
-Deﬁnition 3.3.1 The Raleigh quotient of a symmetric matrix M and a nonzero vector 3 is
-
-p(2:, M) E (:z:‘Ma:)/(z‘a:).
-
-We will also denote this by p(z) when the context makes clear what matrix is involved.
-
-To ﬁnd the approximate invariant subspace most appropriate for our problem, we choose
-each dimension successively, providing justiﬁcation at each step. Thus, start with one observable
-42 on the state space, and let zpl E 111/ 2(¢ — E,,¢). That is, 1/:1 represents a centered (mean zero)
-observable whose ith coordinate is weighted by . First notice that the deﬁnition (which comes
-directly from the deﬁnition of ‘II following equation (2.7)) is such that 1/)1 is a constant vector if
-and only if :15 is oonstant on the state space, in which case 1/); E 0. This fact provides one reason
-observables that are constant on the state space are not interesting. Second, recall that the row
-sums of the matrix P are all one and, therefore, the eigenvector corresponding to the eigenvalue
-Ao(P) = 1 is the constant vector. By deﬁnition of M E 1'11/2Pl'I‘1/2, we see that the constant vector
-is also the eigenvector corresponding to A0  This will play an important role in what follows, as
-it allows us to focus primarily on the subdominant eigenvalue z\max(M) E max{}q (M), [A44 
-rather than on A0  (which we already know is 1).
-
-Now, notice that
-
-lP(¢1,M)l 5 max |P(x,M)l = »\max(M) (3-2)
-% 20.txt
-where the max is taken over all non-constant vectors. Since our interest centers on An-.3_x(M), we
-would like a. 1/11 which makes the left hand side of (3.2) large. This would be achieved if 1/)1 were to
-lie in the space spanned by, say, the first two eigenvectors of the Markov chain (sometimes referred
-to as the slowest modes of the process). However, this subspace spans only two dimensions of the
-entire d-dimensional space, and it is more likely that ¢1, at best, only comes close to lying in the
-subspace of interest. Now, given 1/11, a judicious choice for the second dimension l/)2, and hence
-W; E [1/)1 1/:2], would be that which makes max,,¢o |p(\II2a)| is as large as possible. To establish that
-
-this is indeed the pertinent objective, note the following:
-
-_ a‘\1:5Mx1:2a
-I;1;g<Ip(\I'2a,M)l = 13:3: ————a,W§q,2a
-= M
-mrggf,mIp(z. )1
-5 maxlp(z,M)l (3-3)
-= ).,,m(M)
-
-Again, the max on the right side of (3.3) is over non-constant vectors. In words, we wish to chose
-\II-2 = [1/:1 ¢2] so that there is a vector a 6 R2 making |p(\II2a,M)| close to )«max(M).
-
-Now, p(1/11) changes most rapidly in the direction of the gradient Vp(¢1).
-
-W0  (iiiiiiv---viiiléi)
-= ,,;,l (Mam — p<¢.)«m (3.4)
-
-So, to maximize the left hand side of (3.3), ‘I12 should be chosen so that the subspace ra.n(\II2)
-
-contains the gradient vector. That is, we must choose 1/22 so that
-
-VP(¢'1) E F3l1{¢1y¢2} E F-'%!1(‘I’2) (3-5)
-
-Clearly, equation (3.4) implies Vp(1/:1) E ran{z/11,M1,b1}. Therefore, if ran{1/:1,1/)2} = ra.n{¢1,M¢1},
-
-then (3.5) is satisﬁed.
-% 21.txt
-In general, having chosen \II,, E [1/)1 - -- 1/11,] so that
-
-ran{'pl9¢21' " 1¢k} = ran{¢1vM1[)la ' ‘ ' vMk—11/’k}
-
-we must chose 1/1;.“ so that, for any non-zero vector u 6 R",
-
-VP(‘I’I:a) 6 ra‘n{¢11 ‘21 - ° - 31/Jk-F1} 
-Now,
-vp<~I~.a) = gmm — P(‘I’kﬂ)‘I’k“)
-
-and therefore,
-
-Vp(‘II;,a.) E ran(M\II;,) U ra.n(\IIk) = ra.n{¢1, M1/)1, . . . , Mktﬁl}
-
-Thus, requirement (3.6) is met when ran{1/:1, 1/:2, . . . , 1/4+1} = ran{1/)1, M1/:1, . . . , Mk:/11}.
-
-In conclusion, the p-dimensional approximate invariant subspace which is most suitable
-
-for our objective is
-
-/C(Mn/mp) 2 ran{«/:1,M¢1,... ,M"“«/21}
-
-This is known as the K rylov subspace. Therefore, in answer to the problem posed at the outset
-of this section, if we take the columns of Q to be an orthonormal basis for IC(M,z/)1, p), then the
-
-eigenvalues of T = Q‘MQ should provide good estimates of p extremal eigenvalues of M.
-
-% 22.txt
-\chapter{Lanczos Procedures}
-
-The following will review the usual procedure for generating an orthonormal basis for
-lC(M, «/11, p), and conclude by showing that, for our problem, we can find the matrix Q‘MQ without
-actually carrying out this procedure. Note that this conclusion is essential since the procedure
-
-requires matrix vector multiplication - an operation we have assumed impossible for large enough :1.
-
-4.1 The General Lanczos Algorithm
-Let 21;“) E M"¢ and consider the d x :1 matrix
-«I 2 [«/»,M«/»,  ,M"“¢] 2 [«b‘°>,«/»“’,...,¢<‘-‘>1
-
-Notice that M\II = [1/:(1),... ,1/;(d'1),Md¢r] and, assuming for the moment that \I1 is non-singular,
-
-we can compute the vector h = —\II‘1M"1/J. Thus,
-
-M\II = \II[e2,. . . ,e,1, —h] E IIIH
-
-% 23.txt
-23
-
-Equivalently,
-0 0  0 -—h1
-1 0  0 —h2
-HE   3 ; =\II“M\II (4.1)
-0 "hl1—1
-1 —h,;
-
-H is a companion matrix which means that its characteristic polynomial is p(:r) = 2:‘ + 23:1 h,-:r"‘1.
-Since H is similar to M, ﬁnding the eigenvalues of M is equivalent to ﬁnding the roots of p(a:).
-However, this is of little practical use since ﬁnding h, constructing 12(1), and ﬁnding its roots is
-probably a harder problem than the one we started with. Instead, the value of decomposition (4.1)
-derives from its upper Hessenberg form. We exploit this property below.
-
-Let ‘II = QR be the QR decomposition of \II. Since ‘I! is assumed nonsingular,
-~II;‘M~r.: = (R"Q‘)M(QR) = H
-Therefore,
-Q‘MQ = RHR-1 2 T
-
-Since R and R“1 are both upper triangular and H is upper Hessenberg, T = R.HR'1 is also upper
-Hessenberg. Furthermore, since M is symmetric, it is clear that T‘ = Q‘MQ = T. That is, T is
-
-an upper Hessenberg matrix that is symmetric. Therefore, T is tridiagonal and we can write it as
-
-follows:
-
-T =    (4.2)
-
-ﬁd—1
-
-0  174-1 ad
-
-% 24.txt
-24
-
-Equating columns j on both sides of MQ = QT yields
-M113‘ = B,-—1qj—1 + 0:91 + ﬁjqj+1 (43)
-
-Since the columns of Q are orthonormal, multiplying both sides of (4.3) by q_,- and q,-+1 yields
-oz, = q§Mq,- and ﬂj = q;-+1Mq,-.
-The foregoing justifies what is called the Lanczos algorithm, performed as follows:
-
-The Lanczos algorithm for (partial) reduction to symmetric tridiaganal form
-
-91 = ¢/||¢||2, [30 = 0, 40 = 0
-
-for j=1top
-
-z=Mq_,-
-
-5:‘ = ||Z||2
-if 5, = 0, quit
-¢Ij+1 = 2/51
-
-end for
-
-The qj computed by the Lanczos algorithm are often called the Lanczos vectors. If the
-loop in the algorithm is terminated because ﬁg = 0, this indicates that an exact invariant subspace
-has been computed, and is given by ran{q1 - - - qk}. Otherwise, we usually halt the algorithm after
-p steps, in which case the algorithm converg to approximations of at most 12 eigenvalues.‘
-
-Note that equation (4.3) results from the equation MQ = QT. The latter equation holds
-since, starting with d vectors in our subspace, Q is a d Xd matrix whose columns form an orthonormal
-
-basis for all of R‘, which is clearly an invariant “sub” space. The Lanczos algorithm above, however,
-
-‘A detailed discussion of such convergence (and misconvergence) is given in
-
-% 25.txt
-25
-
-proceeds for only p steps, producing a d x p matrix Q, whose columns form an orthonormal basis
-for the approximate invariant subspace IC(M, ¢, p), and a p X p matrix T, = Q'MQ. In that case,
-IIMQ = QTp|[g =  is non-zero and gives the error bound described in Theorem 3.2.1. Now,
-
-writing the full d x :1 matrix T of (4.2) as
-
-01 51
-/71
-
-(4.4)
-
-allows us to describe the error bound ||E||-2 in terms of the submatrix Tm, and, hence, in terms of
-
-ﬁzz-
-Theorem 4.1.1
-
-HT, and T,“ are the matrices appearing in (4.4), and if the p columns of Q are computed by the
-
-Lanczos algorithm, then there exist pl, . . . ,;t,, E /\(T,) and A1, . . . , A, E A(M) such that
-
-ill» - /\I=i S iiTP'4”2 = 512
-
-fork=1,...,p.
-
-For a proof, see page 365 of [1].
-
-% 26.txt
-26
-
-4.2 A Lanczos Procedure for Markov Chains
-
-Notice that the Lanczos algorithm requires the computation of Mqj. Throughout this
-paper we have assumed that M 6 Rd“, and that d is so large as to make the matrix vector
-multiplication Mqj impossible. Now, the Lanczos algorithm constructs an orthonormal basis for
-the Krylov subspace. For our purposes, this construction is not essential. What is essential is that
-we ﬁnd a way to generate the Lanczos coejﬁcients oz,-, ﬁj, (j = 1, . . . , p), and whence the matrix
-T, without performing the operation Mqj required by the Lanczos algorithm. We now address this
-problem.
-
-Begin with the centered and weighted observable 1/1 = 111/ 2 (qS—E,,¢), and let q1 = 11;/||1/JH2.
-To simplify notation, let ¢(n) E ¢(X,.). The form of the first Lanczos coeﬂicient a1 is straight
-
-forward:
-
-01 = qiMq1
-
-I/1‘M1/I
-¢’¢
-
-and now the special properties of the Markov chain setting play an important role. For, regard the
-
-following
-
-¢‘M«/2 = (¢ — E.¢)‘II%‘MII%(¢ - Em)
-
-Cw (d>(n), ¢(" + 1)) (4-5)
-
-and
-
-W1!’ = (<15 - E«¢)'H(¢ - E«¢)
-
-Varn (¢(")) (4-6)
-
-Recall that the deﬁnition of M implies H%‘MI'I% = HP, and this justiﬁes equation (4.5). Putting
-
-% 27.txt
-27
-
-these equations together, we have a. nice form for :11:
-
-_ c, (run), an + 1))
-"‘ ' Ta:,(¢(n))
-
-The next Lanczos coeﬂicient, B1, is only slightly more complicated.
-
-[31 = ||M91-0t1tI1||2
-
-[(Mq1 — 0l1¢11)i(M¢]1 - 0101)] 1/2
-
-[t1iM2¢I1 — ail 1/2
-
-(4.7)
-
-Again recalling the deﬁnition M E 111/ 2P1'I‘1/2, a. simple manipulation veriﬁes
-
-1'li‘M’l'[% = HP’, whereby it follows that
-
-(«s — E,,¢)‘HP”(¢ — Em)
-
-"‘M2"‘ (¢ — E,¢)*n<¢ — Em)
-
-C: (¢("), ¢(n + 2))
-V3-Ur
-
-Inserting the last expression into equation (4.7) gives
-
-_ c,(¢(n).¢<n+2)) c,(¢<n>,¢(n+1)> ’ "’
-"“ jvar,<¢<n)> "(jvar,<¢<n)) ‘)
-
-Notice that we can estimate these expressions of oz; and ﬂl by running simulations of the Markov
-chain. To do so, we run the Markov chain starting from a random initial state, compute the value of
-the observable at each step of the chain and, from the data, compute estimates of the covariances
-and variance of the observable.
-
-Before generalizing this for the jth Lanczos coefficient, we consider one more coeﬂicient
-in detail. Recall that the second basis element q; in the Lanczos algorithm is given by 1;; =
-
-(Mql — oz1q1)/ﬁl. From this we can derive an expression for (12 which, though not as neat as those
-
-% 28.txt
-28
-
-for oz; and [31, is nonetheless tractable.
-
-I12 ‘IiMq2
-
-1
-[7(M91 — 0t1¢I1)tM(M¢I1 - 11191)
-1
-
-1
-= f,§(qiM3q1 - 2a1qlM2<11 + a’9iM¢11)
-1
-1
-= —2(<1iM3q1 - 2015i ‘ all (4-8)
-
-1.
-The last equality follows from equation (4.7), which can be written as q{M2q1 = off’ + (if. The
-preceding paragraphs deﬁne all but one of the quantities in the expression (4.8). The one remaining
-is q§M3q1, which can also be expressed in terms of covariances of <15, and the analysis is almost
-
-identical to that for q{M2q1:
-
-(¢ — E,¢)'IIP3(¢ — Em)
-
-"lM3"‘ (¢ — E,¢s)‘H(¢ - Em»)
-
-Cu (¢(")a ¢(n + 3))
-Varn(¢(n))
-
-To generalize the foregoing for higher order Lanczos coeﬂicients requires a recurrence
-relation between (aj+1, ﬂj) and (aj, ﬂ,~_1). As we will see, such a relation involves higher order
-Lanczos vectors q,~ — vectors we do not wish to compute. Recall that the only “vector” we can
-handle is our initial observable. Therefore, we must also establish a recurrence among the q,-’s, thus
-making them available inductively from our observable vector.
-
-To begin with the ﬂfs, recall that ﬁj E ||Mqj — ajq_.,- — ,H_,~_1q,-_;|[2. Whence,
-
-Bf (M411 — °‘j‘Ij - ﬁj-1qj—1)’(MlIj - Otjqj - 51-19141)
-
-q;-M291‘ - 0? - B,z_1 (4-9)
-
-Some simple algebra and the deﬁnitions aj E q§Mqj and ﬁJ~_1 E q;-Mq,--1 prove equality (4.9). The
-% 29.txt
-recurrence for 0114.1 is equally straight forward, though the ﬁnal form is not as neat:
-
-2
-0tj+1 ¢Ij+1M'1j+1
-
-1
-§(M¢15 - 11111;‘ - ﬂj—1<Ij—1)'M(Mqj - Otjqj - /3j—1qj—1)
-J
-
-1
-§(q;-M-“*4; — 2a,~q;M*q,- — 2x3;+’_1q;--1M*q,- + «I? + 2a,-5,51 + am (4.10)
-J
-From (4.9), q;~M2q,~ = oz? + ﬂf + ﬂ}_1 and, upon plugging this into (4.10), we arrive at
-1
-a,~+1 = E(q§»M3q_,,- — (1? — 2ozJ~ﬁJ2 — 2ﬂj_1q;»_1M2qj + ﬁ2_1) (4.11)
-J
-
-The expressions above involve the higher order Lanczos vectors qj. In the present context,
-these vectors will not be readily available. However, another recursion — this time relating forms
-involving q,-+1 to forms involving qj and qj__1 — will make these expressions accessible by induction
-on our initial observable. The general forms of interest are q;+1M"qj+1 and qj-+1M"qJ~. Again, we
-
-begin with the simpler of the two:
-
-¢I;'+1Mk9j = ij(Mqj - Otjqj - ﬂj—1t1j—1)'M"¢1j
-= i(<1§M"“¢Ij - C‘j‘I_;'Mk‘Ij - ﬂj—1q§M"qj-1)
-The second is,
-‘I_;'+1Mkqj+1 = ﬂi?(M<Ij - Otjqj - .3j—1<Ij—1)‘Mk(Mqj — Orjqj - ﬂj—1¢1j—1)
-
-1
-ﬁ(‘1§M'°"<1:‘ + °‘:2"I§M'°<J:' + ﬁ?—z<Ii—1M"<I:‘—1
-J
-—2a,-q;-M*+‘q,~ + 2am,--1q;-M*q,--1 — zza,--1q;-M*+‘q]--1) (4.12)
-
-The point of all this is that each pair of coeﬂicients (oz,-+1, /3,-) can be expressed as a
-function of the covariance between ¢(n) and ¢(n+ m), for m = 0, . . . , 21' + 1. Since the expressions
-are based on four recurrence relations, they become quite complicated as 1' gets large. However,
-deﬁning these recursions in a. computer program is trivial, and we use this fact in Section 5.4 when
-
-applying the foregoing to a particular problem.
-
-% 30.txt
-30
-
-Chapter 5
-
-An Application: Convergence to
-
-Gibbs Measure on the Ising Lattice
-
-5.1 Physical Motivation
-
-Before considering the details of the Metropolis algorithm, it helps to understand the
-setting in which it was developed. To do so we consider an typical example, the Ising model.
-Consider a model of a system consisting of 71 “sites”, each site taking values in {-1, +1}. It might
-help to imagine these sites as equally spaced points on a line or a circle, though we are not restricted
-to such cases. Then our state space 5' consists of all possible n—dimensional permutations of -1’s
-and +1’s. We reprent this as S’ = {-1, +1}", and note that |S| = 2". We can think ofeach site as
-a node existing in one of two possible positions, say “on" or “off”. Each state 1' E S is represented
-by a unique permutation a.- of —1’s and +1’s. Let a,-(k) = :|:l denote the position (on or off) of
-the kth node when the system is in state 1'. An observable ¢ = qS(a,-) on this space is a function of
-
-the permutations 47,-, 1' 6 {1,. . . , 2"}.
-
-% 31.txt
-31
-
-The energy of the system, when in state i, is deﬁned by the Hamiltonian
-
-H07-‘) = - Z Ua(i)0s(k)
-
-<j Ic>
-where the sum runs over all neart neighbor pairs. In statistical mechanics we are often concerned
-
-with the Gibbs measure of state i, deﬁned by
-1r(a,-) = Z'1 exp{—[3H(a,~)} (5.1)
-where Z ‘1 is the normalization constant, or partition function. For the Ising model, we put
-
-Z = §:9XP{~ﬁH(<7:‘)}
-i=1
-
-so that 7r is a probability measure. Now, letting 1: denote the so called Boltzman constant, if a
-classical mechanical system is in thermal equilibrium with its surroundings with absolute temper-
-ature T, and is in state i with energy H (05), then the probability density in phase-space of the
-point representing state i is given by (5.1) with [3 = (lcT)“. The central result of ergodic theory
-implies that we can also interpret 1r(o;) as the proportion of time the system spends in state i. If
-
-the system is observed at a random time, the expectation E,,¢ of any observable ¢ is thus
-
-.31 ¢(0a) 9"l>{-ﬂH(0s‘)}
-
-E” = 2?;.exp{—nH(a.-)}
-
-(5.2)
-
-Perhaps the number of sites n in our Ising model is so large that it is impractical or impossible to
-evaluate (5.2). We might instead consider importance sampling and generate states with probability
-
-density
-
-exp{—aH<a.o}
-2?;.exp{—zaH<a.->} (53)
-
-Then «I: is itself an unbiased estimator of (5.2).
-
-1r(0a') =
-
-If evaluation of (5.2) is difficult, there is no reason to believe that evaluation of (5.3) is any
-
-easier. However, Metropolis and his collaborators [9] contrived a method for producing an ergodic
-
-% 32.txt
-32
-
-Markov chain with transition probability matrix K whose elements Ic(:::, y) satisfy
-
-2«<z>k<x.y) = «<2»
-
-I
-
-By Deﬁnition 2.1.4 this means that the Markov chain has stationary distribution 7r. Since the
-drain is ergodic, we know that it will eventually converge to the desired distribution. So, our main
-problem — the rate at which such a Markov chain will converge — is clearly at issue in the application
-of the Metropolis algorithm. Moreover, as we will see below, the Metropolis algorithm ensures that
-the Markov chain has the desired stationary distribution by requiring that the chain satisfy the
-stronger condition of reversibility; i.e. detailed balance. Therefore, Markov chains produced by the
-Metropolis algorithm satisfy the assumptions of this paper, which fact conﬁrms that our theory
-
-can be used to bound convergence rates of all such chains.
-
-5.2 Description of the Metropolis Algorithm
-
-The following explains how the Metropolis Algorithm is carried out. The exposition is
-similar to that given in Hammersley and Handscomb’s book Monte Carlo Methods 
-To begin, choose an arbitrary symmetric Markov chain; that is, a Markov chain whose
-
-transition probability matrix P is symmetric. P is called the proposition matrix, and its elements,
-P071"), Satisfy
-pa, 1) 2 0. pm‘) = pm). 2,-pm‘) = 1 vi e S (5-4)
-
-Using P, we will derive a new transition probability matrix K which satisﬁes
-
-Z vr(z')k(='.:') = arm (5.5)
-
-I
-
-For 1' aé j deﬁne
-
-kw) = p<-yum")/«<¢> it nu)/1r(='> <1 (5.6)
-
-P(i,J') if 7r(J')/7r(i) Z1
-
-% 33.txt
-33
-
-For 2' = j deﬁne
-
-kw) = pm) (1—  p(-3;"), where J. = {j = 2‘ 7‘ an nu)/«(='> < 1}
-
-Notice that 1r(i) > 0 implies k(i,j) _>_ 0 and
-
-;k<«',j> = pow)  { (1 —  pm) +  +]§§p<«",:')
-= P(is'5) + 2 P(i1.1') + Z P(iy.7') (5-7)
-jel.» jug
-
-Equation (5.7) shows that, for each i, 2, k(z', j) = 1, which fact conﬁrms that K is a stochastic
-
-matrix.
-
-Next we check that K satisﬁes the detailed balance condition. First suppose that 1r(z') =
-
-1r(j). Then (5.4), (5.6), and symmetry of P imply
-k(i,J') = p(i2.7') = p(jvi) = k(j1i)
-
-and, since 1r(i) = 1r(j), the detailed balance condition, 1r(i)k(i,j) = 7r(j)k(i,j), clearly holds. Now
-
-suppose that 1r(i) > 1r(j). Then (5.4) and (5.6) imply k(j,i) =p(j, 1') and
-k(i,J') = P(iyJ')7r(J')/7r(i) = P(J3i)7r(J')/7r(i) = k(.1'» i)7r(J')/1r(i)
-
-That is, 1r(i)lc(i,j) = 1r(j)k(i,j), as we set out to prove.
-To complete the justification of the Metropolis algorithm, we note that the detailed balance
-condition, 7r(i)k(i, j) = 1r(j)k(i, j), implies that 1r is indeed the stationary distribution for a Markov
-
-chain with transition probability matrix K. To see this notice that
-
-27r(i)k(i..7') = Zﬂijlkiiajl = 1r(J') Z/€(i,J‘) = 7r(J')
-
-I i
-We now summarize the tasks performed when implementing the Metropolis algorithm.
-
-First, pick as a proposition matrix any symmetric stochastic matrix and denote it by P. Begin
-
-% 34.txt
-34
-
-the simulation of the Markov chain in state i, and then take one step of the proposition Markov
-chain to arrive in state j. That is, choose j according to the mass function p(z', -). Next, compute
-1r(j)/1r(i). If 7r(j)/1r(i) 2 1, accept j as the new state. If1r(j)/1r(i') < 1, then with probability
-1r(j)/7r(z') accept j as the new state; otherwise, with probability 1 — 1r(j)/1r(i), take i as the new
-state. Performing each of these tasks produces one step of the Markov chain represented by the
-
-transition probability matrix K.
-
-5.3 The Ising Model
-
-Returning to the Ising model considered in section 5.1, we consider how the Metropolis
-algorithm is applied to such a mode]. Again, start the simulated chain in state X0 = 1' and let
-
-X1 = j be a state chosen according to p(i, -), where p(i, j) = p(j,i). Next, compute
-
-7"(‘7j)/7"(‘7i) = eXP{-ﬁlH(<7j) - H(0i)l}
-
-Suppose the energy H (Uj) of the new state satisﬁes H (U,-) 3 H (05). Then 1r(a_,-)/1r(a,-) Z 1 and we
-
-move to state j. On the other hand, if H(aj) > H(a;), then 1r(aj)/7r(a,-) < 1 and we take
-
-j with probability e-M"
-
-X1 (5.8)
-
-i with probability 1 — e-W’
-where AH E H (0,) — H (:75) is the change in energy from state j to state 1'.
-There are many ways to choose a proposition matrix P. The simplest derives from the
-so called Glauber dynamics, which proceeds as follows: choose one of the 11. sites at random with
-probability 1/n — let .9 denote the chosen site — and change the position of this site; i.e., multiply
-
-its value by -1. The resulting state j is the permutation of -1’s and +1’s given by
-
--0,-(k) if]: = .9
-a,~(Ic)= (k=1,...,d)
-
-a,-(Ic) if]: 95 s
-
-% 35.txt
-35
-
-The procedure we employ below involves a slight variation: choose site .9 at random uniformly
-from the 11 possible sites. However, instead of changing the value of the chosen site with absolute
-certainty, as in Glauber dynamics, we “ﬂip the switch” with probability 1/2. Therefore, when k yé .9
-
-we have 0',-(k) = a.~(k), and when k = 3,
-
-—a,~(s) with probability 1/2
-01(3) =
-
-a,-(3) with probability 1/2
-The advantage of this procedure is that the proposition matrix, before altering it with the Metropo-
-lis algorithm, is a bit more interesting than that produced by traditional Glauber dynamics. For,
-it has 1/2 along the main diagonal and, thus, is aperiodic and converges to a stationary distribu-
-tion. Glauber dynamics has period 2, so its proposition matrix does not satisfy the hypotheses of
-
-our convergence theorem (Theorem 2.1.2). Keep in mind, however, that any proposition matrix
-
-modiﬁed by the Metropolis algorithm satisﬁes all of our hypotheses.
-
-5.4 Computations
-
-5.4.1 Computer Programs
-
-Constructing a computer program to simulate realizations of the Markov chain described
-above (i.e. with transition matrix K) is straight forward, and we do so using Matlab and the
-program ising.m which appears in the Appendix.
-
-When performing simulations in the present context, there are a few important aspects
-to consider. First, our program begins with a “hot start", which means that the initial state is
-picked by assigning each site to :l:1 with probability 1/2. Perhaps we should instead have picked
-our initial state from the stationary distribution. On the contrary, this would defeat the whole
-
-purpose of our analysis. Recall that the method has been developed to deal with cases for which
-
-% 36.txt
-36
-
-sampl from the stationary distribution are not quickly attainable. Furthermore, we are trying to
-ﬁgure out how long it will take for observations from the simulated chain to represent samples from
-the stationary distribution. Second, the method we have developed depends solely on covariances
-of our observable which we estimate from the data. produced by the simulations. However, such
-estimates require that the data come from the stationary distribution. Therefore, we must discard a
-number of observations before using the data to estimate covariances. The number of observations
-to discard depends on how long must we wait until we have data from the stationary distribution,
-and we are back where we started — almost. When employing our procedure, we can discard a
-conservative (very large) number of observations and perform our analysis, deriving a bound on
-the convergence time. Then, in all future studies, we will know how much data should be discarded
-before samples come (approximately) from the stationary distribution.
-
-In considering what observable to use on the state space described above, perhaps the
-most obvious is simply the number of nodes in the on position (i.e. the number of +l’s). Recalling
-that 0, (k) = i1 denotes the position (on or off) of the kth node when the system is in state j, this
-
-observable is deﬁned by,
-«M = zxca.-ac) = +1). 1 = 1.---.ISl
-I:
-
-where X is the indicator function. The program ising.m simulates the Markov chain, computes the
-values of this observable (as well as the number of 1’s squared) and writes it to a ﬁle called phi . dat.
-The program then proceeds to compute the ﬁrst few Lanczos coefficients using covariances produced
-by Matlab, so that we can check them against the rults we get from the 1anczos.c program,
-which we now describe.
-
-The lanczos .c program (listed in the Appendix along with its dependencies) implements
-the recursive relations described in section 4.2 to compute the Lanczos coefﬁcients by the new
-
-method. It accepts input from the user specifying any observable, any number of iterations, and any
-
-% 37.txt
-37
-
-TABLE 1: ESTIMATES or /\,m(K) = 0.998746 USING ¢1
-
-sim. # /32
-1 0.995107 0.006960 0.995220 0.600243
-2 0.995099 0.005891 0.994863 0.477393
-3 0.994862 0.009671 0.994925 N/ A
-4 0.995321 0.009319 0.995654 N/A
-5 0.994787 0.004872 0.405446 0.994742
-
-number of desired Lauczos coeﬁicients. That is, the observable could have come from a simulation
-of any reversible Markov chain and not just our special Metropolis chain. The output of this
-program is a T matrix containing the maximum number of Lauczos coeﬂicients (up to the number
-requested by the user) before an approximate invariant subspace was reached. The eigenvalues of
-the nonzero part of this matrix, in accordance with the foregoing theory, should well approximate
-
-eigenvalues of the Markov chain’s stochastic matrix.
-
-5.4.2 Results
-
-Table 1 shows the results of 5 simulations each of 100,000 iterations (discarding the ﬁrst
-5,000 observations) for the one dimensional Ising lattice with 10 nodes and [7 = 1. Displayed is
-the estimate of the ﬁrst Lauczos coeﬂicient a1 (which is the eigenvalue estimate after one step; i.e.
-).m,x(T1)), along with its error bound ﬁl. Appearing in the third column is the largest eigenvalue of
-the 2 x 2 matrix T2, A0 (T2), and ﬁg appears in the fourth column when it is was computable. When
-it was not computable, it was probably because an approximate invariant subspace was reached
-at the ﬁrst step, so ﬁg involved expressions which were close to machine epsilon and, thus, could
-not be computed accurately. Table 2 shows the same information when using a second observable,
-¢2 E ¢§ (i.e. the number of +1’s squared).
-
-The true value of the subdominant eigenvalue for this problem is /\max(K) = 0.998746,
-
-and the next largest is A2(K) = 0.993303. These were computed directly from the transition matrix
-
-% 38.txt
-38
-
-TABLE 2: ESTIMATES OF >.,,,,,.,,,(K) = 0.998746 USING z/>2
-
-sim. # B2
-1 0.994262 0.006142 0.994222 N / A
-2 0.994333 0.002974 0.994332 N / A
-3 0.994307 0.008892 0.994374 0.726320
-4 0.994436 0.012884 0.994685 0.361404
-5 0.993789 0.006909 0.993820 0.788710
-
-using the Matlab program tpm.m listed in the Appendix. They should be very accurate, and it
-takes the Matlab routine eig() just over 20 minutes of cpu time on a Sun Ultra to ﬁnd all the
-eigenvalues of K.
-
-As shown by Theorem 4.1.1, the [Ts provide conservative error bounds on the eigenvalue
-estimates from the previous Lanczos step. Unfortunately, for this example, ,6 appears to be too
-conservative as it does not even allow us to bound our estimates away from 1. Furthermore, it
-seems that the estimates generated by the method above fall systematically below the true value of
-the subdominant eigenvalue. In every simulation, an approximate invariant subspace (to machine
-tolerance) was reached after computing at most 2 pairs of Lanczos coefficients. For ¢1 we see that
-it was reached after computing only one pair on simulations 3 and 4, similarly for ¢2 on simulations
-1 and 2. These observations indicate that the chosen observable is close to the slowest mode (or
-eigenfunction) of the process. That is, we should observe fast convergence on an approximate
-invariant space. However, the space found might contain only the second slowest mode of the
-process and, in that case, our eigenvalue estimates would be closer to the third largest eigenvalue,
-instead of the second largest as desired.‘
-
-The matrix T was produced by the program lanczos . c and its eigenvalues were computed
-by Matlab, both operations taking a few seconds. Granted, comparing a few seconds to the 20
-
-minutes it takes Matlab is not justiﬁed since a simple call to the routine eig() is far from the
-
-‘Recall that we explained above why we will never converge to the eigenspace corresponding to the largest eigen-
-value (which is always 1). That would require a (nearly) constant observable function.
-
-% 39.txt
-39
-
-most eﬂicient traditional method for computing only a few eigenvalues of K. However, recall that
-the primary motivation for the new method are those examples where the matrix K is so large
-that we can’t store even a single vector in fast memory. Once we start swapping data to and from
-cheap memory, the time required by traditional algorithms can become years, even lifetimes. Such
-examples are not exceptional. Consider, for instance, the 1000 node problem which produces a.
-
-2“’°° x 21°°° transition matrix.
-
-% 40.txt
-40
-
-Chapter 6
-
-Conclusion
-
-We conclude by noting a few aspects of the special cases to which the foregoing applies.
-We are interested in the subdominant eigenvalue of a large transition probability matrix. The
-
-foregoing can be used to approximate these eigenvalues when the following conditions are satisﬁed:
-o The Markov chain is reversible (detailed balance condition).
-
-0 We can easily simulate realizations of the chain as well as functions (observables) of these
-
-realizations.
-
-When these conditions hold, we can side step the traditional Lanczos algorithm, and get at the
-values of the Lanczos coeﬂicients via the variational properties of observables on the state space.
-There are many ideas in this paper that should be expanded upon in future studies. Per-
-haps most obvious is the desire to ﬁnd analogous techniques for non-reversible (i.e. non-symmetric)
-Markov chains. This is a considerable problem since the Lanczos methods described above are no
-longer applicable. Perhaps the Amoldi algorithm ([1], algorithm 6.9), or the non-symmetric Lanczos
-
-algorithm can also be usefully specialized for stochastic matrices.
-
-% 41.txt
-41
-
-When applying the Lanczos algorithm, we usually choose a single starting vector (in the
-present context, an observable), perform the algorithm for a number of steps, and gather results.
-Then we do the same with a new starting vector and compare the results to those obtained with
-the ﬁrst vector. Proceeding in this way for a number of starting vectors, we can provide statistical
-evidence that we have, indeed, located the eigenvalues of interest. However, when there are two
-observables 451, :15; of particular interest, perhaps we would beneﬁt by considering both observables
-simultaneously. Future studies might develop an algorithm which incorporates both observables,
-
-again via ¢.- = H1/2(¢.' - E,.¢.-). In this case, a Lanczos algorithm would be based on the space
-
-Ic*(M,¢1,¢»2.p) 2 ran{«/»mz»2,M«/»1,M«/:2, . . .,MP—‘¢1.M"—‘«/22}
-
-The Lanczos coeﬁicients would then involve not only the autocovariances C, (¢,<(n),¢,-(n + m)),
-
-but also the cross-covariances C, (¢.~(n), ¢j(n + m)) (deﬁned in (2.6)).
-
-% 42.txt
-Appendix: computer programs
-
-Z Hatlab code ising.m
-
-Z
-Z Written by William J. De Meo on 12/15/97
-
-2 last modified 1/10/98
-
-Z Inputs:
-
-2 d = number of nodes of the ising lattice (e.g. d=100)
-Z n = number of iterations (e.g. n=1000)
-
-Z beta = Annealing schedule (e.g. beta = 2,
-
-1 when beta -> 0 will always go to nee state
-Z vhen beta -> infty will never go to state of higher energy)
-Z
-
-X = zeros(d,1);
-Energy = zeros(n,1);
-
-Ave=0;
-
-X Initialize using hot start
-for i=1:d,
-
-if rand < .5
-
-x(i) = -1;
-
-else
-
-X(i) = 1;
-
-end
-end
-
-Z Initialize at state of high energy (not used)
-
-if 0
-X(1:2:d-1)=1;
-X(2:2:d)=—1;
-and
-
-Z Initialize energy
-
-H=0;
-
-for i=1:(d-1)
-
-H = H - X(i)*X(i+1);
-end
-
-2 Initialize observable
-phi = zeros(n+1,1);
-
-for i=1:d
-if X(i) == 1
-phi(1)= phi(1)+1;
-end
-
-end
-
-42
-
-% 43.txt
-43
-
-for j=1:n
-
-site=0;
-while site==
-
-site = round(rand#d);
-end
-Z Compute new energy (with new X(site) = —x(site)):
-if site ==
-
-Hnew = H + 2:x(1)~x(2);
-elseif site == d
-
-Hnew = H + 2*X(d)tX(d-1);
-else
-
-Hnew = H + 2*X(site)*(X(site-1) + X(site+1));
-end
-
-Z Change to new state in two ways:
-X with probability 1/2, flip the switch
-if rand < .5
-if (Hnew <= H) I (rand < exp(-beta*(Hnew-H)))
-x(site) = -x(site);
-H = Hnew;
-Z Compute new value of observable
-if X(site) -= 1
-phi(j+1) = phi(j) + 1;
-
-else
-phi(j+1) = phi(j) — 1;
-end
-else phi(j+1)=phi(j);
-end
-else phi(j+1)=phi(j);
-end
-Energy(j) = H;
-end
-p1ot(Energy)
-
-% A second observable
-phi2 = zeros(n+1,1);
-for i=1:n+1
-
-phi2(i) = phi(i)*phi(i);
-end
-
-Z Write first observable to phi.dat:
-
-fid = fopen(’/quean.a/masters/deme5025/research/nyuthesis/programs/c/phi.dat’,’w’);
-fprintf(fid,’%f\n’,phi);
-fc1ose(fid);
-
-Z Write second observable to phi2.dat:
-fid = fopen(’/quean.a/masters/deme5025/research/nyuthesis/programs/c/phi2.dat’,’w’);
-
-% 44.txt
-fprintf(fid,’%f\n’,phi2);
-fc1ose(fid);
-
-1 Compute a few Lanczos coefficients by new method
-START=1001;
-
-V1=zeros(2);
-
-C1=zeros(2);
-
-C2=zeros(2);
-
-C3=zeros(2);
-
-V1 = cov(phi(START:n+1),phi(START:n+1));
-
-V1 = V1(1,1)5
-
-C1 = cov(phi(START:n),phi(START+1:n+1));
-C1 = C1(1,2);
-
-C2 = cov(phi(START:n-1),phi(START+2:n+1));
-C2 = C2(1,2);
-
-C3 = cov(phi(START:n-2),phi(START+3:n+1));
-
-C3 = C3(1,2);
-
-a1pha1 = C1/V1;
-
-betai = sqrt(C2/V1 ' (C1/V1)‘2);
-
-a1pha2 = (1/(beta1’2))*(C3/V1 - 2*a1pha1*(C2/V1) + a1pha1‘3);
-
-T1 = Ealphal betal; betal a1pha2]
-
-v1 = cov(phi2(START:n+1),phi2(START:n+1));
-v1 = v1(1,1);
-
-c1 = cov(phi2(START:n),phi2(START+1:n+1));
-c1 = c1(1,2);
-
-c2 = cov(phi2(START:n—1),phi2(START+2:n+1));
-c2 = c2(1,2);
-
-c3 = cov(phi2(START:n-2),phi2(START+3:n+1));
-c3 = c3(1,2);
-
-alphal = C1/V1;
-betal = sqrt(C2/V1 - (C1/V1)‘2);
-a1pha2 = (1/(beta1‘2))#(C3/V1 - 2ta1pha1*(C2/V1) + a1pha1”3);
-
-T2 = [a1pha1 betai; betal a1pha2]
-
-fid = fopen(’Tmats.dat’,’v’);
-fprintf(fid,’%f\n',T1);
-fprintf(fid,’%f\n’,T2);
-fc1ose(fid);
-
-ZZZ end ising.m
-
-44
-% 45.txt
-\% Matlab code tpm.m for generating transition probability matrix
-Z and computing its eigenvalues directly
-
-1
-
-Z Written by Hilliam J. De Heo on 1/10/98
-
-1
-
-Z Inputs:
-
-Z d = number of nodes of the ising lattice (e.g. d=10)
-
-Z beta = Annealing schedule (e.g. beta - 2,
-
-1 when beta -> 0 will always go to new state
-1 when beta -> infty will never go to state of higher energy)
-2
-
-disp(’bui1ding proposition matrix...’)
-
-states = 2‘d;
-
-A = 0;
-
-for i=0:(d-1)
-
-E=eye(2”i);
-
-A = [A E;E A];
-
-end
-
-B = .5#eye(states);
-
-A = B + .5-(1/d)*A;
-
-2 our modified Glauber dynamics requires the B and the .5¥(1/d)
-
-disp(’...done’)
-
-1 display pattern of nonzero entries
-1 spy(A)
-
-Z check that all row sums are 1
-check=O;
-for i=1:states
-
-check = check + (not(sum(A(i,1:states))<.99));
-end
-1 if not all rows sum to 1, print check = (# of rows with sum=1)
-if not(check==states)
-
-check
-
-error(’Row sums are not all 1’)
-end
-
-Z construct matrix of states
-E = zeros(d,states);
-f1ip=-1;
-for i=1:d
-for j=1:2‘(i-1):states
-flip = -1*f1ip;
-for k=0:2“(i-1)
-
-45
-
-% 46.txt
-E(i,j+k) = flip;
-and
-end
-end
-E=E’;
-
-Z display first 64 states
-Z E(1:64,:)
-
-Z compute energy of each state
-H = zeros(states,1);
-for i=1:states
-
-for j=1:d-1
-H(i) = H(i) - E(i,j)tE(i,j+1);
-end
-end
-
-disp(’building Hetropolis transition matrix...')
-
-for i=1:states
-for j=i+1:states
-if not(A(i,j)==0)
-if(H(j)>H(i))
-alt = A(i,j)t(1-exp(-beta*(H(j)-H(i))));
-A(i,i) I A(i,i)+a1t;
-A(i,j) = A(i,j)-alt;
-elseif(H(j)<H(i))
-alt = A(j,i)t(1-exp(—beta#(H(i)-H(j))));
-A(j,j) = A(j,j)+a1t:
-A(j,i) = A(j,i)-alt;
-end
-end
-end
-end
-
-disp(’...done’)
-
-check=0;
-for i=1:states
-
-check = check + (not(sum(A(i,1:states))<.99));
-end
-
-Z if not all rows sum to 1, print check = (# of rows with aum=1)
-
-if not(check==states)
-
-check
-
-error(’Row sums are not all 1’)
-end
-
-46
-
-% 47.txt
-47
-
-disp(’computing eigenvalues of tpm...’)
-cput = cputime;
-
-evals = eig(A);
-
-ecput = tputime - cput;
-
-disp(’the CPU time (in secs) for computing eigenvalues of tpm: ’)
-
-ecput
-ZZZ end tpm.m
-
-/¥#*##¥###*##*#l####¥#‘###$ﬂ###ﬁi##*U3*#**#*##**##¥***ll¥#*##¥ii
-
-* 1anczos.c main program for computing Lanczos coefficients *
-3 #
-* Created by William J. De Heo #
-t on 1/7/98 *
-# Last modified 1/10/98 #
-
-*¥#######**3It¥*#¥¥¥#**##**¥##¥#####**####*#*###*##**###¥##*##*/
-
-#inc1ude <stdlib.h>
-#include <stdio.h>
-#include <math.h>
-#inc1ude "prototypes.h"
-
-/# Machine constants on dino (Sun Ultra 1 at Courant) t/
-#define HACHEPS 1.15828708535533604981e-16
-
-#define SQRTEPS 1.07623746699106147048e-08
-
-ﬂdefine HAX_NAH 100
-
-void read_name(char *);
-
-/* Prototypes from the file functions.c ¥/
-
-double a1pha(int j);
-
-double betasq(int j);
-
-double form1(int j, int k);
-
-double form2(int j, int k);
-
-double moment(doub1e ‘data, long n, double tave, double tvar, double tcov, long k);
-
-/# external variables to be used by functions */
-double *phi,tcov, tvar, tave;
-
-% 48.txt
-main()
-
-{
-
-char tfilename;
-
-FILE *ofp;
-
-double temp=O, *T;
-
-long i, j, nrou, nlanc, START;
-int f1ag=O;
-
-filename = cma11oc(HAx_NAHE);
-
-printf("\nName of file containing observed values: “);
-
-read_name(fi1ename);
-
-printf(“\nTota1 number of observations in file (iterations):
-
-scanf(“Zu”,&nrov);
-
-printf(“\nNumber of leading observations to discard: “);
-
-scanf("Zu",&START);
-
-printf("\nNumber of Lanczos coefficients desired: “);
-
-scanf("%u",&n1anc);
-
-vhi1e(2*n1anc > nrow—START)
-
-{
-printf("\nNot enough data for that many coefficients.\n“);
-printf("\nEnter a smaller number of Lanczos coefficients: “);
-scanf("Xu“,&n1anc);
-
-}
-
-II);
-
-phi = dma11oc(nrow);
-cov=dma11oc(2#n1anc);
-var=dma11oc(1);
-ave=dma11oc(1);
-
-T = dma11oc(n1anc#n1anc); for(i=0;i<n1anc;i++) T[i]=(doub1e)0;
-
-/* observable phi is stored contiguously column-wise (e.g. by HATLAB) t/
-mat1abread(phi, nrow, 1,fi1ename);
-
-/* send the observable, offset by START */
-moment(phi+START,nrou-START,ave, var, cov,2*n1anc-1);
-
-/* First column of T */
-
-T[0] = a1pha(1);
-if((temp=betasq(1))>HACHEPS*10)
-{
-
-T[1]=sqrt(temp);
-
-/* General column of T */
-
-J'=2:
-
-for(i=1;i<n1anc-1, f1ag==O;)
-{
-
-T[itn1anc+i-1]=sqrt(betasq(i));
-
-48
-
-main()
-
-{
-
-char tfilename;
-
-FILE tofp;
-
-double temp=0, #T;
-
-long i, j, nrow, nlanc, START;
-int f1ag=0;
-
-filename = cma11oc(HAx_NAH);
-
-princf("\nName of file containing observed values: “);
-read_name(fi1ename);
-
-printf("\nTota1 number of observations in file (iterations): “);
-scanf("2u“,&nrow);
-
-printf("\nNumber of leading observations to discard: “);
-scanf("Zu",&START);
-
-printf("\nNumber of Lanczos coefficients desired: “);
-scanf("%u“,&n1anc);
-
-whi1e(2*n1anc > nrou-START)
-
-{
-printf("\nNot enough data for that many coefficients.\n");
-
-printf("\nEnter a smaller number of Lanczos coefficients: “);
-scanf("Zu",&n1anc);
-}
-
-phi = dma11oc(urow);
-cov=dma1loc(2*n1anc);
-var=dma11oc(1);
-ave=dma11oc(1);
-
-T = dma11oc(n1anc*n1anc); for(i=0;i<n1anc;i++) T[i]=(doub1e)O;
-
-/* observable phi is stored contiguously column-wise (e.g. by HATLAB) */
-mat1abread(phi, nrov, 1,fi1ename);
-
-/* send the observable, offset by START #/
-moment(phi+START,nrov-START,ave, var, cov,2*n1anc-1);
-
-/# First column of T */
-
-T[0] = a1pha(1);
-if((temp=betasq(1))>HACHEPS#10)
-{
-
-T[1]=sqrt(temp);
-
-/* General column of T */
-
-J‘=2:
-
-for(i-1;i<n1anc-1, f1ag==0;)
-{
-
-T[i*nlanc+i-1]=sqrt(betasq(i));
-
-48
-
-% 49.txt
-T[i*n1anc+i] = a1pha(i+1);
-if((temp = betasq(i+1))>HACmEPS#10)
-{
-
-T[i#n1anc+i+1] = sqrt(temp);
-
-i++;
-}
-else
-
-f1ag=1;
-
-}
-
-/t Last column */
-if(f1ag!=1 it ((temp= betasq(n1anc-1))>HACHEPSt10))
-
-{
-T[n1anc#n1anc-2] = sqrt(temp);
-T[n1anctn1anc-1] = a1pha(n1anc);
-printf("\nbeta(1) = %1f\nbeta(Zd) = Zlf (last beta)",
-T[1],(nlanc-1),T[nlanc*n1anc-2]);
-}
-else
-{
-printf("\nApproximate invariant space reached at step %d.",i);
-printf("\nbeta(1) = Z1f\nbeta(Zd) = Zlf (last accurate beta)“.
-T[1],i,T[itn1anc+i—1]);
-printf("\nbeta(Zd)“2: Zlf (first spurious resu1t)“,i+1,temp);
-}
-
-printf("\nThe matrix T is: \n");
-matprint(T,n1anc,n1anc);
-
-}
-
-else
-
-{
-
-printf("\nmain(): Approximate invariant space reached at first step.“);
-
-49
-
-printf("\na1pha(1) - Z1f\nbeta(1)”2: Zlf (first spurious result)“, T[O], temp);
-
-}
-ofp = fopen("Tmat.m","v“);
-check(ofp);
-mat1aburite(T,n1anc,n1anc,ofp);
-fc1ose(ofp);
-}
-void read_name(char tname)
-{
-int c, i = 0;
-while ((c = getchar()) != EOF it c != ’ ’ it c !- ’\n’)
-name[i++] = c;
-name[i] = ’\0’;
-}
-
-/#t* end 1anczos.c *#*/
-
-% 50.txt
-50
-
-/llnkttttttttinlntint*1!!!##1##!ttitilltit###t¥#tt¥ttItt####t#ttII=##mlltttt
-
-* functions.c functions required by 1anczos.c t
-# t
-* Created by Hilliam J. De Meo #
-* on 1/7/98 *
-t Last modified 1/10/98 :
-
-¥¥##¥$¥##¥¥#$¥¥¥###¥¥$*I#****#**####*######¥#¥#¥#$#¥¥#l¥**#****/
-
-#inc1ude <math.h>
-
-#define START 1000
-
-#define ITER 10000
-
-/t Machine constants on dino (Sun Ultra 1 at Courant) */
-
-#define HACHEPS 1.15828708535533604981e-16
-
-#define SQRTEPS 1.07623746699106147048e-O8
-
-double a1pha(int j);
-
-double betasq(int j);
-
-double form1(int j, int k);
-
-double form2(int j, int k);
-
-double moment(doub1e tdata, long n, double tave, double tvar, double tcov, int k);
-
-/t external variables to be used by functions */
-double *phi,tcov, *var, tave;
-
-double alpha(int j)
-{
-
-/* alpha is never called with j < 1 */
-if(j==1)
-return
-form1(1,1);
-else if(j>1)
-return
-((doub1e)1/betasq(j-1))*
-(form1(j-1,3) — pou(a1pha(j-1),3)
-- 2 * (alpha(j-1)#betasq(j-1) + sqrt(betasq(j-2))#form2(j-1,2))
-+ betasq(j-2));
-}
-
-double betasq(int j)
-
-{
-if(j==0)
-return (doub1e)O;
-else if(j>0)
-return (form1(j,2) - pow(a1pha(j),2) - betasq(j-1));
-}
-
-double form1(int j, int k)
-{
-
-% 51.txt
-51
-
-double form14,alphal,alpha1sq,form12,form13;
-if(j==0)
-
-return (doub1e)0;
-else if(j==1)
-{
-
-/# printf("\nvar = %1f, cov(%d) = %1f \n“,*var,k,cov[k]);t/
-
-return (cov[k])/(*var); /* the only real value */
-}
-else if(j>1)
-{
-
-return ((double)1/betasq(j-1))
-
-* ( form1(j-1,k+2)
-+ pow(a1pha(j-1),2)*form1(j-1,k)
-+ 2t(a1pha(j-1) t sqrt(betasq(j-2)) * form2(j-1,k)
-- a1pha(j-1)*form1(j-1,k+1)
-- sqrt(betasq(j—2))#form2(j-1,k+1) )
-+ betasq(j-2)*form1(j-2,k)):
-}
-}
-
-double form2(int j, int k)
-{
-/* form2 is never called with j < 1 */
-if(j==1)
-return (doub1e)0;
-else if(j>1)
-return
-(pov(betasq(j-1),—.5)) 3
-(form1(j-1,k+1) - a1pha(j-1)*form1(j-1,k)
-- sqrt(betasq(j-2)) # form2(j—1,k));
-}
-
-/* moment() function for computing var and cov(k)
-arguments:
-data = a nxi array of doubles
-n = length of data[]
-ave =(on exit)= the average of data[]
-var =(on exit)= the variance of data[]
-cov =(on exit)= the covariance of data[i] and data[i+j] for i-1,...,k
-k = the max lag for cov above
-*/
-double moment(double tdata, long n, double tave, double tvar, double tcov, int k)
-{
-/* Centered about data[0] algorithm: #/
-
-long i,j;
-double avel, ave2;
-
-% 52.txt
-52
-
-*ave=O;tvar=0; ave1=ave2=O;
-for(j=0;j<=k;j++)
-cov[j]=0;
-
-for(i=1;i<n;i++)
-{
-rave += (data[i] - data[0]);
-tvar += (data[i] - data[0])#(data[i] - data[0]);
-}
-tvar /= (doub1e)(n-1);
-Ivar -= (((*ave)/(doub1e)n) * ‘(*ave)/(doub1e)(n-1)));
-/# *ave = ((*ave)/(doub1e)n) + data[0]; <- that’s the true average (not needed)*/
-
-for(j=0;j<=k;j++)
-{
-
-for(i=0;i<(n-j);i++)
-
-{
-
-avel += (data[i] - data[0]);
-ave2 += (data[i+j] - data[0]);
-cov[j] += (data[i] - data[0])*(data[i+j] — data[0]);
-
-}
-
-avel/=(doub1e)(n-j); ave2/=(doub1e)(n-j);
-
-cov[j] = ((cov[j] - (double)(n-j)tave1¢ave2)/(double)(n-j-1));
-}
-
-}
-
-/trt end funccions.c ###/

### 2013-10-18 tex/MarkovChains.tex
finished chapter on Markov Chains

@@
 Next we assume that our state space $S$ is finite, and suppose for all states
-$i,j \in S$ that $p(i, j) \geq 0$, and $\Sigma_j p(i, j) = 1$ for all $i$. In
+$i,j \in S$ that $p(i, j) \geq 0$, and $\sum_j p(i, j) = 1$ for all $i$. In
 this case, equation~(\ref{eq:2.2}) takes a more intuitive form,
@@
 \[
-\pij \geq 0, \quad \Sigma_j \pij =1, \quad (i, j =1, 2, \dots, d)
+\pij \geq 0, \quad \sum_j \pij =1, \quad (i, j =1, 2, \dots, d).
+\]
 We also refer to $\P$ as the transition probability matrix.
@@
   \label{eq:2.4}
-\Sigma_x \pi(x) p(x,y) = \pi(x).
+\sum_x \pi(x) p(x,y) = \pi(x).
 \end{equation}
@@
 \[
-\tau_x(\epsilon) = \min\{m : \Delta_x(n) \leq \epsilon \text{ for all } n\geq m\}.
+\tau_x(\epsilon) = \min\{m : \Delta_x(n) \leq \epsilon \text{ for all } n\geq
+m\}.
+\]
 
@@
 eigenvalue which is second largest in absolute value, and we denote this
-eigenvalue by $\lambda_{\max} = \max{\lambda_1, |\lambda_{d-1}|}. 
+eigenvalue by $\lambda_{\max} = \max{\lambda_1, |\lambda_{d-1}|}$. 
 The function bounding the rate of convergence of a Markov chain is given by the
-following theorem:
+following theorem ($\log$ denotes the natural logarithm):
 \begin{theorem}
@@
 The quantity $\tau_x(\epsilon)$ satisfies
-1. $\tau_x(\epsilon) \leq (1-\lambda_\max)^{-1}(\log \pi(x)^{-1} + \log \epsilon^{-1})$;
-2. $\max_{x \in S} \tau_x(\epsilon) \geq \frac{1}{2} \lambda_\max(1-\lambda_\max)^{-1} \log(2\epsilon)^{-1}$.
+\begin{enumerate}
+\item 
+$\tau_x(\epsilon) \leq (1-\lambda_{\max})^{-1}(\log \pi(x)^{-1} + \log \epsilon^{-1})$;
+\item $\max_{x \in S} \tau_x(\epsilon) \geq \frac{1}{2} \lambda_{\max}(1-\lambda_{\max})^{-1} \log(2\epsilon)^{-1}$.
+\end{enumerate}
 \end{theorem}
@@
 subdominant eigenvalue and supply error bounds. Together, an approximation and error bounds
-for $\lambda_\max$  provide enough information to make Theorem~\ref{thm-2.1.2} useful.
+for $\lambda_{\max}$  provide enough information to make Theorem~\ref{thm-2.1.2} useful.
 
@@
 \[
-\E_\pi\phi_i = \Sigma_{x\in S} \pi(x)\phi_i(x)
+\E_\pi\phi_i = \sum_{x\in S} \pi(x)\phi_i(x)
 \]
@@
 \C_\pi (\phi_i(X_n),  \phi_i (X_{n+s})) 
-&= \E_\pi[(\phi_i(X_n) - \E_\pi\phi_i)(\phi_i(X_{n+s}) - \E_\pi\phi_i)]\nonumber\\
-&= \Sigma_{x, y\in S} \P_\pi(X_n = x, X_{n+s} = y) (\phi_i(x)- \E_\pi\phi_i)(\phi_i(y) - \E_\pi\phi_i)
+&= \E_\pi[(\phi_i(X_n) - \E_\pi\phi_i)(\phi_i(X_{n+s}) - \E_\pi\phi_i)]\\
+&= \sum_{x, y\in S} \P_\pi(X_n = x, X_{n+s} = y) (\phi_i(x)- \E_\pi\phi_i)(\phi_i(y) - \E_\pi\phi_i).\nonumber
 \end{align}
-which, by the definition of conditional probability, we can write as,
+By the definition of conditional probability, we can write~(\ref{eq:2.4}) as follows:
 \[
-\Sigma_{x, y\in S} \P_\pi(X_n = x) \P_\pi(X_{n+s} = y\mid X_n = x) (\phi_i(x)-
+\sum_{x, y\in S} \P_\pi(X_n = x) \P_\pi(X_{n+s} = y\mid X_n = x) (\phi_i(x)-
 \E_\pi\phi_i)(\phi_i(y) - \E_\pi\phi_i).
@@
 \[
-\Sigma_{x, y\in S} \pi(x)p^s_{xy} (\phi_i(x)-\E_\pi\phi_i)(\phi_i(y) - \E_\pi\phi_i).
+\sum_{x, y\in S} \pi(x)p^s_{xy} (\phi_i(x)-\E_\pi\phi_i)(\phi_i(y) - \E_\pi\phi_i).
 \]
@@
 &= \E_\pi[(\phi_i(X_n) - \E_\pi\phi_i)(\phi_j(X_{n+s}) - \E_\pi\phi_j)]\nonumber\\
-&= \Sigma_{x, y\in S} \pi(x)p^s_{xy} (\phi_i(x)-\E_\pi\phi_i)(\phi_j(y) - \E_\pi\phi_j).
+&= \sum_{x, y\in S} \pi(x)p^s_{xy} (\phi_i(x)-\E_\pi\phi_i)(\phi_j(y) - \E_\pi\phi_j).
 \end{align}
@@
 on the main diagonal and zeros elsewhere. 
-Finally, denoting by $C(s)$ the $p \times p$ covariance matrix
-whose $(i,j)$ element is $C_\pi(\phi_i(X_n), \phi_j(X_{n+s}))$, we have
+Finally, denoting by $\C(s)$ the $p \times p$ covariance matrix
+whose $(i,j)$ element is $\C_\pi(\phi_i(X_n), \phi_j(X_{n+s}))$, we have
 \begin{align*}
-C(0) &= \E(\Phi(X_n) - \<\Phi\>)(\Phi(X_n) - \<\Phi\>)^t\\
+\C(0) &= \E(\Phi(X_n) - \<\Phi\>)(\Phi(X_n) - \<\Phi\>)^t\\
 &= (\Phi - \<\Phi\>)^t\Pi (\Phi-\<\Phi\>), \\
-C(s) &= \E(\Phi(X_n) - \<\Phi\>)(\Phi(X_{n+s}) - \<\Phi\>)^t\\
+\C(s) &= \E(\Phi(X_n) - \<\Phi\>)(\Phi(X_{n+s}) - \<\Phi\>)^t\\
 &= (\Phi - \<\Phi\>)^t \Pi \P^s (\Phi - \<\Phi\>).
@@
 \label{eq:2.7}
-C(s) &= 
+\C(s) &= 
 (\Phi - \<\Phi\>)^t \Pi^{\frac{1}{2}t}\M^s\Pi^{\frac{1}{2}}(\Phi - \<\Phi\>)\nonumber\\
 &=\Psi^t \M^s \Psi,  % Eqn 2.7
-\end{align*}
+\end{align}
 %% \\Phi\Phi‘M’\\Phi\Phi (2.7)
@@
   \M \text{ is symmetric } 
-& \Leftrightarrow 
+& \quad \Longleftrightarrow  \quad
  (\Pi^{1/2}\P\Pi^{-1/2})^t = \Pi^{1/2}\P\Pi^{-1/2}\\
-& \Leftrightarrow 
+& \quad \Longleftrightarrow  \quad
  \Pi^{-\frac{1}{2}t}\P^t\Pi^{\frac{1}{2}t} = \Pi^{1/2}\P\Pi^{-1/2}\\
-& \Leftrightarrow 
+& \quad \Longleftrightarrow  \quad
  \P^t\Pi^t = \Pi\P.

### 2013-10-18 tex/abstract.tex
finished chapter on Markov Chains

@@
 \begin{center}
-{\bf Abstract}\\
-\\
-A Lanczos Procedure for Approximating Eigenvalues of Large Stochastic Matrices\\
-\\
-by\\
-\\
-William J. DeMeo\\
-\\
-Master of Science in Mathematics\\
-\\
-New York University\\
-\\
+\newcommand\skipsize{6pt}
+A Lanczos Procedure for Approximating Eigenvalues of Large Stochastic Matrices\\[\skipsize]
+by\\[\skipsize]
+William J. DeMeo\\[\skipsize]
+Master of Science in Mathematics\\[\skipsize]
+New York University\\[\skipsize]
 Professor Jonathan Goodman, Chair

### 2013-10-18 tex/InvariantSubspaces.tex
finished typsetting InvariantSubspaces chapter

@@
 \section{Invariant Subspaces}
-Definition 3.1.1 A subspace S Q R" with the property that
 
-1:ES'=Ma:€S
+\begin{definition}
+%% Definition 3.1.1
+A subspace $S \subseteq \R^n$ with the property that
+\[
+x \in S \quad \Longrightarrow \quad \M x \in S
+\]
+is said to be \emph{invariant} for $\M$.
+\end{definition}
 
-is said to be invariant for M.
-
-Recall that, having chosen observables <I> = ((151, . . . , qfip), we constructed the covariance
-matrix C(s) = \II‘M"\II. If the column space of \II, which we denote by ran(\II), is an invariant
-subspace for M, the definition implies Mifij E  That is, for each 111,- there exists a vector
-t of coeﬂicients such that M1/Jj = f=1t,-1/)5,-. This is true for all j and, putting each vector of
-coefficients into a matrix T, we see that M\II = \IIT. Conversely, M\II = IIIT implies that M1/Jj is a
-
-linear combination of columns of \II, so ran(\II) is invariant. We have thus proved the following
+Recall that, having chosen observables $\<\Phi\> = (\phi_1, \dots, \phi_p)$, we
+constructed the covariance matrix 
+$\C(s) = \Psi^t \M \Psi$. 
+If the column space of $\Psi$, which we denote by $\ran(\Psi)$, is an invariant
+subspace for $\M$, the definition implies $\M\psi_j \in \ran(\Psi)$.
+That is, for each $\psi_i$ there exists a vector
+$t$ of coeﬂicients such that $\M \psi_j = \sum_{i=1}^p t_i \psi_{ij}$. 
+This is true for all $j$ and, putting each vector of
+coefficients into a matrix $\T$, we see that 
+$\M \Psi = \Psi \T$. Conversely, $\M\Psi = \Psi \T$ implies that $\M\psi_j$ is a
+linear combination of columns of $\Psi$, so $\ran(\Psi)$ is invariant. We have
+thus proved the following 
 % 15.txt
-Fact 3.1.1 The subspace ran(\II) is invariant for M if and only if there exists T E R”? such that
-
-M\II = \IIT.
+\begin{fact}
+%% Fact 3.1.1 
+\label{fact:3.1.1}
+The subspace $\ran(\Psi)$ is invariant for $\M$ if and only if there exists 
+$\T \in \R^{p\times p}$ such that $\M\Psi = \Psi \T$.
+\end{fact}
 
 Consequently,
-
-Fact 3.1.2 )\(T) Q ).(M)
-
-Proof of 3.1.2:
-
-z\E)\(T) _=_ 3v€R"3Tv=).v
-<=> \IITv = A\IIv
-<=> M\II-u = /\\IIv
-< /\ E /\(M)
-
-The second ¢> follows from fact 3.1.1.
-Facts 3.1.1 and 3.1.2 are theoretically useful. To see why, write the equation of fact 3.1.1
-
-as follows:
-
-\IIT = M\II
-\II‘\IIT = \II‘M\II
-T = (\1:'~1r)-‘\1:'M\1:
-
-In this form, we recognize that T = C‘1(0)C(1). That is, using only the covariance of observables
-on the state space, we can generate a matrix T with the property A(T) (_Z ).(M). Recalling that
-M = H1/2Pl'I‘1/2, we see that M is similar to our original stochastic matrix P, and thus ).(M) =
-
-).(P).
+\begin{fact}
+%% Fact 3.1.2
+\label{fact:3.1.2}
+$\lambda(\T) \subseteq \lambda(\M)$.
+\end{fact}
+\begin{proof}
+%Proof of 3.1.2:
+  \begin{align*}
+\lambda \in \lambda(\T)
+& \quad \Longleftrightarrow  \quad
+(\exists v \in \R^p) \T v = \lambda v\\
+& \quad \Longleftrightarrow  \quad
+\Psi \T v = \lambda \Psi v
+& \quad \Longleftrightarrow  \quad
+\M \Psi v = \lambda \Psi v
+& \quad \Longleftrightarrow  \quad
+\lambda \in \lambda(M).
+  \end{align*}
+The second equivalence follows from Fact~\ref{fact:3.1.1}.  %3.1.1.
+\end{proof}
+%Facts~\ref{fact:3.1.1} and~\ref{fact:3.1.2} are theoretically useful. To see why
+To see why Facts~\ref{fact:3.1.1} and~\ref{fact:3.1.2} are theoretically useful,
+consider the equation of Fact~\ref{fact:3.1.1}:
+\begin{align*}
+& \Psi \T = \M \Psi \\
+\Longleftrightarrow \quad & \Psi^t\Psi \T = \Psi^t \M \Psi\\
+\Longleftrightarrow \quad & \T = (\Psi^t\Psi)^t \Psi^t \M\Psi.
+\end{align*}
+In this form, we recognize that $\T = \C^{-1}(0)\C(1)$. That is, using only the
+covariance of observables on the state space, we can generate a matrix $\T$ with
+the property $\lambda(\T) \subseteq \lambda(\M)$. Recalling that 
+$\M = \Pi^{1/2}\P\Pi^{-1/2}$, 
+we see that $\M$ is similar to our original stochastic matrix $\P$, 
+and thus $\lambda(M) = \lambda(P)$.
 
 \section{Approximate Invariant Subspace}
+% Sec 3.2
+\label{sec:appr-invar-subsp}
+%% We qualified the foregoing by stating that the facts are only theoretically
+%% useful. This
+We noted above that Facts~\ref{fact:3.1.1} and~\ref{fact:3.1.2} are
+\emph{theoretically} useful. Speaking practically, however, 
+%% This is because 
+when choosing observables on the state space we may not be sure that
+they will satisfy the 
+%
+%
+% -------------16.txt---------------------------------------------------
+%
+%
+primary assumption underlying these facts. %Facts 3.1.1 and 3.1.2. 
+Recall the assumption: $\ran(\Psi)$ is an invariant
+subspace for $\M$ where $\Psi = \Pi^{1/2}(\Phi - \<\Phi\>)$. It may well be the
+case that there is a $\psi_j\in \ran(\Psi)$ such that 
+$M\psi_j \notin \ran(\Psi)$, thereby violating the assumption. 
+However, lacking an invariant subspace, we 
+might hope that our observables at least provide an \emph{approximate invariant subspace}.
+\begin{definition}
+%% Definition 3.2.1
+If the columns of $\Psi\in \R^{d\times p}$ are independent and the 
+\emph{residual matrix} $\E = \M\Psi - \Psi \T$ is small for some 
+$\T \in \R^{p\times p}$, then
+$\ran(\Psi)$ defines an \emph{approximate invariant subspace}. 
+\end{definition}
+To see how an approximate invariant subspace can be useful for approximating
+eigenvalues of $\M$, we recall a theorem from Golub and Van Loan~\cite{Golub:1996}.
+\begin{theorem}
+% Theorem 3.2.1
+\label{thm:3.2.1}  
+Suppose $\M \in R^{d\times d}$ and $T \in \R^{p\times p}$ are symmetric and let
+$E = \M\Q - \Q\T$, where $\Q \in \R^{d\times p}$ is orthonormal 
+(i.e., $\Q^t\Q = \I$). Then there exist $\mu_1,\dots, \mu_p \in \lambda(\T)$ and
+$\lambda_1, \dots, \lambda_p \in \lambda(\M)$ such that
+\[
+|\mu_k - \lambda_k| \leq \sqrt{2}\|\E\|_2, \quad \text{ for $k = 1, \dots, p$.}
+\]
+\end{theorem}
+If the subspace $\ran(\Q)$ is an approximate invariant subspace, then the definition implies that
+there is a choice $\T$ rendering the error $\|\E\|_2$ small, and thus the
+eigenvalues of $\T$ provide a good approximation to those of $\M$.
 
-We qualiﬁed the foregoing by stating that the facts are only theoretically useful. This
+When considering the foregoing ideas, it is apparent that their application
+presents new---but hopefully less prohibitive---problems. As these problems
+are the focus of the rest of the paper, now is a good time to examine
+them.
 
-is because when choosing observables of the state space we aren't sure that they will satisfy the
-% 16.txt
-primary assumption underlying facts 3.1.1 and 3.1.2. Recall the assumption: ran(\II) is an invariant
-subspace for M, where ‘II = H1/2(<I> —  It may well be the case that there is a 1p; 6 ran(\II) such
-that M1/zj ¢ ran(\II), thereby violating the assumption. However, lacking an invariant subspace, we
+First, the preceding theorem assumes a matrix $\Q$ whose columns form an
+orthonormal basis for the approximate invariant subspace. For our problem, 
+derivation of such a $\Q$ is tricky, and we must prepare for this.
 
-might hope that our observables at least provide an approximate invariant subspace.
+Next, having an approximate invariant subspace at our disposal merely tells us
+that there exists some matrix $\T$ which makes the error 
+%
+%
+% ------------- 17.txt ----------------------------------------------------
+%
+%
+$\|\E\|_2$ small. We must discover the form of such a $\T$. Moreover, it is
+natural to seek that $\T$ which minimizes $\|\E\|_2$ for a given approximate
+invariant subspace.
 
-Deﬁnition 3.2.1 If the columns of ‘II E R“? are independent and the residual matrix E E
+Finally, in order to apply these ideas to realistic eigenvalue problems, we must
+find a practical way to generate %% the most appropriate
+a good approximate invariant subspace.
 
-M\II —— \IIT is small for some T 6 R7”, then ran(\II) deﬁnes an approximate invariant subspace.
+We now address each of these issues in turn.
+%the order raised.
 
-To see how an approximate invariant subspace can be useful in approximating eigenvalues of M,
-we extract a theorem from Golub and Van Loan 
-% Theorem 3.2.1
+Recall the matrix $\Psi = \Pi^{1/2}(\Phi — \<\Phi\>)$. The columns of this
+matrix, though independent (by choice of independent observables), are not
+necessarily orthonormal. However, consider the polar decomposition 
+$\Psi = \Q \Z$, where $\Q^t\Q = \I$ and $\Z^2 = \Psi^t\Psi$ is a symmetric
+positive semidefinite matrix.\footnote{Recall that the polar decompostion is
+  derived from the singular value decomposition, $\Psi = \U\Sigma \V^t$, by
+  letting $\Q = \U \V^t$ and $\Z = \V\Sigma \V^t$.}
+%% ‘Recall that the polar decomposition is derived from the SVD, \Psi = UZJV‘, by letting Q = UV‘ and Z = VEV‘.
+Note that $\Z = (\Psi^t\Psi)^{1/2}$ is nonsingular, so $\Q$ has the form
+\[
+\Q = \Psi \Z^{-1} = \Psi(\Psi^t\Psi)^{-1/2},
+\]
+and it is clear that $\ran(\Q) = \ran(\Psi)$. 
+Perhaps $\ran(\Q)$ is a useful approximation to the invariant
+subspace for $\M$. If $\ran(Q)$ is not itself invariant, we have the error 
+matrix $\E =\M\Q - \Q\T$.  Below we show that the
+$\T$ which minimizes $\|\E\|_2$ is $\T = \Q^t\M\Q$. 
+This yields the following
 \begin{theorem}
-\label{thm-3.2.1}  
-Suppose M 6 Rd” and T 6 RP"? are symmetric and let E E MQ — QT, where Q 6 Rd"? satisﬁes
+%Theorem 3.2.2
+\label{thm:3.2.2}  
+If $\Psi = \Q\Z$ is the polar decomposition of $\Psi$, then the matrix
+\[
+\T = \Q^t\M\Q = (\Psi^t\Psi)^{-1/2}\Psi^t \M^t \Psi (\Psi^t\Psi)^{-1/2}
+\]
+minimizes $\|\E\|_2 = \|\M\Q - \Q\T\|_2$.
+\end{theorem}
+\begin{proof}
+We prove the result by establishing the following
+\\[6pt]
+\underline{Claim:} Given $\M \in \R^{d\times d}$ suppose $Q \in  \R^{d\times d}$
+satisfies $\Q^t\Q =\I$. 
+Then,
+\[
+\min_{\T\in \R^{p\times p}} \|\M\Q - \Q\T\|_2 = \|(\I- \Q\Q^t)\M\Q\|_2,
+\]
+and $\T = \Q^t\M\Q$ is the minimizer.
+%
+%
+% -------------- 18.txt -------------------------------
+%
+%
+The claim is verified by an easy application of the Pythagorean theorem:
+For any $\T \in \R^{p\times p}$ we have
+\begin{align}
+\label{eq:3.1}
+\|\M\Q - \Q\T\|^2_2 &= \|\M\Q - \Q\Q^t\M\Q + \Q\Q^t\M\Q - \Q\T\|^2_2 \nonumber\\
+&=\|(\I - \Q\Q^t)\M\Q + \Q(\Q^t\M\Q - \T)\|^2_2 \nonumber \\
+&=\|(\I - \Q\Q^t)\M\Q\|^2_2 + \|\Q(\Q^t\M\Q - \T)\|^2_2 \\
+&=\|(\I - \Q\Q^t)\M\Q\|^2_2 \nonumber
+\end{align}
+Equality~(\ref{eq:3.1}) holds since $\I - \Q\Q^t$ projects $\M\Q$ onto the
+subspace orthogonal to 
+$\ran(Q)$. Thus, the two terms in the expression on the right are orthogonal,
+and the Pythagorean theorem yields equality. The concluding inequality
+establishes that the minimizing $\T$ is that which annihilates the 
+second term in~(\ref{eq:3.1}), that is, $\T = \Q^t\M\Q$.
 
-Q‘Q = I. Then there exist p1,... ,;z,, E A(T) and A1,. . . , A,, E A(M) such that
-ll‘k - Akl S \/§|lE"
+The second equality in Theorem~\ref{thm:3.2.2} is a consequence of the polar decomposition, in
+which $\Q = \Psi \Z^{-1} = \Psi(\Psi^t\Psi)^{-1/2}$. Therefore,
+$\T = (\Psi^t\Psi)^{-1/2}\Psi^t \M^t \Psi (\Psi^t\Psi)^{-1/2}$
+is the minimizer, as claimed.
+\end{proof}
 
-fork=1,...,p.
-\end{theorem}
+\subsection{The Krylov Subspace}
+% Sec 3.3
+\label{sec:krylov-subspace}
+Suppose the columns of a matrix $\Q \in \R^{d\times p}$ give an orthonormal
+basis for an approximate invariant subspace. Then, as we have seen, 
+\begin{enumerate}
+\item 
+$\T = \Q^t\M\Q$ minimizes $\|\E\|_2 = \|\M\Q - \Q\T\|_2$ and
+\item $\exists \mu_1, \dots, \mu_p\in \lambda(\T)$ and 
+$\lambda_1,\dots, \lambda_p \in \lambda(\M)$ such that
+\[
+\|\mu_k-\lambda_k\| \leq \sqrt{2}\|\E\|_2, \quad \text{ for $k=1,\dots, p$.}
+\]
+\end{enumerate}
+%
+%
+% ---------------- 19.txt --------------------------------------------
+%
+%
+Given an approximate invariant subspace $\ran(\Q)$ of dimension $p$, these facts
+tell us what matrix we should use to approximate $p$ elements of the spectrum of
+$\M$. Now all we lack is a description of $\ran(\Q)$. 
+That is, we have not specified which approximate invariant subspace would best
+suit our objective of approximating the subdominant eigenvalue
+$\lambda_{\max}(\M)$. For this purpose the following definition is useful:
+\begin{definition}
+%% Definition 3.3.1
+The \emph{Raleigh quotient} of a symmetric matrix $\M$ and a nonzero vector $x$ is
+\[
+\rho(x, \M) = \frac{x^t\M x}{x^tx}.
+\]
+\end{definition}
+We will denote Raleigh quotient by $\rho(x)$ when the context makes clear what
+matrix is involved.
 
-If the subspace ran(Q) is an approximate invariant subspace, then the deﬁnition implies that
-there is a choice T rendering the error IIEII2 small, and thus, the eigenvalues of T provide a good
-approximation to those of M.
-
-When considering the preceding ideas, it becomes apparent that their application will
-present new — but hopefully less prohibitive — problems. As these problems summarize the rt of
-the paper, now is a good time to examine them. First, the preceding theorem assumes a matrix Q
-whose columns form an orthonormal basis for the approximate invariant subspace. For our problem,
-derivation of such a Q is tricky, and we must prepare for this. Next, having an approximate invariant
-
-subspace at our disposal merely tells us that there exists some matrix T which makes the error
-% 17.txt
-"EH2 small. We must discover the form of such a T. Moreover, it is natural to seek that T which
-minimizes IIEII2 for a given approximate invariant subspace. Finally, in order to apply these ideas
-to real examples of our eigenvalue problem, we must ﬁnd a practical way to generate the most
-appropriate approximate invariant subspace. We now address each of these issues in the order
-raised.
-
-Recall the matrix I1 E III/2(<I> — (<I>)). The columns of this matrix, though independent
-(by choice of independent observables), are not necessarily orthonormal. However, consider the
-polar decomposition ‘II = QZ where Q‘Q = I and Z2 = \Il‘\II is symmetric positive semideﬁnite.1
-
-Note that Z = (\II‘\II)l/2 is nonsingular, so Q has the form
-Q = ~1:z-1 = \II(\II‘\II)“/2
-
-and it is clear that ran(Q) = ran(\II). Perhaps ra.n(Q) is a. useful approximation to the invariant
-subspace for M. If ran(Q) is not itself invariant, we have the error E = MQ — QT. It is easy to
-show, as we do in the following proof, that the T which minimizes HE“; is T = Q‘MQ. This yields
-
-the following,
-Theorem 3.2.2
-
-If \I1 = QZ is the polar decomposition of \II, then the matrix
-T = Q‘MQ = (\1:*\1:)-1/’\1:‘M\I:(\Ir‘\I:)-1/2
-minimizes IIEII2 = IIMQ - QTII2
-
-Proof:
-We prove the result by establishing the following
-
-(Hal: Given M 6 Rd” suppose Q 6 Rd"? satisﬁes Q'Q =1. Then,
-
-min ||MQ — QTII2 = H(1- QQ')MQll2
-
-TERI‘?
-
-‘Recall that the polar decomposition is derived from the SVD, \II = UZJV‘, by letting Q = UV‘ and Z = VEV‘.
-
-% 18.txt
-and T = Q’MQ is the minimizer.
-To verify the claim, observe the following application of the Pythagorean theorem:
-
-For any T E R.”"” we have
-
-IIMQ — QT||§ = IIMQ - QQ'MQ + QQ'MQ - QT|l§
-||(I - QQ’)MQ + Q(Q'MQ - T) H3
-
-||(1 - QQ‘)MQ||§ + |lQ(Q'MQ - T)||§ (3-1)
-
-2 l|(I — QQ')MQ||§
-
-Equality (3.1) holds since I — QQ' projects MQ onto the subspace orthogonal to ran(Q). Thus,
-the two terms in the expression on the right are orthogonal, and the Pythagorean theorem yields
-equality. The concluding inequality establishes that the minimizing T is that which annihilates the
-second term in (3.1); i.e. T = Q‘MQ.
-
-The second equality in Theorem 3.2.2 is a consequence of the polar decomposition, in
-
-which Q = \IIZ“ = \I:(~I:‘\I:)-1/2. Therefore,
-T = (x1z'x1:)-1/2x1:‘M~1:(\1:*~1z)—1/2
-
-is the minimizer.
-
-3.3 The Krylov Subspace
-
-Suppose the columns of a matrix Q E R“? provide an orthonormal basis for an approxi-
-
-mate invariant subspace. Then we have seen,
-1. T = Q'MQ minimizes ||E||2 = ||MQ — QT||2 and
-2. E|p.1,...,p,, E z\(T) and /\1,...,A, 6 »\(M) such that
-
-ll-‘I: - Ml S \/§|lE||2
-
-% 19.txt
-19
-fork=1,...,p.
-
-Given an approximate invariant subspace ran(Q) of dimension 1), these facts tell us what matrix we
-should use to approximate p elements of M’s spectrum. Now all we lack is a. description of ran(Q).
-That is, we have not speciﬁed which approximate invariant subspace would best suit our objective
-
-of approximating the subdominant eigenvalue A,,.a2:(M). 'I‘o do so requires the following deﬁnition:
-
-Deﬁnition 3.3.1 The Raleigh quotient of a symmetric matrix M and a nonzero vector 3 is
-
-p(2:, M) E (:z:‘Ma:)/(z‘a:).
-
-We will also denote this by p(z) when the context makes clear what matrix is involved.
-
-To ﬁnd the approximate invariant subspace most appropriate for our problem, we choose
-each dimension successively, providing justiﬁcation at each step. Thus, start with one observable
-42 on the state space, and let zpl E 111/ 2(¢ — E,,¢). That is, 1/:1 represents a centered (mean zero)
-observable whose ith coordinate is weighted by . First notice that the deﬁnition (which comes
-directly from the deﬁnition of ‘II following equation (2.7)) is such that 1/)1 is a constant vector if
-and only if :15 is oonstant on the state space, in which case 1/); E 0. This fact provides one reason
-observables that are constant on the state space are not interesting. Second, recall that the row
-sums of the matrix P are all one and, therefore, the eigenvector corresponding to the eigenvalue
-Ao(P) = 1 is the constant vector. By deﬁnition of M E 1'11/2Pl'I‘1/2, we see that the constant vector
-is also the eigenvector corresponding to A0  This will play an important role in what follows, as
-it allows us to focus primarily on the subdominant eigenvalue z\max(M) E max{}q (M), [A44 
-rather than on A0  (which we already know is 1).
+To find the approximate invariant subspace most appropriate for our problem, we
+choose each dimension successively, providing justification at each step. 
+We start with one observable $\phi$ on the state space, and let 
+$\psi_1 = \Pi^{1/2}(\phi -\E_\pi\phi)$. That is, $\psi_1$ is a centered (mean
+zero) observable whose $i$th coordinate is weighted by $\sqrt{\pi(i)}$. 
+Notice that the definition 
+%% (which comes directly from the definition of ^tII
+%% following equation (2.7)) is such that 1/)1 is a constant vector if 
+(which comes from the definition of $\Psi$ following Equation~(\ref{eq:2.7})) 
+is such that $\psi_1$ is a constant vector if and only if $\phi$ is oonstant on the
+state space, in which case $\psi_1$ is the constant zero function. 
+%% This fact
+%% provides one reason observables that are constant on the state space are not
+%% interesting. 
+(Observables that are constant on the state space are not interesting.) 
+Second, recall that the row sums of the matrix $\P$ are all one, therefore the
+eigenvector corresponding to the eigenvalue $\lambda_0(\P) = 1$ is the constant
+vector.  By definition of $\M = \Pi^{1/2}\P\Pi^{-1/2}$, we see that the constant
+vector is also the eigenvector corresponding to $\lambda_0$.  This will play an
+important role in what follows, as it allows us to focus %primarily 
+on the subdominant eigenvalue 
+$\lambda_{\max}(\M) = \max\{\lambda_1(\M), |\lambda_{d-1}(\M)|\}$
+rather than on $\lambda_0(M)$ (which we already know is 1).
 
 Now, notice that
+\begin{equation}
+  \label{eq:3.2}
+|\rho(\psi_1,\M)| \leq \max |\rho(x,\M)| = \lambda_{\max}(\M)
+\end{equation}
+% 20.txt
+where the max is taken over all nonconstant vectors. Since our interest 
+centers on 
+$\lambda_{\max}(\M)$, 
+we would like a $\psi_1$ that makes the left hand side of~(\ref{eq:3.2}) large. 
+This would be achieved if $\psi_1$ were to lie in the space spanned by, say, the
+first two eigenvectors of the Markov chain (sometimes referred 
+to as the \emph{slowest modes} of the process). However, this subspace spans
+only two dimensions of the entire $d$-dimensional space, and it is more likely
+that $\psi_1$ only comes close, at best, to lying in the subspace of
+interest. Now, given $\psi_1$, a judicious choice for the second dimension $\psi_2$, and
+hence $\Psi_2 = [\psi_1\; \psi_2]$, would be that which makes 
+$\max_{a\neq 0} |\rho(\Psi_2 a)|$ as large as possible. To establish that 
+this is indeed the right objective, note the following:
+\begin{align}
+\label{eq:3.3}
+\max_{a\neq 0}|\rho(\Psi_2a,\M)| &= \max_{a\neq 0} \left|\frac{a^t \Psi^t_2 \M \Psi_2 a}{a^t \Psi^t_2 \Psi_2 a}\right|\nonumber\\
+&= \max_{x\in \ran(\Psi_2)}|\rho(x,\M)|\nonumber\\
+&= \max|\rho(x,\M)|\\
+&= \lambdamax\nonumber
+\end{align}
+Again, the max on the right side of~(\ref{eq:3.3}) is over nonconstant vectors. 
+In other words, we wish to chose
+$\Psi_2 = [\psi_1\; \psi_2]$ so that there is a vector $a \in \R^2$ making 
+$|\rho(\Psi_2 a,\M)|$ close to $\lambdamax$.
 
-lP(¢1,M)l 5 max |P(x,M)l = »\max(M) (3-2)
-% 20.txt
-where the max is taken over all non-constant vectors. Since our interest centers on An-.3_x(M), we
-would like a. 1/11 which makes the left hand side of (3.2) large. This would be achieved if 1/)1 were to
-lie in the space spanned by, say, the first two eigenvectors of the Markov chain (sometimes referred
-to as the slowest modes of the process). However, this subspace spans only two dimensions of the
-entire d-dimensional space, and it is more likely that ¢1, at best, only comes close to lying in the
-subspace of interest. Now, given 1/11, a judicious choice for the second dimension l/)2, and hence
-W; E [1/)1 1/:2], would be that which makes max,,¢o |p(\II2a)| is as large as possible. To establish that
+Now, $\rho(\psi_1)$ changes most rapidly in the direction of the gradient
+$\nabla \rho(\psi_1)$.
+\begin{equation}
+  \label{eq:3.4}
+\nabla \rho(\psi_1) = \left(\frac{\partial \rho(\psi_1)}{\partial \psi_1(1)}, \dots,
+\frac{\partial \rho(\psi_1)}{\partial \psi_1(d)}\right)
+ = \frac{2}{\psi_1^t\psi_1} (\M\psi_1 - \rho(\psi_1)\psi_1).
+\end{equation}
+So, to maximize the left hand side of~(\ref{eq:3.3}), $\Psi_2$ should be chosen
+so that the subspace $\ran(\Psi_2)$ contains the gradient vector.
+That is, we must choose $\psi_2$ so that
+\begin{equation}
+\label{eq:3.5}
+\nabla \rho(\psi_1) \in \ran\{\psi_1, \psi_2\} = \ran(\Psi_2).
+\end{equation}
+Clearly, Equation~(\ref{eq:3.4}) implies 
+$\nabla \rho(\psi_1) \in \ran\{\psi_1, \M\psi_1\}$.
+Therefore, if $\ran\{\psi_1, \psi_2\} = \ran\{\psi_1, \M\psi_1\}$, 
+then~(\ref{eq:3.5}) is satisfied.
+%
+%
+% ------------ 21.txt ----------------------------------------------
+%
+%
+In general, having chosen $\Psi_k = [\psi_1 \; \cdots \; \psi_k]$ so that
+\[
+\ran\{\psi_1, \psi_2, \dots, \psi_k\} = \ran\{\psi_1, \M\psi_1, \dots,
+\M\psi_k\},
+\]
+we must chose $\psi_{k+1}$ so that for any nonzero vector $a\in \R^k$,
+\begin{equation}
+\label{eq:3.6}
+\nabla \rho(\Psi_ka) \in \ran\{\psi_1, \psi_2, \dots, \psi_k\}.
+\end{equation}
+Now,
+\[
+\nabla \rho(\Psi_ka)  = 
+ = \frac{2}{a^t\Psi_k^t\Psi_ka} (\M\Psi_k - \rho(\Psi_ka)\Psi_ka).
+\]
+and therefore,
+\[
+\nabla \rho(\Psi_ka)  \in 
+\ran(\M\Psi_k) \cup \ran(\Psi_k) = \ran\{\psi_1, \M\psi_1, \dots, \M^k\psi_1\}.
+\]
+Thus, the requirement~(\ref{eq:3.6}) is satisfied when 
+$\ran\{\psi_1, \psi_2, \dots, \psi_k\} = \ran\{\psi_1, \M\psi_1, \dots, \M^k\psi_1\}$.
 
-this is indeed the pertinent objective, note the following:
-
-_ a‘\1:5Mx1:2a
-I;1;g<Ip(\I'2a,M)l = 13:3: ————a,W§q,2a
-= M
-mrggf,mIp(z. )1
-5 maxlp(z,M)l (3-3)
-= ).,,m(M)
-
-Again, the max on the right side of (3.3) is over non-constant vectors. In words, we wish to chose
-\II-2 = [1/:1 ¢2] so that there is a vector a 6 R2 making |p(\II2a,M)| close to )«max(M).
-
-Now, p(1/11) changes most rapidly in the direction of the gradient Vp(¢1).
-
-W0  (iiiiiiv---viiiléi)
-= ,,;,l (Mam — p<¢.)«m (3.4)
-
-So, to maximize the left hand side of (3.3), ‘I12 should be chosen so that the subspace ra.n(\II2)
-
-contains the gradient vector. That is, we must choose 1/22 so that
-
-VP(¢'1) E F3l1{¢1y¢2} E F-'%!1(‘I’2) (3-5)
-
-Clearly, equation (3.4) implies Vp(1/:1) E ran{z/11,M1,b1}. Therefore, if ran{1/:1,1/)2} = ra.n{¢1,M¢1},
-
-then (3.5) is satisﬁed.
-% 21.txt
-In general, having chosen \II,, E [1/)1 - -- 1/11,] so that
-
-ran{'pl9¢21' " 1¢k} = ran{¢1vM1[)la ' ‘ ' vMk—11/’k}
-
-we must chose 1/1;.“ so that, for any non-zero vector u 6 R",
-
-VP(‘I’I:a) 6 ra‘n{¢11 ‘21 - ° - 31/Jk-F1} 
-Now,
-vp<~I~.a) = gmm — P(‘I’kﬂ)‘I’k“)
-
-and therefore,
-
-Vp(‘II;,a.) E ran(M\II;,) U ra.n(\IIk) = ra.n{¢1, M1/)1, . . . , Mktﬁl}
-
-Thus, requirement (3.6) is met when ran{1/:1, 1/:2, . . . , 1/4+1} = ran{1/)1, M1/:1, . . . , Mk:/11}.
-
-In conclusion, the p-dimensional approximate invariant subspace which is most suitable
-
+In conclusion, the $p$-dimensional approximate invariant subspace that is most suitable
 for our objective is
-
-/C(Mn/mp) 2 ran{«/:1,M¢1,... ,M"“«/21}
-
-This is known as the K rylov subspace. Therefore, in answer to the problem posed at the outset
-of this section, if we take the columns of Q to be an orthonormal basis for IC(M,z/)1, p), then the
-
-eigenvalues of T = Q‘MQ should provide good estimates of p extremal eigenvalues of M.
-
-% 22.txt
+\[
+\sK(\M, \psi_1, p) = \ran\{\psi_1, \M\psi_1, \dots, \M^{p-1}\psi_1\}.
+\]
+This is known as the \emph{Krylov subspace}. Therefore, to answer the problem posed at the outset
+of this section, if we take the columns of $\Q$ to be an orthonormal basis for 
+$\sK(\M,\psi_1, p)$, then the eigenvalues of $\T = \Q^t\M\Q$ should provide good
+estimates of $p$ extremal eigenvalues of $\M$. 
+%
+%
+% --------------- 22.txt ---------------------------------------
+%
+%

### 2013-10-18 tex/InvariantSubspaces.tex
minor mods

@@
 We noted above that Facts~\ref{fact:3.1.1} and~\ref{fact:3.1.2} are
-\emph{theoretically} useful. Speaking practically, however, 
+theoretically useful. Speaking practically now,
 %% This is because 
@@
 %
-primary assumption underlying these facts. %Facts 3.1.1 and 3.1.2. 
+primary assumption underlying the two facts. %Facts 3.1.1 and 3.1.2. 
 Recall the assumption: $\ran(\Psi)$ is an invariant
 subspace for $\M$ where $\Psi = \Pi^{1/2}(\Phi - \<\Phi\>)$. It may well be the
-case that there is a $\psi_j\in \ran(\Psi)$ such that 
-$M\psi_j \notin \ran(\Psi)$, thereby violating the assumption. 
-However, lacking an invariant subspace, we 
-might hope that our observables at least provide an \emph{approximate invariant subspace}.
+case that there exists $\psi_j\in \ran(\Psi)$ such that 
+$\M\psi_j \notin \ran(\Psi)$, thereby violating the assumption. 
+Even if we are lacking an invariant subspace, however, for some applications it
+is reasonable to expect that observables can be chosen to provide at least
+an \emph{approximate invariant subspace}, which is defined as follows:
 \begin{definition}
 %% Definition 3.2.1
-If the columns of $\Psi\in \R^{d\times p}$ are independent and the 
+If the columns of $\Psi\in \R^{d\times p}$ are independent and the norm of the
 \emph{residual matrix} $\E = \M\Psi - \Psi \T$ is small for some  
@@
 \label{thm:3.2.1}  
-Suppose $\M \in R^{d\times d}$ and $T \in \R^{p\times p}$ are symmetric and let
-$E = \M\Q - \Q\T$, where $\Q \in \R^{d\times p}$ is orthonormal 
+Suppose $\M \in R^{d\times d}$ and $\T \in \R^{p\times p}$ are symmetric and let
+$\E = \M\Q - \Q\T$, where $\Q \in \R^{d\times p}$ is orthonormal 
 (i.e., $\Q^t\Q = \I$). Then there exist $\mu_1,\dots, \mu_p \in \lambda(\T)$ and
@@
 The second equality in Theorem~\ref{thm:3.2.2} is a consequence of the polar decomposition, in
-which $\Q = \Psi \Z^{-1} = \Psi(\Psi^t\Psi)^{-1/2}$. Therefore,
+which $\Q = \Psi \Z^{-1} = \Psi(\Psi^t\Psi)^{-1/2}$. Thus,
 $\T = (\Psi^t\Psi)^{-1/2}\Psi^t \M^t \Psi (\Psi^t\Psi)^{-1/2}$
@@
 $\nabla \rho(\psi_1) \in \ran\{\psi_1, \M\psi_1\}$.
-Therefore, if $\ran\{\psi_1, \psi_2\} = \ran\{\psi_1, \M\psi_1\}$, 
+Thus, if $\ran\{\psi_1, \psi_2\} = \ran\{\psi_1, \M\psi_1\}$, 
 then~(\ref{eq:3.5}) is satisfied.
@@
 In general, having chosen $\Psi_k = [\psi_1 \; \cdots \; \psi_k]$ so that
-\[
-\ran\{\psi_1, \psi_2, \dots, \psi_k\} = \ran\{\psi_1, \M\psi_1, \dots,
-\M\psi_k\},
-\]
+$\ran\{\psi_1, \psi_2, \dots, \psi_k\} = \ran\{\psi_1, \M\psi_1, \dots,
+\M\psi_k\}$,
 we must chose $\psi_{k+1}$ so that for any nonzero vector $a\in \R^k$,
@@
 Thus, the requirement~(\ref{eq:3.6}) is satisfied when 
-$\ran\{\psi_1, \psi_2, \dots, \psi_k\} = \ran\{\psi_1, \M\psi_1, \dots, \M^k\psi_1\}$.
+\[
+\ran\{\psi_1, \psi_2, \dots, \psi_k\} = \ran\{\psi_1, \M\psi_1, \dots,
+\M^k\psi_1\}.
+\]
 

### 2013-10-19 tex/InvariantSubspaces.tex
minor mod
work begun on LanczosProcedure chapter

@@
 interest. Now, given $\psi_1$, a judicious choice for the second dimension $\psi_2$, and
-hence $\Psi_2 = [\psi_1\; \psi_2]$, would be that which makes 
+hence $\Psi_2 = [\psi_1, \psi_2]$, would be that which makes 
 $\max_{a\neq 0} |\rho(\Psi_2 a)|$ as large as possible. To establish that 
@@
 In other words, we wish to chose
-$\Psi_2 = [\psi_1\; \psi_2]$ so that there is a vector $a \in \R^2$ making 
+$\Psi_2 = [\psi_1, \psi_2]$ so that there is a vector $a \in \R^2$ making 
 $|\rho(\Psi_2 a,\M)|$ close to $\lambdamax$.
@@
 %
-In general, having chosen $\Psi_k = [\psi_1 \; \cdots \; \psi_k]$ so that
+In general, having chosen $\Psi_k = [\psi_1, \dots,  \psi_k]$ so that
 $\ran\{\psi_1, \psi_2, \dots, \psi_k\} = \ran\{\psi_1, \M\psi_1, \dots,

### 2013-10-19 tex/LanczosProcedures.tex
work begun on LanczosProcedure chapter

@@
 The following will review the usual procedure for generating an orthonormal basis for
-lC(M, «/11, p), and conclude by showing that, for our problem, we can find the matrix Q‘MQ without
-actually carrying out this procedure. Note that this conclusion is essential since the procedure
+$\sK(\M, \psi_1, p)$, 
+and conclude by showing that, for our problem, we can find the matrix 
+$\Q^t\M\Q$
+without actually carrying out this procedure. Note that this conclusion is essential since the procedure
+requires matrix vector multiplication---an operation we have assumed impossible for large enough $d$.
 
-requires matrix vector multiplication - an operation we have assumed impossible for large enough :1.
+\section{The General Lanczos Algorithm}
+% Sec 4.1
+\label{sec:gener-lancz-algor}
+Let $\psip{k} =  \M^k \psi$ and consider the $d\times d$ matrix
+\[
+\Psi = [\psi, \M\psi, \dots, \M^{d-1}\psi] = [\psip{0}, \psip{1}, \dots, \psip{d-1}]
+\]
+Notice that $\M\Psi = [\psip{1}, \psip{2}, \dots, \psip{d-1}, \M^d \psi]$,
+and assuming for the moment that $\Psi$ is nonsingular,
+we can compute the vector $h = -\Psi^{-1}\M^d \psi$. Thus,
+\begin{equation}
+\label{eq:1}
+\M\Psi = \Psi [e_2,\dots,e_d,-h],  
+\end{equation}
+where $e_j$ is the column vector with 1 in
+the $j$th position and zeros elsewhere. We define $H = [e_2,\dots,e_d,-h]$,
+so~(\ref{eq:1}) becomes $\M\Psi = \Psi H$.
+%
+%
+% ----------------- 23.txt ----------------------------------------------
+%
+%
+Equivalently,
+\begin{equation}
+  \label{eq:4.1}
+H = 
+\begin{pmatrix}
+0 & 0      & \cdots & 0 & -h_1 \\
+1 & 0      & \cdots & 0 & -h_2\\
+  & \ddots &        &   & \vdots \\
+  &        & \ddots & 0 & -h_{d-1}\\
+  &        &        & 1 & -h_{d}
+\end{pmatrix} = \Psi^{-1} \M \Psi.
+\end{equation}
+$\H$ is a \emph{companion matrix} which means that its characteristic polynomial
+is $p(x) = x^d + \sum_{i=1}^d h_i x^{i-1}$.
+Since $\H$ is similar to $\M$, finding the eigenvalues of $\M$ is equivalent to
+finding the roots of $p(x)$. However, this is of little practical use since
+finding $h$, constructing $p(x)$, and finding its roots is 
+probably a harder problem than the one we started with. Instead, the value of decomposition (4.1)
+derives from its \emph{upper Hessenberg form}. We exploit this property below.
 
-4.1 The General Lanczos Algorithm
-Let 21;“) E M"¢ and consider the d x :1 matrix
-«I 2 [«/»,M«/»,  ,M"“¢] 2 [«b‘°>,«/»“’,...,¢<‘-‘>1
+Let $\Psi = \Q\R$ be the QR decomposition of $\Psi$. Since $\Psi$ is assumed nonsingular,
+\[
+\Psi_d^{-1}\M \Psi_d = (\R^{-1}\Q^t)\M(\Q\R) = \H
+\]
+Therefore, $\Q^t\M\Q = \R\H\R^{-1}$.  Let $\T \R\H\R^{-1}$.  Then, since $\R$
+and $\R^{-1}$ are both upper triangular and $\H$ is upper Hessenberg $\T = \R\H\R^{-1}$ is also upper
+Hessenberg. Furthermore, since $\M$ is symmetric, it is clear that 
+$\T^t =\Q^t\M\Q = T$. Thus, $\T$ is both upper Hessenberg and symmetric. Therefore,
+$\T$ is tridiagonal and we can write it as follows:
+\begin{equation}
+  \label{eq:4.2}
+T = 
+\begin{pmatrix}
+\alpha_1 & \beta_1 &      0 & \cdots & 0 \\
+\beta_1  & \alpha_2 & \ddots &        & \vdots\\
+         & \ddots & \ddots & \ddots   &\\
+\vdots   &       &  \ddots & \ddots & \beta_{d-1}\\
+0  &   \cdots    &   & \beta_{d-1} &\alpha_d
+\end{pmatrix}.
+\end{equation}
+%
+%
+% ------------ 24.txt ----------------------------------------------
+%
+%
+Equating columns $j$ on both sides of $\M\Q = \Q\T$ yields
+\begin{equation}
+\label{eq:4.3}
+\M q_j = \beta_{j-1}q_{j-1}+\alpha_j q_j + \beta_j q_{j+1}
+\end{equation}
+Since the columns of $\Q$ are orthonormal, multiplying both sides 
+of~(\ref{eq:4.3})
+by $q_j$ and $q_{j+1}$ yields
+$\alpha_j = q_j^t \M q_j$ and $\beta_j = q_{j+1}^t \M q_j$.
+The foregoing justifies what is called the \emph{Lanczos algorithm}, which is performed as follows:
 
-Notice that M\II = [1/:(1),... ,1/;(d'1),Md¢r] and, assuming for the moment that \I1 is non-singular,
-
-we can compute the vector h = —\II‘1M"1/J. Thus,
-
-M\II = \II[e2,. . . ,e,1, —h] E IIIH
-
-% 23.txt
-23
-
-Equivalently,
-0 0  0 -—h1
-1 0  0 —h2
-HE   3 ; =\II“M\II (4.1)
-0 "hl1—1
-1 —h,;
-
-H is a companion matrix which means that its characteristic polynomial is p(:r) = 2:‘ + 23:1 h,-:r"‘1.
-Since H is similar to M, ﬁnding the eigenvalues of M is equivalent to ﬁnding the roots of p(a:).
-However, this is of little practical use since ﬁnding h, constructing 12(1), and ﬁnding its roots is
-probably a harder problem than the one we started with. Instead, the value of decomposition (4.1)
-derives from its upper Hessenberg form. We exploit this property below.
-
-Let ‘II = QR be the QR decomposition of \II. Since ‘I! is assumed nonsingular,
-~II;‘M~r.: = (R"Q‘)M(QR) = H
-Therefore,
-Q‘MQ = RHR-1 2 T
-
-Since R and R“1 are both upper triangular and H is upper Hessenberg, T = R.HR'1 is also upper
-Hessenberg. Furthermore, since M is symmetric, it is clear that T‘ = Q‘MQ = T. That is, T is
-
-an upper Hessenberg matrix that is symmetric. Therefore, T is tridiagonal and we can write it as
-
-follows:
-
-T =    (4.2)
-
-ﬁd—1
-
-0  174-1 ad
-
-% 24.txt
-24
-
-Equating columns j on both sides of MQ = QT yields
-M113‘ = B,-—1qj—1 + 0:91 + ﬁjqj+1 (43)
-
-Since the columns of Q are orthonormal, multiplying both sides of (4.3) by q_,- and q,-+1 yields
-oz, = q§Mq,- and ﬂj = q;-+1Mq,-.
-The foregoing justifies what is called the Lanczos algorithm, performed as follows:
-
-The Lanczos algorithm for (partial) reduction to symmetric tridiaganal form
+\emph{The Lanczos algorithm for (partial) reduction to symmetric tridiaganal form}
 
@@
 
-5:‘ = ||Z||2
+5:^t = ||Z||2
 if 5, = 0, quit
@@
 The qj computed by the Lanczos algorithm are often called the Lanczos vectors. If the
-loop in the algorithm is terminated because ﬁg = 0, this indicates that an exact invariant subspace
+loop in the algorithm is terminated because fig = 0, this indicates that an exact invariant subspace
 has been computed, and is given by ran{q1 - - - qk}. Otherwise, we usually halt the algorithm after
-p steps, in which case the algorithm converg to approximations of at most 12 eigenvalues.‘
+p steps, in which case the algorithm converg to approximations of at most 12 eigenvalues.^t
 
@@
 
-basis for all of R‘, which is clearly an invariant “sub” space. The Lanczos algorithm above, however,
+basis for all of R^t, which is clearly an invariant “sub” space. The Lanczos algorithm above, however,
 
-‘A detailed discussion of such convergence (and misconvergence) is given in
+^tA detailed discussion of such convergence (and misconvergence) is given in
 
@@
 for the approximate invariant subspace IC(M, ¢, p), and a p X p matrix T, = Q'MQ. In that case,
-IIMQ = QTp|[g =  is non-zero and gives the error bound described in Theorem 3.2.1. Now,
+\PsiMQ = QTp|[g =  is non-zero and gives the error bound described in Theorem 3.2.1. Now,
 
@@
 
-ﬁzz-
+fizz-
 Theorem 4.1.1
@@
 the Krylov subspace. For our purposes, this construction is not essential. What is essential is that
-we ﬁnd a way to generate the Lanczos coejﬁcients oz,-, ﬁj, (j = 1, . . . , p), and whence the matrix
+we find a way to generate the Lanczos coejficients oz,-, fij, (j = 1, . . . , p), and whence the matrix
 T, without performing the operation Mqj required by the Lanczos algorithm. We now address this
@@
 
-I/1‘M1/I
+I/1^tM1/I
 ¢’¢
@@
 
-¢‘M«/2 = (¢ — E.¢)‘II%‘MII%(¢ - Em)
+¢^tM«/2 = (¢ — E.¢)^t\Psi%^tM\Psi%(¢ - Em)
 
@@
 
-Recall that the deﬁnition of M implies H%‘MI'I% = HP, and this justiﬁes equation (4.5). Putting
+Recall that the definition of M implies H%^tMI'I% = HP, and this justifies equation (4.5). Putting
 
@@
 _ c, (run), an + 1))
-"‘ ' Ta:,(¢(n))
+"^t ' Ta:,(¢(n))
 
@@
 
-Again recalling the deﬁnition M E 111/ 2P1'I‘1/2, a. simple manipulation veriﬁes
+Again recalling the definition M E 111/ 2P1'I^t1/2, a. simple manipulation verifies
 
-1'li‘M’l'[% = HP’, whereby it follows that
+1'li^tM’l'[% = HP’, whereby it follows that
 
-(«s — E,,¢)‘HP”(¢ — Em)
+(«s — E,,¢)^tHP”(¢ — Em)
 
-"‘M2"‘ (¢ — E,¢)*n<¢ — Em)
+"^tM2"^t (¢ — E,¢)*n<¢ — Em)
 
@@
 _ c,(¢(n).¢<n+2)) c,(¢<n>,¢(n+1)> ’ "’
-"“ jvar,<¢<n)> "(jvar,<¢<n)) ‘)
+"“ jvar,<¢<n)> "(jvar,<¢<n)) ^t)
 
@@
 
-(Mql — oz1q1)/ﬁl. From this we can derive an expression for (12 which, though not as neat as those
+(Mql — oz1q1)/fil. From this we can derive an expression for (12 which, though not as neat as those
 
@@
 
-I12 ‘IiMq2
+I12 ^tIiMq2
 
@@
 1
-= —2(<1iM3q1 - 2015i ‘ all (4-8)
+= —2(<1iM3q1 - 2015i ^t all (4-8)
 
@@
 The last equality follows from equation (4.7), which can be written as q{M2q1 = off’ + (if. The
-preceding paragraphs deﬁne all but one of the quantities in the expression (4.8). The one remaining
+preceding paragraphs define all but one of the quantities in the expression (4.8). The one remaining
 is q§M3q1, which can also be expressed in terms of covariances of <15, and the analysis is almost
@@
 
-(¢ — E,¢)'IIP3(¢ — Em)
+(¢ — E,¢)'\PsiP3(¢ — Em)
 
-"lM3"‘ (¢ — E,¢s)‘H(¢ - Em»)
+"lM3"^t (¢ — E,¢s)^tH(¢ - Em»)
 
@@
 
-To begin with the ﬂfs, recall that ﬁj E ||Mqj — ajq_.,- — ,H_,~_1q,-_;|[2. Whence,
+To begin with the ﬂfs, recall that fij E ||Mqj — ajq_.,- — ,H_,~_1q,-_;|[2. Whence,
 
-Bf (M411 — °‘j‘Ij - ﬁj-1qj—1)’(MlIj - Otjqj - 51-19141)
+Bf (M411 — °^tj^tIj - fij-1qj—1)’(MlIj - Otjqj - 51-19141)
 
-q;-M291‘ - 0? - B,z_1 (4-9)
+q;-M291^t - 0? - B,z_1 (4-9)
 
-Some simple algebra and the deﬁnitions aj E q§Mqj and ﬁJ~_1 E q;-Mq,--1 prove equality (4.9). The
+Some simple algebra and the definitions aj E q§Mqj and fiJ~_1 E q;-Mq,--1 prove equality (4.9). The
 % 29.txt
-recurrence for 0114.1 is equally straight forward, though the ﬁnal form is not as neat:
+recurrence for 0114.1 is equally straight forward, though the final form is not as neat:
 
@@
 1
-§(M¢15 - 11111;‘ - ﬂj—1<Ij—1)'M(Mqj - Otjqj - /3j—1qj—1)
+§(M¢15 - 11111;^t - ﬂj—1<Ij—1)'M(Mqj - Otjqj - /3j—1qj—1)
 J
@@
 1
-a,~+1 = E(q§»M3q_,,- — (1? — 2ozJ~ﬁJ2 — 2ﬂj_1q;»_1M2qj + ﬁ2_1) (4.11)
+a,~+1 = E(q§»M3q_,,- — (1? — 2ozJ~fiJ2 — 2ﬂj_1q;»_1M2qj + fi2_1) (4.11)
 J
@@
 ¢I;'+1Mk9j = ij(Mqj - Otjqj - ﬂj—1t1j—1)'M"¢1j
-= i(<1§M"“¢Ij - C‘j‘I_;'Mk‘Ij - ﬂj—1q§M"qj-1)
+= i(<1§M"“¢Ij - C^tj^tI_;'Mk^tIj - ﬂj—1q§M"qj-1)
 The second is,
-‘I_;'+1Mkqj+1 = ﬂi?(M<Ij - Otjqj - .3j—1<Ij—1)‘Mk(Mqj — Orjqj - ﬂj—1¢1j—1)
+^tI_;'+1Mkqj+1 = ﬂi?(M<Ij - Otjqj - .3j—1<Ij—1)^tMk(Mqj — Orjqj - ﬂj—1¢1j—1)
 
 1
-ﬁ(‘1§M'°"<1:‘ + °‘:2"I§M'°<J:' + ﬁ?—z<Ii—1M"<I:‘—1
+fi(^t1§M'°"<1:^t + °^t:2"I§M'°<J:' + fi?—z<Ii—1M"<I:^t—1
 J
-—2a,-q;-M*+‘q,~ + 2am,--1q;-M*q,--1 — zza,--1q;-M*+‘q]--1) (4.12)
+—2a,-q;-M*+^tq,~ + 2am,--1q;-M*q,--1 — zza,--1q;-M*+^tq]--1) (4.12)
 
@@
 are based on four recurrence relations, they become quite complicated as 1' gets large. However,
-deﬁning these recursions in a. computer program is trivial, and we use this fact in Section 5.4 when
+defining these recursions in a. computer program is trivial, and we use this fact in Section 5.4 when
 

### 2013-10-19 tex/LanczosProcedures.tex
finished LanczosProcedures chapter

@@
   \label{eq:4.2}
-T = 
+\T = 
 \begin{pmatrix}
@@
 The foregoing justifies what is called the \emph{Lanczos algorithm}, which is performed as follows:
+         \begin{algorithm}
+           \caption{The Lanczos algorithm for (partial) reduction to symmetric tridiaganal form}
+           \begin{algorithmic}
+             \STATE $q_1 = \psi/\|\psi\|_2$, $\beta_0 = 0$, $q_0= 0$
+             \FOR $j = 1 \text{ to } p$
+             \STATE $z \leftarrow \M q_j$
+             \STATE $\alpha_j \leftarrow q_j^tz$
+             \STATE $z \leftarrow z - \alpha_j q_j - \beta_{j-1} q_{j-1}$
+             \STATE $\beta_j \leftarrow \|z\|_2$
+             \IF $\beta_j=0$ 
+             \STATE return
+             \ENDIF
+             \STATE $q_{j+1} = z/\beta_j$
+             \ENDFOR
+           \end{algorithmic}
+         \end{algorithm}
+The $q_j$ computed by the Lanczos algorithm are often called the \emph{Lanczos vectors}. If the
+loop in the algorithm is terminated because $\beta_{k}= 0$, this indicates that
+an exact invariant subspace has been computed, and is given by 
+$\ran\{q_1, \dots, q_k\}$. Otherwise, we usually halt the algorithm after 
+$p$ steps, in which case the algorithm converges to approximations of at most
+  $p$ eigenvalues.\footnote{A Detailed discussion of such convergence (and
+    misconvergence) is given in~\cite{Demmel:1997}.}
 
-\emph{The Lanczos algorithm for (partial) reduction to symmetric tridiaganal form}
+Note that Equation~(\ref{eq:4.3}) results from the equation $\M\Q = \Q\T$. The latter equation holds
+since, starting with $d$ vectors in our subspace, $\Q$ is a $d \times d$ matrix
+whose columns form an orthonormal basis for all of $\R^d$, which is clearly an
+invariant subspace. The Lanczos algorithm above, however, 
+%
+%
+% ------------ 25.txt ---------------------------------------------------------
+%
+%
+proceeds for only $p$ steps, producing a $d \times p$ matrix $\Q$, whose columns
+form an orthonormal basis for the \emph{approximate} invariant subspace 
+$\sK(\M, \psi, p)$, and a $p \times p$ matrix $\T, = \Q^t\M\Q$. In that case,
+$\|\M\Q - \Q\T_p\|_2 = \|\E\|_2$ is nonzero and gives the error bound described
+in Theorem~\ref{thm:3.2.1}. Now, writing the full $d \times d$ matrix $\T$
+of~(\ref{eq:4.2}) as
+\begin{equation}
+  \label{eq:4.4}
+\T = 
+\left( \begin{array}{c|c}
+\T_p & \T_{pu}^t \\
+ \hline
+\T_{pu} & \T_{u}
+\end{array}
+\right)
+=
+\left( \begin{array}{cccc|cccc}
+\alpha_1 & \beta_1 &      &                 & & && \\
+\beta_1 & \ddots  & \ddots&                 & & && \\
+        & \ddots  & \ddots  & \beta_{p-1}    &          & && \\
+        &         & \beta_{p-1}  & \alpha_{p} & \beta_{p}& && \\
+\hline
+        &         &             & \beta_{p}  & \alpha_{p+1} & \beta_{p+1} && \\
+        &         &             &           & \beta_{p+1}  & \ddots & \ddots &\\
+        &         &             &           &             & \ddots & \ddots &\beta_{d-1} \\
+        &         &             &           &             & \ddots & \beta_{d-1} &\alpha_{d}
+\end{array}
+\right)
+\end{equation}
+allows us to describe the error bound $\|\E\|_2$ in terms of the submatrix
+$\T_{pu}$, and hence in terms of $\beta_p$.
+\begin{theorem}
+%Theorem 4.1.1
+\label{thm:4.1.1}
+If $\T_p$ and $\T_{pu}$ are the matrices appearing in~(\ref{eq:4.4}), and if the $p$
+columns of $\Q$ are computed by the 
+Lanczos algorithm, then there exist $\mu_1, \dots, \mu_p \in \lambda(\T_p)$ and 
+$\lambda_1, \dots, \lambda_p \in \lambda(\M)$
+such that
+\[
+|\mu_k - \lambda_k| \leq \|\T_{pu}\|_2 = \beta_p, \quad \text{ for $k=1,\dots,p$.}
+\]
+\end{theorem}
+For a proof, see\cite[Page 365]{Demmel:1997}.
+%
+%
+% --------------- 26.txt -------------------------
+%
+%
+\section{A Lanczos Procedure for Markov Chains}
+Notice that the Lanczos algorithm requires the computation of $\M q_j$. Throughout this
+paper we have assumed that $\M \in \R^{d\times d}$, and that $d$ is so large as
+to make the matrix vector multiplication $\M q_j$ impossible. Now, the Lanczos
+algorithm constructs an orthonormal basis for the Krylov subspace. For our
+purposes, this construction is not essential. What is essential is that we find
+a way to generate the Lanczos coeficients $\alpha_i$, $\beta_j$, $(j = 1,\dots, p)$, and
+whence the matrix $\T$, without performing the operation $\M q_j$ required by the
+Lanczos algorithm. We now address this problem.
 
-91 = ¢/||¢||2, [30 = 0, 40 = 0
-
-for j=1top
-
-z=Mq_,-
-
-5:^t = ||Z||2
-if 5, = 0, quit
-¢Ij+1 = 2/51
-
-end for
-
-The qj computed by the Lanczos algorithm are often called the Lanczos vectors. If the
-loop in the algorithm is terminated because fig = 0, this indicates that an exact invariant subspace
-has been computed, and is given by ran{q1 - - - qk}. Otherwise, we usually halt the algorithm after
-p steps, in which case the algorithm converg to approximations of at most 12 eigenvalues.^t
-
-Note that equation (4.3) results from the equation MQ = QT. The latter equation holds
-since, starting with d vectors in our subspace, Q is a d Xd matrix whose columns form an orthonormal
-
-basis for all of R^t, which is clearly an invariant “sub” space. The Lanczos algorithm above, however,
-
-^tA detailed discussion of such convergence (and misconvergence) is given in
-
-% 25.txt
-25
-
-proceeds for only p steps, producing a d x p matrix Q, whose columns form an orthonormal basis
-for the approximate invariant subspace IC(M, ¢, p), and a p X p matrix T, = Q'MQ. In that case,
-\PsiMQ = QTp|[g =  is non-zero and gives the error bound described in Theorem 3.2.1. Now,
-
-writing the full d x :1 matrix T of (4.2) as
-
-01 51
-/71
-
-(4.4)
-
-allows us to describe the error bound ||E||-2 in terms of the submatrix Tm, and, hence, in terms of
-
-fizz-
-Theorem 4.1.1
-
-HT, and T,“ are the matrices appearing in (4.4), and if the p columns of Q are computed by the
-
-Lanczos algorithm, then there exist pl, . . . ,;t,, E /\(T,) and A1, . . . , A, E A(M) such that
-
-ill» - /\I=i S iiTP'4”2 = 512
-
-fork=1,...,p.
-
-For a proof, see page 365 of [1].
-
-% 26.txt
-26
-
-4.2 A Lanczos Procedure for Markov Chains
-
-Notice that the Lanczos algorithm requires the computation of Mqj. Throughout this
-paper we have assumed that M 6 Rd“, and that d is so large as to make the matrix vector
-multiplication Mqj impossible. Now, the Lanczos algorithm constructs an orthonormal basis for
-the Krylov subspace. For our purposes, this construction is not essential. What is essential is that
-we find a way to generate the Lanczos coejficients oz,-, fij, (j = 1, . . . , p), and whence the matrix
-T, without performing the operation Mqj required by the Lanczos algorithm. We now address this
-problem.
-
-Begin with the centered and weighted observable 1/1 = 111/ 2 (qS—E,,¢), and let q1 = 11;/||1/JH2.
-To simplify notation, let ¢(n) E ¢(X,.). The form of the first Lanczos coeﬂicient a1 is straight
-
+Begin with the centered and weighted observable 
+$\psi = \Pi^{1/2} (\phi - \bE_\pi\phi)$,
+and let $q_1 = \psi/\|\psi\|_2$.
+To simplify notation, let $\phi(n) = \phi(X_n)$. 
+The form of the first Lanczos coeﬂicient $\alpha_1$ is straight
 forward:
-
-01 = qiMq1
-
-I/1^tM1/I
-¢’¢
-
-and now the special properties of the Markov chain setting play an important role. For, regard the
-
-following
-
-¢^tM«/2 = (¢ — E.¢)^t\Psi%^tM\Psi%(¢ - Em)
-
-Cw (d>(n), ¢(" + 1)) (4-5)
-
+\[
+\alpha_1 = q_1^t \M q_1  = \frac{\psi^t\M\psi}{\psi^t\psi}.
+\]
+and now the special properties of the Markov chain setting play an important
+role. Indeed, we have
+\begin{align}
+\label{eq:4.5}
+\psi^t\M\psi &= (\phi - \bE_\pi\phi)^t\Pi^{\frac{1}{2}t}\M\Pi^{\frac{1}{2}}(\phi- \bE_\pi \phi)\nonumber\\
+&= \bC_\pi (\phi(n), \phi(n + 1)).
+\end{equation}
 and
-
-W1!’ = (<15 - E«¢)'H(¢ - E«¢)
-
-Varn (¢(")) (4-6)
-
-Recall that the definition of M implies H%^tMI'I% = HP, and this justifies equation (4.5). Putting
-
-% 27.txt
-27
-
-these equations together, we have a. nice form for :11:
-
-_ c, (run), an + 1))
-"^t ' Ta:,(¢(n))
-
-The next Lanczos coeﬂicient, B1, is only slightly more complicated.
-
-[31 = ||M91-0t1tI1||2
-
-[(Mq1 — 0l1¢11)i(M¢]1 - 0101)] 1/2
-
-[t1iM2¢I1 — ail 1/2
-
-(4.7)
-
-Again recalling the definition M E 111/ 2P1'I^t1/2, a. simple manipulation verifies
-
-1'li^tM’l'[% = HP’, whereby it follows that
-
-(«s — E,,¢)^tHP”(¢ — Em)
-
-"^tM2"^t (¢ — E,¢)*n<¢ — Em)
-
-C: (¢("), ¢(n + 2))
-V3-Ur
-
-Inserting the last expression into equation (4.7) gives
-
-_ c,(¢(n).¢<n+2)) c,(¢<n>,¢(n+1)> ’ "’
-"“ jvar,<¢<n)> "(jvar,<¢<n)) ^t)
-
-Notice that we can estimate these expressions of oz; and ﬂl by running simulations of the Markov
-chain. To do so, we run the Markov chain starting from a random initial state, compute the value of
+\begin{align}
+\label{eq:4.6}
+\psi^t\psi &= (\phi - \bE_\pi\phi)^t\Pi(\phi- \bE_\pi \phi)\nonumber\\
+&= \Var_\pi(\phi(n)).
+\end{equation}
+Recall that the definition of $\M$ implies 
+$\Pi^{\frac{1}{2}t}\M\Pi^{\frac{1}{2}}= \Pi \P$, and this justifies 
+Equation~(\ref{eq:4.5}). Putting
+%
+%
+% -------------- 27.txt ------------------------------
+%
+%
+these equations together, we have a nice form for $\alpha_1$:
+\[
+\alpha_1= \frac{\bC_\pi (\phi(n), \phi(n + 1))}{\Var_\pi(\phi(n))}.
+\]
+The next Lanczos coefficient, $\beta_1$, is only slightly more complicated.
+\begin{align}
+\label{eq:4.7}
+\beta_1 &= \|\M q_1 - \alpha_1 q_1\|_2 \nonumber\\
+&= \bigl[(\M q_1 - \alpha_1 q_1)^t(\M q_1 - \alpha_1 q_1)\bigr]^{1/2}\nonumber\\
+&= \bigl[q_1^t\M^2 q_1 - \alpha^2_1\bigr]^{1/2}.
+\end{align}
+Again recalling the definition $\M =\Pi^{1/2}\P\Pi^{1/2}, a simple manipulation verifies
+$\Pi^{\frac{1}{2}t}\M^s\Pi^{\frac{1}{2}} = \Pi \P^s$,
+whereby it follows that
+\begin{align*}
+q_1^t\M^2 q_1  &= \frac{(\phi - \bE_\pi\phi)^t\Pi \P^2(\phi- \bE_\pi \phi)}
+{(\phi - \bE_\pi\phi)^t\Pi (\phi- \bE_\pi \phi)} \\
+&= \frac{\bC_\pi (\phi(n), \phi(n + 2))}{\Var_\pi(\phi(n))}.
+\end{align*}
+Inserting the last expression into Equation~(\ref{eq:4.7}) gives
+\[
+\beta_1 = \left[\frac{\bC_\pi (\phi(n), \phi(n + 2))}{\Var_\pi(\phi(n))}
+-\bigl( \frac{\bC_\pi (\phi(n), \phi(n + 1))}{\Var_\pi(\phi(n))}\bigr)^2
+\right]^{1/2}.
+\]
+Notice that we can estimate these expressions of $\alpha_1$ and $\beta_1$ by
+running simulations of the Markov chain. 
+To do so, we run the Markov chain starting from a random initial state, compute the value of
 the observable at each step of the chain and, from the data, compute estimates of the covariances
@@
 
-Before generalizing this for the jth Lanczos coefficient, we consider one more coeﬂicient
-in detail. Recall that the second basis element q; in the Lanczos algorithm is given by 1;; =
+Before generalizing this for the $j$th Lanczos coefficient, we consider one more coeﬂicient
+in detail. Recall that the second basis element $q_2$ in the Lanczos algorithm is given by 
+$q_2 = (\M q_1 — \alpha_1 q_1)/\beta_1$. 
+From this we can derive an expression for $\alpha_2$ which, while not as neat as those
+%
+%
+% ----------------- 28.txt ----------------------------------------------------------
+%
+%
+for $\alpha_1$ and $\beta_1$, is nonetheless tractable.
+\begin{align}
+\label{eq:4.8}
+\alpha_2 &= q_2^t \M q_2 \nonumber\\
+&= \frac{1}{\beta_1^2(\M q_1 - \alpha_1 q_1)^t\M (\M q_1 - \alpha_1 q_1)\nonumber\\
+&= \frac{1}{\beta_1^2(q_1^t\M^3 q_1 - 2\alpha_1 q_1^t\M^2 q_1 + \alpha_1^2q_1^t\M q_1 )\nonumber\\
+&= \frac{1}{\beta_1^2(q_1^t\M^3 q_1 - 2\alpha_1 \beta_1^2 + \alpha_1^3 )
+&= \bigl[q_1^t\M^2 q_1 - \alpha^2_1\bigr]^{1/2}.
+\end{align}
+Equation~(\ref{eq:4.8}) follows from Equation~(\ref{eq:4.7}), which can be
+written as $q_1^t\M^2 q_1 = \alpha_1^2 + \beta_1^2$.  
+The preceding paragraphs define all but one of the quantities in the last
+exression of~(\ref{eq:4.8}). The one remaining
+is $q_1^t\M^3 q_1$, which can also be expressed in terms of covariances of $\phi$,
+and the analysis is almost identical to that for $q_1^t\M^2q_1$:
+\begin{align*}
+q_1^t\M^3 q_1  &= \frac{(\phi - \bE_\pi\phi)^t\Pi \P^3(\phi- \bE_\pi \phi)}
+{(\phi - \bE_\pi\phi)^t\Pi (\phi- \bE_\pi \phi)} \\
+&= \frac{\bC_\pi (\phi(n), \phi(n + 3))}{\Var_\pi(\phi(n))}.
+\end{align*}
+To generalize the foregoing for higher order Lanczos coeﬂicients requires a recurrence
+relation between $(\alpha_{j+1}, \beta_j)$ and $(\alpha_{j}, \beta_{j-1})$. 
+As we will see, such a relation involves higher order
+Lanczos vectors $q_j$---vectors we do not wish to compute. Recall that the only ``vector'' we can
+handle is our initial observable. Therefore, we must also establish a recurrence
+among the $q_j$'s, thus making them available inductively from our observable vector.
 
-(Mql — oz1q1)/fil. From this we can derive an expression for (12 which, though not as neat as those
+To begin with the $\beta_j$'s, recall that 
+$\beta_j = \|\M q_j - \alpha_jq_j - \beta_{j-1}q_{j-1}\|_2$. 
+Whence,
+\begin{align}
+\label{eq:4.9}
+\beta_j^2 &= (\M q_j - \alpha_jq_j - \beta_{j-1}q_{j-1})^t(\M q_j - \alpha_jq_j - \beta_{j-1}q_{j-1})\nonumber\\
+&= q_j^t\M^2 q_j - \alpha_j^2 - \beta^2_{j-1}.
+\end{align}
+Some simple algebra and the definitions $\alpha_j = q_j^t\M q_j$ and
+$\beta_{j-1} = q_j^t \M q_{j-1}$ prove equality~(\ref{eq:4.9}).  The
+%
+%
+% -------------- 29.txt ------------------------------------------------------
+%
+%
+recurrence for $\alpha_{j+1}$ is equally straight forward, though the final form
+is not as neat as the expression above. Indeed,
+\begin{align}
+\label{eq:4.10}
+\alpha_{j+1} &= q_{j+1}^t \M q_{j+1} \nonumber\\
+&= \frac{1}{\beta_j^2}(\M q_j - \alpha_j q_j - \beta_{j-1} q_{j-1})^t\M 
+  (\M q_j - \alpha_j q_j - \beta_{j-1} q_{j-1})\nonumber\\
+&= \frac{1}{\beta_j^2}(q_j^t\M^3 q_j 
+    - 2\alpha_j q_j^t\M^2 q_j 
+    - 2\beta^2_{j-1} q_{j-1}^t\M^2 q_j 
+    + \alpha_1^3
+    + 2\alpha_j \beta^2_{j-1} + \beta^2_{j-1}).
+\end{align}
+From~(\ref{eq:4.9}), $\alpha_j^2 +\beta_j^2 + \beta^2_{j-1}$,
+and upon plugging this into~(\ref{eq:4.10}), we arrive at 
+\begin{equation}
+\label{eq:4.11}
+\alpha_{j+1}= \frac{1}{\beta_j^2}(q_j^t\M^3 q_j 
+    - \alpha_j^3
+    - 2\alpha_j\beta^2_{j}
+    - 2\beta_{j-1}q_{j-1}^t\M^2q_j + \beta^2_{j-1}).
+\end{equation}
 
-% 28.txt
-28
-
-for oz; and [31, is nonetheless tractable.
-
-I12 ^tIiMq2
-
-1
-[7(M91 — 0t1¢I1)tM(M¢I1 - 11191)
-1
-
-1
-= f,§(qiM3q1 - 2a1qlM2<11 + a’9iM¢11)
-1
-1
-= —2(<1iM3q1 - 2015i ^t all (4-8)
-
-1.
-The last equality follows from equation (4.7), which can be written as q{M2q1 = off’ + (if. The
-preceding paragraphs define all but one of the quantities in the expression (4.8). The one remaining
-is q§M3q1, which can also be expressed in terms of covariances of <15, and the analysis is almost
-
-identical to that for q{M2q1:
-
-(¢ — E,¢)'\PsiP3(¢ — Em)
-
-"lM3"^t (¢ — E,¢s)^tH(¢ - Em»)
-
-Cu (¢(")a ¢(n + 3))
-Varn(¢(n))
-
-To generalize the foregoing for higher order Lanczos coeﬂicients requires a recurrence
-relation between (aj+1, ﬂj) and (aj, ﬂ,~_1). As we will see, such a relation involves higher order
-Lanczos vectors q,~ — vectors we do not wish to compute. Recall that the only “vector” we can
-handle is our initial observable. Therefore, we must also establish a recurrence among the q,-’s, thus
-making them available inductively from our observable vector.
-
-To begin with the ﬂfs, recall that fij E ||Mqj — ajq_.,- — ,H_,~_1q,-_;|[2. Whence,
-
-Bf (M411 — °^tj^tIj - fij-1qj—1)’(MlIj - Otjqj - 51-19141)
-
-q;-M291^t - 0? - B,z_1 (4-9)
-
-Some simple algebra and the definitions aj E q§Mqj and fiJ~_1 E q;-Mq,--1 prove equality (4.9). The
-% 29.txt
-recurrence for 0114.1 is equally straight forward, though the final form is not as neat:
-
-2
-0tj+1 ¢Ij+1M'1j+1
-
-1
-§(M¢15 - 11111;^t - ﬂj—1<Ij—1)'M(Mqj - Otjqj - /3j—1qj—1)
-J
-
-1
-§(q;-M-“*4; — 2a,~q;M*q,- — 2x3;+’_1q;--1M*q,- + «I? + 2a,-5,51 + am (4.10)
-J
-From (4.9), q;~M2q,~ = oz? + ﬂf + ﬂ}_1 and, upon plugging this into (4.10), we arrive at
-1
-a,~+1 = E(q§»M3q_,,- — (1? — 2ozJ~fiJ2 — 2ﬂj_1q;»_1M2qj + fi2_1) (4.11)
-J
-
-The expressions above involve the higher order Lanczos vectors qj. In the present context,
-these vectors will not be readily available. However, another recursion — this time relating forms
-involving q,-+1 to forms involving qj and qj__1 — will make these expressions accessible by induction
-on our initial observable. The general forms of interest are q;+1M"qj+1 and qj-+1M"qJ~. Again, we
-
+The expressions above involve the higher order Lanczos vectors $q_j$. In the present context,
+these vectors will not be readily available. However, another recursion---this time relating forms
+involving $q_{j+1}$ to forms involving $q_j$ and $q_{j-1} $---will make these
+expressions accessible by induction on our initial observable. The general forms
+of interest are $q_{j+1}^t\M^k q_{j+1}$ and $q^t_{j+1}\M^k q_j$. Again, we 
 begin with the simpler of the two:
-
-¢I;'+1Mk9j = ij(Mqj - Otjqj - ﬂj—1t1j—1)'M"¢1j
-= i(<1§M"“¢Ij - C^tj^tI_;'Mk^tIj - ﬂj—1q§M"qj-1)
+\begin{align*}
+  q_{j+1}^tM^kq_j &= 
+ \frac{1}{\beta_j}(\M q_j - \alpha_j q_j - \beta_{j-1} q_{j-1})^t\M^k q_j\\
+ \frac{1}{\beta_j}(q_j^t\M^{k+1} q_j - \alpha_j q_j^t\M^{k} q_j -\beta_{j-1} q_j^t\M^k).
+\end{align*}
 The second is,
-^tI_;'+1Mkqj+1 = ﬂi?(M<Ij - Otjqj - .3j—1<Ij—1)^tMk(Mqj — Orjqj - ﬂj—1¢1j—1)
-
-1
-fi(^t1§M'°"<1:^t + °^t:2"I§M'°<J:' + fi?—z<Ii—1M"<I:^t—1
-J
-—2a,-q;-M*+^tq,~ + 2am,--1q;-M*q,--1 — zza,--1q;-M*+^tq]--1) (4.12)
-
-The point of all this is that each pair of coeﬂicients (oz,-+1, /3,-) can be expressed as a
-function of the covariance between ¢(n) and ¢(n+ m), for m = 0, . . . , 21' + 1. Since the expressions
-are based on four recurrence relations, they become quite complicated as 1' gets large. However,
-defining these recursions in a. computer program is trivial, and we use this fact in Section 5.4 when
-
-applying the foregoing to a particular problem.
-
-% 30.txt
-30
+\begin{align}
+\label{eq:4.12}
+  q_{j+1}^tM^kq_{j+1} 
+&=  \frac{1}{\beta_j^2}(\M q_j - \alpha_j q_j - \beta_{j-1} q_{j-1})^t\M^k 
+  (\M q_j - \alpha_j q_j - \beta_{j-1} q_{j-1})\nonumber\\
+&=  \frac{1}{\beta_j^2}(q_j^t\M^{k+2} q_j + \alpha^2_jq_j^t\M^{k} q_j 
+  + \beta^2_{j-1} q_{j-1}^t\M^k q_{j-1} 
+    - 2\alpha_jq_j^t\M^{k+1} q_j
+    + 2\alpha_j \beta_{j-1}q_j^t\M^k q_{j-1}
+    - 2\beta_{j-1}q_j^t \M^{k+1}q_{j-1}).
+\end{align}
+The point of all this is that each pair of coeﬂicients $(\alpha_{j+1}, \beta_j)$
+can be expressed as a function of the covariance between 
+$\phi(n)$ and $\phi(n+ m)$, for $m = 0, \dots, 2j+1$. Since the expressions
+are based on four recurrence relations, they become quite complicated as $j$ gets large. However,
+defining these recursions in a computer program is trivial, and we use this fact
+in Section~\label{sec:computations} % Sec 5.4
+when applying the foregoing to a particular problem.
+%
+%
+% ----------------- 30.txt ------------------------------------------------------
+%
+%

### 2013-10-19 tex/LanczosProcedures.tex
fixed bugs: Lanzos Procedures section now compiles

@@
 \begin{equation}
-\label{eq:1}
+\label{eq:10000}
 \M\Psi = \Psi [e_2,\dots,e_d,-h],  
@@
 the $j$th position and zeros elsewhere. We define $H = [e_2,\dots,e_d,-h]$,
-so~(\ref{eq:1}) becomes $\M\Psi = \Psi H$.
+so~(\ref{eq:10000}) becomes $\M\Psi = \Psi H$.
 %
@@
 
-Let $\Psi = \Q\R$ be the QR decomposition of $\Psi$. Since $\Psi$ is assumed nonsingular,
+Let $\Psi = \Q\mR$ be the QR decomposition of $\Psi$. Since $\Psi$ is assumed nonsingular,
 \[
-\Psi_d^{-1}\M \Psi_d = (\R^{-1}\Q^t)\M(\Q\R) = \H
+\Psi_d^{-1}\M \Psi_d = (\mR^{-1}\Q^t)\M(\Q\mR) = \H
 \]
-Therefore, $\Q^t\M\Q = \R\H\R^{-1}$.  Let $\T \R\H\R^{-1}$.  Then, since $\R$
-and $\R^{-1}$ are both upper triangular and $\H$ is upper Hessenberg $\T = \R\H\R^{-1}$ is also upper
+Therefore, $\Q^t\M\Q = \mR\H\mR^{-1}$.  Let $\T \mR\H\mR^{-1}$.  Then, since $\mR$
+and $\mR^{-1}$ are both upper triangular and $\H$ is upper Hessenberg $\T = \mR\H\mR^{-1}$ is also upper
 Hessenberg. Furthermore, since $\M$ is symmetric, it is clear that 
@@
          \begin{algorithm}
-           \caption{The Lanczos algorithm for (partial) reduction to symmetric tridiaganal form}
+           \caption{Lanczos algorithm for partial reduction to symmetric
+             tridiaganal form}
            \begin{algorithmic}
              \STATE $q_1 = \psi/\|\psi\|_2$, $\beta_0 = 0$, $q_0= 0$
-             \FOR $j = 1 \text{ to } p$
+             \FOR{$j = 1, \dots, p$}
              \STATE $z \leftarrow \M q_j$
@@
              \STATE $\beta_j \leftarrow \|z\|_2$
-             \IF $\beta_j=0$ 
+             \IF{$\beta_j=0$}
              \STATE return
@@
         &         &             &           &             & \ddots & \ddots &\beta_{d-1} \\
-        &         &             &           &             & \ddots & \beta_{d-1} &\alpha_{d}
+        &         &             &           &             &        & \beta_{d-1} &\alpha_{d}
 \end{array}
@@
 \end{theorem}
-For a proof, see\cite[Page 365]{Demmel:1997}.
+For a proof, see~\cite[Page 365]{Demmel:1997}.
 %
@@
 &= \bC_\pi (\phi(n), \phi(n + 1)).
-\end{equation}
+\end{align}
 and
@@
 &= \Var_\pi(\phi(n)).
-\end{equation}
+\end{align}
 Recall that the definition of $\M$ implies 
@@
 \end{align}
-Again recalling the definition $\M =\Pi^{1/2}\P\Pi^{1/2}, a simple manipulation verifies
+Again recalling the definition $\M =\Pi^{1/2}\P\Pi^{1/2}$, a simple manipulation verifies
 $\Pi^{\frac{1}{2}t}\M^s\Pi^{\frac{1}{2}} = \Pi \P^s$,
@@
 \alpha_2 &= q_2^t \M q_2 \nonumber\\
-&= \frac{1}{\beta_1^2(\M q_1 - \alpha_1 q_1)^t\M (\M q_1 - \alpha_1 q_1)\nonumber\\
-&= \frac{1}{\beta_1^2(q_1^t\M^3 q_1 - 2\alpha_1 q_1^t\M^2 q_1 + \alpha_1^2q_1^t\M q_1 )\nonumber\\
-&= \frac{1}{\beta_1^2(q_1^t\M^3 q_1 - 2\alpha_1 \beta_1^2 + \alpha_1^3 )
+&= \frac{1}{\beta_1^2}(\M q_1 - \alpha_1 q_1)^t\M (\M q_1 - \alpha_1 q_1)\nonumber\\
+&= \frac{1}{\beta_1^2}(q_1^t\M^3 q_1 - 2\alpha_1 q_1^t\M^2 q_1 + \alpha_1^2q_1^t\M q_1 )\nonumber\\
+&= \frac{1}{\beta_1^2}(q_1^t\M^3 q_1 - 2\alpha_1 \beta_1^2 + \alpha_1^3 )
 &= \bigl[q_1^t\M^2 q_1 - \alpha^2_1\bigr]^{1/2}.

### 2013-10-19 tex/MarkovChains.tex
minor formatting mods

@@
 \label{eq:1}
-P(X_{n+1} \in B \mid \sigma(X_0, \dots, X_{n}))
-=P(X_{n+1} \in B \mid \sigma(X_{n})).
+\bP(X_{n+1} \in B \mid \sigma(X_0, \dots, X_{n}))
+=\bP(X_{n+1} \in B \mid \sigma(X_{n})).
 \end{equation}
@@
 \label{eq:2.2}
-P(X_{n+1} \in B \mid  \sigma(X_n)) = p_n(X_n, B)
+\bP(X_{n+1} \in B \mid  \sigma(X_n)) = p_n(X_n, B)
 \end{equation}
@@
 \label{eq:2.3}
-P(X_j\in B_j ,0 \leq j \leq n) = \int_{B_0} \mu(dx_0) \int_{B_1} p_0(x_0, dx_1) 
+\bP(X_j\in B_j ,0 \leq j \leq n) = \int_{B_0} \mu(dx_0) \int_{B_1} p_0(x_0, dx_1) 
 \int_{B_2} p_1(x_1, dx_2) \cdots 
@@
 The Markov chain that we will study on this space is simply $X_n(\omega) = \omega$, the coordinate maps.
-Then, by the Kolmogorov extension theorem, there exists a unique probability measure $P_\mu$ on
+Then, by the Kolmogorov extension theorem, there exists a unique probability measure $\bP_\mu$ on
 $(\Omega, \F)$ so that the $X_n(\omega)$ have finite dimensional distributions~(\ref{eq:2.3}).
@@
 If instead of $\mu$, we begin with the initial distribution $\delta_x$, i.e., point mass at $x$, then we
-denote the probability measure by $P_x$. With such measures defined for each $x$, we can in turn
-define distributions $P_\mu$, given any initial distribution $\mu$, by
+denote the probability measure by $\bP_x$. With such measures defined for each $x$, we can in turn
+define distributions $\bP_\mu$, given any initial distribution $\mu$, by
 \[
-P_\mu(A) = \int \mu(dx)P_x(A).
+\bP_\mu(A) = \int \mu(dx)\bP_x(A).
 \]
@@
 \[
-P(X_{n+1}=j \mid X_n = i) =p(i,j),
+\bP(X_{n+1}=j \mid X_n = i) =p(i,j),
 \]
@@
 
-Equation~(\ref{eq:2.4}) says $P_\pi(X_1 = y) = \pi(y)$, 
-and by induction that $P_\pi(X_n = y) = \pi(y)$ for all $n \geq 1$. 
+Equation~(\ref{eq:2.4}) says $\bP_\pi(X_1 = y) = \pi(y)$, 
+and by induction that $\bP_\pi(X_n = y) = \pi(y)$ for all $n \geq 1$. 
 If $\pi$ is a probability measure, then we call $\pi$ a stationary distribution. It
@@
 \[
-\E_\pi\phi_i = \sum_{x\in S} \pi(x)\phi_i(x)
+\bE_\pi\phi_i = \sum_{x\in S} \pi(x)\phi_i(x)
 \]
@@
 % Eqn 2.5
-\C_\pi (\phi_i(X_n),  \phi_i (X_{n+s})) 
-&= \E_\pi[(\phi_i(X_n) - \E_\pi\phi_i)(\phi_i(X_{n+s}) - \E_\pi\phi_i)]\\
-&= \sum_{x, y\in S} \P_\pi(X_n = x, X_{n+s} = y) (\phi_i(x)- \E_\pi\phi_i)(\phi_i(y) - \E_\pi\phi_i).\nonumber
+\bC_\pi (\phi_i(X_n),  \phi_i (X_{n+s})) 
+&= \bE_\pi[(\phi_i(X_n) - \bE_\pi\phi_i)(\phi_i(X_{n+s}) - \bE_\pi\phi_i)]\\
+&= \sum_{x, y\in S} \bP_\pi(X_n = x, X_{n+s} = y) (\phi_i(x)- \bE_\pi\phi_i)(\phi_i(y) - \bE_\pi\phi_i).\nonumber
 \end{align}
@@
 \[
-\sum_{x, y\in S} \P_\pi(X_n = x) \P_\pi(X_{n+s} = y\mid X_n = x) (\phi_i(x)-
-\E_\pi\phi_i)(\phi_i(y) - \E_\pi\phi_i).
+\sum_{x, y\in S} \bP_\pi(X_n = x) \bP_\pi(X_{n+s} = y\mid X_n = x) (\phi_i(x)-
+\bE_\pi\phi_i)(\phi_i(y) - \bE_\pi\phi_i).
 \]
@@
 \[
-\sum_{x, y\in S} \pi(x)p^s_{xy} (\phi_i(x)-\E_\pi\phi_i)(\phi_i(y) - \E_\pi\phi_i).
+\sum_{x, y\in S} \pi(x)p^s_{xy} (\phi_i(x)-\bE_\pi\phi_i)(\phi_i(y) - \bE_\pi\phi_i).
 \]
@@
 % Eqn 2.6
-\C_\pi (\phi_i(X_n),  \phi_j (X_{n+s})) 
-&= \E_\pi[(\phi_i(X_n) - \E_\pi\phi_i)(\phi_j(X_{n+s}) - \E_\pi\phi_j)]\nonumber\\
-&= \sum_{x, y\in S} \pi(x)p^s_{xy} (\phi_i(x)-\E_\pi\phi_i)(\phi_j(y) - \E_\pi\phi_j).
+\bC_\pi (\phi_i(X_n),  \phi_j (X_{n+s})) 
+&= \bE_\pi[(\phi_i(X_n) - \bE_\pi\phi_i)(\phi_j(X_{n+s}) - \bE_\pi\phi_j)]\nonumber\\
+&= \sum_{x, y\in S} \pi(x)p^s_{xy} (\phi_i(x)-\bE_\pi\phi_i)(\phi_j(y) - \bE_\pi\phi_j).
 \end{align}
 Now let $\<\Phi\>$ denote the matrix of mean vectors whose $j$th column is 
-$\E_\pi\phi_j\one$, where $\one = (1,\dots, 1)^t$,
+$\bE_\pi\phi_j\one$, where $\one = (1,\dots, 1)^t$,
 and let $\Pi = \diag(\pi(\omega_1),\dots, \pi(\omega_d))$ be the $d \times d$
@@
 Finally, denoting by $\C(s)$ the $p \times p$ covariance matrix
-whose $(i,j)$ element is $\C_\pi(\phi_i(X_n), \phi_j(X_{n+s}))$, we have
+whose $(i,j)$ element is $\bC_\pi(\phi_i(X_n), \phi_j(X_{n+s}))$, we have
 \begin{align*}

### 2013-10-19 tex/LanczosProcedures.tex
minor formatting mods
miscellaneous fixes

@@
 \chapter{Lanczos Procedures}
-
+\label{cha:lanczos-procedures}
 The following will review the usual procedure for generating an orthonormal basis for
@@
 \section{A Lanczos Procedure for Markov Chains}
+% Sec 4.2
+\label{sec:lancz-proc-mark}
 Notice that the Lanczos algorithm requires the computation of $\M q_j$. Throughout this
@@
 &= \frac{1}{\beta_1^2}(q_1^t\M^3 q_1 - 2\alpha_1 \beta_1^2 + \alpha_1^3 )
-&= \bigl[q_1^t\M^2 q_1 - \alpha^2_1\bigr]^{1/2}.
+%&= \bigl[q_1^t\M^2 q_1 - \alpha^2_1\bigr]^{1/2}.
 \end{align}
@@
 &=  \frac{1}{\beta_j^2}(q_j^t\M^{k+2} q_j + \alpha^2_jq_j^t\M^{k} q_j 
-  + \beta^2_{j-1} q_{j-1}^t\M^k q_{j-1} 
-    - 2\alpha_jq_j^t\M^{k+1} q_j
+  + \beta^2_{j-1} q_{j-1}^t\M^k q_{j-1} - 2\alpha_jq_j^t\M^{k+1} q_j \nonumber\\
+& \qquad    
     + 2\alpha_j \beta_{j-1}q_j^t\M^k q_{j-1}

### 2013-10-19 tex/intro.tex
miscellaneous fixes

@@
-The rate at which a Markov chain converges to a given probability distribution has long
-been an active area of research. This is not surprising considering this problem’s relevance to the ar-
-eas of statistics, statistical mechanics, and computer science. Markov Chain Monte Carlo (MCMC)
-algorithms provide important examples. These algorithms come in handy when we encounter a
-complicated probability distribution from which we want to draw random samples. In statistical
-mechanics, we might wish to estimate the phase average of a function on the state space. Goodman
-and Sokal [6] examine Monte Carlo methods in this context. Examples from statistics occur in the
-Bayesian paradigm when we are forced to simulate an unwieldy posterior distribution (see, e.g.,
-Geman and Geman
+The rate at which a Markov chain converges to a given probability distribution
+has long been an active area of research. This is not surprising considering
+this problem's relevance to the areas of statistics, statistical mechanics, and
+computer science. Markov Chain Monte Carlo (MCMC) algorithms provide important
+examples. These algorithms come in handy when we encounter a complicated
+probability distribution from which we want to draw random samples. In
+statistical mechanics, we might wish to estimate the phase average of a function
+on the state space. Goodman and Sokal~\cite{GoodmanSokal:1989} examine Monte
+Carlo methods in this context. Examples from statistics occur in the Bayesian
+paradigm when we are forced to simulate an unwieldy posterior distribution (see,
+e.g., Geman and Geman~\cite{Geman:1984}.
 
 To implement the MCMC algorithm, we invent a Markov chain that converges to the
-desired distribution (this is often accomplished using the Metropolis algorithm
-described in Chapter 5). Realizations of the chain will eventually represent
-samples from this distribution. Sometimes ``eventually'' -- meaning all but
-finitely many terms of the chain -- is just not enough. We need more practical
-results. In particular, we want to know how many terms of the chain should be 
-discarded before we are sampling from a distribution that is close (in total variation distance) to
-% 03.txt
-the distribution of interest. This is the purpose of bounding convergence rates for Markov chains.
+desired distribution (this is often accomplished using the \emph{Metropolis algorithm}
+described in Chapter~\ref{cha:an-appl-conv}). Realizations of the chain will
+eventually represent samples from this distribution. Sometimes
+``eventually''---meaning all but finitely many terms of the chain---is just not 
+enough. We need more practical results. In particular, we want to know how many
+terms of the chain should be discarded before we are sampling from a
+distribution that is close (in total variation distance) to 
+%
+%
+% ------------- 03.txt -----------------------------------------------------
+%
+%
+the distribution of interest. This is the purpose of bounding convergence rates
+for Markov chains.
 
-Often the Markov chains encountered in this context satisfy a condition known in the
-physics literature as detailed balance. Probabilists call chains with this property reversible. This
-simply means that the chain has the same probability law whether time moves
-forward or 
-backward.\footnote{This is not a precise definition. In particular the chain must
-  have started from its stationary distribution. Full rigor is postponed until Section 2.1.}
-In this paper, we consider the rate at which such chains converge to a \emph{stationary distribution}.\footnote{This and other italicized terms are defined in Section 2.1.}
+Often the Markov chains encountered in this context satisfy a condition known in
+the physics literature as \emph{detailed balance}. Probabilists call chains with
+this property \emph{reversible}. This simply means that the chain has the same
+probability law whether time moves forward or backward.\footnote{This is not a
+precise definition. In particular the chain must have started from its
+stationary distribution. Full rigor is postponed until
+Section~\ref{sec:general-theory}.} 
+In this paper, we consider the rate at which such chains converge to
+a \emph{stationary distribution}. This and other italicized terms are
+defined in Section~\ref{sec:general-theory}.
 
-There are a number of different methods in common use for bounding convergence rates of
-Markov chains, and a good review of these methods with many references can be found in  More
-recently developed methods, employing logarithmic Sobolev inequalities, are reviewed in  Most
-of the bounds in common use involve the sub—dominant eigenvalue of the Markov chain's transition
-probability matrix, and thus require good approximations to such eigenvalues. In many applications,
-however, the transition probability matrix is so large that it becomes impossible to store even a
-single vector of the matrix in conventional computer memory. These so called out-of-core problems
-are not amenable to traditional eigenvalue algorithms\footnote{By ``traditional
-  eigenvalue algorithms'' we refer to those found, for example, in Golub and Van
-  Loan[5]. See also the book by Demmel [1] for a more recent discussion.}
-without modification. This paper develops 
-such a modification for the Markov chain eigenvalue problem. In particular it develops a method
-for approximating the first few eigenvalues of a transition probability matrix when we know the
-general structure of the underlying Markov chain. The method does not require storage of large
-matrices or vectors. Instead we need only simulate the Markov chain, and conduct a statistical
-analysis of the simulation.
+There are a number of different methods in common use for bounding convergence
+rates of Markov chains, and a good review of these methods with many references
+can be found in~\cite{Rosenthal:1995}.  More recently developed methods
+employing logarithmic Sobolev inequalities are reviewed in~\cite{Diaconis:1996}.  
+Most of the bounds in common use involve the subdominant eigenvalue of the
+Markov chain's transition probability matrix, and thus require good
+approximations to such eigenvalues. In many applications, however, the
+transition probability matrix is so large that it becomes impossible to store
+even a single vector of the matrix in conventional computer memory. These so
+called \emph{out-of-core} problems are not amenable to traditional eigenvalue
+algorithms without modification.\footnote{By ``traditional 
+  eigenvalue algorithms'' we mean those found, for example, in Golub and Van
+  Loan~\cite{Golub:1996}. See also the book by Demmel\cite{Demmel:1997} for more 
+  recent treatment.}
+In this paper we develop such a modification for the Markov chain eigenvalue
+problem. In particular we develop a method for approximating the first few
+eigenvalues of a transition probability matrix when we know the general
+structure of the underlying Markov chain. The method does not require storage
+of large matrices or vectors. Instead we need only simulate the Markov chain,
+and conduct a statistical analysis of the simulation.
 
-Here is a look at what follows. Section 2.1 contains a review of the relevant Markov chain
-theory. Readers conversant in the asymptotic theory of Markov chains might wish to at least skim
-Section 2.1, if only to become familiar with our notation. Section 2.2 describes functions on the state
+Here is a brief summary of the paper. Section~\ref{sec:general-theory} contains
+a review of the relevant Markov chain theory. Readers who already know the
+basics of asymptotic theory of Markov chains might wish to skim 
+Section\ref{sec:general-theory} if only to familiarize themselves with our
+notation. Section~\ref{sec:funct-state-space} describes functions on the state 
 % 04.txt
-space of the Markov process. This section and Chapter 3 develop the context in which we formulate
-the new ideas of the paper. In the last section of Chapter 3, Section 3.3, we present the familiar
-Krylov subspace and explain why this represents our best approximation to a subspace containing
-extremal eigenvectors of the transition probability matrix.\footnote{or, more
-  precisely, a similarity transformation of this matrix.} The first section of
-Chapter 4 describes the \emph{Lanczos algorithm} for generating an orthonormal basis
-for the Krylov subspace. As it stands, this algorithm is useless for an
-out-of-core problem such as ours since, by definition of such problems, 
-it requires too much data movement; all the computing time is spent swapping data between slow
-and fast memory (e.g. between the hard disk and cache). Therefore, we discuss alternatives to
-Lanczos and demonstrate that the \emph{Lanczos coeficients} are readily available through simulations of
-the Markov chain, which fact allows us to avoid the standard algorithm altogether. Following this
-is a chapter describing the Metropolis algorithm used to produce a reversible stochastic matrix.
-It is here that we experiment with the procedure described in Section 4.2 and approximate the
-extremal eigenvalues of the matrix, without storing any of its vectors. Finally, Chapter 6 concludes
-the paper.
+space of the Markov process. This section and Chapter~\ref{cha:invar-appr-invar}
+develop the context in which we formulate the new ideas of the paper. In the
+last section of Chapter~\ref{cha:invar-appr-invar},
+Section~\ref{sec:krylov-subspace}, we describe the \emph{Krylov subspace} and
+explain why this represents our best approximation to a subspace containing 
+extremal eigenvectors of the transition probability matrix (more
+  precisely, a similarity transformation of this matrix). The first section of
+Chapter~\ref{cha:lanczos-procedures} describes the \emph{Lanczos algorithm} for
+generating an orthonormal basis for the Krylov subspace. As it stands, this
+algorithm is useless for an out-of-core problem since, by
+definition of such problems, it requires too much data movement; all the
+computing time is spent swapping data between slow and fast memory (disk to ram
+to cache and back). We develop alternatives to the Lanczos algorithm and
+demonstrate that the \emph{Lanczos coefficients} of the algorithm can be
+obtained by simulations of the Markov chain, and this allows us to avoid
+the standard algorithm altogether. Following this is a chapter describing the
+Metropolis algorithm used to produce a reversible stochastic matrix. It is here
+that we experiment with the procedure described in
+Section~\ref{sec:lancz-proc-mark} and approximate the extremal eigenvalues of
+the matrix, without storing any of its vectors. Finally,
+Chapter~\ref{cha:conclusion} concludes the paper. 
 % 05.txt

### 2013-10-20 tex/Application.tex
finished LanczosProcedures chapter
miscellaneous fixes
finished Application and Conclusion Sections

@@
-Chapter 5
+\chapter{Convergence to Gibbs measure on the Ising lattice}
+%Chapter 5
+\label{cha:an-appl-conv}
 
-An Application: Convergence to
+\section{Physical Motivation}
+% Sec 5.1
+\label{sec:physical-motivation}
+Before considering the details of the Metropolis algorithm, it helps to
+understand the setting in which it was developed. To do so we consider a
+typical example, the \emph{Ising model}. This is a model of a system consisting
+of $n$ ``sites,'' each site taking values in 
+%% $\{-1, +1\}$. 
+$\{0, 1\}$. 
+It might 
+help to imagine these sites as equally spaced points on a line or a circle,
+though we are not restricted to such cases. 
+%% Then our state space $S$ consists of
+%% all possible $n$-dimensional permutations of $-1$'s and $+1$'s. 
+The state space $S$ consists of
+all possible $n$-dimensional permutations of $0$'s and $1$'s. 
+%% Thus, $S = \{-1, +1\}^n$ and $|S| = 2^n$. We can think of each
+Thus, $S = \{0, 1\}^n$ and $|S| = 2^n$. We can think of each
+site as a node existing in one of two possible positions, say ``on'' or
+``off.'' Each state $i \in S$ is %represented by 
+a unique permutation $\sigma_i$ of of $0$'s and $1$'s. 
+Let $\sigma_n(k)$ denote the position %(on or off) 
+of the $k$th node when the system is in state $i$. 
+An observable $\phi = \phi(\sigma_i)$ on this space is a function of
+the permutations $\sigma_i$, $i\in \{1,\dots, 2^n\}$.
+%
+%
+% ---------------- 31.txt ------------------------------------------------------
+%
+%
 
-Gibbs Measure on the Ising Lattice
+The \emph{energy} of the system, when in state $i$, is defined by the Hamiltonian
+\[
+H(\sigma_i) = - \sum_{\<j k\>} \sigma_i(j)\sigma_i(k)
+\]
+where the sum runs over all neart neighbor pairs. In statistical mechanics one is often concerned
+with the \emph{Gibbs measure} of state $i$, which is defined by
+\begin{equation}
+\label{eq:5.1}
+\pi(\sigma_i)= Z^{-1} \exp\{-\beta H(\sigma_i)\}
+\end{equation}
+where $Z^{-1}$ is a normalizing constant, sometimes called the \emph{partition function}. 
+For the Ising model, we put  
+\begin{equation*}
+Z = \sum_{i=1}^{2^n}\exp\{-\beta H(\sigma_i)\}
+\end{equation*}
+so that $\pi$ is a probability measure. 
+Now, letting $k$ denote the so called \emph{Boltzman constant}, if a
+classical mechanical system is in thermal equilibrium with its surroundings with
+absolute temperature $T$, and is in state $i$ with energy $H(\sigma_i)$, then the
+probability density in phase-space of the point representing state $i$ is given by
+(\ref{eq:5.1}) with $\beta = (kT)^{-1}$. A fundamental result of ergodic theory 
+implies that we can also interpret $\pi(\sigma_i)$ as the proportion of time the
+system spends in state $i$. If the system is observed at a random time, the
+expected value $\bE_\pi\phi$ of any observable $\phi$ is thus 
+\begin{equation}
+\label{eq:5.2}
+\bE_\pi\phi = 
+\sum_{i=1}^{2^n}\phi(\sigma_i) \pi(\sigma_i) = 
+%% \frac{\sum_{i=1}^{2^n}\phi(\sigma_i) \exp\{-\beta H(\sigma_i)\}}
+%% {\sum_{i=1}^{2^n} \exp\{-\beta H(\sigma_i)\}}
+Z^{-1}\sum_{i=1}^{2^n}\phi(\sigma_i) \exp\{-\beta H(\sigma_i)\}.
+\end{equation}
 
-5.1 Physical Motivation
+Perhaps the number of sites $n$ in our Ising model is so large that it is
+impractical or impossible to evaluate~(\ref{eq:5.2}).
+We might instead consider importance sampling and generate states with
+the probability density $\pi$ given in~(\ref{eq:5.1}).
+Then $\phi$ is itself an unbiased estimator of~(\ref{eq:5.2}).
 
-Before considering the details of the Metropolis algorithm, it helps to understand the
-setting in which it was developed. To do so we consider an typical example, the Ising model.
-Consider a model of a system consisting of 71 “sites”, each site taking values in {-1, +1}. It might
-help to imagine these sites as equally spaced points on a line or a circle, though we are not restricted
-to such cases. Then our state space 5' consists of all possible n—dimensional permutations of -1’s
-and +1’s. We reprent this as S’ = {-1, +1}", and note that |S| = 2". We can think ofeach site as
-a node existing in one of two possible positions, say “on" or “off”. Each state 1' E S is represented
-by a unique permutation a.- of —1’s and +1’s. Let a,-(k) = :|:l denote the position (on or off) of
-the kth node when the system is in state 1'. An observable ¢ = qS(a,-) on this space is a function of
+If evaluation of~(\ref{eq:5.2}) is difficult, there is no reason to believe that
+evaluation of~(\ref{eq:5.1}) is any easier. However, Metropolis and his
+collaborators~\cite{Metropolis:1953} contrived a method for producing an ergodic 
+%
+%
+% ------------------ 32.txt --------------------------------------------------
+%
+%
+Markov chain with transition probability matrix K whose elements $k(x,y)$ satisfy
+\[
+\sum_x \pi(x) k(x,y) = \pi(x)
+\]
+By 
+%Definition 2.1.4 
+Definition~\ref{def:2.1.4}
+this means that the Markov chain has stationary distribution $\pi$. Since the
+chain is ergodic, we know that it will eventually converge to the desired
+distribution. So, our main 
+problem---the rate at which such a Markov chain will converge---is of primary
+concern in applications of the Metropolis algorithm. Moreover, as we will see
+below, the Metropolis algorithm ensures that the Markov chain has the desired
+stationary distribution by requiring that the chain satisfy the stronger
+condition of reversibility, or ``detailed balance.'' Therefore, Markov chains
+produced by the Metropolis algorithm satisfy the assumptions of this paper,
+and our theory and methods can be used to bound convergence rates of such chains.
 
-the permutations 47,-, 1' 6 {1,. . . , 2"}.
+\section{Description of the Metropolis Algorithm}
+% Sec 5.2 
+\label{sec:descr-metr-algor}
+The following explains how the Metropolis Algorithm is carried out. The exposition is
+similar to that given in Hammersley and Handscomb's book 
+\emph{Monte Carlo Methods} \cite{Hammersley:1964}. We refer the reader desiring
+more details to~\cite[Section 9.3]{Hammersley:1964}.
 
-% 31.txt
-31
+To begin, choose an arbitrary symmetric Markov chain; that is, a Markov chain whose
+transition probability matrix P is symmetric. The matrix P is called the 
+\emph{proposition matrix}, and its elements
+%, $p(i,j)$,
+satisfy, for all $i$ and $j$ in $S$,
+\begin{equation}
+\label{eq:5.4}
+p(i,j) \geq 0, \quad p(i,j) = p(j,i), \quad \sum_m p(i,m)=1.
+\end{equation}
+Using P, we will derive a new transition probability matrix K with elements
+$k(i,j)$ satisfying
+\begin{equation}
+\label{eq:5.5}
+\sum_i \pi(i)k(i,j) = \pi(j)
+\end{equation}
+For $i\neq j$ define
+\begin{equation}
+\label{eq:5.6}
+k(i,j) = 
+\begin{cases}
+p(i,j)\pi(j)/\pi(i), & \text{ if $\pi(j)/\pi(i) < 1$,}\\
+p(i,j), & \text{ if $\pi(j)/\pi(i) \geq 1$.}
+\end{cases}
+\end{equation}
+%
+%
+% ----------------- 33.txt ----------------------------------------------
+%
+%
+For $i=j$ define
+\begin{equation*}
+k(i,j) = 
+p(i,i) + \sum_{j\in J_i}\bigl(1 - \frac{\pi(j)}{\pi(i)}\bigr)p(i,j), 
+\end{equation*}
+where $J_i = \{j : j\neq i, \pi(j)/\pi(i) < 1\}$.
+Notice that $\pi(i) > 0$ implies $k(i,j) \geq 0$, and
+\begin{align}
+\label{eq:5.7}
+\sum_j k(i,j) 
+&= p(i,i) + 
+\sum_{j\in J_i}\left\{\bigl(1 - \frac{\pi(j)}{\pi(i)}\bigr)p(i,j)
++ \frac{p(i,j)\pi(j)}{\pi(i)}\right\} 
++ \sum_{j\in J_i^c}p(i,j)
+\nonumber\\
+&= p(i,i) + 
+\sum_{j\in J_i}p(i,j)
++ \sum_{j\in J_i^c}p(i,j).
+\end{align}
+Equation~(\ref{eq:5.7}) shows that for each $i$ we have $\sum_j k(i,j) = 1$
+(the row sums of K are 1), which confirms that K is a stochastic matrix.
 
-The energy of the system, when in state i, is deﬁned by the Hamiltonian
+Next we check that K satisfies the following \emph{detailed balance} condition:
+\begin{equation}
+\label{eq:detailed-balance}
+\pi(i)k(i,j) = \pi(j)k(j,i).
+\end{equation}
+Obviously this holds when $i=j$, so assume $i\neq j$.
 
-H07-‘) = - Z Ua(i)0s(k)
+First suppose $\pi(i) = \pi(j)$.
+In this case, the definition~(\ref{eq:5.6}) and symmetry of P imply that
+\[
+k(i,j) = p(i,j) = p(j,i) = k(j,i).
+\]
+Therefore, when $\pi(i) = \pi(j)$ the detailed balance
+condition~(\ref{eq:detailed-balance}) clearly holds.
 
-<j Ic>
-where the sum runs over all neart neighbor pairs. In statistical mechanics we are often concerned
+Now suppose that $\pi(i) > \pi(j)$. Then the second case in definition (\ref{eq:5.6}) implies
+$k(j,i) =p(j,i)$, while the first case gives
+\[
+k(i,j) = p(i,j)\pi(j)/\pi(i) =  p(j,i)\pi(j)/\pi(i) = k(j,i)\pi(j)/\pi(i),
+\]
+and we see that 
+$\pi(i)k(i,j) = \pi(j)k(j,i)$ holds.
+Finally, a symmetric argument shows that if $\pi(j) > \pi(i)$,
+then detailed balance is again satisfied.
 
-with the Gibbs measure of state i, deﬁned by
-1r(a,-) = Z'1 exp{—[3H(a,~)} (5.1)
-where Z ‘1 is the normalization constant, or partition function. For the Ising model, we put
+To complete the justification of the Metropolis algorithm, we note
+detailed balance~(\ref{eq:detailed-balance})
+implies that $\pi$ is indeed the stationary distribution for a Markov 
+chain with transition probability matrix K; that is, (\ref{eq:5.5}) is
+satisfied. Indeed, by the detailed balance condition and the fact that
+the row sums of K are 1, we have
+\begin{equation*}
+\sum_i \pi(i)k(i,j)
+=\sum_i \pi(j)k(j,i)
+=\pi(j)\sum_i k(j,i)
+ = \pi(j).
+\end{equation*}
 
-Z = §:9XP{~ﬁH(<7:‘)}
-i=1
-
-so that 7r is a probability measure. Now, letting 1: denote the so called Boltzman constant, if a
-classical mechanical system is in thermal equilibrium with its surroundings with absolute temper-
-ature T, and is in state i with energy H (05), then the probability density in phase-space of the
-point representing state i is given by (5.1) with [3 = (lcT)“. The central result of ergodic theory
-implies that we can also interpret 1r(o;) as the proportion of time the system spends in state i. If
-
-the system is observed at a random time, the expectation E,,¢ of any observable ¢ is thus
-
-.31 ¢(0a) 9"l>{-ﬂH(0s‘)}
-
-E” = 2?;.exp{—nH(a.-)}
-
-(5.2)
-
-Perhaps the number of sites n in our Ising model is so large that it is impractical or impossible to
-evaluate (5.2). We might instead consider importance sampling and generate states with probability
-
-density
-
-exp{—aH<a.o}
-2?;.exp{—zaH<a.->} (53)
-
-Then «I: is itself an unbiased estimator of (5.2).
-
-1r(0a') =
-
-If evaluation of (5.2) is difficult, there is no reason to believe that evaluation of (5.3) is any
-
-easier. However, Metropolis and his collaborators [9] contrived a method for producing an ergodic
-
-% 32.txt
-32
-
-Markov chain with transition probability matrix K whose elements Ic(:::, y) satisfy
-
-2«<z>k<x.y) = «<2»
-
-I
-
-By Deﬁnition 2.1.4 this means that the Markov chain has stationary distribution 7r. Since the
-drain is ergodic, we know that it will eventually converge to the desired distribution. So, our main
-problem — the rate at which such a Markov chain will converge — is clearly at issue in the application
-of the Metropolis algorithm. Moreover, as we will see below, the Metropolis algorithm ensures that
-the Markov chain has the desired stationary distribution by requiring that the chain satisfy the
-stronger condition of reversibility; i.e. detailed balance. Therefore, Markov chains produced by the
-Metropolis algorithm satisfy the assumptions of this paper, which fact conﬁrms that our theory
-
-can be used to bound convergence rates of all such chains.
-
-5.2 Description of the Metropolis Algorithm
-
-The following explains how the Metropolis Algorithm is carried out. The exposition is
-similar to that given in Hammersley and Handscomb’s book Monte Carlo Methods 
-To begin, choose an arbitrary symmetric Markov chain; that is, a Markov chain whose
-
-transition probability matrix P is symmetric. P is called the proposition matrix, and its elements,
-P071"), Satisfy
-pa, 1) 2 0. pm‘) = pm). 2,-pm‘) = 1 vi e S (5-4)
-
-Using P, we will derive a new transition probability matrix K which satisﬁes
-
-Z vr(z')k(='.:') = arm (5.5)
-
-I
-
-For 1' aé j deﬁne
-
-kw) = p<-yum")/«<¢> it nu)/1r(='> <1 (5.6)
-
-P(i,J') if 7r(J')/7r(i) Z1
-
-% 33.txt
-33
-
-For 2' = j deﬁne
-
-kw) = pm) (1—  p(-3;"), where J. = {j = 2‘ 7‘ an nu)/«(='> < 1}
-
-Notice that 1r(i) > 0 implies k(i,j) _>_ 0 and
-
-;k<«',j> = pow)  { (1 —  pm) +  +]§§p<«",:')
-= P(is'5) + 2 P(i1.1') + Z P(iy.7') (5-7)
-jel.» jug
-
-Equation (5.7) shows that, for each i, 2, k(z', j) = 1, which fact conﬁrms that K is a stochastic
-
-matrix.
-
-Next we check that K satisﬁes the detailed balance condition. First suppose that 1r(z') =
-
-1r(j). Then (5.4), (5.6), and symmetry of P imply
-k(i,J') = p(i2.7') = p(jvi) = k(j1i)
-
-and, since 1r(i) = 1r(j), the detailed balance condition, 1r(i)k(i,j) = 7r(j)k(i,j), clearly holds. Now
-
-suppose that 1r(i) > 1r(j). Then (5.4) and (5.6) imply k(j,i) =p(j, 1') and
-k(i,J') = P(iyJ')7r(J')/7r(i) = P(J3i)7r(J')/7r(i) = k(.1'» i)7r(J')/1r(i)
-
-That is, 1r(i)lc(i,j) = 1r(j)k(i,j), as we set out to prove.
-To complete the justification of the Metropolis algorithm, we note that the detailed balance
-condition, 7r(i)k(i, j) = 1r(j)k(i, j), implies that 1r is indeed the stationary distribution for a Markov
-
-chain with transition probability matrix K. To see this notice that
-
-27r(i)k(i..7') = Zﬂijlkiiajl = 1r(J') Z/€(i,J‘) = 7r(J')
-
-I i
 We now summarize the tasks performed when implementing the Metropolis algorithm.
-
-First, pick as a proposition matrix any symmetric stochastic matrix and denote it by P. Begin
-
-% 34.txt
-34
-
-the simulation of the Markov chain in state i, and then take one step of the proposition Markov
-chain to arrive in state j. That is, choose j according to the mass function p(z', -). Next, compute
-1r(j)/1r(i). If 7r(j)/1r(i) 2 1, accept j as the new state. If1r(j)/1r(i') < 1, then with probability
-1r(j)/7r(z') accept j as the new state; otherwise, with probability 1 — 1r(j)/1r(i), take i as the new
-state. Performing each of these tasks produces one step of the Markov chain represented by the
-
+First, pick as a proposition matrix any symmetric stochastic matrix and denote
+it by P. Begin 
+%
+%
+% ----------- 34.txt --------------------------------------------------------
+%
+the simulation of the Markov chain in state $i$, and then take one step of the
+proposition Markov chain to arrive in state $j$. That is, choose state $j$
+according to the probability mass function $p(i, \cdot)$. 
+Next, compute $\pi(j)/\pi(i)$. 
+If $\pi(j)/\pi(i) \geq 1$, accept $j$ as the new state. If $\pi(j)/\pi(i) < 1$,
+then with probability $\pi(j)/\pi(i)$ accept $j$ as the new state; otherwise
+(with probability $1- \pi(j)/\pi(i)$), take $i$ as the new state. Performing
+each of these tasks produces one step of the Markov chain represented by the 
 transition probability matrix K.
 
-5.3 The Ising Model
+\section{The Ising Model}
+% Sec. 5.3 
+\label{sec:ising-model}
+Returning to the Ising model considered in
+Section~\ref{sec:physical-motivation}, we consider how the Metropolis 
+algorithm is applied to such a model. Again, start the simulated chain in state
+$X_0 = i$ and let $X_1 = j$ be a state chosen according to 
+$p(i, \cdot)$, where $p(i,j) = p(j,i)$. Next, compute 
+\[
+\pi(\sigma_j)/\pi(\sigma_i) = \exp\{-\beta[H(\sigma_j) - H(\sigma_i)]\}
+\]
+Suppose the energy $H(\sigma_j)$ of the new state satisfies 
+$H(\sigma_j) \leq H(\sigma_i)$.
+Then $\pi(\sigma_j)/\pi(\sigma_i)$ is at least 1 and we move to state $j$.
+On the other hand, if
+$H(\sigma_j) > H(\sigma_i)$,
+then $\pi(\sigma_j)/\pi(\sigma_i)< 1$ and we take
+\begin{equation}
+\label{eq:5.8}
+X_1 = 
+\begin{cases}
+j & \text{ with probability $e^{-\beta\Delta H}$},\\
+i & \text{ with probability $1-e^{-\beta\Delta H}$}.
+\end{cases}
+\end{equation}
+where $\Delta H = H(\sigma_j) - H(\sigma_i)$ is the change in energy from state $j$ to state $i$.
 
-Returning to the Ising model considered in section 5.1, we consider how the Metropolis
-algorithm is applied to such a mode]. Again, start the simulated chain in state X0 = 1' and let
+There are many ways to choose a proposition matrix P. The simplest derives from the
+so called \emph{Glauber dynamics}, which proceeds as follows: choose one of the
+$n$ sites at random with probability $1/n$---let $s$ denote the chosen
+site---and change the position of this site; i.e., subtract its value from 1.
+%multiply its value by $-1$. 
+The resulting state $j$ is the 
+%permutation of 
+%-1’s and +1’s
+% 0's and 1's 
+binary string
+given by
+\begin{equation*}
+\sigma_j(k) = 
+\begin{cases}
+1-\sigma_i(k) & \text{ if $k =s$},\\
+\sigma_i(k) & \text{ if $k =s$},
+\end{cases}\quad (k = 1, \dots, d).
+\end{equation*}
+%
+%
+% -------------- 35.txt ---------------------------------------------------------
+%
+%
+The procedure we employ below involves a slight variation: choose site $s$ at
+random uniformly from the $n$ possible sites. However, instead of changing the
+value of the chosen site with absolute certainty, as in Glauber dynamics, we
+``flip the switch'' with probability $1/2$. 
+Therefore, when $k \neq s$
+we have $\sigma_j(k) = \sigma_i(k)$, and when $k = s$,
+\begin{equation*}
+\sigma_j(s) = 
+\begin{cases}
+1-\sigma_i(s) & \text{ with probability 1/2},\\
+\sigma_i(s) & \text{ with probability 1/2}.
+\end{cases}
+\end{equation*}
+The advantage of this procedure is that the proposition matrix, before altering
+it with the Metropolis algorithm, is a bit more interesting than that produced
+by traditional Glauber dynamics. For, it has 1/2 along the main diagonal and is
+thus aperiodic and converges to a stationary distribution. Glauber dynamics has
+period 2, so its proposition matrix does not satisfy the hypotheses of our
+convergence theorem (Theorem~\ref{thm-2.1.2}). Keep in mind, however, that any
+proposition matrix modified by the Metropolis algorithm satisfies our hypotheses.
+\section{Computations}
+% Sec 5.4
+\label{sec:computations}
+\subsection{Computer Programs}
+Writing a computer program to simulate realizations of the Markov chain described
+above (i.e. with transition matrix K) is straight forward, and we do so using
+Matlab and the program {\tt ising.m} which appears in the Appendix.
 
-X1 = j be a state chosen according to p(i, -), where p(i, j) = p(j,i). Next, compute
+When performing simulations in the present context, there are a few important
+aspects to consider. First, our program begins with a ``hot start,'' which means
+that the initial state is picked by assigning each site to the value 0 or 1, each
+with probability 1/2.\footnote{We cannot simply draw our initial state
+from the stationary distribution since the point of this work
+is to deal with cases for which samples from the stationary distribution are not
+immediately attainable. Furthermore, we are trying to determine how long it will
+take for observations from the simulated chain to represent samples from the
+stationary distribution.}
+%% figure out how long it will take for observations from the simulated chain to represent samples from
+%% the stationary distribution. 
+%% Perhaps we should instead have picked our initial state
+%% from the stationary distribution. On the contrary, this would defeat the
+%% whole purpose of our analysis.  
+%% Recall that the method has been developed to deal with
+%% cases for which 
+%
+%
+% ----------------- 36.txt --------------------------------------------
+%
+%
+%% sampl from the stationary distribution are not quickly attainable. Furthermore, we are trying to
+%% figure out how long it will take for observations from the simulated chain to represent samples from
+%% the stationary distribution.
 
-7"(‘7j)/7"(‘7i) = eXP{-ﬁlH(<7j) - H(0i)l}
+%% sampl from the stationary distribution are not quickly attainable. Furthermore, we are trying to
+%% figure out how long it will take for observations from the simulated chain to represent samples from
+%% the stationary distribution. 
+%% Second, t
 
-Suppose the energy H (Uj) of the new state satisﬁes H (U,-) 3 H (05). Then 1r(a_,-)/1r(a,-) Z 1 and we
-
-move to state j. On the other hand, if H(aj) > H(a;), then 1r(aj)/7r(a,-) < 1 and we take
-
-j with probability e-M"
-
-X1 (5.8)
-
-i with probability 1 — e-W’
-where AH E H (0,) — H (:75) is the change in energy from state j to state 1'.
-There are many ways to choose a proposition matrix P. The simplest derives from the
-so called Glauber dynamics, which proceeds as follows: choose one of the 11. sites at random with
-probability 1/n — let .9 denote the chosen site — and change the position of this site; i.e., multiply
-
-its value by -1. The resulting state j is the permutation of -1’s and +1’s given by
-
--0,-(k) if]: = .9
-a,~(Ic)= (k=1,...,d)
-
-a,-(Ic) if]: 95 s
-
-% 35.txt
-35
-
-The procedure we employ below involves a slight variation: choose site .9 at random uniformly
-from the 11 possible sites. However, instead of changing the value of the chosen site with absolute
-certainty, as in Glauber dynamics, we “ﬂip the switch” with probability 1/2. Therefore, when k yé .9
-
-we have 0',-(k) = a.~(k), and when k = 3,
-
-—a,~(s) with probability 1/2
-01(3) =
-
-a,-(3) with probability 1/2
-The advantage of this procedure is that the proposition matrix, before altering it with the Metropo-
-lis algorithm, is a bit more interesting than that produced by traditional Glauber dynamics. For,
-it has 1/2 along the main diagonal and, thus, is aperiodic and converges to a stationary distribu-
-tion. Glauber dynamics has period 2, so its proposition matrix does not satisfy the hypotheses of
-
-our convergence theorem (Theorem 2.1.2). Keep in mind, however, that any proposition matrix
-
-modiﬁed by the Metropolis algorithm satisﬁes all of our hypotheses.
-
-5.4 Computations
-
-5.4.1 Computer Programs
-
-Constructing a computer program to simulate realizations of the Markov chain described
-above (i.e. with transition matrix K) is straight forward, and we do so using Matlab and the
-program ising.m which appears in the Appendix.
-
-When performing simulations in the present context, there are a few important aspects
-to consider. First, our program begins with a “hot start", which means that the initial state is
-picked by assigning each site to :l:1 with probability 1/2. Perhaps we should instead have picked
-our initial state from the stationary distribution. On the contrary, this would defeat the whole
-
-purpose of our analysis. Recall that the method has been developed to deal with cases for which
-
-% 36.txt
-36
-
-sampl from the stationary distribution are not quickly attainable. Furthermore, we are trying to
-ﬁgure out how long it will take for observations from the simulated chain to represent samples from
-the stationary distribution. Second, the method we have developed depends solely on covariances
-of our observable which we estimate from the data. produced by the simulations. However, such
-estimates require that the data come from the stationary distribution. Therefore, we must discard a
-number of observations before using the data to estimate covariances. The number of observations
-to discard depends on how long must we wait until we have data from the stationary distribution,
-and we are back where we started — almost. When employing our procedure, we can discard a
+The method we have developed depends solely on covariances
+of our observable, which we estimate from the data produced by the
+simulations. However, such estimates require that the data come from the
+stationary distribution. Therefore, we must discard a number of observations
+before using the data to estimate covariances. The number of observations 
+to discard depends on how long must we wait 
+before we can expect the data to represent samples from the stationary distribution.
+%% until we have data from the stationary distribution,
+%% and we are back where we started — almost.
+When employing our procedure, we can discard a
 conservative (very large) number of observations and perform our analysis, deriving a bound on
 the convergence time. Then, in all future studies, we will know how much data should be discarded
-before samples come (approximately) from the stationary distribution.
+before samples can be assumed to have come from the stationary distribution.
 
 In considering what observable to use on the state space described above, perhaps the
-most obvious is simply the number of nodes in the on position (i.e. the number of +l’s). Recalling
-that 0, (k) = i1 denotes the position (on or off) of the kth node when the system is in state j, this
+most obvious is simply the number of nodes in the ``on'' position (i.e., the
+number of 1's). Recalling that $\sigma_j(k)\in \{0,1\}$ denotes the position of
+the $k$th node when the system is in state $j$, this observable is simply the
+sum $\phi(j) = \sum_k\sigma_j(k)$. 
+%% where X is the indicator function. 
+The program {\tt ising.m} simulates the Markov chain, computes the
+values of this observable (and its square) and writes these data to a file
+called {\tt phi.dat}. 
+The program then computes the first few Lanczos coefficients using covariances
+produced by Matlab, so that we can check them against the results we get from the
+{\tt lanczos.c} program, which we now describe.
 
-observable is deﬁned by,
-«M = zxca.-ac) = +1). 1 = 1.---.ISl
-I:
+The lanczos.c program (listed, along with its dependencies, in the Appendix) implements
+the recursive relations described in Section~\ref{sec:lancz-proc-mark} % Sec 4.2
+to compute the Lanczos coefficients by the new
+method. It accepts input from the user specifying any observable, any number of
+iterations, and any 
+%
+%
+% ---------------  37.txt  -----------------------------------------------
+%
+%
+number of desired Lauczos coefiicients. That is, the observable could have come from a simulation
+of any reversible Markov chain and not just our special Metropolis chain. The output of this
+program is a matrix T containing the maximum number of Lanczos coefficients (up
+to the number requested by the user) before an approximate invariant subspace
+was reached. The eigenvalues of the nonzero part of this matrix, in accordance
+with the theory developed above, should provide a close approximation to the
+eigenvalues of the Markov chain's stochastic matrix.
 
-where X is the indicator function. The program ising.m simulates the Markov chain, computes the
-values of this observable (as well as the number of 1’s squared) and writes it to a ﬁle called phi . dat.
-The program then proceeds to compute the ﬁrst few Lanczos coefficients using covariances produced
-by Matlab, so that we can check them against the rults we get from the 1anczos.c program,
-which we now describe.
+\subsection{Results}
+% Sec. 5.4.2 
+\begin{table}
+\centering
+\label{tab:1}
+\caption{Estimates of $\lambda_{\max}(\K) = 0.998746$ using $\phi_1$.}
+\vskip2mm
+\begin{tabular}{r|c|c|c|c}
+simulation \# & $\alpha_1$ & $\beta_1$ & $\lambda_{\max}(\T_2)$& $\beta_2$\\
+\hline
+1& 0.995107& 0.006960& 0.995220& 0.600243\\
+2& 0.995099& 0.005891& 0.994863& 0.477393\\
+3& 0.994862& 0.009671& 0.994925& N/A\\
+4& 0.995321& 0.009319& 0.995654& N/A\\
+5& 0.994787& 0.004872& 0.405446& 0.994742
+\end{tabular}
+\end{table}
+Table~\ref{tab:1} shows the results of 5 simulations each of 100,000 iterations (discarding the first
+5,000 observations) for the one dimensional Ising lattice with 10 nodes and $\beta= 1$. Displayed is
+the estimate of the first Lanczos coefficient $\alpha_1$ (which is the
+eigenvalue estimate after one step; i.e., $\lambda_{\max}(\T_1)$), along with its
+error bound $\beta_1$. Appearing in the third column is the largest eigenvalue,
+$\lambda_0(T_2)$, of the $2 \times 2$ matrix $\T_2$, and $\beta_2$ appears in the
+fourth column when it is was computable. When it was not computable, it was
+%% probably 
+because an approximate invariant subspace was reached 
+at the first step, so $\beta_2$ involved expressions which were close to machine
+epsilon and thus could not be computed accurately. Table~\ref{tab:2}  shows the same
+information when using a second observable, 
+$\phi_2 = \phi_1^2$ (i.e., the number of 1's squared).
 
-The lanczos .c program (listed in the Appendix along with its dependencies) implements
-the recursive relations described in section 4.2 to compute the Lanczos coefﬁcients by the new
+\begin{table}
+\caption{Estimates of $\lambda_{\max}(K) = 0.998746$ using $\phi_2$}
+\label{tab:2}
+\begin{tabular}{r|c|c|c|c}
+simulation \# & $\alpha_1$ & $\beta$ & $\lambda_{\max}(\T_2)$& $\beta_2$\\
+\hline
+1& 0.994262& 0.006142& 0.994222& N/A\\
+2& 0.994333& 0.002974& 0.994332& N/A\\
+3& 0.994307& 0.008892& 0.994374& 0.726320\\
+4& 0.994436& 0.012884& 0.994685& 0.361404\\
+5& 0.993789& 0.006909& 0.993820& 0.788710
+\end{tabular}
+\end{table}
+The true value of the subdominant eigenvalue for this problem is $\lambda_{\max}(\K) = 0.998746$,
+and the next largest is $\lambda_2(\K) = 0.993303$. These were computed directly
+from the transition matrix 
+%
+%
+% ---------- 38.txt ----------------------------------------------
+%
+%
+using the Matlab program {\tt tpm.m} listed in the Appendix. They should be very accurate, and it
+takes the Matlab routine {\tt eig()} just over 20 minutes of cpu time on a Sun
+Ultra Sparc to find all the eigenvalues of K.
 
-method. It accepts input from the user specifying any observable, any number of iterations, and any
-
-% 37.txt
-37
-
-TABLE 1: ESTIMATES or /\,m(K) = 0.998746 USING ¢1
-
-sim. # /32
-1 0.995107 0.006960 0.995220 0.600243
-2 0.995099 0.005891 0.994863 0.477393
-3 0.994862 0.009671 0.994925 N/ A
-4 0.995321 0.009319 0.995654 N/A
-5 0.994787 0.004872 0.405446 0.994742
-
-number of desired Lauczos coeﬁicients. That is, the observable could have come from a simulation
-of any reversible Markov chain and not just our special Metropolis chain. The output of this
-program is a T matrix containing the maximum number of Lauczos coeﬂicients (up to the number
-requested by the user) before an approximate invariant subspace was reached. The eigenvalues of
-the nonzero part of this matrix, in accordance with the foregoing theory, should well approximate
-
-eigenvalues of the Markov chain’s stochastic matrix.
-
-5.4.2 Results
-
-Table 1 shows the results of 5 simulations each of 100,000 iterations (discarding the ﬁrst
-5,000 observations) for the one dimensional Ising lattice with 10 nodes and [7 = 1. Displayed is
-the estimate of the ﬁrst Lauczos coeﬂicient a1 (which is the eigenvalue estimate after one step; i.e.
-).m,x(T1)), along with its error bound ﬁl. Appearing in the third column is the largest eigenvalue of
-the 2 x 2 matrix T2, A0 (T2), and ﬁg appears in the fourth column when it is was computable. When
-it was not computable, it was probably because an approximate invariant subspace was reached
-at the ﬁrst step, so ﬁg involved expressions which were close to machine epsilon and, thus, could
-not be computed accurately. Table 2 shows the same information when using a second observable,
-¢2 E ¢§ (i.e. the number of +1’s squared).
-
-The true value of the subdominant eigenvalue for this problem is /\max(K) = 0.998746,
-
-and the next largest is A2(K) = 0.993303. These were computed directly from the transition matrix
-
-% 38.txt
-38
-
-TABLE 2: ESTIMATES OF >.,,,,,.,,,(K) = 0.998746 USING z/>2
-
-sim. # B2
-1 0.994262 0.006142 0.994222 N / A
-2 0.994333 0.002974 0.994332 N / A
-3 0.994307 0.008892 0.994374 0.726320
-4 0.994436 0.012884 0.994685 0.361404
-5 0.993789 0.006909 0.993820 0.788710
-
-using the Matlab program tpm.m listed in the Appendix. They should be very accurate, and it
-takes the Matlab routine eig() just over 20 minutes of cpu time on a Sun Ultra to ﬁnd all the
-eigenvalues of K.
-
-As shown by Theorem 4.1.1, the [Ts provide conservative error bounds on the eigenvalue
-estimates from the previous Lanczos step. Unfortunately, for this example, ,6 appears to be too
-conservative as it does not even allow us to bound our estimates away from 1. Furthermore, it
-seems that the estimates generated by the method above fall systematically below the true value of
-the subdominant eigenvalue. In every simulation, an approximate invariant subspace (to machine
-tolerance) was reached after computing at most 2 pairs of Lanczos coefficients. For ¢1 we see that
-it was reached after computing only one pair on simulations 3 and 4, similarly for ¢2 on simulations
-1 and 2. These observations indicate that the chosen observable is close to the slowest mode (or
-eigenfunction) of the process. That is, we should observe fast convergence on an approximate
+As montioned in Section~\ref{sec:gener-lancz-algor} (see
+Theorem~\ref{thm:4.1.1}), the $\beta$ coefficients provide conservative error 
+bounds on the eigenvalue estimates from the previous Lanczos
+step. Unfortunately, for this example $\beta$ appears to be too conservative as
+it does not even allow us to bound our estimates away from 1. Furthermore, it  
+seems that the estimates generated by the method above fall systematically below
+the true value of the subdominant eigenvalue. In every simulation, an
+approximate invariant subspace (to machine tolerance) was reached after
+computing at most 2 pairs of Lanczos coefficients. For $\phi_1$ we see that 
+it was reached after computing only one pair on simulations 3 and 4, similarly
+for $\phi_2$ on simulations 1 and 2. These observations indicate that the chosen
+observable is close to the slowest mode (or 
+eigenfunction) of the process. That is, we should observe fast convergence to an approximate
 invariant space. However, the space found might contain only the second slowest mode of the
 process and, in that case, our eigenvalue estimates would be closer to the third largest eigenvalue,
-instead of the second largest as desired.‘
+instead of the second largest.\footnote{Recall from above that we won't
+converge to the eigenspace corresponding to the largest eigenvalue 
+(which is always 1), unless we choose a nearly constant observable function.}
 
-The matrix T was produced by the program lanczos . c and its eigenvalues were computed
+The matrix T was produced by the program {\tt lanczos.c} and its eigenvalues were computed
 by Matlab, both operations taking a few seconds. Granted, comparing a few seconds to the 20
-
-minutes it takes Matlab is not justiﬁed since a simple call to the routine eig() is far from the
-
-‘Recall that we explained above why we will never converge to the eigenspace corresponding to the largest eigen-
-value (which is always 1). That would require a (nearly) constant observable function.
-
-% 39.txt
-39
-
-most eﬂicient traditional method for computing only a few eigenvalues of K. However, recall that
+minutes it takes Matlab to compute all the eigenvales is not really fair, 
+since calling Matlab's {\tt eig()} routine is not always the
+%
+%
+% --------- 39.txt -----------------------------------------------------
+%
+%
+optimal traditional method for computing only a few eigenvalues of K. However, recall that
 the primary motivation for the new method are those examples where the matrix K is so large
-that we can’t store even a single vector in fast memory. Once we start swapping data to and from
-cheap memory, the time required by traditional algorithms can become years, even lifetimes. Such
-examples are not exceptional. Consider, for instance, the 1000 node problem which produces a.
-
-2“’°° x 21°°° transition matrix.
-
+that we can't store even a single column vector in fast memory. Once we start
+swapping data to and from slow memory, traditional algorithms can become
+intractable, and such examples are easy to come by. For instance, if the Ising
+model described has 1000 nodes, it produces a $2^{1000}\times 2^{1000}$
+transition probability matrix.
+%
+%
 % 40.txt
-40
+%
+%

### 2013-10-20 tex/Conclusion.tex
miscellaneous fixes
finished Application and Conclusion Sections

@@
-Chapter 6
-
-Conclusion
-
-We conclude by noting a few aspects of the special cases to which the foregoing applies.
-We are interested in the subdominant eigenvalue of a large transition probability matrix. The
-
-foregoing can be used to approximate these eigenvalues when the following conditions are satisﬁed:
-o The Markov chain is reversible (detailed balance condition).
-
-0 We can easily simulate realizations of the chain as well as functions (observables) of these
-
-realizations.
-
-When these conditions hold, we can side step the traditional Lanczos algorithm, and get at the
-values of the Lanczos coeﬂicients via the variational properties of observables on the state space.
-There are many ideas in this paper that should be expanded upon in future studies. Per-
-haps most obvious is the desire to ﬁnd analogous techniques for non-reversible (i.e. non-symmetric)
-Markov chains. This is a considerable problem since the Lanczos methods described above are no
-longer applicable. Perhaps the Amoldi algorithm ([1], algorithm 6.9), or the non-symmetric Lanczos
-
-algorithm can also be usefully specialized for stochastic matrices.
-
+\chapter{Conclusion}
+% Chapter 6
+\label{cha:conclusion}
+We conclude with some remarks about the special situations in which our
+method is useful.
+%few aspects of the cases to which the foregoing applies.
+We are interested in the subdominant eigenvalue of a large transition probability matrix. Our
+method can be used to approximate these eigenvalues when the following conditions are satisﬁed:
+\begin{itemize}
+\item 
+The Markov chain is reversible (satisfies detailed balance).
+\item
+We can simulate realizations of the chain as well as functions of these
+realizations (observables on the state space).
+\end{itemize}
+When these conditions hold, we can avoid the traditional Lanczos algorithm, and
+estimate the values of the Lanczos coeﬂicients via the variational properties of
+observables on the state space. 
+There are a number of ideas in this paper that could be expanded upon in future
+studies.  Perhaps most obvious would be to try to find analogous techniques for
+nonreversible (i.e. nonsymmetric) Markov chains. This is a considerable problem
+since the Lanczos methods described above are no longer applicable. Perhaps the
+\emph{Arnoldi algorithm} (see \cite[Algorithm 6.9]{Demmel:1997}), or the
+\emph{nonsymmetric Lanczos algorithm} could also be modified to exploit
+special properties of stochastic matrices. %%  (as we have done with the Lanczos
+%% algorithm in the symmetric case). 
+%
+%
 % 41.txt
-41
+%
+%
 
@@
 Then we do the same with a new starting vector and compare the results to those obtained with
-the ﬁrst vector. Proceeding in this way for a number of starting vectors, we can provide statistical
+the first vector. Proceeding in this way for a number of starting vectors, we can provide statistical
 evidence that we have, indeed, located the eigenvalues of interest. However, when there are two
-observables 451, :15; of particular interest, perhaps we would beneﬁt by considering both observables
-simultaneously. Future studies might develop an algorithm which incorporates both observables,
-
-again via ¢.- = H1/2(¢.' - E,.¢.-). In this case, a Lanczos algorithm would be based on the space
-
-Ic*(M,¢1,¢»2.p) 2 ran{«/»mz»2,M«/»1,M«/:2, . . .,MP—‘¢1.M"—‘«/22}
-
-The Lanczos coeﬁicients would then involve not only the autocovariances C, (¢,<(n),¢,-(n + m)),
-
-but also the cross-covariances C, (¢.~(n), ¢j(n + m)) (deﬁned in (2.6)).
+observables, $\phi_1$ and $\phi_2$, of particular interest, perhaps we would
+benefit from considering both observables simultaneously. Future studies might
+develop an algorithm which incorporates both observables, 
+again via $\psi_i = \Pi^{1/2}(\phi_i - \bE_{\pi}\phi_i)$. %% In this case, a
+The Lanczos algorithm would then be based on the space 
+\[
+\sK^*(M,\psi_1, \psi_2, p) =  \ran\{\psi_1, \psi_2, \M\psi_1, 
+\M\psi_2, \dots, \M^{p-1}\psi_1, \M^{p-1}\psi_2\}
+\]
+The Lanczos coefficients would then involve not only the autocovariances,
+%$\bC_\pi(\phi_i(n), \phi_i(n+m))$
+but also the cross-covariances 
+$\bC_\pi(\phi_i(n), \phi_j(n+m))$ (deﬁned in (2.6)).

### 2013-10-20 tex/LanczosProcedures.tex
finished Application and Conclusion Sections

@@
 \beta_1 = \left[\frac{\bC_\pi (\phi(n), \phi(n + 2))}{\Var_\pi(\phi(n))}
--\bigl( \frac{\bC_\pi (\phi(n), \phi(n + 1))}{\Var_\pi(\phi(n))}\bigr)^2
+-\left( \frac{\bC_\pi (\phi(n), \phi(n + 1))}{\Var_\pi(\phi(n))}\right)^2
 \right]^{1/2}.
@@
 are based on four recurrence relations, they become quite complicated as $j$ gets large. However,
-defining these recursions in a computer program is trivial, and we use this fact
-in Section~\label{sec:computations} % Sec 5.4
-when applying the foregoing to a particular problem.
+defining these recursions in a computer program is trivial, as we demonstrate
+in Section~\ref{sec:computations} % Sec 5.4
+by applying the foregoing to a particular problem.
 %

### 2013-10-20 tex/MarkovChains.tex
miscellaneous fixes
finished Application and Conclusion Sections

@@
 \section{General Theory}
-
+\label{sec:general-theory}
 This review of Markov chain theory can be found in any good probability text. The present
@@
 %% Definition 2.1.4
+\label{def:2.1.4}
 A measure $\pi$ is said to be a \emph{stationary measure} if

### 2013-10-20 tex/acknowledge.tex
finished Application and Conclusion Sections

@@
 Finally, I would like to thank my family, for their support of my interests in research (and
-everything else!), and especially my parents, Barbara and Ted Terry, and Benita and Bill De Meo.
-
-Without them, this paper would not have been written.
+everything else!), and especially my parents, Barbara and Ted Terry, and Benita and Bill DeMeo.
+Needless to say, without them this paper would not have been written.

### 2013-10-20 tex/Appendix.tex
finished Appendix section

@@
-% 42.txt
-Appendix: computer programs
+\singlespacing
 
-Z Hatlab code ising.m
-
-Z
-Z Written by William J. De Meo on 12/15/97
-
-2 last modified 1/10/98
-
-Z Inputs:
-
-2 d = number of nodes of the ising lattice (e.g. d=100)
-Z n = number of iterations (e.g. n=1000)
-
-Z beta = Annealing schedule (e.g. beta = 2,
-
-1 when beta -> 0 will always go to nee state
-Z vhen beta -> infty will never go to state of higher energy)
-Z
-
+{\small 
+\begin{verbatim}
+% Matlab code ising.m
+%
+% Written by William J. DeMeo on 12/15/97
+% last modified 2013.10.19
+% Inputs:
+%         d = number of nodes of the ising lattice (e.g. d=100)
+%         n = number of iterations (e.g. n=1000)
+%         beta = Annealing schedule (e.g. beta = 2,
+%                when beta -> 0 will always go to nee state
+%                when beta -> infty will never go to state of higher energy)
+%
 X = zeros(d,1);
 Energy = zeros(n,1);
-
 Ave=0;
 
-X Initialize using hot start
+% Initialize using hot start
 for i=1:d,
-
   if rand < .5
-
-x(i) = -1;
-
+    X(i) = -1;
   else
-
     X(i) = 1;
-
   end
@@
 
-Z Initialize at state of high energy (not used)
-
+% Initialize at state of high energy (not used)
 if 0
@@
   X(2:2:d)=—1;
-and
+end
 
-Z Initialize energy
-
+% Initialize energy
 H=0;
-
 for i=1:(d-1)
-
   H = H - X(i)*X(i+1);
@@
 
-2 Initialize observable
+% Initialize observable
 phi = zeros(n+1,1);
@@
   end
-
 end
 
-42
-
-% 43.txt
-43
-
 for j=1:n
-
   site=0;
-while site==
-
-site = round(rand#d);
+  while site==0
+    site = round(rand*d);
   end
-Z Compute new energy (with new X(site) = —x(site)):
-if site ==
-
-Hnew = H + 2:x(1)~x(2);
+  % Compute new energy (with new X(site) = -X(site)):
+  if site == 1
+    Hnew = H + 2*X(1)*X(2);
   elseif site == d
-
-Hnew = H + 2*X(d)tX(d-1);
+    Hnew = H + 2*X(d)*X(d-1);
   else
-
     Hnew = H + 2*X(site)*(X(site-1) + X(site+1));
@@
 
-Z Change to new state in two ways:
-X with probability 1/2, flip the switch
+  % Change to new state in two ways:
+  % with probability 1/2, flip the switch
   if rand < .5
-if (Hnew <= H) I (rand < exp(-beta*(Hnew-H)))
-x(site) = -x(site);
+    if (Hnew <= H) | (rand < exp(-beta*(Hnew-H)))
+      X(site) = -X(site);
       H = Hnew;
-Z Compute new value of observable
-if X(site) -= 1
+      % Compute new value of observable
+      if X(site) == 1
         phi(j+1) = phi(j) + 1;
-
       else
-phi(j+1) = phi(j) — 1;
+        phi(j+1) = phi(j) -1;
       end
@@
 end
-p1ot(Energy)
+plot(Energy)
 
@@
 for i=1:n+1
-
   phi2(i) = phi(i)*phi(i);
@@
 
-Z Write first observable to phi.dat:
+% Write first observable to phi.dat:
+fid = fopen('phi.dat','w');
+fprintf(fid,'%f\n',phi);
+fclose(fid);
 
-fid = fopen(’/quean.a/masters/deme5025/research/nyuthesis/programs/c/phi.dat’,’w’);
-fprintf(fid,’%f\n’,phi);
+% Write second observable to phi2.dat:
+fid = fopen('phi2.dat','w');
+fprintf(fid,'%f\n',phi2);
 fc1ose(fid);
 
-Z Write second observable to phi2.dat:
-fid = fopen(’/quean.a/masters/deme5025/research/nyuthesis/programs/c/phi2.dat’,’w’);
-
-% 44.txt
-fprintf(fid,’%f\n’,phi2);
-fc1ose(fid);
-
-1 Compute a few Lanczos coefficients by new method
+% Compute a few Lanczos coefficients by new method
 START=1001;
-
 V1=zeros(2);
-
 C1=zeros(2);
-
 C2=zeros(2);
-
 C3=zeros(2);
@@
 V1 = cov(phi(START:n+1),phi(START:n+1));
-
-V1 = V1(1,1)5
-
+V1 = V1(1,1);
 C1 = cov(phi(START:n),phi(START+1:n+1));
 C1 = C1(1,2);
-
 C2 = cov(phi(START:n-1),phi(START+2:n+1));
 C2 = C2(1,2);
-
 C3 = cov(phi(START:n-2),phi(START+3:n+1));
-
 C3 = C3(1,2);
 
-a1pha1 = C1/V1;
+alpha1 = C1/V1;
+beta1 = sqrt(C2/V1 -(C1/V1)^2);
+alpha2 = (1/(beta1^2))*(C3/V1 - 2*alpha1*(C2/V1) + alpha1^3);
 
-betai = sqrt(C2/V1 ' (C1/V1)‘2);
+T1 = [alphal beta1; beta1 alpha2]
+V1 = cov(phi2(START:n+1),phi2(START:n+1));
+V1 = V1(1,1);
+C1 = cov(phi2(START:n),phi2(START+1:n+1));
+C1 = C1(1,2);
+C2 = cov(phi2(START:n-1),phi2(START+2:n+1));
+C2 = C2(1,2);
+C3 = cov(phi2(START:n-2),phi2(START+3:n+1));
+C3 = C3(1,2);
 
-a1pha2 = (1/(beta1’2))*(C3/V1 - 2*a1pha1*(C2/V1) + a1pha1‘3);
+alpha1 = C1/V1;
+beta1 = sqrt(C2/V1 - (C1/V1)^2);
+alpha2 = (1/(beta1^2))#(C3/V1 - 2*alpha1*(C2/V1) + alpha1^3);
 
-T1 = Ealphal betal; betal a1pha2]
+T2 = [alpha1 beta1; beta1 alpha2]
 
-v1 = cov(phi2(START:n+1),phi2(START:n+1));
-v1 = v1(1,1);
+fid = fopen('Tmats.dat','v');
+fprintf(fid,'%f\n',T1);
+fprintf(fid,'%f\n',T2);
+fclose(fid);
 
-c1 = cov(phi2(START:n),phi2(START+1:n+1));
-c1 = c1(1,2);
+%%% end ising.m
 
-c2 = cov(phi2(START:n—1),phi2(START+2:n+1));
-c2 = c2(1,2);
+\end{verbatim}}
 
-c3 = cov(phi2(START:n-2),phi2(START+3:n+1));
-c3 = c3(1,2);
+\vskip2cm
+%
+%
+% -------------- 45.txt ------------------------------------------
+%
+%
+{\small 
+\begin{verbatim}
+% Matlab code tpm.m for generating transition probability matrix
+% and computing its eigenvalues directly
+%
+% Written by William J. DeMeo on 1/10/98
+%
+% Inputs:
+%        d = number of nodes of the ising lattice (e.g. d=10)
+%     beta = Annealing schedule (e.g. beta - 2,
+%            when beta -> 0 will always go to new state
+%            when beta -> infty will never go to state of higher energy)
 
-alphal = C1/V1;
-betal = sqrt(C2/V1 - (C1/V1)‘2);
-a1pha2 = (1/(beta1‘2))#(C3/V1 - 2ta1pha1*(C2/V1) + a1pha1”3);
+disp('bui1ding proposition matrix...')
 
-T2 = [a1pha1 betai; betal a1pha2]
-
-fid = fopen(’Tmats.dat’,’v’);
-fprintf(fid,’%f\n',T1);
-fprintf(fid,’%f\n’,T2);
-fc1ose(fid);
-
-ZZZ end ising.m
-
-44
-% 45.txt
-\% Matlab code tpm.m for generating transition probability matrix
-Z and computing its eigenvalues directly
-
-1
-
-Z Written by Hilliam J. De Heo on 1/10/98
-
-1
-
-Z Inputs:
-
-Z d = number of nodes of the ising lattice (e.g. d=10)
-
-Z beta = Annealing schedule (e.g. beta - 2,
-
-1 when beta -> 0 will always go to new state
-1 when beta -> infty will never go to state of higher energy)
-2
-
-disp(’bui1ding proposition matrix...’)
-
-states = 2‘d;
-
+states = 2^d;
 A = 0;
-
 for i=0:(d-1)
-
-E=eye(2”i);
-
-A = [A E;E A];
-
+  E=eye(2^i);
+  A = [A E; E A];
 end
 
-B = .5#eye(states);
+B = .5*eye(states);
+A = B + .5*(1/d)*A;
+% our modified Glauber dynamics requires the B and the .5*(1/d)
 
-A = B + .5-(1/d)*A;
+disp('...done')
 
-2 our modified Glauber dynamics requires the B and the .5¥(1/d)
+% display pattern of nonzero entries
+% spy(A)
 
-disp(’...done’)
-
-1 display pattern of nonzero entries
-1 spy(A)
-
-Z check that all row sums are 1
+% check that all row sums are 1
 check=O;
 for i=1:states
-
   check = check + (not(sum(A(i,1:states))<.99));
 end
-1 if not all rows sum to 1, print check = (# of rows with sum=1)
+% if not all rows sum to 1, print check = (# of rows with sum=1)
 if not(check==states)
-
   check
-
-error(’Row sums are not all 1’)
+  error('Row sums are not all 1')
 end
 
-Z construct matrix of states
+% construct matrix of states
 E = zeros(d,states);
-f1ip=-1;
+flip=-1;
 for i=1:d
-for j=1:2‘(i-1):states
-flip = -1*f1ip;
-for k=0:2“(i-1)
-
-45
-
-% 46.txt
+  for j=1:2^(i-1):states
+    flip = -1*flip;
+    for k=0:2^(i-1)
       E(i,j+k) = flip;
-and
       end
   end
-E=E’;
+end
+E=E';
 
-Z display first 64 states
-Z E(1:64,:)
+% display first 64 states
+% E(1:64,:)
 
-Z compute energy of each state
+% compute energy of each state
 H = zeros(states,1);
 for i=1:states
-
   for j=1:d-1
-H(i) = H(i) - E(i,j)tE(i,j+1);
+    H(i) = H(i) - E(i,j)*E(i,j+1);
   end
@@
 
-disp(’building Hetropolis transition matrix...')
+disp('building Hetropolis transition matrix...')
 
@@
       if(H(j)>H(i))
-alt = A(i,j)t(1-exp(-beta*(H(j)-H(i))));
-A(i,i) I A(i,i)+a1t;
+        alt = A(i,j)*(1-exp(-beta*(H(j)-H(i))));
+        A(i,i) = A(i,i)+a1t;
         A(i,j) = A(i,j)-alt;
       elseif(H(j)<H(i))
-alt = A(j,i)t(1-exp(—beta#(H(i)-H(j))));
+        alt = A(j,i)*(1-exp(-beta*(H(i)-H(j))));
         A(j,j) = A(j,j)+a1t:
@@
 
-disp(’...done’)
+disp('...done')
 
@@
 for i=1:states
-
   check = check + (not(sum(A(i,1:states))<.99));
@@
 
-Z if not all rows sum to 1, print check = (# of rows with aum=1)
+% if not all rows sum to 1, print check = (# of rows with aum=1)
 
 if not(check==states)
-
   check
-
-error(’Row sums are not all 1’)
+  error('Row sums are not all 1')
 end
 
-46
+disp('computing eigenvalues of tpm...')
+cput = cputime;
+evals = eig(A);
+ecput = tputime - cput;
+disp('the CPU time (in secs) for computing eigenvalues of tpm: ')
+ecput
 
-% 47.txt
-47
+%%% end tpm.m
+\end{verbatim}}
 
-disp(’computing eigenvalues of tpm...’)
-cput = cputime;
+\vskip2cm
 
-evals = eig(A);
+% 50.txt
 
-ecput = tputime - cput;
+{\small 
+\begin{verbatim}
 
-disp(’the CPU time (in secs) for computing eigenvalues of tpm: ’)
+/*************************************************************
+ * lanczos.c main program for computing Lanczos coefficients *
+ *                                                           *
+ * Created by William J. DeMeo on 1/7/98                     *
+ * Last modified 2013.10.19                                  *
+ ************************************************************/
 
-ecput
-ZZZ end tpm.m
-
-/¥#*##¥###*##*#l####¥#‘###$ﬂ###ﬁi##*U3*#**#*##**##¥***ll¥#*##¥ii
-
-* 1anczos.c main program for computing Lanczos coefficients *
-3 #
-* Created by William J. De Heo #
-t on 1/7/98 *
-# Last modified 1/10/98 #
-
-*¥#######**3It¥*#¥¥¥#**##**¥##¥#####**####*#*###*##**###¥##*##*/
-
-#inc1ude <stdlib.h>
+#include <stdlib.h>
 #include <stdio.h>
 #include <math.h>
-#inc1ude "prototypes.h"
+#include "prototypes.h"
 
-/# Machine constants on dino (Sun Ultra 1 at Courant) t/
-#define HACHEPS 1.15828708535533604981e-16
-
+/* Machine constants on dino (Sun Ultra 1 at Courant) */
+#define MACHEPS 1.15828708535533604981e-16
 #define SQRTEPS 1.07623746699106147048e-08
-
-ﬂdefine HAX_NAH 100
+#define MAX_NAME 100
 
@@
 
-/* Prototypes from the file functions.c ¥/
+/* Prototypes from the file functions.c */
+double alpha(int j);
+double betasq(int j);
+double form1(int j, int k);
+double form2(int j, int k);
+double moment(double *data, long n, double *ave, double *var, double *cov, long k);
 
-double a1pha(int j);
+/* external variables to be used by functions */
+double *phi,*cov, *var, *ave;
 
-double betasq(int j);
-
-double form1(int j, int k);
-
-double form2(int j, int k);
-
-double moment(doub1e ‘data, long n, double tave, double tvar, double tcov, long k);
-
-/# external variables to be used by functions */
-double *phi,tcov, tvar, tave;
-
-% 48.txt
 main()
-
 {
 
-char tfilename;
+  char *filename;
+  FILE *ofp;
+  double temp=O, *T;
+  long i, j, nrow, nlanc, START;
+  int flag=O;
 
-FILE *ofp;
+  filename = cmalloc(MAX_NAME);
 
-double temp=O, *T;
-
-long i, j, nrou, nlanc, START;
-int f1ag=O;
-
-filename = cma11oc(HAx_NAHE);
-
-printf("\nName of file containing observed values: “);
-
-read_name(fi1ename);
-
-printf(“\nTota1 number of observations in file (iterations):
-
-scanf(“Zu”,&nrov);
-
-printf(“\nNumber of leading observations to discard: “);
-
-scanf("Zu",&START);
-
-printf("\nNumber of Lanczos coefficients desired: “);
-
-scanf("%u",&n1anc);
-
-vhi1e(2*n1anc > nrow—START)
-
+  printf("\nName of file containing observed values: ");
+  read_name(filename);
+  printf("\nTota1 number of observations in file (iterations): ");
+  scanf("%u",&nrow);
+  printf("\nNumber of leading observations to discard: ");
+  scanf("%u",&START);
+  printf("\nNumber of Lanczos coefficients desired: ");
+  scanf("%u",&nlanc);
+  while(2*nlanc > nrow-START)
     {
-printf("\nNot enough data for that many coefficients.\n“);
-printf("\nEnter a smaller number of Lanczos coefficients: “);
-scanf("Xu“,&n1anc);
-
+      printf("\nNot enough data for that many coefficients.\n");
+      printf("\nEnter a smaller number of Lanczos coefficients: ");
+      scanf("%u",&nlanc);
     }
 
-II);
+  phi = dmalloc(nrow);
+  cov=dmalloc(2*nlanc);
+  var=dmalloc(1);
+  ave=dmalloc(1);
+  T = dmalloc(nlanc*nlanc); 
+  for(i=0;i<nlanc;i++) 
+    T[i]=(double) 0;
 
-phi = dma11oc(nrow);
-cov=dma11oc(2#n1anc);
-var=dma11oc(1);
-ave=dma11oc(1);
-
-T = dma11oc(n1anc#n1anc); for(i=0;i<n1anc;i++) T[i]=(doub1e)0;
-
-/* observable phi is stored contiguously column-wise (e.g. by HATLAB) t/
-mat1abread(phi, nrow, 1,fi1ename);
+  /* observable phi is stored contiguously column-wise (e.g. by MATLAB) */
+  matlabread(phi, nrow, 1, filename);
 
   /* send the observable, offset by START */
-moment(phi+START,nrou-START,ave, var, cov,2*n1anc-1);
+  moment(phi+START,nrow-START,ave, var, cov,2*nlanc-1);
 
   /* First column of T */
-
-T[0] = a1pha(1);
-if((temp=betasq(1))>HACHEPS*10)
+  T[0] = alpha(1);
+  if((temp=betasq(1))>MACHEPS*10)
     {
-
       T[1]=sqrt(temp);
@@
       /* General column of T */
-
-J'=2:
-
-for(i=1;i<n1anc-1, f1ag==O;)
+      j=2:
+      for(i=1; i<nlanc-1, flag==O;)
 	{
-
-T[itn1anc+i-1]=sqrt(betasq(i));
-
-48
-
-main()
-
+	  T[i*nlanc+i-1]=sqrt(betasq(i));
+	  T[i*nlanc+i] = alpha(i+1);
+	  if((temp = betasq(i+1))>MACHEPS*10)
 	    {
-
-char tfilename;
-
-FILE tofp;
-
-double temp=0, #T;
-
-long i, j, nrow, nlanc, START;
-int f1ag=0;
-
-filename = cma11oc(HAx_NAH);
-
-princf("\nName of file containing observed values: “);
-read_name(fi1ename);
-
-printf("\nTota1 number of observations in file (iterations): “);
-scanf("2u“,&nrow);
-
-printf("\nNumber of leading observations to discard: “);
-scanf("Zu",&START);
-
-printf("\nNumber of Lanczos coefficients desired: “);
-scanf("%u“,&n1anc);
-
-whi1e(2*n1anc > nrou-START)
-
-{
-printf("\nNot enough data for that many coefficients.\n");
-
-printf("\nEnter a smaller number of Lanczos coefficients: “);
-scanf("Zu",&n1anc);
-}
-
-phi = dma11oc(urow);
-cov=dma1loc(2*n1anc);
-var=dma11oc(1);
-ave=dma11oc(1);
-
-T = dma11oc(n1anc*n1anc); for(i=0;i<n1anc;i++) T[i]=(doub1e)O;
-
-/* observable phi is stored contiguously column-wise (e.g. by HATLAB) */
-mat1abread(phi, nrov, 1,fi1ename);
-
-/* send the observable, offset by START #/
-moment(phi+START,nrov-START,ave, var, cov,2*n1anc-1);
-
-/# First column of T */
-
-T[0] = a1pha(1);
-if((temp=betasq(1))>HACHEPS#10)
-{
-
-T[1]=sqrt(temp);
-
-/* General column of T */
-
-J‘=2:
-
-for(i-1;i<n1anc-1, f1ag==0;)
-{
-
-T[i*nlanc+i-1]=sqrt(betasq(i));
-
-48
-
-% 49.txt
-T[i*n1anc+i] = a1pha(i+1);
-if((temp = betasq(i+1))>HACmEPS#10)
-{
-
-T[i#n1anc+i+1] = sqrt(temp);
-
+	      T[i*nlanc+i+1] = sqrt(temp);
 	      i++;
@@
 	  else
-
-f1ag=1;
-
+	    flag=1;
 	}
 
-/t Last column */
-if(f1ag!=1 it ((temp= betasq(n1anc-1))>HACHEPSt10))
-
+      /* Last column */
+      if(flag!=1 && ((temp= betasq(nlanc-1))>MACHEPS*10))
 	{
-T[n1anc#n1anc-2] = sqrt(temp);
-T[n1anctn1anc-1] = a1pha(n1anc);
-printf("\nbeta(1) = %1f\nbeta(Zd) = Zlf (last beta)",
-T[1],(nlanc-1),T[nlanc*n1anc-2]);
+	  T[nlanc*nlanc-2] = sqrt(temp);
+	  T[nlanc*nlanc-1] = alpha(nlanc);
+	  printf("\nbeta(1) = %1f\nbeta(%d) = %lf (last beta)",
+		 T[1],(nlanc-1),T[nlanc*nlanc-2]);
 	}
@@
 	  printf("\nApproximate invariant space reached at step %d.",i);
-printf("\nbeta(1) = Z1f\nbeta(Zd) = Zlf (last accurate beta)“.
-T[1],i,T[itn1anc+i—1]);
-printf("\nbeta(Zd)“2: Zlf (first spurious resu1t)“,i+1,temp);
+	  printf("\nbeta(1) = %lf\nbeta(%d) = %lf (last accurate beta)".
+		 T[1],i,T[i*nlanc+i-1]);
+	  printf("\nbeta(%d)^2: %lf (first spurious resu1t)",i+1,temp);
+	}
+      printf("\nThe matrix T is: \n");
+      matprint(T,nlanc,nlanc);
+    }
+  else
+    {
+      printf("\nmain(): Approximate invariant space reached at first step.");
+      printf("\nalpha(1) - %1f\nbeta(1)^2: %lf (first spurious result)", T[O], temp);
+    }
+  ofp = fopen("Tmat.m","v");
+  check(ofp);
+  matlabwrite(T,nlanc,nlanc,ofp);
+  fclose(ofp);
 }
 
-printf("\nThe matrix T is: \n");
-matprint(T,n1anc,n1anc);
-
+void read_name(char *name)
+{
+  int c, i = 0;
+  while ((c = getchar()) != EOF && c != ' ' && c != '\n')
+    name[i++] = c;
+  name[i] = '\0';
 }
 
-else
+/** end lanczos.c **/
 
-{
+\end{verbatim}}
 
-printf("\nmain(): Approximate invariant space reached at first step.“);
+\vskip2cm
 
-49
+{\small 
+\begin{verbatim}
 
-printf("\na1pha(1) - Z1f\nbeta(1)”2: Zlf (first spurious result)“, T[O], temp);
+/***************************************************
+ * functions.c -- functions required by lanczos.c  *
+ *                                                 *
+ * Created by William J. DeMeo on 1/7/98           *
+ * Last modified 2013/10/19                        *
+ ***************************************************/
 
-}
-ofp = fopen("Tmat.m","v“);
-check(ofp);
-mat1aburite(T,n1anc,n1anc,ofp);
-fc1ose(ofp);
-}
-void read_name(char tname)
-{
-int c, i = 0;
-while ((c = getchar()) != EOF it c != ’ ’ it c !- ’\n’)
-name[i++] = c;
-name[i] = ’\0’;
-}
-
-/#t* end 1anczos.c *#*/
-
-% 50.txt
-50
-
-/llnkttttttttinlntint*1!!!##1##!ttitilltit###t¥#tt¥ttItt####t#ttII=##mlltttt
-
-* functions.c functions required by 1anczos.c t
-# t
-* Created by Hilliam J. De Meo #
-* on 1/7/98 *
-t Last modified 1/10/98 :
-
-¥¥##¥$¥##¥¥#$¥¥¥###¥¥$*I#****#**####*######¥#¥#¥#$#¥¥#l¥**#****/
-
-#inc1ude <math.h>
-
+#include <math.h>
 #define START 1000
-
 #define ITER 10000
 
-/t Machine constants on dino (Sun Ultra 1 at Courant) */
-
-#define HACHEPS 1.15828708535533604981e-16
-
+/* Machine constants on dino (Sun Ultra 1 at Courant) */
+#define MACHEPS 1.15828708535533604981e-16
 #define SQRTEPS 1.07623746699106147048e-O8
 
-double a1pha(int j);
+double alpha(int j);
+double betasq(int j);
+double form1(int j, int k);
+double form2(int j, int k);
+double moment(double *data, long n, double *ave, double *var, double *cov, int k);
 
-double betasq(int j);
-
-double form1(int j, int k);
-
-double form2(int j, int k);
-
-double moment(doub1e tdata, long n, double tave, double tvar, double tcov, int k);
-
-/t external variables to be used by functions */
-double *phi,tcov, *var, tave;
+/* external variables to be used by functions */
+double *phi,*cov, *var, *ave;
 
@@
 {
-
   /* alpha is never called with j < 1 */
   if(j==1)
-return
-form1(1,1);
+    return form1(1,1);
   else if(j>1)
     return
-((doub1e)1/betasq(j-1))*
-(form1(j-1,3) — pou(a1pha(j-1),3)
-- 2 * (alpha(j-1)#betasq(j-1) + sqrt(betasq(j-2))#form2(j-1,2))
+      ((double)1/betasq(j-1)) * (form1(j-1,3) - pow(alpha(j-1),3)
+       - 2 * (alpha(j-1)*betasq(j-1) + sqrt(betasq(j-2))*form2(j-1,2))
        + betasq(j-2));
@@
 double betasq(int j)
-
 {
   if(j==0)
-return (doub1e)O;
+    return (double)O;
   else if(j>0)
-return (form1(j,2) - pow(a1pha(j),2) - betasq(j-1));
+    return (form1(j,2) - pow(alpha(j),2) - betasq(j-1));
 }
@@
 {
-
-% 51.txt
-51
-
-double form14,alphal,alpha1sq,form12,form13;
+  double form14, alpha1, alpha1sq, form12, form13;
   if(j==0)
-
-return (doub1e)0;
+    return (double)0;
   else if(j==1)
     {
-
-/# printf("\nvar = %1f, cov(%d) = %1f \n“,*var,k,cov[k]);t/
-
+      /* printf("\nvar = %lf, cov(%d) = %lf \n",*var,k,cov[k]); */
       return (cov[k])/(*var); /* the only real value */
@@
     {
-
-return ((double)1/betasq(j-1))
-
-* ( form1(j-1,k+2)
-+ pow(a1pha(j-1),2)*form1(j-1,k)
-+ 2t(a1pha(j-1) t sqrt(betasq(j-2)) * form2(j-1,k)
-- a1pha(j-1)*form1(j-1,k+1)
-- sqrt(betasq(j—2))#form2(j-1,k+1) )
+      return 
+	((double)1/betasq(j-1))
+	* ( form1(j-1,k+2) + pow(alpha(j-1),2) * form1(j-1,k)
+	    + 2*(alpha(j-1) * sqrt(betasq(j-2)) * form2(j-1,k)
+		 - alpha(j-1)*form1(j-1,k+1)
+		 - sqrt(betasq(j—2))*form2(j-1,k+1) )
 	    + betasq(j-2)*form1(j-2,k)):
@@
   if(j==1)
-return (doub1e)0;
+    return (double)0;
   else if(j>1)
     return
-(pov(betasq(j-1),—.5)) 3
-(form1(j-1,k+1) - a1pha(j-1)*form1(j-1,k)
-- sqrt(betasq(j-2)) # form2(j—1,k));
+      (pov(betasq(j-1),-.5)) 3
+      (form1(j-1,k+1) - alpha(j-1)*form1(j-1,k)
+       - sqrt(betasq(j-2)) * form2(j—1,k));
 }
@@
 */
-double moment(double tdata, long n, double tave, double tvar, double tcov, int k)
+double moment(double *data, long n, double *ave, double *var, double *cov, int k)
 {
-/* Centered about data[0] algorithm: #/
+  /* Centered about data[0] algorithm: */
+  long i,j;
+  double ave1, ave2;
 
-long i,j;
-double avel, ave2;
-
-% 52.txt
-52
-
-*ave=O;tvar=0; ave1=ave2=O;
+  *ave=O;*var=0; ave1=ave2=O;
   for(j=0;j<=k;j++)
@@
     {
-rave += (data[i] - data[0]);
-tvar += (data[i] - data[0])#(data[i] - data[0]);
+      *ave += (data[i] - data[0]);
+      *var += (data[i] - data[0])*(data[i] - data[0]);
     }
-tvar /= (doub1e)(n-1);
-Ivar -= (((*ave)/(doub1e)n) * ‘(*ave)/(doub1e)(n-1)));
-/# *ave = ((*ave)/(doub1e)n) + data[0]; <- that’s the true average (not needed)*/
+  *var /= (double)(n-1);
+  *var -= (((*ave)/(double)n) * ((*ave)/(double)(n-1)));
+  /* *ave = ((*ave)/(double)n) + data[0]; <- that's the true average (not needed)*/
 
@@
     {
-
       for(i=0;i<(n-j);i++)
-
 	{
-
-avel += (data[i] - data[0]);
+	  ave1 += (data[i] - data[0]);
 	  ave2 += (data[i+j] - data[0]);
-cov[j] += (data[i] - data[0])*(data[i+j] — data[0]);
+	  cov[j] += (data[i] - data[0])*(data[i+j] - data[0]);
 
@@
 
-avel/=(doub1e)(n-j); ave2/=(doub1e)(n-j);
-
-cov[j] = ((cov[j] - (double)(n-j)tave1¢ave2)/(double)(n-j-1));
+      ave1/=(double)(n-j); ave2/=(double)(n-j);
+      cov[j] = ((cov[j] - (double)(n-j)*ave1*ave2)/(double)(n-j-1));
     }
@@
 
-/trt end funccions.c ###/
+/*** end funccions.c ***/
+
+\end{verbatim}}

### 2013-10-20 tex/Conclusion.tex
finished Appendix section

@@
 We are interested in the subdominant eigenvalue of a large transition probability matrix. Our
-method can be used to approximate these eigenvalues when the following conditions are satisﬁed:
+method can be used to approximate these eigenvalues when the following conditions are satisfied:
 \begin{itemize}
@@
 %$\bC_\pi(\phi_i(n), \phi_i(n+m))$
-but also the cross-covariances 
-$\bC_\pi(\phi_i(n), \phi_j(n+m))$ (deﬁned in (2.6)).
+but also the cross-covariances, 
+$\bC_\pi(\phi_i(n), \phi_j(n+m))$, defined in (2.6) above.

### 2013-10-20 tex/Appendix.tex
final revisions before going to single file format

@@
-\singlespacing
+\begin{center}
+  {\Large Appendix: computer programs}
+\end{center}
 
@@
 double form2(int j, int k);
-double moment(double *data, long n, double *ave, double *var, double *cov, long k);
+double moment(double *data, long n, 
+              double *ave, double *var, double *cov, long k);
 
@@
 
-  /* observable phi is stored contiguously column-wise (e.g. by MATLAB) */
+  /* observable phi is stored contiguously column-wise by MATLAB */
   matlabread(phi, nrow, 1, filename);
@@
       printf("\nmain(): Approximate invariant space reached at first step.");
-      printf("\nalpha(1) - %1f\nbeta(1)^2: %lf (first spurious result)", T[O], temp);
+      printf("\nalpha(1) - %lf\nbeta(1)^2: %lf (first spurious result)", 
+             T[O], temp);
     }
@@
 double form2(int j, int k);
-double moment(double *data, long n, double *ave, double *var, double *cov, int k);
+double moment(double *data, long n, 
+              double *ave, double *var, double *cov, int k);
 
@@
 		 - sqrt(betasq(j—2))*form2(j-1,k+1) )
-	    + betasq(j-2)*form1(j-2,k)):
+	    + betasq(j-2)*form1(j-2,k));
     }
@@
 */
-double moment(double *data, long n, double *ave, double *var, double *cov, int k)
+double moment(double *data, long n, 
+              double *ave, double *var, double *cov, int k)
 {
@@
   *var -= (((*ave)/(double)n) * ((*ave)/(double)(n-1)));
-  /* *ave = ((*ave)/(double)n) + data[0]; <- that's the true average (not needed)*/
+  /* *ave = ((*ave)/(double)n) + data[0]; (the true average; not needed)*/
 
@@
 
-\end{verbatim}}
+\end{verbatim}
+}

### 2013-10-20 tex/InvariantSubspaces.tex
miscellaneous fixes
final revisions before going to single file format

@@
 \chapter{Invariant and Approximate Invariant Subspaces}
-
+\label{cha:invar-appr-invar}
 \section{Invariant Subspaces}
@@
 %Proof of 3.1.2:
+~
+\vskip-2cm
   \begin{align*}
@@
 & \quad \Longleftrightarrow  \quad
-\Psi \T v = \lambda \Psi v
+\Psi \T v = \lambda \Psi v\\
 & \quad \Longleftrightarrow  \quad
-\M \Psi v = \lambda \Psi v
+\M \Psi v = \lambda \Psi v\\
 & \quad \Longleftrightarrow  \quad
@@
 To see why Facts~\ref{fact:3.1.1} and~\ref{fact:3.1.2} are theoretically useful,
-consider the equation of Fact~\ref{fact:3.1.1}:
-\begin{align*}
-& \Psi \T = \M \Psi \\
-\Longleftrightarrow \quad & \Psi^t\Psi \T = \Psi^t \M \Psi\\
-\Longleftrightarrow \quad & \T = (\Psi^t\Psi)^t \Psi^t \M\Psi.
-\end{align*}
+%consider the equation of Fact~\ref{fact:3.1.1}:
+%% \begin{align*}
+%% & \Psi \T = \M \Psi \\
+%% \Longleftrightarrow \quad & \Psi^t\Psi \T = \Psi^t \M \Psi\\
+%% \Longleftrightarrow \quad & \T = (\Psi^t\Psi)^t \Psi^t \M\Psi.
+%% \end{align*}
+%% \begin{equation*}
+%% \Psi \T = \M \Psi \quad\Longleftrightarrow \quad \Psi^t\Psi \T = \Psi^t \M \Psi
+%% \Longleftrightarrow \quad  \T = (\Psi^t\Psi)^t \Psi^t \M\Psi.
+%% \end{equation*}
+note that the equation of Fact~\ref{fact:3.1.1}, $\Psi \T = \M \Psi$, is
+equivalent to 
+$\Psi^t\Psi \T = \Psi^t \M \Psi$, which is equivalent to 
+$\T = (\Psi^t\Psi)^t \Psi^t \M\Psi$.
 In this form, we recognize that $\T = \C^{-1}(0)\C(1)$. That is, using only the
@@
 
-\section{Approximate Invariant Subspace}
+\section{Approximate Invariant Subspaces}
 % Sec 3.2
@@
 $\T = \Q^t\M\Q$ minimizes $\|\E\|_2 = \|\M\Q - \Q\T\|_2$ and
-\item $\exists \mu_1, \dots, \mu_p\in \lambda(\T)$ and 
+\item there exist $\mu_1, \dots, \mu_p\in \lambda(\T)$ and 
 $\lambda_1,\dots, \lambda_p \in \lambda(\M)$ such that
@@
 \[
-\nabla \rho(\Psi_ka)  = 
+\nabla \rho(\Psi_ka)   
  = \frac{2}{a^t\Psi_k^t\Psi_ka} (\M\Psi_k - \rho(\Psi_ka)\Psi_ka).

### 2013-10-20 tex/MarkovChains.tex
final revisions before going to single file format

@@
 \label{eq:1}
-\bP(X_{n+1} \in B \mid \sigma(X_0, \dots, X_{n}))
-=\bP(X_{n+1} \in B \mid \sigma(X_{n})).
+P(X_{n+1} \in B \mid \sigma(X_0, \dots, X_{n}))
+=P(X_{n+1} \in B \mid \sigma(X_{n})).
 \end{equation}
@@
 \label{eq:2.2}
-\bP(X_{n+1} \in B \mid  \sigma(X_n)) = p_n(X_n, B)
+P(X_{n+1} \in B \mid  \sigma(X_n)) = p_n(X_n, B)
 \end{equation}
@@
 \label{eq:2.3}
-\bP(X_j\in B_j ,0 \leq j \leq n) = \int_{B_0} \mu(dx_0) \int_{B_1} p_0(x_0, dx_1) 
-\int_{B_2} p_1(x_1, dx_2) \cdots 
+P(X_j\in B_j ,0 \leq j \leq n) = \int_{B_0} \mu(dx_0) \int_{B_1} p_0(x_0, dx_1) 
+%\int_{B_2} p_1(x_1, dx_2) 
+\cdots 
 \int_{B_n} p_{n-1}(x_{n-1}, dx_n).
@@
 \[
-(\Omega, \F) = (S^\omega,\S^\omega), \quad \text{where} \omega = \{0,l,...\}.
+(\Omega, \F) = (S^\omega,\S^\omega), \quad \text{ where } \omega = \{0,1,\dots\}.
 \]
@@
 \[
-S^\omega  = \{(\omega_0, \omega_1, \dots) : \omega_i in S\}\quad \text{ and }
+S^\omega  = \{(\omega_0, \omega_1, \dots) : \omega_i \in S\}\quad \text{ and }
 \quad
@@
 The Markov chain that we will study on this space is simply $X_n(\omega) = \omega$, the coordinate maps.
-Then, by the Kolmogorov extension theorem, there exists a unique probability measure $\bP_\mu$ on
+Then, by the Kolmogorov extension theorem, there exists a unique probability measure $P_\mu$ on
 $(\Omega, \F)$ so that the $X_n(\omega)$ have finite dimensional distributions~(\ref{eq:2.3}).
 
-If instead of $\mu$, we begin with the initial distribution $\delta_x$, i.e., point mass at $x$, then we
-denote the probability measure by $\bP_x$. With such measures defined for each $x$, we can in turn
-define distributions $\bP_\mu$, given any initial distribution $\mu$, by
+If instead of $\mu$ we begin with the initial distribution $\delta_x$, i.e., point mass at $x$, then we
+denote the probability measure by $P_x$. With such measures defined for each $x$, we can in turn
+define distributions $P_\mu$, given any initial distribution $\mu$, by
 \[
-\bP_\mu(A) = \int \mu(dx)\bP_x(A).
+P_\mu(A) = \int \mu(dx)P_x(A).
 \]
@@
 \[
-\bP(X_{n+1}=j \mid X_n = i) =p(i,j),
+P(X_{n+1}=j \mid X_n = i) =p(i,j),
 \]
@@
 
-Equation~(\ref{eq:2.4}) says $\bP_\pi(X_1 = y) = \pi(y)$, 
-and by induction that $\bP_\pi(X_n = y) = \pi(y)$ for all $n \geq 1$. 
+Equation~(\ref{eq:2.4}) says $P_\pi(X_1 = y) = \pi(y)$, 
+and by induction that $P_\pi(X_n = y) = \pi(y)$ for all $n \geq 1$. 
 If $\pi$ is a probability measure, then we call $\pi$ a stationary distribution. It
@@
 &= \bE_\pi[(\phi_i(X_n) - \bE_\pi\phi_i)(\phi_i(X_{n+s}) - \bE_\pi\phi_i)]\\
-&= \sum_{x, y\in S} \bP_\pi(X_n = x, X_{n+s} = y) (\phi_i(x)- \bE_\pi\phi_i)(\phi_i(y) - \bE_\pi\phi_i).\nonumber
+&= \sum_{x, y\in S} P_\pi(X_n = x, X_{n+s} = y) (\phi_i(x)- \bE_\pi\phi_i)(\phi_i(y) - \bE_\pi\phi_i).\nonumber
 \end{align}
@@
 \[
-\sum_{x, y\in S} \bP_\pi(X_n = x) \bP_\pi(X_{n+s} = y\mid X_n = x) (\phi_i(x)-
+\sum_{x, y\in S} P_\pi(X_n = x) P_\pi(X_{n+s} = y\mid X_n = x) (\phi_i(x)-
 \bE_\pi\phi_i)(\phi_i(y) - \bE_\pi\phi_i).

### 2013-10-20 src/functions.c
adding computer program source files

@@ -1,0 +1,130 @@
+/***************************************************
+ * functions.c -- functions required by lanczos.c  *
+ *                                                 *
+ * Created by William J. DeMeo on 1/7/98           *
+ * Last modified 2013/10/19                        *
+ ***************************************************/
+
+#include <math.h>
+#define START 1000
+#define ITER 10000
+
+/* Machine constants on dino (Sun Ultra 1 at Courant) */
+#define MACHEPS 1.15828708535533604981e-16
+#define SQRTEPS 1.07623746699106147048e-O8
+
+double alpha(int j);
+double betasq(int j);
+double form1(int j, int k);
+double form2(int j, int k);
+double moment(double *data, long n, 
+	      double *ave, double *var, double *cov, int k);
+
+/* external variables to be used by functions */
+double *phi,*cov, *var, *ave;
+
+double alpha(int j)
+{
+  /* alpha is never called with j < 1 */
+  if(j==1)
+    return form1(1,1);
+  else if(j>1)
+    return
+      ((double)1/betasq(j-1)) * (form1(j-1,3) - pow(alpha(j-1),3)
+       - 2 * (alpha(j-1)*betasq(j-1) + sqrt(betasq(j-2))*form2(j-1,2))
+       + betasq(j-2));
+}
+
+
+double betasq(int j)
+{
+  if(j==0)
+    return (double)O;
+  else if(j>0)
+    return (form1(j,2) - pow(alpha(j),2) - betasq(j-1));
+}
+
+
+double form1(int j, int k)
+{
+  double form14, alpha1, alpha1sq, form12, form13;
+  if(j==0)
+    return (double)0;
+  else if(j==1)
+    {
+      /* printf("\nvar = %lf, cov(%d) = %lf \n",*var,k,cov[k]); */
+      return (cov[k])/(*var); /* the only real value */
+    }
+  else if(j>1)
+    {
+      return 
+	((double)1/betasq(j-1))
+	* ( form1(j-1,k+2) + pow(alpha(j-1),2) * form1(j-1,k)
+	    + 2*(alpha(j-1) * sqrt(betasq(j-2)) * form2(j-1,k)
+		 - alpha(j-1)*form1(j-1,k+1)
+		 - sqrt(betasq(j—2))*form2(j-1,k+1) )
+	    + betasq(j-2)*form1(j-2,k)):
[…]

### 2013-10-20 src/ising.m

@@ -1,0 +1,137 @@
+% Matlab code ising.m
+%
+% Written by William J. DeMeo on 12/15/97
+% last modified 2013.10.19
+% Inputs:
+%         d = number of nodes of the ising lattice (e.g. d=100)
+%         n = number of iterations (e.g. n=1000)
+%         beta = Annealing schedule (e.g. beta = 2,
+%            when beta -> 0 will always go to nee state
+%            when beta -> infty will never go to state of higher energy)
+%
+X = zeros(d,1);
+Energy = zeros(n,1);
+Ave=0;
+
+% Initialize using hot start
+for i=1:d,
+  if rand < .5
+    X(i) = -1;
+  else
+    X(i) = 1;
+  end
+end
+
+% Initialize at state of high energy (not used)
+if 0
+  X(1:2:d-1)=1;
+  X(2:2:d)=—1;
+end
+
+% Initialize energy
+H=0;
+for i=1:(d-1)
+  H = H - X(i)*X(i+1);
+end
+
+% Initialize observable
+phi = zeros(n+1,1);
+
+for i=1:d
+  if X(i) == 1
+    phi(1)= phi(1)+1;
+  end
+end
+
+for j=1:n
+  site=0;
+  while site==0
+    site = round(rand*d);
+  end
+  % Compute new energy (with new X(site) = -X(site)):
+  if site == 1
+    Hnew = H + 2*X(1)*X(2);
+  elseif site == d
+    Hnew = H + 2*X(d)*X(d-1);
+  else
+    Hnew = H + 2*X(site)*(X(site-1) + X(site+1));
+  end
+
+  % Change to new state in two ways:
+  % with probability 1/2, flip the switch
+  if rand < .5
+    if (Hnew <= H) | (rand < exp(-beta*(Hnew-H)))
+      X(site) = -X(site);
+      H = Hnew;
+      % Compute new value of observable
+      if X(site) == 1
+        phi(j+1) = phi(j) + 1;
+      else
+        phi(j+1) = phi(j) -1;
+      end
+    else phi(j+1)=phi(j);
+    end
+  else phi(j+1)=phi(j);
+  end
+  Energy(j) = H;
[…]

### 2013-10-20 src/lanczos.c

@@ -1,0 +1,130 @@
+/*************************************************************
+ * lanczos.c main program for computing Lanczos coefficients *
+ *                                                           *
+ * Created by William J. DeMeo on 1/7/98                     *
+ * Last modified 2013.10.19                                  *
+ ************************************************************/
+
+#include <stdlib.h>
+#include <stdio.h>
+#include <math.h>
+#include "prototypes.h"
+
+/* Machine constants on dino (Sun Ultra 1 at Courant) */
+#define MACHEPS 1.15828708535533604981e-16
+#define SQRTEPS 1.07623746699106147048e-08
+#define MAX_NAME 100
+
+void read_name(char *);
+
+/* Prototypes from the file functions.c */
+double alpha(int j);
+double betasq(int j);
+double form1(int j, int k);
+double form2(int j, int k);
+double moment(double *data, long n, 
+	      double *ave, double *var, double *cov, long k);
+
+/* external variables to be used by functions */
+double *phi,*cov, *var, *ave;
+
+main()
+{
+
+  char *filename;
+  FILE *ofp;
+  double temp=O, *T;
+  long i, j, nrow, nlanc, START;
+  int flag=O;
+
+  filename = cmalloc(MAX_NAME);
+
+  printf("\nName of file containing observed values: ");
+  read_name(filename);
+  printf("\nTota1 number of observations in file (iterations): ");
+  scanf("%u",&nrow);
+  printf("\nNumber of leading observations to discard: ");
+  scanf("%u",&START);
+  printf("\nNumber of Lanczos coefficients desired: ");
+  scanf("%u",&nlanc);
+  while(2*nlanc > nrow-START)
+    {
+      printf("\nNot enough data for that many coefficients.\n");
+      printf("\nEnter a smaller number of Lanczos coefficients: ");
+      scanf("%u",&nlanc);
+    }
+
+  phi = dmalloc(nrow);
+  cov=dmalloc(2*nlanc);
+  var=dmalloc(1);
+  ave=dmalloc(1);
+  T = dmalloc(nlanc*nlanc); 
+  for(i=0;i<nlanc;i++) 
+    T[i]=(double) 0;
+
[…]

### 2013-10-20 src/tpm.m

@@ -1,0 +1,104 @@
+% Matlab code tpm.m for generating transition probability matrix
+% and computing its eigenvalues directly
+%
+% Written by William J. DeMeo on 1/10/98
+%
+% Inputs:
+%        d = number of nodes of the ising lattice (e.g. d=10)
+%     beta = Annealing schedule (e.g. beta - 2,
+%            when beta -> 0 will always go to new state
+%            when beta -> infty will never go to state of higher energy)
+
+disp('bui1ding proposition matrix...')
+
+states = 2^d;
+A = 0;
+for i=0:(d-1)
+  E=eye(2^i);
+  A = [A E; E A];
+end
+
+B = .5*eye(states);
+A = B + .5*(1/d)*A;
+% our modified Glauber dynamics requires the B and the .5*(1/d)
+
+disp('...done')
+
+% display pattern of nonzero entries
+% spy(A)
+
+% check that all row sums are 1
+check=O;
+for i=1:states
+  check = check + (not(sum(A(i,1:states))<.99));
+end
+% if not all rows sum to 1, print check = (# of rows with sum=1)
+if not(check==states)
+  check
+  error('Row sums are not all 1')
+end
+
+% construct matrix of states
+E = zeros(d,states);
+flip=-1;
+for i=1:d
+  for j=1:2^(i-1):states
+    flip = -1*flip;
+    for k=0:2^(i-1)
+      E(i,j+k) = flip;
+      end
+  end
+end
+E=E';
+
+% display first 64 states
+% E(1:64,:)
+
+% compute energy of each state
+H = zeros(states,1);
+for i=1:states
+  for j=1:d-1
+    H(i) = H(i) - E(i,j)*E(i,j+1);
+  end
+end
+
+disp('building Hetropolis transition matrix...')
+
+for i=1:states
+  for j=i+1:states
+    if not(A(i,j)==0)
+      if(H(j)>H(i))
+        alt = A(i,j)*(1-exp(-beta*(H(j)-H(i))));
+        A(i,i) = A(i,i)+a1t;
+        A(i,j) = A(i,j)-alt;
+      elseif(H(j)<H(i))
+        alt = A(j,i)*(1-exp(-beta*(H(i)-H(j))));
+        A(j,j) = A(j,j)+a1t:
+        A(j,i) = A(j,i)-alt;
+      end
+    end
+  end
+end
+
+disp('...done')
+
+check=0;
+for i=1:states
[…]

### 2013-10-20 msthesis.tex
reorganizing

%% NYU PhD thesis format. Created by Jos Koiller 2007--2008.

%% Use the first of the following lines during production to
%% easily spot "overfull boxes" in the output. Use the second
%% line for the final version.
%\documentclass[12pt,draft,letterpaper]{report}
\documentclass[12pt,letterpaper]{report}

%% Replace the title, name, advisor name, graduation date and dedication below with
%% your own. Graduation months must be January, May or September.
\newcommand{\thesistitle}{A Lanczos Procedure for Approximating Eigenvalues of Large Stochastic Matrices}
\newcommand{\thesisauthor}{William J. DeMeo}
\newcommand{\thesisadvisor}{Professor Jonathan Goodman}
\newcommand{\graddate}{Jan 1999}

%\urladdr{williamdemeo@gmail.com}
%% \urladdr{http://williamdemeo.org}
%% \address{Department of Mathematics\\
%% University of South Carolina\\Columbia 29208\\USA}
%% If you do not want a dedication, scroll down and comment out
%% the appropriate lines in this file.
\newcommand{\thesisdedication}{To my parents, with love and appreciation.}

%% The following makes chapters and sections, but not subsections,
%% appear in the TOC (table of contents). Increase to 2 or 3 to
%% make subsections or subsubsections appear, respectively. It seems
%% to be usual to use the "1" setting, however.
\setcounter{tocdepth}{1}

%% Sectional units up to subsubsections are numbered. To number
%% subsections, but not subsubsections, decrease this counter to 2.
\setcounter{secnumdepth}{3}

%% Page layout (customized to letter paper and NYU requirements):
\setlength{\oddsidemargin}{.6in}
\setlength{\textwidth}{5.8in}
\setlength{\topmargin}{.1in}
\setlength{\headheight}{0in}
\setlength{\headsep}{0in}
\setlength{\textheight}{8.3in}
\setlength{\footskip}{.5in}

%% Use the following commands, if desired, during production.
%% Comment them out for final version.
%\usepackage{layout} % defines the \layout command, see below
%\setlength{\hoffset}{-.75in} % creates a large right margin for notes and \showlabels

%% Controls spacing between lines (\doublespacing, \onehalfspacing, etc.):
\usepackage{setspace}

%% Use the line below for official NYU version, which requires
%% double line spacing. For all other uses, this is unnecessary,
%% so the line can be commented out.
\doublespacing % requires package setspace, invoked above

%% Each of the following lines defines the \com command, which produces
%% a comment (notes for yourself, for instance) in the output file.
%% Example:    \com{this will appear as a comment in the output}
%% Choose (uncomment) only one of the three forms:
%\newcommand{\com}[1]{[/// {#1} ///]}       % between [/// and ///].
\newcommand{\com}[1]{\marginpar{\tiny #1}} % as (tiny) margin notes
%\newcommand{\com}[1]{}                     % suppress all comments.

%% This inputs your auxiliary file with \usepackage's and \newcommand's:
%% It is assumed that that file is called "definitions.tex".
%%
%% \input{wjd_masters.def}
\usepackage{enumerate,amsmath,amssymb,fancyhdr,mathrsfs,amsthm,url,stmaryrd}
\usepackage[colorlinks=true,urlcolor=black,linkcolor=black,citecolor=black]{hyperref}
\usepackage{relsize}
\usepackage{marvosym}
\usepackage{graphicx}
\usepackage{enumerate} 
\usepackage{tikz}
\usetikzlibrary{calc}
\usepackage{scalefnt}
\usepackage{verbatim}
\usepackage{color}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{ifthen}
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{prop}[theorem]{Proposition}
\newtheorem{assumption}[theorem]{Assumption}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{question}[theorem]{Question}
\newcounter{claim}
\newtheorem{claim}[claim]{Claim}
\newcounter{conjecture}
\newtheorem{conjecture}[conjecture]{Conjecture}
\newtheorem*{problem}{Problem}
\newtheorem*{fact}{Fact}
\newtheorem{case}{Case}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\newtheorem*{remarks}{Remarks}
\newtheorem*{notation}{Notation}
\numberwithin{theorem}{chapter}
\numberwithin{claim}{chapter}
\numberwithin{equation}{chapter}
\numberwithin{conjecture}{chapter}
\renewcommand\S{\ensuremath{\mathcal{S}}}
\newcommand\sK{\ensuremath{\mathcal{K}}}
\newcommand\F{\ensuremath{\mathcal{F}}}
\newcommand\R{\ensuremath{\mathbb{R}}}
\newcommand\bC{\ensuremath{\mathbf{C}}}

\newcommand\C{\ensuremath{\mathrm{C}}}
\renewcommand\P{\ensuremath{\mathrm{P}}}
\newcommand\bP{\ensuremath{\mathbf{P}}}
\newcommand\M{\ensuremath{\mathrm{M}}}
\renewcommand\H{\ensuremath{\mathrm{H}}}
\newcommand\T{\ensuremath{\mathrm{T}}}
\newcommand\I{\ensuremath{\mathrm{I}}}
\newcommand\K{\ensuremath{\mathrm{K}}}
\newcommand\Q{\ensuremath{\mathrm{Q}}}
\newcommand\mR{\ensuremath{\mathrm{R}}}
\newcommand\Z{\ensuremath{\mathrm{Z}}}
\newcommand\E{\ensuremath{\mathrm{E}}}
\newcommand\U{\ensuremath{\mathrm{U}}}
\newcommand\V{\ensuremath{\mathrm{V}}}

\newcommand\bE{\ensuremath{\mathbf{E}}}
\newcommand\Var{\ensuremath{\mathbf{Var}}}
\newcommand\ran{\ensuremath{\operatorname{ran}}}
\newcommand\psip[1]{\ensuremath{\psi^{(#1)}}}
\newcommand\pij{\ensuremath{p_{ij}}}
\newcommand\pji{\ensuremath{p_{ji}}}
\newcommand\<{\ensuremath{\langle}}
\renewcommand\>{\ensuremath{\rangle}}
\newcommand\diag{\ensuremath{\operatorname{diag}}}
\newcommand\one{\ensuremath{\mathbf{1}}}
\newcommand\lambdamax{\ensuremath{\lambda_{\max}(\M)}}

%% Cross-referencing utilities. Use one or the other--whichever you prefer--
%% but comment out both lines for final version.
%\usepackage{showlabels}
%\usepackage{showkeys}

\begin{document}
%% Produces a test "layout" page, for "debugging" purposes only.
%% Comment out for final version.
%\layout % requires package layout (see above, on this same file)

%%%%%% Title page %%%%%%%%%%%
%% Sets page numbering to "roman style" i, ii, iii, iv, etc:
\pagenumbering{roman}
%
%% No numbering in the title page:
\thispagestyle{empty}
%
\begin{center}
  {\large\textbf{\thesistitle}}
  \vspace{.7in}

  by
  \vspace{.7in}

  \thesisauthor
  \vfill

\begin{doublespace}
  A dissertation submitted in partial fulfillment\\
  of the requirements for the degree of\\
  Master of Science\\
  Department of Mathematics\\
  New York University\\
  \graddate
\end{doublespace}
\end{center}
\vfill

\noindent\makebox[\textwidth]{\hfill\makebox[2.5in]{\hrulefill}}\\
\makebox[\textwidth]{\hfill\makebox[2.5in]{\hfill\thesisadvisor\hfill}}
\newpage
%%%%%%%%%%%%% Blank page %%%%%%%%%%%%%%%%%%
\thispagestyle{empty}
\vspace*{0in}
\newpage

%%%%%%%%%%%%%% Dedication %%%%%%%%%%%%%%%%%
%% Comment out the following lines if you do not want to dedicate
%% this to anyone...
\vspace*{\fill}
\begin{center}
  \thesisdedication\addcontentsline{toc}{section}{Dedication}
\end{center}
\vfill
\newpage
%%%%%%%%%%%%%% Acknowledgements %%%%%%%%%%%%
%% Comment out the following lines if you do not want to acknowledge
%% anyone's help...
\section*{Acknowledgements}\addcontentsline{toc}{section}{Acknowledgements}
%%% 
%%% \input{acknowledge}
%%%
First, I thank my advisor, Professor Jonathan Goodman, for giving me the opportunity to work on
this problem, and helping me arrive at the following exposition. Next, I wish to thank Professor
Helena Frydman for ﬁrst sparking my interest in Markov chains by giving lucid descriptions of their
many interesting properties and applications. I thank Professors James Demmel and Beresford
Parlett, for answering many questions pertaining to this problem, and Professor Leslie Greengard
for agreeing to review this paper.

Finally, I would like to thank my family, for their support of my interests in research (and
everything else!), and especially my parents, Barbara and Ted Terry, and Benita and Bill DeMeo.
Needless to say, without them this paper would not have been written.

\newpage
%%%% Abstract %%%%%%%%%%%%%%%%%%
\section*{Abstract}\addcontentsline{toc}{section}{Abstract}
%%%----------------------------------------------------------------
%%%
%%% \input{abstract}
%%%
\begin{center}
\newcommand\skipsize{6pt}
A Lanczos Procedure for Approximating Eigenvalues of Large Stochastic Matrices\\[\skipsize]
by\\[\skipsize]
William J. DeMeo\\[\skipsize]
Master of Science in Mathematics\\[\skipsize]
New York University\\[\skipsize]
Professor Jonathan Goodman, Chair
\end{center}

The rate at which a Markov chain converges to a given probability distribution
has long been an active area of research. Well known bounds on this rate of
convergence involve the subdominant eigenvalue of the chain's underlying
transition probability matrix. However, many transition probability matrices are
so large that we are unable to store even a vector of the matrix in fast
computer memory. Thus, traditional methods for approximating eigenvalues are
rendered useless.

In this paper we demonstrate that, if the Markov chain is reversible, and we
understand the structure of the chain, we can derive the coefficients of the
traditional Lanczos algorithm without storing a single vector. We do this by
considering the variational properties of observables on the chain's state
space. In the process we present the classical theory which relates the
information contained in the Lanczos coefficients to the eigenvalues of the
Markov chain. 
%%%----------------------------------------------------------------

\newpage
%%%% Table of Contents %%%%%%%%%%%%
\tableofcontents

\newpage
%%%%% List of Figures %%%%%%%%%%%%%
%% Comment out the following two lines if your thesis does not
%% contain any figures. The list of figures contains only
%% those figures included withing the "figure" environment.
%% \listoffigures\addcontentsline{toc}{section}{List of Figures}
%% \newpage

%%%%% List of Tables %%%%%%%%%%%%%
%% Comment out the following two lines if your thesis does not
%% contain any tables. The list of tables contains only
%% those tables included withing the "table" environment.
%% \listoftables\addcontentsline{toc}{section}{List of Tables}
%% \newpage

%%%%% List of Algorithms %%%%%%%%%%%%%
%% Comment out the following two lines if your thesis does not
%% contain any algorithms. The list of tables contains only
%% those tables included withing the "table" environment.
%% \listofalgorithms\addcontentsline{toc}{section}{List of Algorithms}
%% \newpage

%%%%% Body of thesis starts %%%%%%%%%%%%
\pagenumbering{arabic} % switches page numbering to arabic: 1, 2, 3, etc.
%% Introduction. If your thesis has no introduction, or chapter 1 is
%% meant to be the introduction, then comment out the lines below.
\section*{Introduction}\addcontentsline{toc}{section}{Introduction}
%%%----------------------------------------------------------------
%%%
%%% \input{intro}
%%%
The rate at which a Markov chain converges to a given probability distribution
has long been an active area of research. This is not surprising considering
this problem's relevance to the areas of statistics, statistical mechanics, and
computer science. Markov Chain Monte Carlo (MCMC) algorithms provide important
examples. These algorithms come in handy when we encounter a complicated
probability distribution from which we want to draw random samples. In
statistical mechanics, we might wish to estimate the phase average of a function
on the state space. Goodman and Sokal~\cite{GoodmanSokal:1989} examine Monte
Carlo methods in this context. Examples from statistics occur in the Bayesian
paradigm when we are forced to simulate an unwieldy posterior distribution (see,
e.g., Geman and Geman~\cite{Geman:1984}.

To implement the MCMC algorithm, we invent a Markov chain that converges to the
desired distribution (this is often accomplished using the \emph{Metropolis algorithm}
described in Chapter~\ref{cha:an-appl-conv}). Realizations of the chain will
eventually represent samples from this distribution. Sometimes
``eventually''---meaning all but finitely many terms of the chain---is just not 
enough. We need more practical results. In particular, we want to know how many
terms of the chain should be discarded before we are sampling from a
distribution that is close (in total variation distance) to 
%
%
% ------------- 03.txt -----------------------------------------------------
%
%
the distribution of interest. This is the purpose of bounding convergence rates
for Markov chains.

Often the Markov chains encountered in this context satisfy a condition known in
the physics literature as \emph{detailed balance}. Probabilists call chains with
this property \emph{reversible}. This simply means that the chain has the same
probability law whether time moves forward or backward.\footnote{This is not a
precise definition. In particular the chain must have started from its
stationary distribution. Full rigor is postponed until
Section~\ref{sec:general-theory}.} 
In this paper, we consider the rate at which such chains converge to
a \emph{stationary distribution}. This and other italicized terms are
defined in Section~\ref{sec:general-theory}.

There are a number of different methods in common use for bounding convergence
rates of Markov chains, and a good review of these methods with many references
can be found in~\cite{Rosenthal:1995}.  More recently developed methods
employing logarithmic Sobolev inequalities are reviewed in~\cite{Diaconis:1996}.  
Most of the bounds in common use involve the subdominant eigenvalue of the
Markov chain's transition probability matrix, and thus require good
approximations to such eigenvalues. In many applications, however, the
transition probability matrix is so large that it becomes impossible to store
even a single vector of the matrix in conventional computer memory. These so
called \emph{out-of-core} problems are not amenable to traditional eigenvalue
algorithms without modification.\footnote{By ``traditional 
  eigenvalue algorithms'' we mean those found, for example, in Golub and Van
  Loan~\cite{Golub:1996}. See also the book by Demmel\cite{Demmel:1997} for more 
  recent treatment.}
In this paper we develop such a modification for the Markov chain eigenvalue
problem. In particular we develop a method for approximating the first few
eigenvalues of a transition probability matrix when we know the general
structure of the underlying Markov chain. The method does not require storage
of large matrices or vectors. Instead we need only simulate the Markov chain,
and conduct a statistical analysis of the simulation.

Here is a brief summary of the paper. Section~\ref{sec:general-theory} contains
a review of the relevant Markov chain theory. Readers who already know the
basics of asymptotic theory of Markov chains might wish to skim 
Section\ref{sec:general-theory} if only to familiarize themselves with our
notation. Section~\ref{sec:funct-state-space} describes functions on the state 
% 04.txt
space of the Markov process. This section and Chapter~\ref{cha:invar-appr-invar}
develop the context in which we formulate the new ideas of the paper. In the
last section of Chapter~\ref{cha:invar-appr-invar},
Section~\ref{sec:krylov-subspace}, we describe the \emph{Krylov subspace} and
explain why this represents our best approximation to a subspace containing 
extremal eigenvectors of the transition probability matrix (more
  precisely, a similarity transformation of this matrix). The first section of
Chapter~\ref{cha:lanczos-procedures} describes the \emph{Lanczos algorithm} for
generating an orthonormal basis for the Krylov subspace. As it stands, this
algorithm is useless for an out-of-core problem since, by
definition of such problems, it requires too much data movement; all the
computing time is spent swapping data between slow and fast memory (disk to ram
to cache and back). We develop alternatives to the Lanczos algorithm and
demonstrate that the \emph{Lanczos coefficients} of the algorithm can be
obtained by simulations of the Markov chain, and this allows us to avoid
the standard algorithm altogether. Following this is a chapter describing the
Metropolis algorithm used to produce a reversible stochastic matrix. It is here
that we experiment with the procedure described in
Section~\ref{sec:lancz-proc-mark} and approximate the extremal eigenvalues of
the matrix, without storing any of its vectors. Finally,
Chapter~\ref{cha:conclusion} concludes the paper. 
% 05.txt

%%%===================================================================
%%%-------------------------------------------------------------------

%% If your thesis has different "Parts", use commands such as the following:
\part{Theory\label{part:one}}%

%%%-------------------------------------------------------------------
%%%
%%% \input{MarkovChains} % Chapter 2
%%%
\chapter{Markov Chains}

\section{General Theory}
\label{sec:general-theory}
This review of Markov chain theory can be found in any good probability text. The present
discussion is most similar to that of Durrett~\cite{Durret:1996}, to which we
refer the reader desiring greater detail.

\subsection{The Basic Setup}

Heuristically, a Markov chain is a stochastic process with a lack of memory property. Here
this means that the future of the process, given its past behavior and its present state, depends only
on its present state. This is the probabilistic analogue of a familiar property of classical particle
systems. Given the position and velocities of all particles at time t, the equations of motion can be
completely solved for the future evolution of the system. Thus, information describing the behavior
of the process prior to time t is superﬂuous. To be a bit more precise, if technical, we need the
following definitions.

\begin{definition}
%Definition 2.1.1 
Let $(\S, \S)$ be a measurable space. A sequence $X_n$, $n\geq 0$, of random variables
taking values in \S\ is said to be a Markov chain with respect to the filtration 
$\sigma(X_0, \dots, X_{n})$ if for all $B \in \S$,
\begin{equation}
% Eqn 2.1
\label{eq:1}
P(X_{n+1} \in B \mid \sigma(X_0, \dots, X_{n}))
=P(X_{n+1} \in B \mid \sigma(X_{n})).
\end{equation}
\end{definition}
Equation~(\ref{eq:1}) merely states that if we know the present location or state of $X_{n}$,
then information about earlier locations or states is irrelevant for predicting $X_{n+1}$.

\begin{definition}
%% Definition 2.1.2 
A function $p : S \times \S \rightarrow \R$ is said to be a \emph{transition probability} if:
\begin{enumerate}
\item for each $x \in S$, $A \mapsto p(x, A)$ is a probability measure on $(S, \S)$.
\item for each $A \in \S$, $x \mapsto p(x, A)$ is a measurable function.
\end{enumerate}
\end{definition}
We call $X_n$ a Markov chain with transition probabilities $p_n$ if
\begin{equation}
% Eqn 2.2
\label{eq:2.2}
P(X_{n+1} \in B \mid  \sigma(X_n)) = p_n(X_n, B)
\end{equation}
The spaces $(S, \S)$ that we encounter below are standard Borel spaces, so the existence of the
transition probabilities follows from the existence of regular conditional
probabilities on Borel spaces---a standard measure theory result 
(see e.g.\cite{Durret:1996}). %[3] page 230

Suppose we are given an initial probability distribution $\mu$ on $(S, \S)$ and a sequence $p_n$ of
transition probabilities. We can define a consistent set of finite dimensional distributions by
\begin{equation}
% Eqn 2.3
\label{eq:2.3}
P(X_j\in B_j ,0 \leq j \leq n) = \int_{B_0} \mu(dx_0) \int_{B_1} p_0(x_0, dx_1) 
%\int_{B_2} p_1(x_1, dx_2) 
\cdots 
\int_{B_n} p_{n-1}(x_{n-1}, dx_n).
\end{equation}
Furthermore, denote our probability space by
\[
(\Omega, \F) = (S^\omega,\S^\omega), \quad \text{ where } \omega = \{0,1,\dots\}.
\]
We call this \emph{sequence space} and it is defined more explicitly by
\[
S^\omega  = \{(\omega_0, \omega_1, \dots) : \omega_i \in S\}\quad \text{ and }
\quad
\S^\omega  = \sigma(\omega : \omega_i \in A_i \in \S).
\]
% 07.txt
The Markov chain that we will study on this space is simply $X_n(\omega) = \omega$, the coordinate maps.
Then, by the Kolmogorov extension theorem, there exists a unique probability measure $P_\mu$ on
$(\Omega, \F)$ so that the $X_n(\omega)$ have finite dimensional distributions~(\ref{eq:2.3}).

If instead of $\mu$ we begin with the initial distribution $\delta_x$, i.e., point mass at $x$, then we
denote the probability measure by $P_x$. With such measures defined for each $x$, we can in turn
define distributions $P_\mu$, given any initial distribution $\mu$, by
\[
P_\mu(A) = \int \mu(dx)P_x(A).
\]

That the foregoing construction---which, recall, was derived merely from an
initial distribution $\mu$ and a sequence $p_n$ of transition
probabilities---satisfies Definition~(\ref{eq:2.2}) of a Markov chain is not
obvious, and a proof can be found in~\cite{Durret:1996}.

To state the converse of the foregoing, if $X_n$ is a Markov chain with
transition probabilities $p_n$ and initial distribution $\mu$, then its finite
dimensional distributions are given by~(\ref{eq:2.3}). Proof of this is also found
in~\cite{Durret:1996}.

Now that we have put the theory on a firm, if abstract, foundation, we can bring
the discussion down to earth by making the forgoing a little more
concrete. First, we specialize our study of Markov chains by assuming that our
chain is \emph{temporally homogeneous}, which means that the transition
probabilities do not depend on time; i.e., 
$p_n(\omega_n, B) = p(\omega_n, B)$. (This is the stochastic analogue of a
conservative system.) 
Next we assume that our state space $S$ is finite, and suppose for all states
$i,j \in S$ that $p(i, j) \geq 0$, and $\sum_j p(i, j) = 1$ for all $i$. In
this case, equation~(\ref{eq:2.2}) takes a more intuitive form,
\[
P(X_{n+1}=j \mid X_n = i) =p(i,j),
\]
and our transition probabilities become
\[
p(i,A) = \sigma_{j\in A} p(i,j).
\]
% 08.txt
If $\P$ is a matrix whose $(i, j)$ element is the transition probability 
$p(i, j)$ then P is a \emph{stochastic matrix}; that is, a matrix with elements
$p_{ij}$ satisfying 
\[
\pij \geq 0, \quad \sum_j \pij =1, \quad (i, j =1, 2, \dots, d).
\]
We also refer to $\P$ as the transition probability matrix.

Without loss of generality, we can further suppose our Markov chain is \emph{irreducible}. This
means that, for any states $i, j$, starting in state $i$ the chain will make a
transition to state $j$ at some future time with positive probability. This
state of affairs is often described by saying that all states \emph{communicate}. We
lose no generality with this assumption because any \emph{reducible} Markov chain can
be factored into irreducible classes of states which can each be studied
separately.

The final two conditions we place on the Markov chains considered below will cost us
some generality. Nonetheless, there remain many examples of chains meeting these conditions
and making the present study worthwhile. Furthermore, it may be the case that, with a little
more work, we will be able to drop these conditions in future studies. The first condition is that
the chain is \emph{aperiodic}. If we let $I_x = \{n \geq 1: p^n(x,x) > 0\}$, we
call a Markov chain \emph{aperiodic} if, for any state $x$, the greatest common
divisor of $I_x$ is 1. The second assumption is that our chain is 
\emph{reversible}. This characterization is understood in terms of the following definition.

\begin{definition}
%% Definition 2.1.3 
\label{def:2.1.3}
A measure $\mu$ is called \emph{reversible} if it satisfies
\[
\mu(x)p(x,y) = \mu(y)p(y,x), \quad  \text{ for all $x$ and $y$.}
\]
We call a Markov chain \emph{reversible} if its stationary distribution (defined
in Section~\ref{sec:convergence-theorem}) is reversible. 
\end{definition}

% ---------- 09.txt ------------------------------------------------------------

\subsection{A Convergence Theorem}
% Section 2.1.2
\label{sec:convergence-theorem}
In succeeding arguments, we use some results concerning the asymptotic behavior of
Markov chains. These results require a few more definitions.

\begin{definition}
%% Definition 2.1.4
\label{def:2.1.4}
A measure $\pi$ is said to be a \emph{stationary measure} if
\begin{equation}
% Eqn 2.4
  \label{eq:2.4}
\sum_x \pi(x) p(x,y) = \pi(x).
\end{equation}
\end{definition}

Equation~(\ref{eq:2.4}) says $P_\pi(X_1 = y) = \pi(y)$, 
and by induction that $P_\pi(X_n = y) = \pi(y)$ for all $n \geq 1$. 
If $\pi$ is a probability measure, then we call $\pi$ a stationary distribution. It
represents an equilibrium for the chain in the sense that, if $X_0$ has
distribution $\pi$, then so does $X_n$ for all $n$.

When the Markov chain is irreducible and aperiodic, the distribution of the state at time
$n$ converges pointwise to $\pi$ as $n \rightarrow \infty$, regardless of the
initial state. It is convenient to state this convergence result in terms of the
Markov chain's transition probability matrix $\P$. Before doing so, we note that
irreducibility of a Markov chain is equivalent to irreducibility (in the usual 
matrix theory sense) of its transition probability matrix. Furthermore, it turns
out that a transition probability matrix of an aperiodic Markov chain falls into
that class of matrices often called acyclic, but for simplicity we will call
such stochastic matrices aperiodic. With this terminology, we can state the
convergence theorem in terms of the transition probability matrix \P.

% Theorem 2.1.1
\begin{theorem}
\label{thm-2.1.1}
Suppose $\P$ is irreducible, aperiodic, and has stationary distribution
$\pi$. Then as $n\rightarrow \infty$, $p^n(i,j) \rightarrow \pi(j)$.
\end{theorem}

The notation $p^n(i,j)$ means the $(i,j)$ element of the $n$th power of $\P$.

A Markov chain whose transition probability matrix satisfies the hypotheses of
Theorem~\ref{thm-2.1.1} is called \emph{ergodic}. If we simulate an ergodic chain for
sufficiently many steps, having 
%
%
%
% ---------------10.txt --------------------------------------------------
%
%
%
begun in any initial state, the final state is a sample point from a
distribution that is close to $\pi$.

To make this statement more precise requires that we define ``close.''
\begin{definition}
%% Definition 2.1.5 
Let $\pi$ be a probability measure on $S$. Then the \emph{total variation distance} at time
$n$ with initial state $x$ is given by 
\[
\Delta_x(n) = \|\P^n(x, A) - \pi(A)\|_{TV} = \max_{A\in \S} | \P^n(x,A)-\pi(A)|.
\]
\end{definition}
In what follows, we will measure rate of convergence using the function
$\tau_x(\epsilon)$, defined as the first time after which the total variation
distance is always less than $\epsilon$. That is, 
\[
\tau_x(\epsilon) = \min\{m : \Delta_x(n) \leq \epsilon \text{ for all } n\geq
m\}.
\]

To begin our consideration of the connection between convergence rates of Markov
chains and eigenvalues, we first note that an aperiodic stochastic matrix $\P$ has real
eigenvalues 
$1 = \lambda_0 > \lambda_1 \geq \lambda_2 \geq \cdots \geq \lambda_{d-1} \geq -1$, 
where $d = |S|$ is the dimension of the state space. For an ergodic chain, the
rate of convergence to the stationary distribution $\pi$ is bounded by a
function of the \emph{subdominant} eigenvalue. By subdominant eigenvalue we mean that
eigenvalue which is second largest in absolute value, and we denote this
eigenvalue by $\lambda_{\max} = \max{\lambda_1, |\lambda_{d-1}|}$. 
The function bounding the rate of convergence of a Markov chain is given by the
following theorem ($\log$ denotes the natural logarithm):
\begin{theorem}
% Theorem 2.1.2
\label{thm-2.1.2}
The quantity $\tau_x(\epsilon)$ satisfies
\begin{enumerate}
\item 
$\tau_x(\epsilon) \leq (1-\lambda_{\max})^{-1}(\log \pi(x)^{-1} + \log \epsilon^{-1})$;
\item $\max_{x \in S} \tau_x(\epsilon) \geq \frac{1}{2} \lambda_{\max}(1-\lambda_{\max})^{-1} \log(2\epsilon)^{-1}$.
\end{enumerate}
\end{theorem}
As this theorem shows, if we have an upper bound on the subdominant eigenvalue, then we have
an upper bound on the function $\tau_x(\epsilon)$. In what follows, we will
derive an approximation to the 
% 11.txt
subdominant eigenvalue and supply error bounds. Together, an approximation and error bounds
for $\lambda_{\max}$  provide enough information to make Theorem~\ref{thm-2.1.2} useful.

\section{Functions on the State Space}
% Sec 2.2
\label{sec:funct-state-space}
Recall that $X_n(\omega) = \omega_n\in S$ denotes the state in which the Markov
chain exists at time $n$. 
Suppose that 
$\Phi = \{\phi_1, \dots, \phi_p\}$ 
is a collection of $p$ \emph{observables}, or functions defined on the state
space $S$. Furthermore, let these observables be real valued, 
$\phi_i : S \rightarrow \R$. 
It is often useful to assume that none of the observables is a constant function. Suppose now that
the state space $S$ is finite with $d$ possible states. Then, since an
observable is simply a map of the state space, we can think of each $\phi_i$ as a
vector of $d$ real numbers---the $d$ values that it takes on at the different states.

Now assume the Markov chain is irreducible, and let $\pi$ denote its stationary distribution.
If we start the chain from its stationary distribution---i.e., suppose $X_0$ has
distribution $\pi$---then $X_n$ is a stationary process. 
Furthermore, for each $i$, $\phi_i(X_n)$ is a stationary stochastic process with
\emph{mean}
\[
\bE_\pi\phi_i = \sum_{x\in S} \pi(x)\phi_i(x)
\]
and \emph{autocovariance function}
\begin{align}
% Eqn 2.5
\bC_\pi (\phi_i(X_n),  \phi_i (X_{n+s})) 
&= \bE_\pi[(\phi_i(X_n) - \bE_\pi\phi_i)(\phi_i(X_{n+s}) - \bE_\pi\phi_i)]\\
&= \sum_{x, y\in S} P_\pi(X_n = x, X_{n+s} = y) (\phi_i(x)- \bE_\pi\phi_i)(\phi_i(y) - \bE_\pi\phi_i).\nonumber
\end{align}
By the definition of conditional probability, we can write~(\ref{eq:2.4}) as follows:
\[
\sum_{x, y\in S} P_\pi(X_n = x) P_\pi(X_{n+s} = y\mid X_n = x) (\phi_i(x)-
\bE_\pi\phi_i)(\phi_i(y) - \bE_\pi\phi_i).
\]
Equivalently,
\[
\sum_{x, y\in S} \pi(x)p^s_{xy} (\phi_i(x)-\bE_\pi\phi_i)(\phi_i(y) - \bE_\pi\phi_i).
\]
Here $p^s_{xy}$ denotes the element in row $x$ and column $y$ of $\P^s$, the
$s$th power of the transition probability matrix. 
Similarly, we define the \emph{cross-covariance} between the function 
$\phi_i$ at time $n$ and $\phi_j$ at time $n+s$ as
\begin{align}
% Eqn 2.6
\bC_\pi (\phi_i(X_n),  \phi_j (X_{n+s})) 
&= \bE_\pi[(\phi_i(X_n) - \bE_\pi\phi_i)(\phi_j(X_{n+s}) - \bE_\pi\phi_j)]\nonumber\\
&= \sum_{x, y\in S} \pi(x)p^s_{xy} (\phi_i(x)-\bE_\pi\phi_i)(\phi_j(y) - \bE_\pi\phi_j).
\end{align}
Now let $\<\Phi\>$ denote the matrix of mean vectors whose $j$th column is 
$\bE_\pi\phi_j\one$, where $\one = (1,\dots, 1)^t$,
and let $\Pi = \diag(\pi(\omega_1),\dots, \pi(\omega_d))$ be the $d \times d$
diagonal matrix with stationary probabilities $\pi(\omega)$
on the main diagonal and zeros elsewhere. 
Finally, denoting by $\C(s)$ the $p \times p$ covariance matrix
whose $(i,j)$ element is $\bC_\pi(\phi_i(X_n), \phi_j(X_{n+s}))$, we have
\begin{align*}
\C(0) &= \E(\Phi(X_n) - \<\Phi\>)(\Phi(X_n) - \<\Phi\>)^t\\
&= (\Phi - \<\Phi\>)^t\Pi (\Phi-\<\Phi\>), \\
\C(s) &= \E(\Phi(X_n) - \<\Phi\>)(\Phi(X_{n+s}) - \<\Phi\>)^t\\
&= (\Phi - \<\Phi\>)^t \Pi \P^s (\Phi - \<\Phi\>).
\end{align*}

Below, we will also find it useful to have at our disposal a new matrix that is \emph{similar} to the
transition probability matrix. We have in mind the matrix 
$\M = \Pi^{1/2}\P\Pi^{-1/2}$. 
As is easily verified, this allows us to write the covariance matrix as
\begin{align}
\label{eq:2.7}
\C(s) &= 
(\Phi - \<\Phi\>)^t \Pi^{\frac{1}{2}t}\M^s\Pi^{\frac{1}{2}}(\Phi - \<\Phi\>)\nonumber\\
&=\Psi^t \M^s \Psi,  % Eqn 2.7
\end{align}
%% \\Phi\Phi‘M’\\Phi\Phi (2.7)
where we have defined $\Psi = \Pi^{\frac{1}{2}}(\Phi - \<\Phi\>)$. Recall that
our main motivation is that for out-of-core problems traditional eigenvalue
algorithms are inadequate. With this fact and the form~(\ref{eq:2.7}) % (2.7) 
in 
%
%
% ------------------ 13.txt ---------------------------------------------------
%
%
mind, we will consider using the covariance of observables on the state space to implement the
%% (otherwise useless) 
Rayleigh-Ritz procedure, which we describe below. This procedure requires
that $\M$ be symmetric. As the next fact demonstrates, 
%% this required symmetry is the reason for our
%% interest in the reversibility property of the Markov chain.
this need for symmetry is the reason we insist that the Markov chain be reversible.
\begin{fact}
%% Fact 2.2.1 
The matrix $\M$ is symmetric if and only if the Markov chain is reversible (i.e., iff the
process satisfies the \emph{detailed balance} condition).
\end{fact}
\begin{proof}
  \begin{align*}
  \M \text{ is symmetric } 
& \quad \Longleftrightarrow  \quad
 (\Pi^{1/2}\P\Pi^{-1/2})^t = \Pi^{1/2}\P\Pi^{-1/2}\\
& \quad \Longleftrightarrow  \quad
 \Pi^{-\frac{1}{2}t}\P^t\Pi^{\frac{1}{2}t} = \Pi^{1/2}\P\Pi^{-1/2}\\
& \quad \Longleftrightarrow  \quad
 \P^t\Pi^t = \Pi\P.
  \end{align*}
Elementwise, the final equality is $\pi_i\pij = \pi_j\pji$. 
According to Definition~\ref{def:2.1.3}, this states
that $p$ is a reversible measure.
\end{proof}

% 14.txt

%%%-------------------------------------------------------------------
%%%
%%% \input{InvariantSubspaces} % Chapter 3
%%%
\chapter{Invariant and Approximate Invariant Subspaces}
\label{cha:invar-appr-invar}
\section{Invariant Subspaces}

\begin{definition}
%% Definition 3.1.1
A subspace $S \subseteq \R^n$ with the property that
\[
x \in S \quad \Longrightarrow \quad \M x \in S
\]
is said to be \emph{invariant} for $\M$.
\end{definition}

Recall that, having chosen observables $\<\Phi\> = (\phi_1, \dots, \phi_p)$, we
constructed the covariance matrix 
$\C(s) = \Psi^t \M \Psi$. 
If the column space of $\Psi$, which we denote by $\ran(\Psi)$, is an invariant
subspace for $\M$, the definition implies $\M\psi_j \in \ran(\Psi)$.
That is, for each $\psi_i$ there exists a vector
$t$ of coeﬂicients such that $\M \psi_j = \sum_{i=1}^p t_i \psi_{ij}$. 
This is true for all $j$ and, putting each vector of
coefficients into a matrix $\T$, we see that 
$\M \Psi = \Psi \T$. Conversely, $\M\Psi = \Psi \T$ implies that $\M\psi_j$ is a
linear combination of columns of $\Psi$, so $\ran(\Psi)$ is invariant. We have
thus proved the following 
% 15.txt
\begin{fact}
%% Fact 3.1.1 
\label{fact:3.1.1}
The subspace $\ran(\Psi)$ is invariant for $\M$ if and only if there exists 
$\T \in \R^{p\times p}$ such that $\M\Psi = \Psi \T$.
\end{fact}

Consequently,
\begin{fact}
%% Fact 3.1.2
\label{fact:3.1.2}
$\lambda(\T) \subseteq \lambda(\M)$.
\end{fact}
\begin{proof}
%Proof of 3.1.2:
~
\vskip-2cm
  \begin{align*}
\lambda \in \lambda(\T)
& \quad \Longleftrightarrow  \quad
(\exists v \in \R^p) \T v = \lambda v\\
& \quad \Longleftrightarrow  \quad
\Psi \T v = \lambda \Psi v\\
& \quad \Longleftrightarrow  \quad
\M \Psi v = \lambda \Psi v\\
& \quad \Longleftrightarrow  \quad
\lambda \in \lambda(M).
  \end{align*}
The second equivalence follows from Fact~\ref{fact:3.1.1}.  %3.1.1.
\end{proof}
%Facts~\ref{fact:3.1.1} and~\ref{fact:3.1.2} are theoretically useful. To see why
To see why Facts~\ref{fact:3.1.1} and~\ref{fact:3.1.2} are theoretically useful,
%consider the equation of Fact~\ref{fact:3.1.1}:
%% \begin{align*}
%% & \Psi \T = \M \Psi \\
%% \Longleftrightarrow \quad & \Psi^t\Psi \T = \Psi^t \M \Psi\\
%% \Longleftrightarrow \quad & \T = (\Psi^t\Psi)^t \Psi^t \M\Psi.
%% \end{align*}
%% \begin{equation*}
%% \Psi \T = \M \Psi \quad\Longleftrightarrow \quad \Psi^t\Psi \T = \Psi^t \M \Psi
%% \Longleftrightarrow \quad  \T = (\Psi^t\Psi)^t \Psi^t \M\Psi.
%% \end{equation*}
note that the equation of Fact~\ref{fact:3.1.1}, $\Psi \T = \M \Psi$, is
equivalent to 
$\Psi^t\Psi \T = \Psi^t \M \Psi$, which is equivalent to 
$\T = (\Psi^t\Psi)^t \Psi^t \M\Psi$.
In this form, we recognize that $\T = \C^{-1}(0)\C(1)$. That is, using only the
covariance of observables on the state space, we can generate a matrix $\T$ with
the property $\lambda(\T) \subseteq \lambda(\M)$. Recalling that 
$\M = \Pi^{1/2}\P\Pi^{-1/2}$, 
we see that $\M$ is similar to our original stochastic matrix $\P$, 
and thus $\lambda(M) = \lambda(P)$.

\section{Approximate Invariant Subspaces}
% Sec 3.2
\label{sec:appr-invar-subsp}
%% We qualified the foregoing by stating that the facts are only theoretically
%% useful. This
We noted above that Facts~\ref{fact:3.1.1} and~\ref{fact:3.1.2} are
theoretically useful. Speaking practically now,
%% This is because 
when choosing observables on the state space we may not be sure that
they will satisfy the 
%
%
% -------------16.txt---------------------------------------------------
%
%
primary assumption underlying the two facts. %Facts 3.1.1 and 3.1.2. 
Recall the assumption: $\ran(\Psi)$ is an invariant
subspace for $\M$ where $\Psi = \Pi^{1/2}(\Phi - \<\Phi\>)$. It may well be the
case that there exists $\psi_j\in \ran(\Psi)$ such that 
$\M\psi_j \notin \ran(\Psi)$, thereby violating the assumption. 
Even if we are lacking an invariant subspace, however, for some applications it
is reasonable to expect that observables can be chosen to provide at least
an \emph{approximate invariant subspace}, which is defined as follows:
\begin{definition}
%% Definition 3.2.1
If the columns of $\Psi\in \R^{d\times p}$ are independent and the norm of the
\emph{residual matrix} $\E = \M\Psi - \Psi \T$ is small for some  
$\T \in \R^{p\times p}$, then
$\ran(\Psi)$ defines an \emph{approximate invariant subspace}. 
\end{definition}
To see how an approximate invariant subspace can be useful for approximating
eigenvalues of $\M$, we recall a theorem from Golub and Van Loan~\cite{Golub:1996}.
\begin{theorem}
% Theorem 3.2.1
\label{thm:3.2.1}  
Suppose $\M \in R^{d\times d}$ and $\T \in \R^{p\times p}$ are symmetric and let
$\E = \M\Q - \Q\T$, where $\Q \in \R^{d\times p}$ is orthonormal 
(i.e., $\Q^t\Q = \I$). Then there exist $\mu_1,\dots, \mu_p \in \lambda(\T)$ and
$\lambda_1, \dots, \lambda_p \in \lambda(\M)$ such that
\[
|\mu_k - \lambda_k| \leq \sqrt{2}\|\E\|_2, \quad \text{ for $k = 1, \dots, p$.}
\]
\end{theorem}
If the subspace $\ran(\Q)$ is an approximate invariant subspace, then the definition implies that
there is a choice $\T$ rendering the error $\|\E\|_2$ small, and thus the
eigenvalues of $\T$ provide a good approximation to those of $\M$.

When considering the foregoing ideas, it is apparent that their application
presents new---but hopefully less prohibitive---problems. As these problems
are the focus of the rest of the paper, now is a good time to examine
them.

First, the preceding theorem assumes a matrix $\Q$ whose columns form an
orthonormal basis for the approximate invariant subspace. For our problem, 
derivation of such a $\Q$ is tricky, and we must prepare for this.

Next, having an approximate invariant subspace at our disposal merely tells us
that there exists some matrix $\T$ which makes the error 
%
%
% ------------- 17.txt ----------------------------------------------------
%
%
$\|\E\|_2$ small. We must discover the form of such a $\T$. Moreover, it is
natural to seek that $\T$ which minimizes $\|\E\|_2$ for a given approximate
invariant subspace.

Finally, in order to apply these ideas to realistic eigenvalue problems, we must
find a practical way to generate %% the most appropriate
a good approximate invariant subspace.

We now address each of these issues in turn.
%the order raised.

Recall the matrix $\Psi = \Pi^{1/2}(\Phi — \<\Phi\>)$. The columns of this
matrix, though independent (by choice of independent observables), are not
necessarily orthonormal. However, consider the polar decomposition 
$\Psi = \Q \Z$, where $\Q^t\Q = \I$ and $\Z^2 = \Psi^t\Psi$ is a symmetric
positive semidefinite matrix.\footnote{Recall that the polar decompostion is
  derived from the singular value decomposition, $\Psi = \U\Sigma \V^t$, by
  letting $\Q = \U \V^t$ and $\Z = \V\Sigma \V^t$.}
%% ‘Recall that the polar decomposition is derived from the SVD, \Psi = UZJV‘, by letting Q = UV‘ and Z = VEV‘.
Note that $\Z = (\Psi^t\Psi)^{1/2}$ is nonsingular, so $\Q$ has the form
\[
\Q = \Psi \Z^{-1} = \Psi(\Psi^t\Psi)^{-1/2},
\]
and it is clear that $\ran(\Q) = \ran(\Psi)$. 
Perhaps $\ran(\Q)$ is a useful approximation to the invariant
subspace for $\M$. If $\ran(Q)$ is not itself invariant, we have the error 
matrix $\E =\M\Q - \Q\T$.  Below we show that the
$\T$ which minimizes $\|\E\|_2$ is $\T = \Q^t\M\Q$. 
This yields the following
\begin{theorem}
%Theorem 3.2.2
\label{thm:3.2.2}  
If $\Psi = \Q\Z$ is the polar decomposition of $\Psi$, then the matrix
\[
\T = \Q^t\M\Q = (\Psi^t\Psi)^{-1/2}\Psi^t \M^t \Psi (\Psi^t\Psi)^{-1/2}
\]
minimizes $\|\E\|_2 = \|\M\Q - \Q\T\|_2$.
\end{theorem}
\begin{proof}
We prove the result by establishing the following
\\[6pt]
\underline{Claim:} Given $\M \in \R^{d\times d}$ suppose $Q \in  \R^{d\times d}$
satisfies $\Q^t\Q =\I$. 
Then,
\[
\min_{\T\in \R^{p\times p}} \|\M\Q - \Q\T\|_2 = \|(\I- \Q\Q^t)\M\Q\|_2,
\]
and $\T = \Q^t\M\Q$ is the minimizer.
%
%
% -------------- 18.txt -------------------------------
%
%
The claim is verified by an easy application of the Pythagorean theorem:
For any $\T \in \R^{p\times p}$ we have
\begin{align}
\label{eq:3.1}
\|\M\Q - \Q\T\|^2_2 &= \|\M\Q - \Q\Q^t\M\Q + \Q\Q^t\M\Q - \Q\T\|^2_2 \nonumber\\
&=\|(\I - \Q\Q^t)\M\Q + \Q(\Q^t\M\Q - \T)\|^2_2 \nonumber \\
&=\|(\I - \Q\Q^t)\M\Q\|^2_2 + \|\Q(\Q^t\M\Q - \T)\|^2_2 \\
&=\|(\I - \Q\Q^t)\M\Q\|^2_2 \nonumber
\end{align}
Equality~(\ref{eq:3.1}) holds since $\I - \Q\Q^t$ projects $\M\Q$ onto the
subspace orthogonal to 
$\ran(Q)$. Thus, the two terms in the expression on the right are orthogonal,
and the Pythagorean theorem yields equality. The concluding inequality
establishes that the minimizing $\T$ is that which annihilates the 
second term in~(\ref{eq:3.1}), that is, $\T = \Q^t\M\Q$.

The second equality in Theorem~\ref{thm:3.2.2} is a consequence of the polar decomposition, in
which $\Q = \Psi \Z^{-1} = \Psi(\Psi^t\Psi)^{-1/2}$. Thus,
$\T = (\Psi^t\Psi)^{-1/2}\Psi^t \M^t \Psi (\Psi^t\Psi)^{-1/2}$
is the minimizer, as claimed.
\end{proof}

\subsection{The Krylov Subspace}
% Sec 3.3
\label{sec:krylov-subspace}
Suppose the columns of a matrix $\Q \in \R^{d\times p}$ give an orthonormal
basis for an approximate invariant subspace. Then, as we have seen, 
\begin{enumerate}
\item 
$\T = \Q^t\M\Q$ minimizes $\|\E\|_2 = \|\M\Q - \Q\T\|_2$ and
\item there exist $\mu_1, \dots, \mu_p\in \lambda(\T)$ and 
$\lambda_1,\dots, \lambda_p \in \lambda(\M)$ such that
\[
\|\mu_k-\lambda_k\| \leq \sqrt{2}\|\E\|_2, \quad \text{ for $k=1,\dots, p$.}
\]
\end{enumerate}
%
%
% ---------------- 19.txt --------------------------------------------
%
%
Given an approximate invariant subspace $\ran(\Q)$ of dimension $p$, these facts
tell us what matrix we should use to approximate $p$ elements of the spectrum of
$\M$. Now all we lack is a description of $\ran(\Q)$. 
That is, we have not specified which approximate invariant subspace would best
suit our objective of approximating the subdominant eigenvalue
$\lambda_{\max}(\M)$. For this purpose the following definition is useful:
\begin{definition}
%% Definition 3.3.1
The \emph{Raleigh quotient} of a symmetric matrix $\M$ and a nonzero vector $x$ is
\[
\rho(x, \M) = \frac{x^t\M x}{x^tx}.
\]
\end{definition}
We will denote Raleigh quotient by $\rho(x)$ when the context makes clear what
matrix is involved.

To find the approximate invariant subspace most appropriate for our problem, we
choose each dimension successively, providing justification at each step. 
We start with one observable $\phi$ on the state space, and let 
$\psi_1 = \Pi^{1/2}(\phi -\E_\pi\phi)$. That is, $\psi_1$ is a centered (mean
zero) observable whose $i$th coordinate is weighted by $\sqrt{\pi(i)}$. 
Notice that the definition 
%% (which comes directly from the definition of ^tII
%% following equation (2.7)) is such that 1/)1 is a constant vector if 
(which comes from the definition of $\Psi$ following Equation~(\ref{eq:2.7})) 
is such that $\psi_1$ is a constant vector if and only if $\phi$ is oonstant on the
state space, in which case $\psi_1$ is the constant zero function. 
%% This fact
%% provides one reason observables that are constant on the state space are not
%% interesting. 
(Observables that are constant on the state space are not interesting.) 
Second, recall that the row sums of the matrix $\P$ are all one, therefore the
eigenvector corresponding to the eigenvalue $\lambda_0(\P) = 1$ is the constant
vector.  By definition of $\M = \Pi^{1/2}\P\Pi^{-1/2}$, we see that the constant
vector is also the eigenvector corresponding to $\lambda_0$.  This will play an
important role in what follows, as it allows us to focus %primarily 
on the subdominant eigenvalue 
$\lambda_{\max}(\M) = \max\{\lambda_1(\M), |\lambda_{d-1}(\M)|\}$
rather than on $\lambda_0(M)$ (which we already know is 1).

Now, notice that
\begin{equation}
  \label{eq:3.2}
|\rho(\psi_1,\M)| \leq \max |\rho(x,\M)| = \lambda_{\max}(\M)
\end{equation}
% 20.txt
where the max is taken over all nonconstant vectors. Since our interest 
centers on 
$\lambda_{\max}(\M)$, 
we would like a $\psi_1$ that makes the left hand side of~(\ref{eq:3.2}) large. 
This would be achieved if $\psi_1$ were to lie in the space spanned by, say, the
first two eigenvectors of the Markov chain (sometimes referred 
to as the \emph{slowest modes} of the process). However, this subspace spans
only two dimensions of the entire $d$-dimensional space, and it is more likely
that $\psi_1$ only comes close, at best, to lying in the subspace of
interest. Now, given $\psi_1$, a judicious choice for the second dimension $\psi_2$, and
hence $\Psi_2 = [\psi_1, \psi_2]$, would be that which makes 
$\max_{a\neq 0} |\rho(\Psi_2 a)|$ as large as possible. To establish that 
this is indeed the right objective, note the following:
\begin{align}
\label{eq:3.3}
\max_{a\neq 0}|\rho(\Psi_2a,\M)| &= \max_{a\neq 0} \left|\frac{a^t \Psi^t_2 \M \Psi_2 a}{a^t \Psi^t_2 \Psi_2 a}\right|\nonumber\\
&= \max_{x\in \ran(\Psi_2)}|\rho(x,\M)|\nonumber\\
&= \max|\rho(x,\M)|\\
&= \lambdamax\nonumber
\end{align}
Again, the max on the right side of~(\ref{eq:3.3}) is over nonconstant vectors. 
In other words, we wish to chose
$\Psi_2 = [\psi_1, \psi_2]$ so that there is a vector $a \in \R^2$ making 
$|\rho(\Psi_2 a,\M)|$ close to $\lambdamax$.

Now, $\rho(\psi_1)$ changes most rapidly in the direction of the gradient
$\nabla \rho(\psi_1)$.
\begin{equation}
  \label{eq:3.4}
\nabla \rho(\psi_1) = \left(\frac{\partial \rho(\psi_1)}{\partial \psi_1(1)}, \dots,
\frac{\partial \rho(\psi_1)}{\partial \psi_1(d)}\right)
 = \frac{2}{\psi_1^t\psi_1} (\M\psi_1 - \rho(\psi_1)\psi_1).
\end{equation}
So, to maximize the left hand side of~(\ref{eq:3.3}), $\Psi_2$ should be chosen
so that the subspace $\ran(\Psi_2)$ contains the gradient vector.
That is, we must choose $\psi_2$ so that
\begin{equation}
\label{eq:3.5}
\nabla \rho(\psi_1) \in \ran\{\psi_1, \psi_2\} = \ran(\Psi_2).
\end{equation}
Clearly, Equation~(\ref{eq:3.4}) implies 
$\nabla \rho(\psi_1) \in \ran\{\psi_1, \M\psi_1\}$.
Thus, if $\ran\{\psi_1, \psi_2\} = \ran\{\psi_1, \M\psi_1\}$, 
then~(\ref{eq:3.5}) is satisfied.
%
%
% ------------ 21.txt ----------------------------------------------
%
%
In general, having chosen $\Psi_k = [\psi_1, \dots,  \psi_k]$ so that
$\ran\{\psi_1, \psi_2, \dots, \psi_k\} = \ran\{\psi_1, \M\psi_1, \dots,
\M\psi_k\}$,
we must chose $\psi_{k+1}$ so that for any nonzero vector $a\in \R^k$,
\begin{equation}
\label{eq:3.6}
\nabla \rho(\Psi_ka) \in \ran\{\psi_1, \psi_2, \dots, \psi_k\}.
\end{equation}
Now,
\[
\nabla \rho(\Psi_ka)   
 = \frac{2}{a^t\Psi_k^t\Psi_ka} (\M\Psi_k - \rho(\Psi_ka)\Psi_ka).
\]
and therefore,
\[
\nabla \rho(\Psi_ka)  \in 
\ran(\M\Psi_k) \cup \ran(\Psi_k) = \ran\{\psi_1, \M\psi_1, \dots, \M^k\psi_1\}.
\]
Thus, the requirement~(\ref{eq:3.6}) is satisfied when 
\[
\ran\{\psi_1, \psi_2, \dots, \psi_k\} = \ran\{\psi_1, \M\psi_1, \dots,
\M^k\psi_1\}.
\]

In conclusion, the $p$-dimensional approximate invariant subspace that is most suitable
for our objective is
\[
\sK(\M, \psi_1, p) = \ran\{\psi_1, \M\psi_1, \dots, \M^{p-1}\psi_1\}.
\]
This is known as the \emph{Krylov subspace}. Therefore, to answer the problem posed at the outset
of this section, if we take the columns of $\Q$ to be an orthonormal basis for 
$\sK(\M,\psi_1, p)$, then the eigenvalues of $\T = \Q^t\M\Q$ should provide good
estimates of $p$ extremal eigenvalues of $\M$. 
%
%
% --------------- 22.txt ---------------------------------------
%
%

%%%-------------------------------------------------------------------
%%%
%%% \input{LanczosProcedures} % Chapter 4
%%%
\chapter{Lanczos Procedures}
\label{cha:lanczos-procedures}
The following will review the usual procedure for generating an orthonormal basis for
$\sK(\M, \psi_1, p)$, 
and conclude by showing that, for our problem, we can find the matrix 
$\Q^t\M\Q$
without actually carrying out this procedure. Note that this conclusion is essential since the procedure
requires matrix vector multiplication---an operation we have assumed impossible for large enough $d$.

\section{The General Lanczos Algorithm}
% Sec 4.1
\label{sec:gener-lancz-algor}
Let $\psip{k} =  \M^k \psi$ and consider the $d\times d$ matrix
\[
\Psi = [\psi, \M\psi, \dots, \M^{d-1}\psi] = [\psip{0}, \psip{1}, \dots, \psip{d-1}]
\]
Notice that $\M\Psi = [\psip{1}, \psip{2}, \dots, \psip{d-1}, \M^d \psi]$,
and assuming for the moment that $\Psi$ is nonsingular,
we can compute the vector $h = -\Psi^{-1}\M^d \psi$. Thus,
\begin{equation}
\label{eq:10000}
\M\Psi = \Psi [e_2,\dots,e_d,-h],  
\end{equation}
where $e_j$ is the column vector with 1 in
the $j$th position and zeros elsewhere. We define $H = [e_2,\dots,e_d,-h]$,
so~(\ref{eq:10000}) becomes $\M\Psi = \Psi H$.
%
%
% ----------------- 23.txt ----------------------------------------------
%
%
Equivalently,
\begin{equation}
  \label{eq:4.1}
H = 
\begin{pmatrix}
0 & 0      & \cdots & 0 & -h_1 \\
1 & 0      & \cdots & 0 & -h_2\\
  & \ddots &        &   & \vdots \\
  &        & \ddots & 0 & -h_{d-1}\\
  &        &        & 1 & -h_{d}
\end{pmatrix} = \Psi^{-1} \M \Psi.
\end{equation}
$\H$ is a \emph{companion matrix} which means that its characteristic polynomial
is $p(x) = x^d + \sum_{i=1}^d h_i x^{i-1}$.
Since $\H$ is similar to $\M$, finding the eigenvalues of $\M$ is equivalent to
finding the roots of $p(x)$. However, this is of little practical use since
finding $h$, constructing $p(x)$, and finding its roots is 
probably a harder problem than the one we started with. Instead, the value of decomposition (4.1)
derives from its \emph{upper Hessenberg form}. We exploit this property below.

Let $\Psi = \Q\mR$ be the QR decomposition of $\Psi$. Since $\Psi$ is assumed nonsingular,
\[
\Psi_d^{-1}\M \Psi_d = (\mR^{-1}\Q^t)\M(\Q\mR) = \H
\]
Therefore, $\Q^t\M\Q = \mR\H\mR^{-1}$.  Let $\T \mR\H\mR^{-1}$.  Then, since $\mR$
and $\mR^{-1}$ are both upper triangular and $\H$ is upper Hessenberg $\T = \mR\H\mR^{-1}$ is also upper
Hessenberg. Furthermore, since $\M$ is symmetric, it is clear that 
$\T^t =\Q^t\M\Q = T$. Thus, $\T$ is both upper Hessenberg and symmetric. Therefore,
$\T$ is tridiagonal and we can write it as follows:
\begin{equation}
  \label{eq:4.2}
\T = 
\begin{pmatrix}
\alpha_1 & \beta_1 &      0 & \cdots & 0 \\
\beta_1  & \alpha_2 & \ddots &        & \vdots\\
         & \ddots & \ddots & \ddots   &\\
\vdots   &       &  \ddots & \ddots & \beta_{d-1}\\
0  &   \cdots    &   & \beta_{d-1} &\alpha_d
\end{pmatrix}.
\end{equation}
%
%
% ------------ 24.txt ----------------------------------------------
%
%
Equating columns $j$ on both sides of $\M\Q = \Q\T$ yields
\begin{equation}
\label{eq:4.3}
\M q_j = \beta_{j-1}q_{j-1}+\alpha_j q_j + \beta_j q_{j+1}
\end{equation}
Since the columns of $\Q$ are orthonormal, multiplying both sides 
of~(\ref{eq:4.3})
by $q_j$ and $q_{j+1}$ yields
$\alpha_j = q_j^t \M q_j$ and $\beta_j = q_{j+1}^t \M q_j$.
The foregoing justifies what is called the \emph{Lanczos algorithm}, which is performed as follows:
         \begin{algorithm}
           \caption{The Lanczos algorithm for partial reduction to symmetric
             tridiaganal form.}
           \begin{algorithmic}
             \STATE $q_1 = \psi/\|\psi\|_2$, $\beta_0 = 0$, $q_0= 0$
             \FOR{$j = 1, \dots, p$}
             \STATE $z \leftarrow \M q_j$
             \STATE $\alpha_j \leftarrow q_j^tz$
             \STATE $z \leftarrow z - \alpha_j q_j - \beta_{j-1} q_{j-1}$
             \STATE $\beta_j \leftarrow \|z\|_2$
             \IF{$\beta_j=0$}
             \STATE return
             \ENDIF
             \STATE $q_{j+1} = z/\beta_j$
             \ENDFOR
           \end{algorithmic}
         \end{algorithm}
The $q_j$ computed by the Lanczos algorithm are often called the \emph{Lanczos vectors}. If the
loop in the algorithm is terminated because $\beta_{k}= 0$, this indicates that
an exact invariant subspace has been computed, and is given by 
$\ran\{q_1, \dots, q_k\}$. Otherwise, we usually halt the algorithm after 
$p$ steps, in which case the algorithm converges to approximations of at most
  $p$ eigenvalues.\footnote{A Detailed discussion of such convergence (and
    misconvergence) is given in~\cite{Demmel:1997}.}

Note that Equation~(\ref{eq:4.3}) results from the equation $\M\Q = \Q\T$. The latter equation holds
since, starting with $d$ vectors in our subspace, $\Q$ is a $d \times d$ matrix
whose columns form an orthonormal basis for all of $\R^d$, which is clearly an
invariant subspace. The Lanczos algorithm above, however, 
%
%
% ------------ 25.txt ---------------------------------------------------------
%
%
proceeds for only $p$ steps, producing a $d \times p$ matrix $\Q$, whose columns
form an orthonormal basis for the \emph{approximate} invariant subspace 
$\sK(\M, \psi, p)$, and a $p \times p$ matrix $\T, = \Q^t\M\Q$. In that case,
$\|\M\Q - \Q\T_p\|_2 = \|\E\|_2$ is nonzero and gives the error bound described
in Theorem~\ref{thm:3.2.1}. Now, writing the full $d \times d$ matrix $\T$
of~(\ref{eq:4.2}) as
\begin{equation}
  \label{eq:4.4}
\T = 
\left( \begin{array}{c|c}
\T_p & \T_{pu}^t \\
 \hline
\T_{pu} & \T_{u}
\end{array}
\right)
=
\left( \begin{array}{cccc|cccc}
\alpha_1 & \beta_1 &      &                 & & && \\
\beta_1 & \ddots  & \ddots&                 & & && \\
        & \ddots  & \ddots  & \beta_{p-1}    &          & && \\
        &         & \beta_{p-1}  & \alpha_{p} & \beta_{p}& && \\
\hline
        &         &             & \beta_{p}  & \alpha_{p+1} & \beta_{p+1} && \\
        &         &             &           & \beta_{p+1}  & \ddots & \ddots &\\
        &         &             &           &             & \ddots & \ddots &\beta_{d-1} \\
        &         &             &           &             &        & \beta_{d-1} &\alpha_{d}
\end{array}
\right)
\end{equation}
allows us to describe the error bound $\|\E\|_2$ in terms of the submatrix
$\T_{pu}$, and hence in terms of $\beta_p$.
\begin{theorem}
%Theorem 4.1.1
\label{thm:4.1.1}
If $\T_p$ and $\T_{pu}$ are the matrices appearing in~(\ref{eq:4.4}), and if the $p$
columns of $\Q$ are computed by the 
Lanczos algorithm, then there exist $\mu_1, \dots, \mu_p \in \lambda(\T_p)$ and 
$\lambda_1, \dots, \lambda_p \in \lambda(\M)$
such that
\[
|\mu_k - \lambda_k| \leq \|\T_{pu}\|_2 = \beta_p, \quad \text{ for $k=1,\dots,p$.}
\]
\end{theorem}
For a proof, see~\cite[Page 365]{Demmel:1997}.
%
%
% --------------- 26.txt -------------------------
%
%
\section{A Lanczos Procedure for Markov Chains}
% Sec 4.2
\label{sec:lancz-proc-mark}
Notice that the Lanczos algorithm requires the computation of $\M q_j$. Throughout this
paper we have assumed that $\M \in \R^{d\times d}$, and that $d$ is so large as
to make the matrix vector multiplication $\M q_j$ impossible. Now, the Lanczos
algorithm constructs an orthonormal basis for the Krylov subspace. For our
purposes, this construction is not essential. What is essential is that we find
a way to generate the Lanczos coeficients $\alpha_i$, $\beta_j$, $(j = 1,\dots, p)$, and
whence the matrix $\T$, without performing the operation $\M q_j$ required by the
Lanczos algorithm. We now address this problem.

Begin with the centered and weighted observable 
$\psi = \Pi^{1/2} (\phi - \bE_\pi\phi)$,
and let $q_1 = \psi/\|\psi\|_2$.
To simplify notation, let $\phi(n) = \phi(X_n)$. 
The form of the first Lanczos coeﬂicient $\alpha_1$ is straight
forward:
\[
\alpha_1 = q_1^t \M q_1  = \frac{\psi^t\M\psi}{\psi^t\psi}.
\]
and now the special properties of the Markov chain setting play an important
role. Indeed, we have
\begin{align}
\label{eq:4.5}
\psi^t\M\psi &= (\phi - \bE_\pi\phi)^t\Pi^{\frac{1}{2}t}\M\Pi^{\frac{1}{2}}(\phi- \bE_\pi \phi)\nonumber\\
&= \bC_\pi (\phi(n), \phi(n + 1)).
\end{align}
and
\begin{align}
\label{eq:4.6}
\psi^t\psi &= (\phi - \bE_\pi\phi)^t\Pi(\phi- \bE_\pi \phi)\nonumber\\
&= \Var_\pi(\phi(n)).
\end{align}
Recall that the definition of $\M$ implies 
$\Pi^{\frac{1}{2}t}\M\Pi^{\frac{1}{2}}= \Pi \P$, and this justifies 
Equation~(\ref{eq:4.5}). Putting
%
%
% -------------- 27.txt ------------------------------
%
%
these equations together, we have a nice form for $\alpha_1$:
\[
\alpha_1= \frac{\bC_\pi (\phi(n), \phi(n + 1))}{\Var_\pi(\phi(n))}.
\]
The next Lanczos coefficient, $\beta_1$, is only slightly more complicated.
\begin{align}
\label{eq:4.7}
\beta_1 &= \|\M q_1 - \alpha_1 q_1\|_2 \nonumber\\
&= \bigl[(\M q_1 - \alpha_1 q_1)^t(\M q_1 - \alpha_1 q_1)\bigr]^{1/2}\nonumber\\
&= \bigl[q_1^t\M^2 q_1 - \alpha^2_1\bigr]^{1/2}.
\end{align}
Again recalling the definition $\M =\Pi^{1/2}\P\Pi^{1/2}$, a simple manipulation verifies
$\Pi^{\frac{1}{2}t}\M^s\Pi^{\frac{1}{2}} = \Pi \P^s$,
whereby it follows that
\begin{align*}
q_1^t\M^2 q_1  &= \frac{(\phi - \bE_\pi\phi)^t\Pi \P^2(\phi- \bE_\pi \phi)}
{(\phi - \bE_\pi\phi)^t\Pi (\phi- \bE_\pi \phi)} \\[8pt]
&= \frac{\bC_\pi (\phi(n), \phi(n + 2))}{\Var_\pi(\phi(n))}.
\end{align*}
Inserting the last expression into Equation~(\ref{eq:4.7}) gives
\[
\beta_1 = \left[\frac{\bC_\pi (\phi(n), \phi(n + 2))}{\Var_\pi(\phi(n))}
-\left( \frac{\bC_\pi (\phi(n), \phi(n + 1))}{\Var_\pi(\phi(n))}\right)^2
\right]^{1/2}.
\]
Notice that we can estimate these expressions of $\alpha_1$ and $\beta_1$ by
running simulations of the Markov chain. 
To do so, we run the Markov chain starting from a random initial state, compute the value of
the observable at each step of the chain and, from the data, compute estimates of the covariances
and variance of the observable.

Before generalizing this for the $j$th Lanczos coefficient, we consider one more coeﬂicient
in detail. Recall that the second basis element $q_2$ in the Lanczos algorithm is given by 
$q_2 = (\M q_1 — \alpha_1 q_1)/\beta_1$. 
From this we can derive an expression for $\alpha_2$ which, while not as neat as those
%
%
% ----------------- 28.txt ----------------------------------------------------------
%
%
for $\alpha_1$ and $\beta_1$, is nonetheless tractable.
\begin{align}
\label{eq:4.8}
\alpha_2 &= q_2^t \M q_2 \nonumber\\
&= \frac{1}{\beta_1^2}(\M q_1 - \alpha_1 q_1)^t\M (\M q_1 - \alpha_1 q_1)\nonumber\\
&= \frac{1}{\beta_1^2}(q_1^t\M^3 q_1 - 2\alpha_1 q_1^t\M^2 q_1 + \alpha_1^2q_1^t\M q_1 )\nonumber\\
&= \frac{1}{\beta_1^2}(q_1^t\M^3 q_1 - 2\alpha_1 \beta_1^2 + \alpha_1^3 )
%&= \bigl[q_1^t\M^2 q_1 - \alpha^2_1\bigr]^{1/2}.
\end{align}
Equation~(\ref{eq:4.8}) follows from Equation~(\ref{eq:4.7}), which can be
written as $q_1^t\M^2 q_1 = \alpha_1^2 + \beta_1^2$.  
The preceding paragraphs define all but one of the quantities in the last
exression of~(\ref{eq:4.8}). The one remaining
is $q_1^t\M^3 q_1$, which can also be expressed in terms of covariances of $\phi$,
and the analysis is almost identical to that for $q_1^t\M^2q_1$:
\begin{align*}
q_1^t\M^3 q_1  &= \frac{(\phi - \bE_\pi\phi)^t\Pi \P^3(\phi- \bE_\pi \phi)}
{(\phi - \bE_\pi\phi)^t\Pi (\phi- \bE_\pi \phi)} \\[8pt]
&= \frac{\bC_\pi (\phi(n), \phi(n + 3))}{\Var_\pi(\phi(n))}.
\end{align*}
To generalize the foregoing for higher order Lanczos coeﬂicients requires a recurrence
relation between $(\alpha_{j+1}, \beta_j)$ and $(\alpha_{j}, \beta_{j-1})$. 
As we will see, such a relation involves higher order
Lanczos vectors $q_j$---vectors we do not wish to compute. Recall that the only ``vector'' we can
handle is our initial observable. Therefore, we must also establish a recurrence
among the $q_j$'s, thus making them available inductively from our observable vector.

To begin with the $\beta_j$'s, recall that 
$\beta_j = \|\M q_j - \alpha_jq_j - \beta_{j-1}q_{j-1}\|_2$. 
Whence,
\begin{align}
\label{eq:4.9}
\beta_j^2 &= (\M q_j - \alpha_jq_j - \beta_{j-1}q_{j-1})^t(\M q_j - \alpha_jq_j - \beta_{j-1}q_{j-1})\nonumber\\
&= q_j^t\M^2 q_j - \alpha_j^2 - \beta^2_{j-1}.
\end{align}
Some simple algebra and the definitions $\alpha_j = q_j^t\M q_j$ and
$\beta_{j-1} = q_j^t \M q_{j-1}$ prove equality~(\ref{eq:4.9}).  The
%
%
% -------------- 29.txt ------------------------------------------------------
%
%
recurrence for $\alpha_{j+1}$ is equally straight forward, though the final form
is not as neat as the expression above. Indeed,
\begin{align}
\label{eq:4.10}
\alpha_{j+1} &= q_{j+1}^t \M q_{j+1} \nonumber\\
&= \frac{1}{\beta_j^2}(\M q_j - \alpha_j q_j - \beta_{j-1} q_{j-1})^t\M 
  (\M q_j - \alpha_j q_j - \beta_{j-1} q_{j-1})\nonumber\\
&= \frac{1}{\beta_j^2}(q_j^t\M^3 q_j 
    - 2\alpha_j q_j^t\M^2 q_j 
    - 2\beta^2_{j-1} q_{j-1}^t\M^2 q_j 
    + \alpha_1^3
    + 2\alpha_j \beta^2_{j-1} + \beta^2_{j-1}).
\end{align}
From~(\ref{eq:4.9}), $\alpha_j^2 +\beta_j^2 + \beta^2_{j-1}$,
and upon plugging this into~(\ref{eq:4.10}), we arrive at 
\begin{equation}
\label{eq:4.11}
\alpha_{j+1}= \frac{1}{\beta_j^2}(q_j^t\M^3 q_j 
    - \alpha_j^3
    - 2\alpha_j\beta^2_{j}
    - 2\beta_{j-1}q_{j-1}^t\M^2q_j + \beta^2_{j-1}).
\end{equation}

The expressions above involve the higher order Lanczos vectors $q_j$. In the present context,
these vectors will not be readily available. However, another recursion---this time relating forms
involving $q_{j+1}$ to forms involving $q_j$ and $q_{j-1} $---will make these
expressions accessible by induction on our initial observable. The general forms
of interest are $q_{j+1}^t\M^k q_{j+1}$ and $q^t_{j+1}\M^k q_j$. Again, we 
begin with the simpler of the two:
\begin{align*}
  q_{j+1}^tM^kq_j &= 
 \frac{1}{\beta_j}(\M q_j - \alpha_j q_j - \beta_{j-1} q_{j-1})^t\M^k q_j\\
 \frac{1}{\beta_j}(q_j^t\M^{k+1} q_j - \alpha_j q_j^t\M^{k} q_j -\beta_{j-1} q_j^t\M^k).
\end{align*}
The second is,
\begin{align}
\label{eq:4.12}
  q_{j+1}^tM^kq_{j+1} 
&=  \frac{1}{\beta_j^2}(\M q_j - \alpha_j q_j - \beta_{j-1} q_{j-1})^t\M^k 
  (\M q_j - \alpha_j q_j - \beta_{j-1} q_{j-1})\nonumber\\
&=  \frac{1}{\beta_j^2}(q_j^t\M^{k+2} q_j + \alpha^2_jq_j^t\M^{k} q_j 
  + \beta^2_{j-1} q_{j-1}^t\M^k q_{j-1} - 2\alpha_jq_j^t\M^{k+1} q_j \nonumber\\
& \qquad    
    + 2\alpha_j \beta_{j-1}q_j^t\M^k q_{j-1}
    - 2\beta_{j-1}q_j^t \M^{k+1}q_{j-1}).
\end{align}
The point of all this is that each pair of coeﬂicients $(\alpha_{j+1}, \beta_j)$
can be expressed as a function of the covariance between 
$\phi(n)$ and $\phi(n+ m)$, for $m = 0, \dots, 2j+1$. Since the expressions
are based on four recurrence relations, they become quite complicated as $j$ gets large. However,
defining these recursions in a computer program is trivial, as we demonstrate
in Section~\ref{sec:computations} % Sec 5.4
by applying the foregoing to a particular problem.
%
%
% ----------------- 30.txt ------------------------------------------------------
%
%

%%%===================================================================
%%%-------------------------------------------------------------------
\part{Application\label{part:two}}%
%%%-------------------------------------------------------------------
%%%
%%% \input{Application} % Chapter 5
%%%
\chapter{Convergence to Gibbs measure on the Ising lattice}
%Chapter 5
\label{cha:an-appl-conv}

\section{Physical Motivation}
% Sec 5.1
\label{sec:physical-motivation}
Before considering the details of the Metropolis algorithm, it helps to
understand the setting in which it was developed. To do so we consider a
typical example, the \emph{Ising model}. This is a model of a system consisting
of $n$ ``sites,'' each site taking values in 
%% $\{-1, +1\}$. 
$\{0, 1\}$. 
It might 
help to imagine these sites as equally spaced points on a line or a circle,
though we are not restricted to such cases. 
%% Then our state space $S$ consists of
%% all possible $n$-dimensional permutations of $-1$'s and $+1$'s. 
The state space $S$ consists of
all possible $n$-dimensional permutations of $0$'s and $1$'s. 
%% Thus, $S = \{-1, +1\}^n$ and $|S| = 2^n$. We can think of each
Thus, $S = \{0, 1\}^n$ and $|S| = 2^n$. We can think of each
site as a node existing in one of two possible positions, say ``on'' or
``off.'' Each state $i \in S$ is %represented by 
a unique permutation $\sigma_i$ of of $0$'s and $1$'s. 
Let $\sigma_n(k)$ denote the position %(on or off) 
of the $k$th node when the system is in state $i$. 
An observable $\phi = \phi(\sigma_i)$ on this space is a function of
the permutations $\sigma_i$, $i\in \{1,\dots, 2^n\}$.
%
%
% ---------------- 31.txt ------------------------------------------------------
%
%

The \emph{energy} of the system, when in state $i$, is defined by the Hamiltonian
\[
H(\sigma_i) = - \sum_{\<j k\>} \sigma_i(j)\sigma_i(k)
\]
where the sum runs over all neart neighbor pairs. In statistical mechanics one is often concerned
with the \emph{Gibbs measure} of state $i$, which is defined by
\begin{equation}
\label{eq:5.1}
\pi(\sigma_i)= Z^{-1} \exp\{-\beta H(\sigma_i)\}
\end{equation}
where $Z^{-1}$ is a normalizing constant, sometimes called the \emph{partition function}. 
For the Ising model, we put  
\begin{equation*}
Z = \sum_{i=1}^{2^n}\exp\{-\beta H(\sigma_i)\}
\end{equation*}
so that $\pi$ is a probability measure. 
Now, letting $k$ denote the so called \emph{Boltzman constant}, if a
classical mechanical system is in thermal equilibrium with its surroundings with
absolute temperature $T$, and is in state $i$ with energy $H(\sigma_i)$, then the
probability density in phase-space of the point representing state $i$ is given by
(\ref{eq:5.1}) with $\beta = (kT)^{-1}$. A fundamental result of ergodic theory 
implies that we can also interpret $\pi(\sigma_i)$ as the proportion of time the
system spends in state $i$. If the system is observed at a random time, the
expected value $\bE_\pi\phi$ of any observable $\phi$ is thus 
\begin{equation}
\label{eq:5.2}
\bE_\pi\phi = 
\sum_{i=1}^{2^n}\phi(\sigma_i) \pi(\sigma_i) = 
%% \frac{\sum_{i=1}^{2^n}\phi(\sigma_i) \exp\{-\beta H(\sigma_i)\}}
%% {\sum_{i=1}^{2^n} \exp\{-\beta H(\sigma_i)\}}
Z^{-1}\sum_{i=1}^{2^n}\phi(\sigma_i) \exp\{-\beta H(\sigma_i)\}.
\end{equation}

Perhaps the number of sites $n$ in our Ising model is so large that it is
impractical or impossible to evaluate~(\ref{eq:5.2}).
We might instead consider importance sampling and generate states with
the probability density $\pi$ given in~(\ref{eq:5.1}).
Then $\phi$ is itself an unbiased estimator of~(\ref{eq:5.2}).

If evaluation of~(\ref{eq:5.2}) is difficult, there is no reason to believe that
evaluation of~(\ref{eq:5.1}) is any easier. However, Metropolis and his
collaborators~\cite{Metropolis:1953} contrived a method for producing an ergodic 
%
%
% ------------------ 32.txt --------------------------------------------------
%
%
Markov chain with transition probability matrix K whose elements $k(x,y)$ satisfy
\[
\sum_x \pi(x) k(x,y) = \pi(x)
\]
By 
%Definition 2.1.4 
Definition~\ref{def:2.1.4}
this means that the Markov chain has stationary distribution $\pi$. Since the
chain is ergodic, we know that it will eventually converge to the desired
distribution. So, our main 
problem---the rate at which such a Markov chain will converge---is of primary
concern in applications of the Metropolis algorithm. Moreover, as we will see
below, the Metropolis algorithm ensures that the Markov chain has the desired
stationary distribution by requiring that the chain satisfy the stronger
condition of reversibility, or ``detailed balance.'' Therefore, Markov chains
produced by the Metropolis algorithm satisfy the assumptions of this paper,
and our theory and methods can be used to bound convergence rates of such chains.

\section{Description of the Metropolis Algorithm}
% Sec 5.2 
\label{sec:descr-metr-algor}
The following explains how the Metropolis Algorithm is carried out. The exposition is
similar to that given in Hammersley and Handscomb's book 
\emph{Monte Carlo Methods} \cite{Hammersley:1964}. We refer the reader desiring
more details to~\cite[Section 9.3]{Hammersley:1964}.

To begin, choose an arbitrary symmetric Markov chain; that is, a Markov chain whose
transition probability matrix P is symmetric. The matrix P is called the 
\emph{proposition matrix}, and its elements
%, $p(i,j)$,
satisfy, for all $i$ and $j$ in $S$,
\begin{equation}
\label{eq:5.4}
p(i,j) \geq 0, \quad p(i,j) = p(j,i), \quad \sum_m p(i,m)=1.
\end{equation}
Using P, we will derive a new transition probability matrix K with elements
$k(i,j)$ satisfying
\begin{equation}
\label{eq:5.5}
\sum_i \pi(i)k(i,j) = \pi(j)
\end{equation}
For $i\neq j$ define
\begin{equation}
\label{eq:5.6}
k(i,j) = 
\begin{cases}
p(i,j)\pi(j)/\pi(i), & \text{ if $\pi(j)/\pi(i) < 1$,}\\
p(i,j), & \text{ if $\pi(j)/\pi(i) \geq 1$.}
\end{cases}
\end{equation}
%
%
% ----------------- 33.txt ----------------------------------------------
%
%
For $i=j$ define
\begin{equation*}
k(i,j) = 
p(i,i) + \sum_{j\in J_i}\bigl(1 - \frac{\pi(j)}{\pi(i)}\bigr)p(i,j), 
\end{equation*}
where $J_i = \{j : j\neq i, \pi(j)/\pi(i) < 1\}$.
Notice that $\pi(i) > 0$ implies $k(i,j) \geq 0$, and
\begin{align}
\label{eq:5.7}
\sum_j k(i,j) 
&= p(i,i) + 
\sum_{j\in J_i}\left\{\bigl(1 - \frac{\pi(j)}{\pi(i)}\bigr)p(i,j)
+ \frac{p(i,j)\pi(j)}{\pi(i)}\right\} 
+ \sum_{j\in J_i^c}p(i,j)
\nonumber\\
&= p(i,i) + 
\sum_{j\in J_i}p(i,j)
+ \sum_{j\in J_i^c}p(i,j).
\end{align}
Equation~(\ref{eq:5.7}) shows that for each $i$ we have $\sum_j k(i,j) = 1$
(the row sums of K are 1), which confirms that K is a stochastic matrix.

Next we check that K satisfies the following \emph{detailed balance} condition:
\begin{equation}
\label{eq:detailed-balance}
\pi(i)k(i,j) = \pi(j)k(j,i).
\end{equation}
Obviously this holds when $i=j$, so assume $i\neq j$.

First suppose $\pi(i) = \pi(j)$.
In this case, the definition~(\ref{eq:5.6}) and symmetry of P imply that
\[
k(i,j) = p(i,j) = p(j,i) = k(j,i).
\]
Therefore, when $\pi(i) = \pi(j)$ the detailed balance
condition~(\ref{eq:detailed-balance}) clearly holds.

Now suppose that $\pi(i) > \pi(j)$. Then the second case in definition (\ref{eq:5.6}) implies
$k(j,i) =p(j,i)$, while the first case gives
\[
k(i,j) = p(i,j)\pi(j)/\pi(i) =  p(j,i)\pi(j)/\pi(i) = k(j,i)\pi(j)/\pi(i),
\]
and we see that 
$\pi(i)k(i,j) = \pi(j)k(j,i)$ holds.
Finally, a symmetric argument shows that if $\pi(j) > \pi(i)$,
then detailed balance is again satisfied.

To complete the justification of the Metropolis algorithm, we note
detailed balance~(\ref{eq:detailed-balance})
implies that $\pi$ is indeed the stationary distribution for a Markov 
chain with transition probability matrix K; that is, (\ref{eq:5.5}) is
satisfied. Indeed, by the detailed balance condition and the fact that
the row sums of K are 1, we have
\begin{equation*}
\sum_i \pi(i)k(i,j)
=\sum_i \pi(j)k(j,i)
=\pi(j)\sum_i k(j,i)
 = \pi(j).
\end{equation*}

We now summarize the tasks performed when implementing the Metropolis algorithm.
First, pick as a proposition matrix any symmetric stochastic matrix and denote
it by P. Begin 
%
%
% ----------- 34.txt --------------------------------------------------------
%
the simulation of the Markov chain in state $i$, and then take one step of the
proposition Markov chain to arrive in state $j$. That is, choose state $j$
according to the probability mass function $p(i, \cdot)$. 
Next, compute $\pi(j)/\pi(i)$. 
If $\pi(j)/\pi(i) \geq 1$, accept $j$ as the new state. If $\pi(j)/\pi(i) < 1$,
then with probability $\pi(j)/\pi(i)$ accept $j$ as the new state; otherwise
(with probability $1- \pi(j)/\pi(i)$), take $i$ as the new state. Performing
each of these tasks produces one step of the Markov chain represented by the 
transition probability matrix K.

\section{The Ising Model}
% Sec. 5.3 
\label{sec:ising-model}
Returning to the Ising model considered in
Section~\ref{sec:physical-motivation}, we consider how the Metropolis 
algorithm is applied to such a model. Again, start the simulated chain in state
$X_0 = i$ and let $X_1 = j$ be a state chosen according to 
$p(i, \cdot)$, where $p(i,j) = p(j,i)$. Next, compute 
\[
\pi(\sigma_j)/\pi(\sigma_i) = \exp\{-\beta[H(\sigma_j) - H(\sigma_i)]\}
\]
Suppose the energy $H(\sigma_j)$ of the new state satisfies 
$H(\sigma_j) \leq H(\sigma_i)$.
Then $\pi(\sigma_j)/\pi(\sigma_i)$ is at least 1 and we move to state $j$.
On the other hand, if
$H(\sigma_j) > H(\sigma_i)$,
then $\pi(\sigma_j)/\pi(\sigma_i)< 1$ and we take
\begin{equation}
\label{eq:5.8}
X_1 = 
\begin{cases}
j & \text{ with probability $e^{-\beta\Delta H}$},\\
i & \text{ with probability $1-e^{-\beta\Delta H}$}.
\end{cases}
\end{equation}
where $\Delta H = H(\sigma_j) - H(\sigma_i)$ is the change in energy from state $j$ to state $i$.

There are many ways to choose a proposition matrix P. The simplest derives from the
so called \emph{Glauber dynamics}, which proceeds as follows: choose one of the
$n$ sites at random with probability $1/n$---let $s$ denote the chosen
site---and change the position of this site; i.e., subtract its value from 1.
%multiply its value by $-1$. 
The resulting state $j$ is the 
%permutation of 
%-1’s and +1’s
% 0's and 1's 
binary string
given by
\begin{equation*}
\sigma_j(k) = 
\begin{cases}
1-\sigma_i(k) & \text{ if $k =s$},\\
\sigma_i(k) & \text{ if $k =s$},
\end{cases}\quad (k = 1, \dots, d).
\end{equation*}
%
%
% -------------- 35.txt ---------------------------------------------------------
%
%
The procedure we employ below involves a slight variation: choose site $s$ at
random uniformly from the $n$ possible sites. However, instead of changing the
value of the chosen site with absolute certainty, as in Glauber dynamics, we
``flip the switch'' with probability $1/2$. 
Therefore, when $k \neq s$
we have $\sigma_j(k) = \sigma_i(k)$, and when $k = s$,
\begin{equation*}
\sigma_j(s) = 
\begin{cases}
1-\sigma_i(s) & \text{ with probability 1/2},\\
\sigma_i(s) & \text{ with probability 1/2}.
\end{cases}
\end{equation*}
The advantage of this procedure is that the proposition matrix, before altering
it with the Metropolis algorithm, is a bit more interesting than that produced
by traditional Glauber dynamics. For, it has 1/2 along the main diagonal and is
thus aperiodic and converges to a stationary distribution. Glauber dynamics has
period 2, so its proposition matrix does not satisfy the hypotheses of our
convergence theorem (Theorem~\ref{thm-2.1.2}). Keep in mind, however, that any
proposition matrix modified by the Metropolis algorithm satisfies our hypotheses.
\section{Computations}
% Sec 5.4
\label{sec:computations}
\subsection{Computer Programs}
Writing a computer program to simulate realizations of the Markov chain described
above (i.e. with transition matrix K) is straight forward, and we do so using
Matlab and the program {\tt ising.m} which appears in the Appendix.

When performing simulations in the present context, there are a few important
aspects to consider. First, our program begins with a ``hot start,'' which means
that the initial state is picked by assigning each site to the value 0 or 1, each
with probability 1/2.\footnote{We cannot simply draw our initial state
from the stationary distribution since the point of this work
is to deal with cases for which samples from the stationary distribution are not
immediately attainable. Furthermore, we are trying to determine how long it will
take for observations from the simulated chain to represent samples from the
stationary distribution.}
%% figure out how long it will take for observations from the simulated chain to represent samples from
%% the stationary distribution. 
%% Perhaps we should instead have picked our initial state
%% from the stationary distribution. On the contrary, this would defeat the
%% whole purpose of our analysis.  
%% Recall that the method has been developed to deal with
%% cases for which 
%
%
% ----------------- 36.txt --------------------------------------------
%
%
%% sampl from the stationary distribution are not quickly attainable. Furthermore, we are trying to
%% figure out how long it will take for observations from the simulated chain to represent samples from
%% the stationary distribution.

%% sampl from the stationary distribution are not quickly attainable. Furthermore, we are trying to
%% figure out how long it will take for observations from the simulated chain to represent samples from
%% the stationary distribution. 
%% Second, t

The method we have developed depends solely on covariances
of our observable, which we estimate from the data produced by the
simulations. However, such estimates require that the data come from the
stationary distribution. Therefore, we must discard a number of observations
before using the data to estimate covariances. The number of observations 
to discard depends on how long must we wait 
before we can expect the data to represent samples from the stationary distribution.
%% until we have data from the stationary distribution,
%% and we are back where we started — almost.
When employing our procedure, we can discard a
conservative (very large) number of observations and perform our analysis, deriving a bound on
the convergence time. Then, in all future studies, we will know how much data should be discarded
before samples can be assumed to have come from the stationary distribution.

In considering what observable to use on the state space described above, perhaps the
most obvious is simply the number of nodes in the ``on'' position (i.e., the
number of 1's). Recalling that $\sigma_j(k)\in \{0,1\}$ denotes the position of
the $k$th node when the system is in state $j$, this observable is simply the
sum $\phi(j) = \sum_k\sigma_j(k)$. 
%% where X is the indicator function. 
The program {\tt ising.m} simulates the Markov chain, computes the
values of this observable (and its square) and writes these data to a file
called {\tt phi.dat}. 
The program then computes the first few Lanczos coefficients using covariances
produced by Matlab, so that we can check them against the results we get from the
{\tt lanczos.c} program, which we now describe.

The lanczos.c program (listed, along with its dependencies, in the Appendix) implements
the recursive relations described in Section~\ref{sec:lancz-proc-mark} % Sec 4.2
to compute the Lanczos coefficients by the new
method. It accepts input from the user specifying any observable, any number of
iterations, and any 
%
%
% ---------------  37.txt  -----------------------------------------------
%
%
number of desired Lauczos coefiicients. That is, the observable could have come from a simulation
of any reversible Markov chain and not just our special Metropolis chain. The output of this
program is a matrix T containing the maximum number of Lanczos coefficients (up
to the number requested by the user) before an approximate invariant subspace
was reached. The eigenvalues of the nonzero part of this matrix, in accordance
with the theory developed above, should provide a close approximation to the
eigenvalues of the Markov chain's stochastic matrix.

\subsection{Results}
% Sec. 5.4.2 
\begin{table}
\centering
\label{tab:1}
\caption{Estimates of $\lambda_{\max}(\K) = 0.998746$ using $\phi_1$.}
\vskip4mm
\begin{tabular}{r|c|c|c|c}
simulation \# & $\alpha_1$ & $\beta_1$ & $\lambda_{\max}(\T_2)$& $\beta_2$\\
\hline
1& 0.995107& 0.006960& 0.995220& 0.600243\\
2& 0.995099& 0.005891& 0.994863& 0.477393\\
3& 0.994862& 0.009671& 0.994925& N/A\\
4& 0.995321& 0.009319& 0.995654& N/A\\
5& 0.994787& 0.004872& 0.405446& 0.994742
\end{tabular}
\end{table}
Table~\ref{tab:1} shows the results of 5 simulations each of 100,000 iterations (discarding the first
5,000 observations) for the one dimensional Ising lattice with 10 nodes and $\beta= 1$. Displayed is
the estimate of the first Lanczos coefficient $\alpha_1$ (which is the
eigenvalue estimate after one step; i.e., $\lambda_{\max}(\T_1)$), along with its
error bound $\beta_1$. Appearing in the third column is the largest eigenvalue,
$\lambda_0(T_2)$, of the $2 \times 2$ matrix $\T_2$, and $\beta_2$ appears in the
fourth column when it is was computable. When it was not computable, it was
%% probably 
because an approximate invariant subspace was reached 
at the first step, so $\beta_2$ involved expressions which were close to machine
epsilon and thus could not be computed accurately. Table~\ref{tab:2}  shows the same
information when using a second observable, 
$\phi_2 = \phi_1^2$ (i.e., the number of 1's squared).

\begin{table}
\centering
\label{tab:2}
\caption{Estimates of $\lambda_{\max}(K) = 0.998746$ using $\phi_2$}
\vskip4mm
\begin{tabular}{r|c|c|c|c}
simulation \# & $\alpha_1$ & $\beta$ & $\lambda_{\max}(\T_2)$& $\beta_2$\\
\hline
1& 0.994262& 0.006142& 0.994222& N/A\\
2& 0.994333& 0.002974& 0.994332& N/A\\
3& 0.994307& 0.008892& 0.994374& 0.726320\\
4& 0.994436& 0.012884& 0.994685& 0.361404\\
5& 0.993789& 0.006909& 0.993820& 0.788710
\end{tabular}
\end{table}
The true value of the subdominant eigenvalue for this problem is $\lambda_{\max}(\K) = 0.998746$,
and the next largest is $\lambda_2(\K) = 0.993303$. These were computed directly
from the transition matrix 
%
%
% ---------- 38.txt ----------------------------------------------
%
%
using the Matlab program {\tt tpm.m} listed in the Appendix. They should be very accurate, and it
takes the Matlab routine {\tt eig()} just over 20 minutes of cpu time on a Sun
Ultra Sparc to find all the eigenvalues of K.

As montioned in Section~\ref{sec:gener-lancz-algor} (see
Theorem~\ref{thm:4.1.1}), the $\beta$ coefficients provide conservative error 
bounds on the eigenvalue estimates from the previous Lanczos
step. Unfortunately, for this example $\beta$ appears to be too conservative as
it does not even allow us to bound our estimates away from 1. Furthermore, it  
seems that the estimates generated by the method above fall systematically below
the true value of the subdominant eigenvalue. In every simulation, an
approximate invariant subspace (to machine tolerance) was reached after
computing at most 2 pairs of Lanczos coefficients. For $\phi_1$ we see that 
it was reached after computing only one pair on simulations 3 and 4, similarly
for $\phi_2$ on simulations 1 and 2. These observations indicate that the chosen
observable is close to the slowest mode (or 
eigenfunction) of the process. That is, we should observe fast convergence to an approximate
invariant space. However, the space found might contain only the second slowest mode of the
process and, in that case, our eigenvalue estimates would be closer to the third largest eigenvalue,
instead of the second largest.\footnote{Recall from above that we won't
converge to the eigenspace corresponding to the largest eigenvalue 
(which is always 1), unless we choose a nearly constant observable function.}

The matrix T was produced by the program {\tt lanczos.c} and its eigenvalues were computed
by Matlab, both operations taking a few seconds. Granted, comparing a few seconds to the 20
minutes it takes Matlab to compute all the eigenvales is not really fair, 
since calling Matlab's {\tt eig()} routine is not always the
%
%
% --------- 39.txt -----------------------------------------------------
%
%
optimal traditional method for computing only a few eigenvalues of K. However, recall that
the primary motivation for the new method are those examples where the matrix K is so large
that we can't store even a single column vector in fast memory. Once we start
swapping data to and from slow memory, traditional algorithms can become
intractable, and such examples are easy to come by. For instance, if the Ising
model described has 1000 nodes, it produces a $2^{1000}\times 2^{1000}$
transition probability matrix.
%
%
% 40.txt
%
%

%%%-------------------------------------------------------------------
%%%
%%% \input{Conclusion}
%%%
\chapter{Conclusion}
% Chapter 6
\label{cha:conclusion}
We conclude with some remarks about the special situations in which our
method is useful.
%few aspects of the cases to which the foregoing applies.
We are interested in the subdominant eigenvalue of a large transition probability matrix. Our
method can be used to approximate these eigenvalues when the following conditions are satisfied:
\begin{itemize}
\item 
The Markov chain is reversible (satisfies detailed balance).
\item
We can simulate realizations of the chain as well as functions of these
realizations (observables on the state space).
\end{itemize}
When these conditions hold, we can avoid the traditional Lanczos algorithm, and
estimate the values of the Lanczos coeﬂicients via the variational properties of
observables on the state space. 
There are a number of ideas in this paper that could be expanded upon in future
studies.  Perhaps most obvious would be to try to find analogous techniques for
nonreversible (i.e. nonsymmetric) Markov chains. This is a considerable problem
since the Lanczos methods described above are no longer applicable. Perhaps the
\emph{Arnoldi algorithm} (see \cite[Algorithm 6.9]{Demmel:1997}), or the
\emph{nonsymmetric Lanczos algorithm} could also be modified to exploit
special properties of stochastic matrices. %%  (as we have done with the Lanczos
%% algorithm in the symmetric case). 
%
%
% 41.txt
%
%

When applying the Lanczos algorithm, we usually choose a single starting vector (in the
present context, an observable), perform the algorithm for a number of steps, and gather results.
Then we do the same with a new starting vector and compare the results to those obtained with
the first vector. Proceeding in this way for a number of starting vectors, we can provide statistical
evidence that we have, indeed, located the eigenvalues of interest. However, when there are two
observables, $\phi_1$ and $\phi_2$, of particular interest, perhaps we would
benefit from considering both observables simultaneously. Future studies might
develop an algorithm which incorporates both observables, 
again via $\psi_i = \Pi^{1/2}(\phi_i - \bE_{\pi}\phi_i)$. %% In this case, a
The Lanczos algorithm would then be based on the space 
\[
\sK^*(M,\psi_1, \psi_2, p) =  \ran\{\psi_1, \psi_2, \M\psi_1, 
\M\psi_2, \dots, \M^{p-1}\psi_1, \M^{p-1}\psi_2\}
\]
The Lanczos coefficients would then involve not only the autocovariances,
%$\bC_\pi(\phi_i(n), \phi_i(n+m))$
but also the cross-covariances, 
$\bC_\pi(\phi_i(n), \phi_j(n+m))$, defined in (2.6) above.

\newpage

%%%%% Appendices start %%%%%%%%%%%%%%%%
%% Comment out the following line if your thesis has no appendix
\singlespacing
\appendix
%%%-------------------------------------------------------------------
%%%
%%% \input{Appendix}
%%%
\begin{center}
  {\Large Appendix: computer programs}
\end{center}

{\small 
\begin{verbatim}
% Matlab code ising.m
%
% Written by William J. DeMeo on 12/15/97
% last modified 2013.10.19
% Inputs:
%         d = number of nodes of the ising lattice (e.g. d=100)
%         n = number of iterations (e.g. n=1000)
%         beta = Annealing schedule (e.g. beta = 2,
%            when beta -> 0 will always go to nee state
%            when beta -> infty will never go to state of higher energy)
%
X = zeros(d,1);
Energy = zeros(n,1);
Ave=0;

% Initialize using hot start
for i=1:d,
  if rand < .5
    X(i) = -1;
  else
    X(i) = 1;
  end
end

% Initialize at state of high energy (not used)
if 0
  X(1:2:d-1)=1;
  X(2:2:d)=—1;
end

% Initialize energy
H=0;
for i=1:(d-1)
  H = H - X(i)*X(i+1);
end

% Initialize observable
phi = zeros(n+1,1);

for i=1:d
  if X(i) == 1
    phi(1)= phi(1)+1;
  end
end

for j=1:n
  site=0;
  while site==0
    site = round(rand*d);
  end
  % Compute new energy (with new X(site) = -X(site)):
  if site == 1
    Hnew = H + 2*X(1)*X(2);
  elseif site == d
    Hnew = H + 2*X(d)*X(d-1);
  else
    Hnew = H + 2*X(site)*(X(site-1) + X(site+1));
  end

  % Change to new state in two ways:
  % with probability 1/2, flip the switch
  if rand < .5
    if (Hnew <= H) | (rand < exp(-beta*(Hnew-H)))
      X(site) = -X(site);
      H = Hnew;
      % Compute new value of observable
      if X(site) == 1
        phi(j+1) = phi(j) + 1;
      else
        phi(j+1) = phi(j) -1;
      end
    else phi(j+1)=phi(j);
    end
  else phi(j+1)=phi(j);
  end
  Energy(j) = H;
end
plot(Energy)

% A second observable
phi2 = zeros(n+1,1);
for i=1:n+1
  phi2(i) = phi(i)*phi(i);
end

% Write first observable to phi.dat:
fid = fopen('phi.dat','w');
fprintf(fid,'%f\n',phi);
fclose(fid);

% Write second observable to phi2.dat:
fid = fopen('phi2.dat','w');
fprintf(fid,'%f\n',phi2);
fc1ose(fid);

% Compute a few Lanczos coefficients by new method
START=1001;
V1=zeros(2);
C1=zeros(2);
C2=zeros(2);
C3=zeros(2);

V1 = cov(phi(START:n+1),phi(START:n+1));
V1 = V1(1,1);
C1 = cov(phi(START:n),phi(START+1:n+1));
C1 = C1(1,2);
C2 = cov(phi(START:n-1),phi(START+2:n+1));
C2 = C2(1,2);
C3 = cov(phi(START:n-2),phi(START+3:n+1));
C3 = C3(1,2);

alpha1 = C1/V1;
beta1 = sqrt(C2/V1 -(C1/V1)^2);
alpha2 = (1/(beta1^2))*(C3/V1 - 2*alpha1*(C2/V1) + alpha1^3);

T1 = [alphal beta1; beta1 alpha2]
V1 = cov(phi2(START:n+1),phi2(START:n+1));
V1 = V1(1,1);
C1 = cov(phi2(START:n),phi2(START+1:n+1));
C1 = C1(1,2);
C2 = cov(phi2(START:n-1),phi2(START+2:n+1));
C2 = C2(1,2);
C3 = cov(phi2(START:n-2),phi2(START+3:n+1));
C3 = C3(1,2);

alpha1 = C1/V1;
beta1 = sqrt(C2/V1 - (C1/V1)^2);
alpha2 = (1/(beta1^2))#(C3/V1 - 2*alpha1*(C2/V1) + alpha1^3);

T2 = [alpha1 beta1; beta1 alpha2]

fid = fopen('Tmats.dat','v');
fprintf(fid,'%f\n',T1);
fprintf(fid,'%f\n',T2);
fclose(fid);

%%% end ising.m

\end{verbatim}}

\vskip2cm
%
%
% -------------- 45.txt ------------------------------------------
%
%
{\small 
\begin{verbatim}
% Matlab code tpm.m for generating transition probability matrix
% and computing its eigenvalues directly
%
% Written by William J. DeMeo on 1/10/98
%
% Inputs:
%        d = number of nodes of the ising lattice (e.g. d=10)
%     beta = Annealing schedule (e.g. beta - 2,
%            when beta -> 0 will always go to new state
%            when beta -> infty will never go to state of higher energy)

disp('bui1ding proposition matrix...')

states = 2^d;
A = 0;
for i=0:(d-1)
  E=eye(2^i);
  A = [A E; E A];
end

B = .5*eye(states);
A = B + .5*(1/d)*A;
% our modified Glauber dynamics requires the B and the .5*(1/d)

disp('...done')

% display pattern of nonzero entries
% spy(A)

% check that all row sums are 1
check=O;
for i=1:states
  check = check + (not(sum(A(i,1:states))<.99));
end
% if not all rows sum to 1, print check = (# of rows with sum=1)
if not(check==states)
  check
  error('Row sums are not all 1')
end

% construct matrix of states
E = zeros(d,states);
flip=-1;
for i=1:d
  for j=1:2^(i-1):states
    flip = -1*flip;
    for k=0:2^(i-1)
      E(i,j+k) = flip;
      end
  end
end
E=E';

% display first 64 states
% E(1:64,:)

% compute energy of each state
H = zeros(states,1);
for i=1:states
  for j=1:d-1
    H(i) = H(i) - E(i,j)*E(i,j+1);
  end
end

disp('building Hetropolis transition matrix...')

for i=1:states
  for j=i+1:states
    if not(A(i,j)==0)
      if(H(j)>H(i))
        alt = A(i,j)*(1-exp(-beta*(H(j)-H(i))));
        A(i,i) = A(i,i)+a1t;
        A(i,j) = A(i,j)-alt;
      elseif(H(j)<H(i))
        alt = A(j,i)*(1-exp(-beta*(H(i)-H(j))));
        A(j,j) = A(j,j)+a1t:
        A(j,i) = A(j,i)-alt;
      end
    end
  end
end

disp('...done')

check=0;
for i=1:states
  check = check + (not(sum(A(i,1:states))<.99));
end

% if not all rows sum to 1, print check = (# of rows with aum=1)

if not(check==states)
  check
  error('Row sums are not all 1')
end

disp('computing eigenvalues of tpm...')
cput = cputime;
evals = eig(A);
ecput = tputime - cput;
disp('the CPU time (in secs) for computing eigenvalues of tpm: ')
ecput

%%% end tpm.m
\end{verbatim}}

\vskip2cm

% 50.txt

{\small 
\begin{verbatim}

/*************************************************************
 * lanczos.c main program for computing Lanczos coefficients *
 *                                                           *
 * Created by William J. DeMeo on 1/7/98                     *
 * Last modified 2013.10.19                                  *
 ************************************************************/

#include <stdlib.h>
#include <stdio.h>
#include <math.h>
#include "prototypes.h"

/* Machine constants on dino (Sun Ultra 1 at Courant) */
#define MACHEPS 1.15828708535533604981e-16
#define SQRTEPS 1.07623746699106147048e-08
#define MAX_NAME 100

void read_name(char *);

/* Prototypes from the file functions.c */
double alpha(int j);
double betasq(int j);
double form1(int j, int k);
double form2(int j, int k);
double moment(double *data, long n, 
              double *ave, double *var, double *cov, long k);

/* external variables to be used by functions */
double *phi,*cov, *var, *ave;

main()
{

  char *filename;
  FILE *ofp;
  double temp=O, *T;
  long i, j, nrow, nlanc, START;
  int flag=O;

  filename = cmalloc(MAX_NAME);

  printf("\nName of file containing observed values: ");
  read_name(filename);
  printf("\nTota1 number of observations in file (iterations): ");
  scanf("%u",&nrow);
  printf("\nNumber of leading observations to discard: ");
  scanf("%u",&START);
  printf("\nNumber of Lanczos coefficients desired: ");
  scanf("%u",&nlanc);
  while(2*nlanc > nrow-START)
    {
      printf("\nNot enough data for that many coefficients.\n");
      printf("\nEnter a smaller number of Lanczos coefficients: ");
      scanf("%u",&nlanc);
    }

  phi = dmalloc(nrow);
  cov=dmalloc(2*nlanc);
  var=dmalloc(1);
  ave=dmalloc(1);
  T = dmalloc(nlanc*nlanc); 
  for(i=0;i<nlanc;i++) 
    T[i]=(double) 0;

  /* observable phi is stored contiguously column-wise by MATLAB */
  matlabread(phi, nrow, 1, filename);

  /* send the observable, offset by START */
  moment(phi+START,nrow-START,ave, var, cov,2*nlanc-1);

  /* First column of T */
  T[0] = alpha(1);
  if((temp=betasq(1))>MACHEPS*10)
    {
      T[1]=sqrt(temp);

      /* General column of T */
      j=2:
      for(i=1; i<nlanc-1, flag==O;)
	{
	  T[i*nlanc+i-1]=sqrt(betasq(i));
	  T[i*nlanc+i] = alpha(i+1);
	  if((temp = betasq(i+1))>MACHEPS*10)
	    {
	      T[i*nlanc+i+1] = sqrt(temp);
	      i++;
	    }
	  else
	    flag=1;
	}

      /* Last column */
      if(flag!=1 && ((temp= betasq(nlanc-1))>MACHEPS*10))
	{
	  T[nlanc*nlanc-2] = sqrt(temp);
	  T[nlanc*nlanc-1] = alpha(nlanc);
	  printf("\nbeta(1) = %1f\nbeta(%d) = %lf (last beta)",
		 T[1],(nlanc-1),T[nlanc*nlanc-2]);
	}
      else
	{
	  printf("\nApproximate invariant space reached at step %d.",i);
	  printf("\nbeta(1) = %lf\nbeta(%d) = %lf (last accurate beta)".
		 T[1],i,T[i*nlanc+i-1]);
	  printf("\nbeta(%d)^2: %lf (first spurious resu1t)",i+1,temp);
	}
      printf("\nThe matrix T is: \n");
      matprint(T,nlanc,nlanc);
    }
  else
    {
      printf("\nmain(): Approximate invariant space reached at first step.");
      printf("\nalpha(1) - %lf\nbeta(1)^2: %lf (first spurious result)", 
             T[O], temp);
    }
  ofp = fopen("Tmat.m","v");
  check(ofp);
  matlabwrite(T,nlanc,nlanc,ofp);
  fclose(ofp);
}

void read_name(char *name)
{
  int c, i = 0;
  while ((c = getchar()) != EOF && c != ' ' && c != '\n')
    name[i++] = c;
  name[i] = '\0';
}

/** end lanczos.c **/

\end{verbatim}}

\vskip2cm

{\small 
\begin{verbatim}

/***************************************************
 * functions.c -- functions required by lanczos.c  *
 *                                                 *
 * Created by William J. DeMeo on 1/7/98           *
 * Last modified 2013/10/19                        *
 ***************************************************/

#include <math.h>
#define START 1000
#define ITER 10000

/* Machine constants on dino (Sun Ultra 1 at Courant) */
#define MACHEPS 1.15828708535533604981e-16
#define SQRTEPS 1.07623746699106147048e-O8

double alpha(int j);
double betasq(int j);
double form1(int j, int k);
double form2(int j, int k);
double moment(double *data, long n, 
              double *ave, double *var, double *cov, int k);

/* external variables to be used by functions */
double *phi,*cov, *var, *ave;

double alpha(int j)
{
  /* alpha is never called with j < 1 */
  if(j==1)
    return form1(1,1);
  else if(j>1)
    return
      ((double)1/betasq(j-1)) * (form1(j-1,3) - pow(alpha(j-1),3)
       - 2 * (alpha(j-1)*betasq(j-1) + sqrt(betasq(j-2))*form2(j-1,2))
       + betasq(j-2));
}

double betasq(int j)
{
  if(j==0)
    return (double)O;
  else if(j>0)
    return (form1(j,2) - pow(alpha(j),2) - betasq(j-1));
}

double form1(int j, int k)
{
  double form14, alpha1, alpha1sq, form12, form13;
  if(j==0)
    return (double)0;
  else if(j==1)
    {
      /* printf("\nvar = %lf, cov(%d) = %lf \n",*var,k,cov[k]); */
      return (cov[k])/(*var); /* the only real value */
    }
  else if(j>1)
    {
      return 
	((double)1/betasq(j-1))
	* ( form1(j-1,k+2) + pow(alpha(j-1),2) * form1(j-1,k)
	    + 2*(alpha(j-1) * sqrt(betasq(j-2)) * form2(j-1,k)
		 - alpha(j-1)*form1(j-1,k+1)
		 - sqrt(betasq(j—2))*form2(j-1,k+1) )
	    + betasq(j-2)*form1(j-2,k));
    }
}

double form2(int j, int k)
{
/* form2 is never called with j < 1 */
  if(j==1)
    return (double)0;
  else if(j>1)
    return
      (pov(betasq(j-1),-.5)) 3
      (form1(j-1,k+1) - alpha(j-1)*form1(j-1,k)
       - sqrt(betasq(j-2)) * form2(j—1,k));
}

/* moment() function for computing var and cov(k)
   arguments:
   data = a nxi array of doubles
   n = length of data[]
   ave =(on exit)= the average of data[]
   var =(on exit)= the variance of data[]
   cov =(on exit)= the covariance of data[i] and data[i+j] for i-1,...,k
   k = the max lag for cov above
*/
double moment(double *data, long n, 
              double *ave, double *var, double *cov, int k)
{
  /* Centered about data[0] algorithm: */
  long i,j;
  double ave1, ave2;

  *ave=O;*var=0; ave1=ave2=O;
  for(j=0;j<=k;j++)
    cov[j]=0;

  for(i=1;i<n;i++)
    {
      *ave += (data[i] - data[0]);
      *var += (data[i] - data[0])*(data[i] - data[0]);
    }
  *var /= (double)(n-1);
  *var -= (((*ave)/(double)n) * ((*ave)/(double)(n-1)));
  /* *ave = ((*ave)/(double)n) + data[0]; (the true average; not needed)*/

  for(j=0;j<=k;j++)
    {
      for(i=0;i<(n-j);i++)
	{
	  ave1 += (data[i] - data[0]);
	  ave2 += (data[i+j] - data[0]);
	  cov[j] += (data[i] - data[0])*(data[i+j] - data[0]);

	}

      ave1/=(double)(n-j); ave2/=(double)(n-j);
      cov[j] = ((cov[j] - (double)(n-j)*ave1*ave2)/(double)(n-j-1));
    }

}

/*** end funccions.c ***/

\end{verbatim}
}

%%%-------------------------------------------------------------------
%%%
%%% BIBLIOGRAPHY
%%%

%% Note: If your thesis has more than one appendix, NYU requires a "list of
%% appendices" page before the body of the thesis. I don't provide the tools
%% to create that here, so you're on your own for that one... Sorry.
%\input{app2}
%%%% Input bibliography file %%%%%%%%%%%%%%%
%\input{biblio}

%% \bibliographystyle{plainurl}
%% \bibliography{wjd}
\def\cprime{$'$} \def\cprime{$'$}
  \def\ocirc#1{\ifmmode\setbox0=\hbox{$#1$}\dimen0=\ht0 \advance\dimen0
  by1pt\rlap{\hbox to\wd0{\hss\raise\dimen0
  \hbox{\hskip.2em$\scriptscriptstyle\circ$}\hss}}#1\else {\accent"17 #1}\fi}
\begin{thebibliography}{1}

\bibitem{Demmel:1997}
J.~Demmel.
\newblock {\em Applied Numerical Linear Algebra}.
\newblock Society for Industrial and Applied Mathematics, 1997.

\bibitem{Diaconis:1996}
P.~Diaconis and L.~Saloff-Coste.
\newblock Logarithmic sobolev inequalities for finite markov chains.
\newblock {\em The Annals of Applied Probability}, 6:695--750, 1996.

\bibitem{Durret:1996}
R.~Durrett.
\newblock {\em Probability: Theory and Examples}.
\newblock Duxbury Press, second edition, 1996.

\bibitem{Geman:1984}
S.~Geman and D.~Geman.
\newblock Stochastic relaxation. gibbs distributions and the bayesian
  restoration of images.
\newblock {\em IEEE Transactions on Pattern Analysis and Machine Intelligence},
  6:721--741, 1984.

\bibitem{Golub:1996}
G.~Golub and Charles Van~Loan.
\newblock {\em Matrix Computations}.
\newblock Johns Hopkins University Press, third edition, 1996.

\bibitem{GoodmanSokal:1989}
G.~Goodman and A.~Sokal.
\newblock Multigrid monte carlo method. conceptual foundations.
\newblock {\em Physical Review D}, 40:2035--2071, 1989.

\bibitem{Hammersley:1964}
J.~M. Hammersley and D.~C. Handscomb.
\newblock {\em Monte Carlo Methods}.
\newblock Chapman and Hall, 1964.

\bibitem{Metropolis:1953}
N.~Metropolis, A.~W. Rosenbluth, M.~N. Rosenbluth, A.~H. Teller, and E.~Teller.
\newblock Equations of state calculations by fast computing machines.
\newblock {\em J. Chem. Phys.}, 21:1087--1092, 1953.

\bibitem{Rosenthal:1995}
J.~Rosenthal.
\newblock Convergence rates for markov chains.
\newblock {\em Siam Review}, 37:387--405, 1995.

\end{thebibliography}

\end{document}

\end{document}

### 2013-10-20 msthesis.tex
minor mod
modified title page format to conform with original

@@
 \newcommand{\thesisadvisor}{Professor Jonathan Goodman}
-\newcommand{\graddate}{Jan 1999}
+\newcommand{\graddate}{January 1999}
 
@@
   {\large\textbf{\thesistitle}}
-  \vspace{.7in}
+
+  \vspace{5mm}
 
   by
-  \vspace{.7in}
+
+  \vspace{5mm}
 
   \thesisauthor
+
   \vfill
 
-\begin{doublespace}
-  A dissertation submitted in partial fulfillment\\
-  of the requirements for the degree of\\
-  Master of Science\\
-  Department of Mathematics\\
-  New York University\\
+%% \begin{doublespace}
+  A dissertation submitted in partial satisfaction of the\\[-4pt]
+  requirements for the degree of\\[-4pt]
+  Master of Science\\[6pt]
+  in\\[6pt]
+  Mathematics\\[6pt]
+  in the\\[6pt]
+  GRADUATE DIVISION\\[-4pt]
+  of the \\[-4pt]
+  NEW YORK UNIVERSITY\\
+%% \end{doublespace}
+\end{center}
+Committee in charge:\\
+\phantom{XXX} Professor Jonathan Goodman, Chair\\[-4pt]
+\phantom{XXX} Professor Leslie Greengard
+\begin{center}
   \graddate
-\end{doublespace}
 \end{center}

### 2013-10-20 msthesis.tex
fixed typos and spelling errors

@@
 %% anyone's help...
-\section*{Acknowledgements}\addcontentsline{toc}{section}{Acknowledgements}
+\section*{Acknowledgments}\addcontentsline{toc}{section}{Acknowledgments}
 %%% 
@@
 That is, for each $\psi_i$ there exists a vector
-$t$ of coeﬂicients such that $\M \psi_j = \sum_{i=1}^p t_i \psi_{ij}$. 
+$t$ of coefficients such that $\M \psi_j = \sum_{i=1}^p t_i \psi_{ij}$. 
 This is true for all $j$ and, putting each vector of
@@
 $\Psi = \Q \Z$, where $\Q^t\Q = \I$ and $\Z^2 = \Psi^t\Psi$ is a symmetric
-positive semidefinite matrix.\footnote{Recall that the polar decompostion is
+positive semidefinite matrix.\footnote{Recall that the polar decomposition is
   derived from the singular value decomposition, $\Psi = \U\Sigma \V^t$, by
@@
 (which comes from the definition of $\Psi$ following Equation~(\ref{eq:2.7})) 
-is such that $\psi_1$ is a constant vector if and only if $\phi$ is oonstant on the
+is such that $\psi_1$ is a constant vector if and only if $\phi$ is constant on the
 state space, in which case $\psi_1$ is the constant zero function. 
@@
            \caption{The Lanczos algorithm for partial reduction to symmetric
-             tridiaganal form.}
+             tridiagonal form.}
            \begin{algorithmic}
@@
 purposes, this construction is not essential. What is essential is that we find
-a way to generate the Lanczos coeficients $\alpha_i$, $\beta_j$, $(j = 1,\dots, p)$, and
+a way to generate the Lanczos coefficients $\alpha_i$, $\beta_j$, $(j = 1,\dots, p)$, and
 whence the matrix $\T$, without performing the operation $\M q_j$ required by the
@@
 To simplify notation, let $\phi(n) = \phi(X_n)$. 
-The form of the first Lanczos coeﬂicient $\alpha_1$ is straight
+The form of the first Lanczos coefficient $\alpha_1$ is straight
 forward:
@@
 
-Before generalizing this for the $j$th Lanczos coefficient, we consider one more coeﬂicient
+Before generalizing this for the $j$th Lanczos coefficient, we consider one more coefficient
 in detail. Recall that the second basis element $q_2$ in the Lanczos algorithm is given by 
@@
 The preceding paragraphs define all but one of the quantities in the last
-exression of~(\ref{eq:4.8}). The one remaining
+expression of~(\ref{eq:4.8}). The one remaining
 is $q_1^t\M^3 q_1$, which can also be expressed in terms of covariances of $\phi$,
@@
 \end{align*}
-To generalize the foregoing for higher order Lanczos coeﬂicients requires a recurrence
+To generalize the foregoing for higher order Lanczos coefficients requires a recurrence
 relation between $(\alpha_{j+1}, \beta_j)$ and $(\alpha_{j}, \beta_{j-1})$. 
@@
 \end{align}
-The point of all this is that each pair of coeﬂicients $(\alpha_{j+1}, \beta_j)$
+The point of all this is that each pair of coefficients $(\alpha_{j+1}, \beta_j)$
 can be expressed as a function of the covariance between 
@@
 \]
-where the sum runs over all neart neighbor pairs. In statistical mechanics one is often concerned
+where the sum runs over all nearest neighbor pairs. In statistical mechanics one is often concerned
 with the \emph{Gibbs measure} of state $i$, which is defined by
@@
 so that $\pi$ is a probability measure. 
-Now, letting $k$ denote the so called \emph{Boltzman constant}, if a
+Now, letting $k$ denote the so called \emph{Boltzmann constant}, if a
 classical mechanical system is in thermal equilibrium with its surroundings with
@@
 %
-number of desired Lauczos coefiicients. That is, the observable could have come from a simulation
+number of desired Lanczos coefficients. That is, the observable could have come from a simulation
 of any reversible Markov chain and not just our special Metropolis chain. The output of this
@@
 
-As montioned in Section~\ref{sec:gener-lancz-algor} (see
+As mentioned in Section~\ref{sec:gener-lancz-algor} (see
 Theorem~\ref{thm:4.1.1}), the $\beta$ coefficients provide conservative error 
@@
 by Matlab, both operations taking a few seconds. Granted, comparing a few seconds to the 20
-minutes it takes Matlab to compute all the eigenvales is not really fair, 
+minutes it takes Matlab to compute all the eigenvalues is not really fair, 
 since calling Matlab's {\tt eig()} routine is not always the
@@
 When these conditions hold, we can avoid the traditional Lanczos algorithm, and
-estimate the values of the Lanczos coeﬂicients via the variational properties of
+estimate the values of the Lanczos coefficients via the variational properties of
 observables on the state space. 
@@
 P.~Diaconis and L.~Saloff-Coste.
-\newblock Logarithmic sobolev inequalities for finite markov chains.
+\newblock Logarithmic {S}obolev inequalities for finite {M}arkov chains.
 \newblock {\em The Annals of Applied Probability}, 6:695--750, 1996.
@@
 S.~Geman and D.~Geman.
-\newblock Stochastic relaxation. gibbs distributions and the bayesian
+\newblock Stochastic relaxation. {G}ibbs distributions and the {B}ayesian
   restoration of images.
@@
 G.~Goodman and A.~Sokal.
-\newblock Multigrid monte carlo method. conceptual foundations.
+\newblock Multigrid {M}onte {C}arlo method. conceptual foundations.
 \newblock {\em Physical Review D}, 40:2035--2071, 1989.
@@
 J.~Rosenthal.
-\newblock Convergence rates for markov chains.
+\newblock Convergence rates for {M}arkov chains.
 \newblock {\em Siam Review}, 37:387--405, 1995.

### 2013-10-22 msthesis.tex
fixed various typos

@@
 \newtheorem*{problem}{Problem}
-\newtheorem*{fact}{Fact}
+\newtheorem{fact}{Fact}[chapter]
 \newtheorem{case}{Case}
@@
 %Definition 2.1.1 
-Let $(\S, \S)$ be a measurable space. A sequence $X_n$, $n\geq 0$, of random variables
-taking values in \S\ is said to be a Markov chain with respect to the filtration 
+Let $(S, \S)$ be a measurable space. A sequence $X_n$, $n\geq 0$, of random variables
+taking values in $S$ is said to be a Markov chain with respect to the filtration 
 $\sigma(X_0, \dots, X_{n})$ if for all $B \in \S$,
@@
 \label{eq:2.2}
-P(X_{n+1} \in B \mid  \sigma(X_n)) = p_n(X_n, B)
+P(X_{n+1} \in B \mid  \sigma(X_n)) = p_n(X_n, B).
 \end{equation}
@@
 \[
-p(i,A) = \sigma_{j\in A} p(i,j).
+p(i,A) = \sum_{j\in A} p(i,j).
 \]
@@
 The final two conditions we place on the Markov chains considered below will cost us
-some generality. Nonetheless, there remain many examples of chains meeting these conditions
+some generality. Nonetheless, there are many examples of chains meeting these conditions
 and making the present study worthwhile. Furthermore, it may be the case that, with a little
@@
 eigenvalue which is second largest in absolute value, and we denote this
-eigenvalue by $\lambda_{\max} = \max{\lambda_1, |\lambda_{d-1}|}$. 
+eigenvalue by $\lambda_{\max} = \max\{\lambda_1, |\lambda_{d-1}|\}$. 
 The function bounding the rate of convergence of a Markov chain is given by the
@@
 \begin{align*}
-\C(0) &= \E(\Phi(X_n) - \<\Phi\>)(\Phi(X_n) - \<\Phi\>)^t\\
+\C(0) &= \bE(\Phi(X_n) - \<\Phi\>)(\Phi(X_n) - \<\Phi\>)^t\\
 &= (\Phi - \<\Phi\>)^t\Pi (\Phi-\<\Phi\>), \\
-\C(s) &= \E(\Phi(X_n) - \<\Phi\>)(\Phi(X_{n+s}) - \<\Phi\>)^t\\
+\C(s) &= \bE(\Phi(X_n) - \<\Phi\>)(\Phi(X_{n+s}) - \<\Phi\>)^t\\
 &= (\Phi - \<\Phi\>)^t \Pi \P^s (\Phi - \<\Phi\>).
@@
 According to Definition~\ref{def:2.1.3}, this states
-that $p$ is a reversible measure.
+that $\pi$ is a reversible measure.
 \end{proof}
@@
 %% Definition 3.1.1
-A subspace $S \subseteq \R^n$ with the property that
-\[
-x \in S \quad \Longrightarrow \quad \M x \in S
+A subspace $V \subseteq \R^n$ with the property that
+\[
+v\in V \quad \Longrightarrow \quad \M v \in V
 \]
@@
 \M \Psi v = \lambda \Psi v\\
-& \quad \Longleftrightarrow  \quad
+& \quad \Longrightarrow  \quad
 \lambda \in \lambda(M).
@@
 $\T \in \R^{p\times p}$, then
-$\ran(\Psi)$ defines an \emph{approximate invariant subspace}. 
+$\ran(\Psi)$ defines an \emph{approximate invariant subspace} for $\M$. 
 \end{definition}
@@
 \[
-|\mu_k - \lambda_k| \leq \sqrt{2}\|\E\|_2, \quad \text{ for $k = 1, \dots, p$.}
+|\mu_k - \lambda_k| \leq \sqrt{2}\|\E\|_2, \quad \text{ for } \quad k = 1, \dots, p.
 \]
@@
 
-Recall the matrix $\Psi = \Pi^{1/2}(\Phi — \<\Phi\>)$. The columns of this
+Recall the matrix $\Psi = \Pi^{1/2}(\Phi - \<\Phi\>)$. The columns of this
 matrix, though independent (by choice of independent observables), are not
@@
 in detail. Recall that the second basis element $q_2$ in the Lanczos algorithm is given by 
-$q_2 = (\M q_1 — \alpha_1 q_1)/\beta_1$. 
+$q_2 = (\M q_1 - \alpha_1 q_1)/\beta_1$. 
 From this we can derive an expression for $\alpha_2$ which, while not as neat as those
@@
 %% until we have data from the stationary distribution,
-%% and we are back where we started — almost.
+%% and we are back where we started - almost.
 When employing our procedure, we can discard a
@@
   X(1:2:d-1)=1;
-  X(2:2:d)=—1;
+  X(2:2:d)=-1;
 end
@@
       (form1(j-1,k+1) - alpha(j-1)*form1(j-1,k)
-       - sqrt(betasq(j-2)) * form2(j—1,k));
+       - sqrt(betasq(j-2)) * form2(j-1,k));
 }

### 2013-10-22 src/functions.c

@@ -62,7 +62,7 @@ double form1(int j, int k)
 	* ( form1(j-1,k+2) + pow(alpha(j-1),2) * form1(j-1,k)
 	    + 2*(alpha(j-1) * sqrt(betasq(j-2)) * form2(j-1,k)
 		 - alpha(j-1)*form1(j-1,k+1)
-		 - sqrt(betasq(j—2))*form2(j-1,k+1) )
+		 - sqrt(betasq(j-2))*form2(j-1,k+1) )
 	    + betasq(j-2)*form1(j-2,k)):
     }
 }
@@ -77,7 +77,7 @@ double form2(int j, int k)
     return
       (pov(betasq(j-1),-.5)) 3
       (form1(j-1,k+1) - alpha(j-1)*form1(j-1,k)
-       - sqrt(betasq(j-2)) * form2(j—1,k));
+       - sqrt(betasq(j-2)) * form2(j-1,k));
 }
 
 /* moment() function for computing var and cov(k)

### 2013-10-22 src/ising.m

@@ -25,7 +25,7 @@ end
 % Initialize at state of high energy (not used)
 if 0
   X(1:2:d-1)=1;
-  X(2:2:d)=—1;
+  X(2:2:d)=-1;
 end
 
 % Initialize energy

### 2013-10-22 msthesis.tex
fixed typos

@@
 and it is clear that $\ran(\Q) = \ran(\Psi)$. 
-Perhaps $\ran(\Q)$ is a useful approximation to the invariant
-subspace for $\M$. If $\ran(Q)$ is not itself invariant, we have the error 
+Perhaps $\ran(\Q)$ is a useful approximation to an invariant
+subspace for $\M$. If $\ran(\Q)$ is not itself invariant, we have the error 
 matrix $\E =\M\Q - \Q\T$.  Below we show that the
@@
 %
-The claim is verified by an easy application of the Pythagorean theorem:
+The claim is verified by an easy application of the Pythagorean Theorem:
 For any $\T \in \R^{p\times p}$ we have
@@
 subspace orthogonal to 
-$\ran(Q)$. Thus, the two terms in the expression on the right are orthogonal,
+$\ran(\Q)$. Thus, the two terms in the expression on the right are orthogonal,
 and the Pythagorean theorem yields equality. The concluding inequality
@@
   \label{eq:3.2}
-|\rho(\psi_1,\M)| \leq \max |\rho(x,\M)| = \lambda_{\max}(\M)
+|\rho(\psi_1)| \leq \max |\rho(x)| = \lambda_{\max}(\M)
 \end{equation}
@@
 \label{eq:3.3}
-\max_{a\neq 0}|\rho(\Psi_2a,\M)| &= \max_{a\neq 0} \left|\frac{a^t \Psi^t_2 \M \Psi_2 a}{a^t \Psi^t_2 \Psi_2 a}\right|\nonumber\\
-&= \max_{x\in \ran(\Psi_2)}|\rho(x,\M)|\nonumber\\
-&= \max|\rho(x,\M)|\\
+\max_{a\neq 0}|\rho(\Psi_2a)| &= \max_{a\neq 0} \left|\frac{a^t \Psi^t_2 \M \Psi_2 a}{a^t \Psi^t_2 \Psi_2 a}\right|\nonumber\\
+&= \max_{x\in \ran(\Psi_2)}|\rho(x)|\nonumber\\
+&= \max|\rho(x)|\\
 &= \lambdamax\nonumber
@@
 $\Psi_2 = [\psi_1, \psi_2]$ so that there is a vector $a \in \R^2$ making 
-$|\rho(\Psi_2 a,\M)|$ close to $\lambdamax$.
+$|\rho(\Psi_2 a)|$ close to $\lambdamax$.
 
@@
 where $e_j$ is the column vector with 1 in
-the $j$th position and zeros elsewhere. We define $H = [e_2,\dots,e_d,-h]$,
-so~(\ref{eq:10000}) becomes $\M\Psi = \Psi H$.
+the $j$th position and zeros elsewhere. We define 
+$\H = [e_2,\dots,e_d,-h]$,
+so~(\ref{eq:10000}) becomes $\M\Psi = \Psi \H$.
 %
@@
   \label{eq:4.1}
-H = 
+\H = 
 \begin{pmatrix}
@@
          \end{algorithm}
+
 The $q_j$ computed by the Lanczos algorithm are often called the \emph{Lanczos vectors}. If the
@@
 \begin{align*}
-  q_{j+1}^tM^kq_j &= 
+  q_{j+1}^t\M^kq_j &= 
  \frac{1}{\beta_j}(\M q_j - \alpha_j q_j - \beta_{j-1} q_{j-1})^t\M^k q_j\\
- \frac{1}{\beta_j}(q_j^t\M^{k+1} q_j - \alpha_j q_j^t\M^{k} q_j -\beta_{j-1} q_j^t\M^k).
+&=  \frac{1}{\beta_j}(q_j^t\M^{k+1} q_j - \alpha_j q_j^t\M^{k} q_j -\beta_{j-1} q_j^t\M^k).
 \end{align*}
@@
 \label{eq:4.12}
-  q_{j+1}^tM^kq_{j+1} 
+  q_{j+1}^t\M^kq_{j+1} 
 &=  \frac{1}{\beta_j^2}(\M q_j - \alpha_j q_j - \beta_{j-1} q_{j-1})^t\M^k 

### 2013-10-26 lit/MCMC.tex
initial commit of some background articles

\documentclass{article}

\begin{document}

\begin{center} 
Lecture Notes on Monte Carlo Methods \\
Fall Semester, 2005 \\
Courant Institute of Mathematical Sciences, NYU \\
Jonathan Goodman, goodman@cims.nyu.edu \\ 
\vspace{.6cm}
\large 
Chapter 2:  Dynamic sampling and Markov chain Monte Carlo.
\normalsize
\end{center}

A variety of techniques collectively 
called\footnote{The practice predates the name by several decades.} 
{\em Markov chain Monte Carlo} (MCMC) or {\em dynamic sampling} allow sampling of 
comlex high dimensional distributions not accessablt by simple samplers.
With technical ideas to follow, the rough idea is that if $f$ is the invariant law 
of a nondegenerate Markov chain $X(t)$, then the law of $X(t)$ converges to $f$ as 
$t \to \infty$.
We do not have a direct simple way to sample $f$, so we choose a starting state 
$X(0)$ in an arbitrary way and {\em run} the chain.
Regardless of the distribution of $X(0)$, the law of $X(t)$ will converge to $f$
as $t\to\infty$.
Moreover, for large enough $s$, the samples $X(t)$ and $X(t+s)$ ``decorrelate''
(become independent, actually) so as to become effectively distinct samples of $f$.
Therefore we can estimate $A = E_f[V(X)]$ using
\begin{equation}
\widehat{A} = \frac{1}{n} \sum_{t=1}^n V(X(t)) \; .
\end{equation}
Regardless of the choice of $X(0)$, we have $\widehat{A} \to A$ as $n \to\infty$.
This is called the {\em ergodic theorem} for Markov chains, in anology to ergodic
theorems in statistical mechanics.
It is remarkable that there are many distributions that are impractical to sample
directly but easy (or at least possible) to sample with MCMC.

A rejection sampler is inefficient if its acceptence probability is low.
A dynamic sampler can be inefficient because of large long lasting correlations
between $X(t)$.
The {\em autocorrelation time} is the minimum $s$ before $X(t)$ and $X(t+s)$ 
are effectively independent samples from the point of view of error bars.
A {\em rapidly mixing} chain has small autocorrelation time.
There are tricks for inventing rapidly mixing chains in various situations
(e.g. Multigrid Monte Carlo, many more in Alan Sokal's notes or Jun Liu's book).
We will discuss error bars and autocorrelation time estimates for dynamic Monte Carlo.

\end{document}
